### 2020-01-26 ms/ms.tex
some text from ISF

\documentclass[12pt]{extarticle}
\usepackage{geometry}
\geometry{
a4paper,
total={170mm,257mm},
left=20mm,
top=20mm,
headheight=12pt
}

%\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx} % Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
% TeX will automatically convert eps --> pdf in pdflatex

\usepackage{amssymb,amsmath,amsthm}
\usepackage{commath}
\usepackage{longtable}
\usepackage[hyphens]{url}
\usepackage[dvipsnames]{xcolor}
\usepackage[unicode=true,colorlinks=true,urlcolor=CadetBlue,citecolor=black,linkcolor=black]{hyperref}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage{authblk}

%SetFonts
% newtxtext+newtxmath
\usepackage{newtxtext} %loads helv for ss, txtt for tt
\usepackage{amsmath}
\usepackage[bigdelims]{newtxmath}
\usepackage[T1]{fontenc}
\usepackage{textcomp}
%SetFonts

% less space before sections 
% \@startsection {NAME}{LEVEL}{INDENT}{BEFORESKIP}{AFTERSKIP}{STYLE} 
%            optional * [ALTHEADING]{HEADING} 
\makeatletter
 \renewcommand\section{\@startsection {section}{1}{\z@}%
     {-2.5ex \@plus -1ex \@minus -.2ex}%
     {1.3ex \@plus.2ex}%
    {\Large\bfseries}}

% Species names
%% Meta-Command for defining new species macros
\usepackage{xspace}

\newcommand{\species}[3]{%
  \newcommand{#1}{\gdef#1{\textit{#3}\xspace}\textit{#2}\xspace}}

\species{\yeast}{Saccharomyces cerevisiae}{S.~cerevisiae}
\species{\calbicans}{Candida albicans}{C.~albicans}
\species{\cneoformans}{Cryptococcus neoformans}{C.~neoformans}

% Yoav & Lee commands
\newcommand*{\tr}{^\intercal}
\let\vec\mathbf
\newcommand{\matrx}[1]{{\left[ \stackrel{}{#1}\right]}}
\newcommand{\diag}[1]{\mbox{diag}\matrx{#1}}
\newcommand{\goesto}{\rightarrow}
\newcommand{\dspfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2} }
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{remark}{Remark}
\newtheorem{result}{Result}
\renewcommand\qedsymbol{} % no square at end of proof
\newcommand{\cl}{\mathbf{L}}
\newcommand{\cj}{\mathbf{J}}
\newcommand{\ci}{I}

% NatBib
\usepackage[round,colon,authoryear]{natbib}

% Title page
% Chromosomal duplication is a transient evolutionary solution to stress
\title{Interference between chromosome duplication and mutations: a computational analysis}

\author[a]{Ilia Kohanovski}
\author[b,c]{Avihu H. Yona}
\author[d]{Orna Dahan}
\author[d]{Yitzhak Pilpel}
\author[a,*]{Yoav Ram}

\affil[a]{School of Computer Sciences, IDC Herzliya, P.O. Box 167, Herzliya, 4610101, Israel}
\affil[b]{Physics of Living Systems, Department of Physics, Massachusetts Institute of Technology, Cambridge, MA 02139, USA}{
\affil[b]{Department of Biological Engineering, Massachusetts Institute of Technology, Cambridge, MA 02139, USA}
\affil[d]{Department of Molecular Genetics, Weizmann Institute of Science, Rehovot 76100 Israel}
\affil[*]{Corresponding author: yoav@yoavram.com}

% Document
\begin{document}
\maketitle

% Abstract
\begin{abstract}
Aneuploidy is common in eukaryotes, often leading to decreased cell growth and fitness. However, evidence from yeast and fungi, as well as tumour cells, suggests that aneuploidy can be beneficial under stressful conditions and  lead to elevated growth rates and to adaptation. Thus, it is crucial to develop a quantitative theory for the role of aneuploidy in adaptive evolution. Here, we analyse results from experiments in which \yeast adapted to heat stress. The experimental population first acquired chromosome duplications, only to later revert back to a euploid state. We use several evolutionary models within a approximate Bayesian computation framework to show that clonal interference between chromosome duplications and beneficial mutations explains the experimental results. Our results suggest that aneuploidy can both accelerate and deccelerate adaptation in a non-intuitive manner, creating an evolutionary conflict between the individual and the population and further complicating the process of adaptation in cell populations.
\end{abstract}

\pagebreak
% Introduction
\section*{Introduction}

Aneuploidy is an imbalance in the number of chromosomes in the cell: an incorrect karyotype.
Recent evidence suggests aneuploidy is very common in eukaryotes, e.g. animals~\citep{Santaguida2015, Naylor2016, Bakhoum2017}, and fungi~\citep{Zhu2016, Pavelka2010, Robbins2017, Todd2017}.
Aneuploidy has been implicated in cancer formation and progression~\citep{Schvartzman2010, Boveri2008}: 
90\% of solid tumours and 50\% of blood cancers are aneuploid~\citep{Santaguida2015}.
It is also linked to the emergence of drug resistance~\citep{Selmecki2009} and virulence~\citep{Moller2018} in fungal pathogens, which are under-studied~\citep{Rodrigues2018} despite infecting close to a billion people per year, causing serious infections and significant morbidity in >150 million people per year and killing >1.5 million people per year~\citep{Selmecki2009, Rodrigues2018}.
In addition, aneuploidy is common in protozoan pathogens of the Leishmania genus, a major global health concern~\citep{Mannaert2012}.

The molecular and genetic mechanisms involved in aneuploidy have been explored in the past decade~\citep{Chen2012b, Rancati2013, Gerstein2015, Shor2015, Sheltzer2011, Musacchio2007}.
Experiments with human and mouse embryos found that aneuploidy is usually lethal;
it is also associated with developmental defects and lethality in other multicellular organisms~\citep{Sheltzer2011} for example, aneuploid mouse embryonic cells grew slower than euploid cells20.
Similarly, in unicellular eukaryotes growing in benign conditions, aneuploidy usually leads to slower growth and decreased overall ~\citep{Sheltzer2011, Pavelka2010, Torres2007, Kasuga2016, Niwa2006}, in part due to proteotoxic stress caused by increased expression in aneuploid cells~\citep{Pavelka2010, Santaguida2015, Zhu2018}.
However, aneuploidy can be beneficial under stressful conditions due to the wide range of phenotypes it can generate, some of which are advantageous~\citep{Pavelka2010}.
Thus, aneuploidy can lead to rapid adaptation in unicellular eukaryotes~\citep{Gerstein2015,Torres2010, Hong2014, Rancati2008}, as well as to rapid growth of somatic tumour cells~\citep{Schvartzman2010, Sheltzer2017}.
For example, aneuploidy in \yeast facilitates adaptation to a variety of stress conditions like heat~\citep{Yona2012}, pH~\citep{Yona2012}, copper~\citep{Covo2014, Gerstein2015}, salt~\citep{Dhar2011}, and nutrient-limitation~\citep{Dunham2002, Gresham2008}.
Importantly, aneuploidy also leads to drug resistance in pathogenic fungi such as \calbicans~\citep{Selmecki2010, Selmecki2008, Gerstein2018} and \cneoformans~\citep{Sionov2010}, which cause candidiasis and meningoencephalitis, respectively.

The role of aneuploidy in adaptation has been observed only recently~\citep{Yona2012, Gerstein2015, Sionov2010}, and is largely missing from the literature on evolution and adaptation:
the introductory textbook Evolution by Bergstrom and Dugatkin~\citet{Bergstrom2012} does not mention the word aneuploidy, and the graduate-level book Mutation-Driven Evolution by Nei~\citet{Nei2013} only briefly mentions aneuploidy in the context of speciation, but not adaptation.
In recent reviews of the literature, aneuploidy is suggested to play an important role in fungal adaptation6,7 and cancer evolution~\citep{Santaguida2015, Naylor2016, Sansregret2017}, yet these reviews cite no theoretical studies nor quantitative evolutionary models.
Indeed, evolutionary, ecological, and epidemiological studies mostly assume adaptation occurs via beneficial mutations, recombination, and sex.
Therefore, there is a critical need to develop an evolutionary theory of aneuploidy, like the evolutionary theories of other mechanisms that generate genetic variation: mutation~\citep{Lynch2010}, recombination~\citep{Hartfield2012}, and sex~\citep{Otto2009}.
An evolutionary theory of aneuploidy will be central to the interpretation of experimental and clinical observations and design of new hypotheses, experiments, and treatments~\citep{Carja2014}.
For example, despite the lack of theoretical models, aneuploidy has been invoked in a new strategy to combat pathogens and tumour cells by setting “evolutionary traps”~\citep{Gerstein2015,Chen2015}, in which a condition that predictably leads to emergence of aneuploidy is applied, followed by a condition that specifically selects against aneuploid cells.

The published data indicate that, like mutation, aneuploidy can be both deleterious and beneficial~\citep{Pavelka2010, Sheltzer2011}.
Nevertheless, there are important and fundamental differences between adaptation by aneuploidy
and adaptation by beneficial mutations~\citep{Yona2015}, which make aneuploidy a unique mechanism for generating genetic
variation.
First, the aneuploidy rate (i.e. the frequency of mis-segregation events) is significantly higher than the
mutation rate~\citep{Santaguida2015}.
Thus, everything else being equal, adaptation by aneuploidy will be faster and more frequent.
Second, fitness effects of aneuploidy are larger than those of the majority of mutations, on average, and are rarely
neutral~\citep{Pavelka2010, Yona2012, Sunshine2015}, allowing selection to quickly sort deleterious and beneficial genotypes.
Third, the number of different karyotypes is considerably smaller than the number of different genotypes, and different karyotypes are likely to have different phenotypes~\citep{Pavelka2010}.
Therefore, exploration of the phenotype space by aneuploidy requires smaller populations and a shorter time span.
Fourth, aneuploidy is a reversible state, as the rate of chromosome loss is high and the cost of aneuploidy is significant~\citep{Niwa2006}.
Indeed, aneuploidy often provides a transient solution: under short-term stress conditions, aneuploidy reverts (chromosome number returns to normal) when the stress subsides; under long-term stress conditions, aneuploidy reverts when refined solutions, generated by beneficial mutations, take over~\citep{Yona2012}.
Finally, aneuploidy results in increased genome instability, potentially increasing genetic variation by a positive feedback loop~\citep{Rancati2013, Bouchonville2009, Zhu2012}, while also increasing its own transience.

Here, we develop a series of evolutionary models that consider both mutation and aneuploidy.
These models follow a population of cells characterised by both by their ploidy and by their genotype.
We use these models to examine the role of aneuploidy as a transient adaptive solution~\citep{Yona2015} (\textbf{Fig.~1}) in evolutionary experiments in \yeast adapting to heat and pH stress~\citep{Yona2012}.
Aneuploidy is transient due to its distinct properties mentioned above: aneuploidy usually occurs before mutation, often produces significant improvements in fitness, and yet, because it affects many genes on a chromosome or chromosome fragment, also has fitness costs.
Thus, it is rapidly lost when stress is removed or after beneficial mutations finally occur, because the costs then outweigh any incremental advantage over the mutation.
We use these models to test the hypothesis that aneuploidy accelerates adaptation.

\pagebreak
% Models and Methods
\section*{Models and Methods}

\pagebreak
% Results
\section*{Results}

\pagebreak
% Discussion
\section*{Discussion}

\pagebreak
% Acknowledgements
{\small
\section*{Acknowledgements}
We thank Lilach Hadany, Judith Berman, David Gresham, and Martin Kupiec for discussions and comments.
This work was supported in part by 
the Israel Science Foundation (YR 552/19).
}

\bibliographystyle{agsm}
\bibliography{/Users/yoavram/library.bib}

\end{document}

### 2020-01-26 ms/ms.tex
moved parts of intro to discussion; separated intro to paragraphs; added 'this study' paragraph

@@
 % Chromosomal duplication is a transient evolutionary solution to stress
-\title{Interference between chromosome duplication and mutations: a computational analysis}
+\title{Clonal interference between aneuploidy and mutation}
+
+% Authors
+\renewcommand\Affilfont{\small}
 
@@
 
-\affil[a]{School of Computer Sciences, IDC Herzliya, P.O. Box 167, Herzliya, 4610101, Israel}
-\affil[b]{Physics of Living Systems, Department of Physics, Massachusetts Institute of Technology, Cambridge, MA 02139, USA}{
+\affil[a]{School of Computer Sciences, IDC Herzliya, Herzliya 4610101, Israel}
+\affil[b]{Department of Physics, Massachusetts Institute of Technology, Cambridge, MA 02139, USA}{
 \affil[b]{Department of Biological Engineering, Massachusetts Institute of Technology, Cambridge, MA 02139, USA}
-\affil[d]{Department of Molecular Genetics, Weizmann Institute of Science, Rehovot 76100 Israel}
+\affil[d]{Department of Molecular Genetics, Weizmann Institute of Science, Rehovot 76100, Israel}
 \affil[*]{Corresponding author: yoav@yoavram.com}
@@
 
+\paragraph*{Aneuploidy is common in eukaryotes.}
 Aneuploidy is an imbalance in the number of chromosomes in the cell: an incorrect karyotype.
-Recent evidence suggests aneuploidy is very common in eukaryotes, e.g. animals~\citep{Santaguida2015, Naylor2016, Bakhoum2017}, and fungi~\citep{Zhu2016, Pavelka2010, Robbins2017, Todd2017}.
-Aneuploidy has been implicated in cancer formation and progression~\citep{Schvartzman2010, Boveri2008}: 
-90\% of solid tumours and 50\% of blood cancers are aneuploid~\citep{Santaguida2015}.
-It is also linked to the emergence of drug resistance~\citep{Selmecki2009} and virulence~\citep{Moller2018} in fungal pathogens, which are under-studied~\citep{Rodrigues2018} despite infecting close to a billion people per year, causing serious infections and significant morbidity in >150 million people per year and killing >1.5 million people per year~\citep{Selmecki2009, Rodrigues2018}.
+Recent evidence suggests aneuploidy is very common in eukaryotes, e.g. animals~\citep{Santaguida2015review, Naylor2016, Bakhoum2017}, and fungi~\citep{Pavelka2010, Zhu2016, Robbins2017, Todd2017}.
+Aneuploidy has been implicated in cancer formation and progression~\citep{Boveri2008, Schvartzman2010}: 
+90\% of solid tumours and 50\% of blood cancers are aneuploid~\citep{Santaguida2015review}.
+Aneuploidy is also linked to the emergence of drug resistance~\citep{Selmecki2009} and virulence~\citep{Moller2018} in fungal pathogens, which are under-studied~\citep{Rodrigues2018} despite infecting close to a billion people per year, causing serious infections and significant morbidity in >150 million people per year and killing >1.5 million people per year~\citep{Selmecki2009, Rodrigues2018}.
 In addition, aneuploidy is common in protozoan pathogens of the Leishmania genus, a major global health concern~\citep{Mannaert2012}.
 
-The molecular and genetic mechanisms involved in aneuploidy have been explored in the past decade~\citep{Chen2012b, Rancati2013, Gerstein2015, Shor2015, Sheltzer2011, Musacchio2007}.
+\paragraph*{Aneuploidy is generally deleterious.}
+The molecular and genetic mechanisms involved in aneuploidy have been explored in the past decade~\citep{Musacchio2007, Sheltzer2011, Chen2012b, Rancati2013, Gerstein2015, Shor2015}.
 Experiments with human and mouse embryos found that aneuploidy is usually lethal;
-it is also associated with developmental defects and lethality in other multicellular organisms~\citep{Sheltzer2011} for example, aneuploid mouse embryonic cells grew slower than euploid cells20.
-Similarly, in unicellular eukaryotes growing in benign conditions, aneuploidy usually leads to slower growth and decreased overall ~\citep{Sheltzer2011, Pavelka2010, Torres2007, Kasuga2016, Niwa2006}, in part due to proteotoxic stress caused by increased expression in aneuploid cells~\citep{Pavelka2010, Santaguida2015, Zhu2018}.
-However, aneuploidy can be beneficial under stressful conditions due to the wide range of phenotypes it can generate, some of which are advantageous~\citep{Pavelka2010}.
+it is also associated with developmental defects and lethality in other multicellular organisms~\citep{Sheltzer2011}.
+For example, aneuploid mouse embryonic cells grow slower than euploid cells~\citep{Williams2008}.
+Similarly, in unicellular eukaryotes growing in benign conditions, aneuploidy usually leads to slower growth and decreased overall fitness~\citep{Niwa2006, Torres2007, Pavelka2010, Sheltzer2011, Kasuga2016}, in part due to proteotoxic stress caused by increased expression in aneuploid cells~\citep{Pavelka2010, Santaguida2015, Zhu2018} and hypo-osmotic-like stress~\citep{Tsai2019}.
+
+\paragraph*{Aneuploidy can lead to adaptation.}
+However, aneuploidy can be beneficial under stressful conditions due to the wide range of phenotypes it can produce, some of which are advantageous~\citep{Pavelka2010}.
 Thus, aneuploidy can lead to rapid adaptation in unicellular eukaryotes~\citep{Gerstein2015,Torres2010, Hong2014, Rancati2008}, as well as to rapid growth of somatic tumour cells~\citep{Schvartzman2010, Sheltzer2017}.
-For example, aneuploidy in \yeast facilitates adaptation to a variety of stress conditions like heat~\citep{Yona2012}, pH~\citep{Yona2012}, copper~\citep{Covo2014, Gerstein2015}, salt~\citep{Dhar2011}, and nutrient-limitation~\citep{Dunham2002, Gresham2008}.
-Importantly, aneuploidy also leads to drug resistance in pathogenic fungi such as \calbicans~\citep{Selmecki2010, Selmecki2008, Gerstein2018} and \cneoformans~\citep{Sionov2010}, which cause candidiasis and meningoencephalitis, respectively.
+For example, aneuploidy in \yeast facilitates adaptation to a variety of stressful conditions like heat and pH~\citep{Yona2012}, copper~\citep{Covo2014, Gerstein2015}, salt~\citep{Dhar2011}, and nutrient limitation~\citep{Dunham2002, Gresham2008}.
+Importantly, aneuploidy can also lead to drug resistance in pathogenic fungi such as \calbicans~\citep{Selmecki2008, Selmecki2010, Gerstein2018} and \cneoformans~\citep{Sionov2010}, which cause candidiasis and meningoencephalitis, respectively.
 
-The role of aneuploidy in adaptation has been observed only recently~\citep{Yona2012, Gerstein2015, Sionov2010}, and is largely missing from the literature on evolution and adaptation:
-the introductory textbook Evolution by Bergstrom and Dugatkin~\citet{Bergstrom2012} does not mention the word aneuploidy, and the graduate-level book Mutation-Driven Evolution by Nei~\citet{Nei2013} only briefly mentions aneuploidy in the context of speciation, but not adaptation.
-In recent reviews of the literature, aneuploidy is suggested to play an important role in fungal adaptation6,7 and cancer evolution~\citep{Santaguida2015, Naylor2016, Sansregret2017}, yet these reviews cite no theoretical studies nor quantitative evolutionary models.
-Indeed, evolutionary, ecological, and epidemiological studies mostly assume adaptation occurs via beneficial mutations, recombination, and sex.
-Therefore, there is a critical need to develop an evolutionary theory of aneuploidy, like the evolutionary theories of other mechanisms that generate genetic variation: mutation~\citep{Lynch2010}, recombination~\citep{Hartfield2012}, and sex~\citep{Otto2009}.
-An evolutionary theory of aneuploidy will be central to the interpretation of experimental and clinical observations and design of new hypotheses, experiments, and treatments~\citep{Carja2014}.
-For example, despite the lack of theoretical models, aneuploidy has been invoked in a new strategy to combat pathogens and tumour cells by setting “evolutionary traps”~\citep{Gerstein2015,Chen2015}, in which a condition that predictably leads to emergence of aneuploidy is applied, followed by a condition that specifically selects against aneuploid cells.
+\paragraph*{Transient adaptive solution.} 
+Aneuploidy differs from mutation due to its distinct properties. 
+Chromosome duplication usually occurs more often than mutation and on average produces larger fitness effects.
+Yet, because it affects many genes on a whole chromosome or a chromosome fragment, aneuploidy also carries fitness costs.
+Thus, aneuploidy can be a \emph{transient adaptive solution}: it can rapidly occur and fix in the population under stressful conditions, and can be rapidly lost when the cost outweighs the benefit -- when stress is removed or after beneficial mutations occur.
+Experimental evidence of such a transient role of aneuploidy was demonstrated by \citet{Yona2012}, who evolved populations of \yeast under strong heat or pH stress.
+In these experiments, the populations quickly adapted to the stress, and this adaptation was determined to be due to chromosome duplications.
+Later on, the populations reverted back to an euploid state, while remaining adapted to the stress and accumulating multiple mutations.
+However, under gradual heat stress, aneuploidy was not observed.
+\citet{Yona2012} concluded that aneuploidy serves as a transient adaptive solution, or a ``quick fix'', which is expected to facilitate adaptation.
 
-The published data indicate that, like mutation, aneuploidy can be both deleterious and beneficial~\citep{Pavelka2010, Sheltzer2011}.
-Nevertheless, there are important and fundamental differences between adaptation by aneuploidy
-and adaptation by beneficial mutations~\citep{Yona2015}, which make aneuploidy a unique mechanism for generating genetic
-variation.
-First, the aneuploidy rate (i.e. the frequency of mis-segregation events) is significantly higher than the
-mutation rate~\citep{Santaguida2015}.
-Thus, everything else being equal, adaptation by aneuploidy will be faster and more frequent.
-Second, fitness effects of aneuploidy are larger than those of the majority of mutations, on average, and are rarely
-neutral~\citep{Pavelka2010, Yona2012, Sunshine2015}, allowing selection to quickly sort deleterious and beneficial genotypes.
-Third, the number of different karyotypes is considerably smaller than the number of different genotypes, and different karyotypes are likely to have different phenotypes~\citep{Pavelka2010}.
-Therefore, exploration of the phenotype space by aneuploidy requires smaller populations and a shorter time span.
-Fourth, aneuploidy is a reversible state, as the rate of chromosome loss is high and the cost of aneuploidy is significant~\citep{Niwa2006}.
-Indeed, aneuploidy often provides a transient solution: under short-term stress conditions, aneuploidy reverts (chromosome number returns to normal) when the stress subsides; under long-term stress conditions, aneuploidy reverts when refined solutions, generated by beneficial mutations, take over~\citep{Yona2012}.
-Finally, aneuploidy results in increased genome instability, potentially increasing genetic variation by a positive feedback loop~\citep{Rancati2013, Bouchonville2009, Zhu2012}, while also increasing its own transience.
-
-Here, we develop a series of evolutionary models that consider both mutation and aneuploidy.
-These models follow a population of cells characterised by both by their ploidy and by their genotype.
-We use these models to examine the role of aneuploidy as a transient adaptive solution~\citep{Yona2015} (\textbf{Fig.~1}) in evolutionary experiments in \yeast adapting to heat and pH stress~\citep{Yona2012}.
-Aneuploidy is transient due to its distinct properties mentioned above: aneuploidy usually occurs before mutation, often produces significant improvements in fitness, and yet, because it affects many genes on a chromosome or chromosome fragment, also has fitness costs.
-Thus, it is rapidly lost when stress is removed or after beneficial mutations finally occur, because the costs then outweigh any incremental advantage over the mutation.
-We use these models to test the hypothesis that aneuploidy accelerates adaptation.
+\paragraph*{The present study.}
+Here, we develop a series of computational evolutionary models that include the effects of natural selection, genetic drift, aneuploidy and mutation to examine the role of aneuploidy in adaptation.
+These models follow a population of cells characterised by both their ploidy and their genotype.
+We fit these models to the empirical results of \citet{Yona2012} using approximate Bayesian computation.
+These model fits allow us to infer model parameters, including selection coefficients and transition rates, and to perform model selection between different models, thereby testing different hypotheses about the evolutionary process.
+We find that % TODO CHECK!
+the aneuploidy rate is several orders of magnitude higher than the mutation rate; that a simple model of clonal interference is not enough to explain the transience of aneuploidy; and that aneuploidy rescues the population, albeit at the price of delaying further adaptation.
 
@@
 \section*{Discussion}
+
+\paragraph*{Aneuploidy and mutation: \textit{same-same but different}.}
+The published data indicate that, like mutation, aneuploidy can be both deleterious and beneficial~\citep{Pavelka2010, Sheltzer2011}.
+Nevertheless, there are important and fundamental differences between adaptation by aneuploidy
+and adaptation by beneficial mutations~\citep{Yona2015}, which make aneuploidy a unique mechanism for generating genetic
+variation.
+First, the aneuploidy rate (i.e. the frequency of mis-segregation events) is significantly higher than the
+mutation rate~\citep{Santaguida2015review}.
+Thus, everything else being equal, adaptation by aneuploidy will be faster and more frequent.
+Second, fitness effects of aneuploidy are larger than those of the majority of mutations, on average, and are rarely
+neutral~\citep{Pavelka2010, Yona2012, Sunshine2015}, allowing selection to quickly sort deleterious and beneficial genotypes.
+Third, the number of different karyotypes is considerably smaller than the number of different genotypes, and different karyotypes are likely to have different phenotypes~\citep{Pavelka2010}.
+Therefore, exploration of the phenotype space by aneuploidy requires smaller populations and a shorter time span.
+Fourth, aneuploidy is a reversible state, as the rate of chromosome loss is high and the cost of aneuploidy is significant~\citep{Niwa2006}.
+Indeed, aneuploidy often provides a transient solution: under short-term stress conditions, aneuploidy reverts (chromosome number returns to normal) when the stress subsides; under long-term stress conditions, aneuploidy reverts when refined solutions, generated by beneficial mutations, take over~\citep{Yona2012}.
+Finally, aneuploidy results in increased genome instability, potentially increasing genetic variation by a positive feedback loop~\citep{Rancati2013, Bouchonville2009, Zhu2012}, while also increasing its own transience.
+
+\paragraph*{Evolutionary theory of aneuploidy.}
+The role of aneuploidy in adaptation has only recently been observed~\citep{Sionov2010, Yona2012, Gerstein2015}, and is largely missing from the literature on evolution and adaptation:
+the introductory textbook \emph{Evolution} by~\citet{Bergstrom2012} does not mention the word aneuploidy, and the graduate-level book \emph{Mutation-Driven Evolution} by~\citet{Nei2013} only briefly mentions aneuploidy in the context of speciation, but not adaptation.
+In recent reviews of the literature, aneuploidy is suggested to play an important role in fungal adaptation~\citep{Robbins2017, Todd2017} and cancer evolution~\citep{Santaguida2015review, Naylor2016, Sansregret2017}, yet these reviews cite no theoretical studies nor any quantitative models.
+Indeed, evolutionary, ecological, and epidemiological studies mostly assume adaptation occurs via beneficial mutations, recombination, and sex.
+Therefore, there is a critical need to develop an evolutionary theory of aneuploidy like the evolutionary theories of other mechanisms for generation of genetic variation, e.g.
+mutation~\citep{Lynch2010}, recombination~\citep{Hartfield2012}, and sex~\citep{Otto2009}.
+An evolutionary theory of aneuploidy will be central to the interpretation of experimental and clinical observations and design of new hypotheses, experiments, and treatments~\citep{Carja2014}.
+For example, despite the lack of theoretical models, aneuploidy has been invoked in a new strategy to combat pathogens and tumour cells by setting “evolutionary traps”~\citep{Gerstein2015,Chen2015}, in which a condition that predictably leads to emergence of aneuploidy is applied, followed by a condition that specifically selects against aneuploid cells.
 

### 2021-02-10 ms/ms.tex
revised ISF ack
moved bib library
fix library path
update authors, acks, affil

@@
 
-\author[a]{Ilia Kohanovski}
-\author[b,c]{Avihu H. Yona}
-\author[d]{Orna Dahan}
-\author[d]{Yitzhak Pilpel}
-\author[a,*]{Yoav Ram}
+\author[1]{Ilia Kohanovski}
+%\author[b,c]{Avihu H. Yona}
+%\author[d]{Orna Dahan}
+%\author[d]{Yitzhak Pilpel}
+\author[1,2,*]{Yoav Ram}
 
-\affil[a]{School of Computer Sciences, IDC Herzliya, Herzliya 4610101, Israel}
-\affil[b]{Department of Physics, Massachusetts Institute of Technology, Cambridge, MA 02139, USA}{
-\affil[b]{Department of Biological Engineering, Massachusetts Institute of Technology, Cambridge, MA 02139, USA}
-\affil[d]{Department of Molecular Genetics, Weizmann Institute of Science, Rehovot 76100, Israel}
+\affil[1]{School of Computer Sciences, IDC Herzliya, Herzliya, Israel}
+\affil[2]{School of Zoology, Faculty of Life Sciences, Tel Aviv University, Tel Aviv, Israel}
+%\affil[b]{Department of Physics, Massachusetts Institute of Technology, Cambridge, MA 02139, USA}{
+%\affil[b]{Department of Biological Engineering, Massachusetts Institute of Technology, Cambridge, MA 02139, USA}
+%\affil[d]{Department of Molecular Genetics, Weizmann Institute of Science, Rehovot 76100, Israel}
 \affil[*]{Corresponding author: yoav@yoavram.com}
@@
 This work was supported in part by 
-the Israel Science Foundation (YR 552/19).
+the Israel Science Foundation 552/19 and Minerva
+Stiftung Center for Lab Evolution (YR).
 }
@@
 \bibliographystyle{agsm}
-\bibliography{/Users/yoavram/library.bib}
+\bibliography{ms.bib}
 

### 2021-04-05 ms/ms.tex
added single-mutant model description

@@
 
-%\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
+\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
 \usepackage{graphicx} % Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
@@
 \usepackage{authblk}
+\usepackage{subcaption}
+\captionsetup[subfigure]{labelformat=empty}
+\usepackage[font=small,labelfont=bf]{caption}
 
@@
 
+\paragraph*{Evolutionary Models.} To examine the role of aneuploidy in adaptation, two models were developed - \emph{single-mutant model} and \emph{multiple-locus model}. 
+Both are based on the Wright-Fisher model~\citep{Tenaillon1999} and include the effects of natural selection, genetic drift, aneuploidy, and mutation. Only beneficial mutations assumed, 
+thus neglecting the effects of deleterious and neutral mutations. The \emph{single-mutant model} allows for a single mutation (on specific locus or target loci), while the 
+\emph{multiple-locus model} allows for multiple mutations. Both models allow for a single aneuploid karyotype (e.g. chromosome III duplication).
+
+\begin{figure}[b!]
+  \centering
+  \begin{subfigure}[t]{0.5\textwidth}
+      \centering
+      \includegraphics[height=1.8in]{../figures/Fig1-A.pdf}
+      \caption{
+        \textbf{A. Single-mutant}
+        }
+  \end{subfigure}%
+  ~ 
+  \begin{subfigure}[t]{0.5\textwidth}
+      \centering
+      \includegraphics[height=1.8in]{../figures/Fig1-B.pdf}
+      \caption{
+        \textbf{B. Multiple-locus}
+      }
+  \end{subfigure}
+  \caption{
+    \textbf{Model Schemes.} (A) The population with four possible genotypes and four possible transitions from one genotype to another with the transition rates $\mu$, $\alpha$, 
+    $\alpha_l$, $k\mu$. \emph{2n}: euploid wildtype – the initial genotype; \emph{2n*}: euploid mutant, standard karyotype with one specific beneficial mutation; 
+    \emph{2n+1}: aneuploid wildtype, with an extra chromosome, e.g. following chromosome duplication; \emph{2n+1*}: 
+    aneuploid mutant, with both the beneficial mutation and the extra chromosome. Overall there are two possible paths to reach \emph{2n*} from \emph{2n}.
+    (B) Each genotype can have zero or more beneficial mutations and zero or one extra chromosomes. $\mu$, mutation rate; $\alpha$, aneuploidy gain rate; $\alpha_l$, aneuploidy loss rate
+  }
+  \label{fig:models}
+\end{figure}
+
+\paragraph*{Single-mutant model.} It contains four predefined genotypes whose frequencies change during generations. \emph{2n}: euploid wildtype - the initial genotype; 
+\emph{2n*}: euploid mutant, standard karyotype with one specific beneficial mutation; \emph{2n+1}: aneuploid wildtype, with an extra chromosome, e.g. following chromosome duplication; 
+\emph{2n+1*}: aneuploid mutant, with both the beneficial mutation and the extra chromosome.
+
+The parameters of the model are: 
+$N$, the population size, which is constant in all generations; $\mu$, the mutation rate - the transition rate from \emph{2n} to \emph{2n*}, i.e. the frequency at which cells with genotype \emph{2n*}
+are created from cells with genotype \emph{2n}; $\alpha$, the aneuploidy gain rate - the transition rate from \emph{2n} to \emph{2n+1}; 
+$\alpha_l$, the aneuploidy loss rate - the transition rate from \emph{2n+1*} to \emph{2n*}; $k$, the coefficient, such that $k\cdot\mu$ is the transition rate from \emph{2n+1} to \emph{2n+1*}, 
+where on default $k$ is equals to $33/32$ (euploid \emph{2n} genotyped \emph{S. cerevisiae} contains 32 chromosomes, therefore, assuming each chromosome have equal probability to have the specific 
+beneficial mutation in, aneuploid \emph{2n+1} genotype should have 33/32 times the mutation rate); $s$, the mutation selection coefficient; $s_b$, fitness benefit of aneuploidy; 
+$c$, fitness cost of aneuploidy, where $0<c<1$. The relative finesses of genotypes are:
+\begin{equation} \label{eq:model} \begin{aligned}
+w_{2n} & = 1 \\
+w_{2n+1} & = 1-c+s_b \\
+w_{2n+1*} & = (1+s)(1-c)+s_b \\
+w_{2n*} & = (1+s) \;
+\end{aligned}
+\end{equation}
+
 \pagebreak

### 2021-04-18 ms/ms.tex
added Wright-Fisher steps for single model

@@
 thus neglecting the effects of deleterious and neutral mutations. The \emph{single-mutant model} allows for a single mutation (on specific locus or target loci), while the 
-\emph{multiple-locus model} allows for multiple mutations. Both models allow for a single aneuploid karyotype (e.g. chromosome III duplication).
+\emph{multiple-locus model} allows for multiple mutations. Both models allow for a single aneuploid karyotype (e.g. chromosome III duplication). (\autoref{fig:models})
 
@@
         }
+      \label{fig:model1}
   \end{subfigure}%
@@
       }
+      \label{fig:model2}
   \end{subfigure}
@@
 $c$, fitness cost of aneuploidy, where $0<c<1$. The relative finesses of genotypes are:
-\begin{equation} \label{eq:model} \begin{aligned}
+\begin{equation} \label{eq:single-w} \begin{aligned}
 w_{2n} & = 1 \\
@@
 \end{equation}
+
+\pagebreak
+
+The model follows Wright-Fisher \citep{Tenaillon1999} steps. The first generation is initialized with $N$ cells of the genotype \emph{2n}. At each generation, the frequency $f_g$ of 
+each genotype $g$ is updated:
+\begin{enumerate}
+  \item Selection step: 
+    \begin{equation} \label{eq:single-selection} 
+      f_g\gets{\dfrac{f_g w_g}{\sum_{j}f_j w_j}}
+      %  j\in\{2n, 2n+1, 2n+1*, 2n*\} 
+    \end{equation}
+  \item Mutation gain and aneuploidy gain/loss step:
+    \begin{equation} \label{eq:single-mutation} \begin{aligned}
+      f_{2n} & \gets{f_{2n} - \alpha f_{2n} - \mu f_{2n}}  \\
+      f_{2n+1} & \gets{f_{2n+1} - k\mu f_{2n+1} + \alpha f_{2n}}  \\
+      f_{2n+1*} & \gets{f_{2n+1*} - \alpha_l f_{2n+1*} + k\mu f_{2n+1}}  \\
+      f_{2n*} & \gets{\mu f_{2n} + \alpha_l f_{2n+1}}  \;
+    \end{aligned}
+    \end{equation}
+  \item Random drift step: draw the frequencies of all genotypes from a multinomial distribution
+    \begin{equation} \label{eq:single-mutation} 
+      f \sim Mult(N, (f_{2n},f_{2n+1},f_{2n+1*},f_{2n*}))
+    \end{equation}
+\end{enumerate}
+
+\paragraph*{Multiple-locus model.} It expands the \emph{single-mutant model} by allowing for any number of beneficial mutations.
 

### 2021-05-24 ms/ms.tex
add todo for citation
continue with multiple-locus model

@@
 \paragraph*{Multiple-locus model.} It expands the \emph{single-mutant model} by allowing for any number of beneficial mutations.
+For each mutation $i$, selection coefficient $s_i$ is independently drawn from an exponential distribution with expected value $s$.
+The parameters of the model are $\mu$, the mutation rate; $\alpha$, the aneuploidy gain rate; $\alpha_l$, the aneuploidy loss rate; $s$, the parameter of the exponential distribution; $s_b$, fitness benefit of aneuploidy; $c$, fitness cost of aneuploidy, where $0<c<1$.
+In this model, the population may include any number of genotypes that contain zero or more mutations and zero or one extra chromosomes.
+The fitness $w_g$ of genotype $g$ is:
+
+\begin{equation} 
+w_g= 
+\begin{cases}
+   1, & \text{wildtype} \\
+	1-c+s_b, & \text{aneuploid wildtype} \\
+	\prod_{\text{i is mutation in g}}{(1+s_i)}, & \text{euploid mutant} \\ 
+	s_b+(1-c)\prod_{\text{i is mutation in g}}{(1+s_i)}, & \text{aneuploid mutant}
+\end{cases}
+\end{equation}
+
+Therefore, aneuploidy loss would be favored by selection only if there are enough beneficial mutations or/and the mutation selection coefficients $s_i$ are large enough.
+The intuition is as follows: when the benefit of the accumulated beneficial mutations is small, then the benefit of aneuploidy has a large effect; when the benefit of the accumulated beneficial mutations benefit is large, then aneuploidy doesn’t add much, and its cost becomes significant.
+
+Unlike the single-locus model, the population size for the multiple-locus model is not constant.
+Indeed, a serial-transfer protocol~\citep{Barrick2013} is simulated, where the population is repeatedly diluted by transfer to a fresh medium, starting a new growth cycle, similar to the original experiment.
+In the model the population initial size is $N$, then it is doubled every iteration till reaching $8\cdot{N}$, then it is diluted back to $N$.
+
+Denote by $n_g$ be the number of cells of genotype $g$ in the population. At each generation, genotypes frequencies are changed by the following steps:
+\begin{enumerate}
+	\item Selection step: 
+	\begin{equation} \label{eq:multiple-selection} 
+		n_g\gets{n_g\cdot{w_g}}
+	\end{equation}
+	\item Random drift step:
+	\begin{equation} \label{eq:multiple-rand-drift} \begin{aligned}
+			f \sim Mult(N', (\frac{n_1}{N'},\frac{n_2}{N'},...)) \text{ where } N'=n_1+n_2+... 
+		\end{aligned}
+	\end{equation}
+	\item Normalization after N change (growth or dilution):
+	\begin{equation} \label{eq:multiple-pop-size} 
+		n_g\gets{f_g\cdot{N}}
+	\end{equation}
+\end{enumerate}
 

### 2021-05-25 ms/ms.tex
finished multiple-locus model description

@@
 
-\begin{equation} 
+\begin{equation} \label{eq:multiple-fitness}
 w_g= 
@@
 	\end{equation}
-	\item Normalization after N change (growth or dilution):
+	\item Normalization after population size N change (growth or dilution):
 	\begin{equation} \label{eq:multiple-pop-size} 
@@
 	\end{equation}
+	\item Mutation step: \\
+	For each genotype $g$, draw the number of mutant cells from a binomial distribution $Bin(n_g, \mu)$, and subtract that number from $n_g$.
+For each mutant cell, generate a new genotype and add one cell of this genotype to the population.
+The newly created genotype inherits mutations and ploidy karyotype from the parent genotype $g$, and in addition gains one more mutation, with the selected coefficient drawn from the exponential distribution $Exp(\lambda)$.
+The fitness of the newly created genotype is assigned according to the Formula \ref{eq:multiple-fitness}.
+	\item Aneuploidy gain step: \\
+	For each genotype $g$ that doesn’t already have an aneuploid chromosome, draw the number of new cells that gain aneuploidy from a binomial distribution $z \sim Bin(n_g, \alpha)$.
+If $z > 0$, then subtract it from $n_g$ and create a new genotype and add to the population with $z$ cells of this genotype.
+The genotype inherits all the mutations of genotype $g$ and in addition gains aneuploidy.
+The fitness is calculated appropriately by the Formula \ref{eq:multiple-fitness}.
+	\item Aneuploidy loss step: \\
+	It is similar to the aneuploidy step.
+For each genotype $g$ that has aneuploidy, draw the number of new cells that lose aneuploidy from the binomial distribution $z \sim Bin(n_g,\alpha_l)$.
+If $z>0$, subtract $z$ from $n_g$, then create new genotype, and add to the population $z$ cells of this genotype.
+The genotype inherits all the mutations of $g$ and in addition gains aneuploidy.
+The fitness is calculated appropriately by the Formula \ref{eq:multiple-fitness}.
 \end{enumerate}
+
+\paragraph*{Parameters inference.}
 

### 2021-06-06 ms/ms.tex
added data and likelihood to ms.tex

@@
 
-\paragraph*{Parameters inference.}
+\paragraph{Data.} We inference parameters of the models based on the data of the experiments of Yona at al ~\cite{Yona2012}.
+In \emph{heat-stress} experiment four populations of \emph{S.
+cerevisiae} were evolved under $39^{\circ}$ heat-stress and developed aneuploidy till 450 generations (here and further by development/loss of genotype trait we mean that 95\% of the population are the carriers of the genotype).
+Then, two populations were checked at generations 1700 and 2350, and for one population the aneuploidy were eliminated and has not observed allready at generation 1700 and for the second at generation 2350.
+\\
+Similarly, in \emph{ph-stress} experiment four populations were evolved unde high-ph stress conditions and developed aneuploidy during 150 generations.
+At generation 750 no aneuploidy was observed for all four populations.
+
+\paragraph{Likelihood function.}
+ Denote by $A_{450}$ the event of aneuploidy observation at generation 450; by $L_{1700}$ the event of aneuploidy loss observation at generation 1700; and by $L_{2350^*}$ the event of aneuploidy loss observation at generation 2350 while there's no loss in generation 1700.
+The likelihood function of the model with parameter set $\phi$ for \emph{heat-stress} experiment is:
+ \begin{multline} \label{eq:heatstress-likelihood}
+L(\phi | A_{450}, L_{1700}, L_{2350^*} ) = \\
+P(L_{450})^4 \cdot (1 - P(\lnot{L_{1700}} | A_{450} )^4 - P(\lnot{L_{2350^*}}| A_{450})^4 + P(\lnot{L_{1700}} \land \lnot{L_{2350^*}}| A_{450})^4 )		
+ \end{multline}
 

### 2021-06-07 ms/ms.tex
likelihood functions updated

@@
 \end{abstract}
+
+% TODO cite https://www.biorxiv.org/content/10.1101/2020.10.07.327759v1
 
@@
 Similarly, in \emph{ph-stress} experiment four populations were evolved unde high-ph stress conditions and developed aneuploidy during 150 generations.
-At generation 750 no aneuploidy was observed for all four populations.
+At generation 750 aneuploidy fully eliminated for two populations, and only partly for another two.
 
 \paragraph{Likelihood function.}
- Denote by $A_{450}$ the event of aneuploidy observation at generation 450; by $L_{1700}$ the event of aneuploidy loss observation at generation 1700; and by $L_{2350^*}$ the event of aneuploidy loss observation at generation 2350 while there's no loss in generation 1700.
-The likelihood function of the model with parameter set $\phi$ for \emph{heat-stress} experiment is:
+Denote by $A_{t}$ the event of aneuploidy observation at generation $t$; by $L_{t}$ the event of aneuploidy loss observation at generation $t$; and by $L_{2350^*}$ the event of aneuploidy loss observation at generation 2350 while there's no loss in generation 1700.
+The likelihood function of the model with parameter set $\theta$ for \emph{heat-stress} experiment is:
 \begin{multline} \label{eq:heatstress-likelihood}
-L(\phi | A_{450}, L_{1700}, L_{2350^*} ) = \\
-P(L_{450})^4 \cdot (1 - P(\lnot{L_{1700}} | A_{450} )^4 - P(\lnot{L_{2350^*}}| A_{450})^4 + P(\lnot{L_{1700}} \land \lnot{L_{2350^*}}| A_{450})^4 )		
+L(\theta | A_{450}, L_{1700}, L_{2350^*} ) = \\
+P^4(A_{450}) \cdot (1 - P^4(\lnot{L_{1700}} | A_{450} ) - P^4(\lnot{L_{2350^*}}| A_{450}) + P^4(\lnot{L_{1700}} \land \lnot{L_{2350^*}}| A_{450}) )		
 \end{multline}
+The likelihood function of the model with parameter set $\theta$ for \emph{ph-stress} experiment is:
+\begin{equation} \label{eq:phstress-likelihood}
+ L(\theta | A_{150}, L_{750} ) = \\
+ P^4(L_{450}) \cdot 6 \cdot	P^4(L_{750} | A_{150}) \cdot P^4(\lnot{L_{750}} | A_{150})
+\end{equation}
+
+\paragraph{Parameter inference.} To estimate the parameters of the models we apply approximate Bayesian computation with a sequential Monte-Carlo scheme (ABC-SMC), employing pyabc Python 
+package \citep{Klinger2018}
 

### 2021-06-17 ms/ms.tex
add MP, AHY as authors
started revision

@@
 
+% genotype commands
+\newcommand{\euwt}{\emph{2n}}
+\newcommand{\anwt}{\emph{2n+1}}
+\newcommand{\eumt}{\emph{2n*}}
+\newcommand{\anmt}{\emph{2n+1*}}
+
 % NatBib
@@
 % Chromosomal duplication is a transient evolutionary solution to stress
-\title{Clonal interference between aneuploidy and mutation}
+%\title{Clonal interference between aneuploidy and mutation}
+\title{Adaptive evolution with aneuploidy and mutation}
 
@@
 
-\author[1]{Ilia Kohanovski}
-%\author[b,c]{Avihu H. Yona}
+\author[1,*]{Ilia Kohanovski}
+\author[2,*]{Martin Pontz}
+\author[3]{Avihu H. Yona}
 %\author[d]{Orna Dahan}
 %\author[d]{Yitzhak Pilpel}
-\author[1,2,*]{Yoav Ram}
+\author[1,2,$\dagger$]{Yoav Ram}
 
@@
 \affil[2]{School of Zoology, Faculty of Life Sciences, Tel Aviv University, Tel Aviv, Israel}
-%\affil[b]{Department of Physics, Massachusetts Institute of Technology, Cambridge, MA 02139, USA}{
-%\affil[b]{Department of Biological Engineering, Massachusetts Institute of Technology, Cambridge, MA 02139, USA}
+\affil[3]{Institute of Biochemistry, Food Science and Nutrition,
+Robert H. Smith Faculty of Agriculture, Food and Environment,
+The Hebrew University of Jerusalem, Israel}
 %\affil[d]{Department of Molecular Genetics, Weizmann Institute of Science, Rehovot 76100, Israel}
-\affil[*]{Corresponding author: yoav@yoavram.com}
+\affil[*]{These authors contributed equally to this work}
+\affil[$\dagger$]{Corresponding author: yoav@yoavram.com}
 
@@
 Aneuploidy is common in eukaryotes, often leading to decreased cell growth and fitness.
-However, evidence from yeast and fungi, as well as tumour cells, suggests that aneuploidy can be beneficial under stressful conditions and  lead to elevated growth rates and to adaptation.
-Thus, it is crucial to develop a quantitative theory for the role of aneuploidy in adaptive evolution.
-Here, we analyse results from experiments in which \yeast adapted to heat stress.
+However, evidence from yeast and fungi, as well as human tumour cells, suggests that aneuploidy can be beneficial under stressful conditions and lead to elevated growth rates and adaptation.
+Because aneuploidy differs from mutation in rate, expected effect, and reversibility, it is crucial to develop a quantitative theory for the role of aneuploidy in adaptive evolution.
+Here, we develop evolutionary models for adaptive evolution with both mutation and aneuploidy. 
+These models are used within an approximate Bayesian computation framework to estimate the formation rate and fitness effect of aneuploidy and mutation from empirical results of experiments in which \yeast adapted to heat stress.
 The experimental population first acquired chromosome duplications, only to later revert back to a euploid state.
-We use several evolutionary models within a approximate Bayesian computation framework to show that clonal interference between chromosome duplications and beneficial mutations explains the experimental results.
-Our results suggest that aneuploidy can both accelerate and deccelerate adaptation in a non-intuitive manner, creating an evolutionary conflict between the individual and the population and further complicating the process of adaptation in cell populations.
+We also analyze our models to estimate the effect of the aneuploidy and mutation rates on the expected adaptation time and the probability for adaptation via aneuploidy.
+Our results suggest that aneuploidy is a transient adaptive solution, which can decelerate adaptation in a non-intuitive manner.
+By creating an evolutionary conflict between the individual and the population, aneuploidy further complicates  the process of adaptation in cell populations.
 \end{abstract}
@@
 Aneuploidy is an imbalance in the number of chromosomes in the cell: an incorrect karyotype.
-Recent evidence suggests aneuploidy is very common in eukaryotes, e.g. animals~\citep{Santaguida2015review, Naylor2016, Bakhoum2017}, and fungi~\citep{Pavelka2010, Zhu2016, Robbins2017, Todd2017}.
+Evidence suggests aneuploidy is very common in eukaryotes, e.g. animals~\citep{Santaguida2015review, Naylor2016, Bakhoum2017}, and fungi~\citep{Pavelka2010, Zhu2016, Robbins2017, Todd2017}.
 Aneuploidy has been implicated in cancer formation and progression~\citep{Boveri2008, Schvartzman2010}: 
@@
 \paragraph*{Aneuploidy is generally deleterious.}
-The molecular and genetic mechanisms involved in aneuploidy have been explored in the past decade~\citep{Musacchio2007, Sheltzer2011, Chen2012b, Rancati2013, Gerstein2015, Shor2015}.
-Experiments with human and mouse embryos found that aneuploidy is usually lethal;
-it is also associated with developmental defects and lethality in other multicellular organisms~\citep{Sheltzer2011}.
+The molecular and genetic mechanisms involved in aneuploidy have been explored~\citep{Musacchio2007, Sheltzer2011, Chen2012b, Rancati2013, Gerstein2015, Shor2015}.
+Experiments with human and mouse embryos found that aneuploidy is usually lethal.
+It is also associated with developmental defects and lethality in other multicellular organisms~\citep{Sheltzer2011}.
 For example, aneuploid mouse embryonic cells grow slower than euploid cells~\citep{Williams2008}.
@@
 Yet, because it affects many genes on a whole chromosome or a chromosome fragment, aneuploidy also carries fitness costs.
-Thus, aneuploidy can be a \emph{transient adaptive solution}: it can rapidly occur and fix in the population under stressful conditions, and can be rapidly lost when the cost outweighs the benefit -- when stress is removed or after beneficial mutations occur.
-Experimental evidence of such a transient role of aneuploidy was demonstrated by \citet{Yona2012}, who evolved populations of \yeast under strong heat or pH stress.
-In these experiments, the populations quickly adapted to the stress, and this adaptation was determined to be due to chromosome duplications.
-Later on, the populations reverted back to an euploid state, while remaining adapted to the stress and accumulating multiple mutations.
+Thus, aneuploidy can be a \emph{transient adaptive solution}: it can rapidly occur and fix in the population under stressful conditions, and can be rapidly lost when the cost outweighs the benefit---when stress is removed or after beneficial mutations occur.
+Experimental evidence of such a transient role of aneuploidy was demonstrated by \citet{Yona2012}. They evolved populations of \yeast under strong heat or pH stress.
+In these experiments, the populations adapted to the stress within 450 generations, and this adaptation was determined to be due to chromosome duplications.
+Much later, after more than 1500 generations, the populations reverted back to an euploid state, while remaining adapted to the stress and accumulating multiple mutations.
 However, under gradual heat stress, aneuploidy was not observed.
@@
 \paragraph*{The present study.}
-Here, we develop a series of computational evolutionary models that include the effects of natural selection, genetic drift, aneuploidy and mutation to examine the role of aneuploidy in adaptation.
+Here, we develop evolutionary-genetic models that include the effects of natural selection, genetic drift, aneuploidy, and mutation to examine the role of aneuploidy in adaptive evolution.
 These models follow a population of cells characterised by both their ploidy and their genotype.
-We fit these models to the empirical results of \citet{Yona2012} using approximate Bayesian computation.
-These model fits allow us to infer model parameters, including selection coefficients and transition rates, and to perform model selection between different models, thereby testing different hypotheses about the evolutionary process.
+We fit these models to the experimental results of \citet{Yona2012} using an \emph{approximate Bayesian computation} framework~\citep{Sisson2009, Klinger2018} to infer model parameters, including selection coefficients and rates of aneuploidy and mutation, and to perform model selection between different models, thereby testing different hypotheses about the evolutionary process.
+Furthermore, we analyze these models to estimate the effects of parameters on the adaptation time and the probability for adaptation via aneuploidy.
 We find that % TODO CHECK!
-the aneuploidy rate is several orders of magnitude higher than the mutation rate; that a simple model of clonal interference is not enough to explain the transience of aneuploidy; and that aneuploidy rescues the population, albeit at the price of delaying further adaptation.
+the aneuploidy rate is several orders of magnitude higher than the mutation rate; that a simple model of clonal interference is not enough to explain the transience of aneuploidy; % what does that mean?
+and that aneuploidy is likely to fix in the population and increase its mean fitness, but at the price of delaying further adaptation.
 
@@
 
-\paragraph*{Evolutionary Models.} To examine the role of aneuploidy in adaptation, two models were developed - \emph{single-mutant model} and \emph{multiple-locus model}. 
-Both are based on the Wright-Fisher model~\citep{Tenaillon1999} and include the effects of natural selection, genetic drift, aneuploidy, and mutation. Only beneficial mutations assumed, 
-thus neglecting the effects of deleterious and neutral mutations. The \emph{single-mutant model} allows for a single mutation (on specific locus or target loci), while the 
-\emph{multiple-locus model} allows for multiple mutations. Both models allow for a single aneuploid karyotype (e.g. chromosome III duplication). (\autoref{fig:models})
-
 \begin{figure}[b!]
@@
   \begin{subfigure}[t]{0.5\textwidth}
+      \caption{
+        \textbf{A. Single-locus model}
+      }
       \centering
       \includegraphics[height=1.8in]{../figures/Fig1-A.pdf}      
-      \caption{
-        \textbf{A. Single-mutant}
-        }
       \label{fig:model1}
   \end{subfigure}%
-  ~ 
+  \\
   \begin{subfigure}[t]{0.5\textwidth}
+  	  \caption{
+        \textbf{B. Multi-locus model}
+      }
       \centering
       \includegraphics[height=1.8in]{../figures/Fig1-B.pdf}      
-      \caption{
-        \textbf{B. Multiple-locus}
-      }
       \label{fig:model2}
@@
   \caption{
-    \textbf{Model Schemes.} (A) The population with four possible genotypes and four possible transitions from one genotype to another with the transition rates $\mu$, $\alpha$, 
-    $\alpha_l$, $k\mu$. \emph{2n}: euploid wildtype – the initial genotype; \emph{2n*}: euploid mutant, standard karyotype with one specific beneficial mutation; 
-    \emph{2n+1}: aneuploid wildtype, with an extra chromosome, e.g. following chromosome duplication; \emph{2n+1*}: 
-    aneuploid mutant, with both the beneficial mutation and the extra chromosome. Overall there are two possible paths to reach \emph{2n*} from \emph{2n}.
-    (B) Each genotype can have zero or more beneficial mutations and zero or one extra chromosomes. $\mu$, mutation rate; $\alpha$, aneuploidy gain rate; $\alpha_l$, aneuploidy loss rate
+    \textbf{Model illustrations.}
+    \textbf{(A)} In the single-locus model, the four genotypes are: euploid wild-type, \euwt; euploid mutant, \eumt; aneuploid wild-type, \anwt; and aneuploid mutant, \anmt.
+    Overall there are two possible trajectories from \euwt to \eumt.
+    \textbf{(B)} In the multi-locus model, each genotype is characterized by its karyotype, \euwt or \anwt, and the number of accumulated beneficial mutations, denoted by stars. 
+    In both panels arrows denote transitions between genotypes, with transitions rates: $\mu$, mutation rate; $\alpha$, aneuploidy gain rate; $\alpha_l$, aneuploidy loss rate.
   }
@@
 
-\paragraph*{Single-mutant model.} It contains four predefined genotypes whose frequencies change during generations. \emph{2n}: euploid wildtype - the initial genotype; 
-\emph{2n*}: euploid mutant, standard karyotype with one specific beneficial mutation; \emph{2n+1}: aneuploid wildtype, with an extra chromosome, e.g. following chromosome duplication; 
-\emph{2n+1*}: aneuploid mutant, with both the beneficial mutation and the extra chromosome.
+\paragraph*{Evolutionary Models.}
+We developed two models: a single-locus model and a multi-locus model. 
+Both models are based on the Wright-Fisher model~\citep{Otto2007}, assuming a constant population size $N$ with non-overlapping generations and including the effects of natural selection, genetic drift, aneuploidy, and mutation.
+We focus on beneficial mutations, neglecting the effects of deleterious and neutral mutations. Both models allow for a single aneuploid karyotype (e.g., chromosome III duplication).
+While the single-locus model allows for only a single mutation to occur, the
+multi-locus model allows for multiple mutations to accumulate in the genome~(\autoref{fig:models}).
 
-The parameters of the model are: 
-$N$, the population size, which is constant in all generations; $\mu$, the mutation rate - the transition rate from \emph{2n} to \emph{2n*}, i.e. the frequency at which cells with genotype \emph{2n*}
-are created from cells with genotype \emph{2n}; $\alpha$, the aneuploidy gain rate - the transition rate from \emph{2n} to \emph{2n+1}; 
-$\alpha_l$, the aneuploidy loss rate - the transition rate from \emph{2n+1*} to \emph{2n*}; $k$, the coefficient, such that $k\cdot\mu$ is the transition rate from \emph{2n+1} to \emph{2n+1*}, 
-where on default $k$ is equals to $33/32$ (euploid \emph{2n} genotyped \emph{S. cerevisiae} contains 32 chromosomes, therefore, assuming each chromosome have equal probability to have the specific 
-beneficial mutation in, aneuploid \emph{2n+1} genotype should have 33/32 times the mutation rate); $s$, the mutation selection coefficient; $s_b$, fitness benefit of aneuploidy; 
-$c$, fitness cost of aneuploidy, where $0<c<1$. The relative finesses of genotypes are:
-\begin{equation} \label{eq:single-w} \begin{aligned}
-w_{2n} & = 1 \\
-w_{2n+1} & = 1-c+s_b \\
-w_{2n+1*} & = (1+s)(1-c)+s_b \\
-w_{2n*} & = (1+s) \;
+In the following we describe the
+
+\paragraph*{Single-locus model.}
+This model follows four genotypes: euploid wild-type, \euwt, the initial genotype; 
+euploid mutant, \eumt, with the standard karyotype and a single beneficial mutation; 
+aneuploid wild-type, \anwt, with an extra chromosome, e.g. following chromosome duplication; and
+aneuploid mutant, \anmt, with and extra chromosome and a beneficial mutation.
+
+Beneficial mutations from \euwt\; to \eumt\; occur with probability $\mu$, the mutation rate, and from \anwt\; to \anmt\; with probability $\tau \mu$, where $\tau$ is the fold-change in the rate of beneficial mutations in aneuploid cells.
+By default, we assume $\tau=33/32$, as the wild-type \yeast strains in the experiments by \citep{Yona2012} are diploid, with 32 chromosomes, and the aneuploid strains are trisomic, with 33 chromosomes.
+Aneuploidy is formed by chromosome missegregation, so that cells transition from \euwt\; to \anwt\; or from \eumt\; to \anmt\; with probability $\delta$, the aneuploidy gain rate.
+Aneuploidy is lost, transitioning cells from \anwt\; to \euwt\; or from \anmt\; to \eumt\; with probability $\delta_L$, the aneuploidy loss rate.
+The fitness values of the four genotypes are given by
+\begin{equation} \label{eq:fitness-single}
+\begin{aligned}
+&w_{2n}  = 1 \;\;&
+&w_{2n+1}  = 1-c+\sigma \;\;&
+\\
+&w_{2n+1^*}  = (1-c)(1+s)+\sigma \;\;&
+&w_{2n^*}  = 1+s \;,\;&
 \end{aligned}
 \end{equation}
+such that $s \ge 0$ is the selection coefficient of a beneficial mutation;
+$\sigma \ge 0$ is the selection coefficient of aneuploidy; 
+and $0 \le c \le 1$ is the fitness cost of aneuploidy.
 
-\pagebreak
-
-The model follows Wright-Fisher \citep{Tenaillon1999} steps. The first generation is initialized with $N$ cells of the genotype \emph{2n}. At each generation, the frequency $f_g$ of 
-each genotype $g$ is updated:
-\begin{enumerate}
-  \item Selection step: 
-    \begin{equation} \label{eq:single-selection} 
-      f_g\gets{\dfrac{f_g w_g}{\sum_{j}f_j w_j}}
-      %  j\in\{2n, 2n+1, 2n+1*, 2n*\} 
+The first generation is initialized with $N$ cells with genotype \euwt.
+The effect of natural selection on the frequency $f_i$ of genotype $i = 2n, 2n+1, 2n+1^*, \text{or } 2n^*$ is given by
+    \begin{equation} \label{eq:selection-single} 
+      f^s_i = \frac{f_i w_i}{\bar{w}} \;,
     \end{equation}
-  \item Mutation gain and aneuploidy gain/loss step:
-    \begin{equation} \label{eq:single-mutation} \begin{aligned}
-      f_{2n} & \gets{f_{2n} - \alpha f_{2n} - \mu f_{2n}}  \\
-      f_{2n+1} & \gets{f_{2n+1} - k\mu f_{2n+1} + \alpha f_{2n}}  \\
-      f_{2n+1*} & \gets{f_{2n+1*} - \alpha_l f_{2n+1*} + k\mu f_{2n+1}}  \\
-      f_{2n*} & \gets{\mu f_{2n} + \alpha_l f_{2n+1}}  \;
+where $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
+The effect of mutation and aneuploidy on genotype frequencies is given by
+    \begin{equation} \label{eq:mutation-aneuploidy-single}
+    \begin{aligned}
+      f^m_{2n} & = {f^s_{2n} - \delta f^s_{2n} - \mu f^s_{2n}}  \;,\\
+      f^m_{2n+1} & = {f^s_{2n+1} - \tau \mu f^s_{2n+1} + \delta f^s_{2n}}  \;,\\
+      f^m_{2n+1^*} & = {f^s_{2n+1^*} - \delta_L f^s_{2n+1^*} + \tau \mu f^s_{2n+1}}  \;,\\
+      f^m_{2n^*} & = {\mu f^s_{2n} + \delta_L f^s_{2n+1}}  \;.
     \end{aligned}
     \end{equation}
-  \item Random drift step: draw the frequencies of all genotypes from a multinomial distribution
-    \begin{equation} \label{eq:single-mutation} 
-      f \sim Mult(N, (f_{2n},f_{2n+1},f_{2n+1*},f_{2n*}))
+Finally, the effect of random genetic drift on genotype frequencies is given by
+    \begin{equation} \label{eq:drift-single} 
+      N \cdot (f'_{2n},f'_{2n+1},f'_{2n+1^*},f'_{2n^*}) \sim Mult\Big(N, (f_{2n},f_{2n+1},f_{2n+1^*},f_{2n^*})\Big),
     \end{equation}
-\end{enumerate}
+where $Mult\Big(N, (f_{2n},f_{2n+1},f_{2n+1^*},f_{2n^*}\Big)$ is a multinomial distribution parameterized by the population size $N$ and the genotype frequencies.
+Overall, the change in genotype frequencies from one generation to the next is given by the transformation $f_i \to f'_i$.
 
-\paragraph*{Multiple-locus model.} It expands the \emph{single-mutant model} by allowing for any number of beneficial mutations.
-For each mutation $i$, selection coefficient $s_i$ is independently drawn from an exponential distribution with expected value $s$.
-The parameters of the model are $\mu$, the mutation rate; $\alpha$, the aneuploidy gain rate; $\alpha_l$, the aneuploidy loss rate; $s$, the parameter of the exponential distribution; $s_b$, fitness benefit of aneuploidy; $c$, fitness cost of aneuploidy, where $0<c<1$.
-In this model, the population may include any number of genotypes that contain zero or more mutations and zero or one extra chromosomes.
-The fitness $w_g$ of genotype $g$ is:
-
-\begin{equation} \label{eq:multiple-fitness}
-w_g= 
-\begin{cases}
-   1, & \text{wildtype} \\
-	1-c+s_b, & \text{aneuploid wildtype} \\
-	\prod_{\text{i is mutation in g}}{(1+s_i)}, & \text{euploid mutant} \\ 
-	s_b+(1-c)\prod_{\text{i is mutation in g}}{(1+s_i)}, & \text{aneuploid mutant}
-\end{cases}
-\end{equation}
+\paragraph*{Multi-locus model.}
+This model expands the single-locus model by allowing for the accumulation of beneficial mutations in the genome.
+Thus, a genotype is characterized by its karyotype, \euwt\; or \anwt, and the number of accumulated beneficial mutations, which can be zero or more.
+The selection coefficient of the $i$-th mutation, $s_i$, is drawn from an exponential distribution with expected value $s$, $s_i \sim Exp(s)$.
+The rest of the parameters ($N$, $\mu$, $\tau$, $\delta$, $\delta_L$, $\sigma$, $c$) are the same as in the single-locus model.
+The fitness of the different genotypes is the same as in the single-locus model (\autoref{eq:fitness-single}, except that the fitness contribution of beneficial mutations is the product of all accumulated mutations, $\prod_{i \ge 0}(1+s_i)$, instead of the contribution of the single mutation allowed in the single-locus model, $(1+s)$.
 
@@
 
-\paragraph*{Aneuploidy and mutation: \textit{same-same but different}.}
+\paragraph*{Aneuploidy is not just another type of mutation.}
 The published data indicate that, like mutation, aneuploidy can be both deleterious and beneficial~\citep{Pavelka2010, Sheltzer2011}.
@@
 \section*{Acknowledgements}
-We thank Lilach Hadany, Judith Berman, David Gresham, and Martin Kupiec for discussions and comments.
+We thank Yitzhak Pilpel, Orna Dahan, Lilach Hadany, Judith Berman, David Gresham, Shay Covo, Martin Kupiec, and Tal Simon for discussions and comments.
 This work was supported in part by 
-the Israel Science Foundation 552/19 and Minerva
-Stiftung Center for Lab Evolution (YR).
+the Israel Science Foundation (YR 552/19) and
+Minerva Stiftung Center for Lab Evolution (YR).
 }

### 2021-06-20 ms/ms.tex
finished model, data (evidence?) and likelihood

@@
 \usepackage[font=small,labelfont=bf]{caption}
+\usepackage{longtable}
+\usepackage{siunitx}
 
@@
 \newcommand{\ci}{I}
+\newcommand{\likelihood}{\mathcal{L}}
 
@@
 
-In the following we describe the
-
 \paragraph*{Single-locus model.}
-This model follows four genotypes: euploid wild-type, \euwt, the initial genotype; 
+This model follows four genotypes~(\autoref{fig:models}A): euploid wild-type, \euwt, the initial genotype; 
 euploid mutant, \eumt, with the standard karyotype and a single beneficial mutation; 
@@
 
-Beneficial mutations from \euwt\; to \eumt\; occur with probability $\mu$, the mutation rate, and from \anwt\; to \anmt\; with probability $\tau \mu$, where $\tau$ is the fold-change in the rate of beneficial mutations in aneuploid cells.
+Transitions between the genotypes occur as follows~(\autoref{fig:models}A): Beneficial mutations from \euwt\; to \eumt\; occur with probability $\mu$, the mutation rate, and from \anwt\; to \anmt\; with probability $\tau \mu$, where $\tau$ is the fold-change in the rate of beneficial mutations in aneuploid cells.
 By default, we assume $\tau=33/32$, as the wild-type \yeast strains in the experiments by \citep{Yona2012} are diploid, with 32 chromosomes, and the aneuploid strains are trisomic, with 33 chromosomes.
@@
 Aneuploidy is lost, transitioning cells from \anwt\; to \euwt\; or from \anmt\; to \eumt\; with probability $\delta_L$, the aneuploidy loss rate.
-The fitness values of the four genotypes are given by
-\begin{equation} \label{eq:fitness-single}
-\begin{aligned}
-&w_{2n}  = 1 \;\;&
-&w_{2n+1}  = 1-c+\sigma \;\;&
-\\
-&w_{2n+1^*}  = (1-c)(1+s)+\sigma \;\;&
-&w_{2n^*}  = 1+s \;,\;&
-\end{aligned}
-\end{equation}
-such that $s \ge 0$ is the selection coefficient of a beneficial mutation;
+The fitness values of the four genotypes are given by \autoref{table:single-locus}.
+
+%%% Table: single-locus model
+\begin{table}[h]
+\centering
+\caption{\textbf{Single-locus model fitness values.}}
+\begin{tabular}{lllll}
+\emph{Genotype} $i$ & $2n$ & $2n+1$ & $2n+1^*$ & $2n^*$ \\
+\hline
+\emph{Fitness} $w_i$ & $1$ & $1-c+\sigma$ & $(1-c)(1+s)+\sigma$ & $1+s$               
+\end{tabular}
+\label{table:single-locus}
+\caption*{
+$s \ge 0$ is the selection coefficient of a beneficial mutation;
 $\sigma \ge 0$ is the selection coefficient of aneuploidy; 
 and $0 \le c \le 1$ is the fitness cost of aneuploidy.
+}
+\end{table}
+%%%
 
@@
     \end{equation}
-where $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
+where the fitness values $w_i$ are given in \autoref{table:single-locus} and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
 The effect of mutation and aneuploidy on genotype frequencies is given by
@@
     \begin{aligned}
-      f^m_{2n} & = {f^s_{2n} - \delta f^s_{2n} - \mu f^s_{2n}}  \;,\\
-      f^m_{2n+1} & = {f^s_{2n+1} - \tau \mu f^s_{2n+1} + \delta f^s_{2n}}  \;,\\
-      f^m_{2n+1^*} & = {f^s_{2n+1^*} - \delta_L f^s_{2n+1^*} + \tau \mu f^s_{2n+1}}  \;,\\
-      f^m_{2n^*} & = {\mu f^s_{2n} + \delta_L f^s_{2n+1}}  \;.
+      &f^m_{2n} &=&\; (1 - \delta - \mu) f^s_{2n}  \;,\\
+      &f^m_{2n+1} &=&\; \delta f^s_{2n} + (1 - \tau \mu) f^s_{2n+1}  \;,\\
+      &f^m_{2n+1^*} &=&\; \tau \mu f^s_{2n+1} + (1-\delta_L) f^s_{2n+1^*}  \;,\\
+      &f^m_{2n^*} &=&\; \mu f^s_{2n} + \delta_L f^s_{2n+1} + f^s_{2n^*}  \;.
     \end{aligned}
     \end{equation}
-Finally, the effect of random genetic drift on genotype frequencies is given by
+Finally, random genetic drift is modeled using a multinomial distribution~\citep{Otto2007},
     \begin{equation} \label{eq:drift-single}
-      N \cdot (f'_{2n},f'_{2n+1},f'_{2n+1^*},f'_{2n^*}) \sim Mult\Big(N, (f_{2n},f_{2n+1},f_{2n+1^*},f_{2n^*})\Big),
+      \vec{f'} \sim \frac{1}{N} Mult(N, \vec{f}) \;,
     \end{equation}
-where $Mult\Big(N, (f_{2n},f_{2n+1},f_{2n+1^*},f_{2n^*}\Big)$ is a multinomial distribution parameterized by the population size $N$ and the genotype frequencies.
+where $\vec{f}=(f_{2n},f_{2n+1},f_{2n+1^*},f_{2n^*})$ and $\vec{f'}$ are the frequencies of the genotypes in the current and next generation, respectively, and $Mult(N, \vec{f})$ is a multinomial distribution parameterized by the population size $N$ and the genotype frequencies $\vec{f}$.
 Overall, the change in genotype frequencies from one generation to the next is given by the transformation $f_i \to f'_i$.
@@
 \paragraph*{Multi-locus model.}
-This model expands the single-locus model by allowing for the accumulation of beneficial mutations in the genome.
-Thus, a genotype is characterized by its karyotype, \euwt\; or \anwt, and the number of accumulated beneficial mutations, which can be zero or more.
-The selection coefficient of the $i$-th mutation, $s_i$, is drawn from an exponential distribution with expected value $s$, $s_i \sim Exp(s)$.
+This model expands the single-locus model by allowing for the accumulation of beneficial mutations in the genome and a fluctuating population size.
+
+A genotype is characterized by its karyotype, \euwt\; or \anwt, and the number of accumulated beneficial mutations, which can be zero or more.
+The selection coefficient of the $i$-th accumulated mutation in each individual, $s_i$, is drawn from an exponential distribution with expected value $s$, $s_i \sim Exp(s)$.
 The rest of the parameters ($N$, $\mu$, $\tau$, $\delta$, $\delta_L$, $\sigma$, $c$) are the same as in the single-locus model.
-The fitness of the different genotypes is the same as in the single-locus model (\autoref{eq:fitness-single}, except that the fitness contribution of beneficial mutations is the product of all accumulated mutations, $\prod_{i \ge 0}(1+s_i)$, instead of the contribution of the single mutation allowed in the single-locus model, $(1+s)$.
+The fitness of the different genotypes is the same as in the single-locus model (\autoref{table:single-locus}), except that the fitness contribution of $k$ beneficial mutations is the product of their independent effects, $\prod_{i=1}^k(1+s_i)$, instead of the contribution of the single mutation allowed in the single-locus model, $(1+s)$, see \autoref{table:multi-locus}.
+Therefore, aneuploidy loss would be favored by selection only if there are enough beneficial mutations and/or the selection coefficients $s_i$ are large enough.
+The intuition is that when the benefit of the accumulated beneficial mutations is small, the benefit of aneuploidy has a large effect; when the benefit of the accumulated beneficial mutations benefit is large, then aneuploidy is no longer advantageous because of its significant cost.
 
-Therefore, aneuploidy loss would be favored by selection only if there are enough beneficial mutations or/and the mutation selection coefficients $s_i$ are large enough.
-The intuition is as follows: when the benefit of the accumulated beneficial mutations is small, then the benefit of aneuploidy has a large effect; when the benefit of the accumulated beneficial mutations benefit is large, then aneuploidy doesn’t add much, and its cost becomes significant.
+In contrast to the single-locus model, in the multi-locus model the population size changes in order to model serial-transfer experiment protocol~\citep{Yona2012}: the population is serially diluted by transferring a fraction of the population (1/120) to a fresh medium, starting a new growth cycle.
+In this model, the population initial size is $N_0 = N$, and the population size is doubled every generation, $N_1=2N, N_2=4N, \ldots$, and diluted back to $N$ at every seven generations, $N_7=N$.
 
-Unlike the single-locus model, the population size for the multiple-locus model is not constant.
-Indeed, a serial-transfer protocol~\citep{Barrick2013} is simulated, where the population is repeatedly diluted by transfer to a fresh medium, starting a new growth cycle, similar to the original experiment.
-In the model the population initial size is $N$, then it is doubled every iteration till reaching $8\cdot{N}$, then it is diluted back to $N$.
+The change in frequencies due to selection is exactly the same as in the single-locus model (\autoref{eq:selection-single}), only applied using the fitness values in \autoref{table:multi-locus}.
+The change due to random genetic drift is also the same as in \autoref{eq:drift-single}, except that the frequencies vector is $\vec{f}=(f_{2n}, f_{2n+1}, f_{2n^*}, f_{2n+1^*}, f_{2n^{**}}, f_{2n+1^{**}}, \ldots)$ and that the population size changes between generations, as described above.
 
-Denote by $n_g$ be the number of cells of genotype $g$ in the population. At each generation, genotypes frequencies are changed by the following steps:
-\begin{enumerate}
-	\item Selection step: 
-	\begin{equation} \label{eq:multiple-selection} 
-		n_g\gets{n_g\cdot{w_g}}
-	\end{equation}
-	\item Random drift step:
-	\begin{equation} \label{eq:multiple-rand-drift} \begin{aligned}
-			f \sim Mult(N', (\frac{n_1}{N'},\frac{n_2}{N'},...)) \text{ where } N'=n_1+n_2+... 
-		\end{aligned}
-	\end{equation}
-	\item Normalization after population size N change (growth or dilution):
-	\begin{equation} \label{eq:multiple-pop-size} 
-		n_g\gets{f_g\cdot{N}}
-	\end{equation}
-	\item Mutation step: \\
-	For each genotype $g$, draw the number of mutant cells from a binomial distribution $Bin(n_g, \mu)$, and subtract that number from $n_g$.
-For each mutant cell, generate a new genotype and add one cell of this genotype to the population.
-The newly created genotype inherits mutations and ploidy karyotype from the parent genotype $g$, and in addition gains one more mutation, with the selected coefficient drawn from the exponential distribution $Exp(\lambda)$.
-The fitness of the newly created genotype is assigned according to the Formula \ref{eq:multiple-fitness}.
-	\item Aneuploidy gain step: \\
-	For each genotype $g$ that doesn’t already have an aneuploid chromosome, draw the number of new cells that gain aneuploidy from a binomial distribution $z \sim Bin(n_g, \alpha)$.
-If $z > 0$, then subtract it from $n_g$ and create a new genotype and add to the population with $z$ cells of this genotype.
-The genotype inherits all the mutations of genotype $g$ and in addition gains aneuploidy.
-The fitness is calculated appropriately by the Formula \ref{eq:multiple-fitness}.
-	\item Aneuploidy loss step: \\
-	It is similar to the aneuploidy step.
-For each genotype $g$ that has aneuploidy, draw the number of new cells that lose aneuploidy from the binomial distribution $z \sim Bin(n_g,\alpha_l)$.
-If $z>0$, subtract $z$ from $n_g$, then create new genotype, and add to the population $z$ cells of this genotype.
-The genotype inherits all the mutations of $g$ and in addition gains aneuploidy.
-The fitness is calculated appropriately by the Formula \ref{eq:multiple-fitness}.
-\end{enumerate}
+The effects of mutation and aneuploidy on genotype frequencies is more elaborate than in the single-locus model.
+Genotype $i$ is classified according to their karyotype (\emph{2n} or \emph{2n+1}), the number of accumulated beneficial mutations ($k \ge 0$), and their fitness ($w_i$).
+Each offspring cell inherits these properties from its mother cell.
+Then, with probability $\mu$ or $\tau \mu$ for euploid and aneuploid cells, respectively, a new beneficial mutation is accumulated, such that the number of mutations is $k+1$, and its effect $s_{k+1}$ is drawn from an exponential distribution with expected value $s$, such that the contribution of the mutations to the fitness is $\prod_{j=0}^{k+1}{(1+s_j)}$.
+Next, euploid offspring become aneuploid with probability $\delta$, and aneuploid offspring become euploid with probability $\delta_L$.
 
-\paragraph{Data.} We inference parameters of the models based on the data of the experiments of Yona at al ~\cite{Yona2012}.
-In \emph{heat-stress} experiment four populations of \emph{S.
-cerevisiae} were evolved under $39^{\circ}$ heat-stress and developed aneuploidy till 450 generations (here and further by development/loss of genotype trait we mean that 95\% of the population are the carriers of the genotype).
-Then, two populations were checked at generations 1700 and 2350, and for one population the aneuploidy were eliminated and has not observed allready at generation 1700 and for the second at generation 2350.
-\\
-Similarly, in \emph{ph-stress} experiment four populations were evolved unde high-ph stress conditions and developed aneuploidy during 150 generations.
-At generation 750 aneuploidy fully eliminated for two populations, and only partly for another two.
+%%% Table: multi-locus model
+\begin{table}[h]
+\centering
+\caption{\textbf{Multi-locus model fitness values.}}
+\begin{tabular}{lllll}
+\emph{Genotype} $i$ & $2n$ & $2n+1$ & $2n+1^{*k}$ & $2n^{*k}$ \\
+\hline
+\emph{Fitness} $w_i$ & $1$ & $1-c+\sigma$ & $(1-c)\prod_{j=1}^k(1+s_i)+\sigma$ & $\prod_{j=1}^k(1+s_i)$               
+\end{tabular}
+\label{table:multi-locus}
+\caption*{
+$k$ is the number of accumulated beneficial mutations in the genome;
+$s \ge 0$ is the selection coefficient of a beneficial mutation;
+$\sigma \ge 0$ is the selection coefficient of aneuploidy; 
+and $0 \le c \le 1$ is the fitness cost of aneuploidy.
+}
+\end{table}
+%%%
+
+\paragraph{Empirical evidence.}
+Our inference procedure uses empirical data from evolutionary experiments performed by \citet{Yona2012}.
+In the heat-stress experiment, four populations of \yeast evolved under \SI{39}{\celsius}.
+Aneuploidy fixed in all four population in the first 450 generations (hereafter, fixation or elimination of a genotype means that 95\% or 5\% of the population carry the genotype).
+The experiment continued with two populations, in which aneuploidy was eliminated by generation 1700 and 2350.
+In the pH-stress experiment, four populations of \yeast evolved under high-ph stress (8.6).
+Aneuploidy fixed during the first 150 generations.
+It was fully eliminated in two populations and partly eliminated in two populations by generation 750.
 
 \paragraph{Likelihood function.}
-Denote by $A_{t}$ the event of aneuploidy observation at generation $t$; by $L_{t}$ the event of aneuploidy loss observation at generation $t$; and by $L_{2350^*}$ the event of aneuploidy loss observation at generation 2350 while there's no loss in generation 1700.
-The likelihood function of the model with parameter set $\theta$ for \emph{heat-stress} experiment is:
+Denote by $A_{t}$ the fixation of aneuploidy at generation $t$, by $L_{t}$ the loss of aneuploidy at generation $t$, and by $L^*_{t,T}$ the loss of aneuploidy generation $t$ conditioned on no loss by generation $T$.
+The model likelihood function for parameter vector $\theta$ for the heat-stress experiment is
 \begin{multline} \label{eq:heatstress-likelihood}
-L(\theta | A_{450}, L_{1700}, L_{2350^*} ) = \\
-P^4(A_{450}) \cdot (1 - P^4(\lnot{L_{1700}} | A_{450} ) - P^4(\lnot{L_{2350^*}}| A_{450}) + P^4(\lnot{L_{1700}} \land \lnot{L_{2350^*}}| A_{450}) )		
+\likelihood(\theta | A_{450}, L_{1700}, L^*_{2350, 1700} ) = \\
+P^4(A_{450}) \cdot \Big(1 - P^4(\lnot{L_{1700}} | A_{450} ) - P^4(\lnot{L^*_{2350, 1700}}| A_{450}) + P^4(\lnot{L_{1700}} \land \lnot{L^*_{2350, 1700}}| A_{450})\Big)	\;,	
 \end{multline}
-The likelihood function of the model with parameter set $\theta$ for \emph{ph-stress} experiment is:
+where $P(X)$ is the probability of even $X$ and $\lnot$ means \emph{not}.
+The likelihood function of the model with parameter set $\theta$ for the pH-stress experiment is
 \begin{equation} \label{eq:phstress-likelihood}
- L(\theta | A_{150}, L_{750} ) = \\
- P^4(L_{450}) \cdot 6 \cdot	P^4(L_{750} | A_{150}) \cdot P^4(\lnot{L_{750}} | A_{150})
+\likelihood(\theta | A_{150}, L_{750} ) = \\
+ P^4(L_{450}) \cdot 6 \cdot	P^4(L_{750} | A_{150}) \cdot P^4(\lnot{L_{750}} | A_{150}) \;.
 \end{equation}

### 2021-06-22 ms/ms.tex
continuted with methods

@@
     \textbf{(B)} In the multi-locus model, each genotype is characterized by its karyotype, \euwt or \anwt, and the number of accumulated beneficial mutations, denoted by stars. 
-    In both panels arrows denote transitions between genotypes, with transitions rates: $\mu$, mutation rate; $\alpha$, aneuploidy gain rate; $\alpha_l$, aneuploidy loss rate.
+    In both panels arrows denote transitions between genotypes, with transitions rates: $\mu$, mutation rate; $\delta$, aneuploidy gain rate; $\delta_L$, aneuploidy loss rate.
   }
@@
 \hline
-\emph{Fitness} $w_i$ & $1$ & $1-c+\sigma$ & $(1-c)(1+s)+\sigma$ & $1+s$               
+\emph{Fitness} $w_i$ & $1$ & $1-c+b$ & $(1-c)(1+s)+b$ & $1+s$               
 \end{tabular}
@@
 $s \ge 0$ is the selection coefficient of a beneficial mutation;
-$\sigma \ge 0$ is the selection coefficient of aneuploidy; 
+$b \ge 0$ is the selection coefficient of aneuploidy; 
 and $0 \le c \le 1$ is the fitness cost of aneuploidy.
@@
 The selection coefficient of the $i$-th accumulated mutation in each individual, $s_i$, is drawn from an exponential distribution with expected value $s$, $s_i \sim Exp(s)$.
-The rest of the parameters ($N$, $\mu$, $\tau$, $\delta$, $\delta_L$, $\sigma$, $c$) are the same as in the single-locus model.
+The rest of the parameters ($N$, $\mu$, $\tau$, $\delta$, $\delta_L$, $b$, $c$) are the same as in the single-locus model.
 The fitness of the different genotypes is the same as in the single-locus model (\autoref{table:single-locus}), except that the fitness contribution of $k$ beneficial mutations is the product of their independent effects, $\prod_{i=1}^k(1+s_i)$, instead of the contribution of the single mutation allowed in the single-locus model, $(1+s)$, see \autoref{table:multi-locus}.
@@
 \hline
-\emph{Fitness} $w_i$ & $1$ & $1-c+\sigma$ & $(1-c)\prod_{j=1}^k(1+s_i)+\sigma$ & $\prod_{j=1}^k(1+s_i)$               
+\emph{Fitness} $w_i$ & $1$ & $1-c+b$ & $(1-c)\prod_{j=1}^k(1+s_i)+b$ & $\prod_{j=1}^k(1+s_i)$               
 \end{tabular}
@@
 It was fully eliminated in two populations and partly eliminated in two populations by generation 750.
+These empirical results were published by \citet{Yona2012}.
 
 \paragraph{Likelihood function.}
-Denote by $A_{t}$ the fixation of aneuploidy at generation $t$, by $L_{t}$ the loss of aneuploidy at generation $t$, and by $L^*_{t,T}$ the loss of aneuploidy generation $t$ conditioned on no loss by generation $T$.
+Denote by $A_{t}$ the fixation of aneuploidy at generation $t$, by $L_{t}$ the loss of aneuploidy at generation $t$, and by $L^*_{t,T}$ the loss of aneuploidy at generation $t$ conditioned on no loss by generation $T$.
 The model likelihood function for parameter vector $\theta$ for the heat-stress experiment is
@@
 \end{equation} 
+Because of the complexity of the model, this likelihood function is intractable.
+Therefore, given specific values of the model parameters $\theta$, we simulate the model in many replicated to approximate the value of the likelihood function $\likelihood(\theta)$.
 
-\paragraph{Parameter inference.} To estimate the parameters of the models we apply approximate Bayesian computation with a sequential Monte-Carlo scheme (ABC-SMC), employing pyabc Python 
-package \citep{Klinger2018}
+\paragraph{Parameter inference.} To infer model parameters, we use approximate Bayesian computation~\citep{Sunnaker2013} with a sequential Monte-Carlo scheme, or ABC-SMC~\citep{Sisson2009}, implemented in the \texttt{pyabc} Python
+package~\citep{Klinger2018}.
+Briefly, this approach uses numerical stochastic simulations of the model to infer a posterior distribution over the model parameters.
+This approach is applicable for likelihood-free, simulation-based inference~\citep{Cranmer2020}, that is, for estimating a posterior distribution when a likelihood function cannot be computed.
+It is therefore suitable to our case, in which the likelihood function can only be approximated from simulations, and cannot be directly computed.
+% TODO ILIA please add all configuration details, similar to the relevant section in the NPI paper. Discuss priors, convergence tests, etc.
+
+\paragraph{Prior distributions.} The prior distributions of $w_{2n+1}=1-c+b$, $w_{2n+1^*}=(1+s)(1-c)+b$ and $w_{2n^*}=1+s$ were obtained by estimating prior distributions for $1-c$, $1-c+b$ and $1+s$ from growth curves data.
+These growth curve data were previously obtained by~\citet{Yona2012} using mono-culture growth experiments.
+The raw growth curves data were not published before, but were used to produce figures 3C, 4A, and S2 in \citet{Yona2012}.
+We used \texttt{Curveball}, a dedicated method for predicting results of competition experiments from growth curve data~\citet{Ram2019}.
+\texttt{Curveball} takes growth curves of two strains growing separately in mono-culture and predicts how they would grow in a mixed-culture, that is, it predicts the results of a competition assay.
+From these predictions, relative fitness values can be computed.
+We used growth curves of $2n+1$ and $2n$ in \SI{39}{\celsius} and in \SI{30}{\celsius} to estimate $1-c+b$ and $1-c$, respectively.
+The latter assumes that the cost of aneuploidy is the same in \SI{39}{\celsius} and \SI{30}{\celsius}; this might not be correct, but we only assume this in order to generate a prior distribution for the fitness values.
+We also used growth curves of $2n+1$ and $2n^*$ in \SI{39}{\celsius} in \SI{39}{\celsius} to estimate $1+s$.
+Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix.
+Thus, we sampled 10,000 values for $1-c+b$, $1-c$, and $1+s$, which we used as prior distributions.
+Note that the first and last values correspond to $w_{2n+1}$ and $w_{2n}$, whereas $w_{2n+1^*}=(1-c)w_{2n^*}+w_{2n+1}-(1-c)$.
+%TODO ILIA please report here the point estimate and 95% CI for the three distributions, I only have for 1-c+b: 0.9863 (95% CI 0.9790 - 0.9939).
+%TODO ILIA describe the priors for the mutation and aneuploidy rate.
+
+\paragraph{Model comparison.}  % TODO ILIA what did you do for model comparison/selection?
 

### 2021-06-27 ms/ms.tex
added figures demonstrating growth curve estimation

@@
 \newcommand{\anmt}{\emph{2n+1*}}
+
+% Supplementary
+\newcommand{\beginsupplement}{%
+      	\setcounter{table}{0}
+        \renewcommand{\thetable}{S\arabic{table}}%
+        \setcounter{figure}{0}
+        \renewcommand{\thefigure}{S\arabic{figure}}%
+}
 
@@
 
-\paragraph{Parameter inference.} To infer model parameters, we use approximate Bayesian computation~\citep{Sunnaker2013} with a sequential Monte-Carlo scheme, or ABC-SMC~\citep{Sisson2009}, implemented in the \texttt{pyabc} Python
+\paragraph{Parameter inference.} To infer model parameters, we use approximate Bayesian computation~\citep{Sunnaker2013} with a sequential Monte-Carlo scheme, or ABC-SMC~\citep{Sisson2009}, implemented in the \texttt{pyABC}\footnote{https://pyabc.readthedocs.io} Python
 package~\citep{Klinger2018}.
 Briefly, this approach uses numerical stochastic simulations of the model to infer a posterior distribution over the model parameters.
-This approach is applicable for likelihood-free, simulation-based inference~\citep{Cranmer2020}, that is, for estimating a posterior distribution when a likelihood function cannot be computed.
+It is a method of likelihood-free, simulation-based inference~\citep{Cranmer2020}, that is, for estimating a posterior distribution when a likelihood function cannot be computed.
 It is therefore suitable to our case, in which the likelihood function can only be approximated from simulations, and cannot be directly computed.
@@
 
-\paragraph{Prior distributions.} The prior distributions of $w_{2n+1}=1-c+b$, $w_{2n+1^*}=(1+s)(1-c)+b$ and $w_{2n^*}=1+s$ were obtained by estimating prior distributions for $1-c$, $1-c+b$ and $1+s$ from growth curves data.
+\paragraph{Prior distributions.} The prior distributions of $w_{2n+1}=1-c+b$, $w_{2n+1^*}=(1+s)(1-c)+b$ and $w_{2n^*}=1+s$ were obtained by estimating $1-c$, $1-c+b$ and $1+s$ from growth curves data.
 These growth curve data were previously obtained by~\citet{Yona2012} using mono-culture growth experiments.
 The raw growth curves data were not published before, but were used to produce figures 3C, 4A, and S2 in \citet{Yona2012}.
-We used \texttt{Curveball}, a dedicated method for predicting results of competition experiments from growth curve data~\citet{Ram2019}.
-\texttt{Curveball} takes growth curves of two strains growing separately in mono-culture and predicts how they would grow in a mixed-culture, that is, it predicts the results of a competition assay.
+We used \texttt{Curveball}\footnote{https://curveball.yoavram.com}, a dedicated method for predicting results of competition experiments from growth curve data~\citep{Ram2019}.
+\texttt{Curveball} takes growth curves of two strains growing separately in mono-culture and predicts how they would grow in a mixed culture, that is, it predicts the results of a competition assay.
 From these predictions, relative fitness values can be computed.  
 We used growth curves of $2n+1$ and $2n$ in \SI{39}{\celsius} and in \SI{30}{\celsius} to estimate $1-c+b$ and $1-c$, respectively.
-The latter assumes that the cost of aneuploidy is the same in \SI{39}{\celsius} and \SI{30}{\celsius}; this might not be correct, but we only assume this in order to generate a prior distribution for the fitness values.
-We also used growth curves of $2n+1$ and $2n^*$ in \SI{39}{\celsius} in \SI{39}{\celsius} to estimate $1+s$.
+The latter assumes that the cost of aneuploidy is the same in \SI{39}{\celsius} and \SI{30}{\celsius}; this might be incorrect, but we only assume this to generate a prior distribution for the fitness values.
+We also used growth curves of $2n+1$ and $2n^*$ in \SI{39}{\celsius} to estimate $1+s$.
 Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix.
 Thus, we sampled 10,000 values for $1-c+b$, $1-c$, and $1+s$, which we used as prior distributions.
-Note that the first and last values correspond to $w_{2n+1}$ and $w_{2n}$, whereas $w_{2n+1^*}=(1-c)w_{2n^*}+w_{2n+1}-(1-c)$.
+Note that the first and last values correspond to $w_{2n+1}$ and $w_{2n}$, whereas $w_{2n+1^*}=(1-c)(w_{2n^*}-1)+w_{2n+1}$.
+See Figures~\ref{fig:growth-curves-30deg} and~\ref{fig:growth-curves-39deg}.
 %TODO ILIA please report here the point estimate and 95% CI for the three distributions, I only have for 1-c+b: 0.9863 (95% CI 0.9790 - 0.9939).
@@
 
+\section*{Supplementary Material}
+\beginsupplement % https://support.authorea.com/en-us/article/how-to-create-an-appendix-section-or-supplementary-information-1g25i5a/
+
+%%%% Fig - Curveball 30deg %%%
+% generated with growth_curves.ipynb
+\begin{figure}[h]
+    \centering
+	\includegraphics[width=0.5\textwidth]{../figures/evo39_fitness_30deg.pdf}
+    \caption{
+    \textbf{Fitness estimation from \SI{30}{\celsius}.}
+    } 
+    \label{fig:growth-curves-30deg}
+\end{figure}
+
+%%%% Fig - Curveball 39deg %%%
+% generated with growth_curves.ipynb
+\begin{figure}[h]
+    \centering
+	\includegraphics[width=0.5\textwidth]{../figures/evo39_fitness_39deg.pdf}
+    \caption{
+    \textbf{Fitness estimation from \SI{39}{\celsius}.}
+    } 
+    \label{fig:growth-curves-39deg}
+\end{figure}
+
 \end{document}

### 2021-06-30 ms/ms.tex
IK comments

@@
 Experimental evidence of such a transient role of aneuploidy was demonstrated by \citet{Yona2012}. They evolved populations of \yeast under strong heat or pH stress.
-In these experiments, the populations adapted to the stress within 450 generations, and this adaptation was determined to be due to chromosome duplications.
-Much later, after more than 1500 generations, the populations reverted back to an euploid state, while remaining adapted to the stress and accumulating multiple mutations.
+The populations adapted to the heat and pH stress within 450 and 150 generations, and this adaptation was determined to be due to chromosome duplications.
+Much later, after more than 1500 and 750 generations, for the heat and pH stress, respectively, the populations reverted back to an euploid state, while remaining adapted to the stress and accumulating multiple mutations.
 However, under gradual heat stress, aneuploidy was not observed.
@@
 Furthermore, we analyze these models to estimate the effects of parameters on the adaptation time and the probability for adaptation via aneuploidy.
-We find that % TODO CHECK!
-the aneuploidy rate is several orders of magnitude higher than the mutation rate; that a simple model of clonal interference is not enough to explain the transience of aneuploidy; % what does that mean?
-and that aneuploidy is likely to fix in the population and increase its mean fitness, but at the price of delaying further adaptation.
+We find that % TODO
 
@@
 We developed two models: a single-locus model and a multi-locus model. 
-Both models are based on the Wright-Fisher model~\citep{Otto2007}, assuming a constant population size $N$ with non-overlapping generations and including the effects of natural selection, genetic drift, aneuploidy, and mutation.
+Both models are based on the Wright-Fisher model~\citep{Otto2007}, assuming non-overlapping generations and including the effects of natural selection, genetic drift, aneuploidy, and mutation. 
 We focus on beneficial mutations, neglecting the effects of deleterious and neutral mutations. Both models allow for a single aneuploid karyotype (e.g., chromosome III duplication).
@@
 \paragraph*{Single-locus model.}
-This model follows four genotypes~(\autoref{fig:models}A): euploid wild-type, \euwt, the initial genotype; 
+This model assumes a constant population size $N$ and follows four genotypes~(\autoref{fig:models}A): euploid wild-type, \euwt, the initial genotype; 
 euploid mutant, \eumt, with the standard karyotype and a single beneficial mutation; 
@@
 By default, we assume $\tau=33/32$, as the wild-type \yeast strains in the experiments by \citep{Yona2012} are diploid, with 32 chromosomes, and the aneuploid strains are trisomic, with 33 chromosomes.
-Aneuploidy is formed by chromosome missegregation, so that cells transition from \euwt\; to \anwt\; or from \eumt\; to \anmt\; with probability $\delta$, the aneuploidy gain rate.
-Aneuploidy is lost, transitioning cells from \anwt\; to \euwt\; or from \anmt\; to \eumt\; with probability $\delta_L$, the aneuploidy loss rate.
+Aneuploidy is formed by chromosome mis-segregation, so that cells transition from \euwt\; to \anwt\ with probability $\delta$, the aneuploidy gain rate.
+Aneuploidy is lost, transitioning cells from \anmt\; to \eumt\; with probability $\delta_L$, the aneuploidy loss rate.
 The fitness values of the four genotypes are given by \autoref{table:single-locus}.
@@
     \begin{equation} \label{eq:drift-single}
-      \vec{f'} \sim \frac{1}{N} Mult(N, \vec{f}) \;,
+      \vec{f'} \sim \frac{1}{N} Mult(N, \vec{f^m}) \;,
     \end{equation}
-where $\vec{f}=(f_{2n},f_{2n+1},f_{2n+1^*},f_{2n^*})$ and $\vec{f'}$ are the frequencies of the genotypes in the current and next generation, respectively, and $Mult(N, \vec{f})$ is a multinomial distribution parameterized by the population size $N$ and the genotype frequencies $\vec{f}$.
+where $\vec{f^m}=(f_{2n},f_{2n+1},f_{2n+1^*},f_{2n^*})$ are the frequencies of the genotypes after mutation and aneuploidy, $\vec{f'}$ are the genotype frequencies in the next generation, and $Mult(N, \vec{f})$ is a multinomial distribution parameterized by the population size $N$ and the genotype frequencies $\vec{f}$.
 Overall, the change in genotype frequencies from one generation to the next is given by the transformation $f_i \to f'_i$.
@@
 \paragraph*{Multi-locus model.}
-This model expands the single-locus model by allowing for the accumulation of beneficial mutations in the genome and a fluctuating population size.
+This model expands the single-locus model by allowing for (i) the accumulation of beneficial mutations in the genome, and (ii) a fluctuating population size.
 
@@
 The rest of the parameters ($N$, $\mu$, $\tau$, $\delta$, $\delta_L$, $b$, $c$) are the same as in the single-locus model.
+% ILIA: I don’t know if it should be mentioned here, but the meaning of the mutation rate is a bit differ for multi-locus model.
+For the single-locus mutation rate value is the probability to get specific mutation (maybe one of a few).
+For the multi-locus it is the probability to get some beneficial mutation (one of a many).
+So, we don’t expect to receive same values for mutation rate for the models.
+% ILIA: I don’t have in the code 𝜏 parameter for multi-locus model (it’s like 1 there). Do you think it worth adding it for consistency?
 The fitness of the different genotypes is the same as in the single-locus model (\autoref{table:single-locus}), except that the fitness contribution of $k$ beneficial mutations is the product of their independent effects, $\prod_{i=1}^k(1+s_i)$, instead of the contribution of the single mutation allowed in the single-locus model, $(1+s)$, see \autoref{table:multi-locus}.
@@
 In contrast to the single-locus model, in the multi-locus model the population size changes in order to model serial-transfer experiment protocol~\citep{Yona2012}: the population is serially diluted by transferring a fraction of the population (1/120) to a fresh medium, starting a new growth cycle.
-In this model, the population initial size is $N_0 = N$, and the population size is doubled every generation, $N_1=2N, N_2=4N, \ldots$, and diluted back to $N$ at every seven generations, $N_7=N$.
+In this model, the population initial size is $N_0 = N$, and the population size is doubled every generation, $N_1=2N, N_2=4N, \ldots$, and diluted back to $N$ every eight generations, $N_8=N$.
 
@@
 In the heat-stress experiment, four populations of \yeast evolved under \SI{39}{\celsius}.
-Aneuploidy fixed in all four population in the first 450 generations (hereafter, fixation or elimination of a genotype means that 95\% or 5\% of the population carry the genotype).
+Aneuploidy fixed in all four population in the first 450 generations (hereafter, fixation or elimination of a genotype by generation $t$ means that more than 95\% or less than 5\% of the population carry the genotype at generation $t$, and possibly earlier).
 The experiment continued with two populations, in which aneuploidy was eliminated by generation 1700 and 2350.
@@
 Because of the complexity of the model, this likelihood function is intractable.
-Therefore, given specific values of the model parameters $\theta$, we simulate the model in many replicated to approximate the value of the likelihood function $\likelihood(\theta)$.
+Therefore, given specific values of the model parameters $\theta$, we simulate the model in many replicates to approximate the value of the likelihood function $\likelihood(\theta)$.
 

### 2021-07-01 ms/ms.tex
typo fix
typo

@@
     \end{equation}
-where $\vec{f^m}=(f_{2n},f_{2n+1},f_{2n+1^*},f_{2n^*})$ are the frequencies of the genotypes after mutation and aneuploidy, $\vec{f'}$ are the genotype frequencies in the next generation, and $Mult(N, \vec{f})$ is a multinomial distribution parameterized by the population size $N$ and the genotype frequencies $\vec{f}$.
+where $\vec{f^m}=(f^m_{2n},f^m_{2n+1},f^m_{2n+1^*},f^m_{2n^*})$ are the frequencies of the genotypes after mutation and aneuploidy, $\vec{f'}$ are the genotype frequencies in the next generation, and $Mult(N, \vec{f})$ is a multinomial distribution parameterized by the population size $N$ and the genotype frequencies $\vec{f}$.
 Overall, the change in genotype frequencies from one generation to the next is given by the transformation $f_i \to f'_i$.

### 2021-09-29 ms/ms.tex
added some todos for ilia
added some figures

--- 2021-06-30 ms/ms.tex
@@
 \usepackage{longtable}
-\usepackage{siunitx}
+\usepackage{gensymb,siunitx}
 
@@
 \end{multline}
-where $P(X)$ is the probability of even $X$ and $\lnot$ means \emph{not}.
+where $P(X)$ is the probability of event $X$ and $\lnot$ means \emph{not}.
 The likelihood function of the model with parameter set $\theta$ for the pH-stress experiment is
@@
  P^4(L_{450}) \cdot 6 \cdot	P^4(L_{750} | A_{150}) \cdot P^4(\lnot{L_{750}} | A_{150}) \;.
+%%TODO i think it is wrong because fig 5C shows that only 2 replicas are fixated after 650 generations (see Supplementary also)
 \end{equation} 
@@
 \section*{Results}
+
+First we evaluated simulations of single-locus model to inference the model parameters.
+See \ref{fig:posterior} for the posterior.
+The mutation rate (mean 1.71e-10 (95 CI..)) and trisomy rate (..) corresponds to the literature
+
+\begin{figure}[h]
+  \centering
+\includegraphics[width=0.9\textwidth]{../figures/posterior.pdf}
+  \caption{
+  \textbf{Single-locus model parameters posterior distribution}
+  The mean and the 95\% confindence interval of the parameters: 
+  mutation rate - 1.71e-10 (1.55e-10 - 1.88e-10),
+  trisomy rate - 3.3e-05 (2.94e-05 - 3.66e-05),
+  2n+1 fitness - 1.05 (1.05 - 1.05),
+  2n+1* fitness - 5.24 (4.21 - 6.27),
+  2n* fitness - 5.28 (4.24 - 6.32)
+  } 
+  \label{fig:posterior}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+\includegraphics[width=0.6\textwidth]{../figures/fixation-plot.pdf}
+  \caption{
+  \textbf{Time of 2n* fixation can't be explained without aneuploidy}
+   Distribution of time to fixation of 2n* and elimination of trisomy for $10,000$ simulations with parameters taken from the mean of the posterior. 
+  }
+  \label{fig:2n*-fixation}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+\includegraphics[width=0.6\textwidth]{../figures/ppc-plot.pdf}
+  \caption{
+  \textbf{Evolutionary dynamics with fitted single-locus model} 
+  Frequency of the four genotypes averaged over 10,000 simulations of the model. Each line represent one of the 10 parameter sets drawed from kde of the posterior.
+  }
+  \label{fig:ppc-plot}
+\end{figure}
+
+\begin{figure}[h!]
+  \centering
+\includegraphics[width=0.6\textwidth]{../figures/likelihood-plot.pdf}
+  \caption{
+  \textbf{Fit illustration for single locus-model} 
+  TODO
+  The best fit occurs when first dot is low, last dot is hight, and the inner rectangulars are queal...
+  }
+  \label{fig:likelihood}
+\end{figure}
 

### 2021-10-12 ms/ms.tex
added todo ref
some results for single-locus model were added

--- 2021-06-30 ms/ms.tex
@@
 %SetFonts
+
+\renewcommand{\baselinestretch}{1.5}
 
@@
 It is therefore suitable to our case, in which the likelihood function can only be approximated from simulations, and cannot be directly computed.
+After getting weighted particles of the posterior distribution using ABC-SMC scheme, we calculate Kernel Density Estimate using Gaussian kernels.
+$50,000$ sampled points from KDE represent the posterior distribution over the model parameters, and we use it for plots, median calculation, etc.
 % TODO ILIA please add all configuration details, similar to the relevant section in the NPI paper. Discuss priors, convergence tests, etc.
+% And also number of replicates for different models
 
@@
 \section*{Results}
+\paragraph{Single-locus model parameters inference.} 
+First we inference the parameters of the single-locus model. The mean and credible interval (2.5th percentile and the 97.5th percentile) of the parameters are:
+mutation rate - $1.02\substack{+2.40 \\ -1.34}\times10^{-6}$,
+trisomy rate - $7.36\substack{+3.55 \\ -4.82}\times10^{-4}$,
+2n+1 fitness - $1.022\substack{+0.004 \\ -0.004}$,
+2n+1* fitness - $1.025\substack{+0.004 \\ -0.004}$,
+2n* fitness - $1.028\substack{+0.004 \\ -0.004}$.
+This trisomy rate appropriates to the previous estimates; The mutation rate corresponds to the mutations target size of $10^{4}$ if we assume the mutation rate of single base to be about $2*10^{-10}$ as in previous estimates ~\citep{Zhu2014}.
+We can see from the posterior distribution~(\autoref{fig:posterior}) the correlation between fitness of different genotypes (\anwt, \anmt, \eumt)..(TODO say/plot something about the correlation, is it interesting?).
+
+\begin{figure}[h!]
+  \centering
+\includegraphics[width=0.9\textwidth]{../figures/posterior.pdf}
+  \caption{
+  \textbf{Single-locus model parameters posterior density.}
+%TODO histogram on diagonal
+The posterior density for each parameter is presented on the diagonal plots.
+Other plots represent joint posterior density of two parameters where blue color has the lowest density and yellow has the highest density.
+Red point and the dashed lines represent the median (TODO check that it is median and not mean) values of the densities:
+mutation rate - $1.02\times10^{-6}$,
+trisomy rate - $7.36\times10^{-4}$,
+\anwt fitness - $1.022$,
+\anmt fitness - $1.025$,
+\eumt fitness - $1.028$.
+  } 
+  \label{fig:posterior}
+\end{figure}
+
+\paragraph{Single-locus model fits the data well, but only if contains aneuploidy.} Single-locus model provide good fit to the data~(\autoref{fig:likelihood}A).
+On \autoref{fig:ppc-plot} the dynamics of frequency change of each genotype in time is shown. We can see that \anmt never reach substantial frequency in the population.  
+Sensitivity analysis~(\autoref{fig:sensitivity}) shows that if we change the parameter values the fit degrades.
+Furthermore, we measured fixation time of \anwt in population, and found it appropriate to the unpublished data of the experiment ~\citep{Yona2012} - about 300 generations in average.
+On the other hand, if we don't take into consideration aneuploidy in the model, i.e.
+take aneuploidy rate be equal to zero, then the model can't explain the data: the fit is not so good~(\autoref{fig:likelihood}A).
+Also we see that adaptation is faster without aneuploidy: for the best fit of the model without aneuploidy, the 2n* fixation occurs on average after 153 generation, while for the best fit of the model with aneuploidy it is 1637 generations.
+%TODO: put real data here
+
+\begin{figure}[h]
+  \centering
+\includegraphics[width=0.6\textwidth]{../figures/ppc-plot.pdf}
+  \caption{
+  \textbf{Evolutionary dynamics with fitted single-locus mode.l} 
+  Frequency of the four genotypes averaged over 10,000 simulations of the model. Each line represent one of the 10 parameter sets drawed from the inference posterior.
+  }
+  \label{fig:ppc-plot}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+\includegraphics[width=0.6\textwidth]{../figures/fixation-plot.pdf}
+  \caption{
+  \textbf{Time of 2n* fixation cannot be explained without aneuploidy.}
+   TODO: make plot like in poster that compares model with aneuploidy and without
+  }
+  \label{fig:2n*-fixation}
+\end{figure}
 
@@
 
-%%%% Fig - Curveball 39deg %%%
-% generated with growth_curves.ipynb
-\begin{figure}[h]
+% likelihood plots
+\begin{figure}[b!]
   \centering
-	\includegraphics[width=0.5\textwidth]{../figures/evo39_fitness_39deg.pdf}
+  \begin{subfigure}[t]{0.5\textwidth}
       \caption{
-    \textbf{Fitness estimation from \SI{39}{\celsius}.}
+        \textbf{A. With aneuploidy}
       }
-    \label{fig:growth-curves-39deg}
+      \centering
+      \includegraphics[height=1.8in]{../figures/likelihood-plot.pdf}      
+      \label{fig:likelihood-with-aneuploidy}
+  \end{subfigure}%
+  \\
+  \begin{subfigure}[t]{0.5\textwidth}
+  	  \caption{
+        \textbf{B. Without aneuploidy}
+      }
+      \centering
+      \includegraphics[height=1.8in]{../figures/likelihood-plot.pdf}      
+      \label{fig:likelihood-without-aneuploidy}
+  \end{subfigure}
+  \caption{
+    \textbf{Single-locus model fit.} Cumulative distribution of 10000 simulations are shown.
+    \textbf{(A)} Simulations parameters are from the median of the posterior of the single-locus model fit.
+In close to 50\% of simulations \eumt fixation occurs between 450 and 1700 generations and in about 50\% \eumt fixation occurs between 1700 and 2350 generatyions.
+This results in high likelihood.
+    \textbf{(B)} Simulations parameters are from the median of the posterior of the single-locus model \emph{without} aneuploidy fit. TODO
+  }
+  \label{fig:likelihood}
+\end{figure}
+
+% sensitivity analysis plots
+
+\begin{figure}[p]
+  \centering
+  \begin{subfigure}{0.3\textwidth}
+      \includegraphics[width=\textwidth]{../figures/sensitivity-mutation-rate-plot.pdf}      
+      \label{fig:sensitivity-mutation}
+  \end{subfigure}
+  \begin{subfigure}{0.3\textwidth}
+      \includegraphics[width=\textwidth]{../figures/sensitivity-trisomy-rate-plot.pdf}      
+      \label{fig:sensitivity-aneuploidy}
+  \end{subfigure}
+   \begin{subfigure}{0.3\textwidth}
+      \includegraphics[width=\textwidth]{../figures/sensitivity-w(2n+1)-plot.pdf}      
+      \label{fig:sensitivity-anwt}
+  \end{subfigure}
+    \begin{subfigure}{0.3\textwidth}
+      \includegraphics[width=\textwidth]{../figures/sensitivity-w(2n+1*)-plot.pdf}      
+      \label{fig:sensitivity-anmt}
+  \end{subfigure}
+    \begin{subfigure}{0.3\textwidth}
+      \includegraphics[width=\textwidth]{../figures/sensitivity-w(2n*)-plot.pdf}      
+      \label{fig:sensitivity-eumt}
+  \end{subfigure}
+  \caption{
+    \textbf{Sensitivity analysis.} We take parameters that are median of the posterior.
+Then we change only one parameter at time while measuring the model likelihood of the parameters.
+x-axis label represent the parameter that we change.
+Yellow point indicates the median value.
+  }
+  \label{fig:sensitivity}
 \end{figure}

### 2021-10-12 ms/ms.tex

@@
 % TODO cite https://www.biorxiv.org/content/10.1101/2020.10.07.327759v1
+% TODO cite Yang, F. et al. The fitness costs and benefits of trisomy of each Candida albicans chromosome. Genetics 218, 1–7 (2021). https://academic.oup.com/genetics/article/218/2/iyab056/6218773
 
@@
 
+\subsection*{Model fitting}
+
+\paragraph{Simple model: Inference.}
+%tell point estimates (+- hdi) (fig 2 from poster, add hdi)
+%posterior predictions, sensitivity analysis
+%time to fixation of aneuploidy 2n+1 - fits unpublished data
+%figures
+%- posterior distributions
+%- expected frequency over time (fig 1 from poster)
+%- % simulations to reach 2n+1, 2n*. (fig 3 from poster)
+%- time to 2n* cannot be explained without aneuploidy (fig 4 from poster)
+%- sensitivity analysis (fig 5 from poster)
+%- convergence/ess of pyabc? 
+%  - plot of posterior in each iteration shows convergence; 
+%  - robustness of posterior to initial conditions - MAP+-HDI for different seeds;
+
+\paragraph{Complex model: Inference and Comparison.}
+
+%same for complex model
+%model comparison between models - model probabilities
+
 \pagebreak

### 2021-10-18 ms/ms.tex
merge fix
added fixation-plot and updated results with it

@@
 \section*{Results}
+
+\subsection*{Model fitting}
+
+%\paragraph{Simple model: Inference.}
+%tell point estimates (+- hdi) (fig 2 from poster, add hdi)
+%posterior predictions, sensitivity analysis
+%time to fixation of aneuploidy 2n+1 - fits unpublished data
+%figures
+%- posterior distributions
+%- expected frequency over time (fig 1 from poster)
+%- % simulations to reach 2n+1, 2n*. (fig 3 from poster)
+%- time to 2n* cannot be explained without aneuploidy (fig 4 from poster)
+%- sensitivity analysis (fig 5 from poster)
+%- convergence/ess of pyabc? 
+%  - plot of posterior in each iteration shows convergence; 
+%  - robustness of posterior to initial conditions - MAP+-HDI for different seeds;
+
 \paragraph{Single-locus model parameters inference.} 
@@
 On the other hand, if we don't take into consideration aneuploidy in the model, i.e.
-take aneuploidy rate be equal to zero, then the model can't explain the data: the fit is not so good~(\autoref{fig:likelihood}A).
-Also we see that adaptation is faster without aneuploidy: for the best fit of the model without aneuploidy, the 2n* fixation occurs on average after 153 generation, while for the best fit of the model with aneuploidy it is 1637 generations.
-%TODO: put real data here
+take aneuploidy rate be equal to zero, then the model can't explain the data: the fit is not so good~(\autoref{fig:likelihood}A; \autoref{fig:2n*-fixation}).
+The good fit of the model without aneuploidy to the data, requires much lower mutation rate (TODO provide numbers for the median fit?).
+If mutation rate is not low, the fixation of \eumt happens much quicker (see yellow and blue bars at  \autoref{fig:2n*-fixation}).
 
-\begin{figure}[h]
+\begin{figure}[h!]
   \centering
@@
   \caption{
-  \textbf{Evolutionary dynamics with fitted single-locus mode.l} 
+  \textbf{Evolutionary dynamics with fitted single-locus model.} 
   Frequency of the four genotypes averaged over 10,000 simulations of the model. Each line represent one of the 10 parameter sets drawed from the inference posterior.
@@
 
-\begin{figure}[h]
+\begin{figure}[h!]
   \centering
@@
   \caption{
-  \textbf{Time of 2n* fixation cannot be explained without aneuploidy.}
-   TODO: make plot like in poster that compares model with aneuploidy and without
+  \textbf{Time of \eumt fixation cannot be explained without aneuploidy.}
+Each color represents distribution to fixation of \eumt of $10,000$ simulations for different modifications of single-locus model.
+\emph{Blue} represents the model with aneuploidy with parameters that taken from posterior median when we fit the model to the experimental data.
+\emph{Orange} represents the same model that have no aneuploidy, i.e aneuploidy rate is zero.
+\emph{Green} represents the model without aneuploidy with parameters that taken from posterior median when we fit the model without aneuploidy to the experimental data.
+The last bin contains all the simulations with time greater than 3000.
+Dashed lines show experimental data: generations 1700 and 2350.
+The fit to the data is better when the distribution is such that the difference in number of simulations in ranges 0-1700 and 1700-2350 is small and there's no many simulations after 2350 generations.
+The fit of \emph{blue} is better than of \emph{green}.
+And there's no fit of \emph{red} to the data at all.
   }
@@
 \end{figure}
-
-\subsection*{Model fitting}
-
-\paragraph{Simple model: Inference.}
-%tell point estimates (+- hdi) (fig 2 from poster, add hdi)
-%posterior predictions, sensitivity analysis
-%time to fixation of aneuploidy 2n+1 - fits unpublished data
-%figures
-%- posterior distributions
-%- expected frequency over time (fig 1 from poster)
-%- % simulations to reach 2n+1, 2n*. (fig 3 from poster)
-%- time to 2n* cannot be explained without aneuploidy (fig 4 from poster)
-%- sensitivity analysis (fig 5 from poster)
-%- convergence/ess of pyabc? 
-%  - plot of posterior in each iteration shows convergence; 
-%  - robustness of posterior to initial conditions - MAP+-HDI for different seeds;
 

### 2021-10-20 ms/ms.tex
small todo was added about priors of the model without aneuploidy
added likelihood function for no aneuploidy model, updated likelihood figure

@@
 where $P(X)$ is the probability of event $X$ and $\lnot$ means \emph{not}.
+
 The likelihood function of the model with parameter set $\theta$ for the pH-stress experiment is
@@
  P^4(L_{450}) \cdot 6 \cdot	P^4(L_{750} | A_{150}) \cdot P^4(\lnot{L_{750}} | A_{150}) \;.  
-%%TODO i think it is wrong because fig 5C shows that only 2 replicas are fixated after 650 generations (see Supplementary also)
  \end{equation} 
+%TODO i think it is wrong because fig 5C shows that only 2 replicas are fixated after 650 generations (see Supplementary also) Anyway, I don't think that we use this likelihood, so probably should be removed at all.
+
+For the model without aneuploidy let's denote by $M_{t}$ the fixation of \eumt at generation $t$, and by $M^*_{t,T}$ the fixation of \eumt at generation $t$ conditioned on no fixation at generation $T$.
+The likelihood of the model without aneuploidy with parameter vector $\theta$ for the heat-stress experiment is
+\begin{multline} \label{eq:heatstress-noaneuploidy-likelihood}
+\likelihood(\theta |  M_{1700}, M^*_{2350, 1700} ) = \\
+\Big(1 - P^4(\lnot{M_{1700}}) - P^4(\lnot{M^*_{2350, 1700}}) + P^4(\lnot{M_{1700}} \land \lnot{M^*_{2350, 1700}})\Big)	\;,	
+\end{multline}
+
 Because of the complexity of the model, this likelihood function is intractable.
@@
 %TODO ILIA describe the priors for the mutation and aneuploidy rate.
+% TODO also describe the prior for the model without aneuploidy - the mutation rate is wider here.
 
@@
       \centering
-      \includegraphics[height=1.8in]{../figures/likelihood-plot.pdf}      
+      \includegraphics[height=1.8in]{../figures/likelihood-noaneuploidy-plot.pdf}      
       \label{fig:likelihood-without-aneuploidy}
@@
     \textbf{Single-locus model fit.} Cumulative distribution of 10000 simulations are shown.
-    \textbf{(A)} Simulations parameters are from the median of the posterior of the single-locus model fit.
-In close to 50\% of simulations \eumt fixation occurs between 450 and 1700 generations and in about 50\% \eumt fixation occurs between 1700 and 2350 generatyions.
-This results in high likelihood.
-    \textbf{(B)} Simulations parameters are from the median of the posterior of the single-locus model \emph{without} aneuploidy fit. TODO
+    \textbf{(A)} Simulations parameters are from the median of the posterior of the single-locus model fit. The likelihood for the heat-stress experiment (\autoref{eq:heatstress-likelihood}) is 0.86.
+    \textbf{(B)} Simulations parameters are from the median of the posterior of the single-locus model \emph{without} aneuploidy fit.
+The likelihood for the heat-stress experiment (\autoref{eq:heatstress-noaneuploidy-likelihood}) is 0.75.
   }

### 2021-10-25 ms/ms.tex
updated prior text for fitnesses

@@
 
-\paragraph{Prior distributions.} The prior distributions of $w_{2n+1}=1-c+b$, $w_{2n+1^*}=(1+s)(1-c)+b$ and $w_{2n^*}=1+s$ were obtained by estimating $1-c$, $1-c+b$ and $1+s$ from growth curves data.
-These growth curve data were previously obtained by~\citet{Yona2012} using mono-culture growth experiments.
+\paragraph{Prior distributions.} The prior distributions of $w_{2n+1}=1-c+b$, $w_{2n+1^*}=(1+s)(1-c)+b$ and $w_{2n^*}=1+s$ were obtained by estimating it from growth curves data that were previously obtained by~\citet{Yona2012} using mono-culture growth experiments.
 The raw growth curves data were not published before, but were used to produce figures 3C, 4A, and S2 in \citet{Yona2012}.
@@
 From these predictions, relative fitness values can be computed.
-We used growth curves of $2n+1$ and $2n$ in \SI{39}{\celsius} and in \SI{30}{\celsius} to estimate $1-c+b$ and $1-c$, respectively.
-The latter assumes that the cost of aneuploidy is the same in \SI{39}{\celsius} and \SI{30}{\celsius}; this might be incorrect, but we only assume this to generate a prior distribution for the fitness values.
-We also used growth curves of $2n+1$ and $2n^*$ in \SI{39}{\celsius} to estimate $1+s$.
+
+We used growth curves of $2n+1$ and $2n$ in \SI{39}{\celsius} to estimate $1-c+b$.
+We used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$, assuming that fitness values shouldn't be differ too much.
+We also tested different priors, but this one (let's call it main prior model) was better in sense of WAIC, prediction plots, estimated parameters.
+%TODO provide formula of WAIC and the results.
+To show ppc plots also?.
+\\We tested uninformative uniform prior, $\mathit{Uniform}(1,6)$, for all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ or only for $w_{2n+1^*}$, $w_{2n^*}$.
+In both scenarios the estimated by the model parameters of fitness were looks too high: for the first scenario the estimated medians were  $w_{2n+1}=2.767$, $w_{2n+1^*}=5.79$, $w_{2n^*}=5.827$ and for the second $w_{2n+1}=1.046$, $w_{2n+1^*}=4.642$, $w_{2n^*}=4.684$.
+We don't expect to have such high fitness values.
+Moreover the WAIC for the models with these priors are 0.57 and 1.9 respectively, while WAIC for the main prior model is 0.27.
+Also for the model with uninformative prior for all fitnesses the fixation time of 2n+1 is about 16 in average, that contradicts unpublished observations that were about 250-300.
+\\Also we tried to use additional growth curves.
+We used growth curves of 2n* (\emph{refined} from \citep{Yona2012}) and 2n+1 in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$.
+The same prior was used for $w_{2n*}/w_{2n+1*}$.
+This model not perfect well with WAIC 2.8.
+\\Another try was to use growth curves of $2n+1$ and $2n$ in \SI{30}{\celsius} to estimate $1-c$.
+This estimation assumes that the cost of aneuploidy is the same in \SI{39}{\celsius} and \SI{30}{\celsius}; this might be incorrect, but we only assumed this to generate a prior distribution for the fitness values.
+The prior for $b$ was taken the same as for $c$.
+This model didn't work at all: we didn't receive any parameter that the likelihood with it is greater than zero.
+%TODO it make sense because bla bla?
+
 Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix.
-Thus, we sampled 10,000 values for $1-c+b$, $1-c$, and $1+s$, which we used as prior distributions.
-Note that the first and last values correspond to $w_{2n+1}$ and $w_{2n}$, whereas $w_{2n+1^*}=(1-c)(w_{2n^*}-1)+w_{2n+1}$.
+Thus, we sampled 10,000 values for appropriate fitness, which we used as prior distributions.
 See Figures~\ref{fig:growth-curves-30deg} and~\ref{fig:growth-curves-39deg}.
@@
 
+%%%% Fig - Curveball 39deg %%%
+% generated with growth_curves.ipynb
+\begin{figure}[h]
+    \centering
+	\includegraphics[width=0.5\textwidth]{../figures/evo39_fitness_39deg.pdf}
+    \caption{
+    \textbf{Fitness estimation from \SI{39}{\celsius}.}
+    } 
+    \label{fig:growth-curves-39deg}
+\end{figure}
+
 % likelihood plots

### 2021-10-26 ms/ms.tex
bit more for prior text

@@
 
-We used growth curves of $2n+1$ and $2n$ in \SI{39}{\celsius} to estimate $1-c+b$.
+We used growth curves of $2n+1$ and $2n$ in \SI{39}{\celsius} to estimate  $w_{2n+1}$.
 We used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$, assuming that fitness values shouldn't be differ too much.
-We also tested different priors, but this one (let's call it main prior model) was better in sense of WAIC, prediction plots, estimated parameters.
-%TODO provide formula of WAIC and the results.
+We also tested different priors, but this one (let's call it main prior) was better in sense of WAIC, prediction plots, estimated parameters.
+%TODO provide formula of WAIC.
+Table?
 To show ppc plots also?.
-\\We tested uninformative uniform prior, $\mathit{Uniform}(1,6)$, for all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ or only for $w_{2n+1^*}$, $w_{2n^*}$.
+\\We tested uninformative uniform prior, $\mathit{Uniform}(1,6)$, for all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ and only for $w_{2n+1^*}$, $w_{2n^*}$, leaving $w_{2n+1}$ prior as before .
 In both scenarios the estimated by the model parameters of fitness were looks too high: for the first scenario the estimated medians were  $w_{2n+1}=2.767$, $w_{2n+1^*}=5.79$, $w_{2n^*}=5.827$ and for the second $w_{2n+1}=1.046$, $w_{2n+1^*}=4.642$, $w_{2n^*}=4.684$.
@@
 Moreover the WAIC for the models with these priors are 0.57 and 1.9 respectively, while WAIC for the main prior model is 0.27.
-Also for the model with uninformative prior for all fitnesses the fixation time of 2n+1 is about 16 in average, that contradicts unpublished observations that were about 250-300.
+Also for the model with uninformative prior for all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ the fixation time of 2n+1 is about 16 in average, that contradicts unpublished observations that were about 250-300.
 \\Also we tried to use additional growth curves.
-We used growth curves of 2n* (\emph{refined} from \citep{Yona2012}) and 2n+1 in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$.
+We used growth curves of 2n* (\emph{refined} from \citet{Yona2012}) and 2n+1 in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$.
 The same prior was used for $w_{2n*}/w_{2n+1*}$.
-This model not perfect well with WAIC 2.8.
+This model didn't perform well, resulted in WAIC 2.8.
 \\Another try was to use growth curves of $2n+1$ and $2n$ in \SI{30}{\celsius} to estimate $1-c$.
@@
 See Figures~\ref{fig:growth-curves-30deg} and~\ref{fig:growth-curves-39deg}.
+
+For the mutation rate $\mu$ and aneuploidy rate $\delta$ we used uniform priors, $\mathit{Uniform}(1e-8,1e-5)$ and $\mathit{Uniform}(1e-6,1e-3)$ respectively.
+We have tried also uniform priors with lower than $1e-8$ bound for $\mu$, but anyway the posterior ends with higher values.
+For the single locus model without aneuploidy, for the mutation rate wider uniform prior was used - $\mathit{Uniform}(1e-10,1e-5)$. 
 %TODO ILIA please report here the point estimate and 95% CI for the three distributions, I only have for 1-c+b: 0.9863 (95% CI 0.9790 - 0.9939).

### 2021-11-08 ms/ms.tex
added config description for pyabc

@@
 It is therefore suitable to our case, in which the likelihood function can only be approximated from simulations, and cannot be directly computed.
-After getting weighted particles of the posterior distribution using ABC-SMC scheme, we calculate Kernel Density Estimate using Gaussian kernels.
+%adapted description from Syga et al. 2021:
+In short, the algorithm employs sequential importance sampling over generations.
+In generation $t$, the algorithm draws sets of parameters $\theta$ from a given proposal distribution and consequently simulates data $C$ from the model until $n_t$ instances were accepted ($n_t$ "particles") based on the comparison to observed data $C_{obs}$ via a distance function $D(C, C_obs)$ and acceptance threshold $\epsilon_t$, such that $D(C, C_objs)\leq{\epsilon_t}$.
+As the distance function we chose $1-\likelihood$.
+For the acceptance threshold the median of the distances of the accepted instances of the previous generation were applied (for the first generation the acceptance threshold is defined $0.01$).
+When $n_t$ instances were accepted new proposal distribution is constructed by multivariate normal kernel with adaptive covariance matrix.
+In the first generation the proposal distribution equals to the prior.
+The size of $n_t$ was determined by adaptive population strategy~\citep{Klinger2017}.
+We stop the iterations, when no more sufficient change in acceptance threshold occur, and no change in posterior.
+After receiving weighted particles of the posterior distribution using ABC-SMC scheme, we calculate Kernel Density Estimate using Gaussian kernels.
 $50,000$ sampled points from KDE represent the posterior distribution over the model parameters, and we use it for plots, median calculation, etc.
-% TODO ILIA please add all configuration details, similar to the relevant section in the NPI paper. Discuss priors, convergence tests, etc.
-% And also number of replicates for different models
+% TODO convergence tests, etc. Number of replicates for different models.
 
@@
 %TODO ILIA please report here the point estimate and 95% CI for the three distributions, I only have for 1-c+b: 0.9863 (95% CI 0.9790 - 0.9939).
-%TODO ILIA describe the priors for the mutation and aneuploidy rate.
-% TODO also describe the prior for the model without aneuploidy - the mutation rate is wider here.
 

### 2021-11-17 ms/ms.tex
small grammar fixes

@@
 It is therefore suitable to our case, in which the likelihood function can only be approximated from simulations, and cannot be directly computed.
-%adapted description from Syga et al. 2021:
+%adapted/expanded description from Syga et al. 2021:
 In short, the algorithm employs sequential importance sampling over generations. 
-In generation $t$, the algorithm draws sets of parameters $\theta$ from a given proposal distribution and consequently simulates data $C$ from the model until $n_t$ instances were accepted ($n_t$ "particles") based on the comparison to observed data $C_{obs}$ via a distance function $D(C, C_obs)$ and acceptance threshold $\epsilon_t$, such that $D(C, C_objs)\leq{\epsilon_t}$.
-As the distance function we chose $1-\likelihood$.
-For the acceptance threshold the median of the distances of the accepted instances of the previous generation were applied (for the first generation the acceptance threshold is defined $0.01$).
-When $n_t$ instances were accepted new proposal distribution is constructed by multivariate normal kernel with adaptive covariance matrix.
+In generation $t$, the algorithm draws sets of parameters $\theta_i$ from a given proposal distribution and consequently simulates data $C_i$ from the model until $n_t$ instances were accepted based on the comparison to observed data $C_{obs}$ via a distance function $D(C_i, C_{obs})$ and acceptance threshold $\epsilon_t$, such that $D(C_i, C_{objs})\leq{\epsilon_t}$.
+As the distance function we choose $1-\likelihood$.
+For the acceptance threshold the median of the distances of the accepted instances of the previous generation are applied (for the first generation the acceptance threshold is defined $0.01$).
+When $n_t$ instances are accepted new proposal distribution is constructed by multivariate normal kernel with adaptive covariance matrix.
 In the first generation the proposal distribution equals to the prior.
-The size of $n_t$ was determined by adaptive population strategy~\citep{Klinger2017}.
+The size of $n_t$ is determined by adaptive population strategy~\citep{Klinger2017}.
 We stop the iterations, when no more sufficient change in acceptance threshold occur, and no change in posterior.
@@
 We used growth curves of $2n+1$ and $2n$ in \SI{39}{\celsius} to estimate  $w_{2n+1}$.
-We used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$, assuming that fitness values shouldn't be differ too much.
-We also tested different priors, but this one (let's call it main prior) was better in sense of WAIC, prediction plots, estimated parameters.
+We used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$, assuming that fitness values shouldn't differ too much.
+We also tested different priors, but this one (let's call it main prior) was better in the sense of WAIC, prediction plots, estimated parameters.
 %TODO provide formula of WAIC.
@@
 To show ppc plots also?.
-\\We tested uninformative uniform prior, $\mathit{Uniform}(1,6)$, for all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ and only for $w_{2n+1^*}$, $w_{2n^*}$, leaving $w_{2n+1}$ prior as before .
-In both scenarios the estimated by the model parameters of fitness were looks too high: for the first scenario the estimated medians were  $w_{2n+1}=2.767$, $w_{2n+1^*}=5.79$, $w_{2n^*}=5.827$ and for the second $w_{2n+1}=1.046$, $w_{2n+1^*}=4.642$, $w_{2n^*}=4.684$.
+\\We tested uniform prior (let's call it uninformative), $\mathit{Uniform}(1,6)$, for all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ and only for $w_{2n+1^*}$, $w_{2n^*}$, leaving $w_{2n+1}$ prior as before.
+In both scenarios the estimates by the model parameters of fitness looks too high: for the first scenario the median estimates were  $w_{2n+1}=2.767$, $w_{2n+1^*}=5.79$, $w_{2n^*}=5.827$ and for the second $w_{2n+1}=1.046$, $w_{2n+1^*}=4.642$, $w_{2n^*}=4.684$.
 We don't expect to have such high fitness values.
 Moreover the WAIC for the models with these priors are 0.57 and 1.9 respectively, while WAIC for the main prior model is 0.27.
-Also for the model with uninformative prior for all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ the fixation time of 2n+1 is about 16 in average, that contradicts unpublished observations that were about 250-300.
+Also for the model with uninformative prior for all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ the fixation time of 2n+1 is about 16 in average, which contradicts unpublished observations that were about 250-300.
 \\Also we tried to use additional growth curves.
@@
 The same prior was used for $w_{2n*}/w_{2n+1*}$.
-This model didn't perform well, resulted in WAIC 2.8.
+This model didn't perform well, it resulted in WAIC 2.8.
 \\Another try was to use growth curves of $2n+1$ and $2n$ in \SI{30}{\celsius} to estimate $1-c$.
@@
 The prior for $b$ was taken the same as for $c$.
-This model didn't work at all: we didn't receive any parameter that the likelihood with it is greater than zero.
+This model didn't work at all: no parameter sets were found with the likelihood greater than zero.
 %TODO it make sense because bla bla?
 
-Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix.
+Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution, defined by the maximum-likelihood covariance matrix.
 Thus, we sampled 10,000 values for appropriate fitness, which we used as prior distributions.
@@
 For the mutation rate $\mu$ and aneuploidy rate $\delta$ we used uniform priors, $\mathit{Uniform}(1e-8,1e-5)$ and $\mathit{Uniform}(1e-6,1e-3)$ respectively.
-We have tried also uniform priors with lower than $1e-8$ bound for $\mu$, but anyway the posterior ends with higher values.
-For the single locus model without aneuploidy, for the mutation rate wider uniform prior was used - $\mathit{Uniform}(1e-10,1e-5)$. 
+We tried also uniform priors with lower than the $1e-8$ bound for $\mu$, but anyway the posterior ended with higher values.
+For the single locus model without aneuploidy, wider uniform prior for the mutation rate was used - $\mathit{Uniform}(1e-10,1e-5)$. 
 %TODO ILIA please report here the point estimate and 95% CI for the three distributions, I only have for 1-c+b: 0.9863 (95% CI 0.9790 - 0.9939).
+\paragraph{Model comparison.}  % TODO ILIA what did you do for model comparison/selection? WAIC
 
-\paragraph{Model comparison.}  % TODO ILIA what did you do for model comparison/selection?
-
-\pagebreak
 % Results

### 2021-11-20 ms/ms.tex
some more fixes to the tex

@@
 We fit these models to the experimental results of \citet{Yona2012} using an \emph{approximate Bayesian computation} framework~\citep{Sisson2009, Klinger2018} to infer model parameters, including selection coefficients and rates of aneuploidy and mutation, and to perform model selection between different models, thereby testing different hypotheses about the evolutionary process.
+%TODO do model selection by comparing Bayes factor or ignore this part.
 Furthermore, we analyze these models to estimate the effects of parameters on the adaptation time and the probability for adaptation via aneuploidy.
@@
     \textbf{(A)} In the single-locus model, the four genotypes are: euploid wild-type, \euwt; euploid mutant, \eumt; aneuploid wild-type, \anwt; and aneuploid mutant, \anmt.
-    Overall there are two possible trajectories from \euwt to \eumt.
-    \textbf{(B)} In the multi-locus model, each genotype is characterized by its karyotype, \euwt or \anwt, and the number of accumulated beneficial mutations, denoted by stars. 
+    Overall there are two possible trajectories from \euwt\ to \eumt.
+    \textbf{(B)} In the multi-locus model, each genotype is characterized by its karyotype, \euwt\ or \anwt, and the number of accumulated beneficial mutations, denoted by stars. 
     In both panels arrows denote transitions between genotypes, with transitions rates: $\mu$, mutation rate; $\delta$, aneuploidy gain rate; $\delta_L$, aneuploidy loss rate.
@@
 The likelihood function of the model with parameter set $\theta$ for the pH-stress experiment is
-\begin{equation} \label{eq:phstress-likelihood}
+\begin{equation} \label{eq:phstress-likelihood} % TODO: add ph-Stress model to results or remove it from here
 \likelihood(\theta | A_{150}, L_{750} ) = \\
@@
 
-For the model without aneuploidy let's denote by $M_{t}$ the fixation of \eumt at generation $t$, and by $M^*_{t,T}$ the fixation of \eumt at generation $t$ conditioned on no fixation at generation $T$.
+For the model without aneuploidy, let's denote by $M_{t}$, the fixation of \eumt\ at generation $t$, and by $M^*_{t,T}$ the fixation of \eumt\ at generation $t$ conditioned on no fixation at generation $T$.
 The likelihood of the model without aneuploidy with parameter vector $\theta$ for the heat-stress experiment is
@@
 
-Because of the complexity of the model, this likelihood function is intractable.
+Because of the complexity of the model, these likelihood functions are intractable.
 Therefore, given specific values of the model parameters $\theta$, we simulate the model in many replicates to approximate the value of the likelihood function $\likelihood(\theta)$.
@@
 It is therefore suitable to our case, in which the likelihood function can only be approximated from simulations, and cannot be directly computed.
-%adapted/expanded description from Syga et al. 2021:
+%a bit adapted/expanded description from Syga et al. 2021:
 In short, the algorithm employs sequential importance sampling over generations. 
-In generation $t$, the algorithm draws sets of parameters $\theta_i$ from a given proposal distribution and consequently simulates data $C_i$ from the model until $n_t$ instances were accepted based on the comparison to observed data $C_{obs}$ via a distance function $D(C_i, C_{obs})$ and acceptance threshold $\epsilon_t$, such that $D(C_i, C_{objs})\leq{\epsilon_t}$.
+In generation $t$, the algorithm draws sets of parameters $\theta_i$ from a given proposal distribution and consequently simulates data $C_i$ from the model until $n_t$ instances were accepted based on the comparison to observed data $C_{obs}$ via a distance function $D(C_i, C_{obs})$ and acceptance threshold $\epsilon_t$, such that $D(C_i, C_{obs})\leq{\epsilon_t}$.
 As the distance function we choose $1-\likelihood$.
-For the acceptance threshold the median of the distances of the accepted instances of the previous generation are applied (for the first generation the acceptance threshold is defined $0.01$).
+For the acceptance threshold the median of the distances of the accepted instances of the previous generation are applied (for the first generation we define the acceptance threshold be $0.01$).
 When $n_t$ instances are accepted new proposal distribution is constructed by multivariate normal kernel with adaptive covariance matrix.
-In the first generation the proposal distribution equals to the prior.
+In the first generation the prior (see below section) defines the proposal distribution.
 The size of $n_t$ is determined by adaptive population strategy~\citep{Klinger2017}.
@@
 % TODO convergence tests, etc. Number of replicates for different models.
+
+\paragraph{Model comparison.} We perform model selection using WAIC (widely applicable information criterion)~\citep{gelman2013bayesian},
+\begin{equation} \label{eq:WAIC}
+\begin{aligned}
+\mathit{WAIC}(\theta, \vec{X}) &= -2\log\mathbb{E}[\mathcal{L}(\theta \mid \vec{X})] + 2\mathbb{V}[\log\mathcal{L}(\theta \mid \vec{X})]
+\end{aligned}
+\end{equation}
+where $\mathbb{E}[\cdot]$ and $\mathbb{V}[\cdot]$ are the expectation and variance operators taken over the posterior distribution $P(\theta \mid \vec{X})$.
+WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy~\citep{Kass1995}.
+% TODO do we need to say "and a difference of 2 is a popular threshold for model comparison."? Meanwile WAICs are so small that there's no such difference
+ % TODO do we use more methods? Describe probably posterior plots comparison method, and Bayes factor if it will used.
 
@@
 We used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$, assuming that fitness values shouldn't differ too much.
-We also tested different priors, but this one (let's call it main prior) was better in the sense of WAIC, prediction plots, estimated parameters.
-%TODO provide formula of WAIC.
-Table?
-To show ppc plots also?.
-\\We tested uniform prior (let's call it uninformative), $\mathit{Uniform}(1,6)$, for all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ and only for $w_{2n+1^*}$, $w_{2n^*}$, leaving $w_{2n+1}$ prior as before.
-In both scenarios the estimates by the model parameters of fitness looks too high: for the first scenario the median estimates were  $w_{2n+1}=2.767$, $w_{2n+1^*}=5.79$, $w_{2n^*}=5.827$ and for the second $w_{2n+1}=1.046$, $w_{2n+1^*}=4.642$, $w_{2n^*}=4.684$.
+We also tested different priors, but this one (denote it \emph{main prior}) was better in the sense of WAIC, prediction plots, estimated parameters.
+%TODO To show ppc plots also?
+Write paragraph about prediction plots in model comparison section.
+\\We tested uniform prior (denote \emph{uninformative}), $\mathit{Uniform}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ and (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, leaving $w_{2n+1}$ prior as in main prior.
+In both scenarios the fitness estimates by the model looks too high: for the first scenario the median estimates were  $w_{2n+1}=2.767$, $w_{2n+1^*}=5.79$, $w_{2n^*}=5.827$ and for the second $w_{2n+1}=1.046$, $w_{2n+1^*}=4.642$, $w_{2n^*}=4.684$.
 We don't expect to have such high fitness values.
@@
 This model didn't work at all: no parameter sets were found with the likelihood greater than zero.
-%TODO it make sense because bla bla?
+%TODO it make sense because bla bla (to explain)?
 
@@
 For the single locus model without aneuploidy, wider uniform prior for the mutation rate was used - $\mathit{Uniform}(1e-10,1e-5)$.
+
 %TODO ILIA please report here the point estimate and 95% CI for the three distributions, I only have for 1-c+b: 0.9863 (95% CI 0.9790 - 0.9939).
-\paragraph{Model comparison.}  % TODO ILIA what did you do for model comparison/selection? WAIC
 
@@
 
+% Almost all done from here already
 %\paragraph{Simple model: Inference.}
@@
 trisomy rate - $7.36\substack{+3.55 \\ -4.82}\times10^{-4}$,
-2n+1 fitness - $1.022\substack{+0.004 \\ -0.004}$,
-2n+1* fitness - $1.025\substack{+0.004 \\ -0.004}$,
-2n* fitness - $1.028\substack{+0.004 \\ -0.004}$.
-This trisomy rate appropriates to the previous estimates; The mutation rate corresponds to the mutations target size of $10^{4}$ if we assume the mutation rate of single base to be about $2*10^{-10}$ as in previous estimates ~\citep{Zhu2014}.
-We can see from the posterior distribution~(\autoref{fig:posterior}) the correlation between fitness of different genotypes (\anwt, \anmt, \eumt)..(TODO say/plot something about the correlation, is it interesting?).
+\anwt\ fitness - $1.022\substack{+0.004 \\ -0.004}$,
+\anmt\ fitness - $1.025\substack{+0.004 \\ -0.004}$,
+\eumt\ fitness - $1.028\substack{+0.004 \\ -0.004}$.
+This trisomy rate appropriates to the previous estimates; the mutation rate corresponds to the mutations target size of $10^{4}$ if we assume the mutation rate of single base to be about $2*10^{-10}$ as in previous estimates ~\citep{Zhu2014}.
+We can see from the posterior distribution~(\autoref{fig:posterior}) the correlation between fitness of different genotypes.. (TODO say/plot something about the correlation, is it interesting?).
 
@@
 %TODO histogram on diagonal
-The posterior density for each parameter is presented on the diagonal plots.
+On the diagonal plots, the posterior density for each parameter is presented .
 Other plots represent joint posterior density of two parameters where blue color has the lowest density and yellow has the highest density.
@@
 trisomy rate - $7.36\times10^{-4}$,
-\anwt fitness - $1.022$,
-\anmt fitness - $1.025$,
-\eumt fitness - $1.028$.
+\anwt\ fitness - $1.022$,
+\anmt\ fitness - $1.025$,
+\eumt\ fitness - $1.028$.
   } 
@@
 
-\paragraph{Single-locus model fits the data well, but only if contains aneuploidy.} Single-locus model provide good fit to the data~(\autoref{fig:likelihood}A).
-On \autoref{fig:ppc-plot} the dynamics of frequency change of each genotype in time is shown. We can see that \anmt never reach substantial frequency in the population.  
-Sensitivity analysis~(\autoref{fig:sensitivity}) shows that if we change the parameter values the fit degrades.
-Furthermore, we measured fixation time of \anwt in population, and found it appropriate to the unpublished data of the experiment ~\citep{Yona2012} - about 300 generations in average.
+\paragraph{Single-locus model fits the data well, but only if contains aneuploidy.} Single-locus model fits well to the data~(\autoref{fig:likelihood}A).
+On \autoref{fig:ppc-plot} the dynamics of frequency change of each genotype in time is shown. We can see that \anmt\ never reach substantial frequency in the population.  
+Sensitivity analysis~(\autoref{fig:sensitivity}) shows that changing the parameter values cause the model to fit less.
+Furthermore, we measured fixation time of \anwt\ in population, and found it appropriate to the unpublished data of the experiment ~\citep{Yona2012} - about 300 generations in average, that supports the correctness of the inference.
 On the other hand, if we don't take into consideration aneuploidy in the model, i.e.
 take aneuploidy rate be equal to zero, then the model can't explain the data: the fit is not so good~(\autoref{fig:likelihood}A; \autoref{fig:2n*-fixation}).
-The good fit of the model without aneuploidy to the data, requires much lower mutation rate (TODO provide numbers for the median fit?).
-If mutation rate is not low, the fixation of \eumt happens much quicker (see yellow and blue bars at  \autoref{fig:2n*-fixation}).
+The inference parameter values of the model without aneuploidy are:
+mutation rate - $1.251\substack{+0.146 \\ -0.177}\times10^{-8}$,
+2n+1 fitness - $1.01\substack{+0.001 \\ -0.002}$,
+2n+1* fitness - $1.011\substack{+0.001 \\ -0.001}$,
+2n* fitness - $1.011\substack{+0.001 \\ -0.001}$. 
+We can see that mutation rate for this model is much lower than for the model with aneuploidy.
+Higher mutation rate would cause quicker fixation of \eumt\ than expected by the experiment ( \autoref{fig:2n*-fixation} blue bar).
+Thereby, adding aneuploidy to the model, helps to attain good fit with higher mutation rate.
 
@@
   \caption{
-  \textbf{Time of \eumt fixation cannot be explained without aneuploidy.}
-Each color represents distribution to fixation of \eumt of $10,000$ simulations for different modifications of single-locus model.
-\emph{Blue} represents the model with aneuploidy with parameters that taken from posterior median when we fit the model to the experimental data.
-\emph{Orange} represents the same model that have no aneuploidy, i.e aneuploidy rate is zero.
-\emph{Green} represents the model without aneuploidy with parameters that taken from posterior median when we fit the model without aneuploidy to the experimental data.
-The last bin contains all the simulations with time greater than 3000.
-Dashed lines show experimental data: generations 1700 and 2350.
-The fit to the data is better when the distribution is such that the difference in number of simulations in ranges 0-1700 and 1700-2350 is small and there's no many simulations after 2350 generations.
-The fit of \emph{blue} is better than of \emph{green}.
-And there's no fit of \emph{red} to the data at all.
+  \textbf{Time of \eumt\ fixation cannot be explained without aneuploidy.}
+Each color represents distribution to fixation of \eumt\ of $10,000$ simulations for different modifications of single-locus model.
+\emph{Blue} bars correspond to the model with aneuploidy, where parameters are taken from the median of the posterior of the inference.
+\emph{Orange} bars correspond to the same model with same parameters, except aneuploidy that is zero.
+\emph{Green} bars correspond to the model without aneuploidy, i.e.
+with aneuploidy rate zero, where parameters are taken from the median of the posterior of the inference for this model.
+The last bin contains all the simulations with time equal or greater than 3000.
+Dashed lines represent experimental data: generations 1700 and 2350.
+When the difference in number of simulations between ranges 0-1700 and 1700-2350 is small, and there are no much simulations after generation 2350 - the fit is good.
+We can see that the fit of the model with aneuploidy (\emph{blue}) is the best.
+There's no fit at all for the model that is represented by \emph{yellow} bars.
   }
@@
 
-%same for complex model
+%TODO same for complex model
 %model comparison between models - model probabilities

### 2021-12-13 ms/ms.tex
added ess and seeds figures, added At,T to the likelihood

@@
 Aneuploidy fixed in all four population in the first 450 generations (hereafter, fixation or elimination of a genotype by generation $t$ means that more than 95\% or less than 5\% of the population carry the genotype at generation $t$, and possibly earlier).
+By unpublished observations the aneuploidy was not fixed before at least 200 generations.
 The experiment continued with two populations, in which aneuploidy was eliminated by generation 1700 and 2350.
@@
 \paragraph{Likelihood function.}
-Denote by $A_{t}$ the fixation of aneuploidy at generation $t$, by $L_{t}$ the loss of aneuploidy at generation $t$, and by $L^*_{t,T}$ the loss of aneuploidy at generation $t$ conditioned on no loss by generation $T$.
+Denote by $A_{t,T}$ the fixation of aneuploidy at generation $t$, conditioned on no fixation at generation $T$; by $L_{t}$ the loss of aneuploidy at generation $t$; and by $L^*_{t,T}$ the loss of aneuploidy at generation $t$ conditioned on no loss by generation $T$.
 The model likelihood function for parameter vector $\theta$ for the heat-stress experiment is
 \begin{multline} \label{eq:heatstress-likelihood}
-\likelihood(\theta | A_{450}, L_{1700}, L^*_{2350, 1700} ) = \\
-P^4(A_{450}) \cdot \Big(1 - P^4(\lnot{L_{1700}} | A_{450} ) - P^4(\lnot{L^*_{2350, 1700}}| A_{450}) + P^4(\lnot{L_{1700}} \land \lnot{L^*_{2350, 1700}}| A_{450})\Big)	\;,	
+\likelihood(\theta | A_{450,200}, L_{1700}, L^*_{2350, 1700} ) = \\
+P^4(A_{450,200}) \cdot \Big(1 - P^4(\lnot{L_{1700}} | A_{450,200} ) - P^4(\lnot{L^*_{2350, 1700}}| A_{450,200}) + P^4(\lnot{L_{1700}} \land \lnot{L^*_{2350, 1700}}| A_{450,200})\Big)	\;,	
 \end{multline}
@@
 The likelihood function of the model with parameter set $\theta$ for the pH-stress experiment is
-\begin{equation} \label{eq:phstress-likelihood} % TODO: add ph-Stress model to results or remove it from here
+\begin{equation} \label{eq:phstress-likelihood} % TODO: add ph-Stress model to results or remove it from here (if remained add A_t to notations, it is At,T now only)
 \likelihood(\theta | A_{150}, L_{750} ) = \\
- P^4(L_{450}) \cdot 6 \cdot	P^4(L_{750} | A_{150}) \cdot P^4(\lnot{L_{750}} | A_{150}) \;.  
+ P^4(L_{150}) \cdot 6 \cdot	P^2(L_{600} | A_{150}) \cdot P^2(\lnot{L_{600}} | A_{150}) \;.  
  \end{equation}
-%TODO i think it is wrong because fig 5C shows that only 2 replicas are fixated after 650 generations (see Supplementary also) Anyway, I don't think that we use this likelihood, so probably should be removed at all.
 
@@
 Therefore, given specific values of the model parameters $\theta$, we simulate the model in many replicates to approximate the value of the likelihood function $\likelihood(\theta)$.
+For the single locus model 1000 replicates were used.
+%TODO if we have more models, write how much replicates were used.
 
@@
 We stop the iterations, when no more sufficient change in acceptance threshold occur, and no change in posterior.
+We stop single-locus model after four iterations with epsilon 0.13, when 1074 accepted instances with effective sample size of 700 (Figure~\ref{fig:ess-single}).
+We ran the algorithm with different initialization seeds and different number of replicates for simulations and got similar posterior (Figure~\ref{fig:seeds}).
+%TODO if we have more models, write about them here also.
+
 After receiving weighted particles of the posterior distribution using ABC-SMC scheme, we calculate Kernel Density Estimate using Gaussian kernels.
 $50,000$ sampled points from KDE represent the posterior distribution over the model parameters, and we use it for plots, median calculation, etc.
-% TODO convergence tests, etc. Number of replicates for different models.
 
@@
 Moreover the WAIC for the models with these priors are 0.57 and 1.9 respectively, while WAIC for the main prior model is 0.27.
-Also for the model with uninformative prior for all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ the fixation time of 2n+1 is about 16 in average, which contradicts unpublished observations that were about 250-300.
+Also for the model with uninformative prior for all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ the fixation time of 2n+1 is about 16 in average, which contradicts unpublished observations that were at least 250-300.
+%TODO remove about unpublished observations...we already added it into likelihood function of the model.
 \\Also we tried to use additional growth curves.
@@
 \eumt\ fitness - $1.028\substack{+0.004 \\ -0.004}$.
+
+and new:
+mutation rate - $3.606\substack{+6.842 \\ -5.258}\times10^{-6}$,
+trisomy rate - $5.586\substack{+5.513 \\ -6.112}\times10^{-3}$,
+2n+1 fitness - $1.022\substack{+0.004 \\ -0.004}$,
+2n+1* fitness - $1.025\substack{+0.004 \\ -0.004}$,
+2n* fitness - $1.028\substack{+0.005 \\ -0.004}$,
+
 This trisomy rate appropriates to the previous estimates; the mutation rate corresponds to the mutations target size of $10^{4}$ if we assume the mutation rate of single base to be about $2*10^{-10}$ as in previous estimates ~\citep{Zhu2014}.
@@
 
+\begin{figure}[h]
+    \centering
+	\includegraphics[width=1\textwidth]{../figures/ess.pdf}
+    \caption{
+    \textbf{Single-locus model convergence.} Pyabc run over four iterations.
+The model is converged when acceptance threshold (epsilon) is 0.13, the number of particles (accepted instances) is 1074 giving the effective sample size (ESS) 700.
+    } 
+    \label{fig:ess-single}
+\end{figure}
+
+\begin{figure}[h]
+    \centering
+	\includegraphics[width=0.6\textwidth]{../figures/seeds.pdf}
+    \caption{
+    \textbf{Posterior for different runs} %TODO make more compact; one legend is enough.
+    For each parameter of the single-locus model, the posterior is represented for runs with different initialization seeds and different number of replicas in simulations run. 
+    } 
+    \label{fig:seeds}
+\end{figure}
+
 \end{document}

### 2021-12-15 ms/ms.tex
updated figures and texts related to an0

@@
 
-After receiving weighted particles of the posterior distribution using ABC-SMC scheme, we calculate Kernel Density Estimate using Gaussian kernels.
-$50,000$ sampled points from KDE represent the posterior distribution over the model parameters, and we use it for plots, median calculation, etc.
+After getting weighted particles of the posterior distribution by ABC-SMC scheme, we calculate Kernel Density Estimate using Gaussian kernels and $50,000$ points are sampled from KDE.
+We use the KDE and the points to calculate mode, highest density interval (HDI), for drawing distribution plots, etc.
 
@@
 
-For the mutation rate $\mu$ and aneuploidy rate $\delta$ we used uniform priors, $\mathit{Uniform}(1e-8,1e-5)$ and $\mathit{Uniform}(1e-6,1e-3)$ respectively.
-We tried also uniform priors with lower than the $1e-8$ bound for $\mu$, but anyway the posterior ended with higher values.
+For the mutation rate $\mu$ and aneuploidy rate $\delta$ we used uniform priors, $\mathit{Uniform}(1e-9,1e-5)$ and $\mathit{Uniform}(1e-6,1e-2)$ respectively.
 For the single locus model without aneuploidy, wider uniform prior for the mutation rate was used - $\mathit{Uniform}(1e-10,1e-5)$.
@@
 \paragraph{Single-locus model parameters inference.} 
-First we inference the parameters of the single-locus model. The mean and credible interval (2.5th percentile and the 97.5th percentile) of the parameters are:
-mutation rate - $1.02\substack{+2.40 \\ -1.34}\times10^{-6}$,
-trisomy rate - $7.36\substack{+3.55 \\ -4.82}\times10^{-4}$,
-\anwt\ fitness - $1.022\substack{+0.004 \\ -0.004}$,
-\anmt\ fitness - $1.025\substack{+0.004 \\ -0.004}$,
-\eumt\ fitness - $1.028\substack{+0.004 \\ -0.004}$.
+First we inference the parameters of the single-locus model. The mode and 50\% hdi of the parameters posterior are:
+mutation rate - $3.06\substack{+1.229 \\ -2.84}\times10^{-6}$,
+trisomy rate - $1.61\substack{+1.122 \\ -0.197}\times10^{-3}$,
+2n+1 fitness - $1.022\substack{+0.001 \\ -0.002}$,
+2n+1* fitness - $1.025\substack{+0.001 \\ -0.001}$,
+2n* fitness - $1.028\substack{+0.001 \\ -0.002}$,
 
-and new:
-mutation rate - $3.606\substack{+6.842 \\ -5.258}\times10^{-6}$,
-trisomy rate - $5.586\substack{+5.513 \\ -6.112}\times10^{-3}$,
-2n+1 fitness - $1.022\substack{+0.004 \\ -0.004}$,
-2n+1* fitness - $1.025\substack{+0.004 \\ -0.004}$,
-2n* fitness - $1.028\substack{+0.005 \\ -0.004}$,
-
-This trisomy rate appropriates to the previous estimates; the mutation rate corresponds to the mutations target size of $10^{4}$ if we assume the mutation rate of single base to be about $2*10^{-10}$ as in previous estimates ~\citep{Zhu2014}.
+This trisomy rate appropriates to the previous estimates; the mutation rate corresponds to the mutations target size of $10^{4}$ if we assume the mutation rate of single base to be about $2*10^{-10}$ as estimated by ~\cite{Zhu2014}.
 We can see from the posterior distribution~(\autoref{fig:posterior}) the correlation between fitness of different genotypes.. (TODO say/plot something about the correlation, is it interesting?).
@@
 Other plots represent joint posterior density of two parameters where blue color has the lowest density and yellow has the highest density.
-Red point and the dashed lines represent the median (TODO check that it is median and not mean) values of the densities:
-mutation rate - $1.02\times10^{-6}$,
-trisomy rate - $7.36\times10^{-4}$,
+Red point and the dashed lines represent the maximum a posteriori probability estimate:
+mutation rate - $3.06\times10^{-6}$,
+trisomy rate - $1.61\times10^{-4}$,
 \anwt\ fitness - $1.022$,
@@
   \textbf{Evolutionary dynamics with fitted single-locus model.} 
-  Frequency of the four genotypes averaged over 10,000 simulations of the model. Each line represent one of the 10 parameter sets drawed from the inference posterior.
+  Frequency of the four genotypes averaged over 10,000 simulations of the model. Each line represent one of the 10 parameter sets drawed randomly from the inference posterior.
   }
@@
     \textbf{Single-locus model fit.} Cumulative distribution of 10000 simulations are shown.
-    \textbf{(A)} Simulations parameters are from the median of the posterior of the single-locus model fit. The likelihood for the heat-stress experiment (\autoref{eq:heatstress-likelihood}) is 0.86.
-    \textbf{(B)} Simulations parameters are from the median of the posterior of the single-locus model \emph{without} aneuploidy fit.
+    \textbf{(A)} Simulations parameters are from the maximum a posterior estimation of the single-locus model fit.
+The likelihood for the heat-stress experiment (\autoref{eq:heatstress-likelihood}) is 0.86.
+    \textbf{(B)} Simulations parameters are from the maximum a posterior estimation of the single-locus model \emph{without} aneuploidy fit. 
+    %TODO get it from MAP and not mean
     The likelihood for the heat-stress experiment (\autoref{eq:heatstress-noaneuploidy-likelihood}) is 0.75.
@@
   \caption{
-    \textbf{Sensitivity analysis.} We take parameters that are median of the posterior.
+    \textbf{Sensitivity analysis.} We take parameters that are maximum a posterior (MAP) for single locus model.
 Then we change only one parameter at time while measuring the model likelihood of the parameters.
 x-axis label represent the parameter that we change.
-Yellow point indicates the median value.
+Yellow point indicates the MAP value.
   }

### 2021-12-20 ms/ms.tex
revised abstract, intro, models and methods

@@
 However, evidence from yeast and fungi, as well as human tumour cells, suggests that aneuploidy can be beneficial under stressful conditions and lead to elevated growth rates and adaptation.
-Because aneuploidy differs from mutation in rate, expected effect, and reversibility, it is crucial to develop a quantitative theory for the role of aneuploidy in adaptive evolution.
+Importantly, aneuploidy differs from point mutations in rate, fitness effect, and reversibility.
 Here, we develop evolutionary models for adaptive evolution with both mutation and aneuploidy. 
-These models are used within an approximate Bayesian computation framework to estimate the formation rate and fitness effect of aneuploidy and mutation from empirical results of experiments in which \yeast adapted to heat stress.
-The experimental population first acquired chromosome duplications, only to later revert back to a euploid state.
+These models are used within an approximate Bayesian computation framework to estimate the formation rate and fitness effect of aneuploidy and mutation from  results of evolutionary experiments in which \yeast adapted to heat stress: the experimental populations first acquired chromosome duplications, only to later revert back to a euploid state.
 We also analyze our models to estimate the effect of the aneuploidy and mutation rates on the expected adaptation time and the probability for adaptation via aneuploidy.
-Our results suggest that aneuploidy is a transient adaptive solution, which can decelerate adaptation in a non-intuitive manner.
+Our results suggest that aneuploidy can be a transient adaptive solution, which can decelerate adaptation in a non-intuitive manner.
 By creating an evolutionary conflict between the individual and the population, aneuploidy further complicates  the process of adaptation in cell populations.
@@
 These models follow a population of cells characterised by both their ploidy and their genotype.
-We fit these models to the experimental results of \citet{Yona2012} using an \emph{approximate Bayesian computation} framework~\citep{Sisson2009, Klinger2018} to infer model parameters, including selection coefficients and rates of aneuploidy and mutation, and to perform model selection between different models, thereby testing different hypotheses about the evolutionary process.
+We fit these models to the experimental results of \citet{Yona2012} using an \emph{approximate Bayesian computation} framework~\citep{Sisson2009} to infer model parameters, including selection coefficients and rates of aneuploidy and mutation, and to perform model selection between different models, thereby testing hypotheses about the evolutionary process.
 %TODO do model selection by comparing Bayes factor or ignore this part.
-Furthermore, we analyze these models to estimate the effects of parameters on the adaptation time and the probability for adaptation via aneuploidy.
-We find that % TODO
+We analyze these evolutionary-genetic models to estimate the effects of parameters on the adaptation time and the probability for adaptation via aneuploidy.
+We find that \textbf{TODO} % TODO
 
@@
 
+%%%%%%%% Fig illustation
 \begin{figure}[b!]
-  \centering
-  \begin{subfigure}[t]{0.5\textwidth}
+  %\centering
+  \begin{subfigure}[t]{0.4\textwidth}
       \caption{
@@
       \label{fig:model1}
-  \end{subfigure}%
-  \\
-  \begin{subfigure}[t]{0.5\textwidth}
+  \end{subfigure}
+  %\\
+  \begin{subfigure}[t]{0.4\textwidth}
   	  \caption{
@@
     \textbf{(B)} In the multi-locus model, each genotype is characterized by its karyotype, \euwt\ or \anwt, and the number of accumulated beneficial mutations, denoted by stars. 
-    In both panels arrows denote transitions between genotypes, with transitions rates: $\mu$, mutation rate; $\delta$, aneuploidy gain rate; $\delta_L$, aneuploidy loss rate.
+    In both panels arrows denote transitions between genotypes, with transitions rates: $\mu$ and $\tau \mu$, beneficial mutation rates in euploid and aneuploid cells; $\delta$, aneuploidy gain rate; $\delta_L$, aneuploidy loss rate.
+    % TODO IK: change k mu to tau mu in figure; also k mu is too close to the arrow right now; change alpha to delta in the figure 
   }
@@
 \paragraph*{Evolutionary Models.}
-We developed two models: a single-locus model and a multi-locus model. 
+We model the evolution of a population of cells using two models: a single-locus model and a multi-locus model.
 Both models are based on the Wright-Fisher model~\citep{Otto2007}, assuming non-overlapping generations and including the effects of natural selection, genetic drift, aneuploidy, and mutation. 
 We focus on beneficial mutations, neglecting the effects of deleterious and neutral mutations. Both models allow for a single aneuploid karyotype (e.g., chromosome III duplication).
-While the single-locus model allows for only a single mutation to occur, the
-multi-locus model allows for multiple mutations to accumulate in the genome~(\autoref{fig:models}).
+While the single-locus model allows for only a single mutation to accumulate in the genotype, the multi-locus model allows for multiple mutations to accumulate~(\autoref{fig:models}), as well as for a fluctuating population size.
 
 \paragraph*{Single-locus model.}
-This model assumes a constant population size $N$ and follows four genotypes~(\autoref{fig:models}A): euploid wild-type, \euwt, the initial genotype; 
+This model assumes a constant effective population size $N$ and follows four genotypes~(\autoref{fig:models}A): euploid wild-type, \euwt, the initial genotype; 
 euploid mutant, \eumt, with the standard karyotype and a single beneficial mutation; 
-aneuploid wild-type, \anwt, with an extra chromosome, e.g. following chromosome duplication; and
+aneuploid wild-type, \anwt, with an extra chromosome, i.e., following chromosome duplication; and
 aneuploid mutant, \anmt, with and extra chromosome and a beneficial mutation.
@@
 Transitions between the genotypes occur as follows~(\autoref{fig:models}A): Beneficial mutations from \euwt\; to \eumt\; occur with probability $\mu$, the mutation rate, and from \anwt\; to \anmt\; with probability $\tau \mu$, where $\tau$ is the fold-change in the rate of beneficial mutations in aneuploid cells.
-By default, we assume $\tau=33/32$, as the wild-type \yeast strains in the experiments by \citep{Yona2012} are diploid, with 32 chromosomes, and the aneuploid strains are trisomic, with 33 chromosomes.
+By default, we assume $\tau=33/32$, as the wild-type \yeast strains in the experiments by \citep{Yona2012} are diploid with 32 chromosomes, and the aneuploid strains are trisomic, with 33 chromosomes.
 Aneuploidy is formed by chromosome mis-segregation, so that cells transition from \euwt\; to \anwt\ with probability $\delta$, the aneuploidy gain rate.
@@
 and $0 \le c \le 1$ is the fitness cost of aneuploidy.
+% TODO IK do we assume b>c?
 }
@@
 
-The first generation is initialized with $N$ cells with genotype \euwt.
+The initial population has $N$ cells with genotype \euwt.
 The effect of natural selection on the frequency $f_i$ of genotype $i = 2n, 2n+1, 2n+1^*, \text{or } 2n^*$ is given by
@@
 The effect of mutation and aneuploidy on genotype frequencies is given by
+% TODO IK we are missing 2n+1->2n, 2n*->2n+1*, why? to neglect them we need to assume w(2n+1)>w(2n) and w(2n*)>w(2n+1*), but I thought we don't assume that but rather we infer that. 
     \begin{equation} \label{eq:mutation-aneuploidy-single}
@@
     \begin{equation} \label{eq:drift-single}
-      \vec{f'} \sim \frac{1}{N} Mult(N, \vec{f^m}) \;,
+      \vec{f'} \sim \frac{1}{N} \cdot \mathit{Mult}(N,\ \vec{f^m}) \;,
     \end{equation}
-where $\vec{f^m}=(f^m_{2n},f^m_{2n+1},f^m_{2n+1^*},f^m_{2n^*})$ are the frequencies of the genotypes after mutation and aneuploidy, $\vec{f'}$ are the genotype frequencies in the next generation, and $Mult(N, \vec{f})$ is a multinomial distribution parameterized by the population size $N$ and the genotype frequencies $\vec{f}$.
+where $\vec{f^m}=(f^m_{2n}\ ,\ f^m_{2n+1}\ ,\ f^m_{2n+1^*}\ ,\ f^m_{2n^*})$ are the frequencies of the genotypes after mutation and aneuploidy, $\vec{f'}$ are the genotype frequencies in the next generation, and $Mult(N,\ \vec{f})$ is a multinomial distribution parameterized by the population size $N$ and the genotype frequencies $\vec{f}$.
 Overall, the change in genotype frequencies from one generation to the next is given by the transformation $f_i \to f'_i$.
@@
 \paragraph*{Multi-locus model.}
-This model expands the single-locus model by allowing for (i) the accumulation of beneficial mutations in the genome, and (ii) a fluctuating population size.
+This model expands the single-locus model by allowing for (i) the accumulation of beneficial mutations, and (ii) a fluctuating population size.
 
 A genotype is characterized by its karyotype, \euwt\; or \anwt, and the number of accumulated beneficial mutations, which can be zero or more.
-The selection coefficient of the $i$-th accumulated mutation in each individual, $s_i$, is drawn from an exponential distribution with expected value $s$, $s_i \sim Exp(s)$.
+The selection coefficient of the $i$-th accumulated mutation in each individual, $s_i$, is drawn from an exponential distribution with expected value $s$: $s_i \sim \mathit{Exp{(s)}}$.
 The rest of the parameters ($N$, $\mu$, $\tau$, $\delta$, $\delta_L$, $b$, $c$) are the same as in the single-locus model.
@@
 So, we don’t expect to receive same values for mutation rate for the models.
+% YOAV: let's talk about it, I don't necessarily agree
 % ILIA: I don’t have in the code 𝜏 parameter for multi-locus model (it’s like 1 there). Do you think it worth adding it for consistency?
+% YOAV: if our inference tells us that tau=1 is as good as tau>1, then maybe we can remove this parameter from the model presentation altogether, and simply introduce it when we discuss this result as part of the sensitivity analysis section
 The fitness of the different genotypes is the same as in the single-locus model (\autoref{table:single-locus}), except that the fitness contribution of $k$ beneficial mutations is the product of their independent effects, $\prod_{i=1}^k(1+s_i)$, instead of the contribution of the single mutation allowed in the single-locus model, $(1+s)$, see \autoref{table:multi-locus}.
 Therefore, aneuploidy loss would be favored by selection only if there are enough beneficial mutations and/or the selection coefficients $s_i$ are large enough.
-The intuition is that when the benefit of the accumulated beneficial mutations is small, the benefit of aneuploidy has a large effect; when the benefit of the accumulated beneficial mutations benefit is large, then aneuploidy is no longer advantageous because of its significant cost.
+The intuition is that when the benefit of the accumulated beneficial mutations is small, ten the benefit of aneuploidy has a large effect; when the benefit of the accumulated beneficial mutations is large, then aneuploidy is no longer beneficial due of its fitness cost.
 
-In contrast to the single-locus model, in the multi-locus model the population size changes in order to model serial-transfer experiment protocol~\citep{Yona2012}: the population is serially diluted by transferring a fraction of the population (1/120) to a fresh medium, starting a new growth cycle.
-In this model, the population initial size is $N_0 = N$, and the population size is doubled every generation, $N_1=2N, N_2=4N, \ldots$, and diluted back to $N$ every eight generations, $N_8=N$.
+In contrast to the single-locus model, in the multi-locus model the population size fluctuates to model serial-transfer experimental protocol~\citep{Yona2012}: the population is serially diluted by transferring a fraction of the population (1/120) to a fresh medium approximately every seven generations.
+Thus, the population initial size is $N_0 = N$, and the population size is doubled every generation, $N_1=2N, N_2=4N, \ldots$, and diluted back to $N$ after seven generations such that $N_8=N$.
 
 The change in frequencies due to selection is exactly the same as in the single-locus model (\autoref{eq:selection-single}), only applied using the fitness values in \autoref{table:multi-locus}.
-The change due to random genetic drift is also the same as in \autoref{eq:drift-single}, except that the frequencies vector is $\vec{f}=(f_{2n}, f_{2n+1}, f_{2n^*}, f_{2n+1^*}, f_{2n^{**}}, f_{2n+1^{**}}, \ldots)$ and that the population size changes between generations, as described above.
+The change due to random genetic drift is also the same as in \autoref{eq:drift-single}, except that the frequencies vector is $\vec{f}=(f_{2n}\ ,\ f_{2n+1}\ ,\ f_{2n^*}\ ,\ f_{2n+1^*}\ ,\ f_{2n^{**}}\ ,\ f_{2n+1^{**}}\ ,\ \ldots)$ and that the population size changes between generations, as described above.
 
 The effects of mutation and aneuploidy on genotype frequencies is more elaborate than in the single-locus model.
-Genotype $i$ is classified according to their karyotype (\emph{2n} or \emph{2n+1}), the number of accumulated beneficial mutations ($k \ge 0$), and their fitness ($w_i$).
+Genotype $i$ is classified according to its karyotype (\euwt\ or \anwt), the number of accumulated beneficial mutations ($k \ge 0$), and their fitness ($w_i$).
 Each offspring cell inherits these properties from its mother cell.
-Then, with probability $\mu$ or $\tau \mu$ for euploid and aneuploid cells, respectively, a new beneficial mutation is accumulated, such that the number of mutations is $k+1$, and its effect $s_{k+1}$ is drawn from an exponential distribution with expected value $s$, such that the contribution of the mutations to the fitness is $\prod_{j=0}^{k+1}{(1+s_j)}$.
+Then, with probability $\mu$ or $\tau \mu$ for euploid and aneuploid cells, % TODO no tau in multi-locus model?
+respectively, a new beneficial mutation is accumulated, such that the number of mutations is $k+1$, and its effect $s_{k+1}$ is drawn from an exponential distribution with expected value $s$, such that the contribution of the mutations to the fitness is $\prod_{j=0}^{k+1}{(1+s_j)}$.
 Next, euploid offspring become aneuploid with probability $\delta$, and aneuploid offspring become euploid with probability $\delta_L$.
@@
 \caption*{
-$k$ is the number of accumulated beneficial mutations in the genome;
+$k$ is the number of accumulated beneficial mutations;
 $s \ge 0$ is the selection coefficient of a beneficial mutation;
@@
 and $0 \le c \le 1$ is the fitness cost of aneuploidy.
+%TODO do we assume b>c ?
 }
@@
 \paragraph{Empirical evidence.}
-Our inference procedure uses empirical data from evolutionary experiments performed by \citet{Yona2012}.
-In the heat-stress experiment, four populations of \yeast evolved under \SI{39}{\celsius}.
-Aneuploidy fixed in all four population in the first 450 generations (hereafter, fixation or elimination of a genotype by generation $t$ means that more than 95\% or less than 5\% of the population carry the genotype at generation $t$, and possibly earlier).
-By unpublished observations the aneuploidy was not fixed before at least 200 generations.
+We use empirical results of evolutionary experiments reported by \citet{Yona2012}.
+In their heat-stress experiment, four populations of \yeast evolved under \SI{39}{\celsius}.
+Aneuploidy fixed in all four population in the first 450 generations (hereafter, fixation or elimination of a genotype \emph{by generation $t$} means that more than 95\% or less than 5\% of the population carry the genotype at generation $t$, and possibly earlier).
+From unpublished results, aneuploidy did not fix before at least 200 generations elapsed.
 The experiment continued with two populations, in which aneuploidy was eliminated by generation 1700 and 2350.
-In the pH-stress experiment, four populations of \yeast evolved under high-ph stress (8.6).
+%In the pH-stress experiment, four populations of \yeast evolved under high-ph stress (8.6).
 Aneuploidy fixed during the first 150 generations.
 It was fully eliminated in two populations and partly eliminated in two populations by generation 750.
-These empirical results were published by \citet{Yona2012}.
+%TODO do we use pH?
 
 \paragraph{Likelihood function.}
-Denote by $A_{t,T}$ the fixation of aneuploidy at generation $t$, conditioned on no fixation at generation $T$; by $L_{t}$ the loss of aneuploidy at generation $t$; and by $L^*_{t,T}$ the loss of aneuploidy at generation $t$ conditioned on no loss by generation $T$.
+Denote by $A_{t,T}$ the fixation of aneuploidy by generation $t$ conditioned on no fixation up to generation $T$; by $B_{t}$ the extinction of aneuploidy by generation $t$; and by $C_{t,T}$ the extinction of aneuploidy by generation $t$ conditioned on no extinction by generation $T$.
 The model likelihood function for parameter vector $\theta$ for the heat-stress experiment is
 \begin{multline} \label{eq:heatstress-likelihood}
-\likelihood(\theta | A_{450,200}, L_{1700}, L^*_{2350, 1700} ) = \\
-P^4(A_{450,200}) \cdot \Big(1 - P^4(\lnot{L_{1700}} | A_{450,200} ) - P^4(\lnot{L^*_{2350, 1700}}| A_{450,200}) + P^4(\lnot{L_{1700}} \land \lnot{L^*_{2350, 1700}}| A_{450,200})\Big)	\;,	
+\likelihood(\theta\ |\ A_{450,200}\ ,\ B_{1700}\ ,\ C_{2350, 1700} ) = \\
+P^4(A_{450,200}) \cdot \Big(1 - P^4(\lnot{B_{1700}} | A_{450,200} ) - P^4(\lnot{C_{2350, 1700}}| A_{450,200}) + P^4(\lnot{B_{1700}} \land \lnot{L^*_{2350, 1700}}| A_{450,200})\Big)	\;,	
 \end{multline}
@@
 
-The likelihood function of the model with parameter set $\theta$ for the pH-stress experiment is
-\begin{equation} \label{eq:phstress-likelihood} % TODO: add ph-Stress model to results or remove it from here (if remained add A_t to notations, it is At,T now only)
-\likelihood(\theta | A_{150}, L_{750} ) = \\
- P^4(L_{150}) \cdot 6 \cdot	P^2(L_{600} | A_{150}) \cdot P^2(\lnot{L_{600}} | A_{150}) \;.  
- \end{equation}
+%The likelihood function of the model with parameter set $\theta$ for the pH-stress experiment is
+%\begin{equation} \label{eq:phstress-likelihood} % TODO: add ph-Stress model to results or remove it from here (if remained add A_t to notations, it is At,T now only)
+%\likelihood(\theta | A_{150}, L_{750} ) = \\
+% P^4(B_{150}) \cdot 6 \cdot	P^2(B_{600} | A_{150}) \cdot P^2(\lnot{B_{600}} | A_{150}) \;.  
+% \end{equation}
 
-For the model without aneuploidy, let's denote by $M_{t}$, the fixation of \eumt\ at generation $t$, and by $M^*_{t,T}$ the fixation of \eumt\ at generation $t$ conditioned on no fixation at generation $T$.
-The likelihood of the model without aneuploidy with parameter vector $\theta$ for the heat-stress experiment is
+For a model without aneuploidy (that is, when the aneuploidy rate is fixed at zero, $\delta=0$), denote by $M_{t}$ the fixation of \eumt\ by generation $t$, and by $Q_{t,T}$ the fixation of \eumt\ by generation $t$ conditioned on no fixation at generation $T$.
+The likelihood of a model without aneuploidy with parameter vector $\theta$ for the heat-stress experiment is
 \begin{multline} \label{eq:heatstress-noaneuploidy-likelihood}
-\likelihood(\theta |  M_{1700}, M^*_{2350, 1700} ) = \\
-\Big(1 - P^4(\lnot{M_{1700}}) - P^4(\lnot{M^*_{2350, 1700}}) + P^4(\lnot{M_{1700}} \land \lnot{M^*_{2350, 1700}})\Big)	\;,	
+\likelihood(\theta |  M_{1700}, M^*_{2350, 1700} ) = 
+\Big(1 - P^4(\lnot{M_{1700}}) - P^4(\lnot{Q_{2350, 1700}}) + P^4(\lnot{M_{1700}} \land \lnot{Q_{2350, 1700}})\Big)	\;,	
 \end{multline}
 
-Because of the complexity of the model, these likelihood functions are intractable.
-Therefore, given specific values of the model parameters $\theta$, we simulate the model in many replicates to approximate the value of the likelihood function $\likelihood(\theta)$.
-For the single locus model 1000 replicates were used.
+Because our model, just like the Wright-Fisher model, is non-linear and stochastic, these likelihood functions are intractable.
+We overcome this problem by approximating the likelihood using multiple simulations (1,000 simulations per parameter vector for the single-locus model).
 %TODO if we have more models, write how much replicates were used.
+% TODO ref to place where we show 1,000 is enough
 
-\paragraph{Parameter inference.} To infer model parameters, we use approximate Bayesian computation~\citep{Sunnaker2013} with a sequential Monte-Carlo scheme, or ABC-SMC~\citep{Sisson2009}, implemented in the \texttt{pyABC}\footnote{https://pyabc.readthedocs.io} Python
-package~\citep{Klinger2018}.
-Briefly, this approach uses numerical stochastic simulations of the model to infer a posterior distribution over the model parameters.
-It is a method of likelihood-free, simulation-based inference~\citep{Cranmer2020}, that is, for estimating a posterior distribution when a likelihood function cannot be computed.
+\paragraph{Parameter inference.} To infer model parameters, we use approximate Bayesian computation~\citep{Sunnaker2013} with a sequential Monte-Carlo scheme, or ABC-SMC~\citep{Sisson2009}, implemented in the \texttt{pyABC} Python
+package~\citep[\href{https://pyabc.readthedocs.io}{pyabc.readthedocs.io}]{Klinger2018}.
+This approach uses numerical stochastic simulations of the model to infer a posterior distribution over the model parameters.
+It is a method of likelihood-free, simulation-based inference~\citep{Cranmer2020}, that is, for estimating a posterior distribution when a likelihood function cannot be directly computed.
 It is therefore suitable to our case, in which the likelihood function can only be approximated from simulations, and cannot be directly computed.
-%a bit adapted/expanded description from Syga et al. 2021:
-In short, the algorithm employs sequential importance sampling over generations. 
-In generation $t$, the algorithm draws sets of parameters $\theta_i$ from a given proposal distribution and consequently simulates data $C_i$ from the model until $n_t$ instances were accepted based on the comparison to observed data $C_{obs}$ via a distance function $D(C_i, C_{obs})$ and acceptance threshold $\epsilon_t$, such that $D(C_i, C_{obs})\leq{\epsilon_t}$.
-As the distance function we choose $1-\likelihood$.
-For the acceptance threshold the median of the distances of the accepted instances of the previous generation are applied (for the first generation we define the acceptance threshold be $0.01$).
-When $n_t$ instances are accepted new proposal distribution is constructed by multivariate normal kernel with adaptive covariance matrix.
-In the first generation the prior (see below section) defines the proposal distribution.
-The size of $n_t$ is determined by adaptive population strategy~\citep{Klinger2017}.
-We stop the iterations, when no more sufficient change in acceptance threshold occur, and no change in posterior.
-We stop single-locus model after four iterations with epsilon 0.13, when 1074 accepted instances with effective sample size of 700 (Figure~\ref{fig:ess-single}).
-We ran the algorithm with different initialization seeds and different number of replicates for simulations and got similar posterior (Figure~\ref{fig:seeds}).
+%a bit adapted/expanded description from Syga et al. 2021: % TODO if you adapt from it, then we should cite it.
+
+Briefly, the ABS-SMC algorithm employs sequential importance sampling over multiple iterations. 
+In iteration $t$, the algorithm draws parameter vectors $\theta_i$, also called \emph{particles}, from a given proposal distribution and simulates data $X_i$ from the model until $n_t$ particles are accepted.
+Acceptance is based on a comparison to observed data $X_{obs}$ using a distance function $\Delta(X_i, X_{obs})$ and acceptance threshold $\epsilon_t$, such that $\Delta(X_i, X_{obs}) \leq \epsilon_t$.
+For a distance function, we use $1-\likelihood$. % TODO this is unclear because \likelihood is defined above for X_{obs} so how do you apply it for X_i and X_{obs} together?
+% TODO something about how the weight is computed and used?
+The acceptance threshold $\epsilon_t$ is chosen as the median of the distances of the accepted particles in the previous iteration and $\epsilon_0=0.01$. 
+When $n_t$ particles are accepted, a new proposal distribution is constructed using a multivariate normal kernel with adaptive covariance matrix. % TODO ref
+In the first iteration the prior (see below) is used as a proposal distribution. 
+$n_t$ is adapted at every iteration $t$ using the \emph{adaptive population strategy}~\citep{Klinger2017}.
+The algorithm terminates when no more sufficient change in acceptance threshold occur, and no change in posterior. % TODO unclear
+Inference with the single-locus model terminated after four iterations with $\epsilon=0.13$, 1074 accepted parameter vectors and effective sample size (ESS) 700 (\autoref{fig:ess-single}).
+Running the inference algorithm with different initialization seeds and different number of simulations with produced similar outcomes (\autoref{fig:seeds}).
 %TODO if we have more models, write about them here also.
 
-After getting weighted particles of the posterior distribution by ABC-SMC scheme, we calculate Kernel Density Estimate using Gaussian kernels and $50,000$ points are sampled from KDE.
-We use the KDE and the points to calculate mode, highest density interval (HDI), for drawing distribution plots, etc.
+After producing a set of weighted particles from the the posterior distribution using the above ABC-SMC algorithm, we approximate the posterior using kernel density estimation (KDE) with Gaussian kernels.
+We then sample $50,000$ samples from the posterior KDE to compute the posterior mode, highest density interval (HDI), plotting the posterior, etc.
 
-\paragraph{Model comparison.} We perform model selection using WAIC (widely applicable information criterion)~\citep{gelman2013bayesian},
+\paragraph{Model comparison.} We perform model selection using WAIC, the widely applicable information criterion ~\citep{gelman2013bayesian},
 \begin{equation} \label{eq:WAIC}
 \begin{aligned}
-\mathit{WAIC}(\theta, \vec{X}) &= -2\log\mathbb{E}[\mathcal{L}(\theta \mid \vec{X})] + 2\mathbb{V}[\log\mathcal{L}(\theta \mid \vec{X})]
+\mathit{WAIC}(\theta, \vec{X}) &\ =\ 
+-2\log\mathbb{E}[\mathcal{L}(\theta \mid \vec{X})]\ +\ 2\mathbb{V}[\log\mathcal{L}(\theta \mid \vec{X})]
 \end{aligned}
 \end{equation}
-where $\mathbb{E}[\cdot]$ and $\mathbb{V}[\cdot]$ are the expectation and variance operators taken over the posterior distribution $P(\theta \mid \vec{X})$.
-WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy~\citep{Kass1995}.
+where $\theta$ is a parameter vector, $X$ are the model or experiment results (i.e., fixation of aneuploidy by generation $t$ etc.), and $\mathbb{E}[\cdot]$ and $\mathbb{V}[\cdot]$ are the expectation and variance taken over the posterior distribution $P(\theta \mid \vec{X})$.
+WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy and a difference of 2 is a popular threshold for model comparison~\citep{Kass1995}.
 % TODO do we need to say "and a difference of 2 is a popular threshold for model comparison."? Meanwile WAICs are so small that there's no such difference
+% YOAV: let's discuss
  % TODO do we use more methods? Describe probably posterior plots comparison method, and Bayes factor if it will used.
 
-\paragraph{Prior distributions.} The prior distributions of $w_{2n+1}=1-c+b$, $w_{2n+1^*}=(1+s)(1-c)+b$ and $w_{2n^*}=1+s$ were obtained by estimating it from growth curves data that were previously obtained by~\citet{Yona2012} using mono-culture growth experiments.
-The raw growth curves data were not published before, but were used to produce figures 3C, 4A, and S2 in \citet{Yona2012}.
-We used \texttt{Curveball}\footnote{https://curveball.yoavram.com}, a dedicated method for predicting results of competition experiments from growth curve data~\citep{Ram2019}.
-\texttt{Curveball} takes growth curves of two strains growing separately in mono-culture and predicts how they would grow in a mixed culture, that is, it predicts the results of a competition assay.
+\paragraph{Prior distributions.} We used informative prior distributions for $w_{2n+1}=1-c+b$, $w_{2n+1^*}=(1+s)(1-c)+b$ and $w_{2n^*}=1+s$, which we estimated from growth curves data from mono-culture growth experiments previously reported by \citet[Figs.
+3C, 4A, and S2]{Yona2012}.
+We used \texttt{Curveball}, a method for predicting results of competition experiments from growth curve data~\citep[\href{https://curveball.yoavram.com}{curveball.yoavram.com}]{Ram2019}.
+Briefly, \texttt{Curveball} takes growth curves of two strains growing separately in mono-culture and predicts how they would grow in a mixed culture, that is, it predicts the results of a competition assay.
 From these predictions, relative fitness values can be computed.
+Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix.
+We sampled 10,000 samples to use as a prior distribution~(Figures~\ref{fig:growth-curves-30deg} and~\ref{fig:growth-curves-39deg}).
 
-We used growth curves of $2n+1$ and $2n$ in \SI{39}{\celsius} to estimate  $w_{2n+1}$.
-We used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$, assuming that fitness values shouldn't differ too much.
-We also tested different priors, but this one (denote it \emph{main prior}) was better in the sense of WAIC, prediction plots, estimated parameters.
+We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$.
+In lieu of a better prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$.
+Compared to other priors we tested, this prior produced lower WAIC, better posterior prediction plots, and more stable parameter estimates.
 %TODO To show ppc plots also?
 Write paragraph about prediction plots in model comparison section.
-\\We tested uniform prior (denote \emph{uninformative}), $\mathit{Uniform}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ and (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, leaving $w_{2n+1}$ prior as in main prior.
-In both scenarios the fitness estimates by the model looks too high: for the first scenario the median estimates were  $w_{2n+1}=2.767$, $w_{2n+1^*}=5.79$, $w_{2n^*}=5.827$ and for the second $w_{2n+1}=1.046$, $w_{2n+1^*}=4.642$, $w_{2n^*}=4.684$.
-We don't expect to have such high fitness values.
-Moreover the WAIC for the models with these priors are 0.57 and 1.9 respectively, while WAIC for the main prior model is 0.27.
-Also for the model with uninformative prior for all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ the fixation time of 2n+1 is about 16 in average, which contradicts unpublished observations that were at least 250-300.
+We tested a uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ and (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$.
+In both of these uninformative priors, the fitness estimates were much too high: for the first prior the median estimates were  $w_{2n+1}=2.767$, $w_{2n+1^*}=5.79$, $w_{2n^*}=5.827$ and for the second prior the estimates were $w_{2n+1}=1.046$, $w_{2n+1^*}=4.642$, $w_{2n^*}=4.684$.
+Such high fitness values are unreasonable.
+Moreover the WAIC for the models with these priors were 0.57 and 1.9, respectively, whereas the WAIC with the informative prior was 0.27.
+Also, with the uninformative prior for all fitness values, the fixation time of \anwt is about 16 on average.
 %TODO remove about unpublished observations...we already added it into likelihood function of the model.
-\\Also we tried to use additional growth curves.
-We used growth curves of 2n* (\emph{refined} from \citet{Yona2012}) and 2n+1 in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$.
+% YOAV so remove the entire sentence right?
+We tried to use additional growth curves.
+We used growth curves of \eumt\ (\emph{refined} strain from \citet{Yona2012}) and \anwt\ in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$.
 The same prior was used for $w_{2n*}/w_{2n+1*}$.
-This model didn't perform well, it resulted in WAIC 2.8.
-\\Another try was to use growth curves of $2n+1$ and $2n$ in \SI{30}{\celsius} to estimate $1-c$.
+This prior resulted in WAIC 2.8, compared to 0.27 with the above informative prior.
+We also tried to use growth curves of \anwt\ and \euwt\ in \SI{30}{\celsius} to estimate $1-c$.
 This estimation assumes that the cost of aneuploidy is the same in \SI{39}{\celsius} and \SI{30}{\celsius}; this might be incorrect, but we only assumed this to generate a prior distribution for the fitness values.
 The prior for $b$ was taken the same as for $c$.
-This model didn't work at all: no parameter sets were found with the likelihood greater than zero.
+This prior did not work at all, as no parameter sets were found with an approximate likelihood greater than zero.
 %TODO it make sense because bla bla (to explain)?
+%YOAV: if you have an explanation, you can add it
 
-Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution, defined by the maximum-likelihood covariance matrix.
-Thus, we sampled 10,000 values for appropriate fitness, which we used as prior distributions.
-See Figures~\ref{fig:growth-curves-30deg} and~\ref{fig:growth-curves-39deg}.
-
-For the mutation rate $\mu$ and aneuploidy rate $\delta$ we used uniform priors, $\mathit{Uniform}(1e-9,1e-5)$ and $\mathit{Uniform}(1e-6,1e-2)$ respectively.
-For the single locus model without aneuploidy, wider uniform prior for the mutation rate was used - $\mathit{Uniform}(1e-10,1e-5)$.
+For the mutation rate, $\mu$, and aneuploidy rate, $\delta$, we used uninformative, uniform priors, $\mu \sim \mathit{U}(10^{-9}\ ,\ 10^{-5})$ and $\delta \sim \mathit{U}(10^{-6} , 10^{-2})$.
+For a model without aneuploidy ($\delta=0$), we used a wider uniform prior for the mutation rate, $\mu \sim \mathit{U}(10^{-10},10^{-5})$.
 

### 2021-12-21 ms/ms.tex
small updates

@@
 We fit these models to the experimental results of \citet{Yona2012} using an \emph{approximate Bayesian computation} framework~\citep{Sisson2009} to infer model parameters, including selection coefficients and rates of aneuploidy and mutation, and to perform model selection between different models, thereby testing hypotheses about the evolutionary process.
-%TODO do model selection by comparing Bayes factor or ignore this part.
 We analyze these evolutionary-genetic models to estimate the effects of parameters on the adaptation time and the probability for adaptation via aneuploidy.
@@
 % TODO IK do we assume b>c?
+% IK ANS we assume that 2n+1>2n, see below, so yes.
 }
@@
 % TODO IK we are missing 2n+1->2n, 2n*->2n+1*, why? to neglect them we need to assume w(2n+1)>w(2n) and w(2n*)>w(2n+1*), but I thought we don't assume that but rather we infer that. 
+% IK ANS We know that w(2n+1)>w(2n) because trisomy is fixated in the experiment. We can also assume that w(2n*)>w(2n+1) by Fig4A of Yona paper (refined vs trisomic growth) and because it is fixated. 
+% Actually I assume in code that  w(2n*)>w(2n+1*)>w(2n+1)>w(2n) just for speeding up the simulations (I don't check some parameters combinations in this way).
+But I checked also that when I don't adding this constrain the result was the same.
+However I dind't add reverse direction in this case.
     \begin{equation} \label{eq:mutation-aneuploidy-single}
@@
 We overcome this problem by approximating the likelihood using multiple simulations (1,000 simulations per parameter vector for the single-locus model).
+Denote by $\hat{\likelihood}(X | C)$ the approximate likelihood based on data of simulations $X$ and the expected conditions $C$.
 %TODO if we have more models, write how much replicates were used.
@@
 Acceptance is based on a comparison to observed data $X_{obs}$ using a distance function $\Delta(X_i, X_{obs})$ and acceptance threshold $\epsilon_t$, such that $\Delta(X_i, X_{obs}) \leq \epsilon_t$.
-For a distance function, we use $1-\likelihood$. % TODO this is unclear because \likelihood is defined above for X_{obs} so how do you apply it for X_i and X_{obs} together?
+%TODO cite Syga here?
+In our case $X_i$ is 1000 %TODO check if this is the number for all models
+simulations of the model, and the distance function is $1-\hat{\likelihood}(X_i | X_{obs})$
 % TODO something about how the weight is computed and used?
@@
     \textbf{(B)} Simulations parameters are from the maximum a posterior estimation of the single-locus model \emph{without} aneuploidy fit. 
-    %TODO get it from MAP and not mean
     The likelihood for the heat-stress experiment (\autoref{eq:heatstress-noaneuploidy-likelihood}) is 0.75.
@@
     \caption{
-    \textbf{Posterior for different runs} %TODO make more compact; one legend is enough.
-    For each parameter of the single-locus model, the posterior is represented for runs with different initialization seeds and different number of replicas in simulations run. 
+    \textbf{Posterior for different runs}
+    For each parameter of the single-locus model, the posterior is represented for different runs with different number of simulations.
     } 

### 2021-12-23 ms/ms.tex
revised results and all figures

@@
 For the mutation rate, $\mu$, and aneuploidy rate, $\delta$, we used uninformative, uniform priors, $\mu \sim \mathit{U}(10^{-9}\ ,\ 10^{-5})$ and $\delta \sim \mathit{U}(10^{-6} , 10^{-2})$.
+% TODO what about aneuploidy loss?
 For a model without aneuploidy ($\delta=0$), we used a wider uniform prior for the mutation rate, $\mu \sim \mathit{U}(10^{-10},10^{-5})$.
@@
 
-\subsection*{Model fitting}
+\subsection*{Inference: single-locus model}
 
@@
 
-\paragraph{Single-locus model parameters inference.} 
-First we inference the parameters of the single-locus model. The mode and 50\% hdi of the parameters posterior are:
-mutation rate - $3.06\substack{+1.229 \\ -2.84}\times10^{-6}$,
-trisomy rate - $1.61\substack{+1.122 \\ -0.197}\times10^{-3}$,
-2n+1 fitness - $1.022\substack{+0.001 \\ -0.002}$,
-2n+1* fitness - $1.025\substack{+0.001 \\ -0.001}$,
-2n* fitness - $1.028\substack{+0.001 \\ -0.002}$,
+\paragraph{Parameter estimation.} 
+We used ABC-SMC to infer the posterior distribution of the parameters of the single-locus model.
+The MAP (maximum a posteriori) and 50\% HDI (highest density interval) of the parameters posterior are~(\autoref{fig:posterior}): % TODO I changed "mode" to "MAP", ok?
+mutation rate, $\mu=3.06\substack{+1.229 \\ -2.84}\times10^{-6}$,
+aneuploidy gain rate, $\delta=1.61\substack{+1.122 \\ -0.197}\times10^{-3}$,
+% TODO what about aneuploidy loss?
+fitness of \anwt, $w_{2n+1}=1.022\substack{+0.001 \\ -0.002}$,
+fitness of \anmt, $w_{2n+1*}=1.025\substack{+0.001 \\ -0.001}$,
+and fitness of \eumt, $w_{2n*}=1.028\substack{+0.001 \\ -0.002}$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$.
+This estimated aneuploidy gain rate agrees with previous estimates. % TODO refs
+The mutation rate corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014}. 
+%We can see from the posterior distribution~(\autoref{fig:posterior}) the correlation between fitness of different genotypes. 
+% TODO say/plot something about the correlation, is it interesting?). 
+% YES but the sentence above is not enough.
 
-This trisomy rate appropriates to the previous estimates; the mutation rate corresponds to the mutations target size of $10^{4}$ if we assume the mutation rate of single base to be about $2*10^{-10}$ as estimated by ~\cite{Zhu2014}.
-We can see from the posterior distribution~(\autoref{fig:posterior}) the correlation between fitness of different genotypes.. (TODO say/plot something about the correlation, is it interesting?).
-
+% FIG posterior
 \begin{figure}[h!]
@@
   \caption{
-  \textbf{Single-locus model parameters posterior density.}
+  \textbf{Posterior distribution of single-locus model.}
 %TODO histogram on diagonal
-On the diagonal plots, the posterior density for each parameter is presented .
-Other plots represent joint posterior density of two parameters where blue color has the lowest density and yellow has the highest density.
-Red point and the dashed lines represent the maximum a posteriori probability estimate:
-mutation rate - $3.06\times10^{-6}$,
-trisomy rate - $1.61\times10^{-4}$,
-\anwt\ fitness - $1.022$,
-\anmt\ fitness - $1.025$,
-\eumt\ fitness - $1.028$.
+%TODO mutation rate tick labels should be in 10^-6, not 10^-5, and write 10^-6 not 1e-5; also it looks like you get positive density for mutation rate = 0 which doesnt make sense
+% change trisomy rate to aneuploidy gain rate; what about loss rate?
+% remove the y-ticks from the diagonal, they are meaningless, right?
+% fitness 2n+1 (not 2n+1 fitness)
+% in some cases the MAP doesn't look correct on the plot - did you compute the MAP on the joint for all 5 parameters or on the marginal of each parameter? something is weird here.
+On the diagonal, the posterior kernel density estimate for each parameter. 
+Below the diagonal, joint posterior density of two parameters (dark purple and bright yellow for low and high density, respectively.
+Red markers and orange lines for the MAP estimates.
+% TODO change dashed orange to red and make it a bit bolder.
 } 
@@
 
-\paragraph{Single-locus model fits the data well, but only if contains aneuploidy.} Single-locus model fits well to the data~(\autoref{fig:likelihood}A).
-On \autoref{fig:ppc-plot} the dynamics of frequency change of each genotype in time is shown. We can see that \anmt\ never reach substantial frequency in the population.  
-Sensitivity analysis~(\autoref{fig:sensitivity}) shows that changing the parameter values cause the model to fit less.
-Furthermore, we measured fixation time of \anwt\ in population, and found it appropriate to the unpublished data of the experiment ~\citep{Yona2012} - about 300 generations in average, that supports the correctness of the inference.
-On the other hand, if we don't take into consideration aneuploidy in the model, i.e.
-take aneuploidy rate be equal to zero, then the model can't explain the data: the fit is not so good~(\autoref{fig:likelihood}A; \autoref{fig:2n*-fixation}).
-The inference parameter values of the model without aneuploidy are:
-mutation rate - $1.251\substack{+0.146 \\ -0.177}\times10^{-8}$,
+\paragraph{Model checking and comparison.}
+The single-locus model fits the data well: in simulations using the MAP estimate parameters \euwt\ fixed in 59\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\autoref{fig:2n*-fixation}) and \anwt\ fixed in roughly 300 generations, which agrees with experimental results.
+% TODO this fig shows fixation of 2n*, what about fixation of 2n+1?
+% TODO merge fig 3 and fig:likelihood (both panels of fig:likelihood) 
+Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\autoref{fig:ppc-plot}). % TODO move somewhere else?
+Furthermore, sensitivity analysis shows that changing the parameter values from the MAP estimate reduces the model fit to the experimental results~(\autoref{fig:sensitivity}).
+
+However, a model without aneuploidy, i.e.
+where the aneuploidy gain rate is fixed at zero $\delta=0$, cannot explain the data~(\autoref{fig:2n*-fixation} and \autoref{fig:likelihood}).
+The estimated parameter values without aneuploidy are:
+mutation rate, $\mu=1.251\substack{+0.146 \\ -0.177}\times10^{-8}$,
 2n+1 fitness - $1.01\substack{+0.001 \\ -0.002}$,
 2n+1* fitness - $1.011\substack{+0.001 \\ -0.001}$,
+%TODO why do you have fitness of aneuploidy when there is no aneuploidy?
 2n* fitness - $1.011\substack{+0.001 \\ -0.001}$. 
@@
   \centering
+  % TODO add legend
 \includegraphics[width=0.6\textwidth]{../figures/ppc-plot.pdf}
   \caption{
-  \textbf{Evolutionary dynamics with fitted single-locus model.} 
-  Frequency of the four genotypes averaged over 10,000 simulations of the model. Each line represent one of the 10 parameter sets drawed randomly from the inference posterior.
+  \textbf{Single-locus model posterior genotype frequency dynamics.} 
+  The posterior predicted frequency of the four genotypes. Each line is the average of 10,000 simulations using parameters drawn from the posterior distribution (\autoref{fig:posterior}).
   }
@@
   \caption{
-  \textbf{Time of \eumt\ fixation cannot be explained without aneuploidy.}
-Each color represents distribution to fixation of \eumt\ of $10,000$ simulations for different modifications of single-locus model.
-\emph{Blue} bars correspond to the model with aneuploidy, where parameters are taken from the median of the posterior of the inference.
-\emph{Orange} bars correspond to the same model with same parameters, except aneuploidy that is zero.
-\emph{Green} bars correspond to the model without aneuploidy, i.e.
-with aneuploidy rate zero, where parameters are taken from the median of the posterior of the inference for this model.
+  \textbf{Adaptation time cannot be explained without aneuploidy.}
+In blue, the distribution of adaptation time (fixation of \eumt) in $10,000$ simulations of the single-locus model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: in orange, a model with the same parameter values except $\delta=0$; in green, a model fitted to the data assuming $\delta=0$.
+%, where parameters are taken from the median of the posterior of the inference. % why madian and not MAP?
+In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and the other by generation 2,350 (dashed lines) but not before generation 200.
+Thus, the blue distribution demonstrates a better fit compared to the green, and the yellow demonstrates a very poor fit.
 The last bin contains all the simulations with time equal or greater than 3000.
-Dashed lines represent experimental data: generations 1700 and 2350.
-When the difference in number of simulations between ranges 0-1700 and 1700-2350 is small, and there are no much simulations after generation 2350 - the fit is good.
-We can see that the fit of the model with aneuploidy (\emph{blue}) is the best.
-There's no fit at all for the model that is represented by \emph{yellow} bars.
 }
+% TODO add legend
+% TODO change y label to % simulations
+% change x label to Time to fixation of 2n*
+% makes vertical line black
+% add line for t=200?
+% does x axis start at zero? can you add 0 to x ticks?
   \label{fig:2n*-fixation}
@@
 
-\paragraph{Complex model: Inference and Comparison.}
+\paragraph{Inference: multi-locus model.}
 
@@
 
-%%%% Fig - Curveball 30deg %%%
+% Fig Curveball 30deg
 % generated with growth_curves.ipynb
@@
 \end{figure}
+% TODO add caption
 
-%%%% Fig - Curveball 39deg %%%
+% Fig Curveball 39deg
 % generated with growth_curves.ipynb
@@
 \end{figure}
+% TODO add caption, combine to one figure with the previous one
 
-% likelihood plots
+% Fig single-locus model with and without aneuploidy
 \begin{figure}[b!]
@@
   \caption{
-    \textbf{Single-locus model fit.} Cumulative distribution of 10000 simulations are shown.
-    \textbf{(A)} Simulations parameters are from the maximum a posterior estimation of the single-locus model fit.
-The likelihood for the heat-stress experiment (\autoref{eq:heatstress-likelihood}) is 0.86.
+  % TODO change x label - "Time to fixation of 2n*"
+  % plot both lines on one panel with two colors (blue and red for with and without? choose nice color, not the default 'r' and 'b'!), make the without in dashed.
+    \textbf{Single-locus model fit with and without aneuploidy.} Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy (i.e., $\delta=0$ is fixed).
     \textbf{(B)} Simulations parameters are from the maximum a posterior estimation of the single-locus model \emph{without} aneuploidy fit. 
-    The likelihood for the heat-stress experiment (\autoref{eq:heatstress-noaneuploidy-likelihood}) is 0.75.
+     The MAP likelihood (\autoref{eq:heatstress-likelihood}) is 0.86 and 0.75 with and without aneuploidy. % TODO likelihood ratio test? bayes factor? no sense in just giving the likelihoods.
   }
@@
 
-% sensitivity analysis plots
-
+% Fig likelihood profile
 \begin{figure}[p]
@@
   \caption{
-    \textbf{Sensitivity analysis.} We take parameters that are maximum a posterior (MAP) for single locus model.
-Then we change only one parameter at time while measuring the model likelihood of the parameters.
-x-axis label represent the parameter that we change.
-Yellow point indicates the MAP value.
+    \textbf{Likelihood profiles.} The different curves show how the single-locus model log-likelihood changes when each of the parameter changes while the other parameters remain fixed at their MAP estimates.
   }
+  % TODO mark the MAP parameter on the figure with a vertical line
+  % show the log-likelihood rather than the likelihood
+  % can we add more points? can we do this on joint likelihood of two parameters?
+  % have the fitness parameters on the same row
+  % x labels start with capital letter, should include the parameter (Mutation rate, $\mu$) and the name (Fitness of 2n+1, w_{2n+1}), dont use w(xxx) but rather w_{xxx} as in the main text
+  % y label "Approximate log-likelihood"
+  % trisomy rate -> aneuploidy gain rate
   \label{fig:sensitivity}
@@
 
+% Fig ABC convergence
 \begin{figure}[h]
@@
     \caption{
-    \textbf{Single-locus model convergence.} Pyabc run over four iterations.
-The model is converged when acceptance threshold (epsilon) is 0.13, the number of particles (accepted instances) is 1074 giving the effective sample size (ESS) 700.
+    \textbf{Single-locus model inference convergence.} Measures of convergence for the ABC-SMC algorithm with the single-locus model.
+After iteration number 4, the acceptance threshold was $\epsilon=0.13$, the acceptance rate was XXX, the number of accepted parameter sets ("particles") was 1,074, and the effective sample size (ESS) 700.
+% TODO add missing acceptance rate
     } 
     \label{fig:ess-single}
+    % TODO y label in first panel "Acceptance threshold, $\epsilon$"
+    % add panel labels A B C D
+    % x label - just once
+	% should we do 1-3 more iterations? it is not clear that it converged.
 \end{figure}
 
+% Fig 
 \begin{figure}[h]
@@
     \caption{
-    \textbf{Posterior for different runs}
-    For each parameter of the single-locus model, the posterior is represented for different runs with different number of simulations.
+    \textbf{Posterior distribution validation.}
+    The posterior distributions of the single-locus model parameters are roughly the same regardless of the number of simulations (10-1,000) used to approximate the likelihood.
     } 
@@
 \end{figure}
+% TODO put legend in panel A top right corner
+% y label "Density"; y ticks are meaningless, right? 
+% x labels - see comments above.
 

### 2021-12-26 ms/ms.tex
following phone call between YR and IK

@@
 \pagebreak
+%%%%%%%%%%%%%%%%%%%%%%%%
 % Models and Methods
@@
     \textbf{(A)} In the single-locus model, the four genotypes are: euploid wild-type, \euwt; euploid mutant, \eumt; aneuploid wild-type, \anwt; and aneuploid mutant, \anmt.
-    Overall there are two possible trajectories from \euwt\ to \eumt.
-    \textbf{(B)} In the multi-locus model, each genotype is characterized by its karyotype, \euwt\ or \anwt, and the number of accumulated beneficial mutations, denoted by stars. 
-    In both panels arrows denote transitions between genotypes, with transitions rates: $\mu$ and $\tau \mu$, beneficial mutation rates in euploid and aneuploid cells; $\delta$, aneuploidy gain rate; $\delta_L$, aneuploidy loss rate.
+    Overall there are two possible trajectories from \euwt\; to \eumt.
+    \textbf{(B)} In the multi-locus model, each genotype is characterized by its karyotype, \euwt\; or \anwt, and the number of accumulated beneficial mutations, denoted by stars. 
+    In both panels arrows denote transitions between genotypes, with transitions rates: $\mu$ and $\tau \mu$, beneficial mutation rates in euploid and aneuploid cells; $\delta$, aneuploidy rate.
     % TODO IK: change k mu to tau mu in figure; also k mu is too close to the arrow right now; change alpha to delta in the figure 
@@
 \end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%
 
+%%%%%%%%%%%%%%%%%%%%%%%%
 \paragraph*{Evolutionary Models.}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%
 \paragraph*{Single-locus model.}
@@
 By default, we assume $\tau=33/32$, as the wild-type \yeast strains in the experiments by \citep{Yona2012} are diploid with 32 chromosomes, and the aneuploid strains are trisomic, with 33 chromosomes.
-Aneuploidy is formed by chromosome mis-segregation, so that cells transition from \euwt\; to \anwt\ with probability $\delta$, the aneuploidy gain rate.
-Aneuploidy is lost, transitioning cells from \anmt\; to \eumt\; with probability $\delta_L$, the aneuploidy loss rate.
+We neglect back-mutations (i.e., from \eumt\; to \euwt\; and from \anmt\; to \anwt).
+Aneuploidy is formed by chromosome mis-segregation, so that cells transition from \euwt\; to \anwt\; and from \anmt\ to \eumt\ with probability $\delta$, the aneuploidy rate.
+That is, we assume chromosomes are gained and lost at the same rate, and we neglect events that form a less-fit genotype (i.e., \anwt\; to \euwt\; and \eumt\; to \anmt).
 The fitness values of the four genotypes are given by \autoref{table:single-locus}.
@@
 $s \ge 0$ is the selection coefficient of a beneficial mutation;
-$b \ge 0$ is the selection coefficient of aneuploidy; 
-and $0 \le c \le 1$ is the fitness cost of aneuploidy.
-% TODO IK do we assume b>c?
-% IK ANS we assume that 2n+1>2n, see below, so yes.
+$0 \le c \le 1$ is the fitness cost of aneuploidy;
+and $b \ge c$ is the selection coefficient, or fitness benefit, of aneuploidy.
 }
 \end{table}
-%%%
+%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 The effect of mutation and aneuploidy on genotype frequencies is given by
-% TODO IK we are missing 2n+1->2n, 2n*->2n+1*, why? to neglect them we need to assume w(2n+1)>w(2n) and w(2n*)>w(2n+1*), but I thought we don't assume that but rather we infer that. 
-% IK ANS We know that w(2n+1)>w(2n) because trisomy is fixated in the experiment. We can also assume that w(2n*)>w(2n+1) by Fig4A of Yona paper (refined vs trisomic growth) and because it is fixated. 
-% Actually I assume in code that  w(2n*)>w(2n+1*)>w(2n+1)>w(2n) just for speeding up the simulations (I don't check some parameters combinations in this way).
-But I checked also that when I don't adding this constrain the result was the same.
-However I dind't add reverse direction in this case.
     \begin{equation} \label{eq:mutation-aneuploidy-single}
@@
       &f^m_{2n+1} &=&\; \delta f^s_{2n} + (1 - \tau \mu) f^s_{2n+1}  \;,\\
-      &f^m_{2n+1^*} &=&\; \tau \mu f^s_{2n+1} + (1-\delta_L) f^s_{2n+1^*}  \;,\\
-      &f^m_{2n^*} &=&\; \mu f^s_{2n} + \delta_L f^s_{2n+1} + f^s_{2n^*}  \;.
+      &f^m_{2n+1^*} &=&\; \tau \mu f^s_{2n+1} + (1-\delta) f^s_{2n+1^*}  \;,\\
+      &f^m_{2n^*} &=&\; \mu f^s_{2n} + \delta f^s_{2n+1} + f^s_{2n^*}  \;.
     \end{aligned}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%
 \paragraph*{Multi-locus model.}
@@
 The rest of the parameters ($N$, $\mu$, $\tau$, $\delta$, $\delta_L$, $b$, $c$) are the same as in the single-locus model.
-% ILIA: I don’t know if it should be mentioned here, but the meaning of the mutation rate is a bit differ for multi-locus model.
-For the single-locus mutation rate value is the probability to get specific mutation (maybe one of a few).
-For the multi-locus it is the probability to get some beneficial mutation (one of a many).
-So, we don’t expect to receive same values for mutation rate for the models.
-% YOAV: let's talk about it, I don't necessarily agree
-% ILIA: I don’t have in the code 𝜏 parameter for multi-locus model (it’s like 1 there). Do you think it worth adding it for consistency?
-% YOAV: if our inference tells us that tau=1 is as good as tau>1, then maybe we can remove this parameter from the model presentation altogether, and simply introduce it when we discuss this result as part of the sensitivity analysis section
+However, since the multi-locus model allows several mutations with smaller fitness effects to accumulate, we expect the mutation rate to be higher compared to the single-locus model, which focuses on a single, large-effect mutation.
+% TODO : add tau to model fixed at 33/32.
+
 The fitness of the different genotypes is the same as in the single-locus model (\autoref{table:single-locus}), except that the fitness contribution of $k$ beneficial mutations is the product of their independent effects, $\prod_{i=1}^k(1+s_i)$, instead of the contribution of the single mutation allowed in the single-locus model, $(1+s)$, see \autoref{table:multi-locus}.
@@
 Each offspring cell inherits these properties from its mother cell.
-Then, with probability $\mu$ or $\tau \mu$ for euploid and aneuploid cells, % TODO no tau in multi-locus model?
+Then, with probability $\mu$ or $\tau \mu$ for euploid and aneuploid cells,
 respectively, a new beneficial mutation is accumulated, such that the number of mutations is $k+1$, and its effect $s_{k+1}$ is drawn from an exponential distribution with expected value $s$, such that the contribution of the mutations to the fitness is $\prod_{j=0}^{k+1}{(1+s_j)}$.
@@
 $s \ge 0$ is the selection coefficient of a beneficial mutation;
-$b \ge 0$ is the selection coefficient of aneuploidy; 
-and $0 \le c \le 1$ is the fitness cost of aneuploidy.
-%TODO do we assume b>c ?
+$0 \le c \le 1$ is the fitness cost of aneuploidy;
+and $b \ge c$ is the selection coefficient, or fitness benefit, of aneuploidy.
+% TODO check if b > c in code
 }
 \end{table}
-%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 From unpublished results, aneuploidy did not fix before at least 200 generations elapsed.
-The experiment continued with two populations, in which aneuploidy was eliminated by generation 1700 and 2350.
-%In the pH-stress experiment, four populations of \yeast evolved under high-ph stress (8.6).
-Aneuploidy fixed during the first 150 generations.
-It was fully eliminated in two populations and partly eliminated in two populations by generation 750.
-%TODO do we use pH?
+The experiment continued with two populations, in which aneuploidy was eliminated by generation 1,700 and 2,350.
 
-\paragraph{Likelihood function.}
+\paragraph{Likelihood function.} % TODO fix this .... (YR)
 Denote by $A_{t,T}$ the fixation of aneuploidy by generation $t$ conditioned on no fixation up to generation $T$; by $B_{t}$ the extinction of aneuploidy by generation $t$; and by $C_{t,T}$ the extinction of aneuploidy by generation $t$ conditioned on no extinction by generation $T$.
@@
 \begin{multline} \label{eq:heatstress-likelihood}
-\likelihood(\theta\ |\ A_{450,200}\ ,\ B_{1700}\ ,\ C_{2350, 1700} ) = \\
-P^4(A_{450,200}) \cdot \Big(1 - P^4(\lnot{B_{1700}} | A_{450,200} ) - P^4(\lnot{C_{2350, 1700}}| A_{450,200}) + P^4(\lnot{B_{1700}} \land \lnot{L^*_{2350, 1700}}| A_{450,200})\Big)	\;,	
+\likelihood(\theta) = \\
+P^4(A_{450,200}) \cdot \Big(1 - P^4(\lnot{B_{1700}} | A_{450,200} ) - P^4(\lnot{C_{2350, 1700}}| A_{450,200}) + P^4(\lnot{B_{1700}} \land \lnot{L^*_{2350, 1700}}| A_{450,200})\Big)	\;,	 % TODO fix L* (YR)
 \end{multline}
 where $P(X)$ is the probability of event $X$ and $\lnot$ means \emph{not}.
-
-%The likelihood function of the model with parameter set $\theta$ for the pH-stress experiment is
-%\begin{equation} \label{eq:phstress-likelihood} % TODO: add ph-Stress model to results or remove it from here (if remained add A_t to notations, it is At,T now only)
-%\likelihood(\theta | A_{150}, L_{750} ) = \\
-% P^4(B_{150}) \cdot 6 \cdot	P^2(B_{600} | A_{150}) \cdot P^2(\lnot{B_{600}} | A_{150}) \;.  
-% \end{equation}
 
@@
 \begin{multline} \label{eq:heatstress-noaneuploidy-likelihood}
-\likelihood(\theta |  M_{1700}, M^*_{2350, 1700} ) = 
+\likelihood_{!}(\theta) = 
 \Big(1 - P^4(\lnot{M_{1700}}) - P^4(\lnot{Q_{2350, 1700}}) + P^4(\lnot{M_{1700}} \land \lnot{Q_{2350, 1700}})\Big)	\;,	
@@
 Because our model, just like the Wright-Fisher model, is non-linear and stochastic, these likelihood functions are intractable.
-We overcome this problem by approximating the likelihood using multiple simulations (1,000 simulations per parameter vector for the single-locus model).
-Denote by $\hat{\likelihood}(X | C)$ the approximate likelihood based on data of simulations $X$ and the expected conditions $C$.
-%TODO if we have more models, write how much replicates were used.
+We overcome this problem by approximating the likelihood using simulations: we simulate 1,000 experiments per parameter vector.
+This approximate likelihood is denoted by $\hat{\likelihood}$.
+% TODO if we show results of multi-locus model, write how much replicates were used.
 % TODO ref to place where we show 1,000 is enough
@@
 It is therefore suitable to our case, in which the likelihood function can only be approximated from simulations, and cannot be directly computed.
-%a bit adapted/expanded description from Syga et al. 2021: % TODO if you adapt from it, then we should cite it.
 
-Briefly, the ABS-SMC algorithm employs sequential importance sampling over multiple iterations. 
+Briefly, the ABS-SMC algorithm employs sequential importance sampling over multiple iterations~\citep{Syga2021}. 
 In iteration $t$, the algorithm draws parameter vectors $\theta_i$, also called \emph{particles}, from a given proposal distribution and simulates data $X_i$ from the model until $n_t$ particles are accepted.
+We accept a proposal $\theta_i$ if its approximate likelihood $\hat{\likelihood}(\theta_i)$
 Acceptance is based on a comparison to observed data $X_{obs}$ using a distance function $\Delta(X_i, X_{obs})$ and acceptance threshold $\epsilon_t$, such that $\Delta(X_i, X_{obs}) \leq \epsilon_t$.
-%TODO cite Syga here?
-In our case $X_i$ is 1000 %TODO check if this is the number for all models
-simulations of the model, and the distance function is $1-\hat{\likelihood}(X_i | X_{obs})$
+% TODO replace Delta with approx likelihood (YR)
+In our case $X_i$ is 1,000 simulations of the model, and the distance function is $1-\hat{\likelihood}(X_i | X_{obs})$
 % TODO something about how the weight is computed and used?
 The acceptance threshold $\epsilon_t$ is chosen as the median of the distances of the accepted particles in the previous iteration and $\epsilon_0=0.01$. 
-When $n_t$ particles are accepted, a new proposal distribution is constructed using a multivariate normal kernel with adaptive covariance matrix. % TODO ref
+When $n_t$ particles are accepted, a new proposal distribution is constructed using a multivariate normal kernel with adaptive covariance matrix. % TODO ref - Klinger? Syga? Sisson?
 In the first iteration the prior (see below) is used as a proposal distribution. 
 $n_t$ is adapted at every iteration $t$ using the \emph{adaptive population strategy}~\citep{Klinger2017}.
-The algorithm terminates when no more sufficient change in acceptance threshold occur, and no change in posterior. % TODO unclear
-Inference with the single-locus model terminated after four iterations with $\epsilon=0.13$, 1074 accepted parameter vectors and effective sample size (ESS) 700 (\autoref{fig:ess-single}).
+Acceptance is determined according to the approximate likelihood, which has a maximum value of 0.875.
+Thus, we terminated the inference when $\epsilon \le 0.13$ after four iterations, with 1,074 accepted parameter vectors and effective sample size (ESS) 700 (\autoref{fig:ess-single}).
+% TODO for multi-locus model we need to say something else
 Running the inference algorithm with different initialization seeds and different number of simulations with produced similar outcomes (\autoref{fig:seeds}).
-%TODO if we have more models, write about them here also.
+%TODO for multi-locus model, write about them here also.
 
-After producing a set of weighted particles from the the posterior distribution using the above ABC-SMC algorithm, we approximate the posterior using kernel density estimation (KDE) with Gaussian kernels.
-We then sample $50,000$ samples from the posterior KDE to compute the posterior mode, highest density interval (HDI), plotting the posterior, etc.
+After producing a set of weighted particles from the the posterior distribution using the above ABC-SMC algorithm, we approximate the posterior using kernel density estimation (KDE) with Gaussian kernels, from which we find the MAP (maximum a posteriori) estimate as the maximum of the KDE function.
+We then sample $50,000$ samples from the posterior KDE to compute the HDI (highest density interval) and visualize the posterior distribution with histograms.
 
@@
 WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy and a difference of 2 is a popular threshold for model comparison~\citep{Kass1995}.
-% TODO do we need to say "and a difference of 2 is a popular threshold for model comparison."? Meanwile WAICs are so small that there's no such difference
-% YOAV: let's discuss
- % TODO do we use more methods? Describe probably posterior plots comparison method, and Bayes factor if it will used.
+% TODO Describe posterior plots comparison method - Fig 4 & S3
 
@@
 
-We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$.
-In lieu of a better prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$.
-Compared to other priors we tested, this prior produced lower WAIC, better posterior prediction plots, and more stable parameter estimates.
-%TODO To show ppc plots also?
-Write paragraph about prediction plots in model comparison section.
+We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$. In lieu of a better prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$. 
+%TODO please report here the point estimate and 95% CI for the three distributions, I only have for 1-c+b: 0.9863 (95% CI 0.9790 - 0.9939).
+Compared to other priors we tested, this prior produced lower WAIC, better posterior prediction plots, and more stable parameter estimates, as follows.
 We tested a uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ and (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$.
@@
 Moreover the WAIC for the models with these priors were 0.57 and 1.9, respectively, whereas the WAIC with the informative prior was 0.27.
-Also, with the uninformative prior for all fitness values, the fixation time of \anwt is about 16 on average.
-%TODO remove about unpublished observations...we already added it into likelihood function of the model.
-% YOAV so remove the entire sentence right?
+% TODO if you need to remove this sentence: Also, with the uninformative prior for all fitness values, the fixation time of \anwt is about 16 on average, much lower than the (unpublished) experimental result, which is 200-300 generations.
+
 We tried to use additional growth curves.
-We used growth curves of \eumt\ (\emph{refined} strain from \citet{Yona2012}) and \anwt\ in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$.
+We used growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$.
 The same prior was used for $w_{2n*}/w_{2n+1*}$.
 This prior resulted in WAIC 2.8, compared to 0.27 with the above informative prior.
-We also tried to use growth curves of \anwt\ and \euwt\ in \SI{30}{\celsius} to estimate $1-c$.
+We also tried to use growth curves of \anwt\; and \euwt\; in \SI{30}{\celsius} to estimate $1-c$.
 This estimation assumes that the cost of aneuploidy is the same in \SI{39}{\celsius} and \SI{30}{\celsius}; this might be incorrect, but we only assumed this to generate a prior distribution for the fitness values.
@@
 This prior did not work at all, as no parameter sets were found with an approximate likelihood greater than zero.
-%TODO it make sense because bla bla (to explain)?
-%YOAV: if you have an explanation, you can add it
+% TODO it make sense because bla bla (to explain)
+
+% TODO : Actually I assume in code that  w(2n*)>w(2n+1*)>w(2n+1)>w(2n) just for speeding up the simulations (I don't check some parameters combinations in this way).
+But I checked also that when I don't adding this constrain the result was the same.
+However I dind't add reverse direction in this case.
 
 For the mutation rate, $\mu$, and aneuploidy rate, $\delta$, we used uninformative, uniform priors, $\mu \sim \mathit{U}(10^{-9}\ ,\ 10^{-5})$ and $\delta \sim \mathit{U}(10^{-6} , 10^{-2})$.
-% TODO what about aneuploidy loss?
 For a model without aneuploidy ($\delta=0$), we used a wider uniform prior for the mutation rate, $\mu \sim \mathit{U}(10^{-10},10^{-5})$.
 
-%TODO ILIA please report here the point estimate and 95% CI for the three distributions, I only have for 1-c+b: 0.9863 (95% CI 0.9790 - 0.9939).
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Results
@@
 We used ABC-SMC to infer the posterior distribution of the parameters of the single-locus model.
-The MAP (maximum a posteriori) and 50\% HDI (highest density interval) of the parameters posterior are~(\autoref{fig:posterior}): % TODO I changed "mode" to "MAP", ok?
+The MAP (maximum a posteriori) and 50\% HDI (highest density interval) of the parameters posterior are~(\autoref{fig:posterior}):
 mutation rate, $\mu=3.06\substack{+1.229 \\ -2.84}\times10^{-6}$,
-aneuploidy gain rate, $\delta=1.61\substack{+1.122 \\ -0.197}\times10^{-3}$,
-% TODO what about aneuploidy loss?
+aneuploidy rate, $\delta=1.61\substack{+1.122 \\ -0.197}\times10^{-3}$,
 fitness of \anwt, $w_{2n+1}=1.022\substack{+0.001 \\ -0.002}$,
@@
 and fitness of \eumt, $w_{2n*}=1.028\substack{+0.001 \\ -0.002}$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$.
-This estimated aneuploidy gain rate agrees with previous estimates. % TODO refs
+This estimated aneuploidy rate agrees with previous estimates. % TODO refs
 The mutation rate corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014}. 
 %We can see from the posterior distribution~(\autoref{fig:posterior}) the correlation between fitness of different genotypes. 
-% TODO say/plot something about the correlation, is it interesting?). 
-% YES but the sentence above is not enough.
+% TODO say/plot something about the correlation: which fitness values are correlated? how strong is it? why does it occur?
 
@@
   \textbf{Posterior distribution of single-locus model.}
-%TODO histogram on diagonal
-%TODO mutation rate tick labels should be in 10^-6, not 10^-5, and write 10^-6 not 1e-5; also it looks like you get positive density for mutation rate = 0 which doesnt make sense
-% change trisomy rate to aneuploidy gain rate; what about loss rate?
+% TODO histogram on diagonal, 2d histogram without interpolation below diagonal
+% TODO mutation rate tick labels should be in 10^-6, not 10^-5, and write 10^-6 not 1e-5; also it looks like you get positive density for mutation rate = 0 which doesnt make sense - set limit to 0.
+% change trisomy rate to aneuploidy gain rate
+% remove "mutation rate" from y label on top-left
 % remove the y-ticks from the diagonal, they are meaningless, right?
-% fitness 2n+1 (not 2n+1 fitness)
-% in some cases the MAP doesn't look correct on the plot - did you compute the MAP on the joint for all 5 parameters or on the marginal of each parameter? something is weird here.
+% Fitness of 2n+1, w_{2n+1} (not 2n+1 fitness)
+% TODO change dashed orange to red and make it a bit bolder.
 On the diagonal, the posterior kernel density estimate for each parameter. 
-Below the diagonal, joint posterior density of two parameters (dark purple and bright yellow for low and high density, respectively.
-Red markers and orange lines for the MAP estimates.
-% TODO change dashed orange to red and make it a bit bolder.
+Below the diagonal, joint posterior density of two parameters (dark purple and bright yellow for low and high density, respectively).
+Red markers and orange lines for the joint MAP estimates (which may differ from the marginal maximum density, as the marginal distribution integrates over all other parameters).
 } 
@@
 \end{figure}
+%%%%%%%%%%%%%%%%%%%%%%
 
@@
 Thereby, adding aneuploidy to the model, helps to attain good fit with higher mutation rate.
+% TODO WAIC?
 

### 2021-12-27 ms/ms.tex
revised likelihood and ABC description

@@
 The selection coefficient of the $i$-th accumulated mutation in each individual, $s_i$, is drawn from an exponential distribution with expected value $s$: $s_i \sim \mathit{Exp{(s)}}$.
-The rest of the parameters ($N$, $\mu$, $\tau$, $\delta$, $\delta_L$, $b$, $c$) are the same as in the single-locus model.
+The rest of the parameters ($N$, $\mu$, $\tau$, $\delta$, $b$, $c$) are the same as in the single-locus model.
 However, since the multi-locus model allows several mutations with smaller fitness effects to accumulate, we expect the mutation rate to be higher compared to the single-locus model, which focuses on a single, large-effect mutation.
@@
 \paragraph{Empirical evidence.}
-We use empirical results of evolutionary experiments reported by \citet{Yona2012}.
+
+We use the results of evolutionary experiments reported by \citet{Yona2012}.
 In their heat-stress experiment, four populations of \yeast evolved under \SI{39}{\celsius}.
@@
 The experiment continued with two populations, in which aneuploidy was eliminated by generation 1,700 and 2,350.
+%Thus, the empirical evidence can be summarized by $\vec{X}=\Big(200 \le T_i(2n+1) \le 450,\, i=1,2,3,4;\ T_1(2n^*)=1700;\ T_2(2n^*)=2350\Big)$ where $T_i(g)$ is the fixation time of genotype $g$ in population $i$.
 
-\paragraph{Likelihood function.} % TODO fix this .... (YR)
-Denote by $A_{t,T}$ the fixation of aneuploidy by generation $t$ conditioned on no fixation up to generation $T$; by $B_{t}$ the extinction of aneuploidy by generation $t$; and by $C_{t,T}$ the extinction of aneuploidy by generation $t$ conditioned on no extinction by generation $T$.
-The model likelihood function for parameter vector $\theta$ for the heat-stress experiment is
-\begin{multline} \label{eq:heatstress-likelihood}
-\likelihood(\theta) = \\
-P^4(A_{450,200}) \cdot \Big(1 - P^4(\lnot{B_{1700}} | A_{450,200} ) - P^4(\lnot{C_{2350, 1700}}| A_{450,200}) + P^4(\lnot{B_{1700}} \land \lnot{L^*_{2350, 1700}}| A_{450,200})\Big)	\;,	 % TODO fix L* (YR)
-\end{multline}
-where $P(X)$ is the probability of event $X$ and $\lnot$ means \emph{not}.
+\paragraph{Likelihood function.}
 
-For a model without aneuploidy (that is, when the aneuploidy rate is fixed at zero, $\delta=0$), denote by $M_{t}$ the fixation of \eumt\ by generation $t$, and by $Q_{t,T}$ the fixation of \eumt\ by generation $t$ conditioned on no fixation at generation $T$.
-The likelihood of a model without aneuploidy with parameter vector $\theta$ for the heat-stress experiment is
-\begin{multline} \label{eq:heatstress-noaneuploidy-likelihood}
-\likelihood_{!}(\theta) = 
-\Big(1 - P^4(\lnot{M_{1700}}) - P^4(\lnot{Q_{2350, 1700}}) + P^4(\lnot{M_{1700}} \land \lnot{Q_{2350, 1700}})\Big)	\;,	
-\end{multline}
+Because our model, just like the Wright-Fisher model, is non-linear and stochastic, computing the distribution of fixation time $T(g)$ of genotype $g$ for use in the likelihood function is intractable (it is even hard to use a diffusion-equation approximation due to the model having multiple genotypes, rather than just two).
+We overcome this problem by approximating the likelihood using simulations.
+We simulate 1,000 experiments per parameter vector $\theta = (\mu, \tau, \delta, s, b, c)$, resulting in a set of simulated observations $\tilde{\vec X} = \{\tilde{X}_i\}_{i=1}^{1000}$.
+We then compute the approximate likelihood,
+\begin{equation}\begin{aligned}
+\label{eq:heatstress-likelihood}
+\likelihood(\theta) = &\ P^4(200 \le T(2n+1) \le 450) \cdot 
+	\Big[1 - \\
+	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)- \\
+	&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big)+ \\
+	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big) 
+	\Big]\;,
+\end{aligned}\end{equation}
+where $!\{\ldots\}$ is the "logical not" operator, $P^4(\ldots)$ is the 4th power of $P(\ldots)$, and all probabilities $P_{\tilde{\vec X}}(\ldots)$ are approximated from the results of the simulations $\tilde{\vec X}$.
+For example, $P_{\tilde{\vec X}}\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)$ is approximated by taking simulations in which \anwt\; fixed before generation 450 but not before generation 200, and computing the fraction of such simulations in which \eumt\; did not fix by generation 1,700, and hence aneuploidy did not extinct before generation 1,700.
 
-Because our model, just like the Wright-Fisher model, is non-linear and stochastic, these likelihood functions are intractable.
-We overcome this problem by approximating the likelihood using simulations: we simulate 1,000 experiments per parameter vector.
-This approximate likelihood is denoted by $\hat{\likelihood}$.
+%Denote by $A_{t,T}$ the fixation of aneuploidy by generation $t$ conditioned on no fixation up to generation $T$; by $B_{t}$ the extinction of aneuploidy by generation $t$; and by $C_{t,T}$ the extinction of aneuploidy by generation $t$ conditioned on no extinction by generation $T$.
+%The likelihood function for parameter vector $\theta$ conditioned on the observed data $\vec{X}$ from the heat-stress experiment is
+%\begin{multline} \label{eq:heatstress-likelihood}
+%P^4(A_{450,200}) \cdot \Big(1 - P^4(\lnot{B_{1700}} | A_{450,200} ) - P^4(\lnot{C_{2350, 1700}}| A_{450,200}) + P^4(\lnot{B_{1700}} \land \lnot{C_{2350, 1700}}| A_{450,200})\Big)	\;,	 % TODO fix L* (YR)
+%\end{multline}
+%where $P(X)$ is the probability of event $X$ and $\lnot$ means \emph{not}.
+
+For a model without aneuploidy (that is, when the aneuploidy rate is fixed at zero, $\delta=0$), 
+%denote by $M_{t}$ the fixation of \eumt\ by generation $t$, and by $Q_{t,T}$ the fixation of \eumt\ by generation $t$ conditioned on no fixation at generation $T$.
+the likelihood is similarly approximated by
+% of a model without aneuploidy with parameter vector $\theta$ for the heat-stress experiment is
+\begin{equation}\begin{aligned}
+\label{eq:heatstress-noaneuploidy-likelihood}
+\likelihood_{!}(\theta) = &\ 
+	1 - 
+	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\}\big) - \\
+&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\}\big) + \\
+&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\}\big)
+\;.
+\end{aligned}\end{equation}
+
+%\begin{multline} \label{eq:heatstress-noaneuploidy-likelihood}
+%\likelihood_{!}(\theta) = 
+%\Big(1 - P^4(\lnot{M_{1700}}) - P^4(\lnot{Q_{2350, 1700}}) + P^4(\lnot{M_{1700}} \land \lnot{Q_{2350, 1700}})\Big)	\;,	
+%\end{multline}
+
 % TODO if we show results of multi-locus model, write how much replicates were used.
@@
 
-\paragraph{Parameter inference.} To infer model parameters, we use approximate Bayesian computation~\citep{Sunnaker2013} with a sequential Monte-Carlo scheme, or ABC-SMC~\citep{Sisson2009}, implemented in the \texttt{pyABC} Python
-package~\citep[\href{https://pyabc.readthedocs.io}{pyabc.readthedocs.io}]{Klinger2018}.
+\paragraph{Parameter inference.} To infer model parameters, we use approximate Bayesian computation with a sequential Monte-Carlo scheme, or ABC-SMC~\citep{Sisson2009}, implemented in the \texttt{pyABC} Python package~\citep[\href{https://pyabc.readthedocs.io}{pyabc.readthedocs.io}]{Klinger2018}.
 This approach uses numerical stochastic simulations of the model to infer a posterior distribution over the model parameters.
@@
 Briefly, the ABS-SMC algorithm employs sequential importance sampling over multiple iterations~\citep{Syga2021}. 
-In iteration $t$, the algorithm draws parameter vectors $\theta_i$, also called \emph{particles}, from a given proposal distribution and simulates data $X_i$ from the model until $n_t$ particles are accepted.
-We accept a proposal $\theta_i$ if its approximate likelihood $\hat{\likelihood}(\theta_i)$
-Acceptance is based on a comparison to observed data $X_{obs}$ using a distance function $\Delta(X_i, X_{obs})$ and acceptance threshold $\epsilon_t$, such that $\Delta(X_i, X_{obs}) \leq \epsilon_t$.
-% TODO replace Delta with approx likelihood (YR)
-In our case $X_i$ is 1,000 simulations of the model, and the distance function is $1-\hat{\likelihood}(X_i | X_{obs})$
+In iteration $t$ of the algorithm, we continue to sample a parameter vector $\theta$, also called \emph{particle}, and simulate observations $\tilde{\vec X}$, until $n_t$ particles are accepted. 
+A particle $\theta$ is accepted if its approximate likelihood is high enough, $\likelihood(\theta)>1-\epsilon_t$ (or more commonly, if $1-\likelihood(\theta) < \epsilon_t$), where $\epsilon_t>0$ is the \emph{acceptance threshold}, as higher values of $\epsilon_t$ allow more particles to be accepted.
 % TODO something about how the weight is computed and used?
-The acceptance threshold $\epsilon_t$ is chosen as the median of the distances of the accepted particles in the previous iteration and $\epsilon_0=0.01$. 
-When $n_t$ particles are accepted, a new proposal distribution is constructed using a multivariate normal kernel with adaptive covariance matrix. % TODO ref - Klinger? Syga? Sisson?
-In the first iteration the prior (see below) is used as a proposal distribution. 
-$n_t$ is adapted at every iteration $t$ using the \emph{adaptive population strategy}~\citep{Klinger2017}.
+The acceptance threshold $\epsilon_t$ is chosen as the median of the $1-\likelihood(\theta)$ of the particles accepted in the previous iteration, $t-1$, and $\epsilon_0=0.01$. 
+The proposed particles are sampled from a proposal distribution: for $t=0$, this is the prior distribution, and for $t>0$ the proposal distribution is constructed using a multivariate normal kernel with adaptive covariance matrix from the particles accepted in the previous iteration.
+% TODO ref - Klinger?
+Syga?
+Sisson?
+The number of particles, $n_t$, is adapted at every iteration $t$ using the \emph{adaptive population strategy}~\citep{Klinger2017}.
+
 Acceptance is determined according to the approximate likelihood, which has a maximum value of 0.875.
-Thus, we terminated the inference when $\epsilon \le 0.13$ after four iterations, with 1,074 accepted parameter vectors and effective sample size (ESS) 700 (\autoref{fig:ess-single}).
+Thus, we terminated the inference when $\epsilon \le 0.13$ after four iterations, with $n_4=1,074$ accepted parameter vectors and effective sample size (ESS) 700 (\autoref{fig:ess-single}).
 % TODO for multi-locus model we need to say something else
@@
 After producing a set of weighted particles from the the posterior distribution using the above ABC-SMC algorithm, we approximate the posterior using kernel density estimation (KDE) with Gaussian kernels, from which we find the MAP (maximum a posteriori) estimate as the maximum of the KDE function.
-We then sample $50,000$ samples from the posterior KDE to compute the HDI (highest density interval) and visualize the posterior distribution with histograms.
+We then draw $50,000$ samples from the posterior KDE to compute the HDI (highest density interval) and visualize the posterior distribution with histograms.
 
@@
 \begin{aligned}
-\mathit{WAIC}(\theta, \vec{X}) &\ =\ 
--2\log\mathbb{E}[\mathcal{L}(\theta \mid \vec{X})]\ +\ 2\mathbb{V}[\log\mathcal{L}(\theta \mid \vec{X})]
+\mathit{WAIC}(\theta) &\ =\ 
+-2\log\mathbb{E}[\likelihood(\theta)]\ +\ 2\mathbb{V}[\log\likelihood(\theta)]
 \end{aligned}
 \end{equation}
-where $\theta$ is a parameter vector, $X$ are the model or experiment results (i.e., fixation of aneuploidy by generation $t$ etc.), and $\mathbb{E}[\cdot]$ and $\mathbb{V}[\cdot]$ are the expectation and variance taken over the posterior distribution $P(\theta \mid \vec{X})$.
-WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy and a difference of 2 is a popular threshold for model comparison~\citep{Kass1995}.
+where $\theta$ is a parameter vector, and $\mathbb{E}[\cdot]$ and $\mathbb{V}[\cdot]$ are the expectation and variance taken over the posterior distribution.
+WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy~\citep{Kass1995}.
 % TODO Describe posterior plots comparison method - Fig 4 & S3
@@
 
-For the mutation rate, $\mu$, and aneuploidy rate, $\delta$, we used uninformative, uniform priors, $\mu \sim \mathit{U}(10^{-9}\ ,\ 10^{-5})$ and $\delta \sim \mathit{U}(10^{-6} , 10^{-2})$.
+For the mutation rate, $\mu$, and aneuploidy rate, $\delta$, we used uninformative uniform priors, $\mu \sim \mathit{U}(10^{-9}\ ,\ 10^{-5})$ and $\delta \sim \mathit{U}(10^{-6} , 10^{-2})$.
 For a model without aneuploidy ($\delta=0$), we used a wider uniform prior for the mutation rate, $\mu \sim \mathit{U}(10^{-10},10^{-5})$.
@@
 aneuploidy rate, $\delta=1.61\substack{+1.122 \\ -0.197}\times10^{-3}$,
-fitness of \anwt, $w_{2n+1}=1.022\substack{+0.001 \\ -0.002}$,
-fitness of \anmt, $w_{2n+1*}=1.025\substack{+0.001 \\ -0.001}$,
-and fitness of \eumt, $w_{2n*}=1.028\substack{+0.001 \\ -0.002}$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$.
+and fitness values, $w_{2n+1}=1.022\substack{+0.001 \\ -0.002}$,
+$w_{2n+1*}=1.025\substack{+0.001 \\ -0.001}$,
+$w_{2n*}=1.028\substack{+0.001 \\ -0.002}$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$.
+% describe b and c
 This estimated aneuploidy rate agrees with previous estimates. % TODO refs
@@
 \paragraph{Model checking and comparison.}
-The single-locus model fits the data well: in simulations using the MAP estimate parameters \euwt\ fixed in 59\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\autoref{fig:2n*-fixation}) and \anwt\ fixed in roughly 300 generations, which agrees with experimental results.
+The single-locus model fits the data well: in simulations using the best-fit parameters (MAP estimate) \eumt\ fixed in 59\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\autoref{fig:2n*-fixation}) and \anwt\ fixed in roughly 300 generations on average (green lines in \autoref{fig:ppc-plot}), which agrees with experimental results.
+% TODO replace "roughly 300" with the exact number.
 % TODO this fig shows fixation of 2n*, what about fixation of 2n+1?

### 2021-12-27 ms/ms.tex
moved fig from supp to text
revised results

@@
 
-However, a model without aneuploidy, i.e.
-where the aneuploidy gain rate is fixed at zero $\delta=0$, cannot explain the data~(\autoref{fig:2n*-fixation} and \autoref{fig:likelihood}).
+However, a model without aneuploidy where the aneuploidy rate is fixed at zero, $\delta=0$, cannot explain the experimental observations~(Figures~\ref{fig:2n*-fixation} and \ref{fig:likelihood}).
 The estimated parameter values without aneuploidy are:
 mutation rate, $\mu=1.251\substack{+0.146 \\ -0.177}\times10^{-8}$,
-2n+1 fitness - $1.01\substack{+0.001 \\ -0.002}$,
-2n+1* fitness - $1.011\substack{+0.001 \\ -0.001}$,
+%2n+1 fitness - $1.01\substack{+0.001 \\ -0.002}$,
+%2n+1* fitness - $1.011\substack{+0.001 \\ -0.001}$,
 %TODO why do you have fitness of aneuploidy when there is no aneuploidy?
-2n* fitness - $1.011\substack{+0.001 \\ -0.001}$. 
-We can see that mutation rate for this model is much lower than for the model with aneuploidy.
-Higher mutation rate would cause quicker fixation of \eumt\ than expected by the experiment ( \autoref{fig:2n*-fixation} blue bar).
-Thereby, adding aneuploidy to the model, helps to attain good fit with higher mutation rate.
+fitness of mutant, $w_{2n^*}=1.011\substack{+0.001 \\ -0.001}$. 
+We can see that the mutation rate estimate in this model is much lower than in the model with aneuploidy.
+Higher mutation rate would cause quicker fixation of \eumt\ than in the experimental obvservations (\autoref{fig:2n*-fixation}).
+Thus, even disregarding the appearance of trisomic cells in the experiment, the evidence supports the inclusion of aneuploidy in the model.
 % TODO WAIC?
+
+% Fig single-locus model with and without aneuploidy
+\begin{figure}[b!]
+  \centering
+  \begin{subfigure}[t]{0.45\textwidth}
+      \caption{
+        \textbf{A. With aneuploidy}
+      }
+      \centering
+      \includegraphics[height=1.8in]{../figures/likelihood-plot.pdf}      
+      \label{fig:likelihood-with-aneuploidy}
+  \end{subfigure}
+  \begin{subfigure}[t]{0.45\textwidth}
+  	  \caption{
+        \textbf{B. Without aneuploidy}
+      }
+      \centering
+      \includegraphics[height=1.8in]{../figures/likelihood-noaneuploidy-plot.pdf}      
+      \label{fig:likelihood-without-aneuploidy}
+  \end{subfigure}
+  \caption{
+  % TODO change x label - "Time to fixation of 2n*"
+  % plot both lines on one panel with two colors (blue and red for with and without? choose nice color, not the default 'r' and 'b'!), make the without in dashed.
+    \textbf{Single-locus model fit with and without aneuploidy.} Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy (i.e., $\delta=0$ is fixed).
+    \textbf{(B)} Simulations parameters are from the maximum a posterior estimation of the single-locus model \emph{without} aneuploidy fit. 
+     The MAP likelihood (\autoref{eq:heatstress-likelihood}) is 0.86 and 0.75 with and without aneuploidy. % TODO likelihood ratio test? bayes factor? no sense in just giving the likelihoods.
+  }
+  \label{fig:likelihood}
+\end{figure}
 
@@
 
-% Fig single-locus model with and without aneuploidy
-\begin{figure}[b!]
-  \centering
-  \begin{subfigure}[t]{0.5\textwidth}
-      \caption{
-        \textbf{A. With aneuploidy}
-      }
-      \centering
-      \includegraphics[height=1.8in]{../figures/likelihood-plot.pdf}      
-      \label{fig:likelihood-with-aneuploidy}
-  \end{subfigure}%
-  \\
-  \begin{subfigure}[t]{0.5\textwidth}
-  	  \caption{
-        \textbf{B. Without aneuploidy}
-      }
-      \centering
-      \includegraphics[height=1.8in]{../figures/likelihood-noaneuploidy-plot.pdf}      
-      \label{fig:likelihood-without-aneuploidy}
-  \end{subfigure}
-  \caption{
-  % TODO change x label - "Time to fixation of 2n*"
-  % plot both lines on one panel with two colors (blue and red for with and without? choose nice color, not the default 'r' and 'b'!), make the without in dashed.
-    \textbf{Single-locus model fit with and without aneuploidy.} Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy (i.e., $\delta=0$ is fixed).
-    \textbf{(B)} Simulations parameters are from the maximum a posterior estimation of the single-locus model \emph{without} aneuploidy fit. 
-     The MAP likelihood (\autoref{eq:heatstress-likelihood}) is 0.86 and 0.75 with and without aneuploidy. % TODO likelihood ratio test? bayes factor? no sense in just giving the likelihoods.
-  }
-  \label{fig:likelihood}
-\end{figure}
-
 % Fig likelihood profile

### 2021-12-29 ms/ms.tex
Added pyabc weights description. Added different tau values results. Updated seeds plot

@@
 and $b \ge c$ is the selection coefficient, or fitness benefit, of aneuploidy.
-% TODO check if b > c in code
+% TODO check if b > c in code... ANS: no...but maybe it is good to add, because anyway b<c is not interesting, and it can improve the perfomance.
 }
@@
 A particle $\theta$ is accepted if its approximate likelihood is high enough, $\likelihood(\theta)>1-\epsilon_t$ (or more commonly, if $1-\likelihood(\theta) < \epsilon_t$), where $\epsilon_t>0$ is the \emph{acceptance threshold}, as higher values of $\epsilon_t$ allow more particles to be accepted.
-% TODO something about how the weight is computed and used?
 The acceptance threshold $\epsilon_t$ is chosen as the median of the $1-\likelihood(\theta)$ of the particles accepted in the previous iteration, $t-1$, and $\epsilon_0=0.01$. 
-The proposed particles are sampled from a proposal distribution: for $t=0$, this is the prior distribution, and for $t>0$ the proposal distribution is constructed using a multivariate normal kernel with adaptive covariance matrix from the particles accepted in the previous iteration.
-% TODO ref - Klinger?
-Syga?
-Sisson?
+The proposed particles are sampled from a proposal distribution: for $t=0$, this is the prior distribution, and for $t>0$ the proposal distribution is constructed using the pertrubation kernel $K$, that is a multivariate normal kernel with adaptive covariance matrix from the particles accepted in the previous iteration ~\citep{Klinger2017}.
 The number of particles, $n_t$, is adapted at every iteration $t$ using the \emph{adaptive population strategy}~\citep{Klinger2017}.
+For each accepted particle $\theta$ the weight $w_t(\theta)$ is assigned: for $t=0$ it is 1, and for $t>0$ it is equal to the prior of $\theta$ divided by $\sum_{\theta^*}{w_{t-1}(\theta^*) K(\theta,\theta^*)}$, when $\theta^*$ is a particle accepted in the previous iteration.
+~\citep{Toni2009}.
 
@@
 % TODO say/plot something about the correlation: which fitness values are correlated? how strong is it? why does it occur?
+
+We also tried to inference parameters of the single-locus model with different $\tau$: 1, 2, 5, 10 and 100.
+It was seen that for 1, 2, 5 the posterior was not significantly different from the model with $\tau=33/32$, while for $\tau=10$ the estimated $\mu\ $ was a little lower, $1.654\substack{+0.879 \\ -1.775}\times10^{-6}$, and for $\tau=100$ the estimated $\mu\ $ was already significantly lower and $\delta\ $ had higher variability.
+The MAP and 50\% HDI for $\tau=100$:
+$\mu=5.538\substack{+1.611 \\ -5.311}\times10^{-7}$,
+$\delta=8.292\substack{+8.988 \\ -4.96}\times10^{-4}$,
+$w_{2n+1}=1.023\substack{+0.001 \\ -0.002}$,
+$w_{2n+1*}=1.024\substack{+0.001 \\ -0.001}$,
+$w_{2n*}=1.028\substack{+0.001 \\ -0.002}$.
+%TODO: show joint distributions comparison plot (20-diffk.ipynb). WAIC?
 

### 2021-12-29 ms/ms.tex
revised ABC-SMC description

--- 2021-12-27 ms/ms.tex
@@
 \begin{tabular}{lllll}
-\emph{Genotype} $i$ & $2n$ & $2n+1$ & $2n+1^{*k}$ & $2n^{*k}$ \\
+\emph{Genotype}, $g$ & $2n$ & $2n+1$ & $2n+1^{*k}$ & $2n^{*k}$ \\
 \hline
-\emph{Fitness} $w_i$ & $1$ & $1-c+b$ & $(1-c)\prod_{j=1}^k(1+s_i)+b$ & $\prod_{j=1}^k(1+s_i)$               
+\emph{Fitness}, $w_g$ & $1$ & $1-c+b$ & $(1-c)\prod_{j=1}^k(1+s_i)+b$ & $\prod_{j=1}^k(1+s_i)$               
 \end{tabular}
@@
 and $b \ge c$ is the selection coefficient, or fitness benefit, of aneuploidy.
-% TODO check if b > c in code
 }
@@
 The experiment continued with two populations, in which aneuploidy was eliminated by generation 1,700 and 2,350.
-%Thus, the empirical evidence can be summarized by $\vec{X}=\Big(200 \le T_i(2n+1) \le 450,\, i=1,2,3,4;\ T_1(2n^*)=1700;\ T_2(2n^*)=2350\Big)$ where $T_i(g)$ is the fixation time of genotype $g$ in population $i$.
 
 \paragraph{Likelihood function.} 
-
+% TODO please read!
 Because our model, just like the Wright-Fisher model, is non-linear and stochastic, computing the distribution of fixation time $T(g)$ of genotype $g$ for use in the likelihood function is intractable (it is even hard to use a diffusion-equation approximation due to the model having multiple genotypes, rather than just two).
@@
 
-%Denote by $A_{t,T}$ the fixation of aneuploidy by generation $t$ conditioned on no fixation up to generation $T$; by $B_{t}$ the extinction of aneuploidy by generation $t$; and by $C_{t,T}$ the extinction of aneuploidy by generation $t$ conditioned on no extinction by generation $T$.
-%The likelihood function for parameter vector $\theta$ conditioned on the observed data $\vec{X}$ from the heat-stress experiment is
-%\begin{multline} \label{eq:heatstress-likelihood}
-%P^4(A_{450,200}) \cdot \Big(1 - P^4(\lnot{B_{1700}} | A_{450,200} ) - P^4(\lnot{C_{2350, 1700}}| A_{450,200}) + P^4(\lnot{B_{1700}} \land \lnot{C_{2350, 1700}}| A_{450,200})\Big)	\;,	 % TODO fix L* (YR)
-%\end{multline}
-%where $P(X)$ is the probability of event $X$ and $\lnot$ means \emph{not}.
-
 For a model without aneuploidy (that is, when the aneuploidy rate is fixed at zero, $\delta=0$), 
-%denote by $M_{t}$ the fixation of \eumt\ by generation $t$, and by $Q_{t,T}$ the fixation of \eumt\ by generation $t$ conditioned on no fixation at generation $T$.
 the likelihood is similarly approximated by
-% of a model without aneuploidy with parameter vector $\theta$ for the heat-stress experiment is
 \begin{equation}\begin{aligned}
@@
 
-%\begin{multline} \label{eq:heatstress-noaneuploidy-likelihood}
-%\likelihood_{!}(\theta) = 
-%\Big(1 - P^4(\lnot{M_{1700}}) - P^4(\lnot{Q_{2350, 1700}}) + P^4(\lnot{M_{1700}} \land \lnot{Q_{2350, 1700}})\Big)	\;,	
-%\end{multline}
-
 % TODO if we show results of multi-locus model, write how much replicates were used.
@@
 It is a method of likelihood-free, simulation-based inference~\citep{Cranmer2020}, that is, for estimating a posterior distribution when a likelihood function cannot be directly computed.
-It is therefore suitable to our case, in which the likelihood function can only be approximated from simulations, and cannot be directly computed.
+It is therefore suitable in our case, in which the likelihood function can only be approximated from simulations, and cannot be directly computed.
 
-Briefly, the ABS-SMC algorithm employs sequential importance sampling over multiple iterations~\citep{Syga2021}. 
-In iteration $t$ of the algorithm, we continue to sample a parameter vector $\theta$, also called \emph{particle}, and simulate observations $\tilde{\vec X}$, until $n_t$ particles are accepted. 
-A particle $\theta$ is accepted if its approximate likelihood is high enough, $\likelihood(\theta)>1-\epsilon_t$ (or more commonly, if $1-\likelihood(\theta) < \epsilon_t$), where $\epsilon_t>0$ is the \emph{acceptance threshold}, as higher values of $\epsilon_t$ allow more particles to be accepted.
-% TODO something about how the weight is computed and used?
+%% YR revise
+The ABS-SMC algorithm employs sequential importance sampling over multiple iterations~\citep{Toni2009, Klinger2017, Syga2021}.
+In iteration $t$ of the algorithm, a set of parameter vectors, $\{\theta_{i,t}\}_{i=1}^{n_t}$, also called \emph{particles}, are constructed in the following way.
+A proposal particle, $\theta^*$, is sampled from a proposal distribution, and is either accepted or rejected, until $n_t$ particles are accepted.
+The number of particles, $n_t$, is adapted at every iteration $t$ using the adaptive population strategy.
+For $t=0$, the proposal particle is sampled from the prior distribution, $p(\theta)$.
+For $t>0$, the proposal particle is sampled from the particles accepted in the previous iteration, $\{\theta_{i,t-1}\}_{i=1}^{n_{t-1}}$, each with a probability relative to its weight $w_{t-1}(\theta_{i,t-1})$ (see below).
+The proposal particle is then perturbed using a kernel perturbation kernel, $K_t(\theta^* \mid \theta)$ where $\theta$ is the sample from the previous iteration.
+Then, a set of synthetic observations $\tilde{\vec X}^*$ is simulated, and the proposal particle $\theta^*$ is accepted if its approximate likelihood (\autoref{eq:heatstress-likelihood}) is high enough, $\likelihood(\theta^*)>1-\epsilon_t$ (or more commonly, if $1-\likelihood(\theta^*) < \epsilon_t$), where $\epsilon_t>0$ is the \emph{acceptance threshold}, as higher values of $\epsilon_t$ allow more particles to be accepted.
 The acceptance threshold $\epsilon_t$ is chosen as the median of the $1-\likelihood(\theta)$ of the particles accepted in the previous iteration, $t-1$, and $\epsilon_0=0.01$. 
-The proposed particles are sampled from a proposal distribution: for $t=0$, this is the prior distribution, and for $t>0$ the proposal distribution is constructed using a multivariate normal kernel with adaptive covariance matrix from the particles accepted in the previous iteration.
-% TODO ref - Klinger?
-Syga?
-Sisson?
-The number of particles, $n_t$, is adapted at every iteration $t$ using the \emph{adaptive population strategy}~\citep{Klinger2017}.
+For each accepted particle $\theta_{i,t}$ a weight $w_t(\theta_{i,t})$ is assigned: for $t=0$, $w_0(\theta_{i,0})=1$, and for $t>0$, 
+$w_t(\theta_{i,t}) = p(\theta_{i,t}) / \sum_{i=1}^{n_{t-1}}{w_{t-1}(\theta_{i,t-1}) K_t(\theta_{i,t}, \theta_{i,t-1})}$, where $p(\theta)$ is the prior density of $\theta$ and $K_t(\theta', \theta)$ is the probability of a perturbation from $\theta$ to $\theta'$.
+$K_t(\theta' \mid \theta)$ is a multivariate normal distribution. % TODO construction of K_T from \theta_{i,t-1}
 
@@
 We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$. In lieu of a better prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$.
-%TODO please report here the point estimate and 95% CI for the three distributions, I only have for 1-c+b: 0.9863 (95% CI 0.9790 - 0.9939).
+
 Compared to other priors we tested, this prior produced lower WAIC, better posterior prediction plots, and more stable parameter estimates, as follows.
-We tested a uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ and (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$.
-In both of these uninformative priors, the fitness estimates were much too high: for the first prior the median estimates were  $w_{2n+1}=2.767$, $w_{2n+1^*}=5.79$, $w_{2n^*}=5.827$ and for the second prior the estimates were $w_{2n+1}=1.046$, $w_{2n+1^*}=4.642$, $w_{2n^*}=4.684$.
+%In both of these uninformative priors, the fitness estimates were much too high: for the first prior the median estimates were  $w_{2n+1}=2.767$, $w_{2n+1^*}=5.79$, $w_{2n^*}=5.827$ and for the second prior the estimates were $w_{2n+1}=1.046$, $w_{2n+1^*}=4.642$, $w_{2n^*}=4.684$.
 Such high fitness values are unreasonable.
@@
 The same prior was used for $w_{2n*}/w_{2n+1*}$.
-This prior resulted in WAIC 2.8, compared to 0.27 with the above informative prior.
+This prior resulted in WAIC 0.32, compared to 0.27 with the above informative prior.
+% TODO what are the results ?
+
 We also tried to use growth curves of \anwt\; and \euwt\; in \SI{30}{\celsius} to estimate $1-c$.
@@
 The prior for $b$ was taken the same as for $c$.
-This prior did not work at all, as no parameter sets were found with an approximate likelihood greater than zero.
+This prior did not converge: no parameter sets were found with an approximate likelihood greater than zero.
 % TODO it make sense because bla bla (to explain)
+Finally, we also tested a uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ and (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$.
+In both cases the algorithm failed to converge.
 
@@
 But I checked also that when I don't adding this constrain the result was the same.
-However I dind't add reverse direction in this case.
+However I didn't add reverse direction in this case.
 
@@
 and fitness values, $w_{2n+1}=1.022\substack{+0.001 \\ -0.002}$,
-$w_{2n+1*}=1.025\substack{+0.001 \\ -0.001}$,
-$w_{2n*}=1.028\substack{+0.001 \\ -0.002}$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$.
+$w_{2n+1^*}=1.025\substack{+0.001 \\ -0.001}$,
+$w_{2n^*}=1.028\substack{+0.001 \\ -0.002}$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$.
 % describe b and c
@@
 % TODO say/plot something about the correlation: which fitness values are correlated? how strong is it? why does it occur?
+
+We also tried to inference parameters of the single-locus model with different $\tau$: 1, 2, 5, 10 and 100.
+It was seen that for 1, 2, 5 the posterior was not significantly different from the model with $\tau=33/32$, while for $\tau=10$ the estimated $\mu\ $ was a little lower, $1.654\substack{+0.879 \\ -1.775}\times10^{-6}$, and for $\tau=100$ the estimated $\mu\ $ was already significantly lower and $\delta\ $ had higher variability.
+The MAP and 50\% HDI for $\tau=100$:
+$\mu=5.538\substack{+1.611 \\ -5.311}\times10^{-7}$,
+$\delta=8.292\substack{+8.988 \\ -4.96}\times10^{-4}$,
+$w_{2n+1}=1.023\substack{+0.001 \\ -0.002}$,
+$w_{2n+1^*}=1.024\substack{+0.001 \\ -0.001}$,
+$w_{2n^*}=1.028\substack{+0.001 \\ -0.002}$.
+%TODO: show joint distributions comparison plot (20-diffk.ipynb). WAIC?
+
+% TODO use tau=1 as the default; we also check tau>1 - 33/32, 2, 5, 10, 100.
+% TODO support figure of joint posterior delta vs mu per tau=1, 33/32, 2, 5, 10
 
@@
 % TODO replace "roughly 300" with the exact number.
-% TODO this fig shows fixation of 2n*, what about fixation of 2n+1?
-% TODO merge fig 3 and fig:likelihood (both panels of fig:likelihood) 
 Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\autoref{fig:ppc-plot}). % TODO move somewhere else?
@@
 mutation rate, $\mu=1.251\substack{+0.146 \\ -0.177}\times10^{-8}$,
-%2n+1 fitness - $1.01\substack{+0.001 \\ -0.002}$,
-%2n+1* fitness - $1.011\substack{+0.001 \\ -0.001}$,
-%TODO why do you have fitness of aneuploidy when there is no aneuploidy?
 fitness of mutant, $w_{2n^*}=1.011\substack{+0.001 \\ -0.001}$. 
 We can see that the mutation rate estimate in this model is much lower than in the model with aneuploidy.
-Higher mutation rate would cause quicker fixation of \eumt\ than in the experimental obvservations (\autoref{fig:2n*-fixation}).
+Higher mutation rate would cause quicker fixation of \eumt\ than in the experimental observations (\autoref{fig:2n*-fixation}).
 Thus, even disregarding the appearance of trisomic cells in the experiment, the evidence supports the inclusion of aneuploidy in the model.
@@
   \end{subfigure}
+
   \begin{subfigure}[t]{0.45\textwidth}
@@
   \end{subfigure}
+
   \caption{
@@
   % plot both lines on one panel with two colors (blue and red for with and without? choose nice color, not the default 'r' and 'b'!), make the without in dashed.
+  % place 82% a bit higher
+  % xticks - replace 500, 1000,1500,2000,2500 with 450, 1700, 2350
     \textbf{Single-locus model fit with and without aneuploidy.} Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy (i.e., $\delta=0$ is fixed).
-    \textbf{(B)} Simulations parameters are from the maximum a posterior estimation of the single-locus model \emph{without} aneuploidy fit. 
-     The MAP likelihood (\autoref{eq:heatstress-likelihood}) is 0.86 and 0.75 with and without aneuploidy. % TODO likelihood ratio test? bayes factor? no sense in just giving the likelihoods.
+     The MAP likelihood (\autoref{eq:heatstress-likelihood}) is 0.86 and 0.75 with and without aneuploidy.
 	}
@@
 
-\begin{figure}[h!]
-  \centering
-  % TODO add legend
-\includegraphics[width=0.6\textwidth]{../figures/ppc-plot.pdf}
-  \caption{
-  \textbf{Single-locus model posterior genotype frequency dynamics.} 
-  The posterior predicted frequency of the four genotypes. Each line is the average of 10,000 simulations using parameters drawn from the posterior distribution (\autoref{fig:posterior}).
-  }
-  \label{fig:ppc-plot}
-\end{figure}
+% TODO merge previous and next figure to one figure with 2 panels
 
@@
   \textbf{Adaptation time cannot be explained without aneuploidy.}
-In blue, the distribution of adaptation time (fixation of \eumt) in $10,000$ simulations of the single-locus model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: in orange, a model with the same parameter values except $\delta=0$; in green, a model fitted to the data assuming $\delta=0$.
-%, where parameters are taken from the median of the posterior of the inference. % why madian and not MAP?
+The distribution of adaptation time (fixation of \eumt) in $10,000$ simulations of the single-locus model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange); and a model fitted to the data assuming $\delta=0$ (green).
 In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and the other by generation 2,350 (dashed lines) but not before generation 200.
 Thus, the blue distribution demonstrates a better fit compared to the green, and the yellow demonstrates a very poor fit.
-The last bin contains all the simulations with time equal or greater than 3000.
+The last bin contains all the simulations with time equal or greater than 3,000.
 }
@@
   \label{fig:2n*-fixation}
+\end{figure}
+
+\begin{figure}[h!]
+  \centering
+\includegraphics[width=0.6\textwidth]{../figures/ppc-plot.pdf}
+  \caption{
+  \textbf{Single-locus model posterior genotype frequency dynamics.} 
+  The posterior predicted frequency of the four genotypes. Each line is the average of 10,000 simulations using parameters drawn from the posterior distribution (\autoref{fig:posterior}).
+  }
+  \label{fig:ppc-plot}
+  % TODO add legend
+  % xticks - replace 500, 1000,1500,2000,2500 with 450, 1700, 2350
 \end{figure}
@@
 \end{figure}
-% TODO add caption, combine to one figure with the previous one
+% TODO add caption
+
+%TODO combine to one figure with the previous one and with refined
+%TODO please report here the point estimate and 95% CI for the three distributions, I only have for 1-c+b: 0.9863 (95% CI 0.9790 - 0.9939).
 
@@
   \end{subfigure}
+  \\
    \begin{subfigure}{0.3\textwidth}
@@
   % show the log-likelihood rather than the likelihood
-  % can we add more points? can we do this on joint likelihood of two parameters?
-  % have the fitness parameters on the same row
+  % can we add more points? 
   % x labels start with capital letter, should include the parameter (Mutation rate, $\mu$) and the name (Fitness of 2n+1, w_{2n+1}), dont use w(xxx) but rather w_{xxx} as in the main text
@@
 \end{figure}
+% TODO show posteriors of different iterations (with panel labels!!)
 
-% Fig 
+% Fig Validation of posteriors
 \begin{figure}[h]
@@
 % TODO put legend in panel A top right corner
+% panel labels
 % y label "Density"; y ticks are meaningless, right? 
 % x labels - see comments above.
+% color by # of reps; second run with 1000 reps same color as 1st but with dashed line
 

### 2022-01-05 ms/ms.tex
added heading for analysis results
added convergence figures; added convergence.ipynb

@@
 }
-
 \usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
@@
 \paragraph{Likelihood function.} 
-% TODO please read!
 Because our model, just like the Wright-Fisher model, is non-linear and stochastic, computing the distribution of fixation time $T(g)$ of genotype $g$ for use in the likelihood function is intractable (it is even hard to use a diffusion-equation approximation due to the model having multiple genotypes, rather than just two).
@@
 Acceptance is determined according to the approximate likelihood, which has a maximum value of 0.875.
-Thus, we terminated the inference when $\epsilon \le 0.13$ after four iterations, with $n_4=1,074$ accepted parameter vectors and effective sample size (ESS) 700 (\autoref{fig:ess-single}).
+Thus, we terminated the inference when $\epsilon \le 0.13$ after six iterations, with $n_6=982$ accepted parameter vectors and effective sample size (ESS) 651 (\autoref{fig:convergence}).
 % TODO for multi-locus model we need to say something else
@@
 
-% Fig ABC convergence
+% Fig ABC iterations
 \begin{figure}[h]
+
+  \begin{subfigure}{1\textwidth}
     \centering
+      \includegraphics[width=0.45\textwidth]{../figures/convergence-p1_mr.pdf}      
+      \includegraphics[width=0.45\textwidth]{../figures/convergence-p2_tr.pdf}      
+      \includegraphics[width=0.45\textwidth]{../figures/convergence-p3_w1.pdf}      
+      \includegraphics[width=0.45\textwidth]{../figures/convergence-p4_w2.pdf}      
+      \includegraphics[width=0.45\textwidth]{../figures/convergence-p5_w3.pdf}      
+     \label{fig:convergence-A}
+    \caption{\textbf{A. Posterior densities per iteration}}
+    \end{subfigure}
+    \\
+    \vspace*{4mm}
+
+  \begin{subfigure}{1\textwidth}
   	\includegraphics[width=1\textwidth]{../figures/ess.pdf}
+     \label{fig:convergence-B}
+     \vspace*{-8mm}
+    \caption{\textbf{B. ABC-SMC population properties per iteration}}
+  \end{subfigure}
+
     \caption{
-    \textbf{Single-locus model inference convergence.} Measures of convergence for the ABC-SMC algorithm with the single-locus model.
-After iteration number 4, the acceptance threshold was $\epsilon=0.13$, the acceptance rate was XXX, the number of accepted parameter sets ("particles") was 1,074, and the effective sample size (ESS) 700.
-% TODO add missing acceptance rate
+    \textbf{Single-locus model inference convergence.} 
+    ABC-SMC algorithm was executed to inference single-locus model parameters. \textbf{(A)} For each parameter of the model, there's shown KDE of the posterior density of the inference per iteration.
+We can see that the posterior didn't change significantly already from the first iteration.
+\textbf{(B)} Here are shown measures of convergence for the ABC-SMC algorithm.
+After iteration number 6, the acceptance threshold was $\epsilon=0.13$, the acceptance rate was $0.018$, the number of particles was 982, and the effective sample size (ESS) 651.
 }
-    \label{fig:ess-single}
-    % TODO y label in first panel "Acceptance threshold, $\epsilon$"
-    % add panel labels A B C D
-    % x label - just once
-	% should we do 1-3 more iterations? it is not clear that it converged.
+    \label{fig:convergence}
 \end{figure}
-% TODO show posteriors of different iterations (with panel labels!!)
 

### 2022-01-09 ms/ms.tex
updated 2n* fixation plot; added notebooks; updated estimated MAPs, because default tau=1

@@
 We used ABC-SMC to infer the posterior distribution of the parameters of the single-locus model.
-The MAP (maximum a posteriori) and 50\% HDI (highest density interval) of the parameters posterior are~(\autoref{fig:posterior}):
-mutation rate, $\mu=3.06\substack{+1.229 \\ -2.84}\times10^{-6}$,
-aneuploidy rate, $\delta=1.61\substack{+1.122 \\ -0.197}\times10^{-3}$,
+The MAP (maximum a posteriori) and 50\% HDI (highest density interval) of the parameters posterior are~(\autoref{fig:posterior}): %% generated by MAP-and-hdi.ipynb
+mutation rate, $\mu=2.93\substack{+1.122 \\ -2.824}\times10^{-6}$,
+aneuploidy rate, $\delta=1.716\substack{+0.974 \\ -0.384}\times10^{-3}$,
 and fitness values, $w_{2n+1}=1.022\substack{+0.001 \\ -0.002}$,
 $w_{2n+1^*}=1.025\substack{+0.001 \\ -0.001}$,
-$w_{2n^*}=1.028\substack{+0.001 \\ -0.002}$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$.
-% describe b and c
+$w_{2n^*}=1.028\substack{+0.001 \\ -0.001}$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$.
+% describe b and c TODO? what does it mean?
 This estimated aneuploidy rate agrees with previous estimates. % TODO refs
@@
 It was seen that for 1, 2, 5 the posterior was not significantly different from the model with $\tau=33/32$, while for $\tau=10$ the estimated $\mu\ $ was a little lower, $1.654\substack{+0.879 \\ -1.775}\times10^{-6}$, and for $\tau=100$ the estimated $\mu\ $ was already significantly lower and $\delta\ $ had higher variability.
-The MAP and 50\% HDI for $\tau=100$:
+The MAP and 50\% HDI for $\tau=100$: %TODO add to MAP-and-hdi.ipynb and check the values here
 $\mu=5.538\substack{+1.611 \\ -5.311}\times10^{-7}$,
@@
 However, a model without aneuploidy where the aneuploidy rate is fixed at zero, $\delta=0$, cannot explain the experimental observations~(Figures~\ref{fig:2n*-fixation} and \ref{fig:likelihood}).
-The estimated parameter values without aneuploidy are:
-mutation rate, $\mu=1.251\substack{+0.146 \\ -0.177}\times10^{-8}$,
-fitness of mutant, $w_{2n^*}=1.011\substack{+0.001 \\ -0.001}$. 
+The estimated parameter values without aneuploidy are: %% generated by MAP-and-hdi.ipynb
+mutation rate, $\mu=7.837\substack{+0.078 \\ -0.123}\times10^{-9}$,
+fitness of mutant, $w_{2n^*}=1.013\substack{+0.000 \\ -0.000}$. 
 We can see that the mutation rate estimate in this model is much lower than in the model with aneuploidy.
@@
     \textbf{Single-locus model fit with and without aneuploidy.} Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy (i.e., $\delta=0$ is fixed).
-     The MAP likelihood (\autoref{eq:heatstress-likelihood}) is 0.86 and 0.75 with and without aneuploidy.
+     The MAP likelihood (\autoref{eq:heatstress-likelihood}) is 0.84 and 0.67 with and without aneuploidy.
 	}
@@
 
+%%% generated by with-aneuploidy_vs_no_aneuploidy.ipynb
 \begin{figure}[h!]
   \centering
-\includegraphics[width=0.6\textwidth]{../figures/fixation-plot.pdf}
+\includegraphics[width=0.9\textwidth]{../figures/fixation-plot.pdf}
   \caption{
@@
 }
-% TODO add legend
-% TODO change y label to % simulations
-% change x label to Time to fixation of 2n*
-% makes vertical line black
-% add line for t=200?
-% does x axis start at zero? can you add 0 to x ticks?
   \label{fig:2n*-fixation}
@@
 
-% Fig ABC iterations
+% Fig ABC iterations, convergence
+%% generated by convergence.ipynb
 \begin{figure}[h]

### 2022-01-10 ms/ms.tex
merged and updated 2n*fixation plots

@@
 \paragraph{Model checking and comparison.}
-The single-locus model fits the data well: in simulations using the best-fit parameters (MAP estimate) \eumt\ fixed in 59\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\autoref{fig:2n*-fixation}) and \anwt\ fixed in roughly 300 generations on average (green lines in \autoref{fig:ppc-plot}), which agrees with experimental results.
+The single-locus model fits the data well: in simulations using the best-fit parameters (MAP estimate) \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\autoref{fig:2n*-fixation}-B) and \anwt\ fixed in roughly 300 generations on average (green lines in \autoref{fig:ppc-plot}), which agrees with experimental results.
 % TODO replace "roughly 300" with the exact number.
@@
 
-However, a model without aneuploidy where the aneuploidy rate is fixed at zero, $\delta=0$, cannot explain the experimental observations~(Figures~\ref{fig:2n*-fixation} and \ref{fig:likelihood}).
+However, a model without aneuploidy where the aneuploidy rate is fixed at zero, $\delta=0$, cannot explain the experimental observations~(Figure~\ref{fig:2n*-fixation}).
 The estimated parameter values without aneuploidy are: %% generated by MAP-and-hdi.ipynb
@@
 % Fig single-locus model with and without aneuploidy
+%%% generated by with-aneuploidy_vs_no_aneuploidy.ipynb
+
 \begin{figure}[b!]
+
+  \begin{subfigure}[t]{0.9\textwidth}
       \centering
-  \begin{subfigure}[t]{0.45\textwidth}
-      \caption{
-        \textbf{A. With aneuploidy}
-      }
-      \centering
-      \includegraphics[height=1.8in]{../figures/likelihood-plot.pdf}      
-      \label{fig:likelihood-with-aneuploidy}
+      \includegraphics[height=1.8in]{../figures/fixation-plot-a.pdf}      
+      \label{fig:fit}
+    \caption{\textbf{A. Distribution.}}
   \end{subfigure}
 
-  \begin{subfigure}[t]{0.45\textwidth}
-  	  \caption{
-        \textbf{B. Without aneuploidy}
-      }
+  \vspace*{4mm}
+
+  \begin{subfigure}[t]{0.9\textwidth}
       \centering
-      \includegraphics[height=1.8in]{../figures/likelihood-noaneuploidy-plot.pdf}      
-      \label{fig:likelihood-without-aneuploidy}
+      \includegraphics[height=1.8in]{../figures/fixation-plot-b.pdf}      
+      \label{fig:fit-cumulative}
+    \caption{\textbf{B. Cumulative distribution.}}
   \end{subfigure}
@@
   \caption{
-  % TODO change x label - "Time to fixation of 2n*"
-  % plot both lines on one panel with two colors (blue and red for with and without? choose nice color, not the default 'r' and 'b'!), make the without in dashed.
-  % place 82% a bit higher
-  % xticks - replace 500, 1000,1500,2000,2500 with 450, 1700, 2350
-    \textbf{Single-locus model fit with and without aneuploidy.} Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy (i.e., $\delta=0$ is fixed).
-     The MAP likelihood (\autoref{eq:heatstress-likelihood}) is 0.84 and 0.67 with and without aneuploidy.
-	}
-  \label{fig:likelihood}
-\end{figure}
-
-% TODO merge previous and next figure to one figure with 2 panels
-
-%%% generated by with-aneuploidy_vs_no_aneuploidy.ipynb
-\begin{figure}[h!]
-  \centering
-\includegraphics[width=0.9\textwidth]{../figures/fixation-plot.pdf}
-  \caption{
-  \textbf{Adaptation time cannot be explained without aneuploidy.}
-The distribution of adaptation time (fixation of \eumt) in $10,000$ simulations of the single-locus model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange); and a model fitted to the data assuming $\delta=0$ (green).
+    \textbf{Single-locus model fit with and without aneuploidy.}
+    \textbf{(A)} The distribution of adaptation time (fixation of \eumt) in $10,000$ simulations of the single-locus model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange); and a model fitted to the data assuming $\delta=0$ (green).
 In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and the other by generation 2,350 (dashed lines) but not before generation 200.
@@
 The last bin contains all the simulations with time equal or greater than 3,000.
+    \textbf{(B)}
+	Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy (i.e., $\delta=0$ is fixed).
+     The MAP likelihood (\autoref{eq:heatstress-likelihood}) is 0.84 and 0.67 with and without aneuploidy.
 	}
@@
      \label{fig:convergence-A}
-    \caption{\textbf{A. Posterior densities per iteration}}
+    \caption{\textbf{A. Posterior densities per iteration.}}
     \end{subfigure}
@@
      \vspace*{-8mm}
-    \caption{\textbf{B. ABC-SMC population properties per iteration}}
+    \caption{\textbf{B. ABC-SMC population properties per iteration.}}
   \end{subfigure}

### 2022-01-12 ms/diff.sh
run this to create diff.pdf between ms.tex and ms_prev.tex

@@ -1,0 +1,4 @@
+latexdiff ms_prev.tex ms.tex --no-label --enforce-auto-mbox --type=CFONT > diff.tex
+pdflatex diff.tex
+pdflatex diff.tex
+open diff.pdf

### 2022-01-12 ms/ms.tex
remove tau from model, just mention in results

@@
     \textbf{(B)} In the multi-locus model, each genotype is characterized by its karyotype, \euwt\; or \anwt, and the number of accumulated beneficial mutations, denoted by stars. 
-    In both panels arrows denote transitions between genotypes, with transitions rates: $\mu$ and $\tau \mu$, beneficial mutation rates in euploid and aneuploid cells; $\delta$, aneuploidy rate.
-    % TODO IK: change k mu to tau mu in figure; also k mu is too close to the arrow right now; change alpha to delta in the figure 
+    In both panels arrows denote transitions between genotypes, with transitions rates: $\mu$, beneficial mutation rate; $\delta$, aneuploidy rate.
+    % TODO IK: remove k (not tau either!); also k mu is too close to the arrow right now; change alpha to delta in the figure 
   }
@@
 
-Transitions between the genotypes occur as follows~(\autoref{fig:models}A): Beneficial mutations from \euwt\; to \eumt\; occur with probability $\mu$, the mutation rate, and from \anwt\; to \anmt\; with probability $\tau \mu$, where $\tau$ is the fold-change in the rate of beneficial mutations in aneuploid cells.
-By default, we assume $\tau=33/32$, as the wild-type \yeast strains in the experiments by \citep{Yona2012} are diploid with 32 chromosomes, and the aneuploid strains are trisomic, with 33 chromosomes.
+Transitions between the genotypes occur as follows~(\autoref{fig:models}A): Beneficial mutations from \euwt\; to \eumt\; and from \anwt\; to \anmt\; occur with probability $\mu$, the mutation rate.
 We neglect back-mutations (i.e., from \eumt\; to \euwt\; and from \anmt\; to \anwt).
@@
       &f^m_{2n} &=&\; (1 - \delta - \mu) f^s_{2n}  \;,\\
-      &f^m_{2n+1} &=&\; \delta f^s_{2n} + (1 - \tau \mu) f^s_{2n+1}  \;,\\
-      &f^m_{2n+1^*} &=&\; \tau \mu f^s_{2n+1} + (1-\delta) f^s_{2n+1^*}  \;,\\
+      &f^m_{2n+1} &=&\; \delta f^s_{2n} + (1 - \mu) f^s_{2n+1}  \;,\\
+      &f^m_{2n+1^*} &=&\; \mu f^s_{2n+1} + (1-\delta) f^s_{2n+1^*}  \;,\\
       &f^m_{2n^*} &=&\; \mu f^s_{2n} + \delta f^s_{2n+1} + f^s_{2n^*}  \;.
@@
 The selection coefficient of the $i$-th accumulated mutation in each individual, $s_i$, is drawn from an exponential distribution with expected value $s$: $s_i \sim \mathit{Exp{(s)}}$.
-The rest of the parameters ($N$, $\mu$, $\tau$, $\delta$, $b$, $c$) are the same as in the single-locus model.
+The rest of the parameters ($N$, $\mu$, $\delta$, $b$, $c$) are the same as in the single-locus model.
 However, since the multi-locus model allows several mutations with smaller fitness effects to accumulate, we expect the mutation rate to be higher compared to the single-locus model, which focuses on a single, large-effect mutation.
-% TODO : add tau to model fixed at 33/32.
 
@@
 Each offspring cell inherits these properties from its mother cell.
-Then, with probability $\mu$ or $\tau \mu$ for euploid and aneuploid cells,
-respectively, a new beneficial mutation is accumulated, such that the number of mutations is $k+1$, and its effect $s_{k+1}$ is drawn from an exponential distribution with expected value $s$, such that the contribution of the mutations to the fitness is $\prod_{j=0}^{k+1}{(1+s_j)}$.
+Then, a new beneficial mutation is accumulated with probability $\mu$, such that the number of mutations is $k+1$, and its effect $s_{k+1}$ is drawn from an exponential distribution with expected value $s$, such that the contribution of the mutations to the fitness is $\prod_{j=0}^{k+1}{(1+s_j)}$.
 Next, euploid offspring become aneuploid with probability $\delta$, and aneuploid offspring become euploid with probability $\delta_L$.
@@
 We overcome this problem by approximating the likelihood using simulations.
-We simulate 1,000 experiments per parameter vector $\theta = (\mu, \tau, \delta, s, b, c)$, resulting in a set of simulated observations $\tilde{\vec X} = \{\tilde{X}_i\}_{i=1}^{1000}$.
+We simulate 1,000 experiments per parameter vector $\theta = (\mu, \delta, s, b, c)$, resulting in a set of simulated observations $\tilde{\vec X} = \{\tilde{X}_i\}_{i=1}^{1000}$.
 We then compute the approximate likelihood,
@@
 
-We also tried to inference parameters of the single-locus model with different $\tau$: 1, 2, 5, 10 and 100.
-It was seen that for 1, 2, 5 the posterior was not significantly different from the model with $\tau=33/32$, while for $\tau=10$ the estimated $\mu\ $ was a little lower, $1.654\substack{+0.879 \\ -1.775}\times10^{-6}$, and for $\tau=100$ the estimated $\mu\ $ was already significantly lower and $\delta\ $ had higher variability.
-The MAP and 50\% HDI for $\tau=100$: %TODO add to MAP-and-hdi.ipynb and check the values here
+We also tried to infer parameters of the single-locus model when the mutation rate in aneuploid cells, compared to euploid cells, is increased by a factor $\tau$=1, 33/32 (due to an additional chromosome), 2, 5, 10 or 100 (due to genetic instability).
+We found that for $\tau=$1, 33/32, 2, and 5 the posterior was not significantly different from the model with $\tau=1$, while for $\tau=10$ the estimated mutation rate was slightly lower, $\mu=1.654\substack{+0.879 \\ -1.775}\times10^{-6}$, and for $\tau=100$ the estimated mutation rate was significantly lower and the aneuploidy rate had higher variability.
+The MAP and 50\% HDI with $\tau=100$ were: %
+TODO add to MAP-and-hdi.ipynb and check the values here
 $\mu=5.538\substack{+1.611 \\ -5.311}\times10^{-7}$,
@@
 $w_{2n^*}=1.028\substack{+0.001 \\ -0.002}$.
+We thus continued our analysis under the assumption that $\tau=1$, that is, that the mutation rate in euploid and aneuploid cells is the same.
 %TODO: show joint distributions comparison plot (20-diffk.ipynb). WAIC?
-
-% TODO use tau=1 as the default; we also check tau>1 - 33/32, 2, 5, 10, 100.
 % TODO support figure of joint posterior delta vs mu per tau=1, 33/32, 2, 5, 10
@@
 Minerva Stiftung Center for Lab Evolution (YR).
+% TODO add Minerva for Martin; funding for Avihu?
 }

### 2022-01-12 ms/ms.tex
revised parameter inference section (w->W as w is fitness, ref, phrasing)

@@
 For example, $P_{\tilde{\vec X}}\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)$ is approximated by taking simulations in which \anwt\; fixed before generation 450 but not before generation 200, and computing the fraction of such simulations in which \eumt\; did not fix by generation 1,700, and hence aneuploidy did not extinct before generation 1,700.
+\autoref{fig:seeds} compares results with less and more simulated experiments, demonstrating that 1,000 simulations are likely enough.
 
@@
 % TODO if we show results of multi-locus model, write how much replicates were used.
-% TODO ref to place where we show 1,000 is enough
 
@@
 
-%% YR revise
-The ABS-SMC algorithm employs sequential importance sampling over multiple iterations~\citep{Toni2009, Klinger2017, Syga2021}.
+The ABC-SMC algorithm employs sequential importance sampling over multiple iterations~\citep{Toni2009, Klinger2017, Syga2021}.
 In iteration $t$ of the algorithm, a set of parameter vectors, $\{\theta_{i,t}\}_{i=1}^{n_t}$, also called \emph{particles}, are constructed in the following way.
 A proposal particle, $\theta^*$, is sampled from a proposal distribution, and is either accepted or rejected, until $n_t$ particles are accepted.
-The number of particles, $n_t$, is adapted at every iteration $t$ using the adaptive population strategy.
+The number of particles, $n_t$, is adapted at every iteration $t$ using the adaptive population strategy~\citep[\href{https://pyabc.readthedocs.io}{pyabc.readthedocs.io}]{Klinger2018}.
 For $t=0$, the proposal particle is sampled from the prior distribution, $p(\theta)$.
-For $t>0$, the proposal particle is sampled from the particles accepted in the previous iteration, $\{\theta_{i,t-1}\}_{i=1}^{n_{t-1}}$, each with a probability relative to its weight $w_{t-1}(\theta_{i,t-1})$ (see below).
+For $t>0$, the proposal particle is sampled from the particles accepted in the previous iteration, $\{\theta_{i,t-1}\}_{i=1}^{n_{t-1}}$, each with a probability relative to its weight $W_{t-1}(\theta_{i,t-1})$ (see below).
 The proposal particle is then perturbed using a kernel perturbation kernel, $K_t(\theta^* \mid \theta)$ where $\theta$ is the sample from the previous iteration.
@@
 The acceptance threshold $\epsilon_t$ is chosen as the median of the $1-\likelihood(\theta)$ of the particles accepted in the previous iteration, $t-1$, and $\epsilon_0=0.01$. 
-For each accepted particle $\theta_{i,t}$ a weight $w_t(\theta_{i,t})$ is assigned: for $t=0$, $w_0(\theta_{i,0})=1$, and for $t>0$, 
-$w_t(\theta_{i,t}) = p(\theta_{i,t}) / \sum_{i=1}^{n_{t-1}}{w_{t-1}(\theta_{i,t-1}) K_t(\theta_{i,t}, \theta_{i,t-1})}$, where $p(\theta)$ is the prior density of $\theta$ and $K_t(\theta', \theta)$ is the probability of a perturbation from $\theta$ to $\theta'$.
+For each accepted particle $\theta_{i,t}$ a weight $W_t(\theta_{i,t})$ is assigned: for $t=0$, $W_0(\theta_{i,0})=1$, and for $t>0$, 
+$W_t(\theta_{i,t}) = p(\theta_{i,t}) / \sum_{i=1}^{n_{t-1}}{W_{t-1}(\theta_{i,t-1}) K_t(\theta_{i,t}, \theta_{i,t-1})}$, where $p(\theta)$ is the prior density of $\theta$ and $K_t(\theta', \theta)$ is the probability of a perturbation from $\theta$ to $\theta'$.
 $K_t(\theta' \mid \theta)$ is a multivariate normal distribution. % TODO construction of K_T from \theta_{i,t-1}
 
-Acceptance is determined according to the approximate likelihood, which has a maximum value of 0.875.
-Thus, we terminated the inference when $\epsilon \le 0.13$ after six iterations, with $n_6=982$ accepted parameter vectors and effective sample size (ESS) 651 (\autoref{fig:convergence}).
+Acceptance is determined according to the approximate likelihood~(\autoref{eq:heatstress-likelihood}), which has a maximum value of 0.875.
+Thus, we terminated the inference when $\epsilon \le 0.13$ after six iterations, with $n_6=982$ accepted parameter vectors and effective sample size ESS=651 (\autoref{fig:convergence}).
 % TODO for multi-locus model we need to say something else
-Running the inference algorithm with different initialization seeds and different number of simulations with produced similar outcomes (\autoref{fig:seeds}).
+Running the inference algorithm with different initialization seeds and less or more simulations for approximating the likelihood produced similar posterior distributions~(\autoref{fig:seeds}).
 %TODO for multi-locus model, write about them here also.

### 2022-01-12 ms/ms.tex
revise figures formatting and captions, and a bit of the text, and added some todos

@@
 $W_t(\theta_{i,t}) = p(\theta_{i,t}) / \sum_{i=1}^{n_{t-1}}{W_{t-1}(\theta_{i,t-1}) K_t(\theta_{i,t}, \theta_{i,t-1})}$, where $p(\theta)$ is the prior density of $\theta$ and $K_t(\theta', \theta)$ is the probability of a perturbation from $\theta$ to $\theta'$.
-$K_t(\theta' \mid \theta)$ is a multivariate normal distribution. % TODO construction of K_T from \theta_{i,t-1}
+$K_t(\theta' \mid \theta)$ is a multivariate normal distribution, fitted at iteration $t$ to the particles from the previous iteration, \{$\theta_{i,t-1}\}_{i=1}{n_{t-1}}$, and their weights, $\{W(\theta_{i,t-1})\}_{i=1}^{n_{t-1}}$.
 
@@
 Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix.
-We sampled 10,000 samples to use as a prior distribution~(Figures~\ref{fig:growth-curves-30deg} and~\ref{fig:growth-curves-39deg}).
-
+We sampled 10,000 samples to use as a prior distribution~(\autoref{fig:growth-curves}).
 We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$. In lieu of a better prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$. 
+To increase computational efficiency, we also assumed $w(2n*)>w(2n+1*)>w(2n+1)>w(2n)$; running the inference without this assumption produced similar results. % TODO IK is this correct?
 
@@
 % TODO if you need to remove this sentence: Also, with the uninformative prior for all fitness values, the fixation time of \anwt is about 16 on average, much lower than the (unpublished) experimental result, which is 200-300 generations.
-
 We tried to use additional growth curves.
@@
 % TODO what are the results ?
-
 We also tried to use growth curves of \anwt\; and \euwt\; in \SI{30}{\celsius} to estimate $1-c$.
@@
 In both cases the algorithm failed to converge.
-
-% TODO : Actually I assume in code that  w(2n*)>w(2n+1*)>w(2n+1)>w(2n) just for speeding up the simulations (I don't check some parameters combinations in this way).
-But I checked also that when I don't adding this constrain the result was the same.
-However I didn't add reverse direction in this case.
 
@@
 $w_{2n^*}=1.028\substack{+0.001 \\ -0.001}$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$.
-% describe b and c TODO? what does it mean?
+% TODO we are jumping between b c s and w2n* w2n+1 w2n+1*. we should report all models and results either with b c s or with w. you can compute b c s from w before computing the HDIs.
 This estimated aneuploidy rate agrees with previous estimates. % TODO refs
@@
 However, a model without aneuploidy where the aneuploidy rate is fixed at zero, $\delta=0$, cannot explain the experimental observations~(Figure~\ref{fig:2n*-fixation}).
-The estimated parameter values without aneuploidy are: %% generated by MAP-and-hdi.ipynb
+The estimated parameter values without aneuploidy are:
+% generated by MAP-and-hdi.ipynb
 mutation rate, $\mu=7.837\substack{+0.078 \\ -0.123}\times10^{-9}$,
@@
 
-\begin{figure}[b!]
+\begin{figure}[h!]
 
-  \begin{subfigure}[t]{0.9\textwidth}
+  \begin{subfigure}{0.5\textwidth}
       \centering
-      \includegraphics[height=1.8in]{../figures/fixation-plot-a.pdf}      
+      \includegraphics[width=\textwidth]{../figures/fixation-plot-a.pdf}      
       \label{fig:fit}
-    \caption{\textbf{A. Distribution.}}
   \end{subfigure}
-
-  \vspace*{4mm}
-
-  \begin{subfigure}[t]{0.9\textwidth}
+  \begin{subfigure}{0.5\textwidth}
       \centering
-      \includegraphics[height=1.8in]{../figures/fixation-plot-b.pdf}      
+      \includegraphics[width=\textwidth]{../figures/fixation-plot-b.pdf}      
       \label{fig:fit-cumulative}
-    \caption{\textbf{B. Cumulative distribution.}}
   \end{subfigure}
-
+  % TODO please add A and B to the figures themselves, not using latex.
+  % remove legends
   \caption{
     \textbf{Single-locus model fit with and without aneuploidy.}
-    \textbf{(A)} The distribution of adaptation time (fixation of \eumt) in $10,000$ simulations of the single-locus model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange); and a model fitted to the data assuming $\delta=0$ (green).
+    \textbf{(A)} The distribution of time to fixation of \eumt (i.e., adaptation time) in $10,000$ simulations of the single-locus model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange); and a model fitted to the data assuming $\delta=0$ (green).
 In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and the other by generation 2,350 (dashed lines) but not before generation 200.
-Thus, the blue distribution demonstrates a better fit compared to the green, and the yellow demonstrates a very poor fit.
+Thus, the blue distribution is a better fit compared to the green, and the yellow histogram has a very poor fit.
 The last bin contains all the simulations with time equal or greater than 3,000.
     \textbf{(B)}
-	Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy (i.e., $\delta=0$ is fixed).
-     The MAP likelihood (\autoref{eq:heatstress-likelihood}) is 0.84 and 0.67 with and without aneuploidy.
+	Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy in blue and green, respectively.
+     The MAP likelihood (\autoref{eq:heatstress-likelihood}) is 0.84 and 0.67 with and without aneuploidy, respectively.
      }
@@
   \centering
-\includegraphics[width=0.6\textwidth]{../figures/ppc-plot.pdf}
+\includegraphics[width=0.8\textwidth]{../figures/ppc-plot.pdf}
   \caption{
   \textbf{Single-locus model posterior genotype frequency dynamics.} 
-  The posterior predicted frequency of the four genotypes. Each line is the average of 10,000 simulations using parameters drawn from the posterior distribution (\autoref{fig:posterior}).
+  The posterior predicted frequency of the four genotypes. Each curve is the average of 10,000 simulations using parameters drawn from the posterior distribution (\autoref{fig:posterior}).
   }
@@
 An evolutionary theory of aneuploidy will be central to the interpretation of experimental and clinical observations and design of new hypotheses, experiments, and treatments~\citep{Carja2014}.
-For example, despite the lack of theoretical models, aneuploidy has been invoked in a new strategy to combat pathogens and tumour cells by setting “evolutionary traps”~\citep{Gerstein2015,Chen2015}, in which a condition that predictably leads to emergence of aneuploidy is applied, followed by a condition that specifically selects against aneuploid cells.
+For example, despite the lack of theoretical models, aneuploidy has been invoked in a new strategy to combat pathogens and tumour cells by setting ``evolutionary traps''~\citep{Gerstein2015,Chen2015}, in which a condition that predictably leads to emergence of aneuploidy is applied, followed by a condition that specifically selects against aneuploid cells.
 
@@
 
+\begin{figure}[h]
 % Fig Curveball 30deg
 % generated with growth_curves.ipynb
-\begin{figure}[h]
+\begin{subfigure}{0.5\textwidth}
     \centering
-	\includegraphics[width=0.5\textwidth]{../figures/evo39_fitness_30deg.pdf}
+	\includegraphics[width=\textwidth]{../figures/evo39_fitness_30deg.pdf}
+    \label{fig:growth-curves-30deg}
+\end{subfigure}
+% Fig Curveball 39deg
+% generated with growth_curves.ipynb
+\begin{subfigure}{0.5\textwidth}
+    \centering
+	\includegraphics[width=\textwidth]{../figures/evo39_fitness_39deg.pdf}
+    \label{fig:growth-curves-39deg}
+\end{subfigure}
 \caption{
-    \textbf{Fitness estimation from \SI{30}{\celsius}.}
+    \textbf{Fitness estimation from growth curves.}
+    \textbf{(A)} Fitness estimation from growth curves collected at \SI{30}.{\celsius}
+    \textbf{(B)} Fitness estimation from growth curves collected at \SI{39}.{\celsius}
+    Growth curves previously described in \citet[Figs. 3C, 4A, and S2]{Yona2012}.
+	Fitness estimated from growth curves using \texttt{Curveball}, a method for predicting results of competition experiments from growth curve data~\citep[\href{https://curveball.yoavram.com}{curveball.yoavram.com}]{Ram2019}.
+See \emph{Models and Methods, Prior distributions} for more details.
 } 
-    \label{fig:growth-curves-30deg}
+\label{fig:growth-curves}
 \end{figure}
 % TODO add caption
-
-% Fig Curveball 39deg
-% generated with growth_curves.ipynb
-\begin{figure}[h]
-    \centering
-	\includegraphics[width=0.5\textwidth]{../figures/evo39_fitness_39deg.pdf}
-    \caption{
-    \textbf{Fitness estimation from \SI{39}{\celsius}.}
-    } 
-    \label{fig:growth-curves-39deg}
-\end{figure}
-% TODO add caption
-
-%TODO combine to one figure with the previous one and with refined
 %TODO please report here the point estimate and 95% CI for the three distributions, I only have for 1-c+b: 0.9863 (95% CI 0.9790 - 0.9939).
@@
       \includegraphics[width=0.45\textwidth]{../figures/convergence-p1_mr.pdf}      
-      \includegraphics[width=0.45\textwidth]{../figures/convergence-p2_tr.pdf}      
-      \includegraphics[width=0.45\textwidth]{../figures/convergence-p3_w1.pdf}      
-      \includegraphics[width=0.45\textwidth]{../figures/convergence-p4_w2.pdf}      
-      \includegraphics[width=0.45\textwidth]{../figures/convergence-p5_w3.pdf}      
+      \includegraphics[width=0.45\textwidth]{../figures/convergence-p2_tr.pdf} \\     
+      \includegraphics[width=0.325\textwidth]{../figures/convergence-p3_w1.pdf}      
+      \includegraphics[width=0.325\textwidth]{../figures/convergence-p4_w2.pdf}      
+      \includegraphics[width=0.325\textwidth]{../figures/convergence-p5_w3.pdf}      
      \label{fig:convergence-A}
-    \caption{\textbf{A. Posterior densities per iteration.}}
     \end{subfigure}
     \\    
-    \vspace*{4mm}
-
   \begin{subfigure}{1\textwidth}
-  	\includegraphics[width=1\textwidth]{../figures/ess.pdf}
+  	\includegraphics[width=\textwidth]{../figures/ess.pdf}
      \label{fig:convergence-B}
-     \vspace*{-8mm}
-    \caption{\textbf{B. ABC-SMC population properties per iteration.}}
   \end{subfigure}
-
     \caption{
     \textbf{Single-locus model inference convergence.} 
-    ABC-SMC algorithm was executed to inference single-locus model parameters. \textbf{(A)} For each parameter of the model, there's shown KDE of the posterior density of the inference per iteration.
-We can see that the posterior didn't change significantly already from the first iteration.
-\textbf{(B)} Here are shown measures of convergence for the ABC-SMC algorithm.
-After iteration number 6, the acceptance threshold was $\epsilon=0.13$, the acceptance rate was $0.018$, the number of particles was 982, and the effective sample size (ESS) 651.
+    The ABC-SMC algorithm was used to infer the parameters of the single-locus model.
+\textbf{(A-E)} The approximate posterior densities of each model parameter at each iteration of the ABC-SMC algorithm demonstrates convergence, as the posterior did not significantly change after iteration, $t=1$.
+    \textbf{(B)} ABC-SMC measures of convergence.
+After iteration number 6, the acceptance threshold was $\epsilon=0.13$, the acceptance rate was $0.018$, the number of particles was 982, and the effective sample size ESS=651.
 }
     \label{fig:convergence}
+    % TODO
+    % legend - t=1, t=2, ... (iter->t)
+    % what is lambda on x axis??
+    % use full name of parameter + its symbol
+    % add panel labels - ABCDEFGHI
 \end{figure}
@@
 % TODO put legend in panel A top right corner
-% panel labels
+% panel labels A->E
 % y label "Density"; y ticks are meaningless, right? 
-% x labels - see comments above.
+% x labels - see comments above. add symbols, fitness of 2n*, etc.
 % color by # of reps; second run with 1000 reps same color as 1st but with dashed line

### 2022-01-13 ms/ms.tex
some more figures fixes; updated fig s2 - sensitivity analysis figure

@@
 We also tried to infer parameters of the single-locus model when the mutation rate in aneuploid cells, compared to euploid cells, is increased by a factor $\tau$=1, 33/32 (due to an additional chromosome), 2, 5, 10 or 100 (due to genetic instability).
-We found that for $\tau=$1, 33/32, 2, and 5 the posterior was not significantly different from the model with $\tau=1$, while for $\tau=10$ the estimated mutation rate was slightly lower, $\mu=1.654\substack{+0.879 \\ -1.775}\times10^{-6}$, and for $\tau=100$ the estimated mutation rate was significantly lower and the aneuploidy rate had higher variability.
-The MAP and 50\% HDI with $\tau=100$ were: %
-TODO add to MAP-and-hdi.ipynb and check the values here
+We found that for $\tau=$33/32, 2, and 5 the posterior was not significantly different from the model with $\tau=1$, while for $\tau=10$ the estimated mutation rate was slightly lower, $\mu=1.654\substack{+0.879 \\ -1.775}\times10^{-6}$, and for $\tau=100$ the estimated mutation rate was significantly lower and the aneuploidy rate had higher variability.
+The MAP and 50\% HDI with $\tau=100$ were:
+%TODO add to MAP-and-hdi.ipynb and check the values here
 $\mu=5.538\substack{+1.611 \\ -5.311}\times10^{-7}$,
@@
   \end{subfigure}
-  % TODO please add A and B to the figures themselves, not using latex.
-  % remove legends
   \caption{
     \textbf{Single-locus model fit with and without aneuploidy.}
-    \textbf{(A)} The distribution of time to fixation of \eumt (i.e., adaptation time) in $10,000$ simulations of the single-locus model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange); and a model fitted to the data assuming $\delta=0$ (green).
-In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and the other by generation 2,350 (dashed lines) but not before generation 200.
+    \textbf{(A)} The distribution of time to fixation of \eumt\ (i.e., adaptation time) in $10,000$ simulations of the single-locus model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange); and a model fitted to the data assuming $\delta=0$ (green).
+%TODO_YR is it ok the paragraph below?  Green bars don't relate to aneuploidy lost, but to 2n* fixation...so it should be "the adaptation time of one generation was 1700.."? Correct? 
+In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and the other by generation 2,350 (dashed lines) but not before generation 450.
 Thus, the blue distribution is a better fit compared to the green, and the yellow histogram has a very poor fit.
@@
 	Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy in blue and green, respectively.
+%TODO_YR shouldn't we write here about green that it is like in (A) - same MAP as with aneuploidy but with delta=0
      The MAP likelihood (\autoref{eq:heatstress-likelihood}) is 0.84 and 0.67 with and without aneuploidy, respectively.
@@
   \begin{subfigure}{0.3\textwidth}
-      \includegraphics[width=\textwidth]{../figures/sensitivity-mutation-rate-plot.pdf}      
+      \includegraphics[width=\textwidth]{../figures/sensitivity-A.pdf}      
       \label{fig:sensitivity-mutation}
@@
   \begin{subfigure}{0.3\textwidth}
-      \includegraphics[width=\textwidth]{../figures/sensitivity-trisomy-rate-plot.pdf}      
+      \includegraphics[width=\textwidth]{../figures/sensitivity-B.pdf}      
       \label{fig:sensitivity-aneuploidy}
@@
    \begin{subfigure}{0.3\textwidth}
-      \includegraphics[width=\textwidth]{../figures/sensitivity-w(2n+1)-plot.pdf}      
+      \includegraphics[width=\textwidth]{../figures/sensitivity-C.pdf}      
       \label{fig:sensitivity-anwt}
@@
     \begin{subfigure}{0.3\textwidth}
-      \includegraphics[width=\textwidth]{../figures/sensitivity-w(2n+1*)-plot.pdf}      
+      \includegraphics[width=\textwidth]{../figures/sensitivity-D.pdf}      
       \label{fig:sensitivity-anmt}
@@
     \begin{subfigure}{0.3\textwidth}
-      \includegraphics[width=\textwidth]{../figures/sensitivity-w(2n*)-plot.pdf}      
+      \includegraphics[width=\textwidth]{../figures/sensitivity-E.pdf}      
       \label{fig:sensitivity-eumt}
@@
     \textbf{Likelihood profiles.} The different curves show how the single-locus model log-likelihood changes when each of the parameter changes while the other parameters remain fixed at their MAP estimates.
+Dashed vertical line represents the MAP value.
   }
-  % TODO mark the MAP parameter on the figure with a vertical line
-  % show the log-likelihood rather than the likelihood
-  % can we add more points? 
-  % x labels start with capital letter, should include the parameter (Mutation rate, $\mu$) and the name (Fitness of 2n+1, w_{2n+1}), dont use w(xxx) but rather w_{xxx} as in the main text
-  % y label "Approximate log-likelihood"
   % trisomy rate -> aneuploidy gain rate
+  % TODO_YR: To call it aneuploidy gain rate? Because it is loss rate too. I think we called it aneuploidy rate before.
   \label{fig:sensitivity}
@@
 \textbf{(A-E)} The approximate posterior densities of each model parameter at each iteration of the ABC-SMC algorithm demonstrates convergence, as the posterior did not significantly change after iteration, $t=1$.
-    \textbf{(B)} ABC-SMC measures of convergence.
+    \textbf{(F-I)} ABC-SMC measures of convergence.
 After iteration number 6, the acceptance threshold was $\epsilon=0.13$, the acceptance rate was $0.018$, the number of particles was 982, and the effective sample size ESS=651.

### 2022-01-13 ms/ms.tex
minor fixes, delta not lambda, and t not iter
small fix: approximate likelihood at sensitivity plot

@@
   \caption{
-    \textbf{Likelihood profiles.} The different curves show how the single-locus model log-likelihood changes when each of the parameter changes while the other parameters remain fixed at their MAP estimates.
+    \textbf{Likelihood profiles.} The different curves show how the single-locus model approximate likelihood, $\likelihood_{!}(\theta)+10^{-2}$, changes when each of the parameter changes while the other parameters remain fixed at their MAP estimates.
+We add $10^{-2}$ to the likelihood for proper representation in a log scale.
 Dashed vertical line represents the MAP value.
@@
     \label{fig:convergence}
-    % TODO
-    % legend - t=1, t=2, ... (iter->t)
-    % what is lambda on x axis??
-    % use full name of parameter + its symbol
-    % add panel labels - ABCDEFGHI
 \end{figure}

### 2022-01-17 ms/ms.tex
updated fig1 and dynamics plots
respond to IK Qs; small revisions and comments

@@
 
-\affil[1]{School of Computer Sciences, IDC Herzliya, Herzliya, Israel}
+\affil[1]{School of Computer Sciences, Reichman University, Herzliya, Israel}
 \affil[2]{School of Zoology, Faculty of Life Sciences, Tel Aviv University, Tel Aviv, Israel}
@@
     \textbf{Single-locus model fit with and without aneuploidy.}
-    \textbf{(A)} The distribution of time to fixation of \eumt\ (i.e., adaptation time) in $10,000$ simulations of the single-locus model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange); and a model fitted to the data assuming $\delta=0$ (green).
-%TODO_YR is it ok the paragraph below?  Green bars don't relate to aneuploidy lost, but to 2n* fixation...so it should be "the adaptation time of one generation was 1700.."? Correct? 
-In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and the other by generation 2,350 (dashed lines) but not before generation 450.
+    \textbf{(A)} The distribution of time to fixation of \eumt\ (i.e., adaptation time) in $10,000$ simulations of the single-locus model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange), and a model fitted to the data assuming $\delta=0$ (green).
+In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and another by generation 2,350 (dashed lines) but not before generation 450.
 Thus, the blue distribution is a better fit compared to the green, and the yellow histogram has a very poor fit.
@@
     \textbf{(B)}
-	Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy in blue and green, respectively.
-%TODO_YR shouldn't we write here about green that it is like in (A) - same MAP as with aneuploidy but with delta=0
+	Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy in blue and green, respectively, and corresponding to the blue and green bars in panel A.
 The MAP likelihood (\autoref{eq:heatstress-likelihood}) is 0.84 and 0.67 with and without aneuploidy, respectively.
@@
 \end{figure}
-% TODO add caption
 %TODO please report here the point estimate and 95% CI for the three distributions, I only have for 1-c+b: 0.9863 (95% CI 0.9790 - 0.9939).
+% TODO add panel labels
 
@@
   \caption{
-    \textbf{Likelihood profiles.} The different curves show how the single-locus model approximate likelihood, $\likelihood_{!}(\theta)+10^{-2}$, changes when each of the parameter changes while the other parameters remain fixed at their MAP estimates.
-We add $10^{-2}$ to the likelihood for proper representation in a log scale.
+    \textbf{Likelihood profiles.} Sensitivity of the approximate likelihood, $\likelihood(\theta)$, of the single-locus model to changing a single parameter while the other parameters remain fixed at their MAP estimates.
 Dashed vertical line represents the MAP value.
   }
-  % trisomy rate -> aneuploidy gain rate
-  % TODO_YR: To call it aneuploidy gain rate? Because it is loss rate too. I think we called it aneuploidy rate before.
+  % TODO add 10^-10, not 10^-2
+  % TODO ylabel - L shouldn't have a subscript !, this was used only for the model without aneuploidy
+  % TODO add more points in the peaks
+
   \label{fig:sensitivity}
@@
     The ABC-SMC algorithm was used to infer the parameters of the single-locus model.
-\textbf{(A-E)} The approximate posterior densities of each model parameter at each iteration of the ABC-SMC algorithm demonstrates convergence, as the posterior did not significantly change after iteration, $t=1$.
+\textbf{(A-E)} The approximate posterior distributions of model parameters at each iteration of the ABC-SMC algorithm demonstrates convergence, as the posterior did not significantly change after the first iteration, $t=1$.
     \textbf{(F-I)} ABC-SMC measures of convergence.
@@
     \label{fig:convergence}
+    % TODO legend - t=6, etc (add =)
+	% x labels missing for G H I - either add them or put "Iteration" at the middle of the axes. whatever is easier.
 \end{figure}

### 2022-01-17 ms/ms.tex
update TODOs; fix affil

@@
 
-\affil[1]{School of Computer Sciences, Reichman University, Herzliya, Israel}
+\affil[1]{School of Computer Science, Reichman University, Herzliya, Israel}
 \affil[2]{School of Zoology, Faculty of Life Sciences, Tel Aviv University, Tel Aviv, Israel}
@@
     In both panels arrows denote transitions between genotypes, with transitions rates: $\mu$, beneficial mutation rate; $\delta$, aneuploidy rate.
-    % TODO IK: remove k (not tau either!); also k mu is too close to the arrow right now; change alpha to delta in the figure 
   }
@@
   \centering
-\includegraphics[width=0.8\textwidth]{../figures/ppc-plot.pdf}
+\includegraphics[width=0.8\textwidth]{../figures/dynamics.pdf}
   \caption{
   \textbf{Single-locus model posterior genotype frequency dynamics.} 
-  The posterior predicted frequency of the four genotypes. Each curve is the average of 10,000 simulations using parameters drawn from the posterior distribution (\autoref{fig:posterior}).
+  The posterior predicted frequency of the four genotypes. Each of 20 curves is the average of 10,000 simulations using parameters drawn from the posterior distribution (\autoref{fig:posterior}).
   }
   \label{fig:ppc-plot}
-  % TODO add legend
-  % xticks - replace 500, 1000,1500,2000,2500 with 450, 1700, 2350
+  % TODO put legend inside figure - I think there is room in middle right
 \end{figure}

### 2022-01-17 src/pyabc_visualization_patch/pyabc/___init___.py
updated posterior plot

@@ -1,0 +1,51 @@
+"""
+Visualization
+=============
+
+Visualize results of ABCSMC runs.
+"""
+
+from .kde import (
+    plot_kde_1d,
+    plot_kde_1d_highlevel,
+    plot_kde_2d,
+    plot_kde_2d_highlevel,
+    plot_kde_matrix,
+    plot_kde_matrix2,
+    plot_kde_matrix_for_paper_was,
+    plot_kde_matrix_for_paper,
+    plot_kde_matrix_highlevel)
+from .sample import (
+    plot_sample_numbers,
+    plot_total_sample_numbers,
+    plot_sample_numbers_trajectory,
+    plot_acceptance_rates_trajectory,
+    plot_lookahead_evaluations,
+    plot_lookahead_final_acceptance_fractions,
+    plot_lookahead_acceptance_rates)
+from .epsilon import (
+    plot_epsilons)
+from .histogram import (
+    plot_histogram_1d,
+    plot_histogram_1d_lowlevel,
+    plot_histogram_2d,
+    plot_histogram_2d_lowlevel,
+    plot_histogram_matrix,
+    plot_histogram_matrix_lowlevel)
+from .credible import (
+    plot_credible_intervals,
+    plot_credible_intervals_for_time)
+from .model_probabilities import (
+    plot_model_probabilities)
+from .effective_sample_size import (
+    plot_effective_sample_sizes)
+from .data import (
+    plot_data_callback,
+    plot_data_default)
+from .walltime import (
+    plot_total_walltime,
+    plot_walltime,
+    plot_walltime_lowlevel,
+    plot_eps_walltime,
+    plot_eps_walltime_lowlevel)
+

### 2022-01-17 src/pyabc_visualization_patch/pyabc/kde.py

@@ -1,0 +1,926 @@
+"""Kernel density estimate plots
+
+To plot densities from the weighted importance samples, the visualization
+routines employ a kernel density estimate. Note that this can "over-smoothen"
+so that local structure is lost. If this could be the case, it makes sense
+to in the argument `kde` reduce the `scaling` in the default
+MultivariateNormalTransition(), or to replace it by a GridSearchCV() to
+automatically find a visually good level of smoothness.
+"""
+
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+
+from ..transition import MultivariateNormalTransition
+from ..storage import History
+from .util import format_plot_matrix
+
+
+def kde_1d(df, w, x, xmin=None, xmax=None, numx=50, kde=None):
+    """
+    Calculates a 1 dimensional histogram from a Dataframe and weights.
+
+    For example, a results distribution might be obtained from the history
+    class and plotted as follows::
+
+        df, w = history.get_distribution(0)
+        x, pdf = hist_2d(df, w, "x")
+        plt.plot(x, pdf)
+
+
+    Parameters
+    ----------
+    df: Pandas Dataframe
+        The rows are the observations, the columns the variables
+    w: np.ndarray
+        The corresponding weights
+    x: str
+        The variable for the x-axis
+    xmin: float, optional
+        The lower limit in x for the histogram.
+        If left empty, it is set to the minimum of the ovbservations of the
+        variable to be plotted as x.
+    xmax: float, optional
+        The upper limit in x for the histogram.
+        If left empty, it is set to the maximum of the ovbservations of the
+        variable to be plotted as x.
+    numx: int, optional
+        The number of bins in x direction.
+        Defaults to 50.
+    kde: pyabc.Transition, optional
+        The kernel density estimator to use for creating a smooth density
+        from the sample. If None, a multivariate normal kde with
+        cross-validated scaling is used.
+
+    Returns
+    -------
+    x, pdf: (np.ndarray, np.ndarray)
+        The x and the densities at these points.
+        These can be passed for plotting, for example as
+        plt.plot(x, pdf)
+
+    """
+    if kde is None:
+        kde = MultivariateNormalTransition(scaling=1)
+    kde.fit(df[[x]], w)
+
+    if xmin is None:
+        xmin = df[x].min()
+    if xmax is None:
+        xmax = df[x].max()
[…]

### 2022-01-18 ms/ms.tex
..
fig 2 revised caption and added todos

@@
   \textbf{Posterior distribution of single-locus model.}
-% TODO histogram on diagonal, 2d histogram without interpolation below diagonal
-% TODO mutation rate tick labels should be in 10^-6, not 10^-5, and write 10^-6 not 1e-5; also it looks like you get positive density for mutation rate = 0 which doesnt make sense - set limit to 0.
-% change trisomy rate to aneuploidy gain rate
-% remove "mutation rate" from y label on top-left
-% remove the y-ticks from the diagonal, they are meaningless, right?
-% Fitness of 2n+1, w_{2n+1} (not 2n+1 fitness)
-% TODO change dashed orange to red and make it a bit bolder.
-On the diagonal, the posterior kernel density estimate for each parameter. 
-Below the diagonal, joint posterior density of two parameters (dark purple and bright yellow for low and high density, respectively).
-Red markers and orange lines for the joint MAP estimates (which may differ from the marginal maximum density, as the marginal distribution integrates over all other parameters).
+On the diagonal, the posterior kernel distribution of each model parameter. 
+Below the diagonal, joint posterior distribution of pairs of model parameters (dark purple and bright yellow for low and high density, respectively).
+Red markers and orange lines for the joint MAP estimate (which may differ from the marginal MAP, as the marginal distribution integrates over all other parameters).
 } 
+% TODO less y axis ticks
+% more x axes ticks?
   \label{fig:posterior}

### 2022-01-19 src/pyabc_visualization_patch/pyabc/kde.py
updated plots S2 (likelihood profiles), Fig 4 (dynamics), and Fig 2 (posterior)

@@ -720,20 +720,6 @@ def plot_kde_matrix_for_paper(df, w, limits=None, colorbar=True, height=2.5,
                     # kde=kde,
                     xname=names[x.name], yname=names[y.name], bins=20)
 
-        # plot_kde_2d(df, w,
-        #             x.name, y.name,
-        #             xmin=limits.get(x.name, default)[0],
-        #             xmax=limits.get(x.name, default)[1],
-        #             ymin=limits.get(y.name, default)[0],
-        #             ymax=limits.get(y.name, default)[1],
-        #             numx=numx, numy=numy,
-        #             ax=ax, 
-        #             title=None, 
-        #             colorbar=colorbar,
-        #             refval=refval, refval_color=refval_color,
-        #             kde=kde,
-        #             xname=names[x.name], yname=names[y.name])
-
     def scatter(x, y, ax):
         alpha = w / w.max()
         colors = np.zeros((alpha.size, 4))
@@ -753,19 +739,19 @@ def plot_kde_matrix_for_paper(df, w, limits=None, colorbar=True, height=2.5,
         ax.set_xlim(min(df[x.name])-1e-8) #1e-8 to print 0
         ax.set_yticks([])
         if i==0:
-            ax.text(0.87, -0.13, '$x10^{-6}$', transform=ax.transAxes)
-            ax.set_xticks([0,5e-6, 10e-6])
+            ax.text(0.88, -0.11, '$x10^{-6}$', transform=ax.transAxes)
+            ax.set_xticks([1e-6, 4e-6, 7e-6, 10e-6])
             ax.set_xticklabels([int(a) for a in ax.get_xticks()*1e6]);
         if i==1:
-            ax.set_xticks([0, 0.0025,0.0050])
-            ax.set_xticklabels([0, 0.0025,0.0050])
-            ax.set_xlim(min(df[x.name])-1e-5)
-#         plot_kde_1d(df, w, x.name,
-#                     xmin=limits.get(x.name, default)[0],
-#                     xmax=limits.get(x.name, default)[1],
-#                     numx=numx,
-#                     ax=ax, refval=refval, refval_color=refval_color,
-#                     kde=kde, xname=x.name)
+            ax.text(0.88, -0.11, '$x10^{-3}$', transform=ax.transAxes)
+            ax.set_xticks([1e-3, 2e-3, 3e-3, 4e-3])
[…]

### 2022-01-20 ms/ms.tex
updated plots S2 (likelihood profiles), Fig 4 (dynamics), and Fig 2 (posterior)
updated different runs figure, plus small fixes on convergence plot and ess plot

@@
 We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$. In lieu of a better prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$. 
-To increase computational efficiency, we also assumed $w(2n*)>w(2n+1*)>w(2n+1)>w(2n)$; running the inference without this assumption produced similar results. % TODO IK is this correct?
+To increase computational efficiency, we also assumed $w(2n*)>w(2n+1*)>w(2n+1)>w(2n)$; running the inference without this assumption produced similar results.
 
@@
 % FIG posterior
+%% generated with posterior-plot.ipynb
 \begin{figure}[h!]
@@
 } 
-% TODO less y axis ticks
-% more x axes ticks?
   \label{fig:posterior}
@@
 The estimated parameter values without aneuploidy are:
-% generated by MAP-and-hdi.ipynb
+%% generated with MAP-and-hdi.ipynb
 mutation rate, $\mu=7.837\substack{+0.078 \\ -0.123}\times10^{-9}$,
@@
 % Fig single-locus model with and without aneuploidy
-%%% generated by with-aneuploidy_vs_no_aneuploidy.ipynb
-
+%% generated with with-aneuploidy_vs_no_aneuploidy.ipynb
 \begin{figure}[h!]
-
   \begin{subfigure}{0.5\textwidth}
@@
   \label{fig:ppc-plot}
-  % TODO put legend inside figure - I think there is room in middle right
 \end{figure}
@@
 % Fig likelihood profile
+%% generated by sensitivity-analysis.ipynb
 \begin{figure}[p]
@@
   }
-  % TODO add 10^-10, not 10^-2
-  % TODO ylabel - L shouldn't have a subscript !, this was used only for the model without aneuploidy
-  % TODO add more points in the peaks
 
@@
 % Fig ABC iterations, convergence
-%% generated by convergence.ipynb
+%% generated with convergence.ipynb
 \begin{figure}[h]
@@
     \label{fig:convergence}
-    % TODO legend - t=6, etc (add =)
-	% x labels missing for G H I - either add them or put "Iteration" at the middle of the axes. whatever is easier.
 \end{figure}
@@
 % Fig Validation of posteriors
+%% generated with diff-runs.ipynb
 \begin{figure}[h]
     \centering
-	\includegraphics[width=0.6\textwidth]{../figures/seeds.pdf}
+      \includegraphics[width=0.45\textwidth]{../figures/runs-A.pdf}      
+      \includegraphics[width=0.45\textwidth]{../figures/runs-B.pdf} \\     
+      \includegraphics[width=0.325\textwidth]{../figures/runs-C.pdf}      
+      \includegraphics[width=0.325\textwidth]{../figures/runs-D.pdf}      
+      \includegraphics[width=0.325\textwidth]{../figures/runs-E.pdf} 
        \caption{
     \textbf{Posterior distribution validation.}
-    The posterior distributions of the single-locus model parameters are roughly the same regardless of the number of simulations (10-1,000) used to approximate the likelihood.
+    The posterior distributions of the single-locus model parameters are roughly the same regardless of the number of simulations (4-10,000) used to approximate the likelihood.
     }
+
      \label{fig:seeds}
  \end{figure}
-% TODO put legend in panel A top right corner
-% panel labels A->E
-% y label "Density"; y ticks are meaningless, right? 
-% x labels - see comments above. add symbols, fitness of 2n*, etc.
-% color by # of reps; second run with 1000 reps same color as 1st but with dashed line
 

### 2022-01-24 ms/ms.tex
small revisions

--- 2022-01-18 ms/ms.tex
@@
 % TODO cite Yang, F. et al. The fitness costs and benefits of trisomy of each Candida albicans chromosome. Genetics 218, 1–7 (2021). https://academic.oup.com/genetics/article/218/2/iyab056/6218773
+% cite: https://www.cell.com/developmental-cell/fulltext/S1534-5807(21)00562-1?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS1534580721005621%3Fshowall%3Dtrue
+% cite: https://www.cell.com/developmental-cell/fulltext/S1534-5807(21)00592-X?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS153458072100592X%3Fshowall%3Dtrue
 
@@
   \textbf{Posterior distribution of single-locus model.}
-On the diagonal, the posterior kernel distribution of each model parameter. 
-Below the diagonal, joint posterior distribution of pairs of model parameters (dark purple and bright yellow for low and high density, respectively).
+On the diagonal, the inferred posterior distribution of each model parameter. 
+Below the diagonal, the inferred joint posterior distribution of pairs of model parameters (dark purple and bright yellow for low and high density, respectively).
 Red markers and orange lines for the joint MAP estimate (which may differ from the marginal MAP, as the marginal distribution integrates over all other parameters).
 } 
-% TODO less y axis ticks
-% more x axes ticks?
   \label{fig:posterior}
@@
   \caption{
-  \textbf{Single-locus model posterior genotype frequency dynamics.} 
-  The posterior predicted frequency of the four genotypes. Each of 20 curves is the average of 10,000 simulations using parameters drawn from the posterior distribution (\autoref{fig:posterior}).
+  \textbf{Posterior genotype frequency dynamics for the single-locus model.} 
+  The posterior prediction for the frequencies of the four genotypes over time.
+Each of the 20 curves is the average of 10,000 simulations of the single-locus model using parameters drawn from the posterior distribution.
   }
   \label{fig:ppc-plot}
-  % TODO put legend inside figure - I think there is room in middle right
 \end{figure}
@@
   \caption{
-    \textbf{Likelihood profiles.} Sensitivity of the approximate likelihood, $\likelihood(\theta)$, of the single-locus model to changing a single parameter while the other parameters remain fixed at their MAP estimates.
+    \textbf{Likelihood profiles.} Sensitivity of the approximate likelihood of the single-locus model, $\likelihood(\theta)$, to changing a single parameter while the other parameters remain fixed at their MAP estimates.
 Dashed vertical line represents the MAP value.
   }
-  % TODO add 10^-10, not 10^-2
-  % TODO ylabel - L shouldn't have a subscript !, this was used only for the model without aneuploidy
-  % TODO add more points in the peaks
 
   \label{fig:sensitivity}
+  % TODO what defines the x range? is it the prior distribution range?
 \end{figure}

### 2022-01-24 ms/ms.tex
small revisions fig S4

@@
       \includegraphics[width=0.45\textwidth]{../figures/runs-A.pdf}      
-      \includegraphics[width=0.45\textwidth]{../figures/runs-B.pdf} \\     
+      \includegraphics[width=0.45\textwidth]{../figures/runs-B.pdf}    
       \includegraphics[width=0.325\textwidth]{../figures/runs-C.pdf}      
@@
     \textbf{Posterior distribution validation.}
-    The posterior distributions of the single-locus model parameters are roughly the same regardless of the number of simulations (4-10,000) used to approximate the likelihood.
+    The posterior distribution of the single-locus model parameters is roughly the same regardless of the number of simulations (4-10,000 replicates) used to approximate the likelihood (\autoref{eq:heatstress-likelihood}).
+    % TODO add "Density" to y label in C
     }

### 2022-01-24 ms.tex
added analysis results

--- 2022-01-24 ms/ms.tex
@@
 % TODO cite Yang, F. et al. The fitness costs and benefits of trisomy of each Candida albicans chromosome. Genetics 218, 1–7 (2021). https://academic.oup.com/genetics/article/218/2/iyab056/6218773
-% cite: https://www.cell.com/developmental-cell/fulltext/S1534-5807(21)00562-1?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS1534580721005621%3Fshowall%3Dtrue
-% cite: https://www.cell.com/developmental-cell/fulltext/S1534-5807(21)00592-X?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS153458072100592X%3Fshowall%3Dtrue
 
@@
       \centering
-      \includegraphics[height=1.8in]{../figures/Fig1-A.pdf}      
+      \includegraphics[height=1.8in]{figures/Fig1-A.pdf}      
       \label{fig:model1}
@@
       \centering
-      \includegraphics[height=1.8in]{../figures/Fig1-B.pdf}      
+      \includegraphics[height=1.8in]{figures/Fig1-B.pdf}      
       \label{fig:model2}
@@
   \centering
-\includegraphics[width=0.9\textwidth]{../figures/posterior.pdf}
+\includegraphics[width=0.9\textwidth]{figures/posterior.pdf}
   \caption{
   \textbf{Posterior distribution of single-locus model.}
-On the diagonal, the inferred posterior distribution of each model parameter. 
-Below the diagonal, the inferred joint posterior distribution of pairs of model parameters (dark purple and bright yellow for low and high density, respectively).
-Red markers and orange lines for the joint MAP estimate (which may differ from the marginal MAP, as the marginal distribution integrates over all other parameters).
+% TODO histogram on diagonal, 2d histogram without interpolation below diagonal
+% TODO mutation rate tick labels should be in 10^-6, not 10^-5, and write 10^-6 not 1e-5; also it looks like you get positive density for mutation rate = 0 which doesnt make sense - set limit to 0.
+% change trisomy rate to aneuploidy gain rate
+% remove "mutation rate" from y label on top-left
+% remove the y-ticks from the diagonal, they are meaningless, right?
+% Fitness of 2n+1, w_{2n+1} (not 2n+1 fitness)
+% TODO change dashed orange to red and make it a bit bolder.
+On the diagonal, the posterior kernel density estimate for each parameter. 
+Below the diagonal, joint posterior density of two parameters (dark purple and bright yellow for low and high density, respectively).
+Red markers and orange lines for the joint MAP estimates (which may differ from the marginal maximum density, as the marginal distribution integrates over all other parameters).
 } 
@@
       \centering
-      \includegraphics[width=\textwidth]{../figures/fixation-plot-a.pdf}      
+      \includegraphics[width=\textwidth]{figures/fixation-plot-a.pdf}      
       \label{fig:fit}
@@
       \centering
-      \includegraphics[width=\textwidth]{../figures/fixation-plot-b.pdf}      
+      \includegraphics[width=\textwidth]{figures/fixation-plot-b.pdf}      
       \label{fig:fit-cumulative}
@@
   \centering
-\includegraphics[width=0.8\textwidth]{../figures/dynamics.pdf}
+\includegraphics[width=0.8\textwidth]{figures/dynamics.pdf}
   \caption{
-  \textbf{Posterior genotype frequency dynamics for the single-locus model.} 
-  The posterior prediction for the frequencies of the four genotypes over time.
-Each of the 20 curves is the average of 10,000 simulations of the single-locus model using parameters drawn from the posterior distribution.
+  \textbf{Single-locus model posterior genotype frequency dynamics.} 
+  The posterior predicted frequency of the four genotypes. Each of 20 curves is the average of 10,000 simulations using parameters drawn from the posterior distribution (\autoref{fig:posterior}).
   }
   \label{fig:ppc-plot}
+  % TODO put legend inside figure - I think there is room in middle right
 \end{figure}
+
+\paragraph*{Analysis: single-locus model.}
+Here, we aim to analyse the timing of the single-locus model and the relative importance of the various parameters.
+As a stochastic model, there are two events that can occur from the wild type population.
+One is mutation that gives the adapted population state, or a missegregation event.
+The probability with which these events happen is hugely different.
+We are interested in the average waiting times until one of the two events occurs and lead to fixation.
+Let $T_1$ and $T_2$ denote the random variables for the waiting time for the events mutation and missegregation respectively.
+Due to discrete generation, they are geometrically distributed with parameter compound of the probability that at least one individual in this generation changes its state and goes to fixation:
+\begin{subequations}
+\begin{align}
+T_{2n^*}&\sim \text{Geo}(1-(1-2\mu \Delta w)^N)=\text{Geo}(p_{2n^*})\\
+T_{2n+1}&\sim \text{Geo}(1-(1-2\delta\Delta w)^N)=\text{Geo}(p_{2n+1})
+\end{align}
+\end{subequations}
+Here, $\Delta w$ denotes the relative difference between the fitness of the newly introduced type and the population mean fitness.
+However, since we have competing events, we need to find the minimum of these two processes.
+The minimum of two geometrically distributed processes with parameters $p_{2n^*}$ and $p_{2n+1}$ is again geometrically distributed with parameter $1-(1-p_{2n^*})(1-p_{2n+1})$.
+The average waiting time until the first event happens is the inverse of the paramter of the geometric distribution.
+The probability that aneuploidy will be the first event to happen, is given by
+\begin{equation}
+p_A=\frac{p_{2n+1}}{p_{2n^*}+p_{2n+1}}=\frac{1-(1-2(w_3-w_1)\delta)^N}{2-(1-2(w_3-w_1)\delta)^N-(1-2(w_2-w_1)\mu)^N}.
+\end{equation}
+Since we are interested in the average time to fixation of $2n^*$, the steps that are needed differ depending on the actual event that happens first.
+If mutation happens first, then $2n^*$ is already reached and everything is fine.
+On the other hand, if aneuploidy occurs first, the population is in state $2n+1$ and several other steps have to occur, until $2n^*$ is reached.
+Thus, in this aneuploid trajectory, we also need to add the average waiting times to reach $2n+1^*$ and finally $2n^*$ The formulas are written down straightforward.
+
+In general, we assume, that the waiting time until a new type occurs is much longer than the time the population needs to go to fixation for a certain type, but nonetheless, these times need to be included for a rigorous analysis.
+There are two ways, how to compute them and they differ in the assumption of the underlying process.
+
+ If the population evolves deterministically, then one can derive the formula
+\begin{equation}
+t_\text{fix, i}=2\frac{\log(N)}{\log(1+s)}
+\end{equation}
+ from the classical one-locus two-allele equations.
+As we work with simulations, we assume fixation if it reached 95\% of the total population size.
+This is significantly different from full fixation.
+Thus we need a slightely different formula
+ \begin{equation}
+ t_\text{fix, $f_c$, i}=\frac{\log\left(\frac{f_cN}{1-f_c}\right)}{\log(1+s)}.
+ \end{equation}
+ Here, $f_c$ denotes the fraction for which we assume fixation.
+
+In the second approach, the population is assumed to evolve stochastically. It is based on the diffusion equation and given in Kimura and Ohta 1969 (eq. 17). We denote it by $T_\text{fix, i}$.
+
+Thus, the average time to $2n^*$ over the aneuploidy trajectory is given by
+\begin{subequations}\label{T_aneu_mut}
+\begin{equation}
+T_\text{aneu}=(1-(1-p_{2n^*})(1-p_{2n+1}))^{-1}+p_{2n+1^*}^{-1}+p_{2n^*}^{-1}+T_\text{fix, 2n+1}+T_\text{fix, $2n+1^*$}+T_\text{fix, $2n^*$}.
+\end{equation}
+The average time to $2n^*$ over the mutant trajectory is given by 
+\begin{equation}
+T_\text{mut}=(1-(1-p_{2n^*})(1-p_{2n+1}))^{-1}+T_\text{fix, $2n^*$}.
+\end{equation}
+\end{subequations}
+\textcolor{red}{[MP: $T_\text{fix, $2n^*$}$ is not the same in (11) and (12). Need a better notation.]}
+
+In summary, the average time to $2n^*$ is given by
+\begin{equation}\label{T_total}
+T=p_AT_\text{aneu}+(1-p_A)T_\text{mut}.
+\end{equation}
+
+\begin{remark} There are several implicit assumptions that the equations hold like this.
+\begin{enumerate}
+\item The two processes fixation and occurrence of a new type are sequential and do not interfere with each other.
+\item There is no clonal interference.
+\item The fixation process starts from exactly one individual. 
+\end{enumerate}
+There are certainly combinations of the mutation rate, the missegregation rate, the selection strength and the population size for which at least one of these assumptions is violated.
+Thus, \eqref{T_aneu_mut} and \eqref{T_total} may not yield a good approximation to the actual time to the adapted state.
+\end{remark}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=.49\textwidth]{../Simulations/figinvpNonlymut0.pdf}
+\includegraphics[width=.49\textwidth]{../Simulations/figinvpmuonlymut0.pdf}
+\caption{The average time to $2n^*$ is shown as a function of the population size N in the left panel and as a function of the mutation rate in the right panel.
+The blue circles are the means of the simulation runs and the lines denote the span between the 5th and the 95th percentile.
+The orange line is the respective analytic approximation.
+In these two plots, the aneuploidy rate is zero, so only the mutational trajectory is available.}
+\label{fig:only mut}
+\end{figure}
+
+To get a feeling of the limitations of \eqref{T_total}, we plot simulations with $\delta=0$ in Fig.~\ref{fig:only mut}.
+The other paramters are similar to those infered above.
+Even there, thereare slight deviations for high paramter values.
+Clearly visible in the left panel and not so good to see in the right.
+We conjecture, that in this parameter range more than one individual mutates in one generation and our assumption for the fixation time, which starts from exactly one mutant, does not hold.
 
@@
     \centering
-	\includegraphics[width=\textwidth]{../figures/evo39_fitness_30deg.pdf}
+	\includegraphics[width=\textwidth]{figures/evo39_fitness_30deg.pdf}
     \label{fig:growth-curves-30deg}
@@
     \centering
-	\includegraphics[width=\textwidth]{../figures/evo39_fitness_39deg.pdf}
+	\includegraphics[width=\textwidth]{figures/evo39_fitness_39deg.pdf}
     \label{fig:growth-curves-39deg}
@@
   \begin{subfigure}{0.3\textwidth}
-      \includegraphics[width=\textwidth]{../figures/sensitivity-A.pdf}      
+      \includegraphics[width=\textwidth]{figures/sensitivity-A.pdf}      
       \label{fig:sensitivity-mutation}
@@
   \begin{subfigure}{0.3\textwidth}
-      \includegraphics[width=\textwidth]{../figures/sensitivity-B.pdf}      
+      \includegraphics[width=\textwidth]{figures/sensitivity-B.pdf}      
       \label{fig:sensitivity-aneuploidy}
@@
    \begin{subfigure}{0.3\textwidth}
-      \includegraphics[width=\textwidth]{../figures/sensitivity-C.pdf}      
+      \includegraphics[width=\textwidth]{figures/sensitivity-C.pdf}      
       \label{fig:sensitivity-anwt}
@@
     \begin{subfigure}{0.3\textwidth}
-      \includegraphics[width=\textwidth]{../figures/sensitivity-D.pdf}      
+      \includegraphics[width=\textwidth]{figures/sensitivity-D.pdf}      
       \label{fig:sensitivity-anmt}
@@
     \begin{subfigure}{0.3\textwidth}
-      \includegraphics[width=\textwidth]{../figures/sensitivity-E.pdf}      
+      \includegraphics[width=\textwidth]{figures/sensitivity-E.pdf}      
       \label{fig:sensitivity-eumt}
@@
   \caption{
-    \textbf{Likelihood profiles.} Sensitivity of the approximate likelihood of the single-locus model, $\likelihood(\theta)$, to changing a single parameter while the other parameters remain fixed at their MAP estimates.
+    \textbf{Likelihood profiles.} Sensitivity of the approximate likelihood, $\likelihood(\theta)$, of the single-locus model to changing a single parameter while the other parameters remain fixed at their MAP estimates.
 Dashed vertical line represents the MAP value.
   }
+  % TODO add 10^-10, not 10^-2
+  % TODO ylabel - L shouldn't have a subscript !, this was used only for the model without aneuploidy
+  % TODO add more points in the peaks
 
   \label{fig:sensitivity}
-  % TODO what defines the x range? is it the prior distribution range?
 \end{figure}
@@
     \centering
-      \includegraphics[width=0.45\textwidth]{../figures/convergence-p1_mr.pdf}      
-      \includegraphics[width=0.45\textwidth]{../figures/convergence-p2_tr.pdf} \\     
-      \includegraphics[width=0.325\textwidth]{../figures/convergence-p3_w1.pdf}      
-      \includegraphics[width=0.325\textwidth]{../figures/convergence-p4_w2.pdf}      
-      \includegraphics[width=0.325\textwidth]{../figures/convergence-p5_w3.pdf}      
+      \includegraphics[width=0.45\textwidth]{figures/convergence-p1_mr.pdf}      
+      \includegraphics[width=0.45\textwidth]{figures/convergence-p2_tr.pdf} \\     
+      \includegraphics[width=0.325\textwidth]{figures/convergence-p3_w1.pdf}      
+      \includegraphics[width=0.325\textwidth]{figures/convergence-p4_w2.pdf}      
+      \includegraphics[width=0.325\textwidth]{figures/convergence-p5_w3.pdf}      
      \label{fig:convergence-A}
@@
   \begin{subfigure}{1\textwidth}
-  	\includegraphics[width=\textwidth]{../figures/ess.pdf}
+  	\includegraphics[width=\textwidth]{figures/ess.pdf}
      \label{fig:convergence-B}
@@
     \centering
-	\includegraphics[width=0.6\textwidth]{../figures/seeds.pdf}
+	\includegraphics[width=0.6\textwidth]{figures/seeds.pdf}
     \caption{

### 2022-01-24 ms/ms.tex
added tau comparison plots

@@
 We also tried to infer parameters of the single-locus model when the mutation rate in aneuploid cells, compared to euploid cells, is increased by a factor $\tau$=1, 33/32 (due to an additional chromosome), 2, 5, 10 or 100 (due to genetic instability).
-We found that for $\tau=$33/32, 2, and 5 the posterior was not significantly different from the model with $\tau=1$, while for $\tau=10$ the estimated mutation rate was slightly lower, $\mu=1.654\substack{+0.879 \\ -1.775}\times10^{-6}$, and for $\tau=100$ the estimated mutation rate was significantly lower and the aneuploidy rate had higher variability.
+We found that for $\tau=$33/32, 2, and 5 the posterior was not significantly different from the model with $\tau=1$, while for $\tau=10$ the estimated mutation rate was slightly lower, $\mu=1.654\substack{+0.879 \\ -1.775}\times10^{-6}$, and for $\tau=100$ the estimated mutation rate was significantly lower and the aneuploidy rate had higher variability (\autoref{fig:tau}).
 The MAP and 50\% HDI with $\tau=100$ were:
-%TODO add to MAP-and-hdi.ipynb and check the values here
-$\mu=5.538\substack{+1.611 \\ -5.311}\times10^{-7}$,
-$\delta=8.292\substack{+8.988 \\ -4.96}\times10^{-4}$,
-$w_{2n+1}=1.023\substack{+0.001 \\ -0.002}$,
+%% generated with MAP-and-hdi.ipynb
+$\mu=3.807\substack{+2.919 \\ -3.311}\times10^{-7}$,
+$\delta=7.829\substack{+10.683 \\ -3.555}\times10^{-4}$,
+$w_{2n+1}=1.022\substack{+0.001 \\ -0.001}$,
 $w_{2n+1^*}=1.024\substack{+0.001 \\ -0.001}$,
 $w_{2n^*}=1.028\substack{+0.001 \\ -0.002}$.
-We thus continued our analysis under the assumption that $\tau=1$, that is, that the mutation rate in euploid and aneuploid cells is the same.
-%TODO: show joint distributions comparison plot (20-diffk.ipynb). WAIC?
-% TODO support figure of joint posterior delta vs mu per tau=1, 33/32, 2, 5, 10
+We thus continued our analysis under the assumption that $\tau=1$, that is, that the mutation rate is euploid and aneuploid cells are the same.
+%TODO WAIC for all tau except 100 is about 0.268-0.269 and for tau=100 0.271. I don't think that we should provide it, because the difference is minor.
 
@@
   \label{fig:ppc-plot}
+\end{figure}
+
+% Fig tau comparisons
+%% generated with diff-tau.ipynb
+\begin{figure}[h!]
+  \centering
+  \begin{subfigure}{0.45\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-A.pdf}      
+  \end{subfigure}
+  \begin{subfigure}{0.45\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-B.pdf}      
+  \end{subfigure}
+  \\
+   \begin{subfigure}{0.325\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-C.pdf}      
+  \end{subfigure}
+    \begin{subfigure}{0.325\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-D.pdf}      
+  \end{subfigure}
+    \begin{subfigure}{0.325\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-E.pdf}      
+  \end{subfigure}
+  \\
+  \begin{subfigure}{0.325\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-joint-F.pdf}      
+  \end{subfigure}
+\begin{subfigure}{0.325\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-joint-G.pdf}      
+  \end{subfigure}
+\begin{subfigure}{0.325\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-joint-H.pdf}      
+  \end{subfigure}
+\begin{subfigure}{0.325\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-joint-I.pdf}      
+  \end{subfigure}
+\begin{subfigure}{0.325\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-joint-J.pdf}      
+  \end{subfigure}
+\begin{subfigure}{0.325\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-joint-K.pdf}      
+  \end{subfigure}
+  \caption{
+    \textbf{Single-locus model with different $\tau$ comparison.}  \textbf{(A-E)} The inferred posterior distribution of each model paramter for the single-locus model with different $\tau$.
+For $\tau=10$ and $\tau=100$ the mutation and aneuploidy rates tends to be lower.
+\textbf{(F-K)} The inferred joint posterior distribution of mutation rate and aneuploidy rate for the single locus model with different $\tau$ (dark purple and bright yellow for low and high density, respectively).
+  \label{fig:tau}
+  }
   \end{figure}
@@
   % TODO what defines the x range? is it the prior distribution range?
+  %ANS: yes for mutation and aneuploidy rates. For fitness just around the MAP in range that looks beautiful.
 \end{figure}

### 2022-01-26 ms/ms.tex
.
updated growth curves plots

@@
 We sampled 10,000 samples to use as a prior distribution~(\autoref{fig:growth-curves}).
-We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$. In lieu of a better prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$. 
+We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$ ~(\autoref{fig:growth-curves}-A).
+In lieu of a better prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$.
 To increase computational efficiency, we also assumed $w(2n*)>w(2n+1*)>w(2n+1)>w(2n)$; running the inference without this assumption produced similar results.
@@
 % TODO what are the results ?
-We also tried to use growth curves of \anwt\; and \euwt\; in \SI{30}{\celsius} to estimate $1-c$.
+We also tried to use growth curves of \anwt\; and \euwt\; in \SI{30}{\celsius} ~(\autoref{fig:growth-curves}-B) to estimate $1-c$ .
 This estimation assumes that the cost of aneuploidy is the same in \SI{39}{\celsius} and \SI{30}{\celsius}; this might be incorrect, but we only assumed this to generate a prior distribution for the fitness values.
@@
 We also tried to infer parameters of the single-locus model when the mutation rate in aneuploid cells, compared to euploid cells, is increased by a factor $\tau$=1, 33/32 (due to an additional chromosome), 2, 5, 10 or 100 (due to genetic instability).
-We found that for $\tau=$33/32, 2, and 5 the posterior was not significantly different from the model with $\tau=1$, while for $\tau=10$ the estimated mutation rate was slightly lower, $\mu=1.654\substack{+0.879 \\ -1.775}\times10^{-6}$, and for $\tau=100$ the estimated mutation rate was significantly lower and the aneuploidy rate had higher variability (\autoref{fig:tau}).
+We found that for $\tau=$33/32, 2, and 5 the posterior was not significantly different from the model with $\tau=1$, while for $\tau=10$ the estimated mutation rate was slightly lower, $\mu=1.654\substack{+0.879 \\ -1.775}\times10^{-6}$, and for $\tau=100$ the estimated mutation rate was significantly lower and the aneuploidy rate had higher variability ~(\autoref{fig:tau}).
 The MAP and 50\% HDI with $\tau=100$ were:
@@
 $w_{2n^*}=1.028\substack{+0.001 \\ -0.002}$.
+WAIC for the models with each $\tau$ is similar: about 0.268-0.269, and only for $\tau=100$, 0.271.
 We thus continued our analysis under the assumption that $\tau=1$, that is, that the mutation rate is euploid and aneuploid cells are the same.
-%TODO WAIC for all tau except 100 is about 0.268-0.269 and for tau=100 0.271. I don't think that we should provide it, because the difference is minor.
 
@@
 
+% Fig Curveball 
 \begin{figure}[h]
-% Fig Curveball 30deg
-% generated with growth_curves.ipynb
-\begin{subfigure}{0.5\textwidth}
 	\centering
-	\includegraphics[width=\textwidth]{../figures/evo39_fitness_30deg.pdf}
-    \label{fig:growth-curves-30deg}
-\end{subfigure}
-% Fig Curveball 39deg
-% generated with growth_curves.ipynb
-\begin{subfigure}{0.5\textwidth}
-    \centering
-	\includegraphics[width=\textwidth]{../figures/evo39_fitness_39deg.pdf}
-    \label{fig:growth-curves-39deg}
-\end{subfigure}
+	% generated with growth_curves.ipynb and deg = '39' 
+	\includegraphics[width=0.45\textwidth]{../figures/evo39_fitness_39deg.pdf}
+	% generated with growth_curves.ipynb and deg = '30' 
+	\includegraphics[width=0.45\textwidth]{../figures/evo39_fitness_30deg.pdf}
+	% generated with growth_curves_refined.ipynb and deg = '30' 
+	\includegraphics[width=0.45\textwidth]{../figures/refined_vs_evo39_fitness_39deg.pdf}
 \caption{
     \textbf{Fitness estimation from growth curves.}
-    \textbf{(A)} Fitness estimation from growth curves collected at \SI{30}.{\celsius}
-    \textbf{(B)} Fitness estimation from growth curves collected at \SI{39}.{\celsius}
+    \textbf{(A)} Fitness estimation from growth curves collected at \SI{39}{\celsius}. Mean: 1.024; 95\% CI: 0.959 - 1.115.
+    \textbf{(B)} Fitness estimation from growth curves collected at \SI{30}{\celsius}. Mean: 0.986;  95\% CI: 0.979 - 0.994.
+    \textbf{(C)} Fitness estimation from growth curves collected at \SI{39}{\celsius}. Mean: 1.033; 95\% CI: 1.027 - 1.041.
+    \\
     Growth curves previously described in \citet[Figs. 3C, 4A, and S2]{Yona2012}.
@@
 \end{figure}
-%TODO please report here the point estimate and 95% CI for the three distributions, I only have for 1-c+b: 0.9863 (95% CI 0.9790 - 0.9939).
-% TODO add panel labels
 
@@
     The posterior distribution of the single-locus model parameters is roughly the same regardless of the number of simulations (4-10,000 replicates) used to approximate the likelihood (\autoref{eq:heatstress-likelihood}).
-    % TODO add "Density" to y label in C
     }

### 2022-01-26 ms/ms.tex
revise fig S1 caption, added fig S1 todos

@@
 	% generated with growth_curves.ipynb and deg = '39' 
-	\includegraphics[width=0.45\textwidth]{../figures/evo39_fitness_39deg.pdf}
+	\includegraphics[width=0.45\textwidth]{../figures/evo39_fitness_39deg.pdf} \\
 	% generated with growth_curves.ipynb and deg = '30' 
-	\includegraphics[width=0.45\textwidth]{../figures/evo39_fitness_30deg.pdf}
+	\includegraphics[width=0.45\textwidth]{../figures/evo39_fitness_30deg.pdf} \\
 	% generated with growth_curves_refined.ipynb and deg = '30' 
@@
     \textbf{Fitness estimation from growth curves.}
-    \textbf{(A)} Fitness estimation from growth curves collected at \SI{39}{\celsius}. Mean: 1.024; 95\% CI: 0.959 - 1.115.
-    \textbf{(B)} Fitness estimation from growth curves collected at \SI{30}{\celsius}. Mean: 0.986;  95\% CI: 0.979 - 0.994.
-    \textbf{(C)} Fitness estimation from growth curves collected at \SI{39}{\celsius}. Mean: 1.033; 95\% CI: 1.027 - 1.041.
-    \\
+    % you need to give each panel a label so that you will have A-L; have A-D on same row, then E-H same row, I-L on same row;
+    % caption should explain what each panel shows.
+    \textbf{(A)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{39}{\celsius}: $\hat w_{2n+1}/w_{2n}$=1.024 (95\% CI: 0.959 - 1.115).
+    \textbf{(B)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{30}{\celsius}: $\hat w_{2n+1}/w_{2n}$= 0.986 (95\% CI: 0.979 - 0.994).
+    \textbf{(C)} Fitness estimation from growth curves of \anwt\ and \eumt\ at \SI{39}{\celsius}: $\hat w_{2n*}/w_{2n+1}$=1.033 (95\% CI: 1.027 - 1.041).
     Growth curves previously described in \citet[Figs. 3C, 4A, and S2]{Yona2012}.

### 2022-01-26 ms/ms.tex
mode tau figure to supp, revised relevant text and caption, added todo

@@
 
-We also tried to infer parameters of the single-locus model when the mutation rate in aneuploid cells, compared to euploid cells, is increased by a factor $\tau$=1, 33/32 (due to an additional chromosome), 2, 5, 10 or 100 (due to genetic instability).
-We found that for $\tau=$33/32, 2, and 5 the posterior was not significantly different from the model with $\tau=1$, while for $\tau=10$ the estimated mutation rate was slightly lower, $\mu=1.654\substack{+0.879 \\ -1.775}\times10^{-6}$, and for $\tau=100$ the estimated mutation rate was significantly lower and the aneuploidy rate had higher variability ~(\autoref{fig:tau}).
-The MAP and 50\% HDI with $\tau=100$ were:
-%% generated with MAP-and-hdi.ipynb
-$\mu=3.807\substack{+2.919 \\ -3.311}\times10^{-7}$,
-$\delta=7.829\substack{+10.683 \\ -3.555}\times10^{-4}$,
-$w_{2n+1}=1.022\substack{+0.001 \\ -0.001}$,
-$w_{2n+1^*}=1.024\substack{+0.001 \\ -0.001}$,
-$w_{2n^*}=1.028\substack{+0.001 \\ -0.002}$.
-WAIC for the models with each $\tau$ is similar: about 0.268-0.269, and only for $\tau=100$, 0.271.
-We thus continued our analysis under the assumption that $\tau=1$, that is, that the mutation rate is euploid and aneuploid cells are the same.
-
 % FIG posterior
@@
 Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\autoref{fig:ppc-plot}). % TODO move somewhere else?
-Furthermore, sensitivity analysis shows that changing the parameter values from the MAP estimate reduces the model fit to the experimental results~(\autoref{fig:sensitivity}).
+Furthermore, sensitivity analysis shows that changing a single parameter while keeping the rest fixed at the MAP estimate produces a worse fit to the data~(\autoref{fig:sensitivity}).
 
@@
 % TODO WAIC?
+
+We also inferred model parameters under the assumptuon that the mutation rate in aneuploid cells, compared to euploid cells, is increased by a factor $\tau$=1, 33/32 (due to an additional chromosome), 2, 5, 10 or 100 (due to genetic instability).
+We found that for the posterior distribution was similar for $\tau=$1, 33/32, 2, and 5 (\autoref{fig:tau}; WAIC 0.268-0.269).
+With $\tau=10$, the estimated mutation rate was slightly lower ($\mu=1.654\substack{+0.879 \\ -1.775}\times10^{-6}$). 
+With $\tau=100$, the estimated mutation rate was considerably lower ($\mu=3.807\substack{+2.919 \\ -3.311}\times10^{-7}$) and the aneuploidy rate had higher variance ($\delta=7.829\substack{+10.683 \\ -3.555}\times10^{-4}$).
+The model fit was worst in this case (WAIC=0.271).
+%The MAP and 50\% HDI with $\tau=100$ were: 
+%% generated with MAP-and-hdi.ipynb
+%$\mu=3.807\substack{+2.919 \\ -3.311}\times10^{-7}$,
+%$\delta=7.829\substack{+10.683 \\ -3.555}\times10^{-4}$,
+%$w_{2n+1}=1.022\substack{+0.001 \\ -0.001}$,
+%$w_{2n+1^*}=1.024\substack{+0.001 \\ -0.001}$,
+%$w_{2n^*}=1.028\substack{+0.001 \\ -0.002}$.
+%WAIC for the models with each $\tau$ is similar: about 0.268-0.269, and only for $\tau=100$, 0.271.
+Therefore, we cannot rule out an increased mutation rate in aneuploid cells, but we find no evidence that it does, and unless the effect is strong, it does not seem to affect our results.
 
@@
   \label{fig:ppc-plot}
-\end{figure}
-
-% Fig tau comparisons
-%% generated with diff-tau.ipynb
-\begin{figure}[h!]
-  \centering
-  \begin{subfigure}{0.45\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-A.pdf}      
-  \end{subfigure}
-  \begin{subfigure}{0.45\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-B.pdf}      
-  \end{subfigure}
-  \\
-   \begin{subfigure}{0.325\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-C.pdf}      
-  \end{subfigure}
-    \begin{subfigure}{0.325\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-D.pdf}      
-  \end{subfigure}
-    \begin{subfigure}{0.325\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-E.pdf}      
-  \end{subfigure}
-  \\
-  \begin{subfigure}{0.325\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-joint-F.pdf}      
-  \end{subfigure}
-\begin{subfigure}{0.325\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-joint-G.pdf}      
-  \end{subfigure}
-\begin{subfigure}{0.325\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-joint-H.pdf}      
-  \end{subfigure}
-\begin{subfigure}{0.325\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-joint-I.pdf}      
-  \end{subfigure}
-\begin{subfigure}{0.325\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-joint-J.pdf}      
-  \end{subfigure}
-\begin{subfigure}{0.325\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-joint-K.pdf}      
-  \end{subfigure}
-  \caption{
-    \textbf{Single-locus model with different $\tau$ comparison.}  \textbf{(A-E)} The inferred posterior distribution of each model paramter for the single-locus model with different $\tau$.
-For $\tau=10$ and $\tau=100$ the mutation and aneuploidy rates tends to be lower.
-\textbf{(F-K)} The inferred joint posterior distribution of mutation rate and aneuploidy rate for the single locus model with different $\tau$ (dark purple and bright yellow for low and high density, respectively).
-  \label{fig:tau}
-  }
 \end{figure}
@@
 
+ % Fig tau comparisons
+%% generated with diff-tau.ipynb
+\begin{figure}[h!]
+  \centering
+  \begin{subfigure}{0.45\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-A.pdf}      
+  \end{subfigure}
+  \begin{subfigure}{0.45\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-B.pdf}      
+  \end{subfigure}
+  \\
+   \begin{subfigure}{0.325\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-C.pdf}      
+  \end{subfigure}
+    \begin{subfigure}{0.325\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-D.pdf}      
+  \end{subfigure}
+    \begin{subfigure}{0.325\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-E.pdf}      
+  \end{subfigure}
+  \\
+  \begin{subfigure}{0.325\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-joint-F.pdf}      
+  \end{subfigure}
+\begin{subfigure}{0.325\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-joint-G.pdf}      
+  \end{subfigure}
+\begin{subfigure}{0.325\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-joint-H.pdf}      
+  \end{subfigure}
+\begin{subfigure}{0.325\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-joint-I.pdf}      
+  \end{subfigure}
+\begin{subfigure}{0.325\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-joint-J.pdf}      
+  \end{subfigure}
+\begin{subfigure}{0.325\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-joint-K.pdf}      
+  \end{subfigure}
+  \caption{
+    \textbf{Single-locus model inference with elevated mutation rate in aneuploid cells.}  \textbf{(A-E)} The inferred posterior distributions of model parameters for the single-locus model with different values of $\tau$, the fold-increase in mutation rate in aneuploid cells (\anwt\ and \anmt).
+When the increase in mutation rate is high, $\tau=10$ and $\tau=100$, the inferred mutation (A) and aneuploidy (B) rate tends to be lower.
+    \textbf{(F-K)} The inferred joint posterior distribution of mutation rate ($\mu$) and aneuploidy rate ($\delta$) with different $\tau$ values (dark purple and bright yellow for low and high density, respectively).
+  \label{fig:tau}
+  }
+  % TODO maybe added dashed vertical lines with the same colors to show where the MAP of each distribution is?
+
+  \end{figure}
+
 \end{document}

### 2022-01-26 ms/ms.tex
added MP text from a analysis branch and revised it, added todos

@@
 \usepackage{gensymb,siunitx}
+\usepackage{cleveref}
 
@@
 We focus on beneficial mutations, neglecting the effects of deleterious and neutral mutations. Both models allow for a single aneuploid karyotype (e.g., chromosome III duplication).
-While the single-locus model allows for only a single mutation to accumulate in the genotype, the multi-locus model allows for multiple mutations to accumulate~(\autoref{fig:models}), as well as for a fluctuating population size.
+While the single-locus model allows for only a single mutation to accumulate in the genotype, the multi-locus model allows for multiple mutations to accumulate~(\cref{fig:models}), as well as for a fluctuating population size.
 
@@
 \paragraph*{Single-locus model.}
-This model assumes a constant effective population size $N$ and follows four genotypes~(\autoref{fig:models}A): euploid wild-type, \euwt, the initial genotype; 
+This model assumes a constant effective population size $N$ and follows four genotypes~(\cref{fig:models}A): euploid wild-type, \euwt, the initial genotype; 
 euploid mutant, \eumt, with the standard karyotype and a single beneficial mutation; 
@@
 
-Transitions between the genotypes occur as follows~(\autoref{fig:models}A): Beneficial mutations from \euwt\; to \eumt\; and from \anwt\; to \anmt\; occur with probability $\mu$, the mutation rate.
+Transitions between the genotypes occur as follows~(\cref{fig:models}A): Beneficial mutations from \euwt\; to \eumt\; and from \anwt\; to \anmt\; occur with probability $\mu$, the mutation rate.
 We neglect back-mutations (i.e., from \eumt\; to \euwt\; and from \anmt\; to \anwt).
@@
 That is, we assume chromosomes are gained and lost at the same rate, and we neglect events that form a less-fit genotype (i.e., \anwt\; to \euwt\; and \eumt\; to \anmt).
-The fitness values of the four genotypes are given by \autoref{table:single-locus}.
+The fitness values of the four genotypes are given by \cref{table:single-locus}.
 
@@
     \end{equation}
-where the fitness values $w_i$ are given in \autoref{table:single-locus} and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
+where the fitness values $w_i$ are given in \cref{table:single-locus} and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
 The effect of mutation and aneuploidy on genotype frequencies is given by
@@
 
-The fitness of the different genotypes is the same as in the single-locus model (\autoref{table:single-locus}), except that the fitness contribution of $k$ beneficial mutations is the product of their independent effects, $\prod_{i=1}^k(1+s_i)$, instead of the contribution of the single mutation allowed in the single-locus model, $(1+s)$, see \autoref{table:multi-locus}.
+The fitness of the different genotypes is the same as in the single-locus model (\cref{table:single-locus}), except that the fitness contribution of $k$ beneficial mutations is the product of their independent effects, $\prod_{i=1}^k(1+s_i)$, instead of the contribution of the single mutation allowed in the single-locus model, $(1+s)$, see \cref{table:multi-locus}.
 Therefore, aneuploidy loss would be favored by selection only if there are enough beneficial mutations and/or the selection coefficients $s_i$ are large enough.
@@
 
-The change in frequencies due to selection is exactly the same as in the single-locus model (\autoref{eq:selection-single}), only applied using the fitness values in \autoref{table:multi-locus}.
-The change due to random genetic drift is also the same as in \autoref{eq:drift-single}, except that the frequencies vector is $\vec{f}=(f_{2n}\ ,\ f_{2n+1}\ ,\ f_{2n^*}\ ,\ f_{2n+1^*}\ ,\ f_{2n^{**}}\ ,\ f_{2n+1^{**}}\ ,\ \ldots)$ and that the population size changes between generations, as described above.
+The change in frequencies due to selection is exactly the same as in the single-locus model (\cref{eq:selection-single}), only applied using the fitness values in \cref{table:multi-locus}.
+The change due to random genetic drift is also the same as in \cref{eq:drift-single}, except that the frequencies vector is $\vec{f}=(f_{2n}\ ,\ f_{2n+1}\ ,\ f_{2n^*}\ ,\ f_{2n+1^*}\ ,\ f_{2n^{**}}\ ,\ f_{2n+1^{**}}\ ,\ \ldots)$ and that the population size changes between generations, as described above.
 
@@
 For example, $P_{\tilde{\vec X}}\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)$ is approximated by taking simulations in which \anwt\; fixed before generation 450 but not before generation 200, and computing the fraction of such simulations in which \eumt\; did not fix by generation 1,700, and hence aneuploidy did not extinct before generation 1,700.
-\autoref{fig:seeds} compares results with less and more simulated experiments, demonstrating that 1,000 simulations are likely enough.
+\cref{fig:seeds} compares results with less and more simulated experiments, demonstrating that 1,000 simulations are likely enough.
 
@@
 The proposal particle is then perturbed using a kernel perturbation kernel, $K_t(\theta^* \mid \theta)$ where $\theta$ is the sample from the previous iteration.
-Then, a set of synthetic observations $\tilde{\vec X}^*$ is simulated, and the proposal particle $\theta^*$ is accepted if its approximate likelihood (\autoref{eq:heatstress-likelihood}) is high enough, $\likelihood(\theta^*)>1-\epsilon_t$ (or more commonly, if $1-\likelihood(\theta^*) < \epsilon_t$), where $\epsilon_t>0$ is the \emph{acceptance threshold}, as higher values of $\epsilon_t$ allow more particles to be accepted.
+Then, a set of synthetic observations $\tilde{\vec X}^*$ is simulated, and the proposal particle $\theta^*$ is accepted if its approximate likelihood (\cref{eq:heatstress-likelihood}) is high enough, $\likelihood(\theta^*)>1-\epsilon_t$ (or more commonly, if $1-\likelihood(\theta^*) < \epsilon_t$), where $\epsilon_t>0$ is the \emph{acceptance threshold}, as higher values of $\epsilon_t$ allow more particles to be accepted.
 The acceptance threshold $\epsilon_t$ is chosen as the median of the $1-\likelihood(\theta)$ of the particles accepted in the previous iteration, $t-1$, and $\epsilon_0=0.01$. 
@@
 
-Acceptance is determined according to the approximate likelihood~(\autoref{eq:heatstress-likelihood}), which has a maximum value of 0.875.
-Thus, we terminated the inference when $\epsilon \le 0.13$ after six iterations, with $n_6=982$ accepted parameter vectors and effective sample size ESS=651 (\autoref{fig:convergence}).
+Acceptance is determined according to the approximate likelihood~(\cref{eq:heatstress-likelihood}), which has a maximum value of 0.875.
+Thus, we terminated the inference when $\epsilon \le 0.13$ after six iterations, with $n_6=982$ accepted parameter vectors and effective sample size ESS=651 (\cref{fig:convergence}).
 % TODO for multi-locus model we need to say something else
-Running the inference algorithm with different initialization seeds and less or more simulations for approximating the likelihood produced similar posterior distributions~(\autoref{fig:seeds}).
+Running the inference algorithm with different initialization seeds and less or more simulations for approximating the likelihood produced similar posterior distributions~(\cref{fig:seeds}).
 %TODO for multi-locus model, write about them here also.
@@
 Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix.
-We sampled 10,000 samples to use as a prior distribution~(\autoref{fig:growth-curves}).
-We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$ ~(\autoref{fig:growth-curves}-A).
+We sampled 10,000 samples to use as a prior distribution~(\cref{fig:growth-curves}).
+We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$ ~(\cref{fig:growth-curves}-A).
 In lieu of a better prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$.
@@
 % TODO what are the results ?
-We also tried to use growth curves of \anwt\; and \euwt\; in \SI{30}{\celsius} ~(\autoref{fig:growth-curves}-B) to estimate $1-c$ .
+We also tried to use growth curves of \anwt\; and \euwt\; in \SI{30}{\celsius} ~(\cref{fig:growth-curves}-B) to estimate $1-c$ .
 This estimation assumes that the cost of aneuploidy is the same in \SI{39}{\celsius} and \SI{30}{\celsius}; this might be incorrect, but we only assumed this to generate a prior distribution for the fitness values.
@@
 We used ABC-SMC to infer the posterior distribution of the parameters of the single-locus model.
-The MAP (maximum a posteriori) and 50\% HDI (highest density interval) of the parameters posterior are~(\autoref{fig:posterior}): %% generated by MAP-and-hdi.ipynb
+The MAP (maximum a posteriori) and 50\% HDI (highest density interval) of the parameters posterior are~(\cref{fig:posterior}): %% generated by MAP-and-hdi.ipynb
 mutation rate, $\mu=2.93\substack{+1.122 \\ -2.824}\times10^{-6}$,
@@
 The mutation rate corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014}. 
-%We can see from the posterior distribution~(\autoref{fig:posterior}) the correlation between fitness of different genotypes. 
+%We can see from the posterior distribution~(\cref{fig:posterior}) the correlation between fitness of different genotypes. 
 % TODO say/plot something about the correlation: which fitness values are correlated? how strong is it? why does it occur?
@@
 \paragraph{Model checking and comparison.}
-The single-locus model fits the data well: in simulations using the best-fit parameters (MAP estimate) \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\autoref{fig:2n*-fixation}-B) and \anwt\ fixed in roughly 300 generations on average (green lines in \autoref{fig:ppc-plot}), which agrees with experimental results.
+The single-locus model fits the data well: in simulations using the best-fit parameters (MAP estimate) \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\cref{fig:2n*-fixation}-B) and \anwt\ fixed in roughly 300 generations on average (green lines in \cref{fig:ppc-plot}), which agrees with experimental results.
 % TODO replace "roughly 300" with the exact number.
-Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\autoref{fig:ppc-plot}). % TODO move somewhere else?
-Furthermore, sensitivity analysis shows that changing a single parameter while keeping the rest fixed at the MAP estimate produces a worse fit to the data~(\autoref{fig:sensitivity}).
+Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\cref{fig:ppc-plot}). % TODO move somewhere else?
+Furthermore, sensitivity analysis shows that changing a single parameter while keeping the rest fixed at the MAP estimate produces a worse fit to the data~(\cref{fig:sensitivity}).
 
@@
 We can see that the mutation rate estimate in this model is much lower than in the model with aneuploidy.
-Higher mutation rate would cause quicker fixation of \eumt\ than in the experimental observations (\autoref{fig:2n*-fixation}).
+Higher mutation rate would cause quicker fixation of \eumt\ than in the experimental observations (\cref{fig:2n*-fixation}).
 Thus, even disregarding the appearance of trisomic cells in the experiment, the evidence supports the inclusion of aneuploidy in the model.
@@
 We also inferred model parameters under the assumptuon that the mutation rate in aneuploid cells, compared to euploid cells, is increased by a factor $\tau$=1, 33/32 (due to an additional chromosome), 2, 5, 10 or 100 (due to genetic instability).
-We found that for the posterior distribution was similar for $\tau=$1, 33/32, 2, and 5 (\autoref{fig:tau}; WAIC 0.268-0.269).
+We found that for the posterior distribution was similar for $\tau=$1, 33/32, 2, and 5 (\cref{fig:tau}; WAIC 0.268-0.269).
 With $\tau=10$, the estimated mutation rate was slightly lower ($\mu=1.654\substack{+0.879 \\ -1.775}\times10^{-6}$). 
@@
 	Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy in blue and green, respectively, and corresponding to the blue and green bars in panel A.
-The MAP likelihood (\autoref{eq:heatstress-likelihood}) is 0.84 and 0.67 with and without aneuploidy, respectively.
+The MAP likelihood (\cref{eq:heatstress-likelihood}) is 0.84 and 0.67 with and without aneuploidy, respectively.
      }
@@
 
-\paragraph{Inference: multi-locus model.}
+\pagebreak
+%%%%%%%%%%
+\subsection*{Analysis: single-locus model.}
+
+Here, we aim to approximate the adaptation time (i.e., time to fixation of \eumt) in the single-locus model and how it is affected by the model parameters.
+As a stochastic model, there are two events that can occur in the \euwt\ population.
+The first is a mutation that produces a \eumt\ cell.
+The second is a missegregation event that produces a \anwt\ cell.
+The probability with which these events occur is significantly different.
+We are interested in the average waiting time until one of the two events occurs and then lead to adaptation (in the second case, it will be followed by a mutation to \anmt\ and loss of aneuploidy to \eumt).
+We assume the waiting time for appearance of \eumt\ from \euwt\ and appearance of \anwt\ from \euwt are geometrically distributed with rates $p_{2n^*}$ and $p_{2n+1}$, respectively, where
+\begin{subequations}
+\begin{align}
+p_{2n^*} &= 1-(1-2\mu \Delta w_{2n^*})^N \;, \\
+p_{2n+1} &= 1-(1-2\delta\Delta w_{2n+1})^N \;,
+\end{align}
+\end{subequations}
+where $\Delta w_i$ denotes the relative difference between the fitness of genotype $i$ and the population mean fitness.
+However, we have two competing events: if one event occurs (e.g., mutation) then the other (aneuploidy) is unlikely to occur,  we need to find the minimum of these waiting times.
+The minimum of two geometric random variables with parameters $p_{2n^*}$ and $p_{2n+1}$ is also a geometric random variable with parameter $1-(1-p_{2n^*})(1-p_{2n+1})$, which is the inverse of the expected waiting time until the first event occurs.
+The probability that aneuploidy will be the first event to occur is
+\begin{equation}
+p_A = \frac{p_{2n+1}}{p_{2n^*}+p_{2n+1}} \;.
+\end{equation}
+We are interested in the expected time to fixation of \eumt, and the events leading to this fixation depend on the event that occurs first.
+If \eumt\ appears first and avoids extinction by drift, then it will fix.
+If \anwt\ appears first and avoids extinction, then several other events must occur for fixation of \eumt.
+Thus, in this aneuploid trajectory, we need to approximate the expected waiting time to reach \anmt\ and finally \eumt.
+In general, we assume that the waiting time until a new genotype occurs is much longer than the time to its fixation.
+Nonetheless, these fixation times must be included in a rigorous analysis.
+There are two ways to compute the fixation times and they differ in the assumption of the underlying process.
+
+A deterministic-process approach is given by the classical bi-allelic single-locus equation, which provides an approximation for the expected time for genotype $i$ to reach frequency $x$ when starting from frequency $1-x$ in a population of genotype $j$ is % TODO MP is this correct?
+ % TODO MP ref for equation? Crow and Kimura?
+\begin{equation}
+\tilde T_{i \to j} = 2\frac{\log{\big(\frac{x}{1-x}N\big)}}{\log(w_i/w_j)} \;,
+\end{equation}
+A stochastic-process approach for approximating $\tilde T_i$ is given by the diffusion-equation approximation \citep[eq.~17]{Kimura1969a}. 
+Thus, the expected adaptation time via the aneuploidy trajectory is given by
+\begin{equation}\label{eq:T_aneu}
+ T_A = \frac{1}{1-(1-p_{2n^*})(1-p_{2n+1})} + \frac{1}{p_{2n+1^*}} + \frac{1}{p_{2n^*}} + \tilde T_{2n \to 2n+1} + \tilde T_{2n+1 \to 2n+1^*} + \tilde T_{2n+1^* \to 2n^*} \;,
+ \end{equation}
+the expected adaptation time via the mutation trajectory is given by 
+\begin{equation}\label{eq:T_mut}
+T_{M}=\frac{1}{1-(1-p_{2n^*})(1-p_{2n+1})} + \tilde T_{2n \to 2n^*} \;,
+\end{equation}
+and the expected adaptation time is given by
+ \begin{equation}\label{eq:T_total}
+ T = p_A T_{A}+(1-p_A) T_{M}.
+ \end{equation}
+
+% TODO MP - the following remark should be removed.
+Instead, wherever you use one of these assumptions in the above text, you should explicitly say so (rather then say the assumptions are implicit).
+Then you can discuss here how these assumptions may not be met when the parameters are so and so, and that you will point it out in the results or something like that.
+ %\begin{remark} There are several implicit assumptions that the equations hold like this.
+ %\begin{enumerate}
+ %\item The two processes fixation and occurrence of a new type are sequential and do not interfere with each other.
+ %\item There is no clonal interference.
+ %\item The fixation process starts from exactly one individual. 
+ %\end{enumerate}
+ %There are certainly combinations of the mutation rate, the aneuploidy rate, the selection strength and the population size for which at least one of these assumptions is violated.
+Thus, \cref{eq:T_total} may not yield a good approximation to the actual time to the adapted state.
+ %\end{remark}
+
+To validate the approximation in \cref{eq:T_total}, we performed simulations of the full model (\cref{eq:mutation-aneuploidy-single,eq:drift-single,eq:selection-single})  without aneuploidy, $\delta=0$ (\cref{fig:only mut}), and with the other parameters as inferred above.
+The approximation provides a good fit to simulation results except when the effective population size is larger than $5\cdot 10^6$ cells.
+In that case, the expected number of new mutants is ~0.5, and therefore we can expect that in many generations the fixation event will start with more than a single \eumt\ cell appearing in the same generation, which is in contrast to our assumption that an adaptation event starts from a single mutant cell.
+
+ \begin{figure}[h!]
+ \centering
+ \includegraphics[width=.49\textwidth]{../figures/figinvpNonlymut0.pdf}
+ \includegraphics[width=.49\textwidth]{../figures/figinvpmuonlymut0.pdf}
+ \caption{\textbf{Expected adaptation time without aneuploidy.}
+  The expected time to appearance and fixation of \eumt\ as a function of the population size, $N$ (panel A) and as a function of the mutation rate, $\mu$ (panel B).
+Mean and 95\% CI from simulations in blue.
+Analytic approximation (\cref{eq:T_total}) in orange.
+Here, XXXX.}
+  %TODO MP add panel labels A and B
+  %replace XXXX with parameter values and remove them from the figure titles.
+  %y label "Expected adaptation time"
+  % remove legend
+  % make it a single figure with two panels, then there is no need for y label on panel B
+  % The error bars a quite large, can you make them smaller by running more simulations?
+  % Add more x ticks for panel A (10^6, 5 10 ^6 ...)
+  % maybe start x axis in panel A with lower population size? 10^4 or 10^5 ? does the approximation work well there?
+ \label{fig:only mut}
+ \end{figure}
+
+%\subsection*{Inference: multi-locus model.}
 
@@
     \textbf{Posterior distribution validation.}
-    The posterior distribution of the single-locus model parameters is roughly the same regardless of the number of simulations (4-10,000 replicates) used to approximate the likelihood (\autoref{eq:heatstress-likelihood}).
+    The posterior distribution of the single-locus model parameters is roughly the same regardless of the number of simulations (4-10,000 replicates) used to approximate the likelihood (\cref{eq:heatstress-likelihood}).
     }

### 2022-01-26 ms/ms.tex
updated growth curves plots

@@
 Compared to other priors we tested, this prior produced lower WAIC, better posterior prediction plots, and more stable parameter estimates, as follows.
-%In both of these uninformative priors, the fitness estimates were much too high: for the first prior the median estimates were  $w_{2n+1}=2.767$, $w_{2n+1^*}=5.79$, $w_{2n^*}=5.827$ and for the second prior the estimates were $w_{2n+1}=1.046$, $w_{2n+1^*}=4.642$, $w_{2n^*}=4.684$.
-Such high fitness values are unreasonable.
-Moreover the WAIC for the models with these priors were 0.57 and 1.9, respectively, whereas the WAIC with the informative prior was 0.27.
-% TODO if you need to remove this sentence: Also, with the uninformative prior for all fitness values, the fixation time of \anwt is about 16 on average, much lower than the (unpublished) experimental result, which is 200-300 generations.
 We tried to use additional growth curves.
@@
 With $\tau=10$, the estimated mutation rate was slightly lower ($\mu=1.654\substack{+0.879 \\ -1.775}\times10^{-6}$). 
+%% generated with MAP-and-hdi.ipynb
 With $\tau=100$, the estimated mutation rate was considerably lower ($\mu=3.807\substack{+2.919 \\ -3.311}\times10^{-7}$) and the aneuploidy rate had higher variance ($\delta=7.829\substack{+10.683 \\ -3.555}\times10^{-4}$).
 The model fit was worst in this case (WAIC=0.271).
-%The MAP and 50\% HDI with $\tau=100$ were: 
-%% generated with MAP-and-hdi.ipynb
-%$\mu=3.807\substack{+2.919 \\ -3.311}\times10^{-7}$,
-%$\delta=7.829\substack{+10.683 \\ -3.555}\times10^{-4}$,
-%$w_{2n+1}=1.022\substack{+0.001 \\ -0.001}$,
-%$w_{2n+1^*}=1.024\substack{+0.001 \\ -0.001}$,
-%$w_{2n^*}=1.028\substack{+0.001 \\ -0.002}$.
-%WAIC for the models with each $\tau$ is similar: about 0.268-0.269, and only for $\tau=100$, 0.271.
 Therefore, we cannot rule out an increased mutation rate in aneuploid cells, but we find no evidence that it does, and unless the effect is strong, it does not seem to affect our results.
@@
 	% generated with growth_curves.ipynb and deg = '39' 
-	\includegraphics[width=0.45\textwidth]{../figures/evo39_fitness_39deg.pdf} \\
+	\includegraphics[width=.95\textwidth]{../figures/evo39_fitness_39deg.pdf} 
 	% generated with growth_curves.ipynb and deg = '30' 
-	\includegraphics[width=0.45\textwidth]{../figures/evo39_fitness_30deg.pdf} \\
+	\includegraphics[width=0.95\textwidth]{../figures/evo39_fitness_30deg.pdf} 
 	% generated with growth_curves_refined.ipynb and deg = '30' 
-	\includegraphics[width=0.45\textwidth]{../figures/refined_vs_evo39_fitness_39deg.pdf}
+	\includegraphics[width=0.95\textwidth]{../figures/refined_vs_evo39_fitness_39deg.pdf}
 \caption{
     \textbf{Fitness estimation from growth curves.}
-    % you need to give each panel a label so that you will have A-L; have A-D on same row, then E-H same row, I-L on same row;
-    % caption should explain what each panel shows.
-    \textbf{(A)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{39}{\celsius}: $\hat w_{2n+1}/w_{2n}$=1.024 (95\% CI: 0.959 - 1.115).
-    \textbf{(B)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{30}{\celsius}: $\hat w_{2n+1}/w_{2n}$= 0.986 (95\% CI: 0.979 - 0.994).
-    \textbf{(C)} Fitness estimation from growth curves of \anwt\ and \eumt\ at \SI{39}{\celsius}: $\hat w_{2n*}/w_{2n+1}$=1.033 (95\% CI: 1.027 - 1.041).
+    % TODO caption should explain what each panel shows.
+    \textbf{(A-D)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{39}{\celsius}: $\hat w_{2n+1}/w_{2n}$=1.024 (95\% CI: 0.959 - 1.115).
+    \textbf{(E-H)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{30}{\celsius}: $\hat w_{2n+1}/w_{2n}$= 0.986 (95\% CI: 0.979 - 0.994).
+    \textbf{(I-M)} Fitness estimation from growth curves of \anwt\ and \eumt\ at \SI{39}{\celsius}: $\hat w_{2n*}/w_{2n+1}$=1.033 (95\% CI: 1.027 - 1.041).\\
     Growth curves previously described in \citet[Figs. 3C, 4A, and S2]{Yona2012}.

### 2022-01-31 ms/ms.tex
small fixes

@@
     \textbf{(E-H)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{30}{\celsius}: $\hat w_{2n+1}/w_{2n}$= 0.986 (95\% CI: 0.979 - 0.994).
-    \textbf{(I-M)} Fitness estimation from growth curves of \anwt\ and \eumt\ at \SI{39}{\celsius}: $\hat w_{2n*}/w_{2n+1}$=1.033 (95\% CI: 1.027 - 1.041).\\
+    \textbf{(I-L)} Fitness estimation from growth curves of \anwt\ and \eumt\ at \SI{39}{\celsius}: $\hat w_{2n*}/w_{2n+1}$=1.033 (95\% CI: 1.027 - 1.041).\\
     Growth curves previously described in \citet[Figs. 3C, 4A, and S2]{Yona2012}.

### 2022-02-21 ms/ms.tex
clean up todos
mended some TODOs
update MP todos
results regarding only the aneuploidy trajectory

@@
 We analyze these evolutionary-genetic models to estimate the effects of parameters on the adaptation time and the probability for adaptation via aneuploidy.
-We find that \textbf{TODO} % TODO
+We find that \textbf{} % TODO
 
@@
 WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy~\citep{Kass1995}.
-% TODO Describe posterior plots comparison method - Fig 4 & S3
+% TODO IK Describe posterior plots comparison method - Fig 4 & S3
 
@@
 This prior resulted in WAIC 0.32, compared to 0.27 with the above informative prior.
-% TODO what are the results ?
 We also tried to use growth curves of \anwt\; and \euwt\; in \SI{30}{\celsius} ~(\cref{fig:growth-curves}-B) to estimate $1-c$ .
@@
 This prior did not converge: no parameter sets were found with an approximate likelihood greater than zero.
-% TODO it make sense because bla bla (to explain)
+% TODO IK it makes sense because bla bla (to explain)
 Finally, we also tested a uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ and (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$.
@@
 $w_{2n^*}=1.028\substack{+0.001 \\ -0.001}$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$.
-% TODO we are jumping between b c s and w2n* w2n+1 w2n+1*. we should report all models and results either with b c s or with w. you can compute b c s from w before computing the HDIs.
+% TODO IK we are jumping between b c s and w2n* w2n+1 w2n+1*. we should report all models and results either with b c s or with w. you can compute b c s from w before computing the HDIs.
 This estimated aneuploidy rate agrees with previous estimates. % TODO refs
@@
 %We can see from the posterior distribution~(\cref{fig:posterior}) the correlation between fitness of different genotypes. 
-% TODO say/plot something about the correlation: which fitness values are correlated? how strong is it? why does it occur?
+% TODO IK say/plot something about the correlation: which fitness values are correlated? how strong is it? why does it occur?
 
@@
 The single-locus model fits the data well: in simulations using the best-fit parameters (MAP estimate) \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\cref{fig:2n*-fixation}-B) and \anwt\ fixed in roughly 300 generations on average (green lines in \cref{fig:ppc-plot}), which agrees with experimental results.
-% TODO replace "roughly 300" with the exact number.
+% TODO IK replace "roughly 300" with the exact number.
 Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\cref{fig:ppc-plot}). % TODO move somewhere else?
@@
 Thus, even disregarding the appearance of trisomic cells in the experiment, the evidence supports the inclusion of aneuploidy in the model.
-% TODO WAIC?
+% TODO IK WAIC?
 
@@
 We are interested in the average waiting time until one of the two events occurs and then lead to adaptation (in the second case, it will be followed by a mutation to \anmt\ and loss of aneuploidy to \eumt).
-We assume the waiting time for appearance of \eumt\ from \euwt\ and appearance of \anwt\ from \euwt are geometrically distributed with rates $p_{2n^*}$ and $p_{2n+1}$, respectively, where
+We assume the waiting time for appearance of \eumt\ from \euwt\ and appearance of \anwt\ from \euwt\ are geometrically distributed with rates $p_{2n^*}$ and $p_{2n+1}$, respectively, where
 \begin{subequations}
@@
 where $\Delta w_i$ denotes the relative difference between the fitness of genotype $i$ and the population mean fitness.
-However, we have two competing events: if one event occurs (e.g., mutation) then the other (aneuploidy) is unlikely to occur,  we need to find the minimum of these waiting times.
+However, we have two competing events: if one event occurs (e.g., mutation), then the other (aneuploidy) is unlikely to occur,  we need to find the minimum of these waiting times.
 The minimum of two geometric random variables with parameters $p_{2n^*}$ and $p_{2n+1}$ is also a geometric random variable with parameter $1-(1-p_{2n^*})(1-p_{2n+1})$, which is the inverse of the expected waiting time until the first event occurs.
@@
 \end{equation}
-We are interested in the expected time to fixation of \eumt, and the events leading to this fixation depend on the event that occurs first.
+We are interested in the expected time to fixation of \eumt, and the trajectory leading to this fixation depend on the event that occurs first.
 If \eumt\ appears first and avoids extinction by drift, then it will fix.
@@
  % TODO MP ref for equation? Crow and Kimura?
-\begin{equation}
+\begin{equation} \label{eq:T_fix_x}
 \tilde T_{i \to j} = 2\frac{\log{\big(\frac{x}{1-x}N\big)}}{\log(w_i/w_j)} \;,
@@
 
-To validate the approximation in \cref{eq:T_total}, we performed simulations of the full model (\cref{eq:mutation-aneuploidy-single,eq:drift-single,eq:selection-single})  without aneuploidy, $\delta=0$ (\cref{fig:only mut}), and with the other parameters as inferred above.
+As a first step, to validate the approximation in \cref{eq:T_total}, we performed simulations of the full model (\cref{eq:mutation-aneuploidy-single,eq:drift-single,eq:selection-single})  without aneuploidy, $\delta=0$ (\cref{fig:only mut}), and with the other parameters as inferred above.
 The approximation provides a good fit to simulation results except when the effective population size is larger than $5\cdot 10^6$ cells.
@@
  \centering
- \includegraphics[width=.49\textwidth]{../figures/figinvpNonlymut0.pdf}
- \includegraphics[width=.49\textwidth]{../figures/figinvpmuonlymut0.pdf}
+ \includegraphics[width=\textwidth]{../figures/figgrid_noaneu.pdf}
  \caption{\textbf{Expected adaptation time without aneuploidy.}
@@
 Analytic approximation (\cref{eq:T_total}) in orange.
-Here, XXXX.}
+Here, $w_\anwt=1.021,\,   w_\anmt=1.025$ and $w_\eumt=1.028$.
+In panel A, $\mu=9.6\cdot 10^{-8}$ and in panel B, $N=6\cdot 10^6$.
+  }
   %TODO MP add panel labels A and B
@@
  \end{figure}
+
+ \begin{figure}[h!]
+ %\centering
+ \includegraphics[width=1\textwidth]{../figures/figgrid_nodirectmut_ylog_uncorr0.jpg}
+ \caption{\textbf{Expected adaptation time without mutations from $\euwt$ to $\eumt$.}
+  The expected time to appearance and fixation of \eumt\ as a function of the population size, $N$ (panel A) and as a function of the mutation rate, $\mu$ (panel B) and as a function of the aneuploidy rate, $\delta$ (panel C).
+Mean and 95\% CI from simulations in blue.
+Analytic approximation (\cref{eq:T_total}) in green.
+The red curve is achieved by setting $\tilde{T}_{\euwt \rightarrow \anwt}=\tilde{T}_{\anwt \rightarrow \anmt}=0$.
+The orange line is achieved by setting $x=0.017$ in \cref{eq:T_fix_x} for the fixation time of $\eumt$, such that it is well fitting in panel B.
+Here, $w_\anwt=1.021,\,   w_\anmt=1.025$ and $w_\eumt=1.028$.
+Parameter values in panels where it is not varied, $N=6\cdot 10^6$,\, $\mu=9.6\cdot 10^{-8}$ and $\delta=5.4\cdot 10^{-4}$.
+}
+ \label{fig:no direct mut}
+ \end{figure}
+
+ Second, we ran simulations without state changes from $\euwt$ to $\eumt$.
+Before, we had the optimal scenario of waiting until the first successful mutation appears and then waiting until is reaches fixation.
+Our approximations did quite well there.
+Here, the aneuploid states are transient and fixation could be outpaced by the arrival of the next type.
+This would explain the discrepancy between the green curves in \cref{fig:no direct mut}.
+To asses this further, \cref{fig:fixT vs waitT} provides a clear picture.
+At least for intermediate to high values of the respective range of the parameters, the waiting time until a succesful new type appears is far smaller than that to fixation of the old type.
+Taking this into account, it seems reasonable to ignore the fixation time of $\anwt$ and $\anmt$, which results in the better fitting red curves in \cref{fig:no direct mut}.
+Inferring $x=0.017$ for $\tilde{T}_{\anmt \rightarrow \eumt}$ to get the best fit in panel B, does not imply that it is also a best fit for panels A and C.
+However, we stop simulations in state $\eumt$ if they have frequency $0.95$.
+
+  The discrepancies in \cref{fig:no direct mut}C seem to be particularly odd as the approximations are nearly constant and the simulations decreas log-linearly with increasing $\delta$.
+
+ \begin{figure}[h!]
+ %\centering
+ \includegraphics[width=1\textwidth]{../figures/figgrid_fixT_vs_waitT.jpg}
+ \caption{\textbf{Expected time of fixation and waiting time events on the aneuploid trajectory.}
+  The expected time to appearance and fixation of each step seperately in the aneuploid trajectory as a function of the population size, $N$ (panel A) and as a function of the mutation rate, $\mu$ (panel B) and as a function of the aneuploidy rate, $\delta$ (panel C).
+Solid curves give the fixation time and dashed curves the waiting time.
+Curves of the same color indicate the same base population state;  e.g.
+the blue dashed curve is the waiting time in state $\anwt$ until a succesful individual of type $\anmt$ appears.
+The corresponding solid curve indicates the time to fixation of the population in state $\anwt$.
+Here, $w_\anwt=1.021,\,   w_\anmt=1.025$ and $w_\eumt=1.028$.
+Parameter values in panels where it is not varied, $N=6\cdot 10^6$,\, $\mu=9.6\cdot 10^{-8}$ and $\delta=5.4\cdot 10^{-4}$.
+}
+ \label{fig:fixT vs waitT}
+ \end{figure}
+
+In summary, whereas the two possible trajectories to $\eumt$ do not compete against each other in both \cref{fig:only mut} and \cref{fig:no direct mut}, the approximations are doing well only in restricted parameter ranges.
+These tend to be, where the respective parameters are rather small.
+If they grow bigger, they are violating some assumptions of our approximations.
 
@@
     \textbf{Fitness estimation from growth curves.}
-    % TODO caption should explain what each panel shows.
+    % TODO IK caption should explain what each panel shows.
     \textbf{(A-D)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{39}{\celsius}: $\hat w_{2n+1}/w_{2n}$=1.024 (95\% CI: 0.959 - 1.115).
@@
 Dashed vertical line represents the MAP value.
+The prior distributions for the mutation rate and aneuploidy rate are $\mu \sim U(10^{-9}, 10^{-5})$ and $\delta \sim U(10^{-6}, 10^{-2})$, respectively.
   }
@@
   \label{fig:sensitivity}
-  % TODO what defines the x range? is it the prior distribution range?
-  %ANS: yes for mutation and aneuploidy rates. For fitness just around the MAP in range that looks beautiful.
 \end{figure}
@@
   }
-  % TODO maybe added dashed vertical lines with the same colors to show where the MAP of each distribution is?
+  % TODO IK maybe add dashed vertical lines with the same colors to show where the MAP of each distribution is?
 

### 2022-02-21 ms/ms.tex
changed hdi format and fixed values. Started to write about WAIC

--- 2022-01-31 ms/ms.tex
@@
 \usepackage[unicode=true,colorlinks=true,urlcolor=CadetBlue,citecolor=black,linkcolor=black]{hyperref}
+\usepackage{booktabs}
+\usepackage{pgfplotstable} 
 \PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
@@
 where $\theta$ is a parameter vector, and $\mathbb{E}[\cdot]$ and $\mathbb{V}[\cdot]$ are the expectation and variance taken over the posterior distribution.
+We base the calculation on 50,000 samples from the posterior KDE.
 WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy~\citep{Kass1995}.
-% TODO Describe posterior plots comparison method - Fig 4 & S3
+% TODO IK Describe posterior plots comparison method - Fig 4 & S3
 
@@
 For the mutation rate, $\mu$, and aneuploidy rate, $\delta$, we used uninformative uniform priors, $\mu \sim \mathit{U}(10^{-9}\ ,\ 10^{-5})$ and $\delta \sim \mathit{U}(10^{-6} , 10^{-2})$.
+We also tried wider mutation rate prior $\mu \sim \mathit{U}(10^{-9}\ ,\ 10^{-5})$ and got similar inference.
 For a model without aneuploidy ($\delta=0$), we used a wider uniform prior for the mutation rate, $\mu \sim \mathit{U}(10^{-10},10^{-5})$.
@@
 % Almost all done from here already
-%\paragraph{Simple model: Inference.}
-%tell point estimates (+- hdi) (fig 2 from poster, add hdi)
-%posterior predictions, sensitivity analysis
-%time to fixation of aneuploidy 2n+1 - fits unpublished data
-%figures
-%- posterior distributions
-%- expected frequency over time (fig 1 from poster)
-%- % simulations to reach 2n+1, 2n*. (fig 3 from poster)
-%- time to 2n* cannot be explained without aneuploidy (fig 4 from poster)
-%- sensitivity analysis (fig 5 from poster)
-%- convergence/ess of pyabc? 
-%  - plot of posterior in each iteration shows convergence; 
-%  - robustness of posterior to initial conditions - MAP+-HDI for different seeds;
 
@@
 The MAP (maximum a posteriori) and 50\% HDI (highest density interval) of the parameters posterior are~(\cref{fig:posterior}): %% generated by MAP-and-hdi.ipynb
-mutation rate, $\mu=2.93\substack{+1.122 \\ -2.824}\times10^{-6}$,
-aneuploidy rate, $\delta=1.716\substack{+0.974 \\ -0.384}\times10^{-3}$,
-and fitness values, $w_{2n+1}=1.022\substack{+0.001 \\ -0.002}$,
-$w_{2n+1^*}=1.025\substack{+0.001 \\ -0.001}$,
-$w_{2n^*}=1.028\substack{+0.001 \\ -0.001}$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$.
-% TODO we are jumping between b c s and w2n* w2n+1 w2n+1*. we should report all models and results either with b c s or with w. you can compute b c s from w before computing the HDIs.
-This estimated aneuploidy rate agrees with previous estimates. % TODO refs
-The mutation rate corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014}. 
+mutation rate, $\mu=2.942\times10^{-6} (2.017\times10^{-7}-4.15\times10^{-6})$,
+aneuploidy rate, $\delta=1.722\times10^{-3} (1.394\times10^{-3}-2.754\times10^{-3})$,
+and fitness values, $w_{2n+1}=1.022 (1.021-1.023)$,
+$w_{2n+1^*}=1.025 (1.024-1.026)$,
+$w_{2n^*}=1.028 (1.026-1.029)$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$.
+
+% TODO IK we are jumping between b c s and w2n* w2n+1 w2n+1*. we should report all models and results either with b c s or with w. you can compute b c s from w before computing the HDIs.
+This estimated aneuploidy rate agrees with previous estimates. % TODO refs (9.7*10-5 in Zhu 2014, so it's lower).
+The mutation rate corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014}. % 0.33 * 10^(-9) in Lynch 2008 
 %We can see from the posterior distribution~(\cref{fig:posterior}) the correlation between fitness of different genotypes. 
-% TODO say/plot something about the correlation: which fitness values are correlated? how strong is it? why does it occur?
+% TODO IK say/plot something about the correlation: which fitness values are correlated? how strong is it? why does it occur?
 
@@
 %% generated with MAP-and-hdi.ipynb
-mutation rate, $\mu=7.837\substack{+0.078 \\ -0.123}\times10^{-9}$,
-fitness of mutant, $w_{2n^*}=1.013\substack{+0.000 \\ -0.000}$. 
+mutation rate, $\mu=7.979\times10^{-9} (7.903\times10^{-9}-8.135\times10^{-9})$,
+fitness of mutaion, $w_{2n^*}=1.013 (1.012-1.013)$,
 We can see that the mutation rate estimate in this model is much lower than in the model with aneuploidy.
@@
 Thus, even disregarding the appearance of trisomic cells in the experiment, the evidence supports the inclusion of aneuploidy in the model.
-% TODO WAIC?
 
+Next we tried the model with aneuploidy and fixed mutation rate $\mu$.
+Hence $\mu$ is not well identifiable by the model (the inferred value has high variance with no strong correlation between it and other parameters), we tried the models where $\mu$=$10^{-5}$, $10^{-6}$ and $10^{-7}$.
+We saw that the inference posteriors were similar to the model with free $\mu$ parameter.
+However WAIC was better for the models with fixed $\mu$ ~(\autoref{table:WAIC}).
+Model with the $\mu$=$10^{-5}$ has the best WAIC, and better than the model without aneuploidy.
+
+% TODO IK WAIC for different tau
 We also inferred model parameters under the assumptuon that the mutation rate in aneuploid cells, compared to euploid cells, is increased by a factor $\tau$=1, 33/32 (due to an additional chromosome), 2, 5, 10 or 100 (due to genetic instability).
@@
 Analytic approximation (\cref{eq:T_total}) in orange.
-Here, XXXX.}
+Here, $w_\anwt=1.021,\,   w_\anmt=1.025$ and $w_\eumt=1.028$.
+In panel A, $\mu=9.6\cdot 10^{-8}$ and in panel B, $N=6\cdot 10^6$.
+  }
   %TODO MP add panel labels A and B
-  %replace XXXX with parameter values and remove them from the figure titles.
-  %y label "Expected adaptation time"
-  % remove legend
-  % make it a single figure with two panels, then there is no need for y label on panel B
   % The error bars a quite large, can you make them smaller by running more simulations?
@@
 
+%% generated with fixed-mutation-rate.ipynb
+\begin{table}[h]
+\centering
+\caption{
+\textbf{WAIC values for the different models.}}
+\pgfplotstabletypeset[
+    col sep=semicolon,
+    string type,
+    every head row/.style={before row=\hline,after row=\hline},
+    every last row/.style={after row=\hline},
+    ]{../figures/Table_WAIC.csv} 
+\caption*{
+	TODO caption, add diff tau models
+} 
+\label{table:WAIC}
+\end{table}
+
 \end{document}

### 2022-03-01 ms/ms.tex
revised some of recent IK changes and added todos

@@
 \end{equation}
-where $\theta$ is a parameter vector, and $\mathbb{E}[\cdot]$ and $\mathbb{V}[\cdot]$ are the expectation and variance taken over the posterior distribution.
-We base the calculation on 50,000 samples from the posterior KDE.
+where $\theta$ is a parameter vector, and $\mathbb{E}[\cdot]$ and $\mathbb{V}[\cdot]$ are the expectation and variance taken over the posterior distribution, which in practice are approximated using 50,000 samples from the posterior KDE.
 WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy~\citep{Kass1995}.
@@
 For the mutation rate, $\mu$, and aneuploidy rate, $\delta$, we used uninformative uniform priors, $\mu \sim \mathit{U}(10^{-9}\ ,\ 10^{-5})$ and $\delta \sim \mathit{U}(10^{-6} , 10^{-2})$.
-We also tried wider mutation rate prior $\mu \sim \mathit{U}(10^{-9}\ ,\ 10^{-5})$ and got similar inference.
+A wider mutation rate prior, $\mu \sim \mathit{U}(10^{-9}\ ,\ 10^{-5})$, produced similar results.
 For a model without aneuploidy ($\delta=0$), we used a wider uniform prior for the mutation rate, $\mu \sim \mathit{U}(10^{-10},10^{-5})$.
@@
 % TODO IK we are jumping between b c s and w2n* w2n+1 w2n+1*. we should report all models and results either with b c s or with w. you can compute b c s from w before computing the HDIs.
-This estimated aneuploidy rate agrees with previous estimates. % TODO refs (9.7*10-5 in Zhu 2014, so it's lower).
-The mutation rate corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014}. % 0.33 * 10^(-9) in Lynch 2008 
-%We can see from the posterior distribution~(\cref{fig:posterior}) the correlation between fitness of different genotypes. 
-% TODO IK say/plot something about the correlation: which fitness values are correlated? how strong is it? why does it occur?
+This estimated aneuploidy rate agrees with previous estimates. % TODO refs (9.7*10-5 in Zhu 2014, so it's lower). YR: we need more than one ref; also what is the units of Zho 2014?
+The mutation rate corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
+% TODO IK please add lynch2008 to bib file
 
@@
 %% generated with MAP-and-hdi.ipynb
-mutation rate, $\mu=7.979\times10^{-9} (7.903\times10^{-9}-8.135\times10^{-9})$,
-fitness of mutaion, $w_{2n^*}=1.013 (1.012-1.013)$,
-We can see that the mutation rate estimate in this model is much lower than in the model with aneuploidy.
-Higher mutation rate would cause quicker fixation of \eumt\ than in the experimental observations (\cref{fig:2n*-fixation}).
+mutation rate, $\mu=7.979\substack{+0.156 \\ -0.076}\times10^{-9}$,
+fitness of mutant, $w_{2n^*}=1.013\substack{+0 \\ -0.001}$,
+The estimated mutation rate without aneuploidy is much lower compared to a model with aneuploidy, suggesting a target size of just 40, because a higher mutation rate will lead to faster fixation of \eumt\ than in the experimental observations (\cref{fig:2n*-fixation}).
 Thus, even disregarding the appearance of trisomic cells in the experiment, the evidence supports the inclusion of aneuploidy in the model.
+% TODO IK why do you say that?
+spell it out please - WAIC is lower?
+likelihood is higher?
+ppc looks better?
+what?
 
-Next we tried the model with aneuploidy and fixed mutation rate $\mu$.
+% TODO YR Read this
+Next we tried the model with aneuploidy and a fixed mutation rate $\mu$.
 Hence $\mu$ is not well identifiable by the model (the inferred value has high variance with no strong correlation between it and other parameters), we tried the models where $\mu$=$10^{-5}$, $10^{-6}$ and $10^{-7}$.
@@
 % TODO IK WAIC for different tau
-We also inferred model parameters under the assumptuon that the mutation rate in aneuploid cells, compared to euploid cells, is increased by a factor $\tau$=1, 33/32 (due to an additional chromosome), 2, 5, 10 or 100 (due to genetic instability).
+We also inferred model parameters under the assumption that the mutation rate in aneuploid cells, compared to euploid cells, is increased by a factor $\tau$=1, 33/32 (due to an additional chromosome), 2, 5, 10 or 100 (due to genetic instability).
 We found that for the posterior distribution was similar for $\tau=$1, 33/32, 2, and 5 (\cref{fig:tau}; WAIC 0.268-0.269).
@@
 
+%TODO YR read this 
 %% generated with fixed-mutation-rate.ipynb
@@
 \caption*{
-	TODO caption, add diff tau models
+	%TODO IK caption, add diff tau models
 } 

### 2022-03-02 ms/ms.tex
more fixes, priors sections still under construction

@@
 We sampled 10,000 samples to use as a prior distribution~(\cref{fig:growth-curves}).
-We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$ ~(\cref{fig:growth-curves}-A).
+We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$ ~(\cref{fig:growth-curves}-D).
 In lieu of a better prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$.
@@
 Compared to other priors we tested, this prior produced lower WAIC, better posterior prediction plots, and more stable parameter estimates, as follows.
+%TODO IK what do you mean by better posterior prediction and is more stable parameter estimates still rellevant?
 We tried to use additional growth curves.
-We used growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$.
+We used growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$ ~(\cref{fig:growth-curves}-L).
 The same prior was used for $w_{2n*}/w_{2n+1*}$.
-This prior resulted in WAIC 0.32, compared to 0.27 with the above informative prior.
-We also tried to use growth curves of \anwt\; and \euwt\; in \SI{30}{\celsius} ~(\cref{fig:growth-curves}-B) to estimate $1-c$ .
+This prior resulted in WAIC 71.91, compared to 67.87 with the above informative prior.
+% generated with fixed-mutation-rate-and-WAICs.ipynb TODO IK update the WAIC with 1000 reps run
+We also tried to use growth curves of \anwt\; and \euwt\; in \SI{30}{\celsius} to estimate $1-c$ ~(\cref{fig:growth-curves}-H).
 This estimation assumes that the cost of aneuploidy is the same in \SI{39}{\celsius} and \SI{30}{\celsius}; this might be incorrect, but we only assumed this to generate a prior distribution for the fitness values.
@@
 This prior did not converge: no parameter sets were found with an approximate likelihood greater than zero.
+%TODO IK check that this model had no bugs
 % TODO IK it makes sense because bla bla (to explain)
@@
 For the mutation rate, $\mu$, and aneuploidy rate, $\delta$, we used uninformative uniform priors, $\mu \sim \mathit{U}(10^{-9}\ ,\ 10^{-5})$ and $\delta \sim \mathit{U}(10^{-6} , 10^{-2})$.
-A wider mutation rate prior, $\mu \sim \mathit{U}(10^{-9}\ ,\ 10^{-5})$, produced similar results.
-For a model without aneuploidy ($\delta=0$), we used a wider uniform prior for the mutation rate, $\mu \sim \mathit{U}(10^{-10},10^{-5})$.
+A wider mutation rate prior, $\mu \sim \mathit{U}(10^{-9}\ ,\ 10^{-3})$, produced similar results.
+% TODO IK, maybe wider mutation rate should be tried on a model without aneuploidy
+%For a model without aneuploidy ($\delta=0$), we used a wider uniform prior for the mutation rate, $\mu \sim \mathit{U}(10^{-10},10^{-5})$.
 
@@
 \subsection*{Inference: single-locus model}
-
-% Almost all done from here already
 
@@
 \paragraph{Model checking and comparison.}
-The single-locus model fits the data well: in simulations using the best-fit parameters (MAP estimate) \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\cref{fig:2n*-fixation}-B) and \anwt\ fixed in roughly 300 generations on average (green lines in \cref{fig:ppc-plot}), which agrees with experimental results.
-% TODO IK replace "roughly 300" with the exact number.
+The single-locus model fits the data well: in simulations using the best-fit parameters (MAP estimate) \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\cref{fig:2n*-fixation}-B),
 Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\cref{fig:ppc-plot}). % TODO move somewhere else?
@@
 
-% TODO YR Read this
-Next we tried the model with aneuploidy and a fixed mutation rate $\mu$.
-Hence $\mu$ is not well identifiable by the model (the inferred value has high variance with no strong correlation between it and other parameters), we tried the models where $\mu$=$10^{-5}$, $10^{-6}$ and $10^{-7}$.
-We saw that the inference posteriors were similar to the model with free $\mu$ parameter.
-However WAIC was better for the models with fixed $\mu$ ~(\autoref{table:WAIC}).
-Model with the $\mu$=$10^{-5}$ has the best WAIC, and better than the model without aneuploidy.
-
 % TODO IK WAIC for different tau
@@
 Therefore, we cannot rule out an increased mutation rate in aneuploid cells, but we find no evidence that it does, and unless the effect is strong, it does not seem to affect our results.
+
+We also inferred model parameters under the assumption that the mutation rate in aneuploid cells, compared to euploid cells, is increased by a factor $\tau$=1, 33/32 (due to an additional chromosome), 2, 5, 10 or 100 (due to genetic instability).
+We found that the posterior distribution was similar for $\tau=$1, 33/32, 2, and 5 (\cref{fig:tau}).
+%% generated with MAP-and-hdi.ipynb
+With $\tau=10$, the estimated mutation rate was slightly lower ($\mu=1.674\times10^{-6} (2.501\times10^{-7}-1.741\times10^{-6})$). 
+With $\tau=100$, the estimated mutation rate was considerably lower ($\mu=3.81\times10^{-7} (1.508\times10^{-7}-4.995\times10^{-7})$) and the aneuploidy rate had higher variance ($\delta=7.824\times10^{-4} (6.616\times10^{-4}-1.462\times10^{-3})$).
+%TODO YR to reference to WAIC table also? By this table tau=100 is the best model
+We cannot rule out an increased mutation rate in aneuploid cells, but we find no evidence that it does, and unless the effect is strong, it does not seem to affect our results.
+We also checked the differences in fixation/loss times of  \euwt, \anwt and \eumt\  for the models with different $\tau$.
+We saw that the models can be distinguished using these parameters, especially $\tau=100$ ~(\cref{fig:tau-plots}).
+
+% TODO YR Read this
+Next we tried the model with fixed mutation rate $\mu$.
+Hence $\mu$ is not well identifiable by the model (the inferred value has high variance with no strong correlation between it and other parameters), we tried the models where mutation rate is fixed: $\mu$=$10^{-5}$, $10^{-6}$ and $10^{-7}$.
+We observed similar parameters inference comparing to the model with free $\mu$ parameter.
+However WAIC was better for the models with fixed $\mu$ ~(\autoref{table:WAIC}).
+The lowest WAIC got the model with $\mu$=$10^{-5}$ - 0.6, it is lower than \emph{no aneuploidy} model - 0.87, that implies higher predictive accuracy.
 
@@
   \label{fig:ppc-plot}
+\end{figure}
+
+% Fig tau plots
+%% generated by diff-tau-plots.ipynb
+\begin{figure}[h!]
+  \centering
+  \begin{subfigure}{0.75\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-a.pdf}      
+  \end{subfigure}
+  \begin{subfigure}{0.75\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-b.pdf}      
+  \end{subfigure}
+  \\
+   \begin{subfigure}{0.75\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-c.pdf}      
+  \end{subfigure}
+   \caption{
+    \textbf{Genotype fixations for single-locus models with different $\tau$.} For each model, $10,000$ simulations using MAP parameters were evaluated and a percentage of simulations with fixated genotypes  \euwt, \anwt\ and \eumt\ (plots A, B and C respectively) were measured in time.
+\anmt\ didn't rich fixation, therefore it doesn't appear in figures.
+We can see that $\tau=100$ can be distinguished using data of loss of \euwt\ (plot A) and fixation/loss of \anwt\ (plot B).
+It is harder to distinguish $1\leq\tau\leq10$, however the correlation exists.
+  }
+  \label{fig:tau-plots}
 \end{figure}

### 2022-03-02 ms/ms.tex
small fix

@@
 
-% TODO IK WAIC for different tau
-We also inferred model parameters under the assumption that the mutation rate in aneuploid cells, compared to euploid cells, is increased by a factor $\tau$=1, 33/32 (due to an additional chromosome), 2, 5, 10 or 100 (due to genetic instability).
-We found that for the posterior distribution was similar for $\tau=$1, 33/32, 2, and 5 (\cref{fig:tau}; WAIC 0.268-0.269).
-With $\tau=10$, the estimated mutation rate was slightly lower ($\mu=1.654\substack{+0.879 \\ -1.775}\times10^{-6}$). 
-%% generated with MAP-and-hdi.ipynb
-With $\tau=100$, the estimated mutation rate was considerably lower ($\mu=3.807\substack{+2.919 \\ -3.311}\times10^{-7}$) and the aneuploidy rate had higher variance ($\delta=7.829\substack{+10.683 \\ -3.555}\times10^{-4}$).
-The model fit was worst in this case (WAIC=0.271).
-Therefore, we cannot rule out an increased mutation rate in aneuploid cells, but we find no evidence that it does, and unless the effect is strong, it does not seem to affect our results.
-
 We also inferred model parameters under the assumption that the mutation rate in aneuploid cells, compared to euploid cells, is increased by a factor $\tau$=1, 33/32 (due to an additional chromosome), 2, 5, 10 or 100 (due to genetic instability).

### 2022-03-04 ms/ms.tex
removed information prior that was not converged, because there were bug in it execution

@@
 % generated with fixed-mutation-rate-and-WAICs.ipynb TODO IK update the WAIC with 1000 reps run
-We also tried to use growth curves of \anwt\; and \euwt\; in \SI{30}{\celsius} to estimate $1-c$ ~(\cref{fig:growth-curves}-H).
-This estimation assumes that the cost of aneuploidy is the same in \SI{39}{\celsius} and \SI{30}{\celsius}; this might be incorrect, but we only assumed this to generate a prior distribution for the fitness values.
-The prior for $b$ was taken the same as for $c$.
-This prior did not converge: no parameter sets were found with an approximate likelihood greater than zero.
-%TODO IK check that this model had no bugs
-% TODO IK it makes sense because bla bla (to explain)
-Finally, we also tested a uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ and (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$.
+We also tested a uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ and (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$.
 In both cases the algorithm failed to converge.
@@
 A wider mutation rate prior, $\mu \sim \mathit{U}(10^{-9}\ ,\ 10^{-3})$, produced similar results.
-% TODO IK, maybe wider mutation rate should be tried on a model without aneuploidy
-%For a model without aneuploidy ($\delta=0$), we used a wider uniform prior for the mutation rate, $\mu \sim \mathit{U}(10^{-10},10^{-5})$.
 
@@
 The mutation rate corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
-% TODO IK please add lynch2008 to bib file
 
@@
 	\includegraphics[width=.95\textwidth]{../figures/evo39_fitness_39deg.pdf} 
-	% generated with growth_curves.ipynb and deg = '30' 
-	\includegraphics[width=0.95\textwidth]{../figures/evo39_fitness_30deg.pdf} 
-	% generated with growth_curves_refined.ipynb and deg = '30' 
+	% generated with growth_curves_refined.ipynb
 	\includegraphics[width=0.95\textwidth]{../figures/refined_vs_evo39_fitness_39deg.pdf}
@@
     \textbf{(A-D)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{39}{\celsius}: $\hat w_{2n+1}/w_{2n}$=1.024 (95\% CI: 0.959 - 1.115).
-    \textbf{(E-H)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{30}{\celsius}: $\hat w_{2n+1}/w_{2n}$= 0.986 (95\% CI: 0.979 - 0.994).
-    \textbf{(I-L)} Fitness estimation from growth curves of \anwt\ and \eumt\ at \SI{39}{\celsius}: $\hat w_{2n*}/w_{2n+1}$=1.033 (95\% CI: 1.027 - 1.041).\\
+    \textbf{(E-H)} Fitness estimation from growth curves of \anwt\ and \eumt\ at \SI{39}{\celsius}: $\hat w_{2n*}/w_{2n+1}$=1.033 (95\% CI: 1.027 - 1.041).\\
     Growth curves previously described in \citet[Figs. 3C, 4A, and S2]{Yona2012}.

### 2022-03-07 ms/ms.tex
some more fixes
first pass over IK recent changes

@@
 Thus, we terminated the inference when $\epsilon \le 0.13$ after six iterations, with $n_6=982$ accepted parameter vectors and effective sample size ESS=651 (\cref{fig:convergence}).
-% TODO for multi-locus model we need to say something else
 Running the inference algorithm with different initialization seeds and less or more simulations for approximating the likelihood produced similar posterior distributions~(\cref{fig:seeds}).
-%TODO for multi-locus model, write about them here also.
 
@@
 We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$ ~(\cref{fig:growth-curves}-D).
-In lieu of a better prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$.
+In lieu of a more suitable prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$.
 To increase computational efficiency, we also assumed $w(2n*)>w(2n+1*)>w(2n+1)>w(2n)$; running the inference without this assumption produced similar results.
@@
 Compared to other priors we tested, this prior produced lower WAIC, better posterior prediction plots, and more stable parameter estimates, as follows.
-%TODO IK what do you mean by better posterior prediction and is more stable parameter estimates still rellevant?
 We tried to use additional growth curves.
@@
 
-For the mutation rate, $\mu$, and aneuploidy rate, $\delta$, we used uninformative uniform priors, $\mu \sim \mathit{U}(10^{-9}\ ,\ 10^{-5})$ and $\delta \sim \mathit{U}(10^{-6} , 10^{-2})$.
-A wider mutation rate prior, $\mu \sim \mathit{U}(10^{-9}\ ,\ 10^{-3})$, produced similar results.
+For the mutation rate, $\mu$, and aneuploidy rate, $\delta$, we used uninformative uniform priors, $\mu \sim \mathit{U}(10^{-9},10^{-5})$ and $\delta \sim \mathit{U}(10^{-6},10^{-2})$.
+A wider mutation rate prior, $\mu \sim \mathit{U}(10^{-9},10^{-3})$, produced similar results.
 
@@
 The MAP (maximum a posteriori) and 50\% HDI (highest density interval) of the parameters posterior are~(\cref{fig:posterior}): %% generated by MAP-and-hdi.ipynb
-mutation rate, $\mu=2.942\times10^{-6} (2.017\times10^{-7}-4.15\times10^{-6})$,
-aneuploidy rate, $\delta=1.722\times10^{-3} (1.394\times10^{-3}-2.754\times10^{-3})$,
+mutation rate, $\mu=2.942\cdot10^{-6} (2.017\cdot10^{-7}-4.15\cdot10^{-6})$,
+aneuploidy rate, $\delta=1.722\cdot10^{-3} (1.394\cdot10^{-3}-2.754\cdot10^{-3})$,
 and fitness values, $w_{2n+1}=1.022 (1.021-1.023)$,
@@
 \paragraph{Model checking and comparison.}
-The single-locus model fits the data well: in simulations using the best-fit parameters (MAP estimate) \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\cref{fig:2n*-fixation}-B),
-Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\cref{fig:ppc-plot}). % TODO move somewhere else?
+The single-locus model fits the data well: in simulations using the best-fit parameters (MAP estimate) \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\cref{fig:2n*-fixation}B).
+% TODO IK what happened to the sentence about the fixation on anwt?
+green lines ub fig4?
+Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\cref{fig:ppc-plot}).
 Furthermore, sensitivity analysis shows that changing a single parameter while keeping the rest fixed at the MAP estimate produces a worse fit to the data~(\cref{fig:sensitivity}).
@@
 %% generated with MAP-and-hdi.ipynb
-mutation rate, $\mu=7.979\substack{+0.156 \\ -0.076}\times10^{-9}$,
+mutation rate, $\mu=7.979\substack{+0.156 \\ -0.076}\cdot10^{-9}$,
 fitness of mutant, $w_{2n^*}=1.013\substack{+0 \\ -0.001}$,
@@
 %% generated with MAP-and-hdi.ipynb
-With $\tau=10$, the estimated mutation rate was slightly lower ($\mu=1.674\times10^{-6} (2.501\times10^{-7}-1.741\times10^{-6})$). 
-With $\tau=100$, the estimated mutation rate was considerably lower ($\mu=3.81\times10^{-7} (1.508\times10^{-7}-4.995\times10^{-7})$) and the aneuploidy rate had higher variance ($\delta=7.824\times10^{-4} (6.616\times10^{-4}-1.462\times10^{-3})$).
+With $\tau=10$, the estimated mutation rate was slightly lower ($\mu=1.674\cdot10^{-6}\ [2.501\cdot10^{-7}-1.741\cdot10^{-6}]$). 
+With $\tau=100$, the estimated mutation rate was considerably lower ($\mu=3.81\cdot10^{-7}\ [1.508\cdot10^{-7}-4.995\cdot10^{-7}]$) and the aneuploidy rate had higher variance ($\delta=7.824\cdot10^{-4}\ [6.616\cdot10^{-4}-1.462\cdot10^{-3}]$).
 %TODO YR to reference to WAIC table also? By this table tau=100 is the best model
@@
 We also checked the differences in fixation/loss times of  \euwt, \anwt and \eumt\  for the models with different $\tau$.
-We saw that the models can be distinguished using these parameters, especially $\tau=100$ ~(\cref{fig:tau-plots}).
+We saw that the models can be distinguished using these parameters, especially $\tau=100$~(\cref{fig:tau-plots}).
 
@@
 We observed similar parameters inference comparing to the model with free $\mu$ parameter.
-However WAIC was better for the models with fixed $\mu$ ~(\autoref{table:WAIC}).
-The lowest WAIC got the model with $\mu$=$10^{-5}$ - 0.6, it is lower than \emph{no aneuploidy} model - 0.87, that implies higher predictive accuracy.
+However WAIC was better for the models with fixed $\mu$~(\autoref{table:WAIC}).
+The lowest WAIC, 0.6, was for a model with mutation rate fixed at $\mu$=$10^{-5}$.
+It is lower than \emph{no aneuploidy} model - 0.87, that implies higher predictive accuracy.
 
@@
   \begin{subfigure}{0.75\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-a.pdf}      
+      \includegraphics[width=\textwidth]{../figures/tau-diff-a.pdf}      
   \end{subfigure}
   \begin{subfigure}{0.75\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-b.pdf}      
+      \includegraphics[width=\textwidth]{../figures/tau-diff-b.pdf}      
   \end{subfigure}
-  \\
    \begin{subfigure}{0.75\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-c.pdf}      
+      \includegraphics[width=\textwidth]{../figures/tau-diff-c.pdf}      
   \end{subfigure}
    \caption{
-    \textbf{Genotype fixations for single-locus models with different $\tau$.} For each model, $10,000$ simulations using MAP parameters were evaluated and a percentage of simulations with fixated genotypes  \euwt, \anwt\ and \eumt\ (plots A, B and C respectively) were measured in time.
-\anmt\ didn't rich fixation, therefore it doesn't appear in figures.
-We can see that $\tau=100$ can be distinguished using data of loss of \euwt\ (plot A) and fixation/loss of \anwt\ (plot B).
-It is harder to distinguish $1\leq\tau\leq10$, however the correlation exists.
+    \textbf{Genotype fixations for single-locus models with different $\tau$.} We estimated the parameters for different models, each assuming a different value of $\tau$, the fold-increase in mutation rate in aneuploid cells.
+We then generated 10,000 simulations using the MAP estimate of each model and evaluated the fraction of simulations in which the genotype \euwt\ (\textbf{A}), \anwt\ (\textbf{B}), and \eumt\ (\textbf{C}) is fixed (y-axis) at each generation (x-axis).
+Note that \anmt\ did not fix.
+We can see that $\tau=100$ can be distinguished if time to loss of \euwt\ is known (panel~A) or if time to fixation or loss of \anwt\ is known (panel~B).
+It is harder to distinguish between $1\leq \tau \leq10$.
   }
   \label{fig:tau-plots}
+  % TODO IK axis labels like in figure 3, also somehow the graphics looks different from previous figures
+  % fixed means 100% ?
+
 \end{figure}
@@
 
-%\subsection*{Inference: multi-locus model.}
-
-%TODO same for complex model
-%model comparison between models - model probabilities
-
-\pagebreak
 % Discussion
@@
 
+\newpage
+
 \section*{Supplementary Material}
@@
     % TODO IK caption should explain what each panel shows.
-    \textbf{(A-D)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{39}{\celsius}: $\hat w_{2n+1}/w_{2n}$=1.024 (95\% CI: 0.959 - 1.115).
-    \textbf{(E-H)} Fitness estimation from growth curves of \anwt\ and \eumt\ at \SI{39}{\celsius}: $\hat w_{2n*}/w_{2n+1}$=1.033 (95\% CI: 1.027 - 1.041).\\
+    \textbf{(A-D)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{39}{\celsius}. $\hat w_{2n+1}/w_{2n}$=1.024 (95\% CI: 0.959 - 1.115).
+    \textbf{(E-H)} Fitness estimation from growth curves of \anwt\ and \eumt\ at \SI{39}{\celsius}. $\hat w_{2n*}/w_{2n+1}$=1.033 (95\% CI: 1.027 - 1.041).
     Growth curves previously described in \citet[Figs. 3C, 4A, and S2]{Yona2012}.
@@
 %% generated with fixed-mutation-rate.ipynb 
+% TODO IK run for all updated files
 \begin{table}[h]
@@
 \caption{
-\textbf{WAIC values for the different models.}}
+\textbf{WAIC values for various model specifications.}}
 \pgfplotstabletypeset[
@@
 \caption*{
-	%TODO IK caption, add diff tau models
+	TODO caption
+	%TODO IK caption; what is tau in models 1-5?
 } 

### 2022-03-07 ms/ms.tex
remove multi-locus model

@@
 \begin{figure}[b!]
-  %\centering
-  \begin{subfigure}[t]{0.4\textwidth}
-      \caption{
-        \textbf{A. Single-locus model}
-      }
   \centering
@@
   \label{fig:model1}
-  \end{subfigure}
-  %\\
-  \begin{subfigure}[t]{0.4\textwidth}
   \caption{
-        \textbf{B. Multi-locus model}
-      }
-      \centering
-      \includegraphics[height=1.8in]{../figures/Fig1-B.pdf}      
-      \label{fig:model2}
-  \end{subfigure}
-  \caption{
-    \textbf{Model illustrations.}
-    \textbf{(A)} In the single-locus model, the four genotypes are: euploid wild-type, \euwt; euploid mutant, \eumt; aneuploid wild-type, \anwt; and aneuploid mutant, \anmt.
+    \textbf{Model illustration.}
+    There are four genotypes in our model: euploid wild-type, \euwt; euploid mutant, \eumt; aneuploid wild-type, \anwt; and aneuploid mutant, \anmt.
     Overall there are two possible trajectories from \euwt\; to \eumt.
-    \textbf{(B)} In the multi-locus model, each genotype is characterized by its karyotype, \euwt\; or \anwt, and the number of accumulated beneficial mutations, denoted by stars. 
-    In both panels arrows denote transitions between genotypes, with transitions rates: $\mu$, beneficial mutation rate; $\delta$, aneuploidy rate.
+    Arrows denote transitions between genotypes, with transitions rates: $\mu$, beneficial mutation rate; $\delta$, aneuploidy rate.
   }
@@
 \paragraph*{Evolutionary Models.}
-We model the evolution of a population of cells using two models: a single-locus model and a multi-locus model.
-Both models are based on the Wright-Fisher model~\citep{Otto2007}, assuming non-overlapping generations and including the effects of natural selection, genetic drift, aneuploidy, and mutation. 
-We focus on beneficial mutations, neglecting the effects of deleterious and neutral mutations. Both models allow for a single aneuploid karyotype (e.g., chromosome III duplication).
-While the single-locus model allows for only a single mutation to accumulate in the genotype, the multi-locus model allows for multiple mutations to accumulate~(\cref{fig:models}), as well as for a fluctuating population size.
-
-%%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph*{Single-locus model.}
-This model assumes a constant effective population size $N$ and follows four genotypes~(\cref{fig:models}A): euploid wild-type, \euwt, the initial genotype; 
+We model the evolution of a population of cells using a Wright-Fisher model~\citep{Otto2007}, assuming a constant effective population size $N$, non-overlapping generations, and including the effects of natural selection, genetic drift, aneuploidy, and mutation.
+We focus on beneficial genetic modifications, neglecting the effects of deleterious and neutral mutations or karyotypic changes.
+The model allows for a single aneuploid karyotype (e.g., chromosome III duplication) and a single mutation to accumulate in the genotype.
+Thus, the model follows four genotypes~(\cref{fig:models}): euploid wild-type, \euwt, the initial genotype; 
 euploid mutant, \eumt, with the standard karyotype and a single beneficial mutation; 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph*{Multi-locus model.}
-This model expands the single-locus model by allowing for (i) the accumulation of beneficial mutations, and (ii) a fluctuating population size.
-
-A genotype is characterized by its karyotype, \euwt\; or \anwt, and the number of accumulated beneficial mutations, which can be zero or more.
-The selection coefficient of the $i$-th accumulated mutation in each individual, $s_i$, is drawn from an exponential distribution with expected value $s$: $s_i \sim \mathit{Exp{(s)}}$.
-The rest of the parameters ($N$, $\mu$, $\delta$, $b$, $c$) are the same as in the single-locus model.
-However, since the multi-locus model allows several mutations with smaller fitness effects to accumulate, we expect the mutation rate to be higher compared to the single-locus model, which focuses on a single, large-effect mutation.
-
-The fitness of the different genotypes is the same as in the single-locus model (\cref{table:single-locus}), except that the fitness contribution of $k$ beneficial mutations is the product of their independent effects, $\prod_{i=1}^k(1+s_i)$, instead of the contribution of the single mutation allowed in the single-locus model, $(1+s)$, see \cref{table:multi-locus}.
-Therefore, aneuploidy loss would be favored by selection only if there are enough beneficial mutations and/or the selection coefficients $s_i$ are large enough.
-The intuition is that when the benefit of the accumulated beneficial mutations is small, ten the benefit of aneuploidy has a large effect; when the benefit of the accumulated beneficial mutations is large, then aneuploidy is no longer beneficial due of its fitness cost.
-
-In contrast to the single-locus model, in the multi-locus model the population size fluctuates to model serial-transfer experimental protocol~\citep{Yona2012}: the population is serially diluted by transferring a fraction of the population (1/120) to a fresh medium approximately every seven generations.
-Thus, the population initial size is $N_0 = N$, and the population size is doubled every generation, $N_1=2N, N_2=4N, \ldots$, and diluted back to $N$ after seven generations such that $N_8=N$.
-
-The change in frequencies due to selection is exactly the same as in the single-locus model (\cref{eq:selection-single}), only applied using the fitness values in \cref{table:multi-locus}.
-The change due to random genetic drift is also the same as in \cref{eq:drift-single}, except that the frequencies vector is $\vec{f}=(f_{2n}\ ,\ f_{2n+1}\ ,\ f_{2n^*}\ ,\ f_{2n+1^*}\ ,\ f_{2n^{**}}\ ,\ f_{2n+1^{**}}\ ,\ \ldots)$ and that the population size changes between generations, as described above.
-
-The effects of mutation and aneuploidy on genotype frequencies is more elaborate than in the single-locus model.
-Genotype $i$ is classified according to its karyotype (\euwt\ or \anwt), the number of accumulated beneficial mutations ($k \ge 0$), and their fitness ($w_i$).
-Each offspring cell inherits these properties from its mother cell.
-Then, a new beneficial mutation is accumulated with probability $\mu$, such that the number of mutations is $k+1$, and its effect $s_{k+1}$ is drawn from an exponential distribution with expected value $s$, such that the contribution of the mutations to the fitness is $\prod_{j=0}^{k+1}{(1+s_j)}$.
-Next, euploid offspring become aneuploid with probability $\delta$, and aneuploid offspring become euploid with probability $\delta_L$.
-
-%%% Table: multi-locus model
-\begin{table}[h]
-\centering
-\caption{\textbf{Multi-locus model fitness values.}}
-\begin{tabular}{lllll}
-\emph{Genotype}, $g$ & $2n$ & $2n+1$ & $2n+1^{*k}$ & $2n^{*k}$ \\
-\hline
-\emph{Fitness}, $w_g$ & $1$ & $1-c+b$ & $(1-c)\prod_{j=1}^k(1+s_i)+b$ & $\prod_{j=1}^k(1+s_i)$               
-\end{tabular}
-\label{table:multi-locus}
-\caption*{
-$k$ is the number of accumulated beneficial mutations;
-$s \ge 0$ is the selection coefficient of a beneficial mutation;
-$0 \le c \le 1$ is the fitness cost of aneuploidy;
-and $b \ge c$ is the selection coefficient, or fitness benefit, of aneuploidy.
-}
-\end{table}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \paragraph{Empirical evidence.}
@@
 \end{aligned}\end{equation}
-
-% TODO if we show results of multi-locus model, write how much replicates were used.
 

### 2022-03-07 ms/ms.tex
revised inference results

@@
 \paragraph*{The present study.}
-Here, we develop evolutionary-genetic models that include the effects of natural selection, genetic drift, aneuploidy, and mutation to examine the role of aneuploidy in adaptive evolution.
-These models follow a population of cells characterised by both their ploidy and their genotype.
-We fit these models to the experimental results of \citet{Yona2012} using an \emph{approximate Bayesian computation} framework~\citep{Sisson2009} to infer model parameters, including selection coefficients and rates of aneuploidy and mutation, and to perform model selection between different models, thereby testing hypotheses about the evolutionary process.
-We analyze these evolutionary-genetic models to estimate the effects of parameters on the adaptation time and the probability for adaptation via aneuploidy.
-We find that \textbf{} % TODO
+Here, we develop an evolutionary-genetic model that includes the effects of natural selection, genetic drift, aneuploidy, and mutation to examine the role of aneuploidy in adaptive evolution.
+This model follows a population of cells characterised by both their ploidy and their genotype.
+We fit this model to the experimental results of \citet{Yona2012} using an \emph{approximate Bayesian computation} framework~\citep{Sisson2009} to infer model parameters, including selection coefficients and rates of aneuploidy and mutation, and to test hypotheses about the evolutionary process by performing model selection between different versions of the model.
+We also mathematically analyze this evolutionary-genetic model to estimate the effects of selection, mutation, and aneuploidy on the adaptation time and the probability for adaptation via aneuploidy.
 
-\pagebreak
 %%%%%%%%%%%%%%%%%%%%%%%%
@@
 The model allows for a single aneuploid karyotype (e.g., chromosome III duplication) and a single mutation to accumulate in the genotype.
-Thus, the model follows four genotypes~(\cref{fig:models}): euploid wild-type, \euwt, the initial genotype; 
+Thus, the model follows four genotypes~(\Cref{fig:models}): euploid wild-type, \euwt, the initial genotype; 
 euploid mutant, \eumt, with the standard karyotype and a single beneficial mutation; 
@@
 
-Transitions between the genotypes occur as follows~(\cref{fig:models}A): Beneficial mutations from \euwt\; to \eumt\; and from \anwt\; to \anmt\; occur with probability $\mu$, the mutation rate.
+Transitions between the genotypes occur as follows~(\Cref{fig:models}A): Beneficial mutations from \euwt\; to \eumt\; and from \anwt\; to \anmt\; occur with probability $\mu$, the mutation rate.
 We neglect back-mutations (i.e., from \eumt\; to \euwt\; and from \anmt\; to \anwt).
@@
 That is, we assume chromosomes are gained and lost at the same rate, and we neglect events that form a less-fit genotype (i.e., \anwt\; to \euwt\; and \eumt\; to \anmt).
-The fitness values of the four genotypes are given by \cref{table:single-locus}.
+The fitness values of the four genotypes are given by \Cref{table:single-locus}.
 
@@
     \end{equation}
-where the fitness values $w_i$ are given in \cref{table:single-locus} and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
+where the fitness values $w_i$ are given in \Cref{table:single-locus} and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
 The effect of mutation and aneuploidy on genotype frequencies is given by
@@
 For example, $P_{\tilde{\vec X}}\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)$ is approximated by taking simulations in which \anwt\; fixed before generation 450 but not before generation 200, and computing the fraction of such simulations in which \eumt\; did not fix by generation 1,700, and hence aneuploidy did not extinct before generation 1,700.
-\cref{fig:seeds} compares results with less and more simulated experiments, demonstrating that 1,000 simulations are likely enough.
+\Cref{fig:seeds} compares results with less and more simulated experiments, demonstrating that 1,000 simulations are likely enough.
 
-For a model without aneuploidy (that is, when the aneuploidy rate is fixed at zero, $\delta=0$), 
-the likelihood is similarly approximated by
-\begin{equation}\begin{aligned}
-\label{eq:heatstress-noaneuploidy-likelihood}
-\likelihood_{!}(\theta) = &\ 
-	1 - 
-	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\}\big) - \\
-&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\}\big) + \\
-&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\}\big)
-\;.
-\end{aligned}\end{equation}
+% YR: this whole section doesn't really make sense.
+We know there was aneuploidy.
+If we assume delta=0 that means that we think this aneuploidy was not relevant to the adaptation process.
+But then why do we take generations 1700 and 2300 to be the time for 2n* fixation?
+These times were not observed for fixation of a mutant but rather for elimination of aneuploidy.
+If we disregard aneuploidy, why do we still consider these times?
+%For a model without aneuploidy (that is, when the aneuploidy rate is fixed at zero, $\delta=0$), 
+%the likelihood is similarly approximated by
+%\begin{equation}\begin{aligned}
+%\label{eq:heatstress-noaneuploidy-likelihood}
+%\likelihood_{!}(\theta) = &\ 
+%	1 - 
+%	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\}\big) - \\
+%&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\}\big) + \\
+%&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\}\big)
+%\;.
+%\end{aligned}\end{equation}
 
@@
 The proposal particle is then perturbed using a kernel perturbation kernel, $K_t(\theta^* \mid \theta)$ where $\theta$ is the sample from the previous iteration.
-Then, a set of synthetic observations $\tilde{\vec X}^*$ is simulated, and the proposal particle $\theta^*$ is accepted if its approximate likelihood (\cref{eq:heatstress-likelihood}) is high enough, $\likelihood(\theta^*)>1-\epsilon_t$ (or more commonly, if $1-\likelihood(\theta^*) < \epsilon_t$), where $\epsilon_t>0$ is the \emph{acceptance threshold}, as higher values of $\epsilon_t$ allow more particles to be accepted.
+Then, a set of synthetic observations $\tilde{\vec X}^*$ is simulated, and the proposal particle $\theta^*$ is accepted if its approximate likelihood (\Cref{eq:heatstress-likelihood}) is high enough, $\likelihood(\theta^*)>1-\epsilon_t$ (or more commonly, if $1-\likelihood(\theta^*) < \epsilon_t$), where $\epsilon_t>0$ is the \emph{acceptance threshold}, as higher values of $\epsilon_t$ allow more particles to be accepted.
 The acceptance threshold $\epsilon_t$ is chosen as the median of the $1-\likelihood(\theta)$ of the particles accepted in the previous iteration, $t-1$, and $\epsilon_0=0.01$. 
@@
 
-Acceptance is determined according to the approximate likelihood~(\cref{eq:heatstress-likelihood}), which has a maximum value of 0.875.
-Thus, we terminated the inference when $\epsilon \le 0.13$ after six iterations, with $n_6=982$ accepted parameter vectors and effective sample size ESS=651 (\cref{fig:convergence}).
-Running the inference algorithm with different initialization seeds and less or more simulations for approximating the likelihood produced similar posterior distributions~(\cref{fig:seeds}).
+Acceptance is determined according to the approximate likelihood~(\Cref{eq:heatstress-likelihood}), which has a maximum value of 0.875.
+Thus, we terminated the inference when $\epsilon \le 0.13$ after six iterations, with $n_6=982$ accepted parameter vectors and effective sample size ESS=651 (\Cref{fig:convergence}).
+Running the inference algorithm with different initialization seeds and less or more simulations for approximating the likelihood produced similar posterior distributions~(\Cref{fig:seeds}).
 
@@
 Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix.
-We sampled 10,000 samples to use as a prior distribution~(\cref{fig:growth-curves}).
-We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$ ~(\cref{fig:growth-curves}-D).
+We sampled 10,000 samples to use as a prior distribution~(\Cref{fig:growth-curves}).
+We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$ ~(\Cref{fig:growth-curves}-D).
 In lieu of a more suitable prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$.
@@
 Compared to other priors we tested, this prior produced lower WAIC, better posterior prediction plots, and more stable parameter estimates, as follows.
-We tried to use additional growth curves.
-We used growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$ ~(\cref{fig:growth-curves}-L).
+First, we tried to use additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$ ~(\Cref{fig:growth-curves}L).
 The same prior was used for $w_{2n*}/w_{2n+1*}$.
@@
 % generated with fixed-mutation-rate-and-WAICs.ipynb TODO IK update the WAIC with 1000 reps run
-We also tested a uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ and (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$.
-In both cases the algorithm failed to converge.
+We also tested an uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$, or (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$.
+In both cases the inference algorithm failed to converge.
 
@@
 
-\subsection*{Inference: single-locus model}
+\subsection*{Statistical inference}
 
 \paragraph{Parameter estimation.} 
-We used ABC-SMC to infer the posterior distribution of the parameters of the single-locus model.
-The MAP (maximum a posteriori) and 50\% HDI (highest density interval) of the parameters posterior are~(\cref{fig:posterior}): %% generated by MAP-and-hdi.ipynb
-mutation rate, $\mu=2.942\cdot10^{-6} (2.017\cdot10^{-7}-4.15\cdot10^{-6})$,
-aneuploidy rate, $\delta=1.722\cdot10^{-3} (1.394\cdot10^{-3}-2.754\cdot10^{-3})$,
-and fitness values, $w_{2n+1}=1.022 (1.021-1.023)$,
-$w_{2n+1^*}=1.025 (1.024-1.026)$,
-$w_{2n^*}=1.028 (1.026-1.029)$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$.
-
-% TODO IK we are jumping between b c s and w2n* w2n+1 w2n+1*. we should report all models and results either with b c s or with w. you can compute b c s from w before computing the HDIs.
-This estimated aneuploidy rate agrees with previous estimates. % TODO refs (9.7*10-5 in Zhu 2014, so it's lower). YR: we need more than one ref; also what is the units of Zho 2014?
-The mutation rate corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
+We used ABC-SMC to infer the posterior distribution of model parameters~(\Cref{fig:posterior}). 
+We report parameter estimates using the MAP (maximum a posteriori) and providing the 50\% HDI (highest density interval) in square brackets.
+%% estimates generated by MAP-and-hdi.ipynb
+The estimated aneuploidy rate, $\delta=1.722\cdot10^{-3}\ [1.394\cdot10^{-3}-2.754\cdot10^{-3}]$, agrees with previous estimates.
+% TODO refs (9.7*10-5 in Zhu 2014, so it's lower).
+YR: we need more than one ref; also what is the units of Zho 2014?
+The estimated mutation rate, $\mu=2.942\cdot10^{-6}\ [2.017\cdot10^{-7}-4.15\cdot10^{-6}]$, corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
+The estimated fitness values are $w_{2n+1}=1.022\ [1.021-1.023]$,
+$w_{2n+1^*}=1.025\ [1.024-1.026]$,
+$w_{2n^*}=1.028\ [1.026-1.029]$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$. 
+Thus, we can infer that the benefit and cost of trisomy was $b=2.5\%$ and $c=0.3\%$, and the benefit of the beneficial mutation was $2.8\%$~(\Cref{table:single-locus}).
 
@@
   \caption{
-  \textbf{Posterior distribution of single-locus model.}
+  \textbf{Posterior distribution of model parameters.}
 On the diagonal, the inferred posterior distribution of each model parameter. 
@@
 \paragraph{Model checking and comparison.}
-The single-locus model fits the data well: in simulations using the best-fit parameters (MAP estimate) \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\cref{fig:2n*-fixation}B).
+The model fits the data well: in simulations using the MAP parameter estimates, \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\Cref{fig:2n*-fixation}B).
 % TODO IK what happened to the sentence about the fixation on anwt?
 green lines ub fig4?
-Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\cref{fig:ppc-plot}).
-Furthermore, sensitivity analysis shows that changing a single parameter while keeping the rest fixed at the MAP estimate produces a worse fit to the data~(\cref{fig:sensitivity}).
+Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\Cref{fig:ppc-plot}).
 
-However, a model without aneuploidy where the aneuploidy rate is fixed at zero, $\delta=0$, cannot explain the experimental observations~(Figure~\ref{fig:2n*-fixation}).
-The estimated parameter values without aneuploidy are:
+% commented out, see remark above when describing likelihood of delta=0.
+%Moreover, a model without aneuploidy (where the aneuploidy rate is fixed at zero, $\delta=0$), fails to explain the experimental observations~(Figure~\ref{fig:2n*-fixation}).
 %% generated with MAP-and-hdi.ipynb
-mutation rate, $\mu=7.979\substack{+0.156 \\ -0.076}\cdot10^{-9}$,
-fitness of mutant, $w_{2n^*}=1.013\substack{+0 \\ -0.001}$,
-The estimated mutation rate without aneuploidy is much lower compared to a model with aneuploidy, suggesting a target size of just 40, because a higher mutation rate will lead to faster fixation of \eumt\ than in the experimental observations (\cref{fig:2n*-fixation}).
-Thus, even disregarding the appearance of trisomic cells in the experiment, the evidence supports the inclusion of aneuploidy in the model.
-% TODO IK why do you say that?
-spell it out please - WAIC is lower?
-likelihood is higher?
-ppc looks better?
-what?
+%The estimated mutation rate without aneuploidy is $\mu=7.979\cdot10^{-9}\ [7.903\cdot 10^{-9}-8.135\cdot10^{-9}]$, much lower compared to a model with aneuploidy and suggesting a target size of just 40.
+%The fitness of the mutant is also much lower at $w_{2n^*}=1.013\ [1.012-1.013]$.
+%This is because, without aneuploidy, a high mutation rate or fitness effect will lead to faster appearance and fixation of \eumt\ than in the experimental observations.
+Even with these lower estimates, the model fit is worse than that of a model with aneuploidy~(\Cref{fig:2n*-fixation}).
 
-We also inferred model parameters under the assumption that the mutation rate in aneuploid cells, compared to euploid cells, is increased by a factor $\tau$=1, 33/32 (due to an additional chromosome), 2, 5, 10 or 100 (due to genetic instability).
-We found that the posterior distribution was similar for $\tau=$1, 33/32, 2, and 5 (\cref{fig:tau}).
+We also inferred model parameters under the assumption that the mutation rate increases in aneuploid cells by a factor $\tau$ = 33/32 (due to an additional chromosome), 2, 5, 10, or 100 (due to genetic instability).
+We found that the posterior distribution was similar for $\tau=$1, 33/32, 2, and 5 (\Cref{fig:tau}).
 %% generated with MAP-and-hdi.ipynb
-With $\tau=10$, the estimated mutation rate was slightly lower ($\mu=1.674\cdot10^{-6}\ [2.501\cdot10^{-7}-1.741\cdot10^{-6}]$). 
-With $\tau=100$, the estimated mutation rate was considerably lower ($\mu=3.81\cdot10^{-7}\ [1.508\cdot10^{-7}-4.995\cdot10^{-7}]$) and the aneuploidy rate had higher variance ($\delta=7.824\cdot10^{-4}\ [6.616\cdot10^{-4}-1.462\cdot10^{-3}]$).
-%TODO YR to reference to WAIC table also? By this table tau=100 is the best model
-We cannot rule out an increased mutation rate in aneuploid cells, but we find no evidence that it does, and unless the effect is strong, it does not seem to affect our results.
+With $\tau=100$, the estimated mutation rate was about 7-8-fold lower compared to $\tau=1$ ($\mu=3.81\cdot10^{-7}\ [1.508\cdot10^{-7}-4.995\cdot10^{-7}]$) and the aneuploidy rate was about 2-3-fold lower ($\delta=0.782\cdot10^{-3}\ [0.661\cdot10^{-3}-1.462\cdot10^{-3}]$).
+With $\tau=10$, the estimated mutation rate was only slightly lower compared to $\tau=1$ ($\mu=1.674\cdot10^{-6}\ [2.501\cdot10^{-7}-1.741\cdot10^{-6}]$). 
+WAIC is lowest for $\tau=100$~(\Cref{table:WAIC}), and if we rule such a strong effect of genetic instability a-priori, the next two lowest WAIC values are for $\tau=33/32$ and $\tau=5$.  
+Therefore, we cannot rule out an increased mutation rate in aneuploid cells, but unless the effect is strong ($\tau=100$), it does not seem to affect our inference results.
 We also checked the differences in fixation/loss times of  \euwt, \anwt and \eumt\  for the models with different $\tau$.
-We saw that the models can be distinguished using these parameters, especially $\tau=100$~(\cref{fig:tau-plots}).
+We saw that the models can be distinguished using these parameters, especially $\tau=100$~(\Cref{fig:tau-plots}).
 
-% TODO YR Read this
-Next we tried the model with fixed mutation rate $\mu$.
-Hence $\mu$ is not well identifiable by the model (the inferred value has high variance with no strong correlation between it and other parameters), we tried the models where mutation rate is fixed: $\mu$=$10^{-5}$, $10^{-6}$ and $10^{-7}$.
-We observed similar parameters inference comparing to the model with free $\mu$ parameter.
-However WAIC was better for the models with fixed $\mu$~(\autoref{table:WAIC}).
-The lowest WAIC, 0.6, was for a model with mutation rate fixed at $\mu$=$10^{-5}$.
-It is lower than \emph{no aneuploidy} model - 0.87, that implies higher predictive accuracy.
+Sensitivity analysis shows that changing a single parameter while keeping the rest fixed at the MAP estimate produces a worse fit to the data~(\Cref{fig:sensitivity}).
+Furthermore, we fitted models with a mutation rate fixed at $\mu$=$10^{-5}$, $10^{-6}$ and $10^{-7}$.
+We inferred similar parameters estimates compared to the model with a free $\mu$ parameter.
+% TODO IK where is this shown?
+how do you quantify it?
+WAIC was lower when $\mu$ is fixed~(\Cref{table:WAIC}), but this is not surprising, as WAIC attempt to balance between model fit and model complexity, where the latter takes into account the number of model parameters.
 
@@
 	Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy in blue and green, respectively, and corresponding to the blue and green bars in panel A.
-The MAP likelihood (\cref{eq:heatstress-likelihood}) is 0.84 and 0.67 with and without aneuploidy, respectively.
+The MAP likelihood (\Cref{eq:heatstress-likelihood}) is 0.84 and 0.67 with and without aneuploidy, respectively.
      }
@@
 
-\pagebreak
+\newpage
+
 %%%%%%%%%%
-\subsection*{Analysis: single-locus model.}
+\subsection*{Analysis}
 
@@
  %There are certainly combinations of the mutation rate, the aneuploidy rate, the selection strength and the population size for which at least one of these assumptions is violated.
-Thus, \cref{eq:T_total} may not yield a good approximation to the actual time to the adapted state.
+Thus, \Cref{eq:T_total} may not yield a good approximation to the actual time to the adapted state.
  %\end{remark}
 
-To validate the approximation in \cref{eq:T_total}, we performed simulations of the full model (\cref{eq:mutation-aneuploidy-single,eq:drift-single,eq:selection-single})  without aneuploidy, $\delta=0$ (\cref{fig:only mut}), and with the other parameters as inferred above.
+To validate the approximation in \Cref{eq:T_total}, we performed simulations of the full model (\Cref{eq:mutation-aneuploidy-single,eq:drift-single,eq:selection-single})  without aneuploidy, $\delta=0$ (\Cref{fig:only mut}), and with the other parameters as inferred above.
 The approximation provides a good fit to simulation results except when the effective population size is larger than $5\cdot 10^6$ cells.
@@
 Mean and 95\% CI from simulations in blue.
-Analytic approximation (\cref{eq:T_total}) in orange.
+Analytic approximation (\Cref{eq:T_total}) in orange.
 Here, $w_\anwt=1.021,\,   w_\anmt=1.025$ and $w_\eumt=1.028$.
@@
   \caption{
-    \textbf{Likelihood profiles.} Sensitivity of the approximate likelihood of the single-locus model, $\likelihood(\theta)$, to changing a single parameter while the other parameters remain fixed at their MAP estimates.
+    \textbf{Likelihood profiles.} Sensitivity of the model approximate likelihood, $\likelihood(\theta)$, to changing a single parameter while the other parameters remain fixed at their MAP estimates.
 Dashed vertical line represents the MAP value.
@@
     \caption{
-    \textbf{Single-locus model inference convergence.} 
-    The ABC-SMC algorithm was used to infer the parameters of the single-locus model.
+    \textbf{Inference convergence.} 
+    The ABC-SMC algorithm was used to infer the model parameters.
 \textbf{(A-E)} The approximate posterior distributions of model parameters at each iteration of the ABC-SMC algorithm demonstrates convergence, as the posterior did not significantly change after the first iteration, $t=1$.
@@
     \textbf{Posterior distribution validation.}
-    The posterior distribution of the single-locus model parameters is roughly the same regardless of the number of simulations (4-10,000 replicates) used to approximate the likelihood (\cref{eq:heatstress-likelihood}).
+    The posterior distribution of model parameters is roughly the same regardless of the number of simulations (4-10,000 replicates) used to approximate the likelihood (\Cref{eq:heatstress-likelihood}).
     }

### 2022-03-07 ms/ms.tex
merged new-analysis branch

@@
 %% generated with posterior-plot.ipynb
-\begin{figure}[h!]
+\begin{figure}[p]
   \centering
@@
 %% generated with with-aneuploidy_vs_no_aneuploidy.ipynb
-\begin{figure}[h!]
+\begin{figure}[p]
   \begin{subfigure}{0.5\textwidth}
@@
 
-\begin{figure}[h!]
+\begin{figure}[p]
   \centering
@@
 %% generated by diff-tau-plots.ipynb
-\begin{figure}[h!]
+\begin{figure}[p]
   \centering
@@
   % fixed means 100% ?
-
 \end{figure}
 
-\newpage
-
+\pagebreak
 %%%%%%%%%%
@@
 
-Here, we aim to approximate the adaptation time (i.e., time to fixation of \eumt) in the single-locus model and how it is affected by the model parameters.
+Here, we aim to approximate the adaptation time (i.e., time to fixation of \eumt) and analyse how it is affected by the model parameters.
 As a stochastic model, there are two events that can occur in the \euwt\ population.
@@
  %There are certainly combinations of the mutation rate, the aneuploidy rate, the selection strength and the population size for which at least one of these assumptions is violated.
-Thus, \Cref{eq:T_total} may not yield a good approximation to the actual time to the adapted state.
+Thus, \cref{eq:T_total} may not yield a good approximation to the actual time to the adapted state.
  %\end{remark}
 
-To validate the approximation in \Cref{eq:T_total}, we performed simulations of the full model (\Cref{eq:mutation-aneuploidy-single,eq:drift-single,eq:selection-single})  without aneuploidy, $\delta=0$ (\Cref{fig:only mut}), and with the other parameters as inferred above.
+As a first step, to validate the approximation in \cref{eq:T_total}, we performed simulations of the full model (\cref{eq:mutation-aneuploidy-single,eq:drift-single,eq:selection-single})  without aneuploidy, $\delta=0$ (\cref{fig:only mut}), and with the other parameters as inferred above.
 The approximation provides a good fit to simulation results except when the effective population size is larger than $5\cdot 10^6$ cells.
@@
 Mean and 95\% CI from simulations in blue.
-Analytic approximation (\Cref{eq:T_total}) in orange.
+Analytic approximation (\cref{eq:T_total}) in orange.
 Here, $w_\anwt=1.021,\,   w_\anmt=1.025$ and $w_\eumt=1.028$.
@@
   %TODO MP add panel labels A and B
+  %replace XXXX with parameter values and remove them from the figure titles.
+  %y label "Expected adaptation time"
+  % remove legend
+  % make it a single figure with two panels, then there is no need for y label on panel B
   % The error bars a quite large, can you make them smaller by running more simulations?
@@
 
+ \begin{figure}[h!]
+ %\centering
+ \includegraphics[width=1\textwidth]{../figures/figgrid_nodirectmut_ylog_uncorr0.jpg}
+ \caption{\textbf{Expected adaptation time without mutations from $\euwt$ to $\eumt$.}
+  The expected time to appearance and fixation of \eumt\ as a function of the population size, $N$ (panel A) and as a function of the mutation rate, $\mu$ (panel B) and as a function of the aneuploidy rate, $\delta$ (panel C).
+Mean and 95\% CI from simulations in blue.
+Analytic approximation (\cref{eq:T_total}) in green.
+The red curve is achieved by setting $\tilde{T}_{\euwt \rightarrow \anwt}=\tilde{T}_{\anwt \rightarrow \anmt}=0$.
+The orange line is achieved by setting $x=0.017$ in \cref{eq:T_fix_x} for the fixation time of $\eumt$, such that it is well fitting in panel B.
+Here, $w_\anwt=1.021,\,   w_\anmt=1.025$ and $w_\eumt=1.028$.
+Parameter values in panels where it is not varied, $N=6\cdot 10^6$,\, $\mu=9.6\cdot 10^{-8}$ and $\delta=5.4\cdot 10^{-4}$.
+}
+ \label{fig:no direct mut}
+ \end{figure}
+
+ Second, we ran simulations without state changes from $\euwt$ to $\eumt$.
+Before, we had the optimal scenario of waiting until the first successful mutation appears and then waiting until is reaches fixation.
+Our approximations did quite well there.
+Here, the aneuploid states are transient and fixation could be outpaced by the arrival of the next type.
+This would explain the discrepancy between the green curves in \cref{fig:no direct mut}.
+To asses this further, \cref{fig:fixT vs waitT} provides a clear picture.
+At least for intermediate to high values of the respective range of the parameters, the waiting time until a succesful new type appears is far smaller than that to fixation of the old type.
+Taking this into account, it seems reasonable to ignore the fixation time of $\anwt$ and $\anmt$, which results in the better fitting red curves in \cref{fig:no direct mut}.
+Inferring $x=0.017$ for $\tilde{T}_{\anmt \rightarrow \eumt}$ to get the best fit in panel B, does not imply that it is also a best fit for panels A and C.
+However, we stop simulations in state $\eumt$ if they have frequency $0.95$.
+
+  The discrepancies in \cref{fig:no direct mut}C seem to be particularly odd as the approximations are nearly constant and the simulations decreas log-linearly with increasing $\delta$.
+
+ \begin{figure}[h!]
+ %\centering
+ \includegraphics[width=1\textwidth]{../figures/figgrid_fixT_vs_waitT.jpg}
+ \caption{\textbf{Expected time of fixation and waiting time events on the aneuploid trajectory.}
+  The expected time to appearance and fixation of each step seperately in the aneuploid trajectory as a function of the population size, $N$ (panel A) and as a function of the mutation rate, $\mu$ (panel B) and as a function of the aneuploidy rate, $\delta$ (panel C).
+Solid curves give the fixation time and dashed curves the waiting time.
+Curves of the same color indicate the same base population state;  e.g.
+the blue dashed curve is the waiting time in state $\anwt$ until a succesful individual of type $\anmt$ appears.
+The corresponding solid curve indicates the time to fixation of the population in state $\anwt$.
+Here, $w_\anwt=1.021,\,   w_\anmt=1.025$ and $w_\eumt=1.028$.
+Parameter values in panels where it is not varied, $N=6\cdot 10^6$,\, $\mu=9.6\cdot 10^{-8}$ and $\delta=5.4\cdot 10^{-4}$.
+}
+ \label{fig:fixT vs waitT}
+ \end{figure}
+
+In summary, whereas the two possible trajectories to $\eumt$ do not compete against each other in both \cref{fig:only mut} and \cref{fig:no direct mut}, the approximations are doing well only in restricted parameter ranges.
+These tend to be, where the respective parameters are rather small.
+If they grow bigger, they are violating some assumptions of our approximations.
+
+%%%%%%%%%%
 % Discussion

### 2022-03-07 ms/ms.tex
YR revised analysis up to line 541

--- 2022-03-07 ms/ms.tex
@@
 
-Here, we aim to approximate the adaptation time (i.e., time to fixation of \eumt) in the single-locus model and how it is affected by the model parameters.
-As a stochastic model, there are two events that can occur in the \euwt\ population.
-The first is a mutation that produces a \eumt\ cell.
-The second is a missegregation event that produces a \anwt\ cell.
-The probability with which these events occur is significantly different.
-We are interested in the average waiting time until one of the two events occurs and then lead to adaptation (in the second case, it will be followed by a mutation to \anmt\ and loss of aneuploidy to \eumt).
-We assume the waiting time for appearance of \eumt\ from \euwt\ and appearance of \anwt\ from \euwt are geometrically distributed with rates $p_{2n^*}$ and $p_{2n+1}$, respectively, where
+Here, we aim to approximate the adaptation time (i.e., time to fixation of \eumt) and analyse how it is affected by the model parameters.
+As a stochastic model, there are two events that can occur on the \euwt\ genotype background.
+The first is a mutation that produces a \eumt\ genotype.
+The second is a missegregation event that produces a \anwt\ genotype.
+The probabilities with which these events occur is significantly different.
+We are interested in the average waiting time until one of these two genotype appears and then leads to adaptation, i.e., appearance and fixation of the \eumt; in the second case, appearance of \anwt\ will be followed by a mutation to \anmt\ and loss of aneuploidy to \eumt.
+
+We assume the waiting time for appearance of \eumt\ from \euwt\ and appearance of \anwt\ from \euwt\ are geometrically distributed with rates $p_{2n^*}$ and $p_{2n+1}$, respectively, where
 \begin{subequations}
@@
 \end{subequations}
-where $\Delta w_i$ denotes the relative difference between the fitness of genotype $i$ and the population mean fitness.
-However, we have two competing events: if one event occurs (e.g., mutation) then the other (aneuploidy) is unlikely to occur,  we need to find the minimum of these waiting times.
-The minimum of two geometric random variables with parameters $p_{2n^*}$ and $p_{2n+1}$ is also a geometric random variable with parameter $1-(1-p_{2n^*})(1-p_{2n+1})$, which is the inverse of the expected waiting time until the first event occurs.
-The probability that aneuploidy will be the first event to occur is
+where $\Delta w_i$ denotes the relative difference between the fitness $w_i$ of genotype $i$ and the population mean fitness, $\bar{w}$.
+
+However, we have two competing events: if one event occurs (e.g., mutation), then the other event (e.g., aneuploidy) is unlikely to occur.
+Thus, we need to find the minimum of these waiting times.
+The minimum of two geometric random variables with rates $p_{2n^*}$ and $p_{2n+1}$ is also a geometric random variable with rate $1-(1-p_{2n^*})(1-p_{2n+1})$, which is the inverse of the expected waiting time until the first event occurs.
+
+% TODO MP: this seems like repetition of things written above, right? should we remove this? or if there is something important here, merge it to previous paragraphs?
+%We are interested in the expected time to fixation of \eumt, and the trajectory leading to this fixation depends on the event that occurs first.
+If \eumt\ appears first and avoids extinction by drift, then it will fix.
+If \anwt\ appears first and avoids extinction, then several other events must occur for fixation of \eumt.
+Thus, in this aneuploid trajectory, we need to approximate the expected waiting time to reach \anmt\ and finally \eumt.
+
+Given a mutation occurs first, % TODO MP: I added this sentence, is that right?
+the expected adaptation time via the mutation trajectory is 
+\begin{equation}\label{eq:T_mut}
+T_{M}=\frac{1}{1-(1-p_{2n^*})(1-p_{2n+1})} + T_{2n \to 2n^*} \;,
+\end{equation}
+where $T_{i \to j}$ is the waiting time for fixation of genotype $i$ in a population of genotype $j$, given it starts with a single copy and does not go to extinction by drift.
+Given that aneuploidy occurs first, % TODO MP: I added this sentence, is that right?
+the expected adaptation time via the aneuploidy trajectory is approximated by
+\begin{equation}\label{eq:T_aneu} 
+% TODO MP: p_{2n+1^*} is not defined, and why is there 1/p_{2n^*} here?
+shouldn't it have the waiting time for 2n+1* to appear on 2n+1 and then waiting time for 2n* to appear on 2n+1*?
+something is wrong here.
+ T_A = \frac{1}{1-(1-p_{2n^*})(1-p_{2n+1})} + \frac{1}{p_{2n+1^*}} + \frac{1}{p_{2n^*}} + T_{2n \to 2n+1} + T_{2n+1 \to 2n+1^*} + T_{2n+1^* \to 2n^*} \;.
+ \end{equation} % TODO MP do we need the tilde ?
+Finally, the expected adaptation time is 
+ \begin{equation}\label{eq:T_total}
+ T = p_A T_{A}+(1-p_A) T_{M} \;,
+ \end{equation}
+ where $p_A$ is he probability that aneuploidy will be the first event to occur,
 \begin{equation}
@@
 \end{equation}
-We are interested in the expected time to fixation of \eumt, and the events leading to this fixation depend on the event that occurs first.
-If \eumt\ appears first and avoids extinction by drift, then it will fix.
-If \anwt\ appears first and avoids extinction, then several other events must occur for fixation of \eumt.
-Thus, in this aneuploid trajectory, we need to approximate the expected waiting time to reach \anmt\ and finally \eumt.
-In general, we assume that the waiting time until a new genotype occurs is much longer than the time to its fixation.
-Nonetheless, these fixation times must be included in a rigorous analysis.
+
+We assume that the waiting time until a new genotype occurs is much longer than the waiting time for its fixation in the population.
+Nevertheless, these fixation times must be included in a rigorous analysis.
 There are two ways to compute the fixation times and they differ in the assumption of the underlying process.
-
-A deterministic-process approach is given by the classical bi-allelic single-locus equation, which provides an approximation for the expected time for genotype $i$ to reach frequency $x$ when starting from frequency $1-x$ in a population of genotype $j$ is % TODO MP is this correct?
- % TODO MP ref for equation? Crow and Kimura?
-\begin{equation}
-\tilde T_{i \to j} = 2\frac{\log{\big(\frac{x}{1-x}N\big)}}{\log(w_i/w_j)} \;,
+A deterministic-process approach is given by the classical bi-allelic single-locus equation.
+It approximates the expected time for genotype $i$ to reach frequency $x$ when starting from frequency $1-x$ in a population of genotype $j$,
+ % TODO MP ref for equation? Crow and Kimura? YR: yes
+\begin{equation} \label{eq:T_fix_x}
+ T_{i \to j} \approx 2\frac{\log{\big(\frac{x}{1-x}N\big)}}{\log(w_i/w_j)} \;,
 \end{equation} 
-A stochastic-process approach for approximating $\tilde T_i$ is given by the diffusion-equation approximation \citep[eq.~17]{Kimura1969a}. 
-Thus, the expected adaptation time via the aneuploidy trajectory is given by
-\begin{equation}\label{eq:T_aneu}
- T_A = \frac{1}{1-(1-p_{2n^*})(1-p_{2n+1})} + \frac{1}{p_{2n+1^*}} + \frac{1}{p_{2n^*}} + \tilde T_{2n \to 2n+1} + \tilde T_{2n+1 \to 2n+1^*} + \tilde T_{2n+1^* \to 2n^*} \;,
- \end{equation}
-the expected adaptation time via the mutation trajectory is given by 
-\begin{equation}\label{eq:T_mut}
-T_{M}=\frac{1}{1-(1-p_{2n^*})(1-p_{2n+1})} + \tilde T_{2n \to 2n^*} \;,
-\end{equation}
-and the expected adaptation time is given by
- \begin{equation}\label{eq:T_total}
- T = p_A T_{A}+(1-p_A) T_{M}.
- \end{equation}
+A stochastic-process approach for approximating $T_{i \to j}$ is given by the diffusion-equation approximation \citep[eq.~17]{Kimura1969a}.
 
@@
 
-To validate the approximation in \Cref{eq:T_total}, we performed simulations of the full model (\Cref{eq:mutation-aneuploidy-single,eq:drift-single,eq:selection-single})  without aneuploidy, $\delta=0$ (\Cref{fig:only mut}), and with the other parameters as inferred above.
+First, to validate the approximation in \Cref{eq:T_total}, we performed simulations of the full model (\Cref{eq:mutation-aneuploidy-single,eq:drift-single,eq:selection-single}) without aneuploidy, i.e., $\delta=0$ (\Cref{fig:only mut}), and with the other parameters as inferred above.
 The approximation provides a good fit to simulation results except when the effective population size is larger than $5\cdot 10^6$ cells.
-In that case, the expected number of new mutants is ~0.5, and therefore we can expect that in many generations the fixation event will start with more than a single \eumt\ cell appearing in the same generation, which is in contrast to our assumption that an adaptation event starts from a single mutant cell.
+In that case, the expected number of new mutants per generation is roughly 0.5, and  we can expect that in many generations the fixation event will start with more than a single \eumt\ cell appearing in the same generation, which violates our assumption that an adaptation event starts from a single mutant cell.
 
@@
  \caption{\textbf{Expected adaptation time without aneuploidy.}
-  The expected time to appearance and fixation of \eumt\ as a function of the population size, $N$ (panel A) and as a function of the mutation rate, $\mu$ (panel B).
+  The expected time to appearance and fixation of \eumt\ as a function of the population size, $N$ (\textbf{A}) or as a function of the mutation rate, $\mu$ (\textbf{B}).
 Mean and 95\% CI from simulations in blue.
@@
 Here, $w_\anwt=1.021,\,   w_\anmt=1.025$ and $w_\eumt=1.028$.
-In panel A, $\mu=9.6\cdot 10^{-8}$ and in panel B, $N=6\cdot 10^6$.
+In panel A, $\mu=9.6\cdot 10^{-8}$.
+In panel B, $N=6\cdot 10^6$.
   }
-  %TODO MP add panel labels A and B
+  %TODO MP 
+  % which approximation do you use here for T_i->j ?
   % The error bars a quite large, can you make them smaller by running more simulations?
-  % Add more x ticks for panel A (10^6, 5 10 ^6 ...)
-  % maybe start x axis in panel A with lower population size? 10^4 or 10^5 ? does the approximation work well there?
+  % --> or by making the y axis in log scale!!
  \label{fig:only mut}
@@
 
+ \begin{figure}[h!]
+ %\centering
+ \includegraphics[width=1\textwidth]{../figures/figgrid_nodirectmut_ylog_uncorr0.jpg}
+ \caption{\textbf{Expected adaptation time without mutations from $\euwt$ to $\eumt$.}
+  The expected time to appearance and fixation of \eumt\ as a function of the population size, $N$ (\textbf{A}) and as a function of the mutation rate, $\mu$ (\textbf{B}) and as a function of the aneuploidy rate, $\delta$ ({C}).
+Blue error bars for mean and 95\% CI from simulations.
+Green line for the analytic approximation (\Cref{eq:T_total}).
+% TODO why green?
+in previous figure it was orange.
+  Red curve for the approximation with $T_{\euwt \to \anwt}=T_{\anwt \to \anmt}=0$. 
+  Orange curve for the approximation with $x=0.017$ in \Cref{eq:T_fix_x} for the fixation time of $\eumt$; we found this gives a good fit in panel B.
+  Here, $w_\anwt=1.021,\, w_\anmt=1.025$ and $w_\eumt=1.028$, $N=6\cdot 10^6$,\, $\mu=9.6\cdot 10^{-8}$ and $\delta=5.4\cdot 10^{-4}$ unless parameter is varied on the x-axis.
+  }
+ \label{fig:no direct mut}
+ \end{figure}
+
+Second, we performed simulations without mutations from $\euwt$ to $\eumt$ (\Cref{fig:no direct mut}). 
+% TODO MP: I comment this out - not clear what "Before" refers to, and not sure we need this sentence. what do you think?
+%Before, we had the optimal scenario of waiting until the first successful mutation appears and then waiting until is reaches fixation. Our approximations did quite well there. 
+Here, aneuploidy is transient and fixation of \anwt\ or \anmt\ could be outpaced by the arrival of the next genotype.
+This would explain the discrepancy between the green curves in \Cref{fig:no direct mut}.
+To asses this further, \Cref{fig:fixT vs waitT} provides a clear picture.
+At least for intermediate to high values of the respective range of the parameters, the waiting time until a succesful new type appears is far smaller than that to fixation of the old type.
+Taking this into account, it seems reasonable to ignore the fixation time of $\anwt$ and $\anmt$, which results in the better fitting red curves in \Cref{fig:no direct mut}.
+Inferring $x=0.017$ for $\tilde{T}_{\anmt \rightarrow \eumt}$ to get the best fit in panel B, does not imply that it is also a best fit for panels A and C.
+However, we stop simulations in state $\eumt$ if they have frequency $0.95$.
+
+  The discrepancies in \Cref{fig:no direct mut}C seem to be particularly odd as the approximations are nearly constant and the simulations decreas log-linearly with increasing $\delta$.
+
+ \begin{figure}[h!]
+ %\centering
+ \includegraphics[width=1\textwidth]{../figures/figgrid_fixT_vs_waitT.jpg}
+ \caption{\textbf{Expected time of fixation and waiting time events on the aneuploid trajectory.}
+  The expected time to appearance and fixation of each step seperately in the aneuploid trajectory as a function of the population size, $N$ (panel A) and as a function of the mutation rate, $\mu$ (panel B) and as a function of the aneuploidy rate, $\delta$ (panel C).
+Solid curves give the fixation time and dashed curves the waiting time.
+Curves of the same color indicate the same base population state;  e.g.
+the blue dashed curve is the waiting time in state $\anwt$ until a succesful individual of type $\anmt$ appears.
+The corresponding solid curve indicates the time to fixation of the population in state $\anwt$.
+Here, $w_\anwt=1.021,\,   w_\anmt=1.025$ and $w_\eumt=1.028$.
+Parameter values in panels where it is not varied, $N=6\cdot 10^6$,\, $\mu=9.6\cdot 10^{-8}$ and $\delta=5.4\cdot 10^{-4}$.
+}
+ \label{fig:fixT vs waitT}
+ \end{figure}
+
+In summary, whereas the two possible trajectories to $\eumt$ do not compete against each other in both \Cref{fig:only mut} and \Cref{fig:no direct mut}, the approximations are doing well only in restricted parameter ranges.
+These tend to be, where the respective parameters are rather small.
+If they grow bigger, they are violating some assumptions of our approximations.
+
+%%%%%%%%%%
 % Discussion

### 2022-03-07 ms/ms.tex
YR revised analysis to the end

@@
 
-Second, we performed simulations without mutations from $\euwt$ to $\eumt$ (\Cref{fig:no direct mut}). 
-% TODO MP: I comment this out - not clear what "Before" refers to, and not sure we need this sentence. what do you think?
-%Before, we had the optimal scenario of waiting until the first successful mutation appears and then waiting until is reaches fixation. Our approximations did quite well there. 
-Here, aneuploidy is transient and fixation of \anwt\ or \anmt\ could be outpaced by the arrival of the next genotype.
-This would explain the discrepancy between the green curves in \Cref{fig:no direct mut}.
-To asses this further, \Cref{fig:fixT vs waitT} provides a clear picture.
-At least for intermediate to high values of the respective range of the parameters, the waiting time until a succesful new type appears is far smaller than that to fixation of the old type.
-Taking this into account, it seems reasonable to ignore the fixation time of $\anwt$ and $\anmt$, which results in the better fitting red curves in \Cref{fig:no direct mut}.
-Inferring $x=0.017$ for $\tilde{T}_{\anmt \rightarrow \eumt}$ to get the best fit in panel B, does not imply that it is also a best fit for panels A and C.
-However, we stop simulations in state $\eumt$ if they have frequency $0.95$.
-
-  The discrepancies in \Cref{fig:no direct mut}C seem to be particularly odd as the approximations are nearly constant and the simulations decreas log-linearly with increasing $\delta$.
-
   \begin{figure}[h!]
@@
 
-In summary, whereas the two possible trajectories to $\eumt$ do not compete against each other in both \Cref{fig:only mut} and \Cref{fig:no direct mut}, the approximations are doing well only in restricted parameter ranges.
-These tend to be, where the respective parameters are rather small.
-If they grow bigger, they are violating some assumptions of our approximations.
+Second, we performed simulations without mutations from $\euwt$ to $\eumt$ (\Cref{fig:no direct mut}). 
+% TODO MP: I comment this out - not clear what "Before" refers to, and not sure we need this sentence. what do you think?
+%Before, we had the optimal scenario of waiting until the first successful mutation appears and then waiting until is reaches fixation. Our approximations did quite well there. 
+Here, aneuploidy is transient and fixation of \anwt\ or \anmt\ could be outpaced by the arrival of the next genotype.
+This would explain the discrepancy between the green curves in \Cref{fig:no direct mut}.
+To asses this further, \Cref{fig:fixT vs waitT} provides a clear picture.
+At least for intermediate to high values of the respective range of the parameters, the waiting time until a new genotype appears and does not go to extinction is far shorter compared  to waiting time for fixation of the previous genotype.
+Taking this into account, it seems reasonable to ignore the fixation time of $\anwt$ and $\anmt$, which results in a better fit between our approximation and simulation results (red curves in \Cref{fig:no direct mut}, although still an overestimation.
+% TODO MP : not sure we should say this without explaining why x=0.017. did we use 0.95 in the approximation or not?
+% Inferring $x=0.017$ for $\tilde{T}_{\anmt \rightarrow \eumt}$ to get the best fit in panel B, does not imply that it is also a best fit for panels A and C.
+However, we stop simulations in state $\eumt$ if they have frequency $0.95$.
+The discrepancies in \Cref{fig:no direct mut}C seem to be particularly odd as the approximations are nearly constant whereas the average time from simulations decreases log-linearly with the aneuploidy rate, $\delta$.
+
+In summary, our approximation fit simulation results with only mutation and without aneuploidy (\Cref{fig:only mut}).
+With aneuploidy and without a direct mutation from \euwt\ to \eumt\, our approximation agrees with simulations only in a restricted parameter range (\Cref{fig:no direct mut}), but otherwise it seems that some of our assumptions are violated.
 

### 2022-03-07 ms/ms.tex
added missing parentheses

@@
 At least for intermediate to high values of the respective range of the parameters, the waiting time until a new genotype appears and does not go to extinction is far shorter compared  to waiting time for fixation of the previous genotype.
-Taking this into account, it seems reasonable to ignore the fixation time of $\anwt$ and $\anmt$, which results in a better fit between our approximation and simulation results (red curves in \Cref{fig:no direct mut}, although still an overestimation.
+Taking this into account, it seems reasonable to ignore the fixation time of $\anwt$ and $\anmt$, which results in a better fit between our approximation and simulation results (red curves in \Cref{fig:no direct mut}), although still an overestimation.
 % TODO MP : not sure we should say this without explaining why x=0.017. did we use 0.95 in the approximation or not?

### 2022-03-08 ms/ms.tex
comment
some more todos covered

@@
 
-\paragraph{Model comparison.} We perform model selection using WAIC, the widely applicable information criterion ~\citep{gelman2013bayesian},
+\paragraph{Model comparison.} The main criterion that we use for model selection is WAIC, the widely applicable information criterion ~\citep{gelman2013bayesian},
 \begin{equation} \label{eq:WAIC}
@@
 where $\theta$ is a parameter vector, and $\mathbb{E}[\cdot]$ and $\mathbb{V}[\cdot]$ are the expectation and variance taken over the posterior distribution, which in practice are approximated using 50,000 samples from the posterior KDE.
+%TODO IK WAIC calculation is still under consideration, update it here after
 WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy~\citep{Kass1995}.
-% TODO IK Describe posterior plots comparison method - Fig 4 & S3
+
+We also plot posterior prediction: we get maximum a posteriori (MAP) estimate of the models, and for each model we execute $10,000$ simulations and plot distribution of time to fixation of \eumt, one of key properties that define the likelihood of the model to fit the data.
+This visualization helps to see what model fits data better, at least for it MAP parameter vector.
+
+Also, for similar models we plot the posterior distribution of parameters, and joint distributions of parameters and if they are similar we say that the models are interchangeable.
+We validate it by comparing HDI (highest density interval) of posterior parameters of the models.
 
@@
 The model fits the data well: in simulations using the MAP parameter estimates, \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\Cref{fig:2n*-fixation}B).
-% TODO IK what happened to the sentence about the fixation on anwt?
-green lines ub fig4?
 Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\Cref{fig:ppc-plot}).
@@
 how do you quantify it?
+TODO add like S5 plot and mean+hdi after simulations with 1000 reps will finish to run.
 WAIC was lower when $\mu$ is fixed~(\Cref{table:WAIC}), but this is not surprising, as WAIC attempt to balance between model fit and model complexity, where the latter takes into account the number of model parameters.
@@
   \label{fig:tau-plots}
-  % TODO IK axis labels like in figure 3, also somehow the graphics looks different from previous figures
-  % fixed means 100% ?
 \end{figure}
@@
 %% generated with diff-tau.ipynb
+% TODO IK 10_000 reps is mocked, should be rerunned and fixed
 \begin{figure}[h!]
@@
   \caption{
-    \textbf{Single-locus model inference with elevated mutation rate in aneuploid cells.}  \textbf{(A-E)} The inferred posterior distributions of model parameters for the single-locus model with different values of $\tau$, the fold-increase in mutation rate in aneuploid cells (\anwt\ and \anmt).
+    \textbf{Single-locus model inference with elevated mutation rate in aneuploid cells.}  \textbf{(A-E)} The inferred posterior distributions of model parameters for the model with different values of $\tau$, the fold-increase in mutation rate in aneuploid cells (\anwt\ and \anmt).
 When the increase in mutation rate is high, $\tau=10$ and $\tau=100$, the inferred mutation (A) and aneuploidy (B) rate tends to be lower.

### 2022-03-09 ms/ms.tex
Some additions

--- 2022-03-07 ms/ms.tex
@@
 
-We assume the waiting time for appearance of \eumt\ from \euwt\ and appearance of \anwt\ from \euwt\ are geometrically distributed with rates $p_{2n^*}$ and $p_{2n+1}$, respectively, where
+We assume the waiting time for appearance of \eumt\ from \euwt\, of \anwt\ from \euwt\, of \anmt from \anwt and of \eumt from \anmt are geometrically distributed with rates $p_{2n^*}$, $p_{2n+1}$, $p_{2n+1^*}$ and $p_{\eumt_\delta}$, respectively, where
 \begin{subequations}
@@
 p_{2n^*} &= 1-(1-2\mu \Delta w_{2n^*})^N \;, \\
-p_{2n+1} &= 1-(1-2\delta\Delta w_{2n+1})^N \;,
+p_{2n+1} &= 1-(1-2\delta\Delta w_{2n+1})^N \; \\
+p_{2n+1^*} &= 1-(1-2\mu \Delta w_{2n+1^*})^N \; \\
+p_{\eumt_\delta} &=1-(1-2\delta \Delta w_{\eumt_\delta})^N.
 \end{align}
 \end{subequations}
-where $\Delta w_i$ denotes the relative difference between the fitness $w_i$ of genotype $i$ and the population mean fitness, $\bar{w}$.
+Here, $\Delta w_i$ denotes the relative difference between the fitness $w_i$ of genotype $i$ and the population mean fitness, $\bar{w}$.
 
@@
 
-% TODO MP: this seems like repetition of things written above, right? should we remove this? or if there is something important here, merge it to previous paragraphs?
-%We are interested in the expected time to fixation of \eumt, and the trajectory leading to this fixation depends on the event that occurs first.
-If \eumt\ appears first and avoids extinction by drift, then it will fix.
-If \anwt\ appears first and avoids extinction, then several other events must occur for fixation of \eumt.
-Thus, in this aneuploid trajectory, we need to approximate the expected waiting time to reach \anmt\ and finally \eumt.
-
-Given a mutation occurs first, % TODO MP: I added this sentence, is that right?
+Given a mutation occurs first, % TODO MP: I added this sentence, is that right? I guess.
 the expected adaptation time via the mutation trajectory is 
@@
 something is wrong here.
- T_A = \frac{1}{1-(1-p_{2n^*})(1-p_{2n+1})} + \frac{1}{p_{2n+1^*}} + \frac{1}{p_{2n^*}} + T_{2n \to 2n+1} + T_{2n+1 \to 2n+1^*} + T_{2n+1^* \to 2n^*} \;.
+ T_A = \frac{1}{1-(1-p_{2n^*})(1-p_{2n+1})} + \frac{1}{p_{{2n+1^*}_\delta}} + \frac{1}{p_{2n^*}} + T_{2n \to 2n+1} + T_{2n+1 \to 2n+1^*} + T_{2n+1^* \to 2n^*} \;.
  \end{equation} % TODO MP do we need the tilde ?
@@
 
+Second, we performed simulations without mutations from $\euwt$ to $\eumt$ (\Cref{fig:no direct mut}). 
+% TODO MP: I comment this out - not clear what "Before" refers to, and not sure we need this sentence. what do you think?
+%Before, we had the optimal scenario of waiting until the first successful mutation appears and then waiting until is reaches fixation. Our approximations did quite well there. 
+Here, aneuploidy is transient and fixation of \anwt\ or \anmt\ could be outpaced by the arrival of the next genotype.
+This would explain the discrepancy between the green curves in \Cref{fig:no direct mut}.
+To asses this further, \Cref{fig:fixT vs waitT} provides a clear picture.
+At least for intermediate to high values of the respective range of the parameters, the waiting time until a succesful new type appears is far smaller than that to fixation of the old type.
+Taking this into account, it seems reasonable to ignore the fixation time of $\anwt$ and $\anmt$, which results in the better fitting red curves in \Cref{fig:no direct mut}.
+Inferring $x=0.017$ for $\tilde{T}_{\anmt \rightarrow \eumt}$ to get the best fit in panel B, does not imply that it is also a best fit for panels A and C.
+However, we stop simulations in state $\eumt$ if they have frequency $0.95$.
+
+  The discrepancies in \Cref{fig:no direct mut}C seem to be particularly odd as the approximations are nearly constant and the simulations decreas log-linearly with increasing $\delta$.
+
  \begin{figure}[h!]
@@
 
-Second, we performed simulations without mutations from $\euwt$ to $\eumt$ (\Cref{fig:no direct mut}). 
-% TODO MP: I comment this out - not clear what "Before" refers to, and not sure we need this sentence. what do you think?
-%Before, we had the optimal scenario of waiting until the first successful mutation appears and then waiting until is reaches fixation. Our approximations did quite well there. 
-Here, aneuploidy is transient and fixation of \anwt\ or \anmt\ could be outpaced by the arrival of the next genotype.
-This would explain the discrepancy between the green curves in \Cref{fig:no direct mut}.
-To asses this further, \Cref{fig:fixT vs waitT} provides a clear picture.
-At least for intermediate to high values of the respective range of the parameters, the waiting time until a new genotype appears and does not go to extinction is far shorter compared  to waiting time for fixation of the previous genotype.
-Taking this into account, it seems reasonable to ignore the fixation time of $\anwt$ and $\anmt$, which results in a better fit between our approximation and simulation results (red curves in \Cref{fig:no direct mut}, although still an overestimation.
-% TODO MP : not sure we should say this without explaining why x=0.017. did we use 0.95 in the approximation or not?
-% Inferring $x=0.017$ for $\tilde{T}_{\anmt \rightarrow \eumt}$ to get the best fit in panel B, does not imply that it is also a best fit for panels A and C.
-However, we stop simulations in state $\eumt$ if they have frequency $0.95$.
-The discrepancies in \Cref{fig:no direct mut}C seem to be particularly odd as the approximations are nearly constant whereas the average time from simulations decreases log-linearly with the aneuploidy rate, $\delta$.
-
-In summary, our approximation fit simulation results with only mutation and without aneuploidy (\Cref{fig:only mut}).
-With aneuploidy and without a direct mutation from \euwt\ to \eumt\, our approximation agrees with simulations only in a restricted parameter range (\Cref{fig:no direct mut}), but otherwise it seems that some of our assumptions are violated.
+In summary, whereas the two possible trajectories to $\eumt$ do not compete against each other in both \Cref{fig:only mut} and \Cref{fig:no direct mut}, the approximations are doing well only in restricted parameter ranges.
+These tend to be, where the respective parameters are rather small.
+If they grow bigger, they are violating some assumptions of our approximations.
 

### 2022-03-13 ms/ms.tex
small changes

--- 2022-03-08 ms/ms.tex
@@
 
-We also plot posterior prediction: we get maximum a posteriori (MAP) estimate of the models, and for each model we execute $10,000$ simulations and plot distribution of time to fixation of \eumt, one of key properties that define the likelihood of the model to fit the data.
-This visualization helps to see what model fits data better, at least for it MAP parameter vector.
-
-Also, for similar models we plot the posterior distribution of parameters, and joint distributions of parameters and if they are similar we say that the models are interchangeable.
-We validate it by comparing HDI (highest density interval) of posterior parameters of the models.
+We also plot posterior predictions: for each model we execute $10,000$ simulations using the MAP parameter estimates and plot the distributions of time to fixation of \eumt, one of key properties of the model likelihood.
+This visualizes the fit of each model to the data.
+Also, for similar models we plot the marginal and joint posterior distributions of the parameters; if these are similar, we consider the models are interchangeable.
+We validate this by comparing HDI (highest density interval) of posterior distributions.
 
@@
   \caption{
-    \textbf{Single-locus model inference with elevated mutation rate in aneuploid cells.}  \textbf{(A-E)} The inferred posterior distributions of model parameters for the model with different values of $\tau$, the fold-increase in mutation rate in aneuploid cells (\anwt\ and \anmt).
+    \textbf{Model with elevated mutation rate in aneuploid cells.}  \textbf{(A-E)} The inferred posterior distributions for models with different values of $\tau$, the fold-increase in mutation rate in aneuploid cells (\anwt\ and \anmt).
 When the increase in mutation rate is high, $\tau=10$ and $\tau=100$, the inferred mutation (A) and aneuploidy (B) rate tends to be lower.

### 2022-03-14 ms/ms.tex
WAIC update

@@
 where $\theta$ is a parameter vector, and $\mathbb{E}[\cdot]$ and $\mathbb{V}[\cdot]$ are the expectation and variance taken over the posterior distribution, which in practice are approximated using 50,000 samples from the posterior KDE.
-%TODO IK WAIC calculation is still under consideration, update it here after
 WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy~\citep{Kass1995}.
@@
 The same prior was used for $w_{2n*}/w_{2n+1*}$.
-This prior resulted in WAIC 71.91, compared to 67.87 with the above informative prior.
-% generated with fixed-mutation-rate-and-WAICs.ipynb TODO IK update the WAIC with 1000 reps run
+This prior resulted in WAIC -57, compared to 295 with the above informative prior.
+% generated with WAICs.ipynb TODO IK update the WAIC with 1000 reps run for the new prior
+%TODO IK: rewrite text about alternative prior, it has lower WAIC, different estimates for mutation rate etc...maybe it should be moved to results section
 We also tested an uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$, or (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$.
@@
 %TODO YR read this 
-%% generated with fixed-mutation-rate.ipynb 
-% TODO IK run for all updated files
+%% generated with WAIC.ipynb 
 \begin{table}[h]

### 2022-03-14 ms/ms.tex
Update ms.tex

--- 2022-03-09 ms/ms.tex
@@
    \caption{
-    \textbf{Genotype fixations for single-locus models with different $\tau$.} We estimated the parameters for different models, each assuming a different value of $\tau$, the fold-increase in mutation rate in aneuploid cells.
+    \textbf{Genotype fixations for models with different $\tau$.} We estimated the parameters for different models, each assuming a different value of $\tau$, the fold-increase in mutation rate in aneuploid cells.
 We then generated 10,000 simulations using the MAP estimate of each model and evaluated the fraction of simulations in which the genotype \euwt\ (\textbf{A}), \anwt\ (\textbf{B}), and \eumt\ (\textbf{C}) is fixed (y-axis) at each generation (x-axis).
@@
 \begin{align}
-p_{2n^*} &= 1-(1-2\mu \Delta w_{2n^*})^N \;, \\
-p_{2n+1} &= 1-(1-2\delta\Delta w_{2n+1})^N \; \\
-p_{2n+1^*} &= 1-(1-2\mu \Delta w_{2n+1^*})^N \; \\
-p_{\eumt_\delta} &=1-(1-2\delta \Delta w_{\eumt_\delta})^N.
+p_{2n, 2n^*} &= 1-\Big(1-2\mu \frac{w_{2n^*}-w_{2n}}{w_{2n}}\Big)^N \;, \\
+p_{2n, 2n+1} &= 1-\Big(1-2\delta \frac{w_{2n+1}-w_{2n}}{w_{2n}}\Big)^N \; \\
+p_{2n+1, 2n+1^*} &= 1-\Big(1-2\mu \frac{w_{2n+1^*}-w_{2n+1}}{w_{2n+1}}\Big)^N \; \\
+p_{2n+1^*, 2n^*} &=1-\Big(1-2\delta \frac{w_{2n^*}-w_{2n+1^*}}{w_{2n+1^*}}\Big)^N \;.
 \end{align}
 \end{subequations}
-Here, $\Delta w_i$ denotes the relative difference between the fitness $w_i$ of genotype $i$ and the population mean fitness, $\bar{w}$.
 
@@
 Thus, we need to find the minimum of these waiting times.
-The minimum of two geometric random variables with rates $p_{2n^*}$ and $p_{2n+1}$ is also a geometric random variable with rate $1-(1-p_{2n^*})(1-p_{2n+1})$, which is the inverse of the expected waiting time until the first event occurs.
+The minimum of two geometric random variables with rates $p_{2n,2n^*}$ and $p_{2n,2n+1}$ is also a geometric random variable with rate $1-(1-p_{2n,2n^*})(1-p_{2n,2n+1})$, which is the inverse of the expected waiting time until the first event occurs.
 
@@
 \begin{equation}\label{eq:T_mut}
-T_{M}=\frac{1}{1-(1-p_{2n^*})(1-p_{2n+1})} + T_{2n \to 2n^*} \;,
+T_{M}=\frac{1}{1-(1-p_{2n,2n^*})(1-p_{2n,2n+1})} + T_{2n, 2n^*} \;,
 \end{equation}
-where $T_{i \to j}$ is the waiting time for fixation of genotype $i$ in a population of genotype $j$, given it starts with a single copy and does not go to extinction by drift.
+where $T_{i, j}$ is the waiting time for fixation of genotype $j$ in a population of genotype $i$, given it starts with a single copy and does not go to extinction by drift.
 Given that aneuploidy occurs first, % TODO MP: I added this sentence, is that right?
@@
 something is wrong here.
- T_A = \frac{1}{1-(1-p_{2n^*})(1-p_{2n+1})} + \frac{1}{p_{{2n+1^*}_\delta}} + \frac{1}{p_{2n^*}} + T_{2n \to 2n+1} + T_{2n+1 \to 2n+1^*} + T_{2n+1^* \to 2n^*} \;.
+ T_A = \frac{1}{1-(1-p_{2n,2n^*})(1-p_{2n,2n+1})} + \frac{1}{p_{2n+1,2n+1^*}} + \frac{1}{p_{2n+1^*,2n^*}} + T_{2n, 2n+1} + T_{2n+1, 2n+1^*} + T_{2n+1^*, 2n^*} \;.
  \end{equation} % TODO MP do we need the tilde ?
@@
 \begin{equation}
-p_A = \frac{p_{2n+1}}{p_{2n^*}+p_{2n+1}} \;.
+p_A = \frac{p_{2n,2n+1}}{p_{2n,2n^*}+p_{2n,2n+1}} \;.
 \end{equation}
@@
 A deterministic-process approach is given by the classical bi-allelic single-locus equation.
-It approximates the expected time for genotype $i$ to reach frequency $x$ when starting from frequency $1-x$ in a population of genotype $j$,
+It approximates the expected time for genotype $j$ to reach frequency $x$ when starting from frequency $y$ in a population of genotype $i$,
  % TODO MP ref for equation? Crow and Kimura? YR: yes
 \begin{equation} \label{eq:T_fix_x}
- T_{i \to j} \approx 2\frac{\log{\big(\frac{x}{1-x}N\big)}}{\log(w_i/w_j)} \;,
+ T_{i, j}(x,y) \approx \frac{\log{\Big(\frac{x(1-y)}{(1-x)y} \Big)}}{\log{\Big(\frac{w_j}{w_i}\Big)}} \;,
 \end{equation} 
-A stochastic-process approach for approximating $T_{i \to j}$ is given by the diffusion-equation approximation \citep[eq.~17]{Kimura1969a}.
+and $T_{i,j} = T_{i,j}(1-1/N, 1/N) \approx 2\log{(N)}/\log{(w_j/w_i)}$.
+A stochastic-process approach for approximating $T_{i, j}$ is given by the diffusion-equation approximation \citep[eq.~17]{Kimura1969a}. % TODO MP give details here?
 
@@
 First, to validate the approximation in \Cref{eq:T_total}, we performed simulations of the full model (\Cref{eq:mutation-aneuploidy-single,eq:drift-single,eq:selection-single}) without aneuploidy, i.e., $\delta=0$ (\Cref{fig:only mut}), and with the other parameters as inferred above.
-The approximation provides a good fit to simulation results except when the effective population size is larger than $5\cdot 10^6$ cells.
+The approximation provides a good fit to simulation results except when the effective population size is above $5\cdot 10^6$ cells.
 In that case, the expected number of new mutants per generation is roughly 0.5, and  we can expect that in many generations the fixation event will start with more than a single \eumt\ cell appearing in the same generation, which violates our assumption that an adaptation event starts from a single mutant cell.
@@
   % --> or by making the y axis in log scale!!
+  % more ticks for x axis
  \label{fig:only mut}
@@
  %\centering
- \includegraphics[width=1\textwidth]{../figures/figgrid_nodirectmut_ylog_uncorr0.jpg}
+ \includegraphics[width=1\textwidth]{../figures/figgrid_nodirectmut_ylog_uncorr.jpg}
  \caption{\textbf{Expected adaptation time without mutations from $\euwt$ to $\eumt$.}

### 2022-03-14 ms/ms.tex
fixed mutation rate models comparison was updated

--- 2022-03-14 ms/ms.tex
@@
 Furthermore, we fitted models with a mutation rate fixed at $\mu$=$10^{-5}$, $10^{-6}$ and $10^{-7}$.
-We inferred similar parameters estimates compared to the model with a free $\mu$ parameter.
-% TODO IK where is this shown?
-how do you quantify it?
-TODO add like S5 plot and mean+hdi after simulations with 1000 reps will finish to run.
+We inferred similar parameters estimates for the model with $\mu=10^{-6}$ compared to the model with a free $\mu$ parameter.
+Models with $\mu$=$10^{-5}$ and $\mu$=$10^{-7}$ inferred different parameters estimates~(\Cref{fig:mu}).
 WAIC was lower when $\mu$ is fixed~(\Cref{table:WAIC}), but this is not surprising, as WAIC attempt to balance between model fit and model complexity, where the latter takes into account the number of model parameters.
@@
   % TODO IK maybe add dashed vertical lines with the same colors to show where the MAP of each distribution is?
+  \end{figure}
 
+% Fig mu comparisons
+%% generated with diff-mutation-rate.ipynb
+\begin{figure}[h!]
+  \centering
+  \begin{subfigure}{0.45\textwidth}
+      \includegraphics[width=\textwidth]{../figures/mu-A.pdf}      
+  \end{subfigure}
+  \begin{subfigure}{0.45\textwidth}
+      \includegraphics[width=\textwidth]{../figures/mu-B.pdf}      
+  \end{subfigure}
+  \\
+   \begin{subfigure}{0.45\textwidth}
+      \includegraphics[width=\textwidth]{../figures/mu-C.pdf}      
+  \end{subfigure}
+    \begin{subfigure}{0.45\textwidth}
+      \includegraphics[width=\textwidth]{../figures/mu-D.pdf}      
+    \end{subfigure}
+  \caption{
+    \textbf{Model with fixed mutation rate.}  \textbf{(A-D)} The inferred posterior distributions for models with free and fixed mutation rate, $\mu$.
+The MAP (maximum a posteriori) and 50\% HDI (highest density interval) for each model are following:
+\textbf{free $\boldsymbol{\mu}$:}
+$\delta=2.749\cdot10^{-3}\ [1.476\cdot10^{-3}-2.822\cdot10^{-3}]$,
+$w_{2n+1}=1.022\ [1.021-1.023]$,
+$w_{2n+1^*}=1.025\ [1.023-1.026]$,
+$w_{2n^*}=1.027\ [1.026-1.029]$;
+\textbf{free $\boldsymbol{\mu}$, run 2:}
+$\delta=1.938\cdot10^{-3}\ [1.338\cdot10^{-3}-2.748\cdot10^{-3}]$,
+$w_{2n+1}=1.022\ [1.02-1.023]$,
+$w_{2n+1^*}=1.025\ [1.023-1.026]$,
+$w_{2n^*}=1.027\ [1.026-1.029]$;
+\textbf{$\boldsymbol{\mu=10^{-5}}$:}
+$\delta=3.089\cdot10^{-3}\ [2.412\cdot10^{-3}-3.169\cdot10^{-3}]$,
+$w_{2n+1}=1.022\ [1.021-1.023]$,
+$w_{2n+1^*}=1.024\ [1.023-1.026]$,
+$w_{2n^*}=1.027\ [1.026-1.028]$;
+\textbf{$\boldsymbol{\mu=10^{-6}}$:}
+$\delta=1.413\cdot10^{-3}\ [1.04\cdot10^{-3}-2.529\cdot10^{-3}]$,
+$w_{2n+1}=1.021\ [1.02-1.023]$,
+$w_{2n+1^*}=1.024\ [1.023-1.026]$,
+$w_{2n^*}=1.028\ [1.026-1.029]$;
+\textbf{$\boldsymbol{\mu=10^{-7}}$:}
+$\delta=3.4\cdot10^{-3}\ [2.043\cdot10^{-3}-3.578\cdot10^{-3}]$,
+$w_{2n+1}=1.019\ [1.017-1.02]$,
+$w_{2n+1^*}=1.026\ [1.024-1.027]$,
+$w_{2n^*}=1.027\ [1.026-1.029]$.
+\label{fig:mu}
+  }
   \end{figure}

### 2022-03-15 notebooks_mp/Time_approximations.py
updates and added the scripts

@@ -1,0 +1,272 @@
+import numpy as np
+import json
+import datetime
+import signal,sys
+import matplotlib.pyplot as plt
+from scipy.integrate import quad 
+from functools import partial
+
+
+
+
+    
+### Most basic functions and approximations:
+    
+    # w1 is 2n+1 fitness
+    # w2 is 2n+1* fitness
+    # w3 is 2n* fitness
+    
+#def __init__(self):
+#    self.T_kimura = np.vectorize(self.T_kimura0) 
+
+#various relative fitnesses: s0=w3-1, s1=w1-1, s2=(w2-w1)/w1, s3=(w3-w2)/w2
+def sv(w1,w2,w3):
+    return w3-1, w1-1, (w2-w1)/w1, (w3-w2)/w2
+
+
+#with Haldane fix_prob
+def waitT(N, rate, s):
+    if np.ndim(rate) == 0:
+        if rate ==0:
+            return 0
+        return 1/(1-(1-rate*np.minimum(1,2*s))**N)
+    rate=np.array([rate]).flatten()
+    zerovec=(rate==0)
+    retvec=np.array([0]*len(rate),dtype='float')
+    retvec[~zerovec]=1/(1-(1-rate[~zerovec]*np.minimum(1,2*s))**N)
+    return retvec
+
+def min_waitT(N,mut_rate, aneu_rate, w1, w3):
+    aneu=(1-(1-aneu_rate*np.minimum(1,2*sv(w1,1,w3)[1]))**N)
+    mut=(1-(1-mut_rate*np.minimum(1,2*sv(w1,1,w3)[0]))**N)
+    comp_para=1-(1-aneu)*(1-mut)
+    return 1/comp_para
+               
+
+
+def waitT_K(N, rate, s):
+    if np.ndim(rate) == 0:
+        if rate ==0:
+            return 0
+        return 1/(1-(1-rate*np.minimum(1,2*s))**N)
+    rate=np.array([rate]).flatten()
+    zerovec=(rate==0)
+    retvec=np.array([0]*len(rate))
+    retvec[~zerovec]=1/(1-(1-rate[~zerovec]*np.expm1(-2 * 1 * s)/np.expm1(-2 * N * s))**N)
+    return retvec
+
+
[…]

### 2022-03-15 notebooks_mp/singlelocusmodelextended.py

@@ -1,0 +1,458 @@
+import numpy as np
+import json
+import datetime
+import signal,sys
+import matplotlib.pyplot as plt
+
+class SingleLocusModelExt:
+    # k*μ is mutation rate of 2n genotype 
+    def __init__(self, k):
+        self.k=k
+        
+    # w1 is 2n+1 fitness
+    # w2 is 2n+1* fitness
+    # w3 is 2n* fitness
+    def run_simulations(self, N, mutation_rate, aneuploidy_rate, aneuploidy_loss_rate, w1, w2, w3, repetitions=1000, max_gen=2500, seed=3,clonal_intf=True):
+        curr_rand_state = np.random.get_state()
+        np.random.seed(seed)
+
+        #0 wt
+        #1 tri
+        #2 tri+
+        #3 +
+        Mt = np.diag(np.repeat(0.,5))
+        Mt[0][1] = aneuploidy_rate
+        Mt[1][2] = mutation_rate
+        Mt[2][3] = aneuploidy_loss_rate 
+        Mt[0][4] = self.k*mutation_rate  
+        w = [1., w1, w2, w3, w3]
+        times, p, non_aneu_fix= self._simulation(N, w, Mt, repetitions=repetitions, max_gen=max_gen,clonal_intf=clonal_intf)
+        
+        np.random.set_state(curr_rand_state)
+        return (times,p,non_aneu_fix)   
+
+    
+    #have side effect on M, but not harmfull, can run twice simulation
+    def _simulation(self, N, w, Mt ,repetitions=1000, fixation=0.95, max_gen=2500,clonal_intf=True):
+        assert N > 0
+        N = np.uint64(N)
+        L = len(w)
+        S = np.diag(w)
+        if clonal_intf==False:
+            Mnotr=Mt+1e-5
+            Mnotr=Mnotr-1e-5
+            Mnotr[0][1]=0
+            Mnotr[0][4]=0
+            Mnotr=Mnotr.transpose()
+            np.fill_diagonal(Mnotr,0) #neutralling side effect
+            np.fill_diagonal(Mnotr,1-Mnotr.sum(axis=0)) #side effect
+            Enotr= Mnotr @ S
+
+        Mt = Mt.transpose() 
+        np.fill_diagonal(Mt,0) #neutralling side effect
+        np.fill_diagonal(Mt,1-Mt.sum(axis=0)) #side effect
+
+        Et = Mt @ S
+  
+        # rows are genotypes, cols are repretitions
+        n = np.zeros((L, repetitions))
[…]

### 2022-03-16 ms/ms.tex
add question todo
forgot files

@@
 
-Compared to other priors we tested, this prior produced lower WAIC, better posterior prediction plots, and more stable parameter estimates, as follows.
-First, we tried to use additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$ ~(\Cref{fig:growth-curves}L).
+Also we tried extended informative prior that use additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$ ~(\Cref{fig:growth-curves}L).
 The same prior was used for $w_{2n*}/w_{2n+1*}$.
-This prior resulted in WAIC -57, compared to 295 with the above informative prior.
-% generated with WAICs.ipynb TODO IK update the WAIC with 1000 reps run for the new prior
-%TODO IK: rewrite text about alternative prior, it has lower WAIC, different estimates for mutation rate etc...maybe it should be moved to results section
-We also tested an uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$, or (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$.
+
+Also we tested an uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$, or (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$.
 In both cases the inference algorithm failed to converge.
@@
 
+% FIG posterior alt-prior
+%% generated with alt-prior.ipynb
+%% TODO IK - update the figure to look like Figure 2
+\begin{figure}[p]
+  \centering
+\includegraphics[width=0.9\textwidth]{../figures/posterior-alt.pdf}
+  \caption{
+  \textbf{Posterior distribution of parameters for extended-prior model.}
+On the diagonal, the inferred posterior distribution of each model parameter. 
+Below the diagonal, the inferred joint posterior distribution of pairs of model parameters (dark purple and bright yellow for low and high density, respectively).
+Red markers and orange lines for the joint MAP estimate (which may differ from the marginal MAP, as the marginal distribution integrates over all other parameters).
+} 
+  \label{fig:posterior-alt}
+\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%
+
 \paragraph{Model checking and comparison.}
@@
 WAIC was lower when $\mu$ is fixed~(\Cref{table:WAIC}), but this is not surprising, as WAIC attempt to balance between model fit and model complexity, where the latter takes into account the number of model parameters.
+
+% generated with alt-prior.ipynb
+At the end we inferred parameters for \emph{extended informative prior} model (see \emph{Prior distributions}) with different priors for $w_{2n+1^*}$ and $w_{2n*}$.
+The estimated mutation rate was much lower compared to the basic model, $\mu=2.379\cdot10^{-9}\ [2.267\cdot10^{-9}-2.458\cdot10^{-9}]$.
+Other parameter estimates are: $\delta=2.636\cdot10^{-3}\ [2.245\cdot10^{-3}-2.919\cdot10^{-3}]$,
+$w_{2n+1}=1.022\ [1.021-1.023]$,
+$w_{2n+1^*}=1.032\ [1.03-1.035]$,
+$w_{2n^*}=1.033\ [1.031-1.036]$. 
+The WAIC of the model is much lower: -57 compared to 295~(\Cref{table:WAIC}).
 
@@
     \textbf{Fitness estimation from growth curves.}
-    % TODO IK caption should explain what each panel shows.
-    \textbf{(A-D)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{39}{\celsius}. $\hat w_{2n+1}/w_{2n}$=1.024 (95\% CI: 0.959 - 1.115).
+    \textbf{(A-D)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{39}{\celsius}. $\hat w_{2n+1}/w_{2n}$=1.024 (95\% CI: 0.959 - 1.115). \texttt{Curveball} 
     \textbf{(E-H)} Fitness estimation from growth curves of \anwt\ and \eumt\ at \SI{39}{\celsius}. $\hat w_{2n*}/w_{2n+1}$=1.033 (95\% CI: 1.027 - 1.041).
@@
 See \emph{Models and Methods, Prior distributions} for more details.
+\textbf{(A,B;E,F)} The mean of mono-cultures growth curves, generated by \texttt{Curveball} model fitted to the appropriate data growth curves.
+%TODO YR what is vertical lines?
+95% CI?
+\textbf{(C,G)} The mixed-culture prediction for the strains from A,B and E,F respectively. 6375 generated curves. \textbf{(D,H)} The appropriate relative fitness distribution
 } 
@@
 \caption*{
-	TODO caption
-	%TODO IK caption; what is tau in models 1-5?
+	WAIC (widely applicable information criterion; \autoref{eq:WAIC})~\citep{gelman2013bayesian} values for models: (1) without aneuploidy;  (2-4) with aneuploidy, fixed mutation rate, and $\tau=1$; (5-10) with aneuploidy, free mutation rate model parameter, and different $\tau$.
+WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy and a difference of 2 is a popular threshold for model comparison~\citep{Kass1995}.
 } 

### 2022-03-21 ms/ms.tex
revised a bit

@@
 That is, we assume chromosomes are gained and lost at the same rate, and we neglect events that form a less-fit genotype (i.e., \anwt\; to \euwt\; and \eumt\; to \anmt).
-The fitness values of the four genotypes are given by \Cref{table:single-locus}.
+The fitness values of the four genotypes are given by \Cref{table:fitness}.
 
-%%% Table: single-locus model
+%%% Table: fitness
 \begin{table}[h]
 \centering
-\caption{\textbf{Single-locus model fitness values.}}
+\caption{\textbf{Fitness values.}}
 \begin{tabular}{lllll}
@@
 \end{tabular}
-\label{table:single-locus}
+\label{table:fitness}
 \caption*{
@@
     \end{equation}
-where the fitness values $w_i$ are given in \Cref{table:single-locus} and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
+where the fitness values $w_i$ are given in \Cref{table:fitness} and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
 The effect of mutation and aneuploidy on genotype frequencies is given by
@@
 
-% YR: this whole section doesn't really make sense.
-We know there was aneuploidy.
-If we assume delta=0 that means that we think this aneuploidy was not relevant to the adaptation process.
-But then why do we take generations 1700 and 2300 to be the time for 2n* fixation?
-These times were not observed for fixation of a mutant but rather for elimination of aneuploidy.
-If we disregard aneuploidy, why do we still consider these times?
-%For a model without aneuploidy (that is, when the aneuploidy rate is fixed at zero, $\delta=0$), 
-%the likelihood is similarly approximated by
-%\begin{equation}\begin{aligned}
-%\label{eq:heatstress-noaneuploidy-likelihood}
-%\likelihood_{!}(\theta) = &\ 
-%	1 - 
-%	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\}\big) - \\
-%&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\}\big) + \\
-%&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\}\big)
-%\;.
-%\end{aligned}\end{equation}
+For a model without aneuploidy (that is, when the aneuploidy rate is fixed at zero, $\delta=0$), we disregard the increased expression in chromosome III and the growth advantage measured in generation 450, and focus on the growth advantage measured in later generations, presumably due to a beneficial mutation.
+Therefore, the likelihood is approximated by
+\begin{equation}\begin{aligned}
+\label{eq:heatstress-noaneuploidy-likelihood}
+\likelihood_{!}(\theta) = &\ 
+	1 - 
+	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\}\big) - \\
+&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\}\big) + \\
+&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\}\big)
+\;.
+\end{aligned}\end{equation}
 
@@
 We also plot posterior predictions: for each model we execute $10,000$ simulations using the MAP parameter estimates and plot the distributions of time to fixation of \eumt, one of key properties of the model likelihood.
-This visualizes the fit of each model to the data.
+These plots visualize the fit of each model to the data.
 Also, for similar models we plot the marginal and joint posterior distributions of the parameters; if these are similar, we consider the models are interchangeable.
@@
 Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix.
+
 We sampled 10,000 samples to use as a prior distribution~(\Cref{fig:growth-curves}).
-We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$ ~(\Cref{fig:growth-curves}-D).
+We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$~(\Cref{fig:growth-curves}-D, assuming $w_{2n}=1$).
 In lieu of a more suitable prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$.
 To increase computational efficiency, we also assumed $w(2n*)>w(2n+1*)>w(2n+1)>w(2n)$; running the inference without this assumption produced similar results. 
-
-Also we tried extended informative prior that use additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$ ~(\Cref{fig:growth-curves}L).
+We also examined an extended informative prior that uses additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$~(\Cref{fig:growth-curves}L).
 The same prior was used for $w_{2n*}/w_{2n+1*}$.
 
-Also we tested an uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$, or (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$.
-In both cases the inference algorithm failed to converge.
+As a control, we tested an uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$, or (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$.
+In these cases the inference algorithm failed to converge.
 
@@
 $w_{2n^*}=1.028\ [1.026-1.029]$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$. 
-Thus, we can infer that the benefit and cost of trisomy was $b=2.5\%$ and $c=0.3\%$, and the benefit of the beneficial mutation was $2.8\%$~(\Cref{table:single-locus}).
+Thus, we can infer that the benefit and cost of trisomy was $b=2.5\%$ and $c=0.3\%$, and the benefit of the beneficial mutation was $2.8\%$~(\Cref{table:fitness}).
 
@@
 % commented out, see remark above when describing likelihood of delta=0.
-%Moreover, a model without aneuploidy (where the aneuploidy rate is fixed at zero, $\delta=0$), fails to explain the experimental observations~(Figure~\ref{fig:2n*-fixation}).
+Moreover, a model without aneuploidy (where the aneuploidy rate is fixed at zero, $\delta=0$), fails to explain the experimental observations~(Figure~\ref{fig:2n*-fixation}).
 %% generated with MAP-and-hdi.ipynb
-%The estimated mutation rate without aneuploidy is $\mu=7.979\cdot10^{-9}\ [7.903\cdot 10^{-9}-8.135\cdot10^{-9}]$, much lower compared to a model with aneuploidy and suggesting a target size of just 40.
-%The fitness of the mutant is also much lower at $w_{2n^*}=1.013\ [1.012-1.013]$.
-%This is because, without aneuploidy, a high mutation rate or fitness effect will lead to faster appearance and fixation of \eumt\ than in the experimental observations.
+The estimated mutation rate without aneuploidy is $\mu=7.979\cdot10^{-9}\ [7.903\cdot 10^{-9}-8.135\cdot10^{-9}]$, much lower compared to a model with aneuploidy and suggesting a target size of just 40.
+The fitness of the mutant is also much lower at $w_{2n^*}=1.013\ [1.012-1.013]$.
+This is because, without aneuploidy, a high mutation rate or fitness effect will lead to faster appearance and fixation of \eumt\ than in the experimental observations.
 Even with these lower estimates, the model fit is worse than that of a model with aneuploidy~(\Cref{fig:2n*-fixation}).
@@
 % generated with alt-prior.ipynb
-At the end we inferred parameters for \emph{extended informative prior} model (see \emph{Prior distributions}) with different priors for $w_{2n+1^*}$ and $w_{2n*}$.
-The estimated mutation rate was much lower compared to the basic model, $\mu=2.379\cdot10^{-9}\ [2.267\cdot10^{-9}-2.458\cdot10^{-9}]$.
+Finally, we estimated the parameters under an extended informative prior (see \emph{Prior distributions}), with different priors for $w_{2n+1^*}$ and $w_{2n*}$.
+The estimated mutation rate was much lower compared to the basic prior, with $\mu=2.379\cdot10^{-9}\ [2.267\cdot10^{-9}-2.458\cdot10^{-9}]$.
 Other parameter estimates are: $\delta=2.636\cdot10^{-3}\ [2.245\cdot10^{-3}-2.919\cdot10^{-3}]$,
@@
 
-% Fig single-locus model with and without aneuploidy
+% Fig model with and without aneuploidy
 %% generated with with-aneuploidy_vs_no_aneuploidy.ipynb
@@
   \caption{
-    \textbf{Single-locus model fit with and without aneuploidy.}
-    \textbf{(A)} The distribution of time to fixation of \eumt\ (i.e., adaptation time) in $10,000$ simulations of the single-locus model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange), and a model fitted to the data assuming $\delta=0$ (green).
+    \textbf{Model fit with and without aneuploidy.}
+    \textbf{(A)} The distribution of time to fixation of \eumt\ (i.e., adaptation time) in $10,000$ simulations of the model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange), and a model fitted to the data assuming $\delta=0$ (green).
 In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and another by generation 2,350 (dashed lines) but not before generation 450.
@@
      }
+     % TODO should we remove this figure altoghther? change it? because we removed no-aneuploidy model...? should we merge this with the next figure?
   \label{fig:2n*-fixation}
@@
   \caption{
-  \textbf{Posterior genotype frequency dynamics for the single-locus model.} 
+  \textbf{Posterior genotype frequency dynamics for the model.} 
   The posterior prediction for the frequencies of the four genotypes over time.
-Each of the 20 curves is the average of 10,000 simulations of the single-locus model using parameters drawn from the posterior distribution.
+Each of the 20 curves is the average of 10,000 simulations of the model using parameters drawn from the posterior distribution.
   }
@@
 \subsection*{Analysis}
-
-Here, we aim to approximate the adaptation time (i.e., time to fixation of \eumt) and analyse how it is affected by the model parameters.
-As a stochastic model, there are two events that can occur on the \euwt\ genotype background.
-The first is a mutation that produces a \eumt\ genotype.
-The second is a missegregation event that produces a \anwt\ genotype.
-The probabilities with which these events occur is significantly different.
-We are interested in the average waiting time until one of these two genotype appears and then leads to adaptation, i.e., appearance and fixation of the \eumt; in the second case, appearance of \anwt\ will be followed by a mutation to \anmt\ and loss of aneuploidy to \eumt.
-
-We assume the waiting time for appearance of \eumt\ from \euwt\ and appearance of \anwt\ from \euwt\ are geometrically distributed with rates $p_{2n^*}$ and $p_{2n+1}$, respectively, where
-\begin{subequations}
-\begin{align}
-p_{2n^*} &= 1-(1-2\mu \Delta w_{2n^*})^N \;, \\
-p_{2n+1} &= 1-(1-2\delta\Delta w_{2n+1})^N \;,
-\end{align}
-\end{subequations}
-where $\Delta w_i$ denotes the relative difference between the fitness $w_i$ of genotype $i$ and the population mean fitness, $\bar{w}$.
-
-However, we have two competing events: if one event occurs (e.g., mutation), then the other event (e.g., aneuploidy) is unlikely to occur.
-Thus, we need to find the minimum of these waiting times.
-The minimum of two geometric random variables with rates $p_{2n^*}$ and $p_{2n+1}$ is also a geometric random variable with rate $1-(1-p_{2n^*})(1-p_{2n+1})$, which is the inverse of the expected waiting time until the first event occurs.
-
-% TODO MP: this seems like repetition of things written above, right? should we remove this? or if there is something important here, merge it to previous paragraphs?
-%We are interested in the expected time to fixation of \eumt, and the trajectory leading to this fixation depends on the event that occurs first.
-If \eumt\ appears first and avoids extinction by drift, then it will fix.
-If \anwt\ appears first and avoids extinction, then several other events must occur for fixation of \eumt.
-Thus, in this aneuploid trajectory, we need to approximate the expected waiting time to reach \anmt\ and finally \eumt.
-
-Given a mutation occurs first, % TODO MP: I added this sentence, is that right?
-the expected adaptation time via the mutation trajectory is 
-\begin{equation}\label{eq:T_mut}
-T_{M}=\frac{1}{1-(1-p_{2n^*})(1-p_{2n+1})} + T_{2n \to 2n^*} \;,
-\end{equation}
-where $T_{i \to j}$ is the waiting time for fixation of genotype $i$ in a population of genotype $j$, given it starts with a single copy and does not go to extinction by drift.
-Given that aneuploidy occurs first, % TODO MP: I added this sentence, is that right?
-the expected adaptation time via the aneuploidy trajectory is approximated by
-\begin{equation}\label{eq:T_aneu} 
-% TODO MP: p_{2n+1^*} is not defined, and why is there 1/p_{2n^*} here?
-shouldn't it have the waiting time for 2n+1* to appear on 2n+1 and then waiting time for 2n* to appear on 2n+1*?
-something is wrong here.
- T_A = \frac{1}{1-(1-p_{2n^*})(1-p_{2n+1})} + \frac{1}{p_{2n+1^*}} + \frac{1}{p_{2n^*}} + T_{2n \to 2n+1} + T_{2n+1 \to 2n+1^*} + T_{2n+1^* \to 2n^*} \;.
- \end{equation} % TODO MP do we need the tilde ?
-Finally, the expected adaptation time is 
- \begin{equation}\label{eq:T_total}
- T = p_A T_{A}+(1-p_A) T_{M} \;,
- \end{equation}
- where $p_A$ is he probability that aneuploidy will be the first event to occur,
-\begin{equation}
-p_A = \frac{p_{2n+1}}{p_{2n^*}+p_{2n+1}} \;.
-\end{equation}
-
-We assume that the waiting time until a new genotype occurs is much longer than the waiting time for its fixation in the population.
-Nevertheless, these fixation times must be included in a rigorous analysis.
-There are two ways to compute the fixation times and they differ in the assumption of the underlying process.
-A deterministic-process approach is given by the classical bi-allelic single-locus equation.
-It approximates the expected time for genotype $i$ to reach frequency $x$ when starting from frequency $1-x$ in a population of genotype $j$,
- % TODO MP ref for equation? Crow and Kimura? YR: yes
-\begin{equation} \label{eq:T_fix_x}
- T_{i \to j} \approx 2\frac{\log{\big(\frac{x}{1-x}N\big)}}{\log(w_i/w_j)} \;,
-\end{equation} 
-A stochastic-process approach for approximating $T_{i \to j}$ is given by the diffusion-equation approximation \citep[eq.~17]{Kimura1969a}.
-
-% TODO MP - the following remark should be removed.
-Instead, wherever you use one of these assumptions in the above text, you should explicitly say so (rather then say the assumptions are implicit).
-Then you can discuss here how these assumptions may not be met when the parameters are so and so, and that you will point it out in the results or something like that.
- %\begin{remark} There are several implicit assumptions that the equations hold like this.
- %\begin{enumerate}
- %\item The two processes fixation and occurrence of a new type are sequential and do not interfere with each other.
- %\item There is no clonal interference.
- %\item The fixation process starts from exactly one individual. 
- %\end{enumerate}
- %There are certainly combinations of the mutation rate, the aneuploidy rate, the selection strength and the population size for which at least one of these assumptions is violated.
-Thus, \Cref{eq:T_total} may not yield a good approximation to the actual time to the adapted state.
- %\end{remark}
-
-First, to validate the approximation in \Cref{eq:T_total}, we performed simulations of the full model (\Cref{eq:mutation-aneuploidy-single,eq:drift-single,eq:selection-single}) without aneuploidy, i.e., $\delta=0$ (\Cref{fig:only mut}), and with the other parameters as inferred above.
-The approximation provides a good fit to simulation results except when the effective population size is larger than $5\cdot 10^6$ cells.
-In that case, the expected number of new mutants per generation is roughly 0.5, and  we can expect that in many generations the fixation event will start with more than a single \eumt\ cell appearing in the same generation, which violates our assumption that an adaptation event starts from a single mutant cell.
-
- \begin{figure}[h!]
- \centering
- \includegraphics[width=\textwidth]{../figures/figgrid_noaneu.pdf}
- \caption{\textbf{Expected adaptation time without aneuploidy.}
-  The expected time to appearance and fixation of \eumt\ as a function of the population size, $N$ (\textbf{A}) or as a function of the mutation rate, $\mu$ (\textbf{B}).
-Mean and 95\% CI from simulations in blue.
-Analytic approximation (\Cref{eq:T_total}) in orange.
-Here, $w_\anwt=1.021,\,   w_\anmt=1.025$ and $w_\eumt=1.028$.
-In panel A, $\mu=9.6\cdot 10^{-8}$.
-In panel B, $N=6\cdot 10^6$.
-  }
-  %TODO MP 
-  % which approximation do you use here for T_i->j ?
-  % The error bars a quite large, can you make them smaller by running more simulations?
-  % --> or by making the y axis in log scale!!
- \label{fig:only mut}
- \end{figure}
-
- \begin{figure}[h!]
- %\centering
- \includegraphics[width=1\textwidth]{../figures/figgrid_nodirectmut_ylog_uncorr0.jpg}
- \caption{\textbf{Expected adaptation time without mutations from $\euwt$ to $\eumt$.}
-  The expected time to appearance and fixation of \eumt\ as a function of the population size, $N$ (\textbf{A}) and as a function of the mutation rate, $\mu$ (\textbf{B}) and as a function of the aneuploidy rate, $\delta$ ({C}).
-Blue error bars for mean and 95\% CI from simulations.
-Green line for the analytic approximation (\Cref{eq:T_total}).
-% TODO why green?
-in previous figure it was orange.
-  Red curve for the approximation with $T_{\euwt \to \anwt}=T_{\anwt \to \anmt}=0$. 
-  Orange curve for the approximation with $x=0.017$ in \Cref{eq:T_fix_x} for the fixation time of $\eumt$; we found this gives a good fit in panel B.
-  Here, $w_\anwt=1.021,\, w_\anmt=1.025$ and $w_\eumt=1.028$, $N=6\cdot 10^6$,\, $\mu=9.6\cdot 10^{-8}$ and $\delta=5.4\cdot 10^{-4}$ unless parameter is varied on the x-axis.
-  }
- \label{fig:no direct mut}
- \end{figure}
-
-  \begin{figure}[h!]
- %\centering
- \includegraphics[width=1\textwidth]{../figures/figgrid_fixT_vs_waitT.jpg}
- \caption{\textbf{Expected time of fixation and waiting time events on the aneuploid trajectory.}
-  The expected time to appearance and fixation of each step seperately in the aneuploid trajectory as a function of the population size, $N$ (panel A) and as a function of the mutation rate, $\mu$ (panel B) and as a function of the aneuploidy rate, $\delta$ (panel C).
-Solid curves give the fixation time and dashed curves the waiting time.
-Curves of the same color indicate the same base population state;  e.g.
-the blue dashed curve is the waiting time in state $\anwt$ until a succesful individual of type $\anmt$ appears.
-The corresponding solid curve indicates the time to fixation of the population in state $\anwt$.
-Here, $w_\anwt=1.021,\,   w_\anmt=1.025$ and $w_\eumt=1.028$.
-Parameter values in panels where it is not varied, $N=6\cdot 10^6$,\, $\mu=9.6\cdot 10^{-8}$ and $\delta=5.4\cdot 10^{-4}$.
-}
- \label{fig:fixT vs waitT}
- \end{figure}
-
-Second, we performed simulations without mutations from $\euwt$ to $\eumt$ (\Cref{fig:no direct mut}). 
-% TODO MP: I comment this out - not clear what "Before" refers to, and not sure we need this sentence. what do you think?
-%Before, we had the optimal scenario of waiting until the first successful mutation appears and then waiting until is reaches fixation. Our approximations did quite well there. 
-Here, aneuploidy is transient and fixation of \anwt\ or \anmt\ could be outpaced by the arrival of the next genotype.
-This would explain the discrepancy between the green curves in \Cref{fig:no direct mut}.
-To asses this further, \Cref{fig:fixT vs waitT} provides a clear picture.
-At least for intermediate to high values of the respective range of the parameters, the waiting time until a new genotype appears and does not go to extinction is far shorter compared  to waiting time for fixation of the previous genotype.
-Taking this into account, it seems reasonable to ignore the fixation time of $\anwt$ and $\anmt$, which results in a better fit between our approximation and simulation results (red curves in \Cref{fig:no direct mut}), although still an overestimation.
-% TODO MP : not sure we should say this without explaining why x=0.017. did we use 0.95 in the approximation or not?
-% Inferring $x=0.017$ for $\tilde{T}_{\anmt \rightarrow \eumt}$ to get the best fit in panel B, does not imply that it is also a best fit for panels A and C.
-However, we stop simulations in state $\eumt$ if they have frequency $0.95$.
-The discrepancies in \Cref{fig:no direct mut}C seem to be particularly odd as the approximations are nearly constant whereas the average time from simulations decreases log-linearly with the aneuploidy rate, $\delta$.
-
-In summary, our approximation fit simulation results with only mutation and without aneuploidy (\Cref{fig:only mut}).
-With aneuploidy and without a direct mutation from \euwt\ to \eumt\, our approximation agrees with simulations only in a restricted parameter range (\Cref{fig:no direct mut}), but otherwise it seems that some of our assumptions are violated.
 
@@
 See \emph{Models and Methods, Prior distributions} for more details.
-\textbf{(A,B;E,F)} The mean of mono-cultures growth curves, generated by \texttt{Curveball} model fitted to the appropriate data growth curves.
-%TODO YR what is vertical lines?
-95% CI?
-\textbf{(C,G)} The mixed-culture prediction for the strains from A,B and E,F respectively. 6375 generated curves. \textbf{(D,H)} The appropriate relative fitness distribution
+\textbf{(A,B;E,F)} Mono-culture growth curve data (markers) and best-fit growth models (lines).
+\textbf{(C,G)} The mixed-culture prediction for the strains from A,B and E,F respectively, 6,375 generated curves.
+\textbf{(D,H)} The relative fitness distribution for \anwt\ relative to \euwt\ (panel D) and \eumt\ relative to \anwt\ (panel H).
+Figures generated by \texttt{Curveball}.
+% TODO IK please fix the range of x axis in DIt definitely doesnt need to go beyond 1.25...
 } 

### 2022-03-23 ms/ms.tex
revised a bit

@@
   \caption{
-  \textbf{Posterior distribution of parameters for extended-prior model.}
+  \textbf{Posterior distribution of parameters inferred with the extended prior distribution.}
 On the diagonal, the inferred posterior distribution of each model parameter. 
@@
 
-% commented out, see remark above when describing likelihood of delta=0.
 Moreover, a model without aneuploidy (where the aneuploidy rate is fixed at zero, $\delta=0$), fails to explain the experimental observations~(Figure~\ref{fig:2n*-fixation}).
@@
 Figures generated by \texttt{Curveball}.
-% TODO IK please fix the range of x axis in DIt definitely doesnt need to go beyond 1.25...
+% TODO IK please fix the range of x axis in D; It definitely doesnt need to go beyond 1.25...
 } 

### 2022-03-23 ms/ms.tex
use macro for genotypes

@@
 \begin{tabular}{lllll}
-\emph{Genotype} $i$ & $2n$ & $2n+1$ & $2n+1^*$ & $2n^*$ \\
+\emph{Genotype} $i$ & \euwt & \anwt & \anmt & \eumt \\
 \hline
@@
 
-\paragraph{Prior distributions.} We used informative prior distributions for $w_{2n+1}=1-c+b$, $w_{2n+1^*}=(1+s)(1-c)+b$ and $w_{2n^*}=1+s$, which we estimated from growth curves data from mono-culture growth experiments previously reported by \citet[Figs.
+\paragraph{Prior distributions.} We used informative prior distributions for $w_{\anwt}=1-c+b$, $w_{\anmt}=(1+s)(1-c)+b$ and $w_{\eumt}=1+s$, which we estimated from growth curves data from mono-culture growth experiments previously reported by \citet[Figs.
 3C, 4A, and S2]{Yona2012}.
@@
 We sampled 10,000 samples to use as a prior distribution~(\Cref{fig:growth-curves}).
-We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$~(\Cref{fig:growth-curves}-D, assuming $w_{2n}=1$).
-In lieu of a more suitable prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$.
-To increase computational efficiency, we also assumed $w(2n*)>w(2n+1*)>w(2n+1)>w(2n)$; running the inference without this assumption produced similar results. 
-We also examined an extended informative prior that uses additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$~(\Cref{fig:growth-curves}L).
-The same prior was used for $w_{2n*}/w_{2n+1*}$.
+We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{\anwt}$~(\Cref{fig:growth-curves}-D, assuming $w_{\euwt}=1$).
+In lieu of a more suitable prior, we used the same prior for $w_{\anmt}$ and $w_{\eumt}$.
+To increase computational efficiency, we also assumed $w(\eumt)>w(\anmt)>w(\anwt)>w(\euwt)$; running the inference without this assumption produced similar results. 
+We also examined an extended informative prior that uses additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{\eumt}/w_{\anwt}$~(\Cref{fig:growth-curves}L).
+The same prior was used for $w_{\eumt}/w_{\anmt}$.
 
-As a control, we tested an uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$, or (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$.
+As a control, we tested an uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{\anwt}$, $w_{\anmt}$, $w_{\eumt}$, or (ii) only for $w_{\anmt}$, $w_{\eumt}$, using the above informative prior for $w_{\anwt}$.
 In these cases the inference algorithm failed to converge.
@@
 The estimated mutation rate, $\mu=2.942\cdot10^{-6}\ [2.017\cdot10^{-7}-4.15\cdot10^{-6}]$, corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
-The estimated fitness values are $w_{2n+1}=1.022\ [1.021-1.023]$,
-$w_{2n+1^*}=1.025\ [1.024-1.026]$,
-$w_{2n^*}=1.028\ [1.026-1.029]$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$. 
+The estimated fitness values are $w_{\anwt}=1.022\ [1.021-1.023]$,
+$w_{\anmt}=1.025\ [1.024-1.026]$,
+$w_{\eumt}=1.028\ [1.026-1.029]$, all relative to the fitness of \euwt, which is set to $w_{\euwt}=1$. 
 Thus, we can infer that the benefit and cost of trisomy was $b=2.5\%$ and $c=0.3\%$, and the benefit of the beneficial mutation was $2.8\%$~(\Cref{table:fitness}).
@@
 The estimated mutation rate without aneuploidy is $\mu=7.979\cdot10^{-9}\ [7.903\cdot 10^{-9}-8.135\cdot10^{-9}]$, much lower compared to a model with aneuploidy and suggesting a target size of just 40.
-The fitness of the mutant is also much lower at $w_{2n^*}=1.013\ [1.012-1.013]$.
+The fitness of the mutant is also much lower at $w_{\eumt}=1.013\ [1.012-1.013]$.
 This is because, without aneuploidy, a high mutation rate or fitness effect will lead to faster appearance and fixation of \eumt\ than in the experimental observations.
@@
 % generated with alt-prior.ipynb
-Finally, we estimated the parameters under an extended informative prior (see \emph{Prior distributions}), with different priors for $w_{2n+1^*}$ and $w_{2n*}$.
+Finally, we estimated the parameters under an extended informative prior (see \emph{Prior distributions}), with different priors for $w_{\anmt}$ and $w_{\eumt}$.
 The estimated mutation rate was much lower compared to the basic prior, with $\mu=2.379\cdot10^{-9}\ [2.267\cdot10^{-9}-2.458\cdot10^{-9}]$.
 Other parameter estimates are: $\delta=2.636\cdot10^{-3}\ [2.245\cdot10^{-3}-2.919\cdot10^{-3}]$,
-$w_{2n+1}=1.022\ [1.021-1.023]$,
-$w_{2n+1^*}=1.032\ [1.03-1.035]$,
-$w_{2n^*}=1.033\ [1.031-1.036]$. 
+$w_{\anwt}=1.022\ [1.021-1.023]$,
+$w_{\anmt}=1.032\ [1.03-1.035]$,
+$w_{\eumt}=1.033\ [1.031-1.036]$. 
 The WAIC of the model is much lower: -57 compared to 295~(\Cref{table:WAIC}).
@@
     \textbf{Fitness estimation from growth curves.}
-    \textbf{(A-D)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{39}{\celsius}. $\hat w_{2n+1}/w_{2n}$=1.024 (95\% CI: 0.959 - 1.115). \texttt{Curveball} 
-    \textbf{(E-H)} Fitness estimation from growth curves of \anwt\ and \eumt\ at \SI{39}{\celsius}. $\hat w_{2n*}/w_{2n+1}$=1.033 (95\% CI: 1.027 - 1.041).
+    \textbf{(A-D)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{39}{\celsius}. $\hat w_{\anwt}/w_{\euwt}$=1.024 (95\% CI: 0.959 - 1.115). \texttt{Curveball} 
+    \textbf{(E-H)} Fitness estimation from growth curves of \anwt\ and \eumt\ at \SI{39}{\celsius}. $\hat w_{\eumt}/w_{\anwt}$=1.033 (95\% CI: 1.027 - 1.041).
     Growth curves previously described in \citet[Figs. 3C, 4A, and S2]{Yona2012}.
@@
 $\delta=2.749\cdot10^{-3}\ [1.476\cdot10^{-3}-2.822\cdot10^{-3}]$,
-$w_{2n+1}=1.022\ [1.021-1.023]$,
-$w_{2n+1^*}=1.025\ [1.023-1.026]$,
-$w_{2n^*}=1.027\ [1.026-1.029]$;
+$w_{\anwt}=1.022\ [1.021-1.023]$,
+$w_{\anmt}=1.025\ [1.023-1.026]$,
+$w_{\eumt}=1.027\ [1.026-1.029]$;
 \textbf{free $\boldsymbol{\mu}$, run 2:}
 $\delta=1.938\cdot10^{-3}\ [1.338\cdot10^{-3}-2.748\cdot10^{-3}]$,
-$w_{2n+1}=1.022\ [1.02-1.023]$,
-$w_{2n+1^*}=1.025\ [1.023-1.026]$,
-$w_{2n^*}=1.027\ [1.026-1.029]$;
+$w_{\anwt}=1.022\ [1.02-1.023]$,
+$w_{\anmt}=1.025\ [1.023-1.026]$,
+$w_{\eumt}=1.027\ [1.026-1.029]$;
 \textbf{$\boldsymbol{\mu=10^{-5}}$:}
 $\delta=3.089\cdot10^{-3}\ [2.412\cdot10^{-3}-3.169\cdot10^{-3}]$,
-$w_{2n+1}=1.022\ [1.021-1.023]$,
-$w_{2n+1^*}=1.024\ [1.023-1.026]$,
-$w_{2n^*}=1.027\ [1.026-1.028]$;
+$w_{\anwt}=1.022\ [1.021-1.023]$,
+$w_{\anmt}=1.024\ [1.023-1.026]$,
+$w_{\eumt}=1.027\ [1.026-1.028]$;
 \textbf{$\boldsymbol{\mu=10^{-6}}$:}
 $\delta=1.413\cdot10^{-3}\ [1.04\cdot10^{-3}-2.529\cdot10^{-3}]$,
-$w_{2n+1}=1.021\ [1.02-1.023]$,
-$w_{2n+1^*}=1.024\ [1.023-1.026]$,
-$w_{2n^*}=1.028\ [1.026-1.029]$;
+$w_{\anwt}=1.021\ [1.02-1.023]$,
+$w_{\anmt}=1.024\ [1.023-1.026]$,
+$w_{\eumt}=1.028\ [1.026-1.029]$;
 \textbf{$\boldsymbol{\mu=10^{-7}}$:}
 $\delta=3.4\cdot10^{-3}\ [2.043\cdot10^{-3}-3.578\cdot10^{-3}]$,
-$w_{2n+1}=1.019\ [1.017-1.02]$,
-$w_{2n+1^*}=1.026\ [1.024-1.027]$,
-$w_{2n^*}=1.027\ [1.026-1.029]$.
+$w_{\anwt}=1.019\ [1.017-1.02]$,
+$w_{\anmt}=1.026\ [1.024-1.027]$,
+$w_{\eumt}=1.027\ [1.026-1.029]$.
 \label{fig:mu}

### 2022-03-24 src/pyabc_visualization_patch/pyabc/___init___.py
fixed extended prior plot

@@ -14,6 +14,7 @@ from .kde import (
     plot_kde_matrix2,
     plot_kde_matrix_for_paper_was,
     plot_kde_matrix_for_paper,
+    plot_kde_matrix_for_paper_extended,
     plot_kde_matrix_highlevel)
 from .sample import (
     plot_sample_numbers,

### 2022-03-24 src/pyabc_visualization_patch/pyabc/kde.py

@@ -822,6 +822,121 @@ def plot_kde_matrix_for_paper(df, w, limits=None, colorbar=True, height=2.5,
 
     return arr_ax
 
+def plot_kde_matrix_for_paper_extended(df, w, limits=None, colorbar=True, height=2.5,
+                    numx=50, numy=50, refval=None, refval_color='C1',
+                    kde=None, names: dict = None, arr_ax=None):
+    """
+    Plot a KDE matrix for 1- and 2-dim marginals of the parameter samples.
+
+    Parameters
+    ----------
+    df: Pandas Dataframe
+        The rows are the observations, the columns the variables.
+    w: np.narray
+        The corresponding weights.
+
+    Other parameters: See plot_kde_matrix_highlevel.
+
+    Returns
+    -------
+    arr_ax:
+        Array of the generated plots' axes.
+    """
+
+    n_par = df.shape[1]
+    par_ids = list(df.columns.values)
+
+    if names is None:
+        names = {key: key for key in par_ids}
+    if arr_ax is None:
+        fig, arr_ax = plt.subplots(nrows=n_par, ncols=n_par,
+                                   sharex=False, sharey=False,
+                                   figsize=(height * n_par, height * n_par))
+    else:
+        fig = arr_ax[0, 0].get_figure()
+
+    if limits is None:
+        limits = {}
+    default = (None, None)
+        
+    def hist_2d(x, y, ax):
+        df = pd.concat((x, y), axis=1)
+        histogram.plot_histogram_2d_lowlevel(df, w,
+                    x.name, y.name,
+                    xmin=limits.get(x.name, default)[0],
+                    xmax=limits.get(x.name, default)[1],
+                    ymin=limits.get(y.name, default)[0],
+                    ymax=limits.get(y.name, default)[1],
+                    # numx=numx, numy=numy,
+                    ax=ax, 
+                    # title=None, 
+                    # colorbar=colorbar,
+                    refval=refval, refval_color=refval_color,
+                    # kde=kde,
+                    xname=names[x.name], yname=names[y.name], bins=20)
+#         print(ax.get_yticks())
+        ax.set_yticks(ax.get_yticks()[1:-1])
+
+    def hist_1d(x, ax,i):
+        df = pd.concat((x,), axis=1)
+        import seaborn as sns
[…]

### 2022-03-24 ms/ms.tex
added citation; revised prior explanation

@@
 Similarly, in unicellular eukaryotes growing in benign conditions, aneuploidy usually leads to slower growth and decreased overall fitness~\citep{Niwa2006, Torres2007, Pavelka2010, Sheltzer2011, Kasuga2016}, in part due to proteotoxic stress caused by increased expression in aneuploid cells~\citep{Pavelka2010, Santaguida2015, Zhu2018} and hypo-osmotic-like stress~\citep{Tsai2019}.
+% TODO cite Yang, F. et al. The fitness costs and benefits of trisomy of each Candida albicans chromosome. Genetics 218, 1–7 (2021).
 
@@
 From these predictions, relative fitness values can be computed.
-Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix.
+Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values to be used as a prior distribution by sampling 10,000 samples from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix~(\Cref{fig:growth-curves}).
 
-We sampled 10,000 samples to use as a prior distribution~(\Cref{fig:growth-curves}).
 We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{\anwt}$~(\Cref{fig:growth-curves}-D, assuming $w_{\euwt}=1$).
-In lieu of a more suitable prior, we used the same prior for $w_{\anmt}$ and $w_{\eumt}$.
+In our standard informative prior distribution, we used the same prior for $w_{\anmt}$ and $w_{\eumt}$.
 To increase computational efficiency, we also assumed $w(\eumt)>w(\anmt)>w(\anwt)>w(\euwt)$; running the inference without this assumption produced similar results. 
-We also examined an extended informative prior that uses additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{\eumt}/w_{\anwt}$~(\Cref{fig:growth-curves}L).
+In our extended informative prior distribution, we used additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{\eumt}/w_{\anwt}$~(\Cref{fig:growth-curves}L).
 The same prior was used for $w_{\eumt}/w_{\anmt}$.
@@
 
-% FIG posterior alt-prior
-%% generated with alt-prior.ipynb
-%% TODO IK - update the figure to look like Figure 2
-\begin{figure}[p]
-  \centering
-\includegraphics[width=0.9\textwidth]{../figures/posterior-alt.pdf}
-  \caption{
-  \textbf{Posterior distribution of parameters inferred with the extended prior distribution.}
-On the diagonal, the inferred posterior distribution of each model parameter. 
-Below the diagonal, the inferred joint posterior distribution of pairs of model parameters (dark purple and bright yellow for low and high density, respectively).
-Red markers and orange lines for the joint MAP estimate (which may differ from the marginal MAP, as the marginal distribution integrates over all other parameters).
-} 
-  \label{fig:posterior-alt}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%
-
 \paragraph{Model checking and comparison.}
@@
 
-We also inferred model parameters under the assumption that the mutation rate increases in aneuploid cells by a factor $\tau$ = 33/32 (due to an additional chromosome), 2, 5, 10, or 100 (due to genetic instability).
+% Rancati G, Pavelka N.
+Karyotypic changes as drivers and catalyzers of cellular evolvability: A perspective from non-pathogenic yeasts.
+Semin Cell Dev Biol.
+2013;24(4):332-338.
+doi:10.1016/j.semcdb.2013.01.009
+% Bouchonville K, Forche A, Tang KES, Semple C a M, Berman J.
+Aneuploid chromosomes are highly unstable during DNA transformation of Candida albicans.
+Eukaryot Cell.
+2009;8(10):1554-1566.
+doi:10.1128/EC.00209-09
+% Zhu J, Pavelka N, Bradford WD, Rancati G, Li R. Karyotypic determinants of chromosome instability in aneuploid budding yeast. PLoS Genet. 2012;8(5). doi:10.1371/journal.pgen.1002719
+
+It has been suggested that aneuploidy increased genetic instability~\citep{Sheltzer2011b}.
+Therefore, we inferred model parameters under the assumption that the mutation rate increases in aneuploid cells by a factor $\tau$ = 33/32 (due to an additional chromosome), 2, 5, 10, or 100 (due to genetic instability).
 We found that the posterior distribution was similar for $\tau=$1, 33/32, 2, and 5 (\Cref{fig:tau}).
@@
    \caption{
-    \textbf{Genotype fixations for models with different $\tau$.} We estimated the parameters for different models, each assuming a different value of $\tau$, the fold-increase in mutation rate in aneuploid cells.
+    \textbf{Genotype fixations for models with increased genetic instability.} We estimated the parameters for different models, each assuming a different value of $\tau$, the fold-increase in mutation rate in aneuploid cells.
 We then generated 10,000 simulations using the MAP estimate of each model and evaluated the fraction of simulations in which the genotype \euwt\ (\textbf{A}), \anwt\ (\textbf{B}), and \eumt\ (\textbf{C}) is fixed (y-axis) at each generation (x-axis).
@@
 \label{fig:growth-curves}
+\end{figure}
+
+% FIG posterior alt-prior
+%% generated with alt-prior.ipynb
+%% TODO IK - update the figure to look like Figure 2
+\begin{figure}[p]
+  \centering
+\includegraphics[width=0.9\textwidth]{../figures/posterior-alt.pdf}
+  \caption{
+  \textbf{Posterior distribution of parameters inferred with the extended prior distribution.}
+On the diagonal, the inferred posterior distribution of each model parameter. 
+Below the diagonal, the inferred joint posterior distribution of pairs of model parameters (dark purple and bright yellow for low and high density, respectively).
+Red markers and orange lines for the joint MAP estimate (which may differ from the marginal MAP, as the marginal distribution integrates over all other parameters).
+} 
+  \label{fig:posterior-alt}
 \end{figure}

### 2022-03-27 ms/ms.tex
bit more TODOS finished
Update ms.tex

--- 2022-03-23 ms/ms.tex
@@
 That is, we assume chromosomes are gained and lost at the same rate, and we neglect events that form a less-fit genotype (i.e., \anwt\; to \euwt\; and \eumt\; to \anmt).
-The fitness values of the four genotypes are given by \Cref{table:fitness}.
+The fitness values of the four genotypes are given by \Cref{table:single-locus}.
 
-%%% Table: fitness
+%%% Table: single-locus model
 \begin{table}[h]
 \centering
-\caption{\textbf{Fitness values.}}
+\caption{\textbf{Single-locus model fitness values.}}
 \begin{tabular}{lllll}
@@
 \end{tabular}
-\label{table:fitness}
+\label{table:single-locus}
 \caption*{
@@
     \end{equation}
-where the fitness values $w_i$ are given in \Cref{table:fitness} and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
+where the fitness values $w_i$ are given in \Cref{table:single-locus} and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
 The effect of mutation and aneuploidy on genotype frequencies is given by
@@
 
-For a model without aneuploidy (that is, when the aneuploidy rate is fixed at zero, $\delta=0$), we disregard the increased expression in chromosome III and the growth advantage measured in generation 450, and focus on the growth advantage measured in later generations, presumably due to a beneficial mutation.
-Therefore, the likelihood is approximated by
-\begin{equation}\begin{aligned}
-\label{eq:heatstress-noaneuploidy-likelihood}
-\likelihood_{!}(\theta) = &\ 
-	1 - 
-	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\}\big) - \\
-&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\}\big) + \\
-&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\}\big)
-\;.
-\end{aligned}\end{equation}
+% YR: this whole section doesn't really make sense.
+We know there was aneuploidy.
+If we assume delta=0 that means that we think this aneuploidy was not relevant to the adaptation process.
+But then why do we take generations 1700 and 2300 to be the time for 2n* fixation?
+These times were not observed for fixation of a mutant but rather for elimination of aneuploidy.
+If we disregard aneuploidy, why do we still consider these times?
+%For a model without aneuploidy (that is, when the aneuploidy rate is fixed at zero, $\delta=0$), 
+%the likelihood is similarly approximated by
+%\begin{equation}\begin{aligned}
+%\label{eq:heatstress-noaneuploidy-likelihood}
+%\likelihood_{!}(\theta) = &\ 
+%	1 - 
+%	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\}\big) - \\
+%&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\}\big) + \\
+%&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\}\big)
+%\;.
+%\end{aligned}\end{equation}
 
@@
 
-\paragraph{Model comparison.} The main criterion that we use for model selection is WAIC, the widely applicable information criterion ~\citep{gelman2013bayesian},
+\paragraph{Model comparison.} We perform model selection using WAIC, the widely applicable information criterion ~\citep{gelman2013bayesian},
 \begin{equation} \label{eq:WAIC}
@@
 WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy~\citep{Kass1995}.
-
-We also plot posterior predictions: for each model we execute $10,000$ simulations using the MAP parameter estimates and plot the distributions of time to fixation of \eumt, one of key properties of the model likelihood.
-These plots visualize the fit of each model to the data.
-Also, for similar models we plot the marginal and joint posterior distributions of the parameters; if these are similar, we consider the models are interchangeable.
-We validate this by comparing HDI (highest density interval) of posterior distributions.
+% TODO IK Describe posterior plots comparison method - Fig 4 & S3
 
@@
 Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix.
-
 We sampled 10,000 samples to use as a prior distribution~(\Cref{fig:growth-curves}).
-We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$~(\Cref{fig:growth-curves}-D, assuming $w_{2n}=1$).
+We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$ ~(\Cref{fig:growth-curves}-D).
 In lieu of a more suitable prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$.
 To increase computational efficiency, we also assumed $w(2n*)>w(2n+1*)>w(2n+1)>w(2n)$; running the inference without this assumption produced similar results.
-We also examined an extended informative prior that uses additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$~(\Cref{fig:growth-curves}L).
+
+Compared to other priors we tested, this prior produced lower WAIC, better posterior prediction plots, and more stable parameter estimates, as follows.
+First, we tried to use additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$ ~(\Cref{fig:growth-curves}L).
 The same prior was used for $w_{2n*}/w_{2n+1*}$.
-
-As a control, we tested an uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$, or (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$.
-In these cases the inference algorithm failed to converge.
+This prior resulted in WAIC 71.91, compared to 67.87 with the above informative prior.
+% generated with fixed-mutation-rate-and-WAICs.ipynb TODO IK update the WAIC with 1000 reps run
+We also tested an uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$, or (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$.
+In both cases the inference algorithm failed to converge.
 
@@
 $w_{2n^*}=1.028\ [1.026-1.029]$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$. 
-Thus, we can infer that the benefit and cost of trisomy was $b=2.5\%$ and $c=0.3\%$, and the benefit of the beneficial mutation was $2.8\%$~(\Cref{table:fitness}).
+Thus, we can infer that the benefit and cost of trisomy was $b=2.5\%$ and $c=0.3\%$, and the benefit of the beneficial mutation was $2.8\%$~(\Cref{table:single-locus}).
 
@@
 
-% FIG posterior alt-prior
-%% generated with alt-prior.ipynb
-%% TODO IK - update the figure to look like Figure 2
-\begin{figure}[p]
-  \centering
-\includegraphics[width=0.9\textwidth]{../figures/posterior-alt.pdf}
-  \caption{
-  \textbf{Posterior distribution of parameters inferred with the extended prior distribution.}
-On the diagonal, the inferred posterior distribution of each model parameter. 
-Below the diagonal, the inferred joint posterior distribution of pairs of model parameters (dark purple and bright yellow for low and high density, respectively).
-Red markers and orange lines for the joint MAP estimate (which may differ from the marginal MAP, as the marginal distribution integrates over all other parameters).
-} 
-  \label{fig:posterior-alt}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%
-
 \paragraph{Model checking and comparison.}
 The model fits the data well: in simulations using the MAP parameter estimates, \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\Cref{fig:2n*-fixation}B).
+% TODO IK what happened to the sentence about the fixation on anwt?
+green lines ub fig4?
 Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\Cref{fig:ppc-plot}).
 
-Moreover, a model without aneuploidy (where the aneuploidy rate is fixed at zero, $\delta=0$), fails to explain the experimental observations~(Figure~\ref{fig:2n*-fixation}).
+% commented out, see remark above when describing likelihood of delta=0.
+%Moreover, a model without aneuploidy (where the aneuploidy rate is fixed at zero, $\delta=0$), fails to explain the experimental observations~(Figure~\ref{fig:2n*-fixation}).
 %% generated with MAP-and-hdi.ipynb
-The estimated mutation rate without aneuploidy is $\mu=7.979\cdot10^{-9}\ [7.903\cdot 10^{-9}-8.135\cdot10^{-9}]$, much lower compared to a model with aneuploidy and suggesting a target size of just 40.
-The fitness of the mutant is also much lower at $w_{2n^*}=1.013\ [1.012-1.013]$.
-This is because, without aneuploidy, a high mutation rate or fitness effect will lead to faster appearance and fixation of \eumt\ than in the experimental observations.
+%The estimated mutation rate without aneuploidy is $\mu=7.979\cdot10^{-9}\ [7.903\cdot 10^{-9}-8.135\cdot10^{-9}]$, much lower compared to a model with aneuploidy and suggesting a target size of just 40.
+%The fitness of the mutant is also much lower at $w_{2n^*}=1.013\ [1.012-1.013]$.
+%This is because, without aneuploidy, a high mutation rate or fitness effect will lead to faster appearance and fixation of \eumt\ than in the experimental observations.
 Even with these lower estimates, the model fit is worse than that of a model with aneuploidy~(\Cref{fig:2n*-fixation}).
@@
 Furthermore, we fitted models with a mutation rate fixed at $\mu$=$10^{-5}$, $10^{-6}$ and $10^{-7}$.
-We inferred similar parameters estimates for the model with $\mu=10^{-6}$ compared to the model with a free $\mu$ parameter.
-Models with $\mu$=$10^{-5}$ and $\mu$=$10^{-7}$ inferred different parameters estimates~(\Cref{fig:mu}).
+We inferred similar parameters estimates compared to the model with a free $\mu$ parameter.
+% TODO IK where is this shown?
+how do you quantify it?
 WAIC was lower when $\mu$ is fixed~(\Cref{table:WAIC}), but this is not surprising, as WAIC attempt to balance between model fit and model complexity, where the latter takes into account the number of model parameters.
 
-% generated with alt-prior.ipynb
-Finally, we estimated the parameters under an extended informative prior (see \emph{Prior distributions}), with different priors for $w_{2n+1^*}$ and $w_{2n*}$.
-The estimated mutation rate was much lower compared to the basic prior, with $\mu=2.379\cdot10^{-9}\ [2.267\cdot10^{-9}-2.458\cdot10^{-9}]$.
-Other parameter estimates are: $\delta=2.636\cdot10^{-3}\ [2.245\cdot10^{-3}-2.919\cdot10^{-3}]$,
-$w_{2n+1}=1.022\ [1.021-1.023]$,
-$w_{2n+1^*}=1.032\ [1.03-1.035]$,
-$w_{2n^*}=1.033\ [1.031-1.036]$. 
-The WAIC of the model is much lower: -57 compared to 295~(\Cref{table:WAIC}).
-
-% Fig model with and without aneuploidy
+% Fig single-locus model with and without aneuploidy
 %% generated with with-aneuploidy_vs_no_aneuploidy.ipynb
@@
   \caption{
-    \textbf{Model fit with and without aneuploidy.}
-    \textbf{(A)} The distribution of time to fixation of \eumt\ (i.e., adaptation time) in $10,000$ simulations of the model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange), and a model fitted to the data assuming $\delta=0$ (green).
+    \textbf{Single-locus model fit with and without aneuploidy.}
+    \textbf{(A)} The distribution of time to fixation of \eumt\ (i.e., adaptation time) in $10,000$ simulations of the single-locus model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange), and a model fitted to the data assuming $\delta=0$ (green).
 In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and another by generation 2,350 (dashed lines) but not before generation 450.
@@
      }
-     % TODO should we remove this figure altoghther? change it? because we removed no-aneuploidy model...? should we merge this with the next figure?
   \label{fig:2n*-fixation}
@@
   \caption{
-  \textbf{Posterior genotype frequency dynamics for the model.} 
+  \textbf{Posterior genotype frequency dynamics for the single-locus model.} 
   The posterior prediction for the frequencies of the four genotypes over time.
-Each of the 20 curves is the average of 10,000 simulations of the model using parameters drawn from the posterior distribution.
+Each of the 20 curves is the average of 10,000 simulations of the single-locus model using parameters drawn from the posterior distribution.
   }
@@
   \label{fig:tau-plots}
+  % TODO IK axis labels like in figure 3, also somehow the graphics looks different from previous figures
+  % fixed means 100% ?
 \end{figure}
@@
 \subsection*{Analysis}
+
+\paragraph{Approximation.}
+We aim to approximate the adaptation time (i.e., time to fixation of \eumt) and analyse how it is affected by the model parameters.
+There are two stochastic events that can occur on the \euwt\ genotype background. The first is a mutation that produces a \eumt\ genotype.
+The second is a mis-segregation event that produces a \anwt\ genotype.
+The rate at which these events occur is significantly different, and we assume that if one event occurs (e.g., mutation), then the other event (e.g., aneuploidy) is unlikely to occur.
+The expected waiting time until one of these two genotype appears and then leads to adaptation is therefore
+\begin{equation}\label{eq:T_total}
+T = p_a T_a+(1-p_a) T_m \;,
+\end{equation}
+where $p_a$ is the probability that aneuploidy is the first event to occur, 
+$T_m$ is the expected time for appearance and fixation of \eumt\ given it appears before \anwt, and $T_a$ is the expected time for appearance of \anwt, followed by a mutation to \anmt, and loss of aneuploidy to \eumt, given that \anwt\ appears before \eumt.
+
+We assume the waiting time $T_{i,j}$ for appearance of genotype $j$ in a population of genotype $i$ is geometrically distributed with rate $p_{i,j}$, such that
+\begin{subequations}
+\begin{align}
+p_{2n, 2n^*} &= 1-\Big(1 - 2 \frac{w_{2n^*}-w_{2n}}{w_{2n}} \mu \Big)^N \;, \\
+p_{2n, 2n+1} &= 1-\Big(1 - 2 \frac{w_{2n+1}-w_{2n}}{w_{2n}} \delta \Big)^N \; \\
+p_{2n+1, 2n+1^*} &= 1-\Big(1 - 2 \frac{w_{2n+1^*}-w_{2n+1}}{w_{2n+1}} \mu \Big)^N \; \\
+p_{2n+1^*, 2n^*} &=1-\Big(1 - 2 \frac{w_{2n^*}-w_{2n+1^*}}{w_{2n+1^*}} \delta \Big)^N \;.
+\end{align}
+\end{subequations}
+However, if the mis-segragation supply is high enough, $\delta N > 1$, then \eumt\ may appear before \anmt\ is fixed (\Cref{fig:ppc-plot}), and the latter equation is
+\begin{equation}
+p_{2n+1^*, 2n^*} = 1-\Big(1-2 \frac{w_{2n^*}-w_{2n+1}}{w_{2n+1}} \delta \Big)^{zN} \;,
+\end{equation}
+where $z \ll 1$ is the frequency of \anmt\ when \eumt\ first appears (note that here we use $w_{2n+1}$ rather than $w_{2n+1^*}$).
+
+We assume that if \anwt\ appears in a population of \euwt\ and does not go to extinction then \eumt\ does not appear, and vice versa.
+Thus, we need to find the minimum of the waiting times for appearance of these genotypes.
+The minimum of two geometric random variables with rates $p_{2n,2n^*}$ and $p_{2n,2n+1}$ is also a geometric random variable with rate $p_{min} = 1-(1-p_{2n,2n^*})(1-p_{2n,2n+1})$, which is the inverse of the expected waiting time until the first event occurs, $T_{min}=1/p_{min}$.
+Given that one of these genotypes appeared, the probability that it was \anwt\ rather than \eumt\ is
+$p_A = \frac{p_{2n,2n+1}}{p_{2n,2n^*}+p_{2n,2n+1}}$.
+
+Given a mutation occurs first, the expected adaptation time via the mutation trajectory is 
+\begin{equation}\label{eq:T_a} 
+T_{M} = T_{min} + \tau_{2n, 2n^*} \;,
+\end{equation}
+where $\tau_{i, j}$ is the waiting time for fixation of genotype $j$ in a population of genotype $i$, given it starts with a single copy and does not go to extinction by drift.
+Given that aneuploidy occurs first,
+the expected adaptation time via the aneuploidy trajectory is approximated by
+\begin{equation}\label{eq:T_a} 
+T_a = T_{min} + T_{2n+1, 2n+1^*} + T_{2n+1^*,2n^*} + \tau_{2n, 2n+1} + \tau_{2n+1, 2n+1^*} + \tau_{2n+1^*, 2n^*} \;.
+\end{equation}
+
+We assume that the waiting time until a new genotype occurs, $T{i,j}$, is much longer than the waiting time for its fixation, $\tau_{i,j}$.
+Nevertheless, these fixation times must be included in a rigorous analysis.
+There are two ways to compute the fixation times.
+A deterministic-process approach is given by the classical bi-allelic single-locus equation.
+It approximates the expected time for genotype $j$ to reach frequency $x$ when starting from frequency $y$ in a population of genotype $i$ by
+ % TODO MP ref for equation? Crow and Kimura? YR: yes
+\begin{equation} \label{eq:tau_fix_x}
+\tau_{i, j}(x,y) \approx \frac{\log{\Big(\frac{x(1-y)}{(1-x)y} \Big)}}{\log{\Big(\frac{w_j}{w_i}\Big)}} \;,
+\end{equation} 
+and we use $\tau_{i,j} = \tau_{i,j}(1-1/N, 1/N) \approx 2\log{(N)}/\log{(w_j/w_i)}$, except for $\tau_{i,2n^*} = \tau_{i,2n^*}(0.95, 1/N)$ because we stop our simulations when genotype \eumt\ reaches frequency of 0.95.
+A stochastic-process approach for approximating $T_{i, j}$ is given by the diffusion-equation approximation \citep[eq.~17]{Kimura1969a}. 
+% TODO MP give details here
+
+\paragraph{Validation.}
+To validate our approach we performed simulations of the full model (\Cref{eq:mutation-aneuploidy-single,eq:drift-single,eq:selection-single}) without aneuploidy, i.e., $\delta=0$, and with the other parameters as inferred above.
+The approximation for the expected adaptation time (\Cref{eq:T_total}) provides a good fit to simulation results except when the effective population size is above $5\cdot 10^6$ cells (\Cref{fig:only mut}).
+In that case, the expected number of new mutants per generation is roughly 0.5, and  we it is likely that the fixation event will start with more than a single \eumt\ cell appearing in the same generation, which violates our assumption that a fixation event starts from a single mutant cell (i.e., $y=1/N$ in \Cref{eq:tau_fix_x}).
+
+ \begin{figure}[h!]
+ \centering
+ \includegraphics[width=\textwidth]{../figures/figgrid_noaneu.pdf}
+ \caption{\textbf{Expected adaptation time without aneuploidy.}
+  The expected time to appearance and fixation of \eumt\ as a function of the population size, $N$ (\textbf{A}) or as a function of the mutation rate, $\mu$ (\textbf{B}).
+Mean and 95\% CI from simulations in blue.
+Analytic approximation (\Cref{eq:T_total}) in orange.
+Here, $w_\anwt=1.021,\,   w_\anmt=1.025$ and $w_\eumt=1.028$.
+In panel A, $\mu=9.6\cdot 10^{-8}$.
+In panel B, $N=6\cdot 10^6$.
+  }
+  %TODO MP 
+  % which approximation do you use here for T_i->j ?
+  % The error bars a quite large, can you make them smaller by running more simulations?
+  % --> or by making the y axis in log scale!!
+  % more ticks for x axis
+ \label{fig:only mut}
+ \end{figure}
+
+ \begin{figure}[h!]
+ %\centering
+ \includegraphics[width=1\textwidth]{../figures/figgrid_nodirectmut_ylog_uncorr.jpg}
+ \caption{\textbf{Expected adaptation time without mutations from $\euwt$ to $\eumt$.}
+  The expected time to appearance and fixation of \eumt\ as a function of the population size, $N$ (\textbf{A}) and as a function of the mutation rate, $\mu$ (\textbf{B}) and as a function of the aneuploidy rate, $\delta$ ({C}).
+Blue error bars for mean and 95\% CI from simulations.
+Green line for the analytic approximation (\Cref{eq:T_total}).
+% TODO why green?
+in previous figure it was orange.
+  Red curve for the approximation with $T_{\euwt, \anwt}=T_{\anwt, \anmt}=0$. 
+  Orange curve for the approximation with $x=0.017$ in \Cref{eq:tau_fix_x} for the fixation time of $\eumt$; we found this gives a good fit in panel B.
+  Here, $w_\anwt=1.021,\, w_\anmt=1.025$ and $w_\eumt=1.028$, $N=6\cdot 10^6$,\, $\mu=9.6\cdot 10^{-8}$ and $\delta=5.4\cdot 10^{-4}$ unless parameter is varied on the x-axis.
+  }
+ \label{fig:no direct mut}
+ \end{figure}
+
+Next, we performed simulations without mutations from $\euwt$ to $\eumt$ (\Cref{fig:no direct mut}). 
+% TODO MP: I comment this out - not clear what "Before" refers to, and not sure we need this sentence. what do you think?
+%Before, we had the optimal scenario of waiting until the first successful mutation appears and then waiting until is reaches fixation. Our approximations did quite well there. 
+Here, aneuploidy is transient and fixation of \anwt\ or \anmt\ could be outpaced by the arrival of the next genotype.
+This would explain the discrepancy between the approximations and simulation results in \Cref{fig:no direct mut}.
+To asses this further, \Cref{fig:fixT vs waitT} provides a clear picture.
+At least for intermediate to high values of the respective range of the parameters, the waiting time for appearance, $T_{i,j}$ is far shorter than the time to fixation, $\tau_{i,j}$.
+Indeed, if we ignore the fixation time of $\anwt$ and $\anmt$ ($T_{2n,2n+1}$ and $T_{2n+1,2n+1^*}$) in \Cref{eq:T_a}, we get a better fit to the data (compare red curves and blue markers in \Cref{fig:no direct mut}).
+Furthermore, setting $x=0.017$ for $T_{2n+1^*, 2n^*}$ in \Cref{eq:tau_fix_x}, we get an even better fit in some cases (\Cref{fig:no direct mut}B), but not others (\Cref{fig:no direct mut}A,C).
+% TODO maybe remove these x=0.017 sentences and the next sentence
+However, we stop simulations when genotype $\eumt$ reaches frequency of $0.95>0.017$.
+
+ \begin{figure}[h!]
+ %\centering
+ \includegraphics[width=1\textwidth]{../figures/figgrid_fixT_vs_waitT.jpg}
+ \caption{\textbf{Expected time of fixation and waiting time events on the aneuploid trajectory.}
+  The expected time to appearance and fixation of each step seperately in the aneuploid trajectory as a function of the population size, $N$ (panel A) and as a function of the mutation rate, $\mu$ (panel B) and as a function of the aneuploidy rate, $\delta$ (panel C).
+Solid curves give the fixation time and dashed curves the waiting time.
+Curves of the same color indicate the same base population state;  e.g.
+the blue dashed curve is the waiting time in state $\anwt$ until a succesful individual of type $\anmt$ appears.
+The corresponding solid curve indicates the time to fixation of the population in state $\anwt$.
+Here, $w_\anwt=1.021,\,   w_\anmt=1.025$ and $w_\eumt=1.028$.
+Parameter values in panels where it is not varied, $N=6\cdot 10^6$,\, $\mu=9.6\cdot 10^{-8}$ and $\delta=5.4\cdot 10^{-4}$.
+}
+ \label{fig:fixT vs waitT}
+ \end{figure}
+
+In summary, whereas the two possible trajectories to $\eumt$ do not compete against each other in both \Cref{fig:only mut} and \Cref{fig:no direct mut}, the approximations are doing well only in restricted parameter ranges.
+These tend to be, where the respective parameters are rather small.
+If they grow bigger, they are violating some assumptions of our approximations.
 
@@
     \textbf{Fitness estimation from growth curves.}
-    \textbf{(A-D)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{39}{\celsius}. $\hat w_{2n+1}/w_{2n}$=1.024 (95\% CI: 0.959 - 1.115). \texttt{Curveball} 
+    % TODO IK caption should explain what each panel shows.
+    \textbf{(A-D)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{39}{\celsius}. $\hat w_{2n+1}/w_{2n}$=1.024 (95\% CI: 0.959 - 1.115).
     \textbf{(E-H)} Fitness estimation from growth curves of \anwt\ and \eumt\ at \SI{39}{\celsius}. $\hat w_{2n*}/w_{2n+1}$=1.033 (95\% CI: 1.027 - 1.041).
@@
 See \emph{Models and Methods, Prior distributions} for more details.
-\textbf{(A,B;E,F)} Mono-culture growth curve data (markers) and best-fit growth models (lines).
-\textbf{(C,G)} The mixed-culture prediction for the strains from A,B and E,F respectively, 6,375 generated curves.
-\textbf{(D,H)} The relative fitness distribution for \anwt\ relative to \euwt\ (panel D) and \eumt\ relative to \anwt\ (panel H).
-Figures generated by \texttt{Curveball}.
-% TODO IK please fix the range of x axis in D; It definitely doesnt need to go beyond 1.25...
 } 
@@
 %% generated with diff-tau.ipynb
-% TODO IK 10_000 reps is mocked, should be rerunned and fixed
 \begin{figure}[h!]
@@
   \caption{
-    \textbf{Model with elevated mutation rate in aneuploid cells.}  \textbf{(A-E)} The inferred posterior distributions for models with different values of $\tau$, the fold-increase in mutation rate in aneuploid cells (\anwt\ and \anmt).
+    \textbf{Single-locus model inference with elevated mutation rate in aneuploid cells.}  \textbf{(A-E)} The inferred posterior distributions of model parameters for the single-locus model with different values of $\tau$, the fold-increase in mutation rate in aneuploid cells (\anwt\ and \anmt).
 When the increase in mutation rate is high, $\tau=10$ and $\tau=100$, the inferred mutation (A) and aneuploidy (B) rate tends to be lower.
@@
   % TODO IK maybe add dashed vertical lines with the same colors to show where the MAP of each distribution is?
-  \end{figure}
 
-% Fig mu comparisons
-%% generated with diff-mutation-rate.ipynb
-\begin{figure}[h!]
-  \centering
-  \begin{subfigure}{0.45\textwidth}
-      \includegraphics[width=\textwidth]{../figures/mu-A.pdf}      
-  \end{subfigure}
-  \begin{subfigure}{0.45\textwidth}
-      \includegraphics[width=\textwidth]{../figures/mu-B.pdf}      
-  \end{subfigure}
-  \\
-   \begin{subfigure}{0.45\textwidth}
-      \includegraphics[width=\textwidth]{../figures/mu-C.pdf}      
-  \end{subfigure}
-    \begin{subfigure}{0.45\textwidth}
-      \includegraphics[width=\textwidth]{../figures/mu-D.pdf}      
-    \end{subfigure}
-  \caption{
-    \textbf{Model with fixed mutation rate.}  \textbf{(A-D)} The inferred posterior distributions for models with free and fixed mutation rate, $\mu$.
-The MAP (maximum a posteriori) and 50\% HDI (highest density interval) for each model are following:
-\textbf{free $\boldsymbol{\mu}$:}
-$\delta=2.749\cdot10^{-3}\ [1.476\cdot10^{-3}-2.822\cdot10^{-3}]$,
-$w_{2n+1}=1.022\ [1.021-1.023]$,
-$w_{2n+1^*}=1.025\ [1.023-1.026]$,
-$w_{2n^*}=1.027\ [1.026-1.029]$;
-\textbf{free $\boldsymbol{\mu}$, run 2:}
-$\delta=1.938\cdot10^{-3}\ [1.338\cdot10^{-3}-2.748\cdot10^{-3}]$,
-$w_{2n+1}=1.022\ [1.02-1.023]$,
-$w_{2n+1^*}=1.025\ [1.023-1.026]$,
-$w_{2n^*}=1.027\ [1.026-1.029]$;
-\textbf{$\boldsymbol{\mu=10^{-5}}$:}
-$\delta=3.089\cdot10^{-3}\ [2.412\cdot10^{-3}-3.169\cdot10^{-3}]$,
-$w_{2n+1}=1.022\ [1.021-1.023]$,
-$w_{2n+1^*}=1.024\ [1.023-1.026]$,
-$w_{2n^*}=1.027\ [1.026-1.028]$;
-\textbf{$\boldsymbol{\mu=10^{-6}}$:}
-$\delta=1.413\cdot10^{-3}\ [1.04\cdot10^{-3}-2.529\cdot10^{-3}]$,
-$w_{2n+1}=1.021\ [1.02-1.023]$,
-$w_{2n+1^*}=1.024\ [1.023-1.026]$,
-$w_{2n^*}=1.028\ [1.026-1.029]$;
-\textbf{$\boldsymbol{\mu=10^{-7}}$:}
-$\delta=3.4\cdot10^{-3}\ [2.043\cdot10^{-3}-3.578\cdot10^{-3}]$,
-$w_{2n+1}=1.019\ [1.017-1.02]$,
-$w_{2n+1^*}=1.026\ [1.024-1.027]$,
-$w_{2n^*}=1.027\ [1.026-1.029]$.
-\label{fig:mu}
-  }
   \end{figure}
@@
 %TODO YR read this 
-%% generated with WAIC.ipynb 
+%% generated with fixed-mutation-rate.ipynb 
+% TODO IK run for all updated files
 \begin{table}[h]
@@
 \caption*{
-	WAIC (widely applicable information criterion; \autoref{eq:WAIC})~\citep{gelman2013bayesian} values for models: (1) without aneuploidy;  (2-4) with aneuploidy, fixed mutation rate, and $\tau=1$; (5-10) with aneuploidy, free mutation rate model parameter, and different $\tau$.
-WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy and a difference of 2 is a popular threshold for model comparison~\citep{Kass1995}.
+	TODO caption
+	%TODO IK caption; what is tau in models 1-5?
 } 

### 2022-03-30 notebooks_mp/Time_approximations.py
updated scripts and plots

@@ -108,18 +108,30 @@ def probaneufirst_K(N, mut_rate,aneu_rate,w1,w3):
 def T_haldane_ffix(N, s):
     return 2 * np.log(N) / np.log(1+s)
 
-#deterministic approximation to fixation assumed at frequency 95%
-def T_haldane(N, s, fixation_cutoff=0.95):
+#general deterministic approximation to fixation assumed at frequency 95%
+def T_haldane(N, s, fixation_cutoff=0.95, n0=1):
     if fixation_cutoff < 1:
-        T_ret=np.log(fixation_cutoff*N/(1-fixation_cutoff)) / np.log(1+s)
+        T_ret=np.log(fixation_cutoff*(N-n0)/((1-fixation_cutoff)*n0)) / np.log(1+s)
     else:
         T_ret=2 * np.log(N) / np.log(1+s)
     return  np.maximum(T_ret,0)
 
+#numerical simulation of the time to fixation with recurrent mutations
+def time_evolution(p0, s, rate, tmax, cf=.95):
+    p = [p0]
+
+    for t in range(1, tmax):
+        rec_p = p[t-1] * (1 + s) / (1 + p[t-1] * s)+(1-p[t-1])*2*s*rate
+        p.append( rec_p)
+        if rec_p>=cf:
+            break
+
+    return t
+
 #Fixation time after Kimura
 def integral(f, N, s, a, b):
     f = partial(f, N, s)    
-    return quad(f, a, b)[0]
+    return quad(f, a, b, limit=100)[0]
 def I1(N, s, x):
     if x == 1:
         return 0
@@ -150,11 +162,11 @@ def T_kimura(N, s, fixation_cutoff=.95, n0=1):
 
 ##Two main ones:
 
-def T_mut(N, mut_rate, aneu_rate, w1, w3, fixation=False, fc0=.95):
+def T_mut(N, mut_rate, aneu_rate, w1, w3, fixation=False, fc0=.95, n00=1):
     wait_T=min_waitT(N, mut_rate, aneu_rate, w1, w3)
     if fixation is False:
         return wait_T
[…]

### 2022-03-30 ms/ms.tex
added sentence about WAIC validation
updated ppc plots for extended informative prior model

--- 2022-03-24 ms/ms.tex
@@
 where $\theta$ is a parameter vector, and $\mathbb{E}[\cdot]$ and $\mathbb{V}[\cdot]$ are the expectation and variance taken over the posterior distribution, which in practice are approximated using 50,000 samples from the posterior KDE.
+It was validated that WAIC doesn't change significantly, and WAIC differences between models are presaved when we redraw 50,000 samples.
 WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy~\citep{Kass1995}.
@@
 $w_{\eumt}=1.033\ [1.031-1.036]$. 
-The WAIC of the model is much lower: -57 compared to 295~(\Cref{table:WAIC}).
+The WAIC of the model is much lower: -57 compared to 295~(\Cref{table:WAIC}). However the model fails to explain the experimental observations~(Figure~\ref{fig:2n*-fixation}).
 
@@
     \textbf{Model fit with and without aneuploidy.}
-    \textbf{(A)} The distribution of time to fixation of \eumt\ (i.e., adaptation time) in $10,000$ simulations of the model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange), and a model fitted to the data assuming $\delta=0$ (green).
+    \textbf{(A)} The distribution of time to fixation of \eumt\ (i.e., adaptation time) in $10,000$ simulations of the model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange), and a model fitted to the data assuming $\delta=0$ (green); and to the model with aneuploidy with extended prior distribution (pink).
 In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and another by generation 2,350 (dashed lines) but not before generation 450.
-Thus, the blue distribution is a better fit compared to the green, and the yellow histogram has a very poor fit.
-The last bin contains all the simulations with time equal or greater than 3,000.
+Thus, the blue distribution is a better fit compared to the green and pink, and the yellow histogram has a very poor fit.
+The last bin contains all the simulations with time equal or greater than 3,000 - from pink and green distributions.
     \textbf{(B)}
 	Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy in blue and green, respectively, and corresponding to the blue and green bars in panel A.
-The MAP likelihood (\Cref{eq:heatstress-likelihood}) is 0.84 and 0.67 with and without aneuploidy, respectively.
+The MAP likelihood (\Cref{eq:heatstress-likelihood}) is 0.84, 0.67 with and without aneuploidy, respectively, and 0.14 for the model with aneuploidy and extended prior distribution.
      }
-     % TODO should we remove this figure altoghther? change it? because we removed no-aneuploidy model...? should we merge this with the next figure?
   \label{fig:2n*-fixation}
@@
 Figures generated by \texttt{Curveball}.
-% TODO IK please fix the range of x axis in D; It definitely doesnt need to go beyond 1.25...
 } 
@@
 %% generated with alt-prior.ipynb
-%% TODO IK - update the figure to look like Figure 2
 \begin{figure}[p]

### 2022-03-31 ms/ms.tex
Update ms.tex

--- 2022-03-27 ms/ms.tex
@@
 The proposal particle is then perturbed using a kernel perturbation kernel, $K_t(\theta^* \mid \theta)$ where $\theta$ is the sample from the previous iteration.
-Then, a set of synthetic observations $\tilde{\vec X}^*$ is simulated, and the proposal particle $\theta^*$ is accepted if its approximate likelihood (\Cref{eq:heatstress-likelihood}) is high enough, $\likelihood(\theta^*)>1-\epsilon_t$ (or more commonly, if $1-\likelihood(\theta^*) < \epsilon_t$), where $\epsilon_t>0$ is the \emph{acceptance threshold}, as higher values of $\epsilon_t$ allow more particles to be accepted.
+Then, a set of synthetic observations $\tilde{\vec X}^*$ is simulated, and the proposal particle $\theta^*$ is accepted if its approximate likelihood (\cref{eq:heatstress-likelihood}) is high enough, $\likelihood(\theta^*)>1-\epsilon_t$ (or more commonly, if $1-\likelihood(\theta^*) < \epsilon_t$), where $\epsilon_t>0$ is the \emph{acceptance threshold}, as higher values of $\epsilon_t$ allow more particles to be accepted.
 The acceptance threshold $\epsilon_t$ is chosen as the median of the $1-\likelihood(\theta)$ of the particles accepted in the previous iteration, $t-1$, and $\epsilon_0=0.01$. 
@@
 
-Acceptance is determined according to the approximate likelihood~(\Cref{eq:heatstress-likelihood}), which has a maximum value of 0.875.
+Acceptance is determined according to the approximate likelihood~(\cref{eq:heatstress-likelihood}), which has a maximum value of 0.875.
 Thus, we terminated the inference when $\epsilon \le 0.13$ after six iterations, with $n_6=982$ accepted parameter vectors and effective sample size ESS=651 (\Cref{fig:convergence}).
@@
 	Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy in blue and green, respectively, and corresponding to the blue and green bars in panel A.
-The MAP likelihood (\Cref{eq:heatstress-likelihood}) is 0.84 and 0.67 with and without aneuploidy, respectively.
+The MAP likelihood (\cref{eq:heatstress-likelihood}) is 0.84 and 0.67 with and without aneuploidy, respectively.
      }
@@
 We aim to approximate the adaptation time (i.e., time to fixation of \eumt) and analyse how it is affected by the model parameters.
-There are two stochastic events that can occur on the \euwt\ genotype background. The first is a mutation that produces a \eumt\ genotype.
+There are two genotypes events that can occur on the \euwt\ genotype background. The first is a mutation that produces a \eumt\ genotype.
 The second is a mis-segregation event that produces a \anwt\ genotype.
-The rate at which these events occur is significantly different, and we assume that if one event occurs (e.g., mutation), then the other event (e.g., aneuploidy) is unlikely to occur.
-The expected waiting time until one of these two genotype appears and then leads to adaptation is therefore
+The rate at which these genotypes appear is significantly different. 
+We assume that if one genotype appears (e.g., \eumt), then it will genotype rapidly fix in the population and therefore the other genotype (e.g., \anwt) is unlikely to occur. 
+The expected waiting time until one of the two genotype appears and then leads to adaptation is therefore
 \begin{equation}\label{eq:T_total}
@@
 \end{equation}
-where $p_a$ is the probability that aneuploidy is the first event to occur, 
+where $p_a$ is the probability that \anwt\ is the first to appear without going to extinction, 
 $T_m$ is the expected time for appearance and fixation of \eumt\ given it appears before \anwt, and $T_a$ is the expected time for appearance of \anwt, followed by a mutation to \anmt, and loss of aneuploidy to \eumt, given that \anwt\ appears before \eumt.
+Note that we assume the probability that \eumt\ is the first to appear without going to extinction is $1-p_a$.
 
@@
 
-We assume that the waiting time until a new genotype occurs, $T{i,j}$, is much longer than the waiting time for its fixation, $\tau_{i,j}$.
+We assume that the waiting time until a new genotype occurs, $T_{i,j}$, is much longer than the waiting time for its fixation, $\tau_{i,j}$.
 Nevertheless, these fixation times must be included in a rigorous analysis.
@@
 \paragraph{Validation.}
-To validate our approach we performed simulations of the full model (\Cref{eq:mutation-aneuploidy-single,eq:drift-single,eq:selection-single}) without aneuploidy, i.e., $\delta=0$, and with the other parameters as inferred above.
-The approximation for the expected adaptation time (\Cref{eq:T_total}) provides a good fit to simulation results except when the effective population size is above $5\cdot 10^6$ cells (\Cref{fig:only mut}).
-In that case, the expected number of new mutants per generation is roughly 0.5, and  we it is likely that the fixation event will start with more than a single \eumt\ cell appearing in the same generation, which violates our assumption that a fixation event starts from a single mutant cell (i.e., $y=1/N$ in \Cref{eq:tau_fix_x}).
+To validate our approach we performed simulations of the full model (\cref{eq:mutation-aneuploidy-single,eq:drift-single,eq:selection-single}) without aneuploidy, i.e., $\delta=0$, and with the other parameters as inferred above.
+The approximation for the expected adaptation time $T$ (\cref{eq:T_total}) has a good fit to simulation results except when the effective population size is above $5\cdot 10^6$ cells (\Cref{fig:only mut}A).
+In that case, the expected number of new mutants per generation is roughly 0.5, and   the fixation event is likely to start with more than a single \eumt\ cell appearing in the same generation, which violates our assumption that a fixation event starts from a single mutant cell (i.e., $y=1/N$ in \cref{eq:tau_fix_x}).
+% TODO can we verify that setting y to higher value corrects the approximation?
 
@@
 
+Next, we performed simulations without mutations from \euwt\ to \eumt\ but with mutations from \anwt\ to \eumt\ (\Cref{fig:no direct mut}). 
+% TODO MP: I comment this out - not clear what "Before" refers to, and not sure we need this sentence. what do you think?
+%Before, we had the optimal scenario of waiting until the first successful mutation appears and then waiting until is reaches fixation. Our approximations did quite well there. 
+Here, aneuploidy is transient and fixation of \anwt\ or \anmt\ could be outpaced by the arrival of the next genotype (\anmt\ and \eumt, respectively).
+This would explain the discrepancy between the approximations and simulation results in \Cref{fig:no direct mut}.
+To asses this further, \Cref{fig:fixT vs waitT} provides a clear picture.
+At least for intermediate to high values of the respective range of the parameters, the waiting time for appearance, $T_{i,j}$ is far shorter than the time to fixation, $\tau_{i,j}$.
+Indeed, if we ignore the fixation time of $\anwt$ and $\anmt$ ($T_{2n,2n+1}$ and $T_{2n+1,2n+1^*}$) in \cref{eq:T_a}, we get a better fit to the data (compare red curves and blue markers in \Cref{fig:no direct mut}).
+Furthermore, setting $x=0.017$ for $T_{2n+1^*, 2n^*}$ in \cref{eq:tau_fix_x}, we get an even better fit in some cases (\Cref{fig:no direct mut}B), but not others (\Cref{fig:no direct mut}A,C).
+% TODO maybe remove these x=0.017 sentences and the next sentence
+However, we stop simulations when genotype $\eumt$ reaches frequency of $0.95>0.017$.
+
 \begin{figure}[h!]
@@
 Blue error bars for mean and 95\% CI from simulations.
-Green line for the analytic approximation (\Cref{eq:T_total}).
+Green line for the analytic approximation (\cref{eq:T_total}).
 % TODO why green?
@@
   Red curve for the approximation with $T_{\euwt, \anwt}=T_{\anwt, \anmt}=0$. 
-  Orange curve for the approximation with $x=0.017$ in \Cref{eq:tau_fix_x} for the fixation time of $\eumt$; we found this gives a good fit in panel B.
+  Orange curve for the approximation with $x=0.017$ in \cref{eq:tau_fix_x} for the fixation time of $\eumt$; we found this gives a good fit in panel B.
   Here, $w_\anwt=1.021,\, w_\anmt=1.025$ and $w_\eumt=1.028$, $N=6\cdot 10^6$,\, $\mu=9.6\cdot 10^{-8}$ and $\delta=5.4\cdot 10^{-4}$ unless parameter is varied on the x-axis.
@@
  \end{figure}
-
-Next, we performed simulations without mutations from $\euwt$ to $\eumt$ (\Cref{fig:no direct mut}). 
-% TODO MP: I comment this out - not clear what "Before" refers to, and not sure we need this sentence. what do you think?
-%Before, we had the optimal scenario of waiting until the first successful mutation appears and then waiting until is reaches fixation. Our approximations did quite well there. 
-Here, aneuploidy is transient and fixation of \anwt\ or \anmt\ could be outpaced by the arrival of the next genotype.
-This would explain the discrepancy between the approximations and simulation results in \Cref{fig:no direct mut}.
-To asses this further, \Cref{fig:fixT vs waitT} provides a clear picture.
-At least for intermediate to high values of the respective range of the parameters, the waiting time for appearance, $T_{i,j}$ is far shorter than the time to fixation, $\tau_{i,j}$.
-Indeed, if we ignore the fixation time of $\anwt$ and $\anmt$ ($T_{2n,2n+1}$ and $T_{2n+1,2n+1^*}$) in \Cref{eq:T_a}, we get a better fit to the data (compare red curves and blue markers in \Cref{fig:no direct mut}).
-Furthermore, setting $x=0.017$ for $T_{2n+1^*, 2n^*}$ in \Cref{eq:tau_fix_x}, we get an even better fit in some cases (\Cref{fig:no direct mut}B), but not others (\Cref{fig:no direct mut}A,C).
-% TODO maybe remove these x=0.017 sentences and the next sentence
-However, we stop simulations when genotype $\eumt$ reaches frequency of $0.95>0.017$.
 
@@
     \textbf{Posterior distribution validation.}
-    The posterior distribution of model parameters is roughly the same regardless of the number of simulations (4-10,000 replicates) used to approximate the likelihood (\Cref{eq:heatstress-likelihood}).
+    The posterior distribution of model parameters is roughly the same regardless of the number of simulations (4-10,000 replicates) used to approximate the likelihood (\cref{eq:heatstress-likelihood}).
     }

### 2022-04-03 ms/ms.tex
added vertical lines to diff-tau plots
t
fix a few typos

@@
 Given that one of these genotypes appeared, the probability that it was \anwt\ rather than \eumt\ is
-$p_A = \frac{p_{2n,2n+1}}{p_{2n,2n^*}+p_{2n,2n+1}}$.
+$p_a = \frac{p_{2n,2n+1}}{p_{2n,2n^*}+p_{2n,2n+1}}$.
 
@@
 \begin{equation}\label{eq:T_a} 
-T_{M} = T_{min} + \tau_{2n, 2n^*} \;,
+T_m = T_{min} + \tau_{2n, 2n^*} \;,
 \end{equation}

### 2022-04-04 ms/ms.tex
todo IK
minor editing

--- 2022-03-30 ms/ms.tex
@@
 % TODO cite https://www.biorxiv.org/content/10.1101/2020.10.07.327759v1
-% TODO cite Yang, F. et al. The fitness costs and benefits of trisomy of each Candida albicans chromosome. Genetics 218, 1–7 (2021). https://academic.oup.com/genetics/article/218/2/iyab056/6218773
 % cite: https://www.cell.com/developmental-cell/fulltext/S1534-5807(21)00562-1?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS1534580721005621%3Fshowall%3Dtrue
@@
 We fit this model to the experimental results of \citet{Yona2012} using an \emph{approximate Bayesian computation} framework~\citep{Sisson2009} to infer model parameters, including selection coefficients and rates of aneuploidy and mutation, and to test hypotheses about the evolutionary process by performing model selection between different versions of the model.
-We also mathematically analyze this evolutionary-genetic model to estimate the effects of selection, mutation, and aneuploidy on the adaptation time and the probability for adaptation via aneuploidy.
+%We also mathematically analyze this evolutionary-genetic model to estimate the effects of selection, mutation, and aneuploidy on the adaptation time and the probability for adaptation via aneuploidy.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph*{Evolutionary Models.}
+\paragraph*{Evolutionary Model.}
 We model the evolution of a population of cells using a Wright-Fisher model~\citep{Otto2007}, assuming a constant effective population size $N$, non-overlapping generations, and including the effects of natural selection, genetic drift, aneuploidy, and mutation.
@@
 
-\paragraph{Model comparison.} The main criterion that we use for model selection is WAIC, the widely applicable information criterion ~\citep{gelman2013bayesian},
+\paragraph{Model comparison.} 
+We examine several versions of our evolutionary models, e.g. without aneuploidy or with increased mutation rate in aneuploid cells, as well as several versions of prior distributions (see below).
+To compare these, we use WAIC and posterior predictive plots.
+WAIC, or the widely applicable information criterion ~\citep{gelman2013bayesian}, is defined as
 \begin{equation} \label{eq:WAIC}
@@
 where $\theta$ is a parameter vector, and $\mathbb{E}[\cdot]$ and $\mathbb{V}[\cdot]$ are the expectation and variance taken over the posterior distribution, which in practice are approximated using 50,000 samples from the posterior KDE.
-It was validated that WAIC doesn't change significantly, and WAIC differences between models are presaved when we redraw 50,000 samples.
+We validated that upon resampling WAIC values do not significantly change and that differences in WAIC between models are preserved.
 WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy~\citep{Kass1995}.
@@
 These plots visualize the fit of each model to the data.
-Also, for similar models we plot the marginal and joint posterior distributions of the parameters; if these are similar, we consider the models are interchangeable.
+Also, for similar models we plot the marginal and joint posterior distributions of the parameters; if these are similar, we consider the models interchangeable.
 We validate this by comparing HDI (highest density interval) of posterior distributions.
@@
 In our extended informative prior distribution, we used additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{\eumt}/w_{\anwt}$~(\Cref{fig:growth-curves}L).
-The same prior was used for $w_{\eumt}/w_{\anmt}$.
+The same distribution was used for $w_{\eumt}/w_{\anmt}$.
 
@@
 
-\subsection*{Statistical inference}
-
 \paragraph{Parameter estimation.} 
@@
 The estimated aneuploidy rate, $\delta=1.722\cdot10^{-3}\ [1.394\cdot10^{-3}-2.754\cdot10^{-3}]$, agrees with previous estimates.
-% TODO refs (9.7*10-5 in Zhu 2014, so it's lower).
-YR: we need more than one ref; also what is the units of Zho 2014?
+% TODO IK refs (9.7*10-5 in Zhu 2014, so it's lower).
+ILIA - we need more than one ref; also what is the units of Zho 2014?
 The estimated mutation rate, $\mu=2.942\cdot10^{-6}\ [2.017\cdot10^{-7}-4.15\cdot10^{-6}]$, corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
@@
 % generated with alt-prior.ipynb
-Finally, we estimated the parameters under an extended informative prior (see \emph{Prior distributions}), with different priors for $w_{\anmt}$ and $w_{\eumt}$.
-The estimated mutation rate was much lower compared to the basic prior, with $\mu=2.379\cdot10^{-9}\ [2.267\cdot10^{-9}-2.458\cdot10^{-9}]$.
+Finally, we estimated the parameters under an extended informative prior (see \emph{Prior distributions}), with more informed priors for $w_{\anmt}$ and $w_{\eumt}$.
+The posterior predictive plot show that inference with this extended prior produces a posterior distribution that fails to explain the empirical observations~(Figure~\ref{fig:2n*-fixation}).
+Despite this, it produces a lower WAIC, $-57$ compared to $295$ with our standard prior, as the inferred posterior distribution is considerably narrower (compare \Cref{fig:posterior,fig:posterior-alt}), which greatly affects WAIC due to its dependence on variance (\cref{eq:WAIC}).
+The estimated mutation rate was much lower compared to the previously used prior, with $\mu=2.379\cdot10^{-9}\ [2.267\cdot10^{-9}-2.458\cdot10^{-9}]$.
 Other parameter estimates are: $\delta=2.636\cdot10^{-3}\ [2.245\cdot10^{-3}-2.919\cdot10^{-3}]$,
@@
 $w_{\eumt}=1.033\ [1.031-1.036]$.
-The WAIC of the model is much lower: -57 compared to 295~(\Cref{table:WAIC}). However the model fails to explain the experimental observations~(Figure~\ref{fig:2n*-fixation}).
 
@@
 Thus, the blue distribution is a better fit compared to the green and pink, and the yellow histogram has a very poor fit.
-The last bin contains all the simulations with time equal or greater than 3,000 - from pink and green distributions.
+The last bin contains all values equal or greater than 3,000.
     \textbf{(B)}
 	Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy in blue and green, respectively, and corresponding to the blue and green bars in panel A.
-The MAP likelihood (\Cref{eq:heatstress-likelihood}) is 0.84, 0.67 with and without aneuploidy, respectively, and 0.14 for the model with aneuploidy and extended prior distribution.
+The MAP likelihood (\cref{eq:heatstress-likelihood}) is 0.84, 0.67, and 0.14 for the models represented by blue, green, and pink distributions, respectively.
+% TODO IK check that last sentence is correct, I modified from your phrasing.
      }
@@
 \pagebreak
-%%%%%%%%%%
-\subsection*{Analysis}
 
@@
 % TODO IK 10_000 reps is mocked, should be rerunned and fixed
+% TODO IK the colors are hard to distinguish, let's try something else. maybe YlGnBu?
 \begin{figure}[h!]
@@
     \textbf{Model with elevated mutation rate in aneuploid cells.}  \textbf{(A-E)} The inferred posterior distributions for models with different values of $\tau$, the fold-increase in mutation rate in aneuploid cells (\anwt\ and \anmt).
+Vertical dashed lines represent MAP (maximum a posteriori) of each distribution.
 When the increase in mutation rate is high, $\tau=10$ and $\tau=100$, the inferred mutation (A) and aneuploidy (B) rate tends to be lower.
@@
   }
-  % TODO IK maybe add dashed vertical lines with the same colors to show where the MAP of each distribution is?
   \end{figure}
@@
     \textbf{Model with fixed mutation rate.}  \textbf{(A-D)} The inferred posterior distributions for models with free and fixed mutation rate, $\mu$.
-The MAP (maximum a posteriori) and 50\% HDI (highest density interval) for each model are following:
-\textbf{free $\boldsymbol{\mu}$:}
+The MAP (maximum a posteriori) and 50\% HDI (highest density interval) for each model are:
+\textbf{free $\boldsymbol{\mu}$, run 1:}
 $\delta=2.749\cdot10^{-3}\ [1.476\cdot10^{-3}-2.822\cdot10^{-3}]$,
@@
 \caption*{
-	WAIC (widely applicable information criterion; \autoref{eq:WAIC})~\citep{gelman2013bayesian} values for models: (1) without aneuploidy;  (2-4) with aneuploidy, fixed mutation rate, and $\tau=1$; (5-10) with aneuploidy, free mutation rate model parameter, and different $\tau$.
+	WAIC (widely applicable information criterion, \cref{eq:WAIC})~\citep{gelman2013bayesian} values for models with different configurations.
 WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy and a difference of 2 is a popular threshold for model comparison~\citep{Kass1995}.

### 2022-04-14 ms/ms.tex
a bit todos updated
10k reps plots fixed
some of AY comments

@@
 Aneuploidy fixed in all four population in the first 450 generations (hereafter, fixation or elimination of a genotype \emph{by generation $t$} means that more than 95\% or less than 5\% of the population carry the genotype at generation $t$, and possibly earlier).
-From unpublished results, aneuploidy did not fix before at least 200 generations elapsed.
-The experiment continued with two populations, in which aneuploidy was eliminated by generation 1,700 and 2,350.
+From re-analysis of data not published in the original paper, aneuploidy did not fix before at least 200 generations elapsed.
+The experiment continued with two populations, in which aneuploidy was eliminated by generation 1,700 and 2,350 while still under the same conditions of elevated heat (\SI{39}{\celsius}).
 
@@
 For example, $P_{\tilde{\vec X}}\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)$ is approximated by taking simulations in which \anwt\; fixed before generation 450 but not before generation 200, and computing the fraction of such simulations in which \eumt\; did not fix by generation 1,700, and hence aneuploidy did not extinct before generation 1,700.
-\Cref{fig:seeds} compares results with less and more simulated experiments, demonstrating that 1,000 simulations are likely enough.
+\Cref{fig:seeds} compares results with less and more simulated experiments, demonstrating that 1,000 simulations are likely sufficient.
 
@@
 The estimated aneuploidy rate, $\delta=1.722\cdot10^{-3}\ [1.394\cdot10^{-3}-2.754\cdot10^{-3}]$, agrees with previous estimates.
-% TODO IK refs (9.7*10-5 in Zhu 2014, so it's lower).
+% TODO IK refs (9.7^10-5 in Zhu 2014 aneuploidy event per genome, 6.7^10-6 for chromosome 3 (2/(145*2062); losses only 0.7*10^-5).
 ILIA - we need more than one ref; also what is the units of Zho 2014?
@@
 The MAP likelihood (\cref{eq:heatstress-likelihood}) is 0.84, 0.67, and 0.14 for the models represented by blue, green, and pink distributions, respectively.
-% TODO IK check that last sentence is correct, I modified from your phrasing.
      }
@@
 % TODO IK 10_000 reps is mocked, should be rerunned and fixed
-% TODO IK the colors are hard to distinguish, let's try something else. maybe YlGnBu?
 \begin{figure}[h!]

### 2022-04-16 ms/ms.tex
added references for aneuploidy rate

--- 2022-04-04 ms/ms.tex
@@
 Acceptance is determined according to the approximate likelihood~(\cref{eq:heatstress-likelihood}), which has a maximum value of 0.875.
-Thus, we terminated the inference when $\epsilon \le 0.13$ after six iterations, with $n_6=982$ accepted parameter vectors and effective sample size ESS=651 (\Cref{fig:convergence}).
+Thus, we terminated the inference when $\epsilon \le 0.13$ %TODO IK for extended prior model and probably some more we used another epsilon.
+after six iterations, with $n_6=982$ accepted parameter vectors and effective sample size ESS=651 (\Cref{fig:convergence}).
 Running the inference algorithm with different initialization seeds and less or more simulations for approximating the likelihood produced similar posterior distributions~(\Cref{fig:seeds}).
@@
 %% estimates generated by MAP-and-hdi.ipynb
-The estimated aneuploidy rate, $\delta=1.722\cdot10^{-3}\ [1.394\cdot10^{-3}-2.754\cdot10^{-3}]$, agrees with previous estimates.
-% TODO IK refs (9.7*10-5 in Zhu 2014, so it's lower).
-ILIA - we need more than one ref; also what is the units of Zho 2014?
 The estimated mutation rate, $\mu=2.942\cdot10^{-6}\ [2.017\cdot10^{-7}-4.15\cdot10^{-6}]$, corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
+The estimated aneuploidy rate, $\delta=1.722\cdot10^{-3}\ [1.394\cdot10^{-3}-2.754\cdot10^{-3}]$ is higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimates  $6.7\cdot10^{-6}$ aneuploidy gain rate, \citet{Kumaran2013} estimates $3.0-4.3\cdot10^{-5}$ aneuploidy loss rate (95\% confidence interval).
 The estimated fitness values are $w_{\anwt}=1.022\ [1.021-1.023]$,
@@
 %% generated with diff-tau.ipynb
-% TODO IK 10_000 reps is mocked, should be rerunned and fixed
-% TODO IK the colors are hard to distinguish, let's try something else. maybe YlGnBu?
 \begin{figure}[h!]

### 2022-04-18 ms/ms.tex
moved tables with fitness values inside the model illustration figure; also added table for s,b,c calculation

@@
   \centering
+
+  \begin{subfigure}[t]{0.03\textwidth} \textbf{A} \end{subfigure}
+  \begin{subfigure}{0.75\textwidth}
+    \centering
     \includegraphics[height=1.8in]{../figures/Fig1-A.pdf}      
-  \label{fig:model1}
+  \end{subfigure}
+
+  \begin{subfigure}[t]{0.03\textwidth} \textbf{B} \end{subfigure}
+  \begin{subfigure}{0.75\textwidth}
+	\centering
+	\begin{tabular}{lllll}
+	\emph{Genotype} $i$ & \euwt & \anwt & \anmt & \eumt \\
+	\hline
+	\emph{Fitness} $w_i$ & $1$ & $1-c+b$ & $(1-c)(1+s)+b$ & $1+s$               
+	\end{tabular}
+	\label{table:fitness}
+  \end{subfigure}
+
+    \begin{subfigure}[t]{0.03\textwidth} \textbf{C} \end{subfigure}
+    \begin{subfigure}{0.75\textwidth}
+	\centering
+	\begin{tabular}{llll}
+	\emph{s, b, c} & s & c & b \\
+	\hline
+	 \emph{Calculation} & $w_{\eumt-1}$ & $1-\dfrac{w_\anmt-w_\anwt}{w_\eumt-1}$ & $w_\anwt - 1 + c$               
+	\end{tabular}
+	\label{table:fitness}
+  \end{subfigure}
+
   \caption{
-    \textbf{Model illustration.}
+    \textbf{The model.} 
+    \textbf{(A)} Model Illustration.
     There are four genotypes in our model: euploid wild-type, \euwt; euploid mutant, \eumt; aneuploid wild-type, \anwt; and aneuploid mutant, \anmt.
@@
     Arrows denote transitions between genotypes, with transitions rates: $\mu$, beneficial mutation rate; $\delta$, aneuploidy rate.
+    \textbf{(B)} Fitness values. $s \ge 0$ is the selection coefficient of a beneficial mutation;
+$0 \le c \le 1$ is the fitness cost of aneuploidy;
+and $b \ge c$ is the selection coefficient, or fitness benefit, of aneuploidy.
+   \textbf{(C)} Calculations of $s,b,c$ from $w_i$ values, derived from (B) expressions.
    }
@@
 \end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%
@@
 That is, we assume chromosomes are gained and lost at the same rate, and we neglect events that form a less-fit genotype (i.e., \anwt\; to \euwt\; and \eumt\; to \anmt).
-The fitness values of the four genotypes are given by \Cref{table:fitness}.
-
-%%% Table: fitness
-\begin{table}[h]
-\centering
-\caption{\textbf{Fitness values.}}
-\begin{tabular}{lllll}
-\emph{Genotype} $i$ & \euwt & \anwt & \anmt & \eumt \\
-\hline
-\emph{Fitness} $w_i$ & $1$ & $1-c+b$ & $(1-c)(1+s)+b$ & $1+s$               
-\end{tabular}
-\label{table:fitness}
-\caption*{
-$s \ge 0$ is the selection coefficient of a beneficial mutation;
-$0 \le c \le 1$ is the fitness cost of aneuploidy;
-and $b \ge c$ is the selection coefficient, or fitness benefit, of aneuploidy.
-}
-\end{table}
-%%%%%%%%%%%%%%%%%%%%%%%%
+The fitness values of the four genotypes are given by \Cref{fig:models}B.
 
@@
     \end{equation}
-where the fitness values $w_i$ are given in \Cref{table:fitness} and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
+where the fitness values $w_i$ are given in \Cref{fig:models}B and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
 The effect of mutation and aneuploidy on genotype frequencies is given by
@@
 $w_{\eumt}=1.028\ [1.026-1.029]$, all relative to the fitness of \euwt, which is set to $w_{\euwt}=1$. 
-Thus, we can infer that the benefit and cost of trisomy was $b=2.5\%$ and $c=0.3\%$, and the benefit of the beneficial mutation was $2.8\%$~(\Cref{table:fitness}).
+Thus, we can infer that the benefit and cost of trisomy was $b=92\%$ and $c=89\%$, and the benefit of the beneficial mutation was $2.8\%$~(\Cref{fig:models}C).
 

### 2022-05-01 ms/ms.tex
revisions during meeting - remove fitness table, benefit cost equations, update figures and text

@@
 \author[3]{Avihu H. Yona}
-%\author[d]{Orna Dahan}
-%\author[d]{Yitzhak Pilpel}
 \author[1,2,$\dagger$]{Yoav Ram}
@@
 The Hebrew University of Jerusalem, Israel}
-%\affil[d]{Department of Molecular Genetics, Weizmann Institute of Science, Rehovot 76100, Israel}
 \affil[*]{These authors contributed equally to this work}
@@
   \centering
-
-  \begin{subfigure}[t]{0.03\textwidth} \textbf{A} \end{subfigure}
-  \begin{subfigure}{0.75\textwidth}
-    \centering
     \includegraphics[height=1.8in]{../figures/Fig1-A.pdf}      
-  \end{subfigure}
-
-  \begin{subfigure}[t]{0.03\textwidth} \textbf{B} \end{subfigure}
-  \begin{subfigure}{0.75\textwidth}
-	\centering
-	\begin{tabular}{lllll}
-	\emph{Genotype} $i$ & \euwt & \anwt & \anmt & \eumt \\
-	\hline
-	\emph{Fitness} $w_i$ & $1$ & $1-c+b$ & $(1-c)(1+s)+b$ & $1+s$               
-	\end{tabular}
-	\label{table:fitness}
-  \end{subfigure}
-
-    \begin{subfigure}[t]{0.03\textwidth} \textbf{C} \end{subfigure}
-    \begin{subfigure}{0.75\textwidth}
-	\centering
-	\begin{tabular}{llll}
-	\emph{s, b, c} & s & c & b \\
-	\hline
-	 \emph{Calculation} & $w_{\eumt-1}$ & $1-\dfrac{w_\anmt-w_\anwt}{w_\eumt-1}$ & $w_\anwt - 1 + c$               
-	\end{tabular}
-	\label{table:fitness}
-  \end{subfigure}
-
   \caption{
-    \textbf{The model.} 
-    \textbf{(A)} Model Illustration.
+    \textbf{Model Illustration.}
     There are four genotypes in our model: euploid wild-type, \euwt; euploid mutant, \eumt; aneuploid wild-type, \anwt; and aneuploid mutant, \anmt.
     Overall there are two possible trajectories from \euwt\; to \eumt.
-    Arrows denote transitions between genotypes, with transitions rates: $\mu$, beneficial mutation rate; $\delta$, aneuploidy rate.
-    \textbf{(B)} Fitness values. $s \ge 0$ is the selection coefficient of a beneficial mutation;
-$0 \le c \le 1$ is the fitness cost of aneuploidy;
-and $b \ge c$ is the selection coefficient, or fitness benefit, of aneuploidy.
-   \textbf{(C)} Calculations of $s,b,c$ from $w_i$ values, derived from (B) expressions.
+    Arrows denote transitions between genotypes, with transitions rates $\mu$ for the beneficial mutation rate and $\delta$ for the aneuploidy rate.
    }
@@
 
-Transitions between the genotypes occur as follows~(\Cref{fig:models}A): Beneficial mutations from \euwt\; to \eumt\; and from \anwt\; to \anmt\; occur with probability $\mu$, the mutation rate.
+Transitions between the genotypes occur as follows~(\Cref{fig:models}): Beneficial mutations from \euwt\; to \eumt\; and from \anwt\; to \anmt\; occur with probability $\mu$, the mutation rate.
 We neglect back-mutations (i.e., from \eumt\; to \euwt\; and from \anmt\; to \anwt).
@@
 That is, we assume chromosomes are gained and lost at the same rate, and we neglect events that form a less-fit genotype (i.e., \anwt\; to \euwt\; and \eumt\; to \anmt).
-The fitness values of the four genotypes are given by \Cref{fig:models}B.
 
@@
     \end{equation}
-where the fitness values $w_i$ are given in \Cref{fig:models}B and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
+where $w_i$ is the fitness of genotype $i$ and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
 The effect of mutation and aneuploidy on genotype frequencies is given by
@@
 Acceptance is determined according to the approximate likelihood~(\cref{eq:heatstress-likelihood}), which has a maximum value of 0.875.
-Thus, we terminated the inference when $\epsilon \le 0.13$ %TODO IK for extended prior model and probably some more we used another epsilon.
-after six iterations, with $n_6=982$ accepted parameter vectors and effective sample size ESS=651 (\Cref{fig:convergence}).
+We terminated the inference iterations when the change in $\epsilon$ value from one iteration to the next was small
+With our standard prior and model, we reached $\epsilon=0.13$ after six iterations, with $n_6=982$ accepted parameter vectors and effective sample size ESS=651 (\Cref{fig:convergence}).
 Running the inference algorithm with different initialization seeds and less or more simulations for approximating the likelihood produced similar posterior distributions~(\Cref{fig:seeds}).
@@
 \paragraph{Model comparison.} 
-We examine several versions of our evolutionary models, e.g. without aneuploidy or with increased mutation rate in aneuploid cells, as well as several versions of prior distributions (see below).
-To compare these, we use WAIC and posterior predictive plots.
-WAIC, or the widely applicable information criterion ~\citep{gelman2013bayesian}, is defined as
+We examine several versions of our evolutionary models, e.g. without aneuploidy or with increased mutation rate in aneuploid cells, as well as several different prior distributions (see below).
+To compare these, we plot posterior predictions: for each model we execute $10,000$ simulations using the MAP parameter estimates and plot the distributions of time to fixation of \eumt, one of key properties of the model likelihood.
+These plots visualize the fit of each model to the data.
+Also, for similar models we plot the marginal and joint posterior distributions of the parameters; if these are similar, we consider the models interchangeable.
+We validate this by comparing HDI (highest density interval) of posterior distributions.
+
+Where posterior plots are very similar, we use WAIC, or the widely applicable information criterion ~\citep{gelman2013bayesian}, defined as
 \begin{equation} \label{eq:WAIC}
@@
 
-We also plot posterior predictions: for each model we execute $10,000$ simulations using the MAP parameter estimates and plot the distributions of time to fixation of \eumt, one of key properties of the model likelihood.
-These plots visualize the fit of each model to the data.
-Also, for similar models we plot the marginal and joint posterior distributions of the parameters; if these are similar, we consider the models interchangeable.
-We validate this by comparing HDI (highest density interval) of posterior distributions.
-
 \paragraph{Prior distributions.} We used informative prior distributions for $w_{\anwt}=1-c+b$, $w_{\anmt}=(1+s)(1-c)+b$ and $w_{\eumt}=1+s$, which we estimated from growth curves data from mono-culture growth experiments previously reported by \citet[Figs.
@@
 We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{\anwt}$~(\Cref{fig:growth-curves}-D, assuming $w_{\euwt}=1$).
-In our standard informative prior distribution, we used the same prior for $w_{\anmt}$ and $w_{\eumt}$.
-To increase computational efficiency, we also assumed $w(\eumt)>w(\anmt)>w(\anwt)>w(\euwt)$; running the inference without this assumption produced similar results. 
+In our basic informative prior distribution, we used the same prior for $w_{\anmt}$ and $w_{\eumt}$.
+To increase computational efficiency, we also assumed $w_\eumt>w_\anmt>w_\anwt>w_\euwt$; running the inference without this assumption produced similar results. 
 In our extended informative prior distribution, we used additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{\eumt}/w_{\anwt}$~(\Cref{fig:growth-curves}L).
 The same distribution was used for $w_{\eumt}/w_{\anmt}$.
+To summarize, our basic informative prior uses a single prior distribution for all fitness values, whereas the extended informative prior uses one distribution for \anwt, and another distribution for both \anmt\ and \eumt.
 
@@
 The estimated mutation rate, $\mu=2.942\cdot10^{-6}\ [2.017\cdot10^{-7}-4.15\cdot10^{-6}]$, corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
-The estimated aneuploidy rate, $\delta=1.722\cdot10^{-3}\ [1.394\cdot10^{-3}-2.754\cdot10^{-3}]$ is higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimates  $6.7\cdot10^{-6}$ aneuploidy gain rate, \citet{Kumaran2013} estimates $3.0-4.3\cdot10^{-5}$ aneuploidy loss rate (95\% confidence interval).
+The estimated aneuploidy rate, $\delta=1.722\cdot10^{-3}\ [1.394\cdot10^{-3}-2.754\cdot10^{-3}]$ is higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimate $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
 The estimated fitness values are $w_{\anwt}=1.022\ [1.021-1.023]$,
@@
 $w_{\eumt}=1.028\ [1.026-1.029]$, all relative to the fitness of \euwt, which is set to $w_{\euwt}=1$. 
-Thus, we can infer that the benefit and cost of trisomy was $b=92\%$ and $c=89\%$, and the benefit of the beneficial mutation was $2.8\%$~(\Cref{fig:models}C).
+Thus, we can infer that the cost of trisomy is $c = w_{\eumt}-w_{\anmt} = 0.003$ (or 0.3\%) and the benefit of trisomy is $w_{\anwt}-1-c=0.019$ (1.9\%), whereas the benefit of beneficial mutation is $w_{\eumt}-1=0.028$ (2.8\%).
 
@@
 Finally, we estimated the parameters under an extended informative prior (see \emph{Prior distributions}), with more informed priors for $w_{\anmt}$ and $w_{\eumt}$.
-The posterior predictive plot show that inference with this extended prior produces a posterior distribution that fails to explain the empirical observations~(Figure~\ref{fig:2n*-fixation}).
+Inference took much longer to run but the posterior distribution seemed to converge as it did not change much in the final iterations. 
+The posterior predictive plot shows that inference with this extended prior produces a posterior distribution that fails to explain the empirical observations~(Figure~\ref{fig:2n*-fixation}).
 Despite this, it produces a lower WAIC, $-57$ compared to $295$ with our standard prior, as the inferred posterior distribution is considerably narrower (compare \Cref{fig:posterior,fig:posterior-alt}), which greatly affects WAIC due to its dependence on variance (\cref{eq:WAIC}).
-The estimated mutation rate was much lower compared to the previously used prior, with $\mu=2.379\cdot10^{-9}\ [2.267\cdot10^{-9}-2.458\cdot10^{-9}]$.
-Other parameter estimates are: $\delta=2.636\cdot10^{-3}\ [2.245\cdot10^{-3}-2.919\cdot10^{-3}]$,
-$w_{\anwt}=1.022\ [1.021-1.023]$,
-$w_{\anmt}=1.032\ [1.03-1.035]$,
-$w_{\eumt}=1.033\ [1.031-1.036]$.
+The estimated mutation rate was much lower compared to the previously used prior, with $\mu=2.475\cdot10^{-9}\ [2.42\cdot10^{-9}-2.609\cdot10^{-9}]$.
+Other parameter estimates are: $\delta=2.705\cdot10^{-3}\ [2.112\cdot10^{-3}-3.107\cdot10^{-3}]$,
+$w_{\anwt}=1.022\ [1.021-1.024]$,
+$w_{\anmt}=1.029\ [1.027-1.03]$,
+$w_{\eumt}=1.03\ [1.029-1.031]$. 
+Notably, the maximum posterior ratio $w_{\eumt}/w_{\anwt}=1.007$ is much lower than the maximum prior ratio of $1.033$ (\Cref{fig:growth-curves}H) and closer to the ratio of $1$ that we assume in our standard prior.
 
@@
     \textbf{Model fit with and without aneuploidy.}
-    \textbf{(A)} The distribution of time to fixation of \eumt\ (i.e., adaptation time) in $10,000$ simulations of the model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange), and a model fitted to the data assuming $\delta=0$ (green); and to the model with aneuploidy with extended prior distribution (pink).
+    \textbf{(A)} The distribution of time to fixation of \eumt\ (i.e., adaptation time) in $10,000$ simulations of the model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange), and a model fitted to the data assuming $\delta=0$ (green); and to the model with aneuploidy and an extended prior distribution (pink).
 In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and another by generation 2,350 (dashed lines) but not before generation 450.
 Thus, the blue distribution is a better fit compared to the green and pink, and the yellow histogram has a very poor fit.
-The last bin contains all values equal or greater than 3,000.
+The last bin contains all values equal or greater than 3,000 and the green bar is partially concealed by the pink bar.
     \textbf{(B)}
@@
     \textbf{Fitness estimation from growth curves.}
-    \textbf{(A-D)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{39}{\celsius}. $\hat w_{\anwt}/w_{\euwt}$=1.024 (95\% CI: 0.959 - 1.115). \texttt{Curveball} 
-    \textbf{(E-H)} Fitness estimation from growth curves of \anwt\ and \eumt\ at \SI{39}{\celsius}. $\hat w_{\eumt}/w_{\anwt}$=1.033 (95\% CI: 1.027 - 1.041).
+    \textbf{(A-D)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{39}{\celsius}. $ w_{\anwt}/w_{\euwt}$=1.024 (95\% CI: 0.959 - 1.115). \texttt{Curveball} 
+    \textbf{(E-H)} Fitness estimation from growth curves of \anwt\ and \eumt\ at \SI{39}{\celsius}. $ w_{\eumt}/w_{\anwt}$=1.033 (95\% CI: 1.027 - 1.041).
     Growth curves previously described in \citet[Figs. 3C, 4A, and S2]{Yona2012}.
@@
 
-%TODO YR read this 
 %% generated with WAIC.ipynb 
@@
 \caption*{
-	WAIC (widely applicable information criterion, \cref{eq:WAIC})~\citep{gelman2013bayesian} values for models with different configurations.
+	WAIC (widely applicable information criterion, \cref{eq:WAIC}, ~\citep{gelman2013bayesian}) values for models with different configurations.
 WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy and a difference of 2 is a popular threshold for model comparison~\citep{Kass1995}.

### 2022-05-03 src/pyabc_visualization_patch/pyabc/kde.py
updated alt-prior plot with t=18

@@ -880,15 +880,15 @@ def plot_kde_matrix_for_paper_extended(df, w, limits=None, colorbar=True, height
     def hist_1d(x, ax,i):
         df = pd.concat((x,), axis=1)
         import seaborn as sns
-        sns.histplot(df[x.name], ax=ax, bins=30, color='darkgray')
+        sns.histplot(df[x.name], ax=ax, bins=20)
         ax.ticklabel_format(style='plain')
         ax.axvline(refval[x.name], color=refval_color, linestyle='dotted',linewidth=2)
 #         ax.set_xlim(min(df[x.name])-1e-9) #1e-8 to print 0
         ax.set_yticks([])
         if i==0:
-            ax.text(0.92, -0.2, '$x10^{-9}$', transform=ax.transAxes)
-            ax.set_xticks([2e-9, 2.5e-9, 3e-9])
-            ax.set_xticklabels([float(a) for a in ax.get_xticks()*1e9]);
+            ax.text(0.92, -0.17, '$x10^{-9}$', transform=ax.transAxes)
+            ax.set_xticks([2.25e-9, 2.5e-9, 2.75e-9])
+            ax.set_xticklabels([round(float(a),2) for a in ax.get_xticks()*1e9]);
         if i==1:
             ax.set_xticks([0.002,0.004, 0.006])
             ax.set_xticklabels([0.002,0.004, 0.006])
@@ -912,9 +912,9 @@ def plot_kde_matrix_for_paper_extended(df, w, limits=None, colorbar=True, height
             hist_2d(x, y, ax)
             if j == 0 : #mutation rate
                 ax.ticklabel_format(style='plain')
-                ax.text(0.92, -0.2, '$x10^{-9}$', transform=ax.transAxes)
-                ax.set_xticks([2e-9, 2.5e-9, 3e-9])
-                ax.set_xticklabels([float(a) for a in ax.get_xticks()*1e9]);
+                ax.text(0.92, -0.17, '$x10^{-9}$', transform=ax.transAxes)
+                ax.set_xticks([2.25e-9, 2.5e-9, 2.75e-9])
+                ax.set_xticklabels([round(float(a),2) for a in ax.get_xticks()*1e9]);
             if j == 1: #aneuploidy rate
                 ax.set_xticks([0.002,0.004, 0.006])
                 ax.set_xticklabels([0.002,0.004, 0.006])

### 2022-05-06 ms/ms.tex
small TODO was added

@@
 \paragraph*{Evolutionary Model.}
-We model the evolution of a population of cells using a Wright-Fisher model~\citep{Otto2007}, assuming a constant effective population size $N$, non-overlapping generations, and including the effects of natural selection, genetic drift, aneuploidy, and mutation.
+We model the evolution of a population of cells using a Wright-Fisher model~\citep{Otto2007}, assuming a constant effective population size $N$, %TODO we need to write in some place that it is 6.425*10**6 in our simulations: 1.6M to 2004.8M in Yona experiment
+ non-overlapping generations, and including the effects of natural selection, genetic drift, aneuploidy, and mutation. 
 We focus on beneficial genetic modifications, neglecting the effects of deleterious and neutral mutations or karyotypic changes.
@@
 
+% Fig dynamics
+%% generated with dynamics-plot.ipynb
 \begin{figure}[p]

### 2022-05-08 ms/ms.tex
population size
epsilon is 1-lik, again

--- 2022-05-01 ms/ms.tex
@@
 
+In the experiment by \citet{Yona2012}, the population was grown every day from $1.6 \cdot 10^6$ cells until reaching stationary phase and then diluted 1:120.
+Thus, we set the population size to $N=6.425 \cdot 10^6$, the harmonic mean of $\{2^k \cdot 1.6 \cdot 10^6\}_{k=0}^{7}$~\citep{Crow1970}.
 The initial population has $N$ cells with genotype \euwt. 
@@
 
-Acceptance is determined according to the approximate likelihood~(\cref{eq:heatstress-likelihood}), which has a maximum value of 0.875.
-We terminated the inference iterations when the change in $\epsilon$ value from one iteration to the next was small
-With our standard prior and model, we reached $\epsilon=0.13$ after six iterations, with $n_6=982$ accepted parameter vectors and effective sample size ESS=651 (\Cref{fig:convergence}).
+Acceptance is determined according to the approximate likelihood~(\cref{eq:heatstress-likelihood}), which has a maximum value of $\likelihood_{max}=0.875$ (giving a minimal value of $\epsilon_{min}=0.125$).
+We terminated the inference iterations when the change in $\epsilon$ value from one iteration to the next was small.
+With our standard prior and model, we reached $\epsilon=0.13$ (or $\likelihood=0.87$) after six iterations, with $n_6=982$ accepted parameter vectors and effective sample size ESS=651 (\Cref{fig:convergence}).
 Running the inference algorithm with different initialization seeds and less or more simulations for approximating the likelihood produced similar posterior distributions~(\Cref{fig:seeds}).
@@
     \textbf{(F-I)} ABC-SMC measures of convergence.
-After iteration number 6, the acceptance threshold was $\epsilon=0.13$, the acceptance rate was $0.018$, the number of particles was 982, and the effective sample size ESS=651.
+After iteration number 6, the acceptance threshold was $\epsilon=0.13$ (i.e., $\likelihood=0.87$, \cref{eq:heatstress-likelihood}), the acceptance rate was $0.018$, the number of particles was 982, and the effective sample size ESS=651.
 }

### 2022-05-08 ms/ms.tex
started reorganizing results

@@
 
-\paragraph{Parameter estimation.} 
+\paragraph{Rates and fitness effects of aneuploidy and mutation.} 
 We used ABC-SMC to infer the posterior distribution of model parameters~(\Cref{fig:posterior}). 
@@
 %% estimates generated by MAP-and-hdi.ipynb
-The estimated mutation rate, $\mu=2.942\cdot10^{-6}\ [2.017\cdot10^{-7}-4.15\cdot10^{-6}]$, corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
-The estimated aneuploidy rate, $\delta=1.722\cdot10^{-3}\ [1.394\cdot10^{-3}-2.754\cdot10^{-3}]$ is higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimate $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
+The estimated mutation rate, $\mu=2.942\cdot10^{-6}\ [2.017\cdot10^{-7}-4.15\cdot10^{-6}]$, corresponds to a mutation target size of $\sim 10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
+The estimated aneuploidy rate, $\delta=1.722\cdot10^{-3}\ [1.394\cdot10^{-3}-2.754\cdot10^{-3}]$ is much higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
 The estimated fitness values are $w_{\anwt}=1.022\ [1.021-1.023]$,
@@
 %% generated with posterior-plot.ipynb
-\begin{figure}[p]
+\begin{figure}[h!]
   \centering
@@
 
-\paragraph{Model checking and comparison.}
-The model fits the data well: in simulations using the MAP parameter estimates, \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\Cref{fig:2n*-fixation}B).
-Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\Cref{fig:ppc-plot}).
+\paragraph{Role of aneuploidy in the evolutionary dynamics.}
 
-Moreover, a model without aneuploidy (where the aneuploidy rate is fixed at zero, $\delta=0$), fails to explain the experimental observations~(Figure~\ref{fig:2n*-fixation}).
+Our model fits the data well: in simulations using the MAP parameter estimates, \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\Cref{fig:2n*-fixation}B).
+
+However, a model without aneuploidy (where the aneuploidy rate is fixed at zero, $\delta=0$), fails to explain the experimental observations~(Figure~\ref{fig:2n*-fixation}).
 %% generated with MAP-and-hdi.ipynb
@@
 Even with these lower estimates, the model fit is worse than that of a model with aneuploidy~(\Cref{fig:2n*-fixation}).
+We also checked a model in which aneuploidy is neutral compared to the wild type, that is, $w_{\anwt} = w_{\euwt}$ and $w_{\anmt} = w_{\eumt}$ but $\delta>0$.
+This model fits the data better than the model with no aneuploidy (in which $\delta=0$), but worse than a model with positive selection for aneuploidy, in which $w_{\anwt}>w_{\euwt}$~(\Cref{fig:2n*-fixation}).
+Thus, our results suggest that aneuploidy plays a significant role in the evolutionary dynamics \citep{Yona2012,Yona2015}.
+
+% Fig model comparison 
+%% generated with with-aneuploidy_vs_no_aneuploidy.ipynb
+\begin{figure}[h!]
+  \begin{subfigure}{0.5\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{../figures/fixation-plot-a.pdf}      
+      \label{fig:fit}
+  \end{subfigure}
+  \begin{subfigure}{0.5\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{../figures/fixation-plot-b.pdf}      
+      \label{fig:fit-cumulative}
+  \end{subfigure}
+  \caption{
+    \textbf{Model fit with and without aneuploidy.}
+    \textbf{(A)} The distribution of time to fixation of \eumt\ (i.e., adaptation time) in $10,000$ simulations of the model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange), and a model fitted to the data assuming $\delta=0$ (green); and to the model with aneuploidy and an extended prior distribution (pink).
+In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and another by generation 2,350 (dashed lines) but not before generation 450.
+Thus, the blue distribution is a better fit compared to the green and pink, and the yellow histogram has a very poor fit.
+The last bin contains all values equal or greater than 3,000 and the green bar is partially concealed by the pink bar.
+    \textbf{(B)}
+	Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy in blue and green, respectively, and corresponding to the blue and green bars in panel A.
+The MAP likelihood (\cref{eq:heatstress-likelihood}) is 0.84, 0.67, and 0.14 for the models represented by blue, green, and pink distributions, respectively.
+     }
+  \label{fig:2n*-fixation}
+\end{figure}
 
@@
 
-% Fig model with and without aneuploidy
-%% generated with with-aneuploidy_vs_no_aneuploidy.ipynb
-\begin{figure}[p]
-  \begin{subfigure}{0.5\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../figures/fixation-plot-a.pdf}      
-      \label{fig:fit}
-  \end{subfigure}
-  \begin{subfigure}{0.5\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../figures/fixation-plot-b.pdf}      
-      \label{fig:fit-cumulative}
-  \end{subfigure}
-  \caption{
-    \textbf{Model fit with and without aneuploidy.}
-    \textbf{(A)} The distribution of time to fixation of \eumt\ (i.e., adaptation time) in $10,000$ simulations of the model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange), and a model fitted to the data assuming $\delta=0$ (green); and to the model with aneuploidy and an extended prior distribution (pink).
-In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and another by generation 2,350 (dashed lines) but not before generation 450.
-Thus, the blue distribution is a better fit compared to the green and pink, and the yellow histogram has a very poor fit.
-The last bin contains all values equal or greater than 3,000 and the green bar is partially concealed by the pink bar.
-    \textbf{(B)}
-	Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy in blue and green, respectively, and corresponding to the blue and green bars in panel A.
-The MAP likelihood (\cref{eq:heatstress-likelihood}) is 0.84, 0.67, and 0.14 for the models represented by blue, green, and pink distributions, respectively.
-     }
-  \label{fig:2n*-fixation}
-\end{figure}
-
-% Fig dynamics
-%% generated with dynamics-plot.ipynb
-\begin{figure}[p]
-  \centering
-\includegraphics[width=0.8\textwidth]{../figures/dynamics.pdf}
-  \caption{
-  \textbf{Posterior genotype frequency dynamics for the model.} 
-  The posterior prediction for the frequencies of the four genotypes over time.
-Each of the 20 curves is the average of 10,000 simulations of the model using parameters drawn from the posterior distribution.
-  }
-  \label{fig:ppc-plot}
-\end{figure}
-
 % Fig tau plots
 %% generated by diff-tau-plots.ipynb
-\begin{figure}[p]
+\begin{figure}[p!]
   \centering
@@
   \label{fig:tau-plots}
+\end{figure}
+
+\paragraph*{Genotype frequency dynamics.}
+Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\Cref{fig:ppc-plot}).
+% TODO new figure by Martin and discussion of the probability of 2n* arriving via 2n+1.
+
+% Fig dynamics
+%% generated with dynamics-plot.ipynb
+\begin{figure}[h!]
+  \centering
+\includegraphics[width=0.8\textwidth]{../figures/dynamics.pdf}
+  \caption{
+  \textbf{Posterior genotype frequency dynamics for the model.} 
+  The posterior prediction for the frequencies of the four genotypes over time.
+Each of the 20 curves is the average of 10,000 simulations of the model using parameters drawn from the posterior distribution.
+  }
+  \label{fig:ppc-plot}
 \end{figure}

### 2022-05-08 ms/ms.tex
cont revising results

@@
 Even with these lower estimates, the model fit is worse than that of a model with aneuploidy~(\Cref{fig:2n*-fixation}).
-We also checked a model in which aneuploidy is neutral compared to the wild type, that is, $w_{\anwt} = w_{\euwt}$ and $w_{\anmt} = w_{\eumt}$ but $\delta>0$.
-This model fits the data better than the model with no aneuploidy (in which $\delta=0$), but worse than a model with positive selection for aneuploidy, in which $w_{\anwt}>w_{\euwt}$~(\Cref{fig:2n*-fixation}).
+
+We also checked a model in which aneuploidy occurs but is adaptively neutral compared to the wild type, that is, $w_{\anwt} = w_{\euwt}$ and $w_{\anmt} = w_{\eumt}$ but $\delta>0$.
+This model fits the data better than the model with no aneuploidy (in which $\delta=0$), but worse than a model with positive selection for aneuploidy, in which $w_{\euwt}<w_{\anwt}<w_{\anmt}<w_{\eumt}$~(\Cref{fig:2n*-fixation}).
 Thus, our results suggest that aneuploidy plays a significant role in the evolutionary dynamics \citep{Yona2012,Yona2015}.
@@
     \textbf{Model fit with and without aneuploidy.}
-    \textbf{(A)} The distribution of time to fixation of \eumt\ (i.e., adaptation time) in $10,000$ simulations of the model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange), and a model fitted to the data assuming $\delta=0$ (green); and to the model with aneuploidy and an extended prior distribution (pink).
+    The distribution of time to fixation of \eumt\ (i.e., adaptation time) in $10,000$ simulations using MAP parameters of the model with beneficial aneuploidy (blue; $\delta>0, \; w_{\euwt}<w_{\anwt}<w_{\anmt}<w_{\eumt}$) compared to alternative models: a model with the same parameter values but without aneuploidy (gray, $\delta=0$, concentrated at $t=450$); a model fitted to the data assuming no aneuploidy (green, $\delta=0$); a model fitted to the data assuming neutral aneuploidy (yellow, $\delta>0, \; w_{\anwt}=w_{\euwt}, \; w_{\anmt}=w_{\eumt}$); and a model with beneficial aneuploidy and an extended prior distribution (pink).
 In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and another by generation 2,350 (dashed lines) but not before generation 450.
-Thus, the blue distribution is a better fit compared to the green and pink, and the yellow histogram has a very poor fit.
-The last bin contains all values equal or greater than 3,000 and the green bar is partially concealed by the pink bar.
-    \textbf{(B)}
-	Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy in blue and green, respectively, and corresponding to the blue and green bars in panel A.
-The MAP likelihood (\cref{eq:heatstress-likelihood}) is 0.84, 0.67, and 0.14 for the models represented by blue, green, and pink distributions, respectively.
+Thus, the blue distribution has a better fit compared to the other distributions (the gray distribution has a particulary poor fit).
+The MAP likelihood (\cref{eq:heatstress-likelihood}) is 0.84, 0.78, 0.67, and 0.14 for the models represented by blue, yellow, green, and pink distributions, respectively. 
+\textbf{(A)} Histogram of the time to fixation of \eumt. The last bin contains all values equal or greater than 3,000.
+\textbf{(B)} Cumulative distribution of the time to fixation. 
      }
@@
 
+\paragraph*{Genetic instability in aneuploid cells.}
 It has been suggested that aneuploidy increased genetic instability~\citep{Sheltzer2011b}.
-Therefore, we inferred model parameters under the assumption that the mutation rate increases in aneuploid cells by a factor $\tau$ = 33/32 (due to an additional chromosome), 2, 5, 10, or 100 (due to genetic instability).
+Therefore, we inferred model parameters under the assumption that the mutation rate increases in aneuploid cells by a factor $\tau$ = 1, 33/32 (due to an additional chromosome), 2, 5, 10, or 100 (due to genetic instability).
 We found that the posterior distribution was similar for $\tau=$1, 33/32, 2, and 5 (\Cref{fig:tau}).
@@
     \textbf{Model with elevated mutation rate in aneuploid cells.}  \textbf{(A-E)} The inferred posterior distributions for models with different values of $\tau$, the fold-increase in mutation rate in aneuploid cells (\anwt\ and \anmt).
-Vertical dashed lines represent MAP (maximum a posteriori) of each distribution.
-When the increase in mutation rate is high, $\tau=10$ and $\tau=100$, the inferred mutation (A) and aneuploidy (B) rate tends to be lower.
+Vertical dashed lines represent the MAP (maximum a posteriori) of each distribution.
+When the increase in mutation rate is high, $\tau=10$ and $\tau=100$, the inferred mutation (A) and aneuploidy (B) rates tend to be lower.
     \textbf{(F-K)} The inferred joint posterior distribution of mutation rate ($\mu$) and aneuploidy rate ($\delta$) with different $\tau$ values (dark purple and bright yellow for low and high density, respectively).

### 2022-05-08 ms/ms.tex
finished revising results; reorder supp figures; move sensitivity and extended prior to supp

@@
 
-\paragraph{Prior distributions.} We used informative prior distributions for $w_{\anwt}=1-c+b$, $w_{\anmt}=(1+s)(1-c)+b$ and $w_{\eumt}=1+s$, which we estimated from growth curves data from mono-culture growth experiments previously reported by \citet[Figs.
+\paragraph{Prior distributions.} \label{par:prior}
+We used informative prior distributions for $w_{\anwt}=1-c+b$, $w_{\anmt}=(1+s)(1-c)+b$ and $w_{\eumt}=1+s$, which we estimated from growth curves data from mono-culture growth experiments previously reported by \citet[Figs.
 3C, 4A, and S2]{Yona2012}.
@@
 
-We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{\anwt}$~(\Cref{fig:growth-curves}-D, assuming $w_{\euwt}=1$).
-In our basic informative prior distribution, we used the same prior for $w_{\anmt}$ and $w_{\eumt}$.
+We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate an informative prior distribution for $w_{\anwt}$~(\Cref{fig:growth-curves}-D, assuming $w_{\euwt}=1$).
+In this prior distribution, we used the same prior for $w_{\anmt}$ and $w_{\eumt}$. 
 To increase computational efficiency, we also assumed $w_\eumt>w_\anmt>w_\anwt>w_\euwt$; running the inference without this assumption produced similar results.
-In our extended informative prior distribution, we used additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{\eumt}/w_{\anwt}$~(\Cref{fig:growth-curves}L).
-The same distribution was used for $w_{\eumt}/w_{\anmt}$.
-To summarize, our basic informative prior uses a single prior distribution for all fitness values, whereas the extended informative prior uses one distribution for \anwt, and another distribution for both \anmt\ and \eumt.
+See \emph{supporting material} for an extended informative prior distribution that uses growth curves of \eumt\ and \anwt\ growing in \SI{39}{\celsius}; this prior distribution proved to be less useful.
 
@@
 We used ABC-SMC to infer the posterior distribution of model parameters~(\Cref{fig:posterior}). 
-We report parameter estimates using the MAP (maximum a posteriori) and providing the 50\% HDI (highest density interval) in square brackets.
+We report parameter estimates using the MAP (maximum a posteriori) and providing the 50\% HDI (highest density interval) in square brackets. See \emph{Supporting material} for sensitivity analysis.
 %% estimates generated by MAP-and-hdi.ipynb
@@
 With $\tau=10$, the estimated mutation rate was only slightly lower compared to $\tau=1$ ($\mu=1.674\cdot10^{-6}\ [2.501\cdot10^{-7}-1.741\cdot10^{-6}]$). 
-WAIC is lowest for $\tau=100$~(\Cref{table:WAIC}), and if we rule such a strong effect of genetic instability a-priori, the next two lowest WAIC values are for $\tau=33/32$ and $\tau=5$.  
-Therefore, we cannot rule out an increased mutation rate in aneuploid cells, but unless the effect is strong ($\tau=100$), it does not seem to affect our inference results.
-We also checked the differences in fixation/loss times of  \euwt, \anwt and \eumt\  for the models with different $\tau$.
-We saw that the models can be distinguished using these parameters, especially $\tau=100$~(\Cref{fig:tau-plots}).
-
-Sensitivity analysis shows that changing a single parameter while keeping the rest fixed at the MAP estimate produces a worse fit to the data~(\Cref{fig:sensitivity}).
-Furthermore, we fitted models with a mutation rate fixed at $\mu$=$10^{-5}$, $10^{-6}$ and $10^{-7}$.
-We inferred similar parameters estimates for the model with $\mu=10^{-6}$ compared to the model with a free $\mu$ parameter.
-Models with $\mu$=$10^{-5}$ and $\mu$=$10^{-7}$ inferred different parameters estimates~(\Cref{fig:mu}).
-WAIC was lower when $\mu$ is fixed~(\Cref{table:WAIC}), but this is not surprising, as WAIC attempt to balance between model fit and model complexity, where the latter takes into account the number of model parameters.
-
-% generated with alt-prior.ipynb
-Finally, we estimated the parameters under an extended informative prior (see \emph{Prior distributions}), with more informed priors for $w_{\anmt}$ and $w_{\eumt}$.
-Inference took much longer to run but the posterior distribution seemed to converge as it did not change much in the final iterations. 
-The posterior predictive plot shows that inference with this extended prior produces a posterior distribution that fails to explain the empirical observations~(Figure~\ref{fig:2n*-fixation}).
-Despite this, it produces a lower WAIC, $-57$ compared to $295$ with our standard prior, as the inferred posterior distribution is considerably narrower (compare \Cref{fig:posterior,fig:posterior-alt}), which greatly affects WAIC due to its dependence on variance (\cref{eq:WAIC}).
-The estimated mutation rate was much lower compared to the previously used prior, with $\mu=2.475\cdot10^{-9}\ [2.42\cdot10^{-9}-2.609\cdot10^{-9}]$.
-Other parameter estimates are: $\delta=2.705\cdot10^{-3}\ [2.112\cdot10^{-3}-3.107\cdot10^{-3}]$,
-$w_{\anwt}=1.022\ [1.021-1.024]$,
-$w_{\anmt}=1.029\ [1.027-1.03]$,
-$w_{\eumt}=1.03\ [1.029-1.031]$. 
-Notably, the maximum posterior ratio $w_{\eumt}/w_{\anwt}=1.007$ is much lower than the maximum prior ratio of $1.033$ (\Cref{fig:growth-curves}H) and closer to the ratio of $1$ that we assume in our standard prior.
-
-% Fig tau plots
-%% generated by diff-tau-plots.ipynb
-\begin{figure}[p!]
-  \centering
-  \begin{subfigure}{0.75\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-diff-a.pdf}      
-  \end{subfigure}
-  \begin{subfigure}{0.75\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-diff-b.pdf}      
-  \end{subfigure}
-   \begin{subfigure}{0.75\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-diff-c.pdf}      
-  \end{subfigure}
-   \caption{
-    \textbf{Genotype fixations for models with increased genetic instability.} We estimated the parameters for different models, each assuming a different value of $\tau$, the fold-increase in mutation rate in aneuploid cells.
-We then generated 10,000 simulations using the MAP estimate of each model and evaluated the fraction of simulations in which the genotype \euwt\ (\textbf{A}), \anwt\ (\textbf{B}), and \eumt\ (\textbf{C}) is fixed (y-axis) at each generation (x-axis).
-Note that \anmt\ did not fix.
-We can see that $\tau=100$ can be distinguished if time to loss of \euwt\ is known (panel~A) or if time to fixation or loss of \anwt\ is known (panel~B).
-It is harder to distinguish between $1\leq \tau \leq10$.
-  }
-  \label{fig:tau-plots}
-\end{figure}
+WAIC is lowest for $\tau=33/32$ and $\tau=1$.
+Therefore, evidence does not support an increase in mutation rate in aneuploid cells, and moreover, unless the increase is strong ($\tau \ge 10$), it does not seem to affect our inference.
+We also checked the differences in genotype frequency dynamics for different $\tau$ values.
+We observe $\tau=100$ could be distinguished if accurate data was available for the waiting time until the frequency of \euwt\ to decrease below 95\%~(\Cref{fig:tau-plots}A) or for waiting time for the frequency of \anwt\ to either reach or go below 95\%~(\Cref{fig:tau-plots}B).
 
@@
 
-\section*{Supplementary Material}
+\section*{Supplementary material}
 \beginsupplement % https://support.authorea.com/en-us/article/how-to-create-an-appendix-section-or-supplementary-information-1g25i5a/
 
-% Fig Curveball 
-\begin{figure}[h]
-	\centering
-	% generated with growth_curves.ipynb and deg = '39' 
-	\includegraphics[width=.95\textwidth]{../figures/evo39_fitness_39deg.pdf} 
-	% generated with growth_curves_refined.ipynb
-	\includegraphics[width=0.95\textwidth]{../figures/refined_vs_evo39_fitness_39deg.pdf}
-\caption{
-    \textbf{Fitness estimation from growth curves.}
-    \textbf{(A-D)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{39}{\celsius}. $ w_{\anwt}/w_{\euwt}$=1.024 (95\% CI: 0.959 - 1.115). \texttt{Curveball} 
-    \textbf{(E-H)} Fitness estimation from growth curves of \anwt\ and \eumt\ at \SI{39}{\celsius}. $ w_{\eumt}/w_{\anwt}$=1.033 (95\% CI: 1.027 - 1.041).
-    Growth curves previously described in \citet[Figs. 3C, 4A, and S2]{Yona2012}.
-	Fitness estimated from growth curves using \texttt{Curveball}, a method for predicting results of competition experiments from growth curve data~\citep[\href{https://curveball.yoavram.com}{curveball.yoavram.com}]{Ram2019}.
-See \emph{Models and Methods, Prior distributions} for more details.
-\textbf{(A,B;E,F)} Mono-culture growth curve data (markers) and best-fit growth models (lines).
-\textbf{(C,G)} The mixed-culture prediction for the strains from A,B and E,F respectively, 6,375 generated curves.
-\textbf{(D,H)} The relative fitness distribution for \anwt\ relative to \euwt\ (panel D) and \eumt\ relative to \anwt\ (panel H).
-Figures generated by \texttt{Curveball}.
-} 
-\label{fig:growth-curves}
-\end{figure}
+\subsection*{Supplementary analysis}
+\label{sec:supp_analysis}
 
-% FIG posterior alt-prior
-%% generated with alt-prior.ipynb
+\paragraph*{Sensitivity analysis.} 
+Changing a single parameter while keeping the rest fixed at the MAP estimate produces a worse fit to the data~(\Cref{fig:sensitivity}).
+Furthermore, we fitted models with a mutation rate fixed at $\mu$=$10^{-5}$, $10^{-6}$ and $10^{-7}$.
+We inferred similar parameters estimates for the model with $\mu=10^{-6}$ compared to the model with a free $\mu$ parameter, in which the inferred mutation rate is $\mu \approx 3\cdot10^{-6}$.
+Models with $\mu$=$10^{-5}$ and $\mu$=$10^{-7}$ inferred different parameters estimates, including higher aneuploidy rate and lower fitness values for \anwt\ and \eumt~(\Cref{fig:mu}).
+
+\paragraph*{Extended informative prior distribution.}
+In an extended informative prior distribution, we used additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{\eumt}/w_{\anwt}$~(\Cref{fig:growth-curves}L).
+The same distribution was used for $w_{\eumt}/w_{\anmt}$.
+Thus, our main informative prior uses a single prior distribution for fitness values of \anwt, \anmt, and \eumt, whereas the extended informative prior uses one distribution for \anwt, and another distribution for both \anmt\ and \eumt.
+
+% generated with alt-prior.ipynb
+We estimated the parameters under this extended informative prior.
+Inference took much longer to run but the posterior distribution seemed to converge, as it did not change much in the final iterations. 
+The posterior predictive plot shows that inference with this extended prior produces a posterior distribution that fails to explain the empirical observations~(pink in Figure~\ref{fig:2n*-fixation}).
+However, the inferred posterior distribution is considerably narrower (compare \Cref{fig:posterior,fig:posterior-alt}) and therefore parameter estimates are less variable.
+The estimated mutation rate was much lower compared to the main informative prior, with $\mu=2.475\cdot10^{-9}\ [2.42\cdot10^{-9}-2.609\cdot10^{-9}]$.
+Other parameter estimates are: $\delta=2.705\cdot10^{-3}\ [2.112\cdot10^{-3}-3.107\cdot10^{-3}]$,
+$w_{\anwt}=1.022\ [1.021-1.024]$,
+$w_{\anmt}=1.029\ [1.027-1.03]$,
+$w_{\eumt}=1.03\ [1.029-1.031]$. 
+Notably, the maximum posterior ratio $w_{\eumt}/w_{\anwt}=1.007$ is much lower than the maximum prior ratio of $1.033$ (\Cref{fig:growth-curves}H) and closer to the ratio of $1$ that we assume in our standard prior.
+Together with the posterior predictive results, we conclude that the main informative prior is preferable over the extended informative prior.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
+\subsection*{Supplementary figures \& tables}
+
+% Fig Validation of posteriors
+%% generated with diff-runs.ipynb
 \begin{figure}[p]
     \centering
-\includegraphics[width=0.9\textwidth]{../figures/posterior-alt.pdf}
+      \includegraphics[width=0.45\textwidth]{../figures/runs-A.pdf}      
+      \includegraphics[width=0.45\textwidth]{../figures/runs-B.pdf}    
+      \includegraphics[width=0.325\textwidth]{../figures/runs-C.pdf}      
+      \includegraphics[width=0.325\textwidth]{../figures/runs-D.pdf}      
+      \includegraphics[width=0.325\textwidth]{../figures/runs-E.pdf} 
        \caption{
-  \textbf{Posterior distribution of parameters inferred with the extended prior distribution.}
-On the diagonal, the inferred posterior distribution of each model parameter. 
-Below the diagonal, the inferred joint posterior distribution of pairs of model parameters (dark purple and bright yellow for low and high density, respectively).
-Red markers and orange lines for the joint MAP estimate (which may differ from the marginal MAP, as the marginal distribution integrates over all other parameters).
-} 
-  \label{fig:posterior-alt}
-\end{figure}
-
-% Fig likelihood profile
-%% generated by sensitivity-analysis.ipynb
-\begin{figure}[p]
-  \centering
-  \begin{subfigure}{0.3\textwidth}
-      \includegraphics[width=\textwidth]{../figures/sensitivity-A.pdf}      
-      \label{fig:sensitivity-mutation}
-  \end{subfigure}
-  \begin{subfigure}{0.3\textwidth}
-      \includegraphics[width=\textwidth]{../figures/sensitivity-B.pdf}      
-      \label{fig:sensitivity-aneuploidy}
-  \end{subfigure}
-  \\
-   \begin{subfigure}{0.3\textwidth}
-      \includegraphics[width=\textwidth]{../figures/sensitivity-C.pdf}      
-      \label{fig:sensitivity-anwt}
-  \end{subfigure}
-    \begin{subfigure}{0.3\textwidth}
-      \includegraphics[width=\textwidth]{../figures/sensitivity-D.pdf}      
-      \label{fig:sensitivity-anmt}
-  \end{subfigure}
-    \begin{subfigure}{0.3\textwidth}
-      \includegraphics[width=\textwidth]{../figures/sensitivity-E.pdf}      
-      \label{fig:sensitivity-eumt}
-  \end{subfigure}
-  \caption{
-    \textbf{Likelihood profiles.} Sensitivity of the model approximate likelihood, $\likelihood(\theta)$, to changing a single parameter while the other parameters remain fixed at their MAP estimates.
-Dashed vertical line represents the MAP value.
-The prior distributions for the mutation rate and aneuploidy rate are $\mu \sim U(10^{-9}, 10^{-5})$ and $\delta \sim U(10^{-6}, 10^{-2})$, respectively.
+    \textbf{Posterior distribution validation.}
+    The posterior distribution of model parameters is roughly the same regardless of the number of simulations (4-10,000 replicates) used to approximate the likelihood (\cref{eq:heatstress-likelihood}).
     }
 
-  \label{fig:sensitivity}
+     \label{fig:seeds}
  \end{figure}
@@
 %% generated with convergence.ipynb
-\begin{figure}[h]
+\begin{figure}[p]
 
@@
 
-% Fig Validation of posteriors
-%% generated with diff-runs.ipynb
-\begin{figure}[h]
+% Fig Curveball 
+\begin{figure}[p]
 	\centering
-      \includegraphics[width=0.45\textwidth]{../figures/runs-A.pdf}      
-      \includegraphics[width=0.45\textwidth]{../figures/runs-B.pdf}    
-      \includegraphics[width=0.325\textwidth]{../figures/runs-C.pdf}      
-      \includegraphics[width=0.325\textwidth]{../figures/runs-D.pdf}      
-      \includegraphics[width=0.325\textwidth]{../figures/runs-E.pdf} 
+	% generated with growth_curves.ipynb and deg = '39' 
+	\includegraphics[width=.95\textwidth]{../figures/evo39_fitness_39deg.pdf} 
+	% generated with growth_curves_refined.ipynb
+	\includegraphics[width=0.95\textwidth]{../figures/refined_vs_evo39_fitness_39deg.pdf}
 \caption{
-    \textbf{Posterior distribution validation.}
-    The posterior distribution of model parameters is roughly the same regardless of the number of simulations (4-10,000 replicates) used to approximate the likelihood (\cref{eq:heatstress-likelihood}).
+    \textbf{Fitness estimation from growth curves.}
+    \textbf{(A-D)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{39}{\celsius}. $ w_{\anwt}/w_{\euwt}$=1.024 (95\% CI: 0.959 - 1.115). \texttt{Curveball} 
+    \textbf{(E-H)} Fitness estimation from growth curves of \anwt\ and \eumt\ at \SI{39}{\celsius}. $ w_{\eumt}/w_{\anwt}$=1.033 (95\% CI: 1.027 - 1.041).
+    Growth curves previously described in \citet[Figs. 3C, 4A, and S2]{Yona2012}.
+	Fitness estimated from growth curves using \texttt{Curveball}, a method for predicting results of competition experiments from growth curve data~\citep[\href{https://curveball.yoavram.com}{curveball.yoavram.com}]{Ram2019}.
+See \emph{Models and Methods, Prior distributions} for more details.
+\textbf{(A,B;E,F)} Mono-culture growth curve data (markers) and best-fit growth models (lines).
+\textbf{(C,G)} The mixed-culture prediction for the strains from A,B and E,F respectively, 6,375 generated curves.
+\textbf{(D,H)} The relative fitness distribution for \anwt\ relative to \euwt\ (panel D) and \eumt\ relative to \anwt\ (panel H).
+Figures generated by \texttt{Curveball}.
 } 
-
-     \label{fig:seeds}
+\label{fig:growth-curves}
 \end{figure}
@@
 %% generated with diff-tau.ipynb
-\begin{figure}[h!]
+\begin{figure}[p]
   \centering
@@
 
+% Fig tau plots
+%% generated by diff-tau-more-plots.ipynb
+\begin{figure}[p]
+  \centering
+  \begin{subfigure}{0.75\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-diff-a.pdf}      
+  \end{subfigure}
+  \begin{subfigure}{0.75\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-diff-b.pdf}      
+  \end{subfigure}
+   \begin{subfigure}{0.75\textwidth}
+      \includegraphics[width=\textwidth]{../figures/tau-diff-c.pdf}      
+  \end{subfigure}
+   \caption{
+    \textbf{Genotype fixations for models with increased genetic instability.} We estimated the parameters for different models, each assuming a different value of $\tau$, the fold-increase in mutation rate in aneuploid cells.
+We then generated 10,000 simulations using the MAP estimate of each model and evaluated the fraction of simulations in which the frequency of genotype \euwt\ (\textbf{A}), \anwt\ (\textbf{B}), and \eumt\ (\textbf{C}) is above 95\% (y-axis) at each generation (x-axis).
+Note that \anmt\ did not fix.
+We can see that $\tau=100$ can be distinguished if the waiting time for $f_{\euwt}<95\%$ is known (panel~A) or if the waiting time for $f_{\anwt}>95\%$ or $f_{\anwt}<95\%$ is known (panel~B).
+It is harder to distinguish between $1\leq \tau \leq10$.
+  }
+  \label{fig:tau-plots}
+\end{figure}
+
+% Fig likelihood profile
+%% generated by sensitivity-analysis.ipynb
+\begin{figure}[p]
+  \centering
+  \begin{subfigure}{0.3\textwidth}
+      \includegraphics[width=\textwidth]{../figures/sensitivity-A.pdf}      
+      \label{fig:sensitivity-mutation}
+  \end{subfigure}
+  \begin{subfigure}{0.3\textwidth}
+      \includegraphics[width=\textwidth]{../figures/sensitivity-B.pdf}      
+      \label{fig:sensitivity-aneuploidy}
+  \end{subfigure}
+  \\
+   \begin{subfigure}{0.3\textwidth}
+      \includegraphics[width=\textwidth]{../figures/sensitivity-C.pdf}      
+      \label{fig:sensitivity-anwt}
+  \end{subfigure}
+    \begin{subfigure}{0.3\textwidth}
+      \includegraphics[width=\textwidth]{../figures/sensitivity-D.pdf}      
+      \label{fig:sensitivity-anmt}
+  \end{subfigure}
+    \begin{subfigure}{0.3\textwidth}
+      \includegraphics[width=\textwidth]{../figures/sensitivity-E.pdf}      
+      \label{fig:sensitivity-eumt}
+  \end{subfigure}
+  \caption{
+    \textbf{Likelihood profiles.} Sensitivity of the model approximate likelihood, $\likelihood(\theta)$, to changing a single parameter while the other parameters remain fixed at their MAP estimates.
+Dashed vertical line represents the MAP value.
+The prior distributions for the mutation rate and aneuploidy rate are $\mu \sim U(10^{-9}, 10^{-5})$ and $\delta \sim U(10^{-6}, 10^{-2})$, respectively.
+  }
+
+  \label{fig:sensitivity}
+\end{figure}
+
 % Fig mu comparisons
 %% generated with diff-mutation-rate.ipynb
-\begin{figure}[h!]
+\begin{figure}[p]
   \centering
@@
 
+% FIG posterior alt-prior
+%% generated with alt-prior.ipynb
+\begin{figure}[p]
+  \centering
+\includegraphics[width=0.9\textwidth]{../figures/posterior-alt.pdf}
+  \caption{
+  \textbf{Posterior distribution of parameters inferred with the extended prior distribution.}
+On the diagonal, the inferred posterior distribution of each model parameter. 
+Below the diagonal, the inferred joint posterior distribution of pairs of model parameters (dark purple and bright yellow for low and high density, respectively).
+Red markers and orange lines for the joint MAP estimate (which may differ from the marginal MAP, as the marginal distribution integrates over all other parameters).
+} 
+  \label{fig:posterior-alt}
+\end{figure}
+
 %% generated with WAIC.ipynb 

### 2022-05-08 ms/ms.tex
waic only for tau comparison

@@
 
-Where posterior plots are very similar, we use WAIC, or the widely applicable information criterion ~\citep{gelman2013bayesian}, defined as
+Where posterior plots are very similar and the number of parameters is the same, we use WAIC, or the widely applicable information criterion ~\citep{gelman2013bayesian}, defined as
 \begin{equation} \label{eq:WAIC}
@@
 Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\Cref{fig:ppc-plot}).
-% TODO new figure by Martin and discussion of the probability of 2n* arriving via 2n+1.
+% TODO new figure by Martin and discussion of the probability of 2n* arriving via 2n+1 - transient solution etc.
 
@@
 \caption{
-\textbf{WAIC values for various model specifications.}}
+\textbf{WAIC values for different $\tau$ values.}}
 \pgfplotstabletypeset[
@@
 \caption*{
-	WAIC (widely applicable information criterion, \cref{eq:WAIC}, ~\citep{gelman2013bayesian}) values for models with different configurations.
-WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy and a difference of 2 is a popular threshold for model comparison~\citep{Kass1995}.
+	WAIC defined in \cref{eq:WAIC}.
 } 

### 2022-05-08 ms/ms.tex
revised discussion, introduction, abstract

@@
 Importantly, aneuploidy differs from point mutations in rate, fitness effect, and reversibility.
-Here, we develop evolutionary models for adaptive evolution with both mutation and aneuploidy. 
-These models are used within an approximate Bayesian computation framework to estimate the formation rate and fitness effect of aneuploidy and mutation from  results of evolutionary experiments in which \yeast adapted to heat stress: the experimental populations first acquired chromosome duplications, only to later revert back to a euploid state.
-We also analyze our models to estimate the effect of the aneuploidy and mutation rates on the expected adaptation time and the probability for adaptation via aneuploidy.
-Our results suggest that aneuploidy can be a transient adaptive solution, which can decelerate adaptation in a non-intuitive manner.
-By creating an evolutionary conflict between the individual and the population, aneuploidy further complicates the process of adaptation in cell populations.
+Here, we develop an evolutionary model for adaptive evolution with both aneuploidy and  mutation, and apply it within an Bayesian inference framework.
+We estimate the formation rate and fitness effect of aneuploidy and mutation in an evolutionary experiment in which \yeast adapted to heat stress.
+Our results suggest that aneuploidy is more frequent than previously estimated and that experimental observations cannot be explained without aneuploidy.
+We do not find evidence for genetic instability in aneuploid cells.
+Our results support the hypothesis that aneuploidy is a transient adaptive hypothesis: the experimental populations adapted by first acquiring a chromosome duplication; then by acquiring a beneficial mutation that likely appeared in aneuploid cells, rather than euploid cells; and finally by rapidly losing the extra chromosome to avoid paying the fitness cost of aneuploidy.
 \end{abstract}
@@
 This model follows a population of cells characterised by both their ploidy and their genotype.
-We fit this model to the experimental results of \citet{Yona2012} using an \emph{approximate Bayesian computation} framework~\citep{Sisson2009} to infer model parameters, including selection coefficients and rates of aneuploidy and mutation, and to test hypotheses about the evolutionary process by performing model selection between different versions of the model.
-%We also mathematically analyze this evolutionary-genetic model to estimate the effects of selection, mutation, and aneuploidy on the adaptation time and the probability for adaptation via aneuploidy.
+We fit this model to the experimental results of \citet{Yona2012} using an \emph{approximate Bayesian computation} framework~\citep{Sisson2009} to infer model parameters, including rates and fitness effects of aneuploidy and mutation, and to test hypotheses about the evolutionary process by comparing different versions of the model and predicting the genotype frequency dynamics using the estimated parameter values.
 
@@
 
+\paragraph*{Results of current study.}
+We inferred the rates and fitness effects of aneuploidy and mutation in a previously published evolutionary experiment in which yeast populations adapted to heat stress~\citep{Yona2012}. 
+First, we estimate that the aneuploidy rate (i.e., number of chromosome gains per generation) is $1.7 \cdot 10^{-3}$, is higher than a previous estimate of $6.7 \cdot 1e0^{-6}$~\citep{Zhu2016}.
+% TODO explain?
+Second, the experiments results cannot be explained without aneuploidy ($\delta=0$) or if aneuploidy is neutral ($w_{\anwt}=w_{\euwt}$, $w_{\anmt}=w_{\eumt}$), and therefore we conclude that aneuploidy plays a significant role in the dynamics.
+Third, we find no evidence for increased mutation rates in aneuploid cells, but more data would have allowed a more conclusive statement.
+Therefore, the effect of aneuploidy on genetic instability remains to be explored in future work.
+Fourth, we simulated genotype frequency dynamics using inferred rates and fitness effects of aneuploidy and mutation.
+We found that the aneuploid mutant, \anmt, never reaches high frequency, in a process similar to \emph{stochastic tunneling}~\citep{Iwasa2004}.
+% TODO ref komarova & nowak
+Furthermore, we find that XXXX. % TODO transient solution?
+
 \paragraph*{Aneuploidy is not just another type of mutation.}
@@
 
+% TODO do we want the following two paragraphs?
 \paragraph*{Evolutionary theory of aneuploidy.}

### 2022-05-10 results/neutral-aneuploidy/pyabc-single-neutral-aneuploidy.py
started to organize notebooks and results

@@ -1,0 +1,130 @@
+import pyabc
+import scipy
+import tempfile
+import os
+os.environ["OMP_NUM_THREADS"] = "1" #for numpy
+import numpy as np
+from model1.singleLocusModel import SingleLocusModel
+import sys
+from datetime import timedelta, datetime
+from shutil import copyfile, copytree, rmtree
+from multiprocessing import  Value
+
+
+N = 6.425*10**6 # 1.6M - 2004.8M in Yona experiment 8/(1.0/(np.array([2**i for i in range(8)])*1.6)).sum()
+GLOBAL_SEED = 123
+# K = 1.03125
+# A,B,C = 50, 300, 2200
+
+PRIOR_THRESHOLD = 0.01
+MAX_POPULATIONS = 50 # stop criterion: num of abc iterations
+A,B,C = 450, 1700, 2350
+REPS1 = int(sys.argv[1])
+SIM_SEED = int(sys.argv[2]) #was 20
+K = float(sys.argv[3])
+PRIOR_TYPE = int(sys.argv[4])
+MIN_EPSILON = float(sys.argv[5]) #stop criterion
+DIR_NAME = sys.argv[6]
+CPUS = int(sys.argv[7])
+
+model1 = SingleLocusModel(k=K)
+
+class W_RV(pyabc.RVBase):
+    def __init__(self):
+        l = np.load('./data/evo39_fitness_39deg.npz')
+        self.kde = scipy.stats.gaussian_kde(l['arr_0'])
+    def rvs(self, *args, **kwargs):
+        return self.kde.resample(1)[0][0]
+    def pdf(self, x, *args, **kwargs):
+        return self.kde.pdf(x)[0]
+    def copy(self):
+        raise NotImplementedError('copy')
+    def pmf(self, x, *args, **kwargs):
+        raise NotImplementedError('pmf')
+    def cdf(self, x, *args, **kwargs):
+        raise NotImplementedError('cdf')
+
+class D_RV(pyabc.RVBase):
+    def __init__(self):
+        l = np.load('./data/refined_vs_evo39_fitness_39deg.npz')
+        self.kde = scipy.stats.gaussian_kde(l['arr_0'])
+    def rvs(self, *args, **kwargs):
+        return self.kde.resample(1)[0][0]
+    def pdf(self, x, *args, **kwargs):
+        return self.kde.pdf(x)[0]
+    def copy(self):
+        raise NotImplementedError('copy')
[…]

### 2022-05-11 results/fixedmr_7/pyabc-single-w1w1.py
more results, updated WAIC.ipynb

@@ -1,0 +1,167 @@
+import pyabc
+import scipy
+import tempfile
+import os
+os.environ["OMP_NUM_THREADS"] = "1" #for numpy
+import numpy as np
+from model1.singleLocusModel import SingleLocusModel
+import sys
+from datetime import timedelta, datetime
+from shutil import copyfile, copytree, rmtree
+from multiprocessing import  Value
+
+
+N = 6.425*10**6 # 1.6M - 2004.8M in Yona experiment 8/(1.0/(np.array([2**i for i in range(8)])*1.6)).sum()
+GLOBAL_SEED = 123
+# K = 1.03125
+# A,B,C = 50, 300, 2200
+
+PRIOR_THRESHOLD = 0.01
+MAX_POPULATIONS = 50 # stop criterion: num of abc iterations
+A,B,C = 450, 1700, 2350
+REPS1 = int(sys.argv[1])
+SIM_SEED = int(sys.argv[2]) #was 20
+K = float(sys.argv[3])
+PRIOR_TYPE = int(sys.argv[4])
+MIN_EPSILON = float(sys.argv[5]) #stop criterion
+DIR_NAME = sys.argv[6]
+CPUS = int(sys.argv[7])
+
+model1 = SingleLocusModel(k=K)
+
+class W_RV(pyabc.RVBase):
+    def __init__(self):
+        l = np.load('./data/evo39_fitness_39deg.npz')
+        self.kde = scipy.stats.gaussian_kde(l['arr_0'])
+    def rvs(self, *args, **kwargs):
+        return self.kde.resample(1)[0][0]
+    def pdf(self, x, *args, **kwargs):
+        return self.kde.pdf(x)[0]
+    def copy(self):
+        raise NotImplementedError('copy')
+    def pmf(self, x, *args, **kwargs):
+        raise NotImplementedError('pmf')
+    def cdf(self, x, *args, **kwargs):
+        raise NotImplementedError('cdf')
+
+class D_RV(pyabc.RVBase):
+    def __init__(self):
+        l = np.load('./data/refined_vs_evo39_fitness_39deg.npz')
+        self.kde = scipy.stats.gaussian_kde(l['arr_0'])
+    def rvs(self, *args, **kwargs):
+        return self.kde.resample(1)[0][0]
+    def pdf(self, x, *args, **kwargs):
+        return self.kde.pdf(x)[0]
+    def copy(self):
+        raise NotImplementedError('copy')
+    def pmf(self, x, *args, **kwargs):
[…]

### 2022-05-11 results/tau10/singleLocusModel.py

@@ -1,0 +1,189 @@
+import numpy as np
+import json
+import datetime
+import signal,sys
+import matplotlib.pyplot as plt
+
+class SingleLocusModel:
+    # k*μ is mutation rate of 2n+1 genotype 
+    def __init__(self, k, seed=20, stopon=2500, fixation=0.95):
+        self.k=k
+        self.seed = seed
+        self.stopon = stopon
+        self.fixation = fixation
+        
+    # w1 is 2n+1 fitness
+    # w2 is 2n+1* fitness
+    # w3 is 2n* fitness
+    def run_simulations(self, N, mutation_rate, aneuploidy_rate, aneuploidy_loss_rate, w1, w2, w3, repetitions=1000, seed=None):
+        curr_rand_state = np.random.get_state()
+        if seed is None:
+            seed = self.seed
+        np.random.seed(seed)
+
+        #0 wt
+        #1 tri
+        #2 tri+
+        #3 +
+        M = np.diag(np.repeat(0.,4))
+        M[0][1] = aneuploidy_rate
+        M[1][2] = self.k*mutation_rate
+        M[2][3] = aneuploidy_loss_rate 
+        M[0][3] = mutation_rate  
+        w = [1., w1, w2, w3]
+        times, p = self._simulation(N, w, M, repetitions=repetitions)
+        
+        np.random.set_state(curr_rand_state)
+        return (times,p)   
+
+    def grade_function2(self, times_p, a0=100, a=450,b=1700,c=2350):
+        fixation = 0.95
+        times , p = times_p
+        nparr = np.array(p)
+        s = nparr.shape[2] #total num of replicas
+        if nparr.shape[0]<a:
+            return 0 
+        Tidx0 = (nparr[a0-1][1]+nparr[a0-1][2])<0.95  
+        Tidx = (nparr[a-1][1]+nparr[a-1][2])>0.95  #fixated 2n+1 and 2n+1*
+        Tidx = Tidx0 &Tidx
+        Xidx = np.full(s,True) if nparr.shape[0]<b else (nparr[b-1,3,:]>0.95) 
+        Yidx = np.full(s,True) if nparr.shape[0]<c else (nparr[c-1,3,:]>0.95)
+        Yidx = ~Xidx&Yidx
+        Tnum = Tidx.sum()
+        if Tnum==0:
+            return 0
+
+        T = (Tnum/s)
+        notXgivenT = (~Xidx&Tidx).sum()/Tnum
+        notYgivenT = (~Yidx&Tidx).sum()/Tnum
[…]

### 2022-05-11 results/tau100/pyabc-single-w1w1.py

@@ -60,7 +60,6 @@ class D_RV(pyabc.RVBase):
         raise NotImplementedError('cdf')
 
 passed_count = Value('i', 0)
-total_count = Value('i', 0)
 
 class CustomDistribution1(pyabc.Distribution):
     def rvs(self):
@@ -80,32 +79,55 @@ class CustomDistribution1(pyabc.Distribution):
 
 
 def abc_model1(parameter, reps=REPS1, toprint=True):
-    global total_count
     #trisomy gain and loss are the same, therefore parameter.p2_tr twice
+    w1 = parameter.p3_w1
+    w2 = parameter.p4_w2
     w3 = parameter.p5_w3
-    x = [parameter.p1_mr, parameter.p2_tr, parameter.p2_tr, 1, w3, w3]
+    if PRIOR_TYPE==4:
+        w2 = w1*w2
+        w3 = w1*w3
+    x = [parameter.p1_mr, parameter.p2_tr, parameter.p2_tr, w1, w2, w3]
     # if toprint:
         # print(x)
-    if x[0]<0 or x[1]<0 or x[2]<0 or x[4]<1 or x[5]<1:
+    if not x[3]<x[4]<x[5] or x[0]<0 or x[1]<0 or x[2]<0 or x[3]<1 or x[4]<1 or x[5]<1:
+        # x[3]<x[4] because otherwise trisomy will not be reached.  x[3]<x[5] by Fig4A.
         return {'res':1}
 
     times_p = model1.run_simulations(N, x[0], x[1], x[2], x[3], x[4], x[5], repetitions=REPS1, seed=SIM_SEED)
     grade = 1+model1.grade_function2(times_p, a0=200, a=A, b=B, c=C)
-    with total_count.get_lock():
-        total_count.value += 1
-    print('model1',total_count.value, 1-grade, x, flush=True)
+    print('model1', 1-grade, x, flush=True)
     return {'res':grade}
 
 def distance(x, y):
     return x['res']
 
 if PRIOR_TYPE==1:
-    prior1 = CustomDistribution1(p1_mr=pyabc.RV("uniform", 10.0**-9, 10.0**-5-10.0**-9),
-                                 p2_tr=pyabc.RV("uniform", 10.0**-6, 10.0**-2-10.0**-6),
-                                 p5_w3=W_RV())
+    prior1 = CustomDistribution1(p1_mr=pyabc.RV("uniform", 10.0**-9, 10.0**-5-10.0**-9)
[…]

### 2022-06-01 ms/ms.tex
formatting
added martin's new plots and a basic discussion of these figures

@@
 This model fits the data better than the model with no aneuploidy (in which $\delta=0$), but worse than a model with positive selection for aneuploidy, in which $w_{\euwt}<w_{\anwt}<w_{\anmt}<w_{\eumt}$~(\Cref{fig:2n*-fixation}).
-Thus, our results suggest that aneuploidy plays a significant role in the evolutionary dynamics \citep{Yona2012,Yona2015}.
+Thus, our results suggest that aneuploidy significantly affects the evolutionary dynamics.
 
@@
 
-\paragraph*{Genotype frequency dynamics.}
-Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\Cref{fig:ppc-plot}).
-% TODO new figure by Martin and discussion of the probability of 2n* arriving via 2n+1 - transient solution etc.
+\paragraph*{Competition between mutation and aneuploidy.}
+
+We simulated 50 % TODO is it 50?
+ replicate genotype frequency dynamics using the MAP estimate parameters. 
+The MAP frequency dynamics (\Cref{fig:ppc-plot}A) show the four genotypes but also the frequencies of \eumt\ arose from either the \anwt\ genotype via mutation and chromosome loss events, or direct from the \euwt\ genotype via mutation.
+We observe that \anmt\ never reaches substantial frequency.
+But more importantly, we observe that the majority % TODO add number
+of \eumt\ cells are descendant of \euwt\ cells and arose via direct mutation, despite the fact that the \anwt\ genotype reaches high frequencies in the population. % TODO add number
+This finding is not unique to the MAP parameter estimate. 
+We simulated genotype frequency dynamics using 100,000 parameter samples from the posterior distribution (\Cref{fig:posterior}), and computed the average frequency of \eumt\ arising from \anwt at the end of the simulation, $F_A$, over 50 replicates.
+$F_A$, was % TODO add number
+and only in 681 (0.6\%) of the 100,000 posterior samples was it larger than 0.5
+(see \Cref{fig:ppc-plot}B) for the full histogram). 
+Overall, the frequency of \eumt\ descending from \anwt\ increases with the aneuploidy rate, $\delta$, and decreases with the mutation rate (\Cref{fig:ppc-plot}C).
+The cases in which $F_A>0.5$ can be roughly divided to two clusters.
+The first, with a low formation rate ratio ($\delta/\mu \approx 500$) and a small fitness difference between \eumt\ and \anwt\ ($w_{\eumt}/w_{\anwt}-1 \approx 0.0055$).
+In these cases, aneuploidy is not generated much faster than mutation, but this is compensated by similar fitness values that decrease the fixation probability of \eumt\ and limit the strength of clonal interference.
+The second cluster, with high formation rate ratio ($\delta/\mu \approx 10^5$) and a higher fitness difference between \eumt\ and \anwt\ ($w_{\eumt}/w_{\anwt}-1 \approx 0.007$). 
+In these cases, the larger fitness difference between \anwt\ and \eumt\ means that once \eumt\ appears it can more easily avoid extinction and compete with \anwt; this is compensated by much faster generation of \anwt\ and can therefore lead to high $F_A$ (note that we only use samples from the posterior, that is, we only compute $F_A$ for parameter choices that reasonably agree with the experimental results.)
+
+%
 
 % Fig dynamics
+\begin{figure}[h!]
+  \begin{subfigure}{0.5\textwidth}
+      \centering
       %% generated with dynamics-plot.ipynb
-\begin{figure}[h!]
+      \includegraphics[width=\textwidth]{../figures/dynamics.pdf}      
+  \end{subfigure}
+  \begin{subfigure}{0.5\textwidth}
       \centering
-\includegraphics[width=0.8\textwidth]{../figures/dynamics.pdf}
+      \includegraphics[width=\textwidth]{../figures/histo_100000.pdf}
+  \end{subfigure}
+  \\
+  \begin{subfigure}{0.33\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{../figures/scatter100000_log_star.pdf}
+  \end{subfigure}
+  \begin{subfigure}{0.33\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{../figures/scatter_dirmut_log_star_byr.pdf}
+  \end{subfigure}
+  \begin{subfigure}{0.33\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{../figures/scatter100000_rater_fitr.pdf}
+  \end{subfigure}
   \caption{
-  \textbf{Posterior genotype frequency dynamics for the model.} 
-  The posterior prediction for the frequencies of the four genotypes over time.
+    \textbf{Posterior genotype frequency dynamics.} 
+    \textbf{(A)} The posterior prediction for the frequencies of the four genotypes over time.
 Each of the 20 curves is the average of 10,000 simulations of the model using parameters drawn from the posterior distribution.
@@
 
-\pagebreak
+%\pagebreak
 % Acknowledgements
@@
 
-\section*{Supplementary material}
+\section*{Supplementary Material}
 \beginsupplement % https://support.authorea.com/en-us/article/how-to-create-an-appendix-section-or-supplementary-information-1g25i5a/
 
-\subsection*{Supplementary analysis}
+\subsection*{Supplementary Analysis}
 \label{sec:supp_analysis}
@@
 \newpage
-\subsection*{Supplementary figures \& tables}
+\subsection*{Supplementary Figures \& Tables}
 
@@
 %% generated with diff-runs.ipynb
-\begin{figure}[p]
+\begin{figure}[h]
     \centering

### 2022-06-01 ms/ms.tex
revised new results section

--- 2022-05-08 ms/ms.tex
@@
 
-\paragraph*{Genotype frequency dynamics.}
-Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\Cref{fig:ppc-plot}).
-% TODO new figure by Martin and discussion of the probability of 2n* arriving via 2n+1 - transient solution etc.
+\paragraph*{Competition between mutation and aneuploidy.}
+
+We simulated 50 % TODO is it 50?
+ replicate genotype frequency dynamics using the MAP estimate parameters. 
+\Cref{fig:ppc-plot}A shows the simulated frequencies of the four genotypes (\euwt, \anwt, \anmt\ and \eumt), as well as the frequencies of \eumt\ cells that arose from either \anwt\ cells via a sequences of mutation and chromosome loss events, or directly from \euwt\ cells via a mutation event.
+We find that \anmt\ never reaches substantial frequency as it is quickly replaced by \eumt.
+
+\citet{Yona2012} have suggested that aneuploidy is a transient adaptive solution, arising quickly due to increased fitness and high formation rate, but 
+ultimately lost as beneficial mutations appear and the extra chromosome is lost due to the cost of aneuploidy. 
+We tested this scenario by estimating $F_A$, the expected frequency of \eumt\ that arose from \anwt\ (as the average frequency of such cells at the end of 50 replicate simulations).
+Surprisingly, we observe that the majority ($F_A=XX$) % TODO add number
+of \eumt\ cells are a product of a direct mutation in \euwt\ cells, rather than descending from \anwt\ cells. 
+This is despite the fact that the \anwt\ genotype reaches high frequencies in the population (\Cref{fig:ppc-plot}A). % TODO add number
+This result is not unique to the MAP parameter estimate. 
+We simulated genotype frequency dynamics using parameter samples from the posterior distribution (\Cref{fig:posterior}), and computed the posterior distribution of $F_A$ (\Cref{fig:ppc-plot}B). 
+The mean $F_A$ was just XX [XX-XX 95\% CI], % TODO add number
+such that if we pick up a cell from the evolved population it is more likely to have descended directly from a euploid cell than from an aneuploid cell.
+As expected, the probability of \eumt\ descending from \anwt\ ($F_A$) increases with the aneuploidy rate, $\delta$, and decreases with the mutation rate, $\mu$ (\Cref{fig:ppc-plot}C and D).
+
+Only in 681 of 100,000 posterior samples (0.6\%) $F_A$ was  larger than 0.5, that is, \eumt\ was more likely to occur via aneuploidy lineage.  
+These parameter samples can be roughly divided to two clusters (\Cref{fig:ppc-plot}C and E).
+In the first, aneuploidy formation is not much faster than mutation ($\delta/\mu \approx 500$), but this is compensated by a small fitness difference between \eumt\ and \anwt\ ($w_{\eumt}/w_{\anwt}-1 \approx 0.0055$), which results in a low fixation probability for \eumt\ once \anwt\ is at a high frequency, and limits the strength of clonal interference.
+In the second cluster, aneuploidy formation is much faster than mutation ($\delta/\mu \approx 10^5$), but the fitness difference between \eumt\ and \anwt\ is also larger ($w_{\eumt}/w_{\anwt}-1 \approx 0.007$).
+Thus, once \eumt\ appears it more easily avoids extinction and competes with \anwt.
+This is compensated by the high aneuploidy rate (note that we only use samples from the posterior, that is, we only compute $F_A$ for parameter choices that reasonably agree with the experimental results.)
 
 % Fig dynamics
+\begin{figure}[h!]
+  \begin{subfigure}{0.5\textwidth}
+      \centering
       %% generated with dynamics-plot.ipynb
-\begin{figure}[h!]
+      \includegraphics[width=\textwidth]{../figures/dynamics.pdf}      
+  \end{subfigure}
+  \begin{subfigure}{0.5\textwidth}
       \centering
-\includegraphics[width=0.8\textwidth]{../figures/dynamics.pdf}
+      \includegraphics[width=\textwidth]{../figures/histo_100000.pdf}
+  \end{subfigure}
+  \\
+  \begin{subfigure}{0.33\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{../figures/scatter100000_log_star.png}
+  \end{subfigure}
+  \begin{subfigure}{0.33\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{../figures/scatter_dirmut_log_star_byr.png}
+  \end{subfigure}
+  \begin{subfigure}{0.33\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{../figures/scatter100000_rater_fitr.png}
+  \end{subfigure}
   \caption{
-  \textbf{Posterior genotype frequency dynamics for the model.} 
-  The posterior prediction for the frequencies of the four genotypes over time.
-Each of the 20 curves is the average of 10,000 simulations of the model using parameters drawn from the posterior distribution.
+    \textbf{Posterior genotype frequency dynamics.}     
+    \textbf{(A)} The posterior prediction for the frequencies of the four genotypes over time, including the source of \eumt.
+Each of the curves is the average of 50 simulations  using parameters drawn from the posterior distribution.
+% TODO check if true
+    \textbf{(B)} Marginal posterior distribution of $F_A$, the expected frequency of \eumt\ cells that arose from \anwt cells, computed as the average frequency at the end of 50 simulations for each of 100,000 samples from the parameter posterior distribution.
+Solid and dashed lines show the mean and 95\% CI.
+	\textbf{(C-E)} $F_A$ values (color coded) from panel B, with their corresponding parameter values on the x- and y-axes. White start shows the MAP estimate. Panel D only shows $F_A<0.5$.
      }

### 2022-06-02 ms/ms.tex
revised FA histogram and added FA values

@@
 We simulated genotype frequency dynamics using parameter samples from the posterior distribution (\Cref{fig:posterior}), and computed the posterior distribution of $F_A$ (\Cref{fig:ppc-plot}B). 
-The mean $F_A$ was just XX [XX-XX 95\% CI], % TODO add number
+The mean $F_A$ was just 0.1758 [0.0148-0.4114 95\% CI],
 such that if we pick up a cell from the evolved population it is more likely to have descended directly from a euploid cell than from an aneuploid cell.

### 2022-06-02 ms/ms.tex
revised FA results after fixing a bug

--- 2022-06-01 ms/ms.tex
@@
 
-We simulated 50 % TODO is it 50?
- replicate genotype frequency dynamics using the MAP estimate parameters. 
-The MAP frequency dynamics (\Cref{fig:ppc-plot}A) show the four genotypes but also the frequencies of \eumt\ arose from either the \anwt\ genotype via mutation and chromosome loss events, or direct from the \euwt\ genotype via mutation.
-We observe that \anmt\ never reaches substantial frequency.
-But more importantly, we observe that the majority % TODO add number
-of \eumt\ cells are descendant of \euwt\ cells and arose via direct mutation, despite the fact that the \anwt\ genotype reaches high frequencies in the population. % TODO add number
-This finding is not unique to the MAP parameter estimate. 
-We simulated genotype frequency dynamics using 100,000 parameter samples from the posterior distribution (\Cref{fig:posterior}), and computed the average frequency of \eumt\ arising from \anwt at the end of the simulation, $F_A$, over 50 replicates.
-$F_A$, was % TODO add number
-and only in 681 (0.6\%) of the 100,000 posterior samples was it larger than 0.5
-(see \Cref{fig:ppc-plot}B) for the full histogram). 
-Overall, the frequency of \eumt\ descending from \anwt\ increases with the aneuploidy rate, $\delta$, and decreases with the mutation rate (\Cref{fig:ppc-plot}C).
-The cases in which $F_A>0.5$ can be roughly divided to two clusters.
-The first, with a low formation rate ratio ($\delta/\mu \approx 500$) and a small fitness difference between \eumt\ and \anwt\ ($w_{\eumt}/w_{\anwt}-1 \approx 0.0055$).
-In these cases, aneuploidy is not generated much faster than mutation, but this is compensated by similar fitness values that decrease the fixation probability of \eumt\ and limit the strength of clonal interference.
-The second cluster, with high formation rate ratio ($\delta/\mu \approx 10^5$) and a higher fitness difference between \eumt\ and \anwt\ ($w_{\eumt}/w_{\anwt}-1 \approx 0.007$). 
-In these cases, the larger fitness difference between \anwt\ and \eumt\ means that once \eumt\ appears it can more easily avoid extinction and compete with \anwt; this is compensated by much faster generation of \anwt\ and can therefore lead to high $F_A$ (note that we only use samples from the posterior, that is, we only compute $F_A$ for parameter choices that reasonably agree with the experimental results.)
+We simulated 50 replicate genotype frequency dynamics using the MAP estimate parameters. 
+\Cref{fig:ppc-plot}A shows the simulated frequencies of the four genotypes (\euwt, \anwt, \anmt\ and \eumt), as well as the frequencies of \eumt\ cells that arose from either \anwt\ cells via a sequences of mutation and chromosome loss events ($2n^*_A$), or directly from \euwt\ cells via a mutation event ($2n^*_M$).
+We find that \anmt\ never reaches substantial frequency as it is quickly replaced by \eumt.
 
-%
+\citet{Yona2012} have suggested that aneuploidy is a transient adaptive solution, arising quickly due to increased fitness and high formation rate, but 
+ultimately lost due to the cost of aneuploidy as beneficial mutations take over.
+We tested this hypothesis by estimating $F_A$, the expected frequency of \eumt\ that arose from \anwt, as the average frequency of such cells at the end of simulations using the MAP estimate parameters.
+Surprisingly, we observe that the majority of \eumt\ cells are a product of a direct mutation in \euwt\ cells, rather than descending from \anwt\ cells ($F_A^{MAP} = 0.106$, \Cref{fig:ppc-plot}A). 
+This is despite the fact that the \anwt\ genotype reaches high frequencies in the population (at least 0.98, \Cref{fig:ppc-plot}A).
+
+This result is not unique to the MAP parameter estimate. 
+We simulated genotype frequency dynamics using parameter samples from the posterior distribution (\Cref{fig:posterior}), and computed the posterior distribution of $F_A$ (\Cref{fig:ppc-plot}B). 
+The mean $F_A$ was just 0.1675 [0.0142 - 0.3758 95\% CI] and only in 408 of 86,349 posterior samples (0.47\%) $F_A$ was  larger than 0.5, that is, \eumt\ was more likely to arise from an aneuploid cell.
+Thus, if we pick up a cell from the evolved \eumt\ population, it is more likely to have descended directly from an euploid cell than from an aneuploid cell.
+The probability of \eumt\ descending from \anwt\ ($F_A$) increases with the aneuploidy rate, $\delta$, and decreases with the mutation rate, $\mu$, and in some cases can also be affected by the fitness parameters (\Cref{fig:ppc-plot}C and \Cref{fig:fig:posterior_FA_supp}).
 
@@
 \begin{figure}[h!]
-  \begin{subfigure}{0.5\textwidth}
+  \begin{subfigure}{\textwidth}
       \centering
       %% generated with dynamics-plot.ipynb
-      \includegraphics[width=\textwidth]{../figures/dynamics.pdf}      
+      \includegraphics[width=7in,height=2.5in]{../figures/dynamic_MAP.pdf}      
   \end{subfigure}
@@
   \end{subfigure}
-  \\
-  \begin{subfigure}{0.33\textwidth}
+  \begin{subfigure}{0.5\textwidth}
       \centering
@@
   \end{subfigure}
-  \begin{subfigure}{0.33\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../figures/scatter_dirmut_log_star_byr.pdf}
-  \end{subfigure}
-  \begin{subfigure}{0.33\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../figures/scatter100000_rater_fitr.pdf}
-  \end{subfigure}
   \caption{
     \textbf{Posterior genotype frequency dynamics.}     
-    \textbf{(A)} The posterior prediction for the frequencies of the four genotypes over time.
-Each of the 20 curves is the average of 10,000 simulations of the model using parameters drawn from the posterior distribution.
+    \textbf{(A)} Posterior predicted genotype frequencies over time, including the source of \eumt. Shown are 50 simulations using the MAP estimate parameters.
+    \textbf{(B)} Marginal posterior distribution of $F_A$, the expected frequency of \eumt\ cells that arose from \anwt cells, computed as the average frequency at the end of 50 simulations for 86,349 samples from the parameter posterior distribution.
+Solid and dashed lines show the mean and 95\% CI.
+	\textbf{(C)} $F_A$ values (color coded) from panel B, with their corresponding mutation rate $\mu$ on x-axis and aneuploidy rate $\delta$ on the y-axisWhite star shows the MAP estimate.
+See also \Cref{fig:posterior_FA_supp}.
      }
@@
 
-\pagebreak
+%\pagebreak
 
@@
 
+% FIG posterior F_A supplementary
+\begin{figure}[p]
+	\begin{subfigure}{0.5\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{../figures/scatter100000_rater_fitr_from_wt.pdf}      
+	\end{subfigure}
+	  \begin{subfigure}{0.5\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{../figures/scatter100000_rater_fitr_mut.pdf}
+  \end{subfigure}
+    \begin{subfigure}{0.5\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{../figures/scatter100000_rater_means_log.pdf}
+    \end{subfigure}    
+  \begin{subfigure}{0.5\textwidth}
+      \centering
+      \includegraphics[width=\textwidth]{../figures/scatter100000_rater_means.pdf}
+  \end{subfigure}
+	\caption{\textbf{Posterior distribution of $F_A$.}
+	\textbf{(A,B)} $F_A$ values (color coded) as in \Cref{fig:ppc-plot} for different parameter choices on the x- and y-axes. White start denotes the MAP estimate.
+	\textbf{(C,D)} $F_A$ values (y-axis) as a function of $\delta/\mu$, the aneuploidy-to-mutation rate ratio.
+	}
+    \label{fig:posterior_FA_supp}
+\end{figure}
+
 %% generated with WAIC.ipynb 

### 2022-06-07 notebooks_mp/singlelocusmodelextended.py
plots update

@@ -107,7 +107,7 @@ class SingleLocusModelExt:
         #return T, P, nona
     
     # simulations only with waiting time as return
-    def run_simulations_time(self, N, mutation_rate, aneuploidy_rate, aneuploidy_loss_rate, w1, w2, w3, repetitions=1000, max_gen=2500, fixation=0.95, seed=5,clonal_intf=True):
+    def run_simulations_time(self, N, mutation_rate, aneuploidy_rate, aneuploidy_loss_rate, w1, w2, w3, repetitions=1000, max_gen=2500, fixation=0.95, seed=5,clonal_intf=True, fix_frequ=False):
         curr_rand_state = np.random.get_state()
         np.random.seed(seed)
 
@@ -121,14 +121,17 @@ class SingleLocusModelExt:
         M[2][3] = aneuploidy_loss_rate 
         M[0][4] = self.k*mutation_rate  
         w = [1., w1, w2, w3, w3]
-        times, non_aneu_fix= self._simulation_time(N, w, M, repetitions=repetitions, fixation=fixation, max_gen=max_gen,clonal_intf=clonal_intf)
+        times, non_aneu_fix, P= self._simulation_time(N, w, M, repetitions=repetitions, fixation=fixation, max_gen=max_gen,clonal_intf=clonal_intf, fix_frequ=fix_frequ)
         
         np.random.set_state(curr_rand_state)
-        return (times, non_aneu_fix)
+        if fix_frequ is True:
+            return (times, non_aneu_fix, P)
+        else:
+            return (times, non_aneu_fix)
 
     
     #have side effect on M, but not harmfull, can run twice simulation
-    def _simulation_time(self, N, w, M ,repetitions=1000, fixation=0.95, max_gen=2500,clonal_intf=True):
+    def _simulation_time(self, N, w, M ,repetitions=1000, fixation=0.95, max_gen=2500,clonal_intf=True, fix_frequ=False):
         assert N > 0
         N = np.uint64(N)
         L = len(w)
@@ -198,11 +201,12 @@ class SingleLocusModelExt:
             
         #last generation update
         p = n/N 
-        #P.append(p)
+        if fix_frequ is True:
+            P.append(p[3:5])
         nona=(~directmut).nonzero()[0]
         combfix=(~combfix).nonzero()[0]
 
-        return T, (nona,combfix)
+        return T, (nona,combfix), P
         #return T, nona
     
         
[…]

### 2022-06-07 ms/ms.tex
posterior and hdi with truncated prior

@@
 %% estimates generated by MAP-and-hdi.ipynb
-The estimated mutation rate, $\mu=2.942\cdot10^{-6}\ [2.017\cdot10^{-7}-4.15\cdot10^{-6}]$, corresponds to a mutation target size of $\sim 10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
-The estimated aneuploidy rate, $\delta=1.722\cdot10^{-3}\ [1.394\cdot10^{-3}-2.754\cdot10^{-3}]$ is much higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
+The estimated mutation rate, $\mu=2.986\cdot10^{-6}\ [3.734\cdot10^{-7}-3.725\cdot10^{-6}]$, corresponds to a mutation target size of $\sim 10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
+The estimated aneuploidy rate, $\delta=1.724\cdot10^{-3}\ [1.451\cdot10^{-3}-2.766\cdot10^{-3}]$ is much higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
 The estimated fitness values are $w_{\anwt}=1.022\ [1.021-1.023]$,
@@
 %% generated with MAP-and-hdi.ipynb
-The estimated mutation rate without aneuploidy is $\mu=7.979\cdot10^{-9}\ [7.903\cdot 10^{-9}-8.135\cdot10^{-9}]$, much lower compared to a model with aneuploidy and suggesting a target size of just 40.
+The estimated mutation rate without aneuploidy is $\mu=7.98\cdot10^{-9}\ [7.91\cdot10^{-9}-8.14\cdot10^{-9}]$, much lower compared to a model with aneuploidy and suggesting a target size of just 40. 
 The fitness of the mutant is also much lower at $w_{\eumt}=1.013\ [1.012-1.013]$.
@@
 %% generated with MAP-and-hdi.ipynb
-With $\tau=100$, the estimated mutation rate was about 7-8-fold lower compared to $\tau=1$ ($\mu=3.81\cdot10^{-7}\ [1.508\cdot10^{-7}-4.995\cdot10^{-7}]$) and the aneuploidy rate was about 2-3-fold lower ($\delta=0.782\cdot10^{-3}\ [0.661\cdot10^{-3}-1.462\cdot10^{-3}]$).
-With $\tau=10$, the estimated mutation rate was only slightly lower compared to $\tau=1$ ($\mu=1.674\cdot10^{-6}\ [2.501\cdot10^{-7}-1.741\cdot10^{-6}]$). 
+With $\tau=100$, the estimated mutation rate was about 7-8-fold lower compared to $\tau=1$ ($\mu=3.84\cdot10^{-7}\ [6.53\cdot10^{-8}-6.207\cdot10^{-7}]$) and the aneuploidy rate was about 2-3-fold lower ($\delta=0.764\cdot10^{-3}\ [0.475\cdot10^{-3}-1.818\cdot10^{-3}]$).
+With $\tau=10$, the estimated mutation rate was only slightly lower compared to $\tau=1$ ($\mu=1.66\cdot10^{-6}\ [9.033\cdot10^{-8}-2.353\cdot10^{-6}]$). 
 WAIC is lowest for $\tau=33/32$ and $\tau=1$.

### 2022-06-07 ms/ms.tex
update FA results

--- 2022-06-02 ms/ms.tex
@@
 Thus, if we pick up a cell from the evolved \eumt\ population, it is more likely to have descended directly from an euploid cell than from an aneuploid cell.
-The probability of \eumt\ descending from \anwt\ ($F_A$) increases with the aneuploidy rate, $\delta$, and decreases with the mutation rate, $\mu$, and in some cases can also be affected by the fitness parameters (\Cref{fig:ppc-plot}C and \Cref{fig:fig:posterior_FA_supp}).
+The probability of \eumt\ descending from \anwt\ ($F_A$) increases with the aneuploidy rate, $\delta$, and decreases with the mutation rate, $\mu$, and in some cases can also be affected by the fitness parameters (\Cref{fig:ppc-plot}C and \Cref{fig:posterior_FA_supp}).
 
@@
       \centering
-      \includegraphics[width=\textwidth]{../figures/scatter100000_rater_means.pdf}
+      \includegraphics[width=\textwidth]{../figures/plot_rateratios_means.pdf}
   \end{subfigure}

### 2022-06-10 ms/ms.tex
updated truncated prior for kde of hdi and posterior plots

--- 2022-06-07 ms/ms.tex
@@
 %% estimates generated by MAP-and-hdi.ipynb
-The estimated mutation rate, $\mu=2.986\cdot10^{-6}\ [3.734\cdot10^{-7}-3.725\cdot10^{-6}]$, corresponds to a mutation target size of $\sim 10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
-The estimated aneuploidy rate, $\delta=1.724\cdot10^{-3}\ [1.451\cdot10^{-3}-2.766\cdot10^{-3}]$ is much higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
+The estimated mutation rate, $\mu=2.965\cdot10^{-6}\ [2.718\cdot10^{-7}-3.589\cdot10^{-6}]$, corresponds to a mutation target size of $\sim 10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
+The estimated aneuploidy rate, $\delta=1.72\cdot10^{-3}\ [1.47\cdot10^{-3}-2.786\cdot10^{-3}]$ is much higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
 The estimated fitness values are $w_{\anwt}=1.022\ [1.021-1.023]$,
@@
 %% generated with MAP-and-hdi.ipynb
-The estimated mutation rate without aneuploidy is $\mu=7.98\cdot10^{-9}\ [7.91\cdot10^{-9}-8.14\cdot10^{-9}]$, much lower compared to a model with aneuploidy and suggesting a target size of just 40. 
+The estimated mutation rate without aneuploidy is $\mu=7.98\cdot10^{-9}\ [7.906\cdot10^{-9}-8.138\cdot10^{-9}]$, much lower compared to a model with aneuploidy and suggesting a target size of just 40.
 The fitness of the mutant is also much lower at $w_{\eumt}=1.013\ [1.012-1.013]$.
@@
 %% generated with MAP-and-hdi.ipynb
-With $\tau=100$, the estimated mutation rate was about 7-8-fold lower compared to $\tau=1$ ($\mu=3.84\cdot10^{-7}\ [6.53\cdot10^{-8}-6.207\cdot10^{-7}]$) and the aneuploidy rate was about 2-3-fold lower ($\delta=0.764\cdot10^{-3}\ [0.475\cdot10^{-3}-1.818\cdot10^{-3}]$).
-With $\tau=10$, the estimated mutation rate was only slightly lower compared to $\tau=1$ ($\mu=1.66\cdot10^{-6}\ [9.033\cdot10^{-8}-2.353\cdot10^{-6}]$). 
+With $\tau=100$, the estimated mutation rate was about 7-8-fold lower compared to $\tau=1$ ($\mu=4.094\cdot10^{-7}\ [6.252\cdot10^{-8}-6.046\cdot10^{-7}]$) and the aneuploidy rate was about 2-3-fold lower ($\delta=0.744\cdot10^{-3}\ [0.506\cdot10^{-3}-1.827\cdot10^{-3}]$).
+With $\tau=10$, the estimated mutation rate was only slightly lower compared to $\tau=1$ ($\mu=1.67\cdot10^{-6}\ [2.836\cdot10^{-8}-2.245\cdot10^{-6}]$). 
 WAIC is lowest for $\tau=33/32$ and $\tau=1$.
@@
 However, the inferred posterior distribution is considerably narrower (compare \Cref{fig:posterior,fig:posterior-alt}) and therefore parameter estimates are less variable.
-The estimated mutation rate was much lower compared to the main informative prior, with $\mu=2.475\cdot10^{-9}\ [2.42\cdot10^{-9}-2.609\cdot10^{-9}]$.
-Other parameter estimates are: $\delta=2.705\cdot10^{-3}\ [2.112\cdot10^{-3}-3.107\cdot10^{-3}]$,
+The estimated mutation rate was much lower compared to the main informative prior, 
+% generated with MAP-and-hdi.ipynb
+with $\mu=2.475\cdot10^{-9}\ [2.419\cdot10^{-9}-2.609\cdot10^{-9}]$. Other parameter estimates are: $\delta=2.707\cdot10^{-3}\ [2.093\cdot10^{-3}-3.092\cdot10^{-3}]$,
 $w_{\anwt}=1.022\ [1.021-1.024]$,

### 2022-06-19 ms/ms.tex
updated plots with truncated kde and organized results files and notebooks

@@
 
-After producing a set of weighted particles from the the posterior distribution using the above ABC-SMC algorithm, we approximate the posterior using kernel density estimation (KDE) with Gaussian kernels, from which we find the MAP (maximum a posteriori) estimate as the maximum of the KDE function.
-We then draw $50,000$ samples from the posterior KDE to compute the HDI (highest density interval) and visualize the posterior distribution with histograms.
+After producing a set of weighted particles from the the posterior distribution using the above ABC-SMC algorithm, we approximate the posterior using kernel density estimation (KDE) with Gaussian kernels.
+Since KDE can have positive posterior probability for values with the prior zero, we truncate KDE, giving zero probability for the ranges outside of the prior.
+From maximum of the KDE function we get the MAP (maximum a posteriori) estimate, and then draw $5,000,000$ samples from the posterior KDE to compute the HDI (highest density interval) and draw $50,000$ samples from it to visualize the posterior distribution with histograms.
 

### 2022-06-22 ms/ms.tex
update affil

@@
 
-\author[1,*]{Ilia Kohanovski}
-\author[2,*]{Martin Pontz}
+\author[1,2,*]{Ilia Kohanovski}
+\author[1,*]{Martin Pontz}
 \author[3]{Avihu H. Yona}
-\author[1,2,$\dagger$]{Yoav Ram}
+\author[1,$\dagger$]{Yoav Ram}
 
-\affil[1]{School of Computer Science, Reichman University, Herzliya, Israel}
-\affil[2]{School of Zoology, Faculty of Life Sciences, Tel Aviv University, Tel Aviv, Israel}
+\affil[1]{School of Zoology, Faculty of Life Sciences, Tel Aviv University, Tel Aviv, Israel}
+\affil[2]{School of Computer Science, Reichman University, Herzliya, Israel}
 \affil[3]{Institute of Biochemistry, Food Science and Nutrition,

### 2022-06-27 ms/ms.tex

@@
 After producing a set of weighted particles from the the posterior distribution using the above ABC-SMC algorithm, we approximate the posterior using kernel density estimation (KDE) with Gaussian kernels.
-Since KDE can have positive posterior probability for values with the prior zero, we truncate KDE, giving zero probability for the ranges outside of the prior.
-From maximum of the KDE function we get the MAP (maximum a posteriori) estimate, and then draw $5,000,000$ samples from the posterior KDE to compute the HDI (highest density interval) and draw $50,000$ samples from it to visualize the posterior distribution with histograms.
+We truncate the estimated posterior to avoid positive posterior density for values with zero prior density.
+The MAP (maximum a posteriori) estimate is computed as the the maximum of the estimated joint posterior density.
+We then draw $5,000,000$ samples from the posterior distribution to compute the HDI (highest density interval) and draw $50,000$ samples to visualize the posterior distribution with histograms.
 
@@
   \textbf{Posterior distribution of model parameters.}
-On the diagonal, the inferred posterior distribution of each model parameter. 
-Below the diagonal, the inferred joint posterior distribution of pairs of model parameters (dark purple and bright yellow for low and high density, respectively).
+On the diagonal, the marginal posterior distribution of each model parameter. 
+Below the diagonal, the joint posterior distribution of pairs of model parameters (dark purple and bright yellow for low and high density, respectively).
 Red markers and orange lines for the joint MAP estimate (which may differ from the marginal MAP, as the marginal distribution integrates over all other parameters).
@@
 Surprisingly, we observe that the majority of \eumt\ cells are a product of a direct mutation in \euwt\ cells, rather than descending from \anwt\ cells ($F_A^{MAP} = 0.106$, \Cref{fig:ppc-plot}A).
+% TODO update numbers
 This is despite the fact that the \anwt\ genotype reaches high frequencies in the population (at least 0.98, \Cref{fig:ppc-plot}A).
@@
       %% generated with dynamics-plot.ipynb
-      \includegraphics[width=7in,height=2.5in]{../figures/dynamic_MAP.pdf}      
+      \includegraphics[width=7in,height=2.5in]{../figures/full_dynamic_MAP.pdf}      
   \end{subfigure}
@@
     \textbf{Posterior genotype frequency dynamics.}     
-    \textbf{(A)} Posterior predicted genotype frequencies over time, including the source of \eumt. Shown are 50 simulations using the MAP estimate parameters.
-    \textbf{(B)} Marginal posterior distribution of $F_A$, the expected frequency of \eumt\ cells that arose from \anwt cells, computed as the average frequency at the end of 50 simulations for 86,349 samples from the parameter posterior distribution.
+    \textbf{(A)} Posterior predicted genotype frequencies over time, including the source of \eumt: $2n^*_A$ arose from $2n+1$, where as $2n^*_M$ arose directly from $2n$.
+Shown are 50 simulations using the MAP estimate parameters.
+    \textbf{(B)} Posterior distribution of $F_A$, the expected frequency of \eumt\ cells that arose from \anwt cells, computed as the average frequency at the end of 50 simulations for 86,349 samples from the parameter posterior distribution.
     Solid and dashed lines show the mean and 95\% CI.
-	\textbf{(C)} $F_A$ values (color coded) from panel B, with their corresponding mutation rate $\mu$ on x-axis and aneuploidy rate $\delta$ on the y-axisWhite star shows the MAP estimate.
-See also \Cref{fig:posterior_FA_supp}.
-     }
+	\textbf{(C)} $F_A$ values (color coded) from panel B, with their corresponding mutation rate $\mu$ on x-axis and aneuploidy rate $\delta$ on the y-axis.
+	White star shows the MAP estimate. See also \Cref{fig:posterior_FA_supp}.
+     } % TODO update numbers
   \label{fig:ppc-plot}

### 2022-06-27 ms/ms.tex

@@
 Surprisingly, we observe that the majority of \eumt\ cells are a product of a direct mutation in \euwt\ cells, rather than descending from \anwt\ cells ($F_A^{MAP} = 0.106$, \Cref{fig:ppc-plot}A). 
-% TODO update numbers
 This is despite the fact that the \anwt\ genotype reaches high frequencies in the population (at least 0.98, \Cref{fig:ppc-plot}A).
@@
 We simulated genotype frequency dynamics using parameter samples from the posterior distribution (\Cref{fig:posterior}), and computed the posterior distribution of $F_A$ (\Cref{fig:ppc-plot}B). 
-The mean $F_A$ was just 0.1675 [0.0142 - 0.3758 95\% CI] and only in 408 of 86,349 posterior samples (0.47\%) $F_A$ was  larger than 0.5, that is, \eumt\ was more likely to arise from an aneuploid cell.
+The mean $F_A$ was just 0.1673 [0.0154-0.370 95\% CI] and only in 489 of 100,000 posterior samples (0.489\%) $F_A$ was  larger than 0.5, that is, \eumt\ was more likely to arise from an aneuploid cell.
 Thus, if we pick up a cell from the evolved \eumt\ population, it is more likely to have descended directly from an euploid cell than from an aneuploid cell.
@@
     \textbf{Posterior genotype frequency dynamics.}     
-    \textbf{(A)} Posterior predicted genotype frequencies over time, including the source of \eumt: $2n^*_A$ arose from $2n+1$, where as $2n^*_M$ arose directly from $2n$.
+    \textbf{(A)} Posterior predicted genotype frequencies over time, including the source of \eumt: $2n^*_A$ arose from $2n+1$, whereas $2n^*_M$ arose directly from $2n$.
 Shown are 50 simulations using the MAP estimate parameters.
-    \textbf{(B)} Posterior distribution of $F_A$, the expected frequency of \eumt\ cells that arose from \anwt cells, computed as the average frequency at the end of 50 simulations for 86,349 samples from the parameter posterior distribution.
+    \textbf{(B)} Posterior distribution of $F_A$, the expected frequency of \eumt\ cells descended from \anwt\ cells, computed as the average frequency at the end of 100 simulations for 100,000 samples from the parameter posterior distribution.
     Solid and dashed lines show the mean and 95\% CI.
@@
 	White star shows the MAP estimate. See also \Cref{fig:posterior_FA_supp}.
-     } % TODO update numbers
+     }
   \label{fig:ppc-plot}
@@
 We inferred the rates and fitness effects of aneuploidy and mutation in a previously published evolutionary experiment in which yeast populations adapted to heat stress~\citep{Yona2012}. 
-First, we estimate that the aneuploidy rate (i.e., number of chromosome gains per generation) is $1.7 \cdot 10^{-3}$, is higher than a previous estimate of $6.7 \cdot 1e0^{-6}$~\citep{Zhu2016}.
+First, we estimate that the aneuploidy rate (i.e., number of chromosome gains per generation) is $1.7 \cdot 10^{-3}$, is higher than a previous estimate of $6.7 \cdot 10^{-6}$~\citep{Zhu2016}.
 % TODO explain?
@@
 \section*{Acknowledgements}
-We thank Yitzhak Pilpel, Orna Dahan, Lilach Hadany, Judith Berman, David Gresham, Shay Covo, Martin Kupiec, and Tal Simon for discussions and comments.
+We thank Yitzhak Pilpel, Orna Dahan, Lilach Hadany, Judith Berman, David Gresham, Shay Covo, Martin Kupiec, Uri Obolski, and Tal Simon for discussions and comments.
 This work was supported in part by 
-the Israel Science Foundation (YR 552/19) and
-Minerva Stiftung Center for Lab Evolution (YR).
-% TODO add Minerva for Martin; funding for Avihu?
+the Israel Science Foundation (YR 552/19),
+Minerva Stiftung Center for Lab Evolution (YR),
+Minerva Stiftung short-term research grant (MP).
+% TODO funding for Avihu
 }
@@
   \end{subfigure}
-    \begin{subfigure}{0.5\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../figures/scatter100000_rater_means_log.pdf}
-    \end{subfigure}    
-  \begin{subfigure}{0.5\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../figures/plot_rateratios_means.pdf}
-  \end{subfigure}
 	\caption{\textbf{Posterior distribution of $F_A$.}
-	\textbf{(A,B)} $F_A$ values (color coded) as in \Cref{fig:ppc-plot} for different parameter choices on the x- and y-axes. White start denotes the MAP estimate.
-	\textbf{(C,D)} $F_A$ values (y-axis) as a function of $\delta/\mu$, the aneuploidy-to-mutation rate ratio.
+	\textbf{(A,B)} $F_A$ values (color coded) as in \Cref{fig:ppc-plot} for different parameter choices on the x- and y-axes. White star denotes the MAP estimate.
 	}

### 2022-06-28 ms/ms.tex

@@
       %% generated with dynamics-plot.ipynb
-      \includegraphics[width=7in,height=2.5in]{../figures/full_dynamic_MAP.pdf}      
+      \includegraphics[width=\textwidth]{../figures/MAP_freq_dynamics.pdf}      
   \end{subfigure}
@@
       \centering
-      \includegraphics[width=\textwidth]{../figures/histo_100000.pdf}
+      \includegraphics[width=\textwidth]{../figures/F_A_posterior.pdf}
   \end{subfigure}
@@
       \centering
-      \includegraphics[width=\textwidth]{../figures/scatter100000_log_star.pdf}
+      \includegraphics[width=\textwidth]{../figures/F_A_mu_delta.pdf}
   \end{subfigure}
@@
       \centering
-      \includegraphics[width=\textwidth]{../figures/scatter100000_rater_fitr_from_wt.pdf}      
+      \includegraphics[width=\textwidth]{../figures/F_A_delta-mu-ratio_1.pdf}      
 	\end{subfigure}
@@
       \centering
-      \includegraphics[width=\textwidth]{../figures/scatter100000_rater_fitr_mut.pdf}
+      \includegraphics[width=\textwidth]{../figures/F_A_delta-mu-ratio_2.pdf}
   \end{subfigure}

### 2022-06-29 src/inference.py
inference scripts were added

@@ -1,0 +1,187 @@
+inference.py
+import sys
+
+if len(sys.argv)==1:
+    print('usage:\npython inference.py num_reps seed tau prior_type stop_epsilon dir_name cpus')
+    print('num_reps - the number of simulations that are executed per each parameter vector')
+    print('seed - seed for random number generator that is used for each parameter vector simulations execution')
+    print('tau - mutation rate increase factor in aneuploid cells')
+    print('prior_type (see more description in the manuscript):')
+    print('   1 - informative prior for every fitness')
+    print('   2 - uninformative uniform prior for every fitness')
+    print('   3 - uninformative uniform prior for w(2n+1*) and w(2n*)')
+    print('   4 - extended informative prior')
+    print('   5 - informative prior; fixed mutation rate = 1e-5')
+    print('   6 - informative prior; fixed mutation rate = 1e-6')
+    print('   7 - informative prior; fixed mutation rate = 1e-7')
+    print('epsilon_stop - epsilon stop bound for pyabc run')
+    print('dir_name - the result files are persisted at ~/sim-data/single-model-abc/+dir_name')
+    print('cpus - num of cpu cores to use.')
+    exit()
+
+import pyabc
+import scipy
+import tempfile
+import os
+os.environ["OMP_NUM_THREADS"] = "1" #for numpy
+import numpy as np
+from models.singleLocusModel import SingleLocusModel
+from datetime import timedelta, datetime
+from shutil import copyfile, copytree, rmtree
+from multiprocessing import  Value
+
+N = 6.425*10**6 # 1.6M - 2004.8M in Yona experiment 8/(1.0/(np.array([2**i for i in range(8)])*1.6)).sum()
+GLOBAL_SEED = 123
+# K = 1.03125
+# A,B,C = 50, 300, 2200
+
+
+
+
+PRIOR_THRESHOLD = 0.01
+MAX_POPULATIONS = 50 # stop criterion: num of abc iterations
+A,B,C = 450, 1700, 2350
+REPS1 = int(sys.argv[1])
+SIM_SEED = int(sys.argv[2])
+K = float(sys.argv[3])
+PRIOR_TYPE = int(sys.argv[4])
+MIN_EPSILON = float(sys.argv[5]) #stop criterion
[…]

### 2022-06-30 src/models/singlelocusmodelextended.py
checked and cleaned the simulations file

@@ -12,6 +12,8 @@ class SingleLocusModelExt:
     # w1 is 2n+1 fitness
     # w2 is 2n+1* fitness
     # w3 is 2n* fitness
+    
+    #This method simulates follows the 4 states and records their frequency for every generation
     def run_simulations(self, N, mutation_rate, aneuploidy_rate, aneuploidy_loss_rate, w1, w2, w3, repetitions=1000, fixation=0.95, max_gen=2500, seed=3, clonal_intf=True):
         curr_rand_state = np.random.get_state()
         np.random.seed(seed)
@@ -19,7 +21,8 @@ class SingleLocusModelExt:
         #0 wt
         #1 tri
         #2 tri+
-        #3 +
+        #3 + (from tri+)
+        #4 + (from wt)
         Mt = np.diag(np.repeat(0.,5))
         Mt[0][1] = aneuploidy_rate
         Mt[1][2] = mutation_rate
@@ -32,8 +35,8 @@ class SingleLocusModelExt:
         return (times, p, non_aneu_fix)   
 
     
-    #have side effect on M, but not harmfull, can run twice simulation
-    def _simulation(self, N, w, Mt ,repetitions=1000, fixation=0.95, max_gen=2500,clonal_intf=True):
+    
+    def _simulation(self, N, w, Mt ,repetitions=1000, fixation=0.95, max_gen=2500, clonal_intf=True):
         assert N > 0
         N = np.uint64(N)
         L = len(w)
@@ -61,14 +64,12 @@ class SingleLocusModelExt:
         # which columns to update
         update = np.array([True] * repetitions)
         directmut = np.array([True] * repetitions)
-        #clonal_int=np.array([False]*repetitions)
         combfix=np.array([True] * repetitions)
-        fixbool=np.array([False]*repetitions)
+
         # follow time
         T = np.zeros(repetitions)
         t = 0
         # follow population mean fitness
-        W = []
         P = []
         E=Et
 
@@ -79,10 +80,8 @@ class SingleLocusModelExt:
             t += 1
             T[update] = t        
             p = n/N  # counts to frequencies
-    #         W.append(w.reshape(1, L) @ p)  # mean fitness
             P.append(p)
             p = E @ p  # natural selection + mutation   
-            #p[-1,clonal_int]=0
             p /= p.sum(axis=0)  # mean fitness
             for j in update.nonzero()[0]:
                 # random genetic drift
@@ -91,22 +90,21 @@ class SingleLocusModelExt:
             if clonal_intf==False:
[…]

### 2022-06-30 ms/ms.tex
todo comment from AY
AY short paragraph for results

@@
 Therefore, evidence does not support an increase in mutation rate in aneuploid cells, and moreover, unless the increase is strong ($\tau \ge 10$), it does not seem to affect our inference.
+% TODO Avihu - in the literature it is suggested elevated mut rate is due to stress; here wt is also stressed so maybe that's why we dont see an effect
 We also checked the differences in genotype frequency dynamics for different $\tau$ values.
@@
 The probability of \eumt\ descending from \anwt\ ($F_A$) increases with the aneuploidy rate, $\delta$, and decreases with the mutation rate, $\mu$, and in some cases can also be affected by the fitness parameters (\Cref{fig:ppc-plot}C and \Cref{fig:posterior_FA_supp}).
+
+%TODO revise AY text:
+At the first stage there are more 2n cells than 2n+1 cells because of the advantage of aneuploidy under heat.
+Accordingly, at the seconds stage, there are more 2n+1* cells than 2n*.
+However, in most cases 2n*M dominate the population (and not 2n*A), because the fitness advantage of a refined mutation that emerge in 2n+1 background in hindered by the cost of aneuploidy and by the extra step needed to eliminate the extra copy of the chromosome.
 

### 2022-07-18 ms/ms.tex
start reorganize, add mutation results

@@
 % NatBib
-\usepackage[round,colon,authoryear]{natbib}
+\usepackage[super,comma]{natbib}
+%\usepackage[round,colon,authoryear]{natbib}
 
@@
 % Chromosomal duplication is a transient evolutionary solution to stress
-%\title{Clonal interference between aneuploidy and mutation}
-\title{Adaptive evolution with aneuploidy and mutation}
+\title{Aneuploidy is an evolutionary sidetrack on the path to adaptation}
 
@@
 
-\paragraph*{Aneuploidy is common in eukaryotes.}
+%\paragraph*{Aneuploidy is common in eukaryotes.}
 Aneuploidy is an imbalance in the number of chromosomes in the cell: an incorrect karyotype.
 Evidence suggests aneuploidy is very common in eukaryotes, e.g. animals~\citep{Santaguida2015review, Naylor2016, Bakhoum2017}, and fungi~\citep{Pavelka2010, Zhu2016, Robbins2017, Todd2017}.
-Aneuploidy has been implicated in cancer formation and progression~\citep{Boveri2008, Schvartzman2010}: 
-90\% of solid tumours and 50\% of blood cancers are aneuploid~\citep{Santaguida2015review}.
-Aneuploidy is also linked to the emergence of drug resistance~\citep{Selmecki2009} and virulence~\citep{Moller2018} in fungal pathogens, which are under-studied~\citep{Rodrigues2018} despite infecting close to a billion people per year, causing serious infections and significant morbidity in >150 million people per year and killing >1.5 million people per year~\citep{Selmecki2009, Rodrigues2018}.
-In addition, aneuploidy is common in protozoan pathogens of the Leishmania genus, a major global health concern~\citep{Mannaert2012}.
+Aneuploidy has been implicated in cancer formation and progression~\citep{Boveri2008, Schvartzman2010,Santaguida2015review}.
+%90\% of solid tumours and 50\% of blood cancers are aneuploid~\citep{Santaguida2015review}.
+It is also common in protozoan pathogens of the Leishmania genus, a major global health concern~\citep{Mannaert2012}, and contributes to the emergence of drug resistance~\citep{Selmecki2009} and virulence~\citep{Moller2018} in fungal pathogens, which are under-studied~\citep{Rodrigues2018}, despite infecting a billion people per year, causing significant morbidity in >150 million and death in >1.5 million people per year~\citep{Selmecki2009, Rodrigues2018}.
 
-\paragraph*{Aneuploidy is generally deleterious.}
-The molecular and genetic mechanisms involved in aneuploidy have been explored~\citep{Musacchio2007, Sheltzer2011, Chen2012b, Rancati2013, Gerstein2015, Shor2015}.
+%\paragraph*{Aneuploidy is generally deleterious.}
+%The molecular and genetic mechanisms involved in aneuploidy have been explored~\citep{Musacchio2007, Sheltzer2011, Chen2012b, Rancati2013, Gerstein2015, Shor2015}.
 Experiments with human and mouse embryos found that aneuploidy is usually lethal.
@@
 For example, aneuploid mouse embryonic cells grow slower than euploid cells~\citep{Williams2008}.
-Similarly, in unicellular eukaryotes growing in benign conditions, aneuploidy usually leads to slower growth and decreased overall fitness~\citep{Niwa2006, Torres2007, Pavelka2010, Sheltzer2011, Kasuga2016}, in part due to proteotoxic stress caused by increased expression in aneuploid cells~\citep{Pavelka2010, Santaguida2015, Zhu2018} and hypo-osmotic-like stress~\citep{Tsai2019}.
-% TODO cite Yang, F. et al. The fitness costs and benefits of trisomy of each Candida albicans chromosome. Genetics 218, 1–7 (2021).
+Similarly, in unicellular eukaryotes growing in benign conditions, aneuploidy usually leads to slower growth and decreased overall fitness~\citep{Niwa2006, Torres2007, Pavelka2010, Sheltzer2011, Kasuga2016, Yang2021}, in part due to proteotoxic stress caused by increased expression in aneuploid cells~\citep{Pavelka2010, Santaguida2015, Zhu2018} and hypo-osmotic-like stress~\citep{Tsai2019}.
 
-\paragraph*{Aneuploidy can lead to adaptation.}
-However, aneuploidy can be beneficial under stressful conditions due to the wide range of phenotypes it can produce, some of which are advantageous~\citep{Pavelka2010}.
+%\paragraph*{Aneuploidy can lead to adaptation.}
+However, aneuploidy can be beneficial under stressful conditions due to the wide range of phenotypes it can produce, some of which are advantageous~\citep{Pavelka2010, Yang2021}.
 Thus, aneuploidy can lead to rapid adaptation in unicellular eukaryotes~\citep{Gerstein2015,Torres2010, Hong2014, Rancati2008}, as well as to rapid growth of somatic tumour cells~\citep{Schvartzman2010, Sheltzer2017}.
-For example, aneuploidy in \yeast facilitates adaptation to a variety of stressful conditions like heat and pH~\citep{Yona2012}, copper~\citep{Covo2014, Gerstein2015}, salt~\citep{Dhar2011}, and nutrient limitation~\citep{Dunham2002, Gresham2008}.
+For example, aneuploidy in \yeast facilitates adaptation to a variety of stressful conditions like heat and pH~\citep{Yona2012}, copper~\citep{Covo2014, Gerstein2015}, salt~\citep{Dhar2011}, and nutrient limitation~\citep{Dunham2002, Gresham2008, Avecilla2022}, with similar results in \calbicans~\citep{Yang2021}.
 Importantly, aneuploidy can also lead to drug resistance in pathogenic fungi such as \calbicans~\citep{Selmecki2008, Selmecki2010, Gerstein2018} and \cneoformans~\citep{Sionov2010}, which cause candidiasis and meningoencephalitis, respectively.
 
-\paragraph*{Transient adaptive solution.} 
-Aneuploidy differs from mutation due to its distinct properties. 
-Chromosome duplication usually occurs more often than mutation and on average produces larger fitness effects.
-Yet, because it affects many genes on a whole chromosome or a chromosome fragment, aneuploidy also carries fitness costs.
-Thus, aneuploidy can be a \emph{transient adaptive solution}: it can rapidly occur and fix in the population under stressful conditions, and can be rapidly lost when the cost outweighs the benefit---when stress is removed or after beneficial mutations occur.
-Experimental evidence of such a transient role of aneuploidy was demonstrated by \citet{Yona2012}. They evolved populations of \yeast under strong heat or pH stress.
-The populations adapted to the heat and pH stress within 450 and 150 generations, and this adaptation was determined to be due to chromosome duplications.
-Much later, after more than 1500 and 750 generations, for the heat and pH stress, respectively, the populations reverted back to an euploid state, while remaining adapted to the stress and accumulating multiple mutations.
-However, under gradual heat stress, aneuploidy was not observed.
-\citet{Yona2012} concluded that aneuploidy serves as a transient adaptive solution, or a ``quick fix'', which is expected to facilitate adaptation.
+%\paragraph*{Transient adaptive solution.} 
+Experimental evidence for the role of aneuploidy in adaptive evolution was demonstrated by \citet{Yona2012}. They evolved populations of \yeast under strong heat stress.
+The populations adapted to the heat stress within 450 generations, and this adaptation was determined to be due a duplication of chromosome III.
+Later on, after more than 1,500 generations, the populations reverted back to an euploid state, while remaining adapted to the heat stress. 
+The authors hypothesized that aneuploidy is a \emph{transient adaptive solution}, because it can rapidly appear and fixate in the population under stressful conditions, and can then be rapidly lost when the cost of aneuploidy outweighs its benefit---after the stress is removed, or after "refined" beneficial mutations appear and fixate.
 
-\paragraph*{The present study.}
-Here, we develop an evolutionary-genetic model that includes the effects of natural selection, genetic drift, aneuploidy, and mutation to examine the role of aneuploidy in adaptive evolution.
-This model follows a population of cells characterised by both their ploidy and their genotype.
-We fit this model to the experimental results of \citet{Yona2012} using an \emph{approximate Bayesian computation} framework~\citep{Sisson2009} to infer model parameters, including rates and fitness effects of aneuploidy and mutation, and to test hypotheses about the evolutionary process by comparing different versions of the model and predicting the genotype frequency dynamics using the estimated parameter values.
+%\paragraph*{The present study.}
+Here, we test the hypothesis that aneuploidy is a \emph{transient adaptive solution}.
+First, we analyze previously unpublished sequencing data from the original experimental populations of \citet{Yona2012} to assess if the aneuploid population and the evolved euploid population carry the same mutant alleles, as expected if the later is descended from the former.
+Second, we develop an evolutionary genetic model and fit it to the experimental results of \citet{Yona2012} in order to predict the genotype frequency dynamics in the experimental populations, thereby estimating the frequency of evolved euploid cells that descended from aneuploid cells.
+Our results show that, despite aneuploidy having reached high frequencies in the experimental populations, the majority of cells in the evolved euploid populations likely did not descend from aneuploid cells, but rather directly from wild type euploid cells.
+These results suggest that aneuploidy may have been a sidetrack, rather than a stepping stone, on the path to adaptation.
 
-%%%%%%%%%%%%%%%%%%%%%%%%
-% Models and Methods
-\section*{Models and Methods}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Results
+\section*{Results}
+
+\paragraph{DNA sequencing.}
+
+In the heat-stress experiment of \citet{Yona2012}, four populations of \yeast evolved under \SI{39}{\celsius}.
+Aneuploidy fixed in all four experimental repetitions in the first 450 generations.
+Two of the repetitions, marked $H2$ and $H4$, carried no large-scale duplications other than a chromosome III trisomy.
+These two repetitions continued to evolve under the same conditions, wherein aneuploidy was eliminated by generation 1,700 and 2,350.
+For each of these two populations we sequenced the ancestral population (generation zero), the aneuploid population (generation 450), and the evolved euploid population (generation 1,700 or 2,350) to estimate the mutant allele frequencies (\Cref{table:mutations}).
+Overall, between 100 and 173 mutant alleles were detected with at least a single read in the six population that were sampled. 
+Disregarding 45 and 40 alleles that were present in the ancestral populations at a frequency >10\%, the aneuploid and euploid populations carried a large number of mutant alleles: 82 and 95, respectively, in repetition $H2$, and 60 and 66 in repetition $H4$.
+
+Surprisingly, out of all these mutant alleles, none was present at a frequency >20\% in both the aneuploid and the evolved euploid populations. 
+Moreover, a high mutant allele frequency in the aneuploid population implies a low frequency in the evolved euploid population, and vice-versa (Spearman's correlation coefficient $\rho=-0.64$ and $-0.66$ in the two experimental repetitions; \Cref{fig:mutation_frequencies}).
+Specifically, for the 18 mutant alleles with high frequency in the aneuploid populations (>20\%), the highest frequencies in the euploid populations were 15.4\%, 16\%, 16.3\% and 19.6\% (the rest were below 15\%).
+Similarly, for the 48 mutant alleles with high frequency in the evolved euploid populations, the highest frequencies in the aneuploid populations were 2.7\%, 7.7\%, and 11.1\% (the rest were below 1\%).
+Thus, mutant alleles frequent in the aneuploid populations decreased in frequency when aneuploidy was lost, and most evolved euploid cells are likely not descendent from aneuploid~cells.
+
+%%%%%%%% Fig mutation frequencies
+\begin{figure}[h!]
+  \centering
+    \includegraphics[width=\textwidth]{../figures/mutation_frequencies.pdf}      
+  \caption{
+    \textbf{Frequencies of mutant alleles in the experimental populations are negatively correlated.} 
+    Frequencies of mutant alleles when trisomy was widespread in the population (y-axis) and after it was eliminated (x-axis) in two experimental repetitions (circles for $H2$ and triangles for $H4$) from \citet{Yona2012}.
+Mutant alleles with >20\% in the aneuploid population were <20\% in the euploid population, and vice versa (the upper-right quadrant is empty), suggesting that the majority of evolved euploid cells did not descend from the most common aneuploid genotypes.
+Alleles with frequency below and above 10\% in the ancestral populations are in black and gray, respectively. 
+Solid black line is a linear orthogonal distance regression line (slope=$-0.559$, intercept=$0.164$).
+Dashed vertical and horizontal lines show allele frequencies 20\%. Error bars show standard error of the mean accounting for the number of reads.
+}
+  \label{fig:mutation_frequencies}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\paragraph{Evolutionary genetic model.} 
+% TODO continue here
+
+We develop an evolutionary genetic model that that includes the effects of natural selection, genetic drift, aneuploidy, and mutation to examine the role of aneuploidy in adaptive evolution.
+The model follows a population of cells characterised by both their ploidy and their genotype, fit this model to the experimental results of \citet{Yona2012} using an \emph{approximate Bayesian computation} framework~\citep{Sisson2009} to infer model parameters, including rates and fitness effects of aneuploidy and mutation, and to test hypotheses about the evolutionary process by comparing different versions of the model and predicting the genotype frequency dynamics using the estimated parameter values.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph*{Evolutionary Model.}
-We model the evolution of a population of cells using a Wright-Fisher model~\citep{Otto2007}, assuming a constant effective population size $N$, non-overlapping generations, and including the effects of natural selection, genetic drift, aneuploidy, and mutation.
-We focus on beneficial genetic modifications, neglecting the effects of deleterious and neutral mutations or karyotypic changes.
-The model allows for a single aneuploid karyotype (e.g., chromosome III duplication) and a single mutation to accumulate in the genotype.
-Thus, the model follows four genotypes~(\Cref{fig:models}): euploid wild-type, \euwt, the initial genotype; 
-euploid mutant, \eumt, with the standard karyotype and a single beneficial mutation; 
-aneuploid wild-type, \anwt, with an extra chromosome, i.e., following chromosome duplication; and
-aneuploid mutant, \anmt, with and extra chromosome and a beneficial mutation.
-
-Transitions between the genotypes occur as follows~(\Cref{fig:models}): Beneficial mutations from \euwt\; to \eumt\; and from \anwt\; to \anmt\; occur with probability $\mu$, the mutation rate.
-We neglect back-mutations (i.e., from \eumt\; to \euwt\; and from \anmt\; to \anwt).
-Aneuploidy is formed by chromosome mis-segregation, so that cells transition from \euwt\; to \anwt\; and from \anmt\ to \eumt\ with probability $\delta$, the aneuploidy rate.
-That is, we assume chromosomes are gained and lost at the same rate, and we neglect events that form a less-fit genotype (i.e., \anwt\; to \euwt\; and \eumt\; to \anmt).
-
-In the experiment by \citet{Yona2012}, the population was grown every day from $1.6 \cdot 10^6$ cells until reaching stationary phase and then diluted 1:120.
-Thus, we set the population size to $N=6.425 \cdot 10^6$, the harmonic mean of $\{2^k \cdot 1.6 \cdot 10^6\}_{k=0}^{7}$~\citep{Crow1970}.
-The initial population has $N$ cells with genotype \euwt. 
-The effect of natural selection on the frequency $f_i$ of genotype $i = 2n, 2n+1, 2n+1^*, \text{or } 2n^*$ is given by
-    \begin{equation} \label{eq:selection-single} 
-      f^s_i = \frac{f_i w_i}{\bar{w}} \;,
-    \end{equation}
-where $w_i$ is the fitness of genotype $i$ and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
-The effect of mutation and aneuploidy on genotype frequencies is given by
-    \begin{equation} \label{eq:mutation-aneuploidy-single}
-    \begin{aligned}
-      &f^m_{2n} &=&\; (1 - \delta - \mu) f^s_{2n}  \;,\\
-      &f^m_{2n+1} &=&\; \delta f^s_{2n} + (1 - \mu) f^s_{2n+1}  \;,\\
-      &f^m_{2n+1^*} &=&\; \mu f^s_{2n+1} + (1-\delta) f^s_{2n+1^*}  \;,\\
-      &f^m_{2n^*} &=&\; \mu f^s_{2n} + \delta f^s_{2n+1} + f^s_{2n^*}  \;.
-    \end{aligned}
-    \end{equation}
-Finally, random genetic drift is modeled using a multinomial distribution~\citep{Otto2007},
-    \begin{equation} \label{eq:drift-single}
-      \vec{f'} \sim \frac{1}{N} \cdot \mathit{Mult}(N,\ \vec{f^m}) \;,
-    \end{equation}
-where $\vec{f^m}=(f^m_{2n}\ ,\ f^m_{2n+1}\ ,\ f^m_{2n+1^*}\ ,\ f^m_{2n^*})$ are the frequencies of the genotypes after mutation and aneuploidy, $\vec{f'}$ are the genotype frequencies in the next generation, and $Mult(N,\ \vec{f})$ is a multinomial distribution parameterized by the population size $N$ and the genotype frequencies $\vec{f}$.
-Overall, the change in genotype frequencies from one generation to the next is given by the transformation $f_i \to f'_i$.
-
-\paragraph{Empirical evidence.}
-
-We use the results of evolutionary experiments reported by \citet{Yona2012}.
-In their heat-stress experiment, four populations of \yeast evolved under \SI{39}{\celsius}.
-Aneuploidy fixed in all four population in the first 450 generations (hereafter, fixation or elimination of a genotype \emph{by generation $t$} means that more than 95\% or less than 5\% of the population carry the genotype at generation $t$, and possibly earlier).
-From re-analysis of data not published in the original paper, aneuploidy did not fix before at least 200 generations elapsed.
-The experiment continued with two populations, in which aneuploidy was eliminated by generation 1,700 and 2,350 while still under the same conditions of elevated heat (\SI{39}{\celsius}).
-
-\paragraph{Likelihood function.} 
-Because our model, just like the Wright-Fisher model, is non-linear and stochastic, computing the distribution of fixation time $T(g)$ of genotype $g$ for use in the likelihood function is intractable (it is even hard to use a diffusion-equation approximation due to the model having multiple genotypes, rather than just two).
-We overcome this problem by approximating the likelihood using simulations.
-We simulate 1,000 experiments per parameter vector $\theta = (\mu, \delta, s, b, c)$, resulting in a set of simulated observations $\tilde{\vec X} = \{\tilde{X}_i\}_{i=1}^{1000}$.
-We then compute the approximate likelihood,
-\begin{equation}\begin{aligned}
-\label{eq:heatstress-likelihood}
-\likelihood(\theta) = &\ P^4(200 \le T(2n+1) \le 450) \cdot 
-	\Big[1 - \\
-	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)- \\
-	&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big)+ \\
-	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big) 
-	\Big]\;,
-\end{aligned}\end{equation}
-where $!\{\ldots\}$ is the "logical not" operator, $P^4(\ldots)$ is the 4th power of $P(\ldots)$, and all probabilities $P_{\tilde{\vec X}}(\ldots)$ are approximated from the results of the simulations $\tilde{\vec X}$.
-For example, $P_{\tilde{\vec X}}\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)$ is approximated by taking simulations in which \anwt\; fixed before generation 450 but not before generation 200, and computing the fraction of such simulations in which \eumt\; did not fix by generation 1,700, and hence aneuploidy did not extinct before generation 1,700.
-\Cref{fig:seeds} compares results with less and more simulated experiments, demonstrating that 1,000 simulations are likely sufficient.
-
-For a model without aneuploidy (that is, when the aneuploidy rate is fixed at zero, $\delta=0$), we disregard the increased expression in chromosome III and the growth advantage measured in generation 450, and focus on the growth advantage measured in later generations, presumably due to a beneficial mutation.
-Therefore, the likelihood is approximated by
-\begin{equation}\begin{aligned}
-\label{eq:heatstress-noaneuploidy-likelihood}
-\likelihood_{!}(\theta) = &\ 
-	1 - 
-	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\}\big) - \\
-&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\}\big) + \\
-&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\}\big)
-\;.
-\end{aligned}\end{equation}
-
-\paragraph{Parameter inference.} To infer model parameters, we use approximate Bayesian computation with a sequential Monte-Carlo scheme, or ABC-SMC~\citep{Sisson2009}, implemented in the \texttt{pyABC} Python package~\citep[\href{https://pyabc.readthedocs.io}{pyabc.readthedocs.io}]{Klinger2018}.
-This approach uses numerical stochastic simulations of the model to infer a posterior distribution over the model parameters.
-It is a method of likelihood-free, simulation-based inference~\citep{Cranmer2020}, that is, for estimating a posterior distribution when a likelihood function cannot be directly computed.
-It is therefore suitable in our case, in which the likelihood function can only be approximated from simulations, and cannot be directly computed.
-
-The ABC-SMC algorithm employs sequential importance sampling over multiple iterations~\citep{Toni2009, Klinger2017, Syga2021}.
-In iteration $t$ of the algorithm, a set of parameter vectors, $\{\theta_{i,t}\}_{i=1}^{n_t}$, also called \emph{particles}, are constructed in the following way.
-A proposal particle, $\theta^*$, is sampled from a proposal distribution, and is either accepted or rejected, until $n_t$ particles are accepted.
-The number of particles, $n_t$, is adapted at every iteration $t$ using the adaptive population strategy~\citep[\href{https://pyabc.readthedocs.io}{pyabc.readthedocs.io}]{Klinger2018}.
-For $t=0$, the proposal particle is sampled from the prior distribution, $p(\theta)$.
-For $t>0$, the proposal particle is sampled from the particles accepted in the previous iteration, $\{\theta_{i,t-1}\}_{i=1}^{n_{t-1}}$, each with a probability relative to its weight $W_{t-1}(\theta_{i,t-1})$ (see below).
-The proposal particle is then perturbed using a kernel perturbation kernel, $K_t(\theta^* \mid \theta)$ where $\theta$ is the sample from the previous iteration.
-Then, a set of synthetic observations $\tilde{\vec X}^*$ is simulated, and the proposal particle $\theta^*$ is accepted if its approximate likelihood (\cref{eq:heatstress-likelihood}) is high enough, $\likelihood(\theta^*)>1-\epsilon_t$ (or more commonly, if $1-\likelihood(\theta^*) < \epsilon_t$), where $\epsilon_t>0$ is the \emph{acceptance threshold}, as higher values of $\epsilon_t$ allow more particles to be accepted.
-The acceptance threshold $\epsilon_t$ is chosen as the median of the $1-\likelihood(\theta)$ of the particles accepted in the previous iteration, $t-1$, and $\epsilon_0=0.01$. 
-For each accepted particle $\theta_{i,t}$ a weight $W_t(\theta_{i,t})$ is assigned: for $t=0$, $W_0(\theta_{i,0})=1$, and for $t>0$, 
-$W_t(\theta_{i,t}) = p(\theta_{i,t}) / \sum_{i=1}^{n_{t-1}}{W_{t-1}(\theta_{i,t-1}) K_t(\theta_{i,t}, \theta_{i,t-1})}$, where $p(\theta)$ is the prior density of $\theta$ and $K_t(\theta', \theta)$ is the probability of a perturbation from $\theta$ to $\theta'$.
-$K_t(\theta' \mid \theta)$ is a multivariate normal distribution, fitted at iteration $t$ to the particles from the previous iteration, \{$\theta_{i,t-1}\}_{i=1}{n_{t-1}}$, and their weights, $\{W(\theta_{i,t-1})\}_{i=1}^{n_{t-1}}$.
-
-Acceptance is determined according to the approximate likelihood~(\cref{eq:heatstress-likelihood}), which has a maximum value of $\likelihood_{max}=0.875$ (giving a minimal value of $\epsilon_{min}=0.125$).
-We terminated the inference iterations when the change in $\epsilon$ value from one iteration to the next was small.
-With our standard prior and model, we reached $\epsilon=0.13$ (or $\likelihood=0.87$) after six iterations, with $n_6=982$ accepted parameter vectors and effective sample size ESS=651 (\Cref{fig:convergence}).
-Running the inference algorithm with different initialization seeds and less or more simulations for approximating the likelihood produced similar posterior distributions~(\Cref{fig:seeds}).
-
-After producing a set of weighted particles from the the posterior distribution using the above ABC-SMC algorithm, we approximate the posterior using kernel density estimation (KDE) with Gaussian kernels.
-We truncate the estimated posterior to avoid positive posterior density for values with zero prior density.
-The MAP (maximum a posteriori) estimate is computed as the the maximum of the estimated joint posterior density.
-We then draw $5,000,000$ samples from the posterior distribution to compute the HDI (highest density interval) and draw $50,000$ samples to visualize the posterior distribution with histograms.
-
-\paragraph{Model comparison.} 
-We examine several versions of our evolutionary models, e.g. without aneuploidy or with increased mutation rate in aneuploid cells, as well as several different prior distributions (see below).
-To compare these, we plot posterior predictions: for each model we execute $10,000$ simulations using the MAP parameter estimates and plot the distributions of time to fixation of \eumt, one of key properties of the model likelihood.
-These plots visualize the fit of each model to the data.
-Also, for similar models we plot the marginal and joint posterior distributions of the parameters; if these are similar, we consider the models interchangeable.
-We validate this by comparing HDI (highest density interval) of posterior distributions.
-
-Where posterior plots are very similar and the number of parameters is the same, we use WAIC, or the widely applicable information criterion ~\citep{gelman2013bayesian}, defined as
-\begin{equation} \label{eq:WAIC}
-\begin{aligned}
-\mathit{WAIC}(\theta) &\ =\ 
--2\log\mathbb{E}[\likelihood(\theta)]\ +\ 2\mathbb{V}[\log\likelihood(\theta)]
-\end{aligned}
-\end{equation}
-where $\theta$ is a parameter vector, and $\mathbb{E}[\cdot]$ and $\mathbb{V}[\cdot]$ are the expectation and variance taken over the posterior distribution, which in practice are approximated using 50,000 samples from the posterior KDE.
-We validated that upon resampling WAIC values do not significantly change and that differences in WAIC between models are preserved.
-WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy~\citep{Kass1995}.
-
-\paragraph{Prior distributions.} \label{par:prior}
-We used informative prior distributions for $w_{\anwt}=1-c+b$, $w_{\anmt}=(1+s)(1-c)+b$ and $w_{\eumt}=1+s$, which we estimated from growth curves data from mono-culture growth experiments previously reported by \citet[Figs.
-3C, 4A, and S2]{Yona2012}.
-We used \texttt{Curveball}, a method for predicting results of competition experiments from growth curve data~\citep[\href{https://curveball.yoavram.com}{curveball.yoavram.com}]{Ram2019}.
-Briefly, \texttt{Curveball} takes growth curves of two strains growing separately in mono-culture and predicts how they would grow in a mixed culture, that is, it predicts the results of a competition assay.
-From these predictions, relative fitness values can be computed.
-Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values to be used as a prior distribution by sampling 10,000 samples from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix~(\Cref{fig:growth-curves}).
-
-We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate an informative prior distribution for $w_{\anwt}$~(\Cref{fig:growth-curves}-D, assuming $w_{\euwt}=1$).
-In this prior distribution, we used the same prior for $w_{\anmt}$ and $w_{\eumt}$. 
-To increase computational efficiency, we also assumed $w_\eumt>w_\anmt>w_\anwt>w_\euwt$; running the inference without this assumption produced similar results.
-See \emph{supporting material} for an extended informative prior distribution that uses growth curves of \eumt\ and \anwt\ growing in \SI{39}{\celsius}; this prior distribution proved to be less useful.
-
-As a control, we tested an uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{\anwt}$, $w_{\anmt}$, $w_{\eumt}$, or (ii) only for $w_{\anmt}$, $w_{\eumt}$, using the above informative prior for $w_{\anwt}$.
-In these cases the inference algorithm failed to converge.
-
-For the mutation rate, $\mu$, and aneuploidy rate, $\delta$, we used uninformative uniform priors, $\mu \sim \mathit{U}(10^{-9},10^{-5})$ and $\delta \sim \mathit{U}(10^{-6},10^{-2})$.
-A wider mutation rate prior, $\mu \sim \mathit{U}(10^{-9},10^{-3})$, produced similar results.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Results
-\section*{Results}
 
@@
 
+% quote: This scenario may thus imply that the duplication of a cer- tain chromosome may provide the evolutionary time window needed for the population to search for solutions not only on the duplicated chromosome but throughout the genome.
+
+\citet{Yona2012} have suggested that the aneuploid cells that come to dominate the experimental populations were later replaced by ``refined strains, which eliminated the cost associated with the trisomy, exhibit improved growth both at the permissive and the high temperatures compared with their trisomic ancestors'' and that ``it is expected that the original contribution of trisomy to heat tolerance was replaced by alternative solutions.''.
+However, our results suggest that most of the cells in the evolved euploid population do not carry mutations that were common in the aneuploid populations, but rather a different set of mutations.
+Therefore, the majority of the evolved euploid cells are not descended from aneuploid cells, but rather are direct descendants of the ancestral population.
+
 \paragraph*{Results of current study.}
@@
 Furthermore, we find that XXXX. % TODO transient solution?
+
+\cite{Yona2012} have analyzed the expression levels of 14 heat-shock proteins (HSP).
+They found that their expression level changed ``very modestly'' in the aneuploid populations, but that most of them ``showed significant up-regulation'' in the evolved euploid population.
 
@@
 
-%\pagebreak
+%%%%%%%%%%%%%%%%%%%%%%%%
+% Models and Methods
+\section*{Models and Methods}
+
+%%%%%%%%%%%%%%%%%%%%%%%%
+\paragraph*{Evolutionary Model.}
+We model the evolution of a population of cells using a Wright-Fisher model~\citep{Otto2007}, assuming a constant effective population size $N$, non-overlapping generations, and including the effects of natural selection, genetic drift, aneuploidy, and mutation.
+We focus on beneficial genetic modifications, neglecting the effects of deleterious and neutral mutations or karyotypic changes.
+The model allows for a single aneuploid karyotype (e.g., chromosome III duplication) and a single mutation to accumulate in the genotype.
+Thus, the model follows four genotypes~(\Cref{fig:models}): euploid wild-type, \euwt, the initial genotype; 
+euploid mutant, \eumt, with the standard karyotype and a single beneficial mutation; 
+aneuploid wild-type, \anwt, with an extra chromosome, i.e., following chromosome duplication; and
+aneuploid mutant, \anmt, with and extra chromosome and a beneficial mutation.
+
+Transitions between the genotypes occur as follows~(\Cref{fig:models}): Beneficial mutations from \euwt\; to \eumt\; and from \anwt\; to \anmt\; occur with probability $\mu$, the mutation rate.
+We neglect back-mutations (i.e., from \eumt\; to \euwt\; and from \anmt\; to \anwt).
+Aneuploidy is formed by chromosome mis-segregation, so that cells transition from \euwt\; to \anwt\; and from \anmt\ to \eumt\ with probability $\delta$, the aneuploidy rate.
+That is, we assume chromosomes are gained and lost at the same rate, and we neglect events that form a less-fit genotype (i.e., \anwt\; to \euwt\; and \eumt\; to \anmt).
+
+In the experiment by \citet{Yona2012}, the population was grown every day from $1.6 \cdot 10^6$ cells until reaching stationary phase and then diluted 1:120.
+Thus, we set the population size to $N=6.425 \cdot 10^6$, the harmonic mean of $\{2^k \cdot 1.6 \cdot 10^6\}_{k=0}^{7}$~\citep{Crow1970}.
+The initial population has $N$ cells with genotype \euwt. 
+The effect of natural selection on the frequency $f_i$ of genotype $i = 2n, 2n+1, 2n+1^*, \text{or } 2n^*$ is given by
+    \begin{equation} \label{eq:selection-single} 
+      f^s_i = \frac{f_i w_i}{\bar{w}} \;,
+    \end{equation}
+where $w_i$ is the fitness of genotype $i$ and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
+The effect of mutation and aneuploidy on genotype frequencies is given by
+    \begin{equation} \label{eq:mutation-aneuploidy-single}
+    \begin{aligned}
+      &f^m_{2n} &=&\; (1 - \delta - \mu) f^s_{2n}  \;,\\
+      &f^m_{2n+1} &=&\; \delta f^s_{2n} + (1 - \mu) f^s_{2n+1}  \;,\\
+      &f^m_{2n+1^*} &=&\; \mu f^s_{2n+1} + (1-\delta) f^s_{2n+1^*}  \;,\\
+      &f^m_{2n^*} &=&\; \mu f^s_{2n} + \delta f^s_{2n+1} + f^s_{2n^*}  \;.
+    \end{aligned}
+    \end{equation}
+Finally, random genetic drift is modeled using a multinomial distribution~\citep{Otto2007},
+    \begin{equation} \label{eq:drift-single}
+      \vec{f'} \sim \frac{1}{N} \cdot \mathit{Mult}(N,\ \vec{f^m}) \;,
+    \end{equation}
+where $\vec{f^m}=(f^m_{2n}\ ,\ f^m_{2n+1}\ ,\ f^m_{2n+1^*}\ ,\ f^m_{2n^*})$ are the frequencies of the genotypes after mutation and aneuploidy, $\vec{f'}$ are the genotype frequencies in the next generation, and $Mult(N,\ \vec{f})$ is a multinomial distribution parameterized by the population size $N$ and the genotype frequencies $\vec{f}$.
+Overall, the change in genotype frequencies from one generation to the next is given by the transformation $f_i \to f'_i$.
+
+\paragraph{Empirical evidence.}
+
+We use the results of evolutionary experiments reported by \citet{Yona2012}.
+In their heat-stress experiment, four populations of \yeast evolved under \SI{39}{\celsius}.
+Aneuploidy fixed in all four population in the first 450 generations (hereafter, fixation or elimination of a genotype \emph{by generation $t$} means that more than 95\% or less than 5\% of the population carry the genotype at generation $t$, and possibly earlier).
+From re-analysis of data not published in the original paper, aneuploidy did not fix before at least 200 generations elapsed.
+The experiment continued with two populations, in which aneuploidy was eliminated by generation 1,700 and 2,350 while still under the same conditions of elevated heat (\SI{39}{\celsius}).
+
+\paragraph{Likelihood function.} 
+Because our model, just like the Wright-Fisher model, is non-linear and stochastic, computing the distribution of fixation time $T(g)$ of genotype $g$ for use in the likelihood function is intractable (it is even hard to use a diffusion-equation approximation due to the model having multiple genotypes, rather than just two).
+We overcome this problem by approximating the likelihood using simulations.
+We simulate 1,000 experiments per parameter vector $\theta = (\mu, \delta, s, b, c)$, resulting in a set of simulated observations $\tilde{\vec X} = \{\tilde{X}_i\}_{i=1}^{1000}$.
+We then compute the approximate likelihood,
+\begin{equation}\begin{aligned}
+\label{eq:heatstress-likelihood}
+\likelihood(\theta) = &\ P^4(200 \le T(2n+1) \le 450) \cdot 
+	\Big[1 - \\
+	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)- \\
+	&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big)+ \\
+	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big) 
+	\Big]\;,
+\end{aligned}\end{equation}
+where $!\{\ldots\}$ is the "logical not" operator, $P^4(\ldots)$ is the 4th power of $P(\ldots)$, and all probabilities $P_{\tilde{\vec X}}(\ldots)$ are approximated from the results of the simulations $\tilde{\vec X}$.
+For example, $P_{\tilde{\vec X}}\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)$ is approximated by taking simulations in which \anwt\; fixed before generation 450 but not before generation 200, and computing the fraction of such simulations in which \eumt\; did not fix by generation 1,700, and hence aneuploidy did not extinct before generation 1,700.
+\Cref{fig:seeds} compares results with less and more simulated experiments, demonstrating that 1,000 simulations are likely sufficient.
+
+For a model without aneuploidy (that is, when the aneuploidy rate is fixed at zero, $\delta=0$), we disregard the increased expression in chromosome III and the growth advantage measured in generation 450, and focus on the growth advantage measured in later generations, presumably due to a beneficial mutation.
+Therefore, the likelihood is approximated by
+\begin{equation}\begin{aligned}
+\label{eq:heatstress-noaneuploidy-likelihood}
+\likelihood_{!}(\theta) = &\ 
+	1 - 
+	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\}\big) - \\
+&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\}\big) + \\
+&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\}\big)
+\;.
+\end{aligned}\end{equation}
+
+\paragraph{Parameter inference.} To infer model parameters, we use approximate Bayesian computation with a sequential Monte-Carlo scheme, or ABC-SMC~\citep{Sisson2009}, implemented in the \texttt{pyABC} Python package~\citep[\href{https://pyabc.readthedocs.io}{pyabc.readthedocs.io}]{Klinger2018}.
+This approach uses numerical stochastic simulations of the model to infer a posterior distribution over the model parameters.
+It is a method of likelihood-free, simulation-based inference~\citep{Cranmer2020}, that is, for estimating a posterior distribution when a likelihood function cannot be directly computed.
+It is therefore suitable in our case, in which the likelihood function can only be approximated from simulations, and cannot be directly computed.
+
+The ABC-SMC algorithm employs sequential importance sampling over multiple iterations~\citep{Toni2009, Klinger2017, Syga2021}.
+In iteration $t$ of the algorithm, a set of parameter vectors, $\{\theta_{i,t}\}_{i=1}^{n_t}$, also called \emph{particles}, are constructed in the following way.
+A proposal particle, $\theta^*$, is sampled from a proposal distribution, and is either accepted or rejected, until $n_t$ particles are accepted.
+The number of particles, $n_t$, is adapted at every iteration $t$ using the adaptive population strategy~\citep[\href{https://pyabc.readthedocs.io}{pyabc.readthedocs.io}]{Klinger2018}.
+For $t=0$, the proposal particle is sampled from the prior distribution, $p(\theta)$.
+For $t>0$, the proposal particle is sampled from the particles accepted in the previous iteration, $\{\theta_{i,t-1}\}_{i=1}^{n_{t-1}}$, each with a probability relative to its weight $W_{t-1}(\theta_{i,t-1})$ (see below).
+The proposal particle is then perturbed using a kernel perturbation kernel, $K_t(\theta^* \mid \theta)$ where $\theta$ is the sample from the previous iteration.
+Then, a set of synthetic observations $\tilde{\vec X}^*$ is simulated, and the proposal particle $\theta^*$ is accepted if its approximate likelihood (\cref{eq:heatstress-likelihood}) is high enough, $\likelihood(\theta^*)>1-\epsilon_t$ (or more commonly, if $1-\likelihood(\theta^*) < \epsilon_t$), where $\epsilon_t>0$ is the \emph{acceptance threshold}, as higher values of $\epsilon_t$ allow more particles to be accepted.
+The acceptance threshold $\epsilon_t$ is chosen as the median of the $1-\likelihood(\theta)$ of the particles accepted in the previous iteration, $t-1$, and $\epsilon_0=0.01$. 
+For each accepted particle $\theta_{i,t}$ a weight $W_t(\theta_{i,t})$ is assigned: for $t=0$, $W_0(\theta_{i,0})=1$, and for $t>0$, 
+$W_t(\theta_{i,t}) = p(\theta_{i,t}) / \sum_{i=1}^{n_{t-1}}{W_{t-1}(\theta_{i,t-1}) K_t(\theta_{i,t}, \theta_{i,t-1})}$, where $p(\theta)$ is the prior density of $\theta$ and $K_t(\theta', \theta)$ is the probability of a perturbation from $\theta$ to $\theta'$.
+$K_t(\theta' \mid \theta)$ is a multivariate normal distribution, fitted at iteration $t$ to the particles from the previous iteration, \{$\theta_{i,t-1}\}_{i=1}{n_{t-1}}$, and their weights, $\{W(\theta_{i,t-1})\}_{i=1}^{n_{t-1}}$.
+
+Acceptance is determined according to the approximate likelihood~(\cref{eq:heatstress-likelihood}), which has a maximum value of $\likelihood_{max}=0.875$ (giving a minimal value of $\epsilon_{min}=0.125$).
+We terminated the inference iterations when the change in $\epsilon$ value from one iteration to the next was small.
+With our standard prior and model, we reached $\epsilon=0.13$ (or $\likelihood=0.87$) after six iterations, with $n_6=982$ accepted parameter vectors and effective sample size ESS=651 (\Cref{fig:convergence}).
+Running the inference algorithm with different initialization seeds and less or more simulations for approximating the likelihood produced similar posterior distributions~(\Cref{fig:seeds}).
+
+After producing a set of weighted particles from the the posterior distribution using the above ABC-SMC algorithm, we approximate the posterior using kernel density estimation (KDE) with Gaussian kernels.
+We truncate the estimated posterior to avoid positive posterior density for values with zero prior density.
+The MAP (maximum a posteriori) estimate is computed as the the maximum of the estimated joint posterior density.
+We then draw $5,000,000$ samples from the posterior distribution to compute the HDI (highest density interval) and draw $50,000$ samples to visualize the posterior distribution with histograms.
+
+\paragraph{Model comparison.} 
+We examine several versions of our evolutionary models, e.g. without aneuploidy or with increased mutation rate in aneuploid cells, as well as several different prior distributions (see below).
+To compare these, we plot posterior predictions: for each model we execute $10,000$ simulations using the MAP parameter estimates and plot the distributions of time to fixation of \eumt, one of key properties of the model likelihood.
+These plots visualize the fit of each model to the data.
+Also, for similar models we plot the marginal and joint posterior distributions of the parameters; if these are similar, we consider the models interchangeable.
+We validate this by comparing HDI (highest density interval) of posterior distributions.
+
+Where posterior plots are very similar and the number of parameters is the same, we use WAIC, or the widely applicable information criterion ~\citep{gelman2013bayesian}, defined as
+\begin{equation} \label{eq:WAIC}
+\begin{aligned}
+\mathit{WAIC}(\theta) &\ =\ 
+-2\log\mathbb{E}[\likelihood(\theta)]\ +\ 2\mathbb{V}[\log\likelihood(\theta)]
+\end{aligned}
+\end{equation}
+where $\theta$ is a parameter vector, and $\mathbb{E}[\cdot]$ and $\mathbb{V}[\cdot]$ are the expectation and variance taken over the posterior distribution, which in practice are approximated using 50,000 samples from the posterior KDE.
+We validated that upon resampling WAIC values do not significantly change and that differences in WAIC between models are preserved.
+WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy~\citep{Kass1995}.
+
+\paragraph{Prior distributions.} \label{par:prior}
+We used informative prior distributions for $w_{\anwt}=1-c+b$, $w_{\anmt}=(1+s)(1-c)+b$ and $w_{\eumt}=1+s$, which we estimated from growth curves data from mono-culture growth experiments previously reported by \citet[Figs.
+3C, 4A, and S2]{Yona2012}.
+We used \texttt{Curveball}, a method for predicting results of competition experiments from growth curve data~\citep[\href{https://curveball.yoavram.com}{curveball.yoavram.com}]{Ram2019}.
+Briefly, \texttt{Curveball} takes growth curves of two strains growing separately in mono-culture and predicts how they would grow in a mixed culture, that is, it predicts the results of a competition assay.
+From these predictions, relative fitness values can be computed.
+Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values to be used as a prior distribution by sampling 10,000 samples from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix~(\Cref{fig:growth-curves}).
+
+We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate an informative prior distribution for $w_{\anwt}$~(\Cref{fig:growth-curves}-D, assuming $w_{\euwt}=1$).
+In this prior distribution, we used the same prior for $w_{\anmt}$ and $w_{\eumt}$. 
+To increase computational efficiency, we also assumed $w_\eumt>w_\anmt>w_\anwt>w_\euwt$; running the inference without this assumption produced similar results.
+See \emph{supporting material} for an extended informative prior distribution that uses growth curves of \eumt\ and \anwt\ growing in \SI{39}{\celsius}; this prior distribution proved to be less useful.
+
+As a control, we tested an uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{\anwt}$, $w_{\anmt}$, $w_{\eumt}$, or (ii) only for $w_{\anmt}$, $w_{\eumt}$, using the above informative prior for $w_{\anwt}$.
+In these cases the inference algorithm failed to converge.
+
+For the mutation rate, $\mu$, and aneuploidy rate, $\delta$, we used uninformative uniform priors, $\mu \sim \mathit{U}(10^{-9},10^{-5})$ and $\delta \sim \mathit{U}(10^{-6},10^{-2})$.
+A wider mutation rate prior, $\mu \sim \mathit{U}(10^{-9},10^{-3})$, produced similar results.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 % Acknowledgements
@@
 This work was supported in part by 
-the Israel Science Foundation (YR 552/19),
-Minerva Stiftung Center for Lab Evolution (YR),
+the Israel Science Foundation (ISF, YR 552/19),
+the US--Israel Binational Science Foundation (BSF, YR 2021276)
+Minerva Stiftung center for Lab Evolution (YR),
 Minerva Stiftung short-term research grant (MP).
-% TODO funding for Avihu
 }
@@
 
+%TODO table:mutations
+
 \end{document}

### 2022-07-18 ms/diff.sh

@@ -1,4 +1,0 @@
-latexdiff ms_prev.tex ms.tex --no-label --enforce-auto-mbox --type=CFONT > diff.tex
-pdflatex diff.tex
-pdflatex diff.tex
-open diff.pdf

### 2022-07-18 ms/ms.tex
short par from AY with intuition

@@
 \begin{abstract}
-Aneuploidy is common in eukaryotes, often leading to decreased cell growth and fitness.
+BLA
+%Aneuploidy is common in eukaryotes, often leading to decreased cell growth and fitness.
 However, evidence from yeast and fungi, as well as human tumour cells, suggests that aneuploidy can be beneficial under stressful conditions and lead to elevated growth rates and adaptation.
-Importantly, aneuploidy differs from point mutations in rate, fitness effect, and reversibility.
-Here, we develop an evolutionary model for adaptive evolution with both aneuploidy and  mutation, and apply it within an Bayesian inference framework.
-We estimate the formation rate and fitness effect of aneuploidy and mutation in an evolutionary experiment in which \yeast adapted to heat stress.
-Our results suggest that aneuploidy is more frequent than previously estimated and that experimental observations cannot be explained without aneuploidy.
-We do not find evidence for genetic instability in aneuploid cells.
-Our results support the hypothesis that aneuploidy is a transient adaptive hypothesis: the experimental populations adapted by first acquiring a chromosome duplication; then by acquiring a beneficial mutation that likely appeared in aneuploid cells, rather than euploid cells; and finally by rapidly losing the extra chromosome to avoid paying the fitness cost of aneuploidy.
+%Importantly, aneuploidy differs from point mutations in rate, fitness effect, and reversibility.
+%Here, we develop an evolutionary model for adaptive evolution with both aneuploidy and  mutation, and apply it within an Bayesian inference framework.
+%We estimate the formation rate and fitness effect of aneuploidy and mutation in an evolutionary experiment in which \yeast adapted to heat stress.
+%Our results suggest that aneuploidy is more frequent than previously estimated and that experimental observations cannot be explained without aneuploidy.
+%We do not find evidence for genetic instability in aneuploid cells.
+%Our results support the hypothesis that aneuploidy is a transient adaptive hypothesis: the experimental populations adapted by first acquiring a chromosome duplication; then by acquiring a beneficial mutation that likely appeared in aneuploid cells, rather than euploid cells; and finally by rapidly losing the extra chromosome to avoid paying the fitness cost of aneuploidy.
 \end{abstract}
@@
 % quote: This scenario may thus imply that the duplication of a cer- tain chromosome may provide the evolutionary time window needed for the population to search for solutions not only on the duplicated chromosome but throughout the genome.
+
+% avihu: At the first stage there are more 2n cells than 2n+1 cells because of the advantage of aneuploidy under heat.
+Accordingly, at the seconds stage, there are more 2n+1* cells than 2n*.
+However, in most cases 2n*M dominate the population (and not 2n*A), because the fitness advantage of a refined mutation that emerge in 2n+1 background in hindered by the cost of aneuploidy and by the extra step needed to eliminate the extra copy of the chromosome.
 

### 2022-07-18 ms/ms.tex
continued reorganizing and revising

@@
 % Chromosomal duplication is a transient evolutionary solution to stress
-\title{Aneuploidy is an evolutionary sidetrack on the path to adaptation}
+\title{Aneuploidy may be an evolutionary sidetrack on the path to adaptation}
 
@@
 
-% Abstract
+% Abstract 250 words
 \begin{abstract} 
@@
 
+% TODO Significance statement 120 words
+
+Classifications. Biological Sciences: Evolution, Genetics, Microbiology, Population Biology.
+
+Keywords: aneuploidy, evolutionary model, adaptive evolution
+
 % TODO cite https://www.biorxiv.org/content/10.1101/2020.10.07.327759v1
@@
 
-%\paragraph*{Aneuploidy is common in eukaryotes.}
+%\paragraph{Aneuploidy is common in eukaryotes.}
 Aneuploidy is an imbalance in the number of chromosomes in the cell: an incorrect karyotype.
@@
 
-%\paragraph*{Aneuploidy is generally deleterious.}
+%\paragraph{Aneuploidy is generally deleterious.}
 %The molecular and genetic mechanisms involved in aneuploidy have been explored~\citep{Musacchio2007, Sheltzer2011, Chen2012b, Rancati2013, Gerstein2015, Shor2015}.
@@
 
-%\paragraph*{Aneuploidy can lead to adaptation.}
+%\paragraph{Aneuploidy can lead to adaptation.}
 However, aneuploidy can be beneficial under stressful conditions due to the wide range of phenotypes it can produce, some of which are advantageous~\citep{Pavelka2010, Yang2021}.
@@
 
-%\paragraph*{Transient adaptive solution.} 
+%\paragraph{Transient adaptive solution.} 
 Experimental evidence for the role of aneuploidy in adaptive evolution was demonstrated by \citet{Yona2012}. They evolved populations of \yeast under strong heat stress.
@@
 
-%\paragraph*{The present study.}
+%\paragraph{The present study.}
 Here, we test the hypothesis that aneuploidy is a \emph{transient adaptive solution}.
@@
 Second, we develop an evolutionary genetic model and fit it to the experimental results of \citet{Yona2012} in order to predict the genotype frequency dynamics in the experimental populations, thereby estimating the frequency of evolved euploid cells that descended from aneuploid cells.
-Our results show that, despite aneuploidy having reached high frequencies in the experimental populations, the majority of cells in the evolved euploid populations likely did not descend from aneuploid cells, but rather directly from wild type euploid cells.
+Our results show that, despite aneuploidy having reached high frequencies in the experimental populations, the majority of cells in the evolved euploid populations likely did not descend from aneuploid cells, but rather directly from wild-type euploid cells.
 These results suggest that aneuploidy may have been a sidetrack, rather than a stepping stone, on the path to adaptation.
@@
 \section*{Results}
-
-\paragraph{DNA sequencing.}
-
 In the heat-stress experiment of \citet{Yona2012}, four populations of \yeast evolved under \SI{39}{\celsius}.
@@
 These two repetitions continued to evolve under the same conditions, wherein aneuploidy was eliminated by generation 1,700 and 2,350.
-For each of these two populations we sequenced the ancestral population (generation zero), the aneuploid population (generation 450), and the evolved euploid population (generation 1,700 or 2,350) to estimate the mutant allele frequencies (\Cref{table:mutations}).
+
+\paragraph{Empirical frequencies of mutant alleles.} 
+Our first test of the \emph{transient adaptive solution hypothesis} employs previously unpublished sequencing data.
+For each of two evolved populations ($H2$ and $H4$) we sequenced the ancestral population (generation zero), the aneuploid population (generation 450), and the evolved euploid population (generation 1,700 or 2,350) to estimate the mutant allele frequencies (\Cref{table:mutations}).
 Overall, between 100 and 173 mutant alleles were detected with at least a single read in the six population that were sampled. 
@@
 
-Surprisingly, out of all these mutant alleles, none was present at a frequency >20\% in both the aneuploid and the evolved euploid populations. 
-Moreover, a high mutant allele frequency in the aneuploid population implies a low frequency in the evolved euploid population, and vice-versa (Spearman's correlation coefficient $\rho=-0.64$ and $-0.66$ in the two experimental repetitions; \Cref{fig:mutation_frequencies}).
-Specifically, for the 18 mutant alleles with high frequency in the aneuploid populations (>20\%), the highest frequencies in the euploid populations were 15.4\%, 16\%, 16.3\% and 19.6\% (the rest were below 15\%).
-Similarly, for the 48 mutant alleles with high frequency in the evolved euploid populations, the highest frequencies in the aneuploid populations were 2.7\%, 7.7\%, and 11.1\% (the rest were below 1\%).
-Thus, mutant alleles frequent in the aneuploid populations decreased in frequency when aneuploidy was lost, and most evolved euploid cells are likely not descendent from aneuploid~cells.
-
 %%%%%%%% Fig mutation frequencies
-\begin{figure}[h!]
+\begin{figure}[h]
   \centering
@@
 
+Surprisingly, out of all these mutant alleles, none was present at a frequency >20\% in both the aneuploid and the evolved euploid populations. 
+Furthermore, a high mutant allele frequency in the aneuploid population implies a low frequency in the evolved euploid population, and vice-versa (Spearman's correlation coefficient $\rho=-0.64$ and $-0.66$ in the two experimental repetitions; \Cref{fig:mutation_frequencies}), such that mutant alleles frequent in the aneuploid populations decreased in frequency when aneuploidy was lost.
+Moreover, for the 18 mutant alleles with high frequency in the aneuploid populations (>20\%), the highest frequencies in the euploid populations were 15.4\%, 16\%, 16.3\% and 19.6\% (the rest were below 15\%).
+Similarly, for the 48 mutant alleles with high frequency in the evolved euploid populations, the highest frequencies in the aneuploid populations were 2.7\%, 7.7\%, and 11.1\% (the rest were below 1\%).
+These results suggest evolved euploid cells are unlikely to descend from aneuploid~cells.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \paragraph{Evolutionary genetic model.} 
-% TODO continue here
+To further test the \emph{transient adaptive solution hypothesis}, we developed an evolutionary genetic model, fitted the model to empirical data, and used it to predict the genotype frequency dynamics, or specifically, the fraction of the evolved euploid population descended from aneuploid cells.
 
-We develop an evolutionary genetic model that that includes the effects of natural selection, genetic drift, aneuploidy, and mutation to examine the role of aneuploidy in adaptive evolution.
-The model follows a population of cells characterised by both their ploidy and their genotype, fit this model to the experimental results of \citet{Yona2012} using an \emph{approximate Bayesian computation} framework~\citep{Sisson2009} to infer model parameters, including rates and fitness effects of aneuploidy and mutation, and to test hypotheses about the evolutionary process by comparing different versions of the model and predicting the genotype frequency dynamics using the estimated parameter values.
+The model includes the effects of natural selection, genetic drift, aneuploidy, and mutation, and follows a population of cells characterized by both their genotype: 
+euploid wild-type, \euwt, is the ancestral diploid genotype; 
+euploid mutant, \eumt, has a diploid karyotype and a single beneficial mutation; 
+aneuploid wild-type, \anwt, has an extra chromosome due to a chromosome duplication event; and
+aneuploid mutant, \anmt, has and extra chromosome and a beneficial mutation.
+Fitness values of the different genotypes are denoted by $w_{\euwt}$, $w_{\eumt}$, $w_{\anwt}$, and $w_{\anmt}$, and the rate of mutation and aneuploidy are denoted by $\mu$ and $\delta$.
+See \Cref{fig:models} for an illustration of the model.
 
-%%%%%%%% Fig illustation
-\begin{figure}[b!]
+We fitted this model to the experimental results\citep{Yona2012} -- time for fixation and for loss of aneuploidy -- using approximate Bayesian computation with sequential Monte Carol, or ABC-SMC~\citep{Sisson2009}, thereby inferring the model parameters: rates and fitness effects of aneuploidy and mutation.
+We then sampled posterior predictions for the genotype frequency dynamics using the estimated parameter values and compared different versions of the model to test additional hypotheses about the evolutionary process.
+
+%%%%%%%% Fig illustration
+\begin{figure}[h]
   \centering
@@
 
-\paragraph{Rates and fitness effects of aneuploidy and mutation.} 
-We used ABC-SMC to infer the posterior distribution of model parameters~(\Cref{fig:posterior}). 
+\paragraph{Estimated rates and fitness effects of aneuploidy and mutation.} 
+We inferred the posterior distribution of model parameters~(\Cref{fig:posterior}). 
 We report parameter estimates using the MAP (maximum a posteriori) and providing the 50\% HDI (highest density interval) in square brackets. See \emph{Supporting material} for sensitivity analysis.
@@
 %% generated with posterior-plot.ipynb
-\begin{figure}[h!]
+\begin{figure}[h]
   \centering
@@
 
-\paragraph{Role of aneuploidy in the evolutionary dynamics.}
+\paragraph{Model comparison and goodness-of-fit.}
 
@@
 
-We also checked a model in which aneuploidy occurs but is adaptively neutral compared to the wild type, that is, $w_{\anwt} = w_{\euwt}$ and $w_{\anmt} = w_{\eumt}$ but $\delta>0$.
+We also checked a model in which aneuploidy occurs but is adaptively neutral compared to the wild-type, that is, $w_{\anwt} = w_{\euwt}$ and $w_{\anmt} = w_{\eumt}$ but $\delta>0$.
 This model fits the data better than the model with no aneuploidy (in which $\delta=0$), but worse than a model with positive selection for aneuploidy, in which $w_{\euwt}<w_{\anwt}<w_{\anmt}<w_{\eumt}$~(\Cref{fig:2n*-fixation}).
-Thus, our results suggest that aneuploidy significantly affects the evolutionary dynamics.
 
@@
 
-% Rancati G, Pavelka N.
-Karyotypic changes as drivers and catalyzers of cellular evolvability: A perspective from non-pathogenic yeasts.
-Semin Cell Dev Biol.
-2013;24(4):332-338.
-doi:10.1016/j.semcdb.2013.01.009
-% Bouchonville K, Forche A, Tang KES, Semple C a M, Berman J.
-Aneuploid chromosomes are highly unstable during DNA transformation of Candida albicans.
-Eukaryot Cell.
-2009;8(10):1554-1566.
-doi:10.1128/EC.00209-09
-% Zhu J, Pavelka N, Bradford WD, Rancati G, Li R. Karyotypic determinants of chromosome instability in aneuploid budding yeast. PLoS Genet. 2012;8(5). doi:10.1371/journal.pgen.1002719
-
-\paragraph*{Genetic instability in aneuploid cells.}
-It has been suggested that aneuploidy increased genetic instability~\citep{Sheltzer2011b}.
-Therefore, we inferred model parameters under the assumption that the mutation rate increases in aneuploid cells by a factor $\tau$ = 1, 33/32 (due to an additional chromosome), 2, 5, 10, or 100 (due to genetic instability).
-We found that the posterior distribution was similar for $\tau=$1, 33/32, 2, and 5 (\Cref{fig:tau}).
-%% generated with MAP-and-hdi.ipynb
-With $\tau=100$, the estimated mutation rate was about 7-8-fold lower compared to $\tau=1$ ($\mu=4.094\cdot10^{-7}\ [6.252\cdot10^{-8}-6.046\cdot10^{-7}]$) and the aneuploidy rate was about 2-3-fold lower ($\delta=0.744\cdot10^{-3}\ [0.506\cdot10^{-3}-1.827\cdot10^{-3}]$).
-With $\tau=10$, the estimated mutation rate was only slightly lower compared to $\tau=1$ ($\mu=1.67\cdot10^{-6}\ [2.836\cdot10^{-8}-2.245\cdot10^{-6}]$). 
-WAIC is lowest for $\tau=33/32$ and $\tau=1$.
-Therefore, evidence does not support an increase in mutation rate in aneuploid cells, and moreover, unless the increase is strong ($\tau \ge 10$), it does not seem to affect our inference.
-% TODO Avihu - in the literature it is suggested elevated mut rate is due to stress; here wt is also stressed so maybe that's why we dont see an effect
-We also checked the differences in genotype frequency dynamics for different $\tau$ values.
-We observe $\tau=100$ could be distinguished if accurate data was available for the waiting time until the frequency of \euwt\ to decrease below 95\%~(\Cref{fig:tau-plots}A) or for waiting time for the frequency of \anwt\ to either reach or go below 95\%~(\Cref{fig:tau-plots}B).
-
-\paragraph*{Competition between mutation and aneuploidy.}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\paragraph{Model predictions of genotype frequency dynamics.}
 
@@
 \Cref{fig:ppc-plot}A shows the simulated frequencies of the four genotypes (\euwt, \anwt, \anmt\ and \eumt), as well as the frequencies of \eumt\ cells that arose from either \anwt\ cells via a sequences of mutation and chromosome loss events ($2n^*_A$), or directly from \euwt\ cells via a mutation event ($2n^*_M$).
-We find that \anmt\ never reaches substantial frequency as it is quickly replaced by \eumt.
+We find that \anmt\ never reaches substantial frequency as it is quickly replaced by \eumt\ in a process similar to \emph{stochastic tunneling}~\citep{Iwasa2004,Komarova2003}.
 
-\citet{Yona2012} have suggested that aneuploidy is a transient adaptive solution, arising quickly due to increased fitness and high formation rate, but 
-ultimately lost due to the cost of aneuploidy as beneficial mutations take over.
-We tested this hypothesis by estimating $F_A$, the expected frequency of \eumt\ that arose from \anwt, as the average frequency of such cells at the end of simulations using the MAP estimate parameters.
+We tested the \emph{transient adaptive solution hypothesis} by estimating $F_A$, the expected frequency of \eumt\ that arose from \anwt, computed as the average frequency of such cells at the end of simulations using the MAP estimate parameters.
 Surprisingly, we observe that the majority of \eumt\ cells are a product of a direct mutation in \euwt\ cells, rather than descending from \anwt\ cells ($F_A^{MAP} = 0.106$, \Cref{fig:ppc-plot}A). 
@@
 The mean $F_A$ was just 0.1673 [0.0154-0.370 95\% CI] and only in 489 of 100,000 posterior samples (0.489\%) $F_A$ was  larger than 0.5, that is, \eumt\ was more likely to arise from an aneuploid cell.
-Thus, if we pick up a cell from the evolved \eumt\ population, it is more likely to have descended directly from an euploid cell than from an aneuploid cell.
+Thus, if we sample a random cell from the evolved \eumt\ population, it is more likely to have descended directly from an euploid cell than from an aneuploid cell.
 The probability of \eumt\ descending from \anwt\ ($F_A$) increases with the aneuploidy rate, $\delta$, and decreases with the mutation rate, $\mu$, and in some cases can also be affected by the fitness parameters (\Cref{fig:ppc-plot}C and \Cref{fig:posterior_FA_supp}).
 
-%TODO revise AY text:
-At the first stage there are more 2n cells than 2n+1 cells because of the advantage of aneuploidy under heat.
-Accordingly, at the seconds stage, there are more 2n+1* cells than 2n*.
-However, in most cases 2n*M dominate the population (and not 2n*A), because the fitness advantage of a refined mutation that emerge in 2n+1 background in hindered by the cost of aneuploidy and by the extra step needed to eliminate the extra copy of the chromosome.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
   \caption{
-    \textbf{Posterior genotype frequency dynamics.}     
+    \textbf{Simulated genotype frequency dynamics.}     
     \textbf{(A)} Posterior predicted genotype frequencies over time, including the source of \eumt: $2n^*_A$ arose from $2n+1$, whereas $2n^*_M$ arose directly from $2n$.
@@
 \end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-%\pagebreak
+% Rancati G, Pavelka N.
+Karyotypic changes as drivers and catalyzers of cellular evolvability: A perspective from non-pathogenic yeasts.
+Semin Cell Dev Biol.
+2013;24(4):332-338.
+doi:10.1016/j.semcdb.2013.01.009
+% Bouchonville K, Forche A, Tang KES, Semple C a M, Berman J.
+Aneuploid chromosomes are highly unstable during DNA transformation of Candida albicans.
+Eukaryot Cell.
+2009;8(10):1554-1566.
+doi:10.1128/EC.00209-09
+% Zhu J, Pavelka N, Bradford WD, Rancati G, Li R. Karyotypic determinants of chromosome instability in aneuploid budding yeast. PLoS Genet. 2012;8(5). doi:10.1371/journal.pgen.1002719
 
-%%%%%%%%%%
+\paragraph{Genetic instability in aneuploid cells.}
+It has been suggested that aneuploidy increased genetic instability~\citep{Sheltzer2011b}.
+Therefore, we inferred model parameters under the assumption that the mutation rate increases in aneuploid cells by a factor $\tau$ = 1, 33/32 (due to an additional chromosome), 2, 5, 10, or 100 (due to genetic instability).
+We found that the posterior distribution was similar for $\tau=$1, 33/32, 2, and 5 (\Cref{fig:tau}).
+%% generated with MAP-and-hdi.ipynb
+With $\tau=100$, the estimated mutation rate was about 7-8-fold lower compared to $\tau=1$ ($\mu=4.094\cdot10^{-7}\ [6.252\cdot10^{-8}-6.046\cdot10^{-7}]$) and the aneuploidy rate was about 2-3-fold lower ($\delta=0.744\cdot10^{-3}\ [0.506\cdot10^{-3}-1.827\cdot10^{-3}]$).
+With $\tau=10$, the estimated mutation rate was only slightly lower compared to $\tau=1$ ($\mu=1.67\cdot10^{-6}\ [2.836\cdot10^{-8}-2.245\cdot10^{-6}]$). 
+WAIC (lower is better, see Methods) is lowest for $\tau=33/32$ and $\tau=1$ (\Cref{table:WAIC}).
+Therefore, evidence does not support an increase in mutation rate in aneuploid cells, and moreover, unless the increase is strong ($\tau \ge 10$), it does not seem to affect our inference.
+We also checked the differences in genotype frequency dynamics for different $\tau$ values.
+We observe $\tau=100$ could be distinguished if accurate data was available for the waiting time until the frequency of \euwt\ to decrease below 95\%~(\Cref{fig:tau-plots}A) or for waiting time for the frequency of \anwt\ to either reach or go below 95\%~(\Cref{fig:tau-plots}B).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Discussion
@@
 
-% quote: This scenario may thus imply that the duplication of a cer- tain chromosome may provide the evolutionary time window needed for the population to search for solutions not only on the duplicated chromosome but throughout the genome.
+\citet{Yona2012} have suggested that the aneuploid cells that come to dominate the experimental populations were later replaced by ``refined strains, which eliminated the cost associated with the trisomy, exhibit improved growth both at the permissive and the high temperatures compared with their trisomic ancestors'' and that ``it is expected that the original contribution of trisomy to heat tolerance was replaced by alternative solutions.''.
+However, our results suggest that most of the cells in the evolved euploid population do not carry mutations that were common in the aneuploid populations, but rather a different set of mutations.
+Therefore, the majority of the evolved euploid cells are not descended from aneuploid cells, but rather are direct descendants of the ancestral population.
+
+% quote: This scenario may thus imply that the duplication of a certain chromosome may provide the evolutionary time window needed for the population to search for solutions not only on the duplicated chromosome but throughout the genome.
 
@@
 
-\citet{Yona2012} have suggested that the aneuploid cells that come to dominate the experimental populations were later replaced by ``refined strains, which eliminated the cost associated with the trisomy, exhibit improved growth both at the permissive and the high temperatures compared with their trisomic ancestors'' and that ``it is expected that the original contribution of trisomy to heat tolerance was replaced by alternative solutions.''.
-However, our results suggest that most of the cells in the evolved euploid population do not carry mutations that were common in the aneuploid populations, but rather a different set of mutations.
-Therefore, the majority of the evolved euploid cells are not descended from aneuploid cells, but rather are direct descendants of the ancestral population.
-
-\paragraph*{Results of current study.}
-We inferred the rates and fitness effects of aneuploidy and mutation in a previously published evolutionary experiment in which yeast populations adapted to heat stress~\citep{Yona2012}. 
-First, we estimate that the aneuploidy rate (i.e., number of chromosome gains per generation) is $1.7 \cdot 10^{-3}$, is higher than a previous estimate of $6.7 \cdot 10^{-6}$~\citep{Zhu2016}.
-% TODO explain?
-Second, the experiments results cannot be explained without aneuploidy ($\delta=0$) or if aneuploidy is neutral ($w_{\anwt}=w_{\euwt}$, $w_{\anmt}=w_{\eumt}$), and therefore we conclude that aneuploidy plays a significant role in the dynamics.
-Third, we find no evidence for increased mutation rates in aneuploid cells, but more data would have allowed a more conclusive statement.
-Therefore, the effect of aneuploidy on genetic instability remains to be explored in future work.
-Fourth, we simulated genotype frequency dynamics using inferred rates and fitness effects of aneuploidy and mutation.
-We found that the aneuploid mutant, \anmt, never reaches high frequency, in a process similar to \emph{stochastic tunneling}~\citep{Iwasa2004}.
-% TODO ref komarova & nowak
-Furthermore, we find that XXXX. % TODO transient solution?
-
 \cite{Yona2012} have analyzed the expression levels of 14 heat-shock proteins (HSP).
@@
 
-\paragraph*{Aneuploidy is not just another type of mutation.}
-The published data indicate that, like mutation, aneuploidy can be both deleterious and beneficial~\citep{Pavelka2010, Sheltzer2011}.
-Nevertheless, there are important and fundamental differences between adaptation by aneuploidy
-and adaptation by beneficial mutations~\citep{Yona2015}, which make aneuploidy a unique mechanism for generating genetic
-variation.
-First, the aneuploidy rate (i.e. the frequency of mis-segregation events) is significantly higher than the
-mutation rate~\citep{Santaguida2015review}.
-Thus, everything else being equal, adaptation by aneuploidy will be faster and more frequent.
-Second, fitness effects of aneuploidy are larger than those of the majority of mutations, on average, and are rarely
-neutral~\citep{Pavelka2010, Yona2012, Sunshine2015}, allowing selection to quickly sort deleterious and beneficial genotypes.
-Third, the number of different karyotypes is considerably smaller than the number of different genotypes, and different karyotypes are likely to have different phenotypes~\citep{Pavelka2010}.
-Therefore, exploration of the phenotype space by aneuploidy requires smaller populations and a shorter time span.
-Fourth, aneuploidy is a reversible state, as the rate of chromosome loss is high and the cost of aneuploidy is significant~\citep{Niwa2006}.
-Indeed, aneuploidy often provides a transient solution: under short-term stress conditions, aneuploidy reverts (chromosome number returns to normal) when the stress subsides; under long-term stress conditions, aneuploidy reverts when refined solutions, generated by beneficial mutations, take over~\citep{Yona2012}.
-Finally, aneuploidy results in increased genome instability, potentially increasing genetic variation by a positive feedback loop~\citep{Rancati2013, Bouchonville2009, Zhu2012}, while also increasing its own transience.
+%\paragraph{Rate and fitness effect of aneuploidy and mutation.}
+We inferred the rates and fitness effects of aneuploidy and mutation in a previously published evolutionary experiment in which yeast populations adapted to heat stress~\citep{Yona2012}. 
+We estimate that the aneuploidy rate (i.e., number of chromosome gains per generation) is $1.7 \cdot 10^{-3}$, is higher than a previous estimate of $6.7 \cdot 10^{-6}$~\citep{Zhu2016}.
+% TODO explain?
+We find no evidence for increased mutation rates in aneuploid cells. 
+% TODO Avihu - in the literature it is suggested elevated mut rate is due to stress; here wt is also stressed so maybe that's why we dont see an effect
 
-% TODO do we want the following two paragraphs?
-\paragraph*{Evolutionary theory of aneuploidy.}
-The role of aneuploidy in adaptation has only recently been observed~\citep{Sionov2010, Yona2012, Gerstein2015}, and is largely missing from the literature on evolution and adaptation:
-the introductory textbook \emph{Evolution} by~\citet{Bergstrom2012} does not mention the word aneuploidy, and the graduate-level book \emph{Mutation-Driven Evolution} by~\citet{Nei2013} only briefly mentions aneuploidy in the context of speciation, but not adaptation.
-In recent reviews of the literature, aneuploidy is suggested to play an important role in fungal adaptation~\citep{Robbins2017, Todd2017} and cancer evolution~\citep{Santaguida2015review, Naylor2016, Sansregret2017}, yet these reviews cite no theoretical studies nor any quantitative models.
-Indeed, evolutionary, ecological, and epidemiological studies mostly assume adaptation occurs via beneficial mutations, recombination, and sex.
-Therefore, there is a critical need to develop an evolutionary theory of aneuploidy like the evolutionary theories of other mechanisms for generation of genetic variation, e.g.
-mutation~\citep{Lynch2010}, recombination~\citep{Hartfield2012}, and sex~\citep{Otto2009}.
-An evolutionary theory of aneuploidy will be central to the interpretation of experimental and clinical observations and design of new hypotheses, experiments, and treatments~\citep{Carja2014}.
-For example, despite the lack of theoretical models, aneuploidy has been invoked in a new strategy to combat pathogens and tumour cells by setting ``evolutionary traps''~\citep{Gerstein2015,Chen2015}, in which a condition that predictably leads to emergence of aneuploidy is applied, followed by a condition that specifically selects against aneuploid cells.
+\paragraph{Conclusions.}
+% TODO YR
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph*{Evolutionary Model.}
+\paragraph{DNA sequencing.}
+% TODO AY
+
+\paragraph{Evolutionary genetic model.}
 We model the evolution of a population of cells using a Wright-Fisher model~\citep{Otto2007}, assuming a constant effective population size $N$, non-overlapping generations, and including the effects of natural selection, genetic drift, aneuploidy, and mutation.
@@
 
-\paragraph{Empirical evidence.}
+\paragraph{Empirical data for model inference.}
 
@@
 
-\paragraph*{Sensitivity analysis.} 
+\paragraph{Sensitivity analysis.} 
 Changing a single parameter while keeping the rest fixed at the MAP estimate produces a worse fit to the data~(\Cref{fig:sensitivity}).
@@
 
-\paragraph*{Extended informative prior distribution.}
+\paragraph{Extended informative prior distribution.}
 In an extended informative prior distribution, we used additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{\eumt}/w_{\anwt}$~(\Cref{fig:growth-curves}L).

### 2022-07-20 ms/ms.tex
finished draft of discussion I think

@@
 \newcommand{\eumt}{\emph{2n*}}
+\newcommand{\eumtM}{\emph{$2n^*_M$}}
+\newcommand{\eumtA}{\emph{$2n^*_A$}}
 \newcommand{\anmt}{\emph{2n+1*}}
@@
 Alleles with frequency below and above 10\% in the ancestral populations are in black and gray, respectively. 
-Solid black line is a linear orthogonal distance regression line (slope=$-0.559$, intercept=$0.164$).
+Solid black line is a linear orthogonal distance regression line (slope=$-0.559$, intercept=$0.164$; a regression through alleles that reach at least 20\% in one of the populations has slope=$-0.645$ and intercept=$0.297$).
 Dashed vertical and horizontal lines show allele frequencies 20\%. Error bars show standard error of the mean accounting for the number of reads.
@@
 We simulated 50 replicate genotype frequency dynamics using the MAP estimate parameters. 
-\Cref{fig:ppc-plot}A shows the simulated frequencies of the four genotypes (\euwt, \anwt, \anmt\ and \eumt), as well as the frequencies of \eumt\ cells that arose from either \anwt\ cells via a sequences of mutation and chromosome loss events ($2n^*_A$), or directly from \euwt\ cells via a mutation event ($2n^*_M$).
+\Cref{fig:ppc-plot}A shows the simulated frequencies of the four genotypes (\euwt, \anwt, \anmt\ and \eumt), as well as the frequencies of \eumt\ cells that arose from either \anwt\ cells via a sequences of mutation and chromosome loss events ($2n^*_A$), or directly from \euwt\ cells via a mutation event (\eumtM).
 We find that \anmt\ never reaches substantial frequency as it is quickly replaced by \eumt\ in a process similar to \emph{stochastic tunneling}~\citep{Iwasa2004,Komarova2003}.
@@
 
-\citet{Yona2012} have suggested that the aneuploid cells that come to dominate the experimental populations were later replaced by ``refined strains, which eliminated the cost associated with the trisomy, exhibit improved growth both at the permissive and the high temperatures compared with their trisomic ancestors'' and that ``it is expected that the original contribution of trisomy to heat tolerance was replaced by alternative solutions.''.
-However, our results suggest that most of the cells in the evolved euploid population do not carry mutations that were common in the aneuploid populations, but rather a different set of mutations.
-Therefore, the majority of the evolved euploid cells are not descended from aneuploid cells, but rather are direct descendants of the ancestral population.
+In a landmark study on the role of chromosome duplication in adaptive evolution, 
+\citet{Yona2012} found that a chromosome III trisomy was acquired by \yeast populations evolving under heat stress, only to be later replaced by euploid mutant cells.
+Moreover, when they initiated evolutionary experiments with a population in which all cells carry a trisomy, such a replacement occurred within 1,000 generations.
+
+They hypothesized that these evolved euploid were a consequence of (i) mutations in genes that confer heat resistance, followed by (ii) reversion of trisomy by a chromosome loss event. 
+If indeed the evolved euploid population is descended from the aneuploid population, then mutant alleles that were common in the aneuploid populations should also be common in the evolved euploid population.
+
+However, we found that this is not the case (\Cref{fig:mutation_frequencies}): mutant allele frequencies in the aneuploid and euploid populations are negatively correlated, and common alleles in the former are rare in the later.
+Furthermore, we developed an evolutionary genetic model of adaptive evolution by aneuploidy and mutation, fitted it to the experimental results of \citet{Yona2012}, and used it to predict the genotype frequency dynamics.
+The model predicted that only about 10\% of the evolved euploid cells descended from aneuploid cells--that is, the majority of the evolved euploid cells are not descended from aneuploid cells, but rather are direct descendants of the ancestral population.
+This is despite aneuploidy reaching a very high frequency in the population of >95\% (\Cref{fig:ppc-plot}).
+
+This is not an intuitive result, as conventional evolutionary reason would be that once the aneuploid genotype reaches high frequency, it will have a better chance at producing "refined" solutions via mutations.
+We therefore suggest a different explanation. 
+At the first stage, \anwt\ cells appear and reach a high frequency due to their advantage under heat.
+Accordingly, \anmt\ cells appear more frequently than \eumtM\ (which arises directly from \euwt).
+However, in most cases \eumtM\ dominates the population (and not \eumtA, which arises from \anmt), because the fitness advantage of a "refined" mutation that emerges in the \anmt\ genotype in hindered by the cost of aneuploidy as well as the extra waiting time needed to eliminate the extra copy of the chromosome by a chromosome loss event.
+Although such chromosome loss occurs fast relative to mutation, it still leaves a time window for \eumtM\ to arise and reach high frequencies.
 
@@
 
-% avihu: At the first stage there are more 2n cells than 2n+1 cells because of the advantage of aneuploidy under heat.
-Accordingly, at the seconds stage, there are more 2n+1* cells than 2n*.
-However, in most cases 2n*M dominate the population (and not 2n*A), because the fitness advantage of a refined mutation that emerge in 2n+1 background in hindered by the cost of aneuploidy and by the extra step needed to eliminate the extra copy of the chromosome.
-
-\cite{Yona2012} have analyzed the expression levels of 14 heat-shock proteins (HSP).
+%\citet{Yona2012} have analyzed the expression levels of 14 heat-shock proteins (HSP).
 They found that their expression level changed ``very modestly'' in the aneuploid populations, but that most of them ``showed significant up-regulation'' in the evolved euploid population.
 
-%\paragraph{Rate and fitness effect of aneuploidy and mutation.}
+\paragraph{Rate and fitness effect of aneuploidy and mutation.}
 We inferred the rates and fitness effects of aneuploidy and mutation in a previously published evolutionary experiment in which yeast populations adapted to heat stress~\citep{Yona2012}. 
@@
 \paragraph{Conclusions.}
-% TODO YR
+BLA BLA BLA
 
@@
 \paragraph{DNA sequencing.}
+BLA BLA BLA
 % TODO AY

### 2022-07-24 ms/ms.tex
collected F_A plots to one figure and formatted it

@@
 \begin{figure}[h!]
-  \begin{subfigure}{\textwidth}
       \centering
-      %% generated with dynamics-plot.ipynb
-      \includegraphics[width=\textwidth]{../figures/MAP_freq_dynamics.pdf}      
-  \end{subfigure}
-  \begin{subfigure}{0.5\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../figures/F_A_posterior.pdf}
-  \end{subfigure}
-  \begin{subfigure}{0.5\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../figures/F_A_mu_delta.pdf}
-  \end{subfigure}
+      \includegraphics[width=\textwidth]{../figures/FA.pdf}      
   \caption{
@@
 	White star shows the MAP estimate. See also \Cref{fig:posterior_FA_supp}.
+	\textbf{(D)} $F_A$ as a function of the population size, $N$, in posterior predictions with MAP parameters.
+Markers show $F_A$ in 250 simulations per population size.
+Error bars show mean $F_A$ with 95\% CI (bootstrap, $n=10,000$).
+Vertical dashed line for population size in the experiment, $6.425\cdot10^6$.
+Horizontal line for $F_A^{MAP}=0.106$.
      }
@@
 In a landmark study on the role of chromosome duplication in adaptive evolution, 
-\citet{Yona2012} found that a chromosome III trisomy was acquired by \yeast populations evolving under heat stress, only to be later replaced by euploid mutant cells.
-Moreover, when they initiated evolutionary experiments with a population in which all cells carry a trisomy, such a replacement occurred within 1,000 generations.
+\citet{Yona2012} found that a chromosome III trisomy was acquired by \yeast populations evolving under heat stress, only to be later replaced by euploid mutant cells that carry "refined" solutions to the stress.
+Additionally, such a replacement also occurred when they initiated evolutionary experiments with a population in which all cells carry a chromosome III trisomy.
+They hypothesized that these evolved euploid cells were a consequence of mutations in genes that confer heat resistance, followed by reversion of trisomy by a chromosome loss event.
 
-They hypothesized that these evolved euploid were a consequence of (i) mutations in genes that confer heat resistance, followed by (ii) reversion of trisomy by a chromosome loss event. 
 If indeed the evolved euploid population is descended from the aneuploid population, then mutant alleles that were common in the aneuploid populations should also be common in the evolved euploid population.
+However, we found that this is not the case (\Cref{fig:mutation_frequencies}): mutant allele frequencies in the aneuploid and euploid populations are negatively correlated, such that common alleles in the former are rare in the later.
+Furthermore, we developed an evolutionary genetic model of adaptive evolution by aneuploidy and mutation (\Cref{fig:models}), fitted it to the experimental results of \citet{Yona2012}, and used it to predict the genotype frequency dynamics.
+The model predicted that only about 10\% of the evolved euploid population descended from aneuploid cells--that is, the majority of the euploid population are not descended from aneuploid cells, but rather are direct descendants of the ancestral population (\Cref{fig:ppc-plot}).
 
-However, we found that this is not the case (\Cref{fig:mutation_frequencies}): mutant allele frequencies in the aneuploid and euploid populations are negatively correlated, and common alleles in the former are rare in the later.
-Furthermore, we developed an evolutionary genetic model of adaptive evolution by aneuploidy and mutation, fitted it to the experimental results of \citet{Yona2012}, and used it to predict the genotype frequency dynamics.
-The model predicted that only about 10\% of the evolved euploid cells descended from aneuploid cells--that is, the majority of the evolved euploid cells are not descended from aneuploid cells, but rather are direct descendants of the ancestral population.
-This is despite aneuploidy reaching a very high frequency in the population of >95\% (\Cref{fig:ppc-plot}).
+This happens despite aneuploidy reaching a high frequency in the population (>95\%). 
+Conventional wisdom might suggest that once the aneuploid genotype \anwt\ reaches high frequency, it will have a better chance at producing "refined" solutions via mutations, and its descendants will come to dominate the population: the frequency \eumtA\ (which arises from \anmt) will be higher than the frequency of \eumtM\ (which arises directly from \euwt).
+So how does \eumtM\ prevail?
+At the first stage, \anwt\ cells appear and reach a high frequency due to their advantage under heat.
+Accordingly, \anmt\ cells appear more frequently than \eumtM.
+However, the fitness advantage of a "refined" mutation that emerges in the \anmt\ genotype in hindered by the cost of aneuploidy as well as the extra waiting time needed to eliminate the extra copy of the chromosome by a chromosome loss event.
+Although such chromosome loss occurs fast relative to mutation, it still leaves a time window for \eumtM\ to arise and reach high frequencies.
+Indeed, we estimate that the advantage of \anmt\ over \anwt\ is similar to the advantage of \eumt\ over \anmt (about 0.3\%), and that the rate of chromosome loss is roughly $1-3\cdot10^{-3}$.
 
-This is not an intuitive result, as conventional evolutionary reason would be that once the aneuploid genotype reaches high frequency, it will have a better chance at producing "refined" solutions via mutations.
-We therefore suggest a different explanation. 
-At the first stage, \anwt\ cells appear and reach a high frequency due to their advantage under heat.
-Accordingly, \anmt\ cells appear more frequently than \eumtM\ (which arises directly from \euwt).
-However, in most cases \eumtM\ dominates the population (and not \eumtA, which arises from \anmt), because the fitness advantage of a "refined" mutation that emerges in the \anmt\ genotype in hindered by the cost of aneuploidy as well as the extra waiting time needed to eliminate the extra copy of the chromosome by a chromosome loss event.
-Although such chromosome loss occurs fast relative to mutation, it still leaves a time window for \eumtM\ to arise and reach high frequencies.
-
-% quote: This scenario may thus imply that the duplication of a certain chromosome may provide the evolutionary time window needed for the population to search for solutions not only on the duplicated chromosome but throughout the genome.
-
-%\citet{Yona2012} have analyzed the expression levels of 14 heat-shock proteins (HSP).
-They found that their expression level changed ``very modestly'' in the aneuploid populations, but that most of them ``showed significant up-regulation'' in the evolved euploid population.
+% TODO effect of population size?
 
@@
 % TODO explain?
-We find no evidence for increased mutation rates in aneuploid cells. 
-% TODO Avihu - in the literature it is suggested elevated mut rate is due to stress; here wt is also stressed so maybe that's why we dont see an effect
+In addition, we find no evidence for increased mutation rates in aneuploid cells. 
+Previous empirical studies have suggested that genetic instability (e.g., elevated mutation rates) in aneuploid cells is due to stress associated with the aneuploid state~\citep{Bouchonville2009,Chen2012b,Zhu2012}.
+However, in the experiment of \citet{Yona2012}, both the wild-type and the aneuploid were under heat stress, which may explain why we did not find evidence for an increased mutation rate.
 
 \paragraph{Conclusions.}
-BLA BLA BLA
+Here, we tested the hypothesis that aneuploid cells are an evolutionary adaptive intermediate between wild-type euploid cells and mutant euploid cells. 
+Our results suggest this is only true at the population level, as the population does go from euploid to aneuploid and back--but not at the individual level: we estimate that only about 10-15\% of the euploid cells descended from aneuploid cells, whereas the rest are direct descendent of the wild-type euploid cells.
+
+% TODO finish conclusions
 

### 2022-07-26 ms/ms.tex
revised FA with deterministic dynamics, log-log in supp

@@
 %% estimates generated by MAP-and-hdi.ipynb
-The estimated mutation rate, $\mu=2.965\cdot10^{-6}\ [2.718\cdot10^{-7}-3.589\cdot10^{-6}]$, corresponds to a mutation target size of $\sim 10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
+The estimated mutation rate, $\mu=2.965\cdot10^{-6}\ [2.718\cdot10^{-7}-3.589\cdot10^{-6}]$, corresponds to a mutation target size of $\sim 10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$ (ref.~\citep{Zhu2014}) or $3.3\cdot10^{-10}$ (ref.~\citep{Lynch2008}).
 The estimated aneuploidy rate, $\delta=1.72\cdot10^{-3}\ [1.47\cdot10^{-3}-2.786\cdot10^{-3}]$ is much higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
@@
 We simulated 50 replicate genotype frequency dynamics using the MAP estimate parameters. 
-\Cref{fig:ppc-plot}A shows the simulated frequencies of the four genotypes (\euwt, \anwt, \anmt\ and \eumt), as well as the frequencies of \eumt\ cells that arose from either \anwt\ cells via a sequences of mutation and chromosome loss events ($2n^*_A$), or directly from \euwt\ cells via a mutation event (\eumtM).
+\Cref{fig:FA}A shows the simulated frequencies of the four genotypes (\euwt, \anwt, \anmt\ and \eumt), as well as the frequencies of \eumt\ cells that arose from either \anwt\ cells via a sequences of mutation and chromosome loss events ($2n^*_A$), or directly from \euwt\ cells via a mutation event (\eumtM).
 We find that \anmt\ never reaches substantial frequency as it is quickly replaced by \eumt\ in a process similar to \emph{stochastic tunneling}~\citep{Iwasa2004,Komarova2003}.
@@
 We tested the \emph{transient adaptive solution hypothesis} by estimating $F_A$, the expected frequency of \eumt\ that arose from \anwt, computed as the average frequency of such cells at the end of simulations using the MAP estimate parameters.
-Surprisingly, we observe that the majority of \eumt\ cells are a product of a direct mutation in \euwt\ cells, rather than descending from \anwt\ cells ($F_A^{MAP} = 0.106$, \Cref{fig:ppc-plot}A). 
-This is despite the fact that the \anwt\ genotype reaches high frequencies in the population (at least 0.98, \Cref{fig:ppc-plot}A).
+Surprisingly, we observe that the majority of \eumt\ cells are a product of a direct mutation in \euwt\ cells, rather than descending from \anwt\ cells ($F_A^{MAP} = 0.106$, \Cref{fig:FA}A). 
+This is despite the fact that the \anwt\ genotype reaches high frequencies in the population (at least 0.98, \Cref{fig:FA}A).
 
 This result is not unique to the MAP parameter estimate. 
-We simulated genotype frequency dynamics using parameter samples from the posterior distribution (\Cref{fig:posterior}), and computed the posterior distribution of $F_A$ (\Cref{fig:ppc-plot}B). 
+We simulated genotype frequency dynamics using parameter samples from the posterior distribution (\Cref{fig:posterior}), and computed the posterior distribution of $F_A$ (\Cref{fig:FA}B). 
 The mean $F_A$ was just 0.1673 [0.0154-0.370 95\% CI] and only in 489 of 100,000 posterior samples (0.489\%) $F_A$ was  larger than 0.5, that is, \eumt\ was more likely to arise from an aneuploid cell.
 Thus, if we sample a random cell from the evolved \eumt\ population, it is more likely to have descended directly from an euploid cell than from an aneuploid cell.
-The probability of \eumt\ descending from \anwt\ ($F_A$) increases with the aneuploidy rate, $\delta$, and decreases with the mutation rate, $\mu$, and in some cases can also be affected by the fitness parameters (\Cref{fig:ppc-plot}C and \Cref{fig:posterior_FA_supp}).
+The probability of \eumt\ descending from \anwt\ ($F_A$) increases with the aneuploidy rate, $\delta$, and decreases with the mutation rate, $\mu$, and in some cases can also be affected by the fitness parameters (\Cref{fig:FA}C and \Cref{fig:posterior_FA_supp}).
 
@@
   \caption{
-    \textbf{Simulated genotype frequency dynamics.}     
+    \textbf{Predicted frequency of aneuploid-descended cells.}     
     \textbf{(A)} Posterior predicted genotype frequencies over time, including the source of \eumt: $2n^*_A$ arose from $2n+1$, whereas $2n^*_M$ arose directly from $2n$.
-Shown are 50 simulations using the MAP estimate parameters.
+Colored curves are 50 simulations using the MAP estimate parameters.
+Black dashed curves are the expected genotype frequencies without genetic drift.
+     See \Cref{fig:MAP_freq_dynamics_log} for log-log scale, in which the sequence of events is easier to observe.
     \textbf{(B)} Posterior distribution of $F_A$, the expected frequency of \eumt\ cells descended from \anwt\ cells, computed as the average frequency at the end of 100 simulations for 100,000 samples from the parameter posterior distribution.
@@
      }
-  \label{fig:ppc-plot}
+  \label{fig:FA}
 \end{figure}
@@
 Furthermore, we developed an evolutionary genetic model of adaptive evolution by aneuploidy and mutation (\Cref{fig:models}), fitted it to the experimental results of \citet{Yona2012}, and used it to predict the genotype frequency dynamics.
-The model predicted that only about 10\% of the evolved euploid population descended from aneuploid cells--that is, the majority of the euploid population are not descended from aneuploid cells, but rather are direct descendants of the ancestral population (\Cref{fig:ppc-plot}).
+The model predicted that only about 10\% of the evolved euploid population descended from aneuploid cells--that is, the majority of the euploid population are not descended from aneuploid cells, but rather are direct descendants of the ancestral population (\Cref{fig:FA}).
 
@@
 
+% FIG MAP frequency dynamics in log scale
+\begin{figure}[p]
+    \centering
+	\includegraphics[width=\textwidth]{../figures/MAP_freq_dynamics.pdf}
+	\caption{\textbf{Posterior predicted genotype frequencies in log-log scale.}
+	Frequency dynamics of the different genotypes with MAP parameter estimates, same as \Cref{fig:FA}A, but in log-log scale. Black dashed curves for a deterministic model without genetic drift. 
+	Clearly, appearance of \anwt\ and \eumtM\ is deterministic.
+	Appearance of \anmt, and therefore \eumtA, is stochastic, but their frequency dynamics after reaching frequency of roughly 0.001 is deterministic. 
+	}
+    \label{fig:MAP_freq_dynamics_log}    
+\end{figure}
+
 % FIG posterior F_A supplementary
@@
 	\caption{\textbf{Posterior distribution of $F_A$.}
-	\textbf{(A,B)} $F_A$ values (color coded) as in \Cref{fig:ppc-plot} for different parameter choices on the x- and y-axes. White star denotes the MAP estimate.
+	\textbf{(A,B)} $F_A$ values (color coded) as in \Cref{fig:FA} for different parameter choices on the x- and y-axes. White star denotes the MAP estimate.
 	}

### 2022-07-26 ms/ms.tex
added some todos
ack DBW

@@
 
+% TODO F_A vs N
+
+% TODO log-log scale of freq dynamics
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
       \centering
-      \includegraphics[width=\textwidth]{../figures/FA.pdf}      
+      \includegraphics[width=0.9\textwidth]{../figures/FA.pdf}      
   \caption{
@@
 So how does \eumtM\ prevail?
+% TODO log-log scale of freq dynamics
 At the first stage, \anwt\ cells appear and reach a high frequency due to their advantage under heat.
@@
 \section*{Acknowledgements}
-We thank Yitzhak Pilpel, Orna Dahan, Lilach Hadany, Judith Berman, David Gresham, Shay Covo, Martin Kupiec, Uri Obolski, and Tal Simon for discussions and comments.
+We thank Yitzhak Pilpel, Orna Dahan, Lilach Hadany, Judith Berman, David Gresham, Shay Covo, Martin Kupiec, Uri Obolski, Daniel Weissman, and Tal Simon for discussions and comments.
 This work was supported in part by 

### 2022-09-08 ms/ms.tex
updated text with back transition results

@@
 % Chromosomal duplication is a transient evolutionary solution to stress
-\title{Aneuploidy may be an evolutionary sidetrack on the path to adaptation}
+\title{Chromosomal duplication can be an evolutionary detour on the road to adaptation}
 
@@
 Second, we develop an evolutionary genetic model and fit it to the experimental results of \citet{Yona2012} in order to predict the genotype frequency dynamics in the experimental populations, thereby estimating the frequency of evolved euploid cells that descended from aneuploid cells.
-Our results show that, despite aneuploidy having reached high frequencies in the experimental populations, the majority of cells in the evolved euploid populations likely did not descend from aneuploid cells, but rather directly from wild-type euploid cells.
-These results suggest that aneuploidy may have been a sidetrack, rather than a stepping stone, on the path to adaptation.
+Our results show that aneuploidy reached high frequencies in the experimental populations, but nevertheless, the majority of cells in the evolved euploid populations likely did not descend from aneuploid cells, but rather directly from wild-type euploid cells.
+These results suggest that aneuploidy can be an evolutionary detour, rather than a stepping stone, on the road to adaptation.
 
@@
 For each of two evolved populations ($H2$ and $H4$) we sequenced the ancestral population (generation zero), the aneuploid population (generation 450), and the evolved euploid population (generation 1,700 or 2,350) to estimate the mutant allele frequencies (\Cref{table:mutations}).
+% TODO add table!!
 Overall, between 100 and 173 mutant alleles were detected with at least a single read in the six population that were sampled. 
@@
 
-We fitted this model to the experimental results\citep{Yona2012} -- time for fixation and for loss of aneuploidy -- using approximate Bayesian computation with sequential Monte Carol, or ABC-SMC~\citep{Sisson2009}, thereby inferring the model parameters: rates and fitness effects of aneuploidy and mutation.
+We fitted this model to the experimental results\citep{Yona2012} -- time for fixation and for loss of aneuploidy -- using approximate Bayesian computation with sequential Monte Carol (ABC-SMC)~\citep{Sisson2009}, thereby inferring the model parameters: rates and fitness effects of aneuploidy and mutation.
 We then sampled posterior predictions for the genotype frequency dynamics using the estimated parameter values and compared different versions of the model to test additional hypotheses about the evolutionary process.
@@
 %% estimates generated by MAP-and-hdi.ipynb
+
+% without back transitions:
 The estimated mutation rate, $\mu=2.965\cdot10^{-6}\ [2.718\cdot10^{-7}-3.589\cdot10^{-6}]$, corresponds to a mutation target size of $\sim 10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$ (ref.~\citep{Zhu2014}) or $3.3\cdot10^{-10}$ (ref.~\citep{Lynch2008}).
@@
 Thus, we can infer that the cost of trisomy is $c = w_{\eumt}-w_{\anmt} = 0.003$ (or 0.3\%) and the benefit of trisomy is $w_{\anwt}-1-c=0.019$ (1.9\%), whereas the benefit of beneficial mutation is $w_{\eumt}-1=0.028$ (2.8\%).
+
+If we allow for transitions (mutation, chromosome loss and gain) to less-fit genotypes (e.g., \eumt\ to \anmt), then we infer similar but slightly different values, see Supplementary Material.
 
@@
 We simulated genotype frequency dynamics using parameter samples from the posterior distribution (\Cref{fig:posterior}), and computed the posterior distribution of $F_A$ (\Cref{fig:FA}B). 
-The mean $F_A$ was just 0.1673 [0.0154-0.370 95\% CI] and only in 489 of 100,000 posterior samples (0.489\%) $F_A$ was  larger than 0.5, that is, \eumt\ was more likely to arise from an aneuploid cell.
+The mean $F_A$ was just 0.1673 [0.0154-0.370 95\% CI] and only in 489 of 100,000 posterior samples (0.489\%) $F_A$ was  larger than 0.5, that is, \eumt\ was more likely to arise from an aneuploid cell (see Supporting Material for results when transitions to less-fit genotypes are allowed, such as \eumt\ to \anmt).
 Thus, if we sample a random cell from the evolved \eumt\ population, it is more likely to have descended directly from an euploid cell than from an aneuploid cell.
@@
       \centering
-      \includegraphics[width=0.9\textwidth]{../figures/FA.pdf}      
+      \includegraphics[width=0.8\textwidth]{../figures/FA.pdf}      
   \caption{
@@
 Colored curves are 50 simulations using the MAP estimate parameters.
-Black dashed curves are the expected genotype frequencies without genetic drift.
+Black dashed curves are the expected genotype frequencies without genetic drift (from a deterministic model).
      See \Cref{fig:MAP_freq_dynamics_log} for log-log scale, in which the sequence of events is easier to observe.
@@
 Horizontal line for $F_A^{MAP}=0.106$.
+	\textbf{(E)} Expected genotype frequencies (without drift) using MAP estimate parameters.
+Solid lines for dynamics with aneuploidy, where \eumt\ fixates (frequency >99\%) at generation 1,900.
+Dashed lines for dynamics without aneuploidy ($\delta=0$), where \eumt\ fixates much earlier, at generation 500.
      }
@@
 This work was supported in part by 
-the Israel Science Foundation (ISF, YR 552/19),
-the US--Israel Binational Science Foundation (BSF, YR 2021276)
-Minerva Stiftung center for Lab Evolution (YR),
-Minerva Stiftung short-term research grant (MP).
+the Israel Science Foundation~(ISF, YR 552/19),
+the US--Israel Binational Science Foundation~(BSF, YR 2021276)
+Minerva Stiftung center for Lab Evolution~(YR),
+Minerva Stiftung short-term research grant~(MP).
 }
@@
 Together with the posterior predictive results, we conclude that the main informative prior is preferable over the extended informative prior.
+
+\subsection*{Supplementary Results}
+
+We also estimated the parameters of a version of the model that includes transitions (mutation, chromosome loss and gain) to less-fit genotypes (e.g., \eumt\ to \anmt),
+\begin{equation} \label{eq:mutation-aneuploidy-single}
+    \begin{aligned}
+      &f^m_{2n} &=&\; (1 - \delta - \mu) f^s_{2n} + \delta f^s_{2n+1} + \mu f^s_{2n^*} \;,\\
+      &f^m_{2n+1} &=&\; \delta f^s_{2n} + (1 - \delta - \mu) f^s_{2n+1} + \mu f^s_{2n+1^*}  \;,\\
+      &f^m_{2n+1^*} &=&\; \mu f^s_{2n+1} + (1 - \delta - \mu) f^s_{2n+1^*} + \delta f^s_{2n^*}  \;,\\
+      &f^m_{2n^*} &=&\; \mu f^s_{2n} + \delta f^s_{2n+1^*} + (1 - \delta - \mu) f^s_{2n^*}  \;.
+    \end{aligned}
+    \end{equation}
+
+The inferred values are slightly different.
+% with back transitions
+The estimated mutation rate, $\mu=1.036\cdot10^{-7}\ [8.01\cdot10^{-8}-1.339\cdot10^{-7}]$, corresponds to a mutation target size of $\sim 300-500$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$ (ref.~\citep{Zhu2014}) or $3.3\cdot10^{-10}$ (ref.~\citep{Lynch2008}).
+The estimated aneuploidy rate, $\delta=2.358\cdot10^{-4}\ [1.766\cdot10^{-4}-2.837\cdot10^{-4}]$ is 5-35-fold higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
+The estimated fitness values are 
+$w_{\anwt}=1.024\ [1.023-1.025]$,
+$w_{\anmt}=1.025\ [1.024-1.026]$,
+$w_{\eumt}=1.032\ [1.031-1.033]$, all relative to the fitness of \euwt, which is set to $w_{\euwt}=1$. 
+Thus, we can infer that the cost of trisomy is $c = w_{\eumt}-w_{\anmt} = 0.007$ (or 0.7\%) and the benefit of trisomy is $w_{\anwt}-1-c=0.017$ (1.7\%), whereas the benefit of beneficial mutation is $w_{\eumt}-1=0.032$ (3.2\%).
+
+We simulated genotype frequency dynamics using parameter samples from the posterior distribution, and computed the posterior distribution of $F_A$. 
+The mean $F_A$ in this case is just 0.0189 [0.0004 - 0.1214 95\% CI], lower than without the transitions to less-fit genotypes.
+Here, $F_A$ is the sum of frequencies of both \eumtA\ and $2n+1^*_A$, which reaches a frequency of 0.0007.
+Out of 100,000 posterior samples, none had $F_A$ above 0.05 (i.e., 5\% of the population).
 
@@
 	Clearly, appearance of \anwt\ and \eumtM\ is deterministic.
-	Appearance of \anmt, and therefore \eumtA, is stochastic, but their frequency dynamics after reaching frequency of roughly 0.001 is deterministic. 
+	Appearance of \anmt, and therefore \eumtA, is stochastic, however, the frequency dynamics are deterministic above a frequency of roughly 0.001. 
 	}

### 2022-09-08 ms/ms.tex
add new paragraphs on new results

@@
 \begin{abstract} 
-BLA
+BLA % TODO revise
 %Aneuploidy is common in eukaryotes, often leading to decreased cell growth and fitness.
@@
 %\paragraph{The present study.}
+% TODO revise
 Here, we test the hypothesis that aneuploidy is a \emph{transient adaptive solution}.
@@
 
-% TODO F_A vs N
-
-% TODO log-log scale of freq dynamics
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Horizontal line for $F_A^{MAP}=0.106$.
-	\textbf{(E)} Expected genotype frequencies (without drift) using MAP estimate parameters.
-Solid lines for dynamics with aneuploidy, where \eumt\ fixates (frequency >99\%) at generation 1,900.
-Dashed lines for dynamics without aneuploidy ($\delta=0$), where \eumt\ fixates much earlier, at generation 500.
+	\textbf{(E)} Population mean fitness in a model without drift using MAP estimate parameters.
+Solid lines for mean fitness with aneuploidy ($\delta>0$), where the population reaches adaptation (mean fitness at 99.99\% of maximum value) at generation 1,802.
+Dashed lines for mean fitness without aneuploidy ($\delta=0$), where the population adapts much earlier, at generation 535.
      }
@@
 Conventional wisdom might suggest that once the aneuploid genotype \anwt\ reaches high frequency, it will have a better chance at producing "refined" solutions via mutations, and its descendants will come to dominate the population: the frequency \eumtA\ (which arises from \anmt) will be higher than the frequency of \eumtM\ (which arises directly from \euwt).
+
 So how does \eumtM\ prevail?
-% TODO log-log scale of freq dynamics
-At the first stage, \anwt\ cells appear and reach a high frequency due to their advantage under heat.
-Accordingly, \anmt\ cells appear more frequently than \eumtM.
+The supply rates of \anwt\ and \eumtM\ are $N \delta \approx 11,000$ and $N \mu \approx 19$, respectively (assuming MAP parameter estimates). 
+Therefore, both genotypes are expected to appear immediately at the beginning of the experiment (\Cref{fig:MAP_freq_dynamics_log}).
+However, \anwt\ appears at a much higher frequency as $\delta \gg \mu$ by 2-3 orders of magnitude.
+After they first appear, \eumtM\ has higher fitness.
+But as long as the frequency of \euwt\ is high, the supply rate of \anwt\ is higher than that of \eumtM, again due to $\delta \gg \mu$. 
+However, because both genotypes are produced from \euwt.
+Therefore, when the frequency of \euwt\ decreases, mainly due to the increase in the frequency of \anwt, both supply rates diminish.
+At this point, the higher fitness of \eumtM\ comes into play and it starts to take over the population, which at this stage is mainly composed of \anwt.
+For the aneuploid lineage to compete with the mutant lineage, it must produce \eumtA\ via a mutation followed by chromosome loss.
+Although this is a stochastic process (due to drift), our results show that the time until \eumtA\ reaches a frequency of ~0.001 is roughly 450 generation, without a lot of variation (intersection of purple lines and vertical dashed line in \Cref{fig:MAP_freq_dynamics_log}).
+However, by that time, \eumtM\ is already at a much higher frequency, ~0.18, and since both mutants have the same fitness, their relative frequency remains roughly the same until the end of the experiment.
+%At the first stage, \anwt\ cells appear and reach a high frequency due to their advantage under heat.
+%Accordingly, \anmt\ cells appear more frequently than \eumtM.
 However, the fitness advantage of a "refined" mutation that emerges in the \anmt\ genotype in hindered by the cost of aneuploidy as well as the extra waiting time needed to eliminate the extra copy of the chromosome by a chromosome loss event.
-Although such chromosome loss occurs fast relative to mutation, it still leaves a time window for \eumtM\ to arise and reach high frequencies.
-Indeed, we estimate that the advantage of \anmt\ over \anwt\ is similar to the advantage of \eumt\ over \anmt (about 0.3\%), and that the rate of chromosome loss is roughly $1-3\cdot10^{-3}$.
+%Although such chromosome loss occurs fast relative to mutation, it still leaves a time window for \eumtM\ to arise and reach high frequencies.
+%Indeed, we estimate that the advantage of \anmt\ over \anwt\ is similar to the advantage of \eumt\ over \anmt (about 0.3\%), and that the rate of chromosome loss is roughly $1-3\cdot10^{-3}$.
 
-% TODO effect of population size?
+% F_A vs N
+\paragraph{Prediction for a small populations.}
+We examined the effect of the population size, $N$, on the frequency of \anwt\ descendants in the evolved population, $F_A$.
+We found that $F_A$ is expected to decrease as the population size increases (\Cref{fig:FA}D), ranging from about 0.9 when the population size is 10,000, to about 0.1 when the population size is above 1,000,000 (less than the experimental population size, which was 6,425,000).
+Thus, our model provides a testable prediction: if the experiment was repeated under a lower population size (via stronger daily dilutions or with in a smaller volume), then the fraction of the population descending from aneuploid cells would likely be much higher.
+
+% mean fitness
+\paragraph{Aneuploidy increases adaptation time.}
+We found that aneuploidy quickly appears and reaches high frequencies but is then outcompeted by adaptive mutant lineages. 
+An additional interesting outcome is that aneuploidy increases rather than decreases the adaptation time (expected time until the population mean fitness is at 99.99\% of its maximum value, which is 1,802 and 535 generation with and without aneuploidy, respectively; \Cref{fig:FA}E).
+This happens despite the fact that the mean fitness initially increases faster in the presence of aneuploidy (\Cref{fig:FA}E). 
+The reason is that once \anwt\ is common, selection for the mutant strain (\anmt\ or \eumt) is weaker compared to when \eumt\ competes directly with \euwt.
 
@@
 \paragraph{Conclusions.}
+% TODO revise
 Here, we tested the hypothesis that aneuploid cells are an evolutionary adaptive intermediate between wild-type euploid cells and mutant euploid cells. 

### 2022-09-11 ms/ms.tex
revise results, intro, discussion; wrote abstract

--- 2022-09-08 ms/ms.tex
@@
 % Chromosomal duplication is a transient evolutionary solution to stress
-\title{Chromosomal duplication can be an evolutionary detour on the road to adaptation}
+\title{Chromosomal duplication can be an evolutionary detour on the path to adaptation}
 
@@
 
+Classifications. Biological Sciences: Evolution, Genetics, Microbiology, Population Biology.
+
+Keywords: aneuploidy, evolutionary model, adaptive evolution
+
+\pagebreak
+
 % Abstract 250 words
 \begin{abstract} 
-BLA
-%Aneuploidy is common in eukaryotes, often leading to decreased cell growth and fitness.
-However, evidence from yeast and fungi, as well as human tumour cells, suggests that aneuploidy can be beneficial under stressful conditions and lead to elevated growth rates and adaptation.
-%Importantly, aneuploidy differs from point mutations in rate, fitness effect, and reversibility.
-%Here, we develop an evolutionary model for adaptive evolution with both aneuploidy and  mutation, and apply it within an Bayesian inference framework.
-%We estimate the formation rate and fitness effect of aneuploidy and mutation in an evolutionary experiment in which \yeast adapted to heat stress.
-%Our results suggest that aneuploidy is more frequent than previously estimated and that experimental observations cannot be explained without aneuploidy.
-%We do not find evidence for genetic instability in aneuploid cells.
-%Our results support the hypothesis that aneuploidy is a transient adaptive hypothesis: the experimental populations adapted by first acquiring a chromosome duplication; then by acquiring a beneficial mutation that likely appeared in aneuploid cells, rather than euploid cells; and finally by rapidly losing the extra chromosome to avoid paying the fitness cost of aneuploidy.
+Aneuploidy is common in eukaryotes, often leading to decreased growth and fitness.
+However, evidence from yeast and fungi, as well as human tumour cells, suggests that aneuploidy can be beneficial under stressful conditions and facilitate  adaptation.
+In a prominent example, an evolutionary experiment with \yeast, populations evolving under heat stress had become aneuploid, only to later revert back to euploid after genetic mutations have accumulated.
+It has therefore been suggested that aneuploidy serves as a "stepping stone" on the path to adaptation.
+Here, we test this hypothesis. First, we apply DNA sequencing to show that mutant alleles common in aneuploid cells are uncommon in the evolved euploid population. 
+Second, we develop an evolutionary model with both aneuploidy and  mutation, and fit it to the results of the experiment using a Bayesian inference framework.
+We then predict the genotype frequency dynamics during the experiment, demonstrating that the majority of the evolved euploid population likely did not descend from aneuploid cells, but rather directly from the euploid wild-type population.
+Together, our results suggest that aneuploidy is an evolutionary "detour" rather than a "stepping stone": it delays rather than facilitates the adaptation of the population, and cells that become aneuploid leave less descendants compared to cells that remain diploid.
 \end{abstract}
@@
 
-Classifications. Biological Sciences: Evolution, Genetics, Microbiology, Population Biology.
+\pagebreak
 
-Keywords: aneuploidy, evolutionary model, adaptive evolution
-
-% TODO cite https://www.biorxiv.org/content/10.1101/2020.10.07.327759v1
-% cite: https://www.cell.com/developmental-cell/fulltext/S1534-5807(21)00562-1?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS1534580721005621%3Fshowall%3Dtrue
-% cite: https://www.cell.com/developmental-cell/fulltext/S1534-5807(21)00592-X?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS153458072100592X%3Fshowall%3Dtrue
-
-\pagebreak
 % Introduction
@@
 Evidence suggests aneuploidy is very common in eukaryotes, e.g. animals~\citep{Santaguida2015review, Naylor2016, Bakhoum2017}, and fungi~\citep{Pavelka2010, Zhu2016, Robbins2017, Todd2017}.
-Aneuploidy has been implicated in cancer formation and progression~\citep{Boveri2008, Schvartzman2010,Santaguida2015review}.
-%90\% of solid tumours and 50\% of blood cancers are aneuploid~\citep{Santaguida2015review}.
+Aneuploidy has been implicated in cancer formation, progression, and resistance~\citep{Boveri2008, Schvartzman2010, Santaguida2015review, Ippolito2021b}.
 It is also common in protozoan pathogens of the Leishmania genus, a major global health concern~\citep{Mannaert2012}, and contributes to the emergence of drug resistance~\citep{Selmecki2009} and virulence~\citep{Moller2018} in fungal pathogens, which are under-studied~\citep{Rodrigues2018}, despite infecting a billion people per year, causing significant morbidity in >150 million and death in >1.5 million people per year~\citep{Selmecki2009, Rodrigues2018}.
@@
 %\paragraph{Aneuploidy is generally deleterious.}
-%The molecular and genetic mechanisms involved in aneuploidy have been explored~\citep{Musacchio2007, Sheltzer2011, Chen2012b, Rancati2013, Gerstein2015, Shor2015}.
 Experiments with human and mouse embryos found that aneuploidy is usually lethal.
@@
 %\paragraph{Transient adaptive solution.} 
-Experimental evidence for the role of aneuploidy in adaptive evolution was demonstrated by \citet{Yona2012}. They evolved populations of \yeast under strong heat stress.
+\citet{Yona2012} demonstrated experimentally the importance of aneuploidy in adaptive evolution. They evolved populations of \yeast under strong heat stress.
 The populations adapted to the heat stress within 450 generations, and this adaptation was determined to be due a duplication of chromosome III.
 Later on, after more than 1,500 generations, the populations reverted back to an euploid state, while remaining adapted to the heat stress. 
-The authors hypothesized that aneuploidy is a \emph{transient adaptive solution}, because it can rapidly appear and fixate in the population under stressful conditions, and can then be rapidly lost when the cost of aneuploidy outweighs its benefit---after the stress is removed, or after "refined" beneficial mutations appear and fixate.
+Aneuploidy was therefore suggested to be a \emph{transient adaptive solution}, because it can rapidly appear and fixate in the population under stressful conditions, and can then be rapidly lost when the cost of aneuploidy outweighs its benefit---after the stress is removed, or after "refined" beneficial mutations appear and fixate~\citep{Yona2012}.
+Furthermore, it has been suggested that aneuploidy is an evolutionary "stepping stone" that facilitates future adaptation by genetic mutations, which require more time to evolve~\citep{Yona2012,Yona2015}.
 
 %\paragraph{The present study.}
-Here, we test the hypothesis that aneuploidy is a \emph{transient adaptive solution}.
-First, we analyze previously unpublished sequencing data from the original experimental populations of \citet{Yona2012} to assess if the aneuploid population and the evolved euploid population carry the same mutant alleles, as expected if the later is descended from the former.
+Here, we test the hypothesis that aneuploidy is a an \emph{evolutionary stepping stone} that facilitates adaptive evolution by genetic mutations.
+First, we analyze previously unpublished sequencing data from the original experimental populations of \citet{Yona2012} to assess if the evolved euploid population is descended from the aneuploid population.
 Second, we develop an evolutionary genetic model and fit it to the experimental results of \citet{Yona2012} in order to predict the genotype frequency dynamics in the experimental populations, thereby estimating the frequency of evolved euploid cells that descended from aneuploid cells.
 Our results show that aneuploidy reached high frequencies in the experimental populations, but nevertheless, the majority of cells in the evolved euploid populations likely did not descend from aneuploid cells, but rather directly from wild-type euploid cells.
-These results suggest that aneuploidy can be an evolutionary detour, rather than a stepping stone, on the road to adaptation.
+These results suggest that aneuploidy can be an evolutionary detour, rather than a stepping stone, on the path to adaptation.
 
@@
 \paragraph{Empirical frequencies of mutant alleles.} 
-Our first test of the \emph{transient adaptive solution hypothesis} employs previously unpublished sequencing data.
-For each of two evolved populations ($H2$ and $H4$) we sequenced the ancestral population (generation zero), the aneuploid population (generation 450), and the evolved euploid population (generation 1,700 or 2,350) to estimate the mutant allele frequencies (\Cref{table:mutations}).
+For each of two evolved populations ($H2$ and $H4$) we sequenced the ancestral population (generation 0), the aneuploid population (generation 450), and the evolved euploid population (generation 1,700 or 2,350) to estimate the mutant allele frequencies (\Cref{table:mutations}).
 % TODO add table!!
-Overall, between 100 and 173 mutant alleles were detected with at least a single read in the six population that were sampled. 
+Overall, between 100 and 173 mutant alleles were detected with at least a single read in the six populations that were sampled. 
 Disregarding 45 and 40 alleles that were present in the ancestral populations at a frequency >10\%, the aneuploid and euploid populations carried a large number of mutant alleles: 82 and 95, respectively, in repetition $H2$, and 60 and 66 in repetition $H4$.
@@
 \paragraph{Evolutionary genetic model.} 
-To further test the \emph{transient adaptive solution hypothesis}, we developed an evolutionary genetic model, fitted the model to empirical data, and used it to predict the genotype frequency dynamics, or specifically, the fraction of the evolved euploid population descended from aneuploid cells.
+To explore the dynamics during the evolutionary experiments, we developed an evolutionary genetic model, fitted the model to empirical data, and used it to predict the genotype frequency dynamics, or specifically, the fraction of the evolved euploid population descended from aneuploid cells.
 
@@
 
-We fitted this model to the experimental results\citep{Yona2012} -- time for fixation and for loss of aneuploidy -- using approximate Bayesian computation with sequential Monte Carol (ABC-SMC)~\citep{Sisson2009}, thereby inferring the model parameters: rates and fitness effects of aneuploidy and mutation.
+We fitted this model to the experimental results\citep{Yona2012} -- time for fixation (>95\%) and for loss (<5\%) of aneuploidy -- using approximate Bayesian computation with sequential Monte Carol (ABC-SMC)~\citep{Sisson2009}, thereby inferring the model parameters: rates aneuploidy and mutation and the fitness of all genotypes.
 We then sampled posterior predictions for the genotype frequency dynamics using the estimated parameter values and compared different versions of the model to test additional hypotheses about the evolutionary process.
@@
 %%%%%%%% Fig illustration
-\begin{figure}[h]
+\begin{figure}[t]
   \centering
@@
 We inferred the posterior distribution of model parameters~(\Cref{fig:posterior}). 
-We report parameter estimates using the MAP (maximum a posteriori) and providing the 50\% HDI (highest density interval) in square brackets. See \emph{Supporting material} for sensitivity analysis.
+We report parameter estimates using the MAP (maximum a posteriori) and providing the 50\% HDI (highest density interval) in square brackets. See Supplementary Material for sensitivity analysis.
 %% estimates generated by MAP-and-hdi.ipynb
@@
 The estimated mutation rate, $\mu=2.965\cdot10^{-6}\ [2.718\cdot10^{-7}-3.589\cdot10^{-6}]$, corresponds to a mutation target size of $\sim 10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$ (ref.~\citep{Zhu2014}) or $3.3\cdot10^{-10}$ (ref.~\citep{Lynch2008}).
-The estimated aneuploidy rate, $\delta=1.72\cdot10^{-3}\ [1.47\cdot10^{-3}-2.786\cdot10^{-3}]$ is much higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
+The estimated aneuploidy rate, $\delta=1.72\cdot10^{-3}\ [1.47\cdot10^{-3}-2.786\cdot10^{-3}]$ is higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
 The estimated fitness values are $w_{\anwt}=1.022\ [1.021-1.023]$,
@@
 $w_{\eumt}=1.028\ [1.026-1.029]$, all relative to the fitness of \euwt, which is set to $w_{\euwt}=1$. 
-Thus, we can infer that the cost of trisomy is $c = w_{\eumt}-w_{\anmt} = 0.003$ (or 0.3\%) and the benefit of trisomy is $w_{\anwt}-1-c=0.019$ (1.9\%), whereas the benefit of beneficial mutation is $w_{\eumt}-1=0.028$ (2.8\%).
+Thus, we can infer that the cost of trisomy is $c = w_{\eumt}-w_{\anmt} = 0.003$ (or 0.3\%) and the benefit of trisomy is $w_{\anwt}-1-c=0.019$ (1.9\%), whereas the benefit of the beneficial mutation is $w_{\eumt}-1=0.028$ (2.8\%).
 
@@
 \paragraph{Model comparison and goodness-of-fit.}
-
-Our model fits the data well: in simulations using the MAP parameter estimates, \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\Cref{fig:2n*-fixation}B).
-
-However, a model without aneuploidy (where the aneuploidy rate is fixed at zero, $\delta=0$), fails to explain the experimental observations~(Figure~\ref{fig:2n*-fixation}).
-%% generated with MAP-and-hdi.ipynb
-The estimated mutation rate without aneuploidy is $\mu=7.98\cdot10^{-9}\ [7.906\cdot10^{-9}-8.138\cdot10^{-9}]$, much lower compared to a model with aneuploidy and suggesting a target size of just 40.
-The fitness of the mutant is also much lower at $w_{\eumt}=1.013\ [1.012-1.013]$.
-This is because, without aneuploidy, a high mutation rate or fitness effect will lead to faster appearance and fixation of \eumt\ than in the experimental observations.
-Even with these lower estimates, the model fit is worse than that of a model with aneuploidy~(\Cref{fig:2n*-fixation}).
-
-We also checked a model in which aneuploidy occurs but is adaptively neutral compared to the wild-type, that is, $w_{\anwt} = w_{\euwt}$ and $w_{\anmt} = w_{\eumt}$ but $\delta>0$.
-This model fits the data better than the model with no aneuploidy (in which $\delta=0$), but worse than a model with positive selection for aneuploidy, in which $w_{\euwt}<w_{\anwt}<w_{\anmt}<w_{\eumt}$~(\Cref{fig:2n*-fixation}).
 
@@
 
+Our model fits the data well: in simulations using the MAP parameter estimates, \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\Cref{fig:2n*-fixation}B).
+
+However, a model without aneuploidy (where the aneuploidy rate is fixed at zero, $\delta=0$), fails to explain the experimental observations~(Figure~\ref{fig:2n*-fixation}).
+%% generated with MAP-and-hdi.ipynb
+The estimated mutation rate without aneuploidy is $\mu=7.98\cdot10^{-9}\ [7.906\cdot10^{-9}-8.138\cdot10^{-9}]$, much lower compared to a model with aneuploidy and suggesting a target size of just 40.
+The fitness of the mutant is also much lower at $w_{\eumt}=1.013\ [1.012-1.013]$.
+This is because, without aneuploidy, a high mutation rate or fitness effect will lead to faster appearance and fixation of \eumt\ than in the experimental observations.
+Even with these lower estimates, the model fit is worse than that of a model with aneuploidy~(\Cref{fig:2n*-fixation}).
+
+We also checked a model in which aneuploidy occurs but is adaptively neutral compared to the wild-type, that is, $w_{\anwt} = w_{\euwt}$ and $w_{\anmt} = w_{\eumt}$ but $\delta>0$.
+This model fits the data better than the model with no aneuploidy (in which $\delta=0$), but worse than a model with positive selection for aneuploidy, in which $w_{\euwt}<w_{\anwt}<w_{\anmt}<w_{\eumt}$~(\Cref{fig:2n*-fixation}).
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-We tested the \emph{transient adaptive solution hypothesis} by estimating $F_A$, the expected frequency of \eumt\ that arose from \anwt, computed as the average frequency of such cells at the end of simulations using the MAP estimate parameters.
-Surprisingly, we observe that the majority of \eumt\ cells are a product of a direct mutation in \euwt\ cells, rather than descending from \anwt\ cells ($F_A^{MAP} = 0.106$, \Cref{fig:FA}A). 
+To test the hypothesis that aneuploidy facilitates adaptation, we estimated $F_A$, the expected frequency of \eumt\ that arose from \anwt, computed as the average frequency of such \eumtA\ cells at the end of simulations using the MAP estimate parameters.
+Surprisingly, we observe that the majority of \eumt\ cells are \eumtM, a product of a direct mutation in \euwt\ cells, rather than descending from \anwt\ cells ($F_A^{MAP} = 0.106$, \Cref{fig:FA}A). 
 This is despite the fact that the \anwt\ genotype reaches high frequencies in the population (at least 0.98, \Cref{fig:FA}A).
@@
 We simulated genotype frequency dynamics using parameter samples from the posterior distribution (\Cref{fig:posterior}), and computed the posterior distribution of $F_A$ (\Cref{fig:FA}B). 
-The mean $F_A$ was just 0.1673 [0.0154-0.370 95\% CI] and only in 489 of 100,000 posterior samples (0.489\%) $F_A$ was  larger than 0.5, that is, \eumt\ was more likely to arise from an aneuploid cell (see Supporting Material for results when transitions to less-fit genotypes are allowed, such as \eumt\ to \anmt).
+The mean $F_A$ was just 0.1673 [0.0154-0.370 95\% CI] and only in 489 of 100,000 posterior samples (0.489\%) $F_A$ was  larger than 0.5 (see Supporting Material for results when transitions to less-fit genotypes are allowed, such as \eumt\ to \anmt).
 Thus, if we sample a random cell from the evolved \eumt\ population, it is more likely to have descended directly from an euploid cell than from an aneuploid cell.
-The probability of \eumt\ descending from \anwt\ ($F_A$) increases with the aneuploidy rate, $\delta$, and decreases with the mutation rate, $\mu$, and in some cases can also be affected by the fitness parameters (\Cref{fig:FA}C and \Cref{fig:posterior_FA_supp}).
-
-% TODO F_A vs N
-
-% TODO log-log scale of freq dynamics
+The probability of \eumt\ descending from \anwt\ ($F_A$) increases with the aneuploidy rate, $\delta$, and decreases with the mutation rate, $\mu$, and population size $N$ (\Cref{fig:FA}C,D).
+In some cases it can also be affected by the fitness parameters (\Cref{fig:posterior_FA_supp}).
 
@@
 
-% Rancati G, Pavelka N.
-Karyotypic changes as drivers and catalyzers of cellular evolvability: A perspective from non-pathogenic yeasts.
-Semin Cell Dev Biol.
-2013;24(4):332-338.
-doi:10.1016/j.semcdb.2013.01.009
-% Bouchonville K, Forche A, Tang KES, Semple C a M, Berman J.
-Aneuploid chromosomes are highly unstable during DNA transformation of Candida albicans.
-Eukaryot Cell.
-2009;8(10):1554-1566.
-doi:10.1128/EC.00209-09
-% Zhu J, Pavelka N, Bradford WD, Rancati G, Li R. Karyotypic determinants of chromosome instability in aneuploid budding yeast. PLoS Genet. 2012;8(5). doi:10.1371/journal.pgen.1002719
-
 \paragraph{Genetic instability in aneuploid cells.}
-It has been suggested that aneuploidy increased genetic instability~\citep{Sheltzer2011b}.
+It has been suggested that aneuploidy increases genetic instability~\citep{Sheltzer2011b, Ippolito2021b}.
 Therefore, we inferred model parameters under the assumption that the mutation rate increases in aneuploid cells by a factor $\tau$ = 1, 33/32 (due to an additional chromosome), 2, 5, 10, or 100 (due to genetic instability).
@@
 Additionally, such a replacement also occurred when they initiated evolutionary experiments with a population in which all cells carry a chromosome III trisomy.
-They hypothesized that these evolved euploid cells were a consequence of mutations in genes that confer heat resistance, followed by reversion of trisomy by a chromosome loss event.
+They hypothesized that the euploid mutant cells evolved by heat-resistance mutations in aneuploid cells followed by reversion of trisomy due to a chromosome loss event.
 
@@
 Furthermore, we developed an evolutionary genetic model of adaptive evolution by aneuploidy and mutation (\Cref{fig:models}), fitted it to the experimental results of \citet{Yona2012}, and used it to predict the genotype frequency dynamics.
-The model predicted that only about 10\% of the evolved euploid population descended from aneuploid cells--that is, the majority of the euploid population are not descended from aneuploid cells, but rather are direct descendants of the ancestral population (\Cref{fig:FA}).
+The model predicted that only about 10-15\% of the evolved euploid population descended from aneuploid cells--that is, the majority of the euploid population are not descended from aneuploid cells, but rather are direct descendants of the ancestral wild-type population (\Cref{fig:FA}).
 
 This happens despite aneuploidy reaching a high frequency in the population (>95\%). 
-Conventional wisdom might suggest that once the aneuploid genotype \anwt\ reaches high frequency, it will have a better chance at producing "refined" solutions via mutations, and its descendants will come to dominate the population: the frequency \eumtA\ (which arises from \anmt) will be higher than the frequency of \eumtM\ (which arises directly from \euwt).
+Conventional wisdom might suggest that once the aneuploid genotype \anwt\ reaches high frequency, it will have a better chance at producing "refined" solutions via mutations, and its descendants will come to dominate the population: the frequency of \eumtA\ (which arises from \anmt) will be higher than the frequency of \eumtM\ (which arises directly from \euwt).
+
 So how does \eumtM\ prevail?
-% TODO log-log scale of freq dynamics
-At the first stage, \anwt\ cells appear and reach a high frequency due to their advantage under heat.
-Accordingly, \anmt\ cells appear more frequently than \eumtM.
-However, the fitness advantage of a "refined" mutation that emerges in the \anmt\ genotype in hindered by the cost of aneuploidy as well as the extra waiting time needed to eliminate the extra copy of the chromosome by a chromosome loss event.
-Although such chromosome loss occurs fast relative to mutation, it still leaves a time window for \eumtM\ to arise and reach high frequencies.
-Indeed, we estimate that the advantage of \anmt\ over \anwt\ is similar to the advantage of \eumt\ over \anmt (about 0.3\%), and that the rate of chromosome loss is roughly $1-3\cdot10^{-3}$.
+Initially, the supply rates of \anwt\ and \eumtM\ are $N \delta \approx 11,000$ and $N \mu \approx 19$, respectively (assuming MAP parameter estimates). 
+Therefore, both genotypes are expected to appear immediately at the beginning of the experiment (\Cref{fig:MAP_freq_dynamics_log}).
+However, \anwt\ appears at a much higher frequency as $\delta \gg \mu$ by 2-3 orders of magnitude.
+After they first appear, \eumtM\ has higher fitness.
+But as long as the frequency of \euwt\ is high, the supply rate of \anwt\ is higher than that of \eumtM, again due to $\delta \gg \mu$. 
+However, supply rates of both genotypes decreases with the frequency of \euwt.
+Therefore, when the latter decreases, mainly due to the increase in the frequency of \anwt, both supply rates diminish.
+At this stage, the higher fitness of \eumtM\ comes into play and it starts to take over the population, which is mainly composed of \anwt.
+For the aneuploid lineage to compete with the mutant lineage, it must produce \eumtA\ via a mutation followed by chromosome loss.
+Although this is a stochastic process (due to drift), our results show that the time until \eumtA\ reaches a frequency of ~0.1\% is roughly 450 generations, without much variation (intersection of purple lines and vertical dashed line in \Cref{fig:MAP_freq_dynamics_log}).
+However, by that time \eumtM\ is already at a roughly 10-fold higher frequency (1.86\%), and since both mutants have the same fitness, their relative frequency remains roughly the same until the end of the experiment.
 
-% TODO effect of population size?
+% population size
+\paragraph{Predictions for small populations.}
+We examined the effect of the population size, $N$, on the frequency of \anwt\ descendants in the evolved population, $F_A$.
+We found that $F_A$ is expected to decrease as the population size increases (\Cref{fig:FA}D), ranging from about 90\% when the population size is 10,000, to about 10\% when the population size is above 1,000,000 (less than the experimental population size, which was 6,425,000).
+Thus, our model provides a testable prediction: if the experiment was repeated under a lower population size (via stronger daily dilutions or in a smaller volume), than the fraction of the population descending from aneuploid cells would be much higher.
 
+% mean fitness
+\paragraph{Aneuploidy delays rather than facilitates adaptation.}
+An additional interesting result of our study is that aneuploidy increases, rather than decreases, the adaptation time (\Cref{fig:FA}E). 
+This happens despite the fact that the mean fitness initially increases faster in the presence of aneuploidy (\Cref{fig:FA}E). 
+This is because once \anwt\ is common, selection for the mutant strain (\anmt\ or \eumt) is weaker compared to when \eumt\ competes directly with \euwt.
+
+% parameter estimates
 \paragraph{Rate and fitness effect of aneuploidy and mutation.}
-We inferred the rates and fitness effects of aneuploidy and mutation in a previously published evolutionary experiment in which yeast populations adapted to heat stress~\citep{Yona2012}. 
-We estimate that the aneuploidy rate (i.e., number of chromosome gains per generation) is $1.7 \cdot 10^{-3}$, is higher than a previous estimate of $6.7 \cdot 10^{-6}$~\citep{Zhu2016}.
-% TODO explain?
+We inferred the rates of aneuploidy and mutation and their effects on fitness. 
+We estimate that the aneuploidy rate (i.e., number of chromosome gains per generation) is $1.7 \cdot 10^{-3}$, higher than a previous estimate of $6.7 \cdot 10^{-6}$~\citep{Zhu2016}.
+This may be due to genetic instability caused by heat stress~\citep{Chen2012a}.
 In addition, we find no evidence for increased mutation rates in aneuploid cells. 
-Previous empirical studies have suggested that genetic instability (e.g., elevated mutation rates) in aneuploid cells is due to stress associated with the aneuploid state~\citep{Bouchonville2009,Chen2012b,Zhu2012}.
+Previous empirical studies have suggested that genetic instability (e.g., elevated mutation rates) in aneuploid cells is due to stress associated with the aneuploid state~\citep{Bouchonville2009, Chen2012b, Zhu2012, Ippolito2021b}.
 However, in the experiment of \citet{Yona2012}, both the wild-type and the aneuploid were under heat stress, which may explain why we did not find evidence for an increased mutation rate.
 
+% Conclusions
 \paragraph{Conclusions.}
-Here, we tested the hypothesis that aneuploid cells are an evolutionary adaptive intermediate between wild-type euploid cells and mutant euploid cells. 
-Our results suggest this is only true at the population level, as the population does go from euploid to aneuploid and back--but not at the individual level: we estimate that only about 10-15\% of the euploid cells descended from aneuploid cells, whereas the rest are direct descendent of the wild-type euploid cells.
-
-% TODO finish conclusions
+Here, we tested the hypothesis that aneuploid cells are an evolutionary "stepping stone", or adaptive intermediate, between wild-type euploid cells and mutant euploid cells. 
+Our results suggest that, although it seems the population goes from euploid to aneuploid and back, this is not the case at the individual level.
+We estimate that only about 10-15\% of the euploid cells descended from aneuploid cells, whereas the rest are direct descendants of the wild-type euploid cells.
+This surprising result reinforces the importance of models when making interpretations on evolutionary processes, and emphasizes the unintuitive outcomes of clonal interference during adaptive evolution.
 
@@
 In their heat-stress experiment, four populations of \yeast evolved under \SI{39}{\celsius}.
-Aneuploidy fixed in all four population in the first 450 generations (hereafter, fixation or elimination of a genotype \emph{by generation $t$} means that more than 95\% or less than 5\% of the population carry the genotype at generation $t$, and possibly earlier).
+Aneuploidy fixed in all four population in the first 450 generations.
+Hereafter, fixation or elimination of a genotype \emph{by generation $t$} means that more than 95\% or less than 5\% of the population carry the genotype at generation $t$, and possibly earlier.
 From re-analysis of data not published in the original paper, aneuploidy did not fix before at least 200 generations elapsed.

### 2022-09-11 ms/ms.tex
added mutation tables

@@
 \usepackage{booktabs}
-\usepackage{pgfplotstable} 
+\usepackage{pgfplotstable} % http://mirrors.ctan.org/graphics/pgf/contrib/pgfplots/doc/pgfplotstable.pdf
 \PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
@@
 \paragraph{Empirical frequencies of mutant alleles.} 
-For each of two evolved populations ($H2$ and $H4$) we sequenced the ancestral population (generation 0), the aneuploid population (generation 450), and the evolved euploid population (generation 1,700 or 2,350) to estimate the mutant allele frequencies (\Cref{table:mutations}).
-% TODO add table!!
+For each of two evolved populations ($H2$ and $H4$) we sequenced the ancestral population (generation 0), the aneuploid population (generation 450), and the evolved euploid population (generation 1,700 or 2,350) to estimate the mutant allele frequencies (\Cref{table:mutations_H2,table:mutations_H4}).
 Overall, between 100 and 173 mutant alleles were detected with at least a single read in the six populations that were sampled. 
 Disregarding 45 and 40 alleles that were present in the ancestral populations at a frequency >10\%, the aneuploid and euploid populations carried a large number of mutant alleles: 82 and 95, respectively, in repetition $H2$, and 60 and 66 in repetition $H4$.
+
+Surprisingly, out of all these mutant alleles, none was present at a frequency >20\% in both the aneuploid and the evolved euploid populations. 
+Furthermore, a high mutant allele frequency in the aneuploid population implies a low frequency in the evolved euploid population, and vice-versa (Spearman's correlation coefficient $\rho=-0.64$ and $-0.66$ in the two experimental repetitions; \Cref{fig:mutation_frequencies}), such that mutant alleles frequent in the aneuploid populations decreased in frequency when aneuploidy was lost.
+Moreover, for the 18 mutant alleles with high frequency in the aneuploid populations (>20\%), the highest frequencies in the euploid populations were 15.4\%, 16\%, 16.3\% and 19.6\% (the rest were below 15\%).
+Similarly, for the 48 mutant alleles with high frequency in the evolved euploid populations, the highest frequencies in the aneuploid populations were 2.7\%, 7.7\%, and 11.1\% (the rest were below 1\%).
+These results suggest evolved euploid cells are unlikely to descend from aneuploid~cells.
 
@@
 \end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-Surprisingly, out of all these mutant alleles, none was present at a frequency >20\% in both the aneuploid and the evolved euploid populations. 
-Furthermore, a high mutant allele frequency in the aneuploid population implies a low frequency in the evolved euploid population, and vice-versa (Spearman's correlation coefficient $\rho=-0.64$ and $-0.66$ in the two experimental repetitions; \Cref{fig:mutation_frequencies}), such that mutant alleles frequent in the aneuploid populations decreased in frequency when aneuploidy was lost.
-Moreover, for the 18 mutant alleles with high frequency in the aneuploid populations (>20\%), the highest frequencies in the euploid populations were 15.4\%, 16\%, 16.3\% and 19.6\% (the rest were below 15\%).
-Similarly, for the 48 mutant alleles with high frequency in the evolved euploid populations, the highest frequencies in the aneuploid populations were 2.7\%, 7.7\%, and 11.1\% (the rest were below 1\%).
-These results suggest evolved euploid cells are unlikely to descend from aneuploid~cells.
 
@@
 
+\pagebreak
+
+% Table - mutations
+%% generated with mutation_analysis.ipynb
+\begin{table}[h]
+\centering
+\caption{
+\textbf{Mutant alleles in population $H2$.}}
+\caption*{
+	Mutant alleles identified in the ancestor (generation 0), aneuploid (generation 450), and evolved (generation 1,700) of population $H2$. See supplementary file.
+} 
+\label{table:mutations_H2}
+\end{table}
+
+% Table - mutations
+%% generated with mutation_analysis.ipynb
+\begin{table}[h]
+\centering
+\caption{
+\textbf{Mutant alleles in population $H4$.}}
+\caption*{
+	Mutant alleles identified in the ancestor (generation 0), aneuploid (generation 450), and evolved (generation 2,350) of population $H4$. See supplementary file.
+} 
+\label{table:mutations_H4}
+\end{table}
+
 %% generated with WAIC.ipynb 
@@
 
-%TODO table:mutations
-
 \end{document}

### 2022-09-11 ms/ms.tex
cover letter; line num

@@
 
+% line numbers
+ \usepackage[displaymath, mathlines]{lineno}
+ \renewcommand\linenumberfont{\normalfont\small\sffamily}
+\linenumbers
+\modulolinenumbers[2]
+
 % genotype commands
@@
 However, evidence from yeast and fungi, as well as human tumour cells, suggests that aneuploidy can be beneficial under stressful conditions and facilitate  adaptation.
-In a prominent example, an evolutionary experiment with \yeast, populations evolving under heat stress had become aneuploid, only to later revert back to euploid after genetic mutations have accumulated.
+In a prominent example, an evolutionary experiment with yeast, populations evolving under heat stress had become aneuploid, only to later revert back to euploid after genetic mutations have accumulated.
 It has therefore been suggested that aneuploidy serves as a "stepping stone" on the path to adaptation.

### 2022-09-12 results/all_trans1/pyabc-single-w1w1.py
added all transactions db files and MAP and HDI calculation for all transactions

@@ -1,0 +1,186 @@
+import sys
+
+if len(sys.argv)==1:
+    print('usage:\npyabc-single-w1w1.py num_reps seed tau prior_type stop_epsilon dir_name cpus')
+    print('num_reps - the number of simulations that are executed per each parameter vector')
+    print('seed - seed for random number generator that is used for each parameter vector simulations execution')
+    print('tau - mutation rate increase factor in aneuploid cells')
+    print('prior_type (see more description in the manuscript):')
+    print('   1 - informative prior for every fitness')
+    print('   2 - uninformative uniform prior for every fitness')
+    print('   3 - uninformative uniform prior for w(2n+1*) and w(2n*)')
+    print('   4 - extended informative prior')
+    print('   5 - informative prior; fixed mutation rate = 1e-5')
+    print('   6 - informative prior; fixed mutation rate = 1e-6')
+    print('   7 - informative prior; fixed mutation rate = 1e-7')
+    print('epsilon_stop - epsilon stop bound for pyabc run')
+    print('dir_name - the result files are persisted at ~/sim-data/single-model-abc/+dir_name')
+    print('cpus - num of cpu cores to use.')
+    exit()
+
+import pyabc
+import scipy
+import tempfile
+import os
+os.environ["OMP_NUM_THREADS"] = "1" #for numpy
+import numpy as np
+from model1.singleLocusModel import SingleLocusModel
+from datetime import timedelta, datetime
+from shutil import copyfile, copytree, rmtree
+from multiprocessing import  Value
+
+N = 6.425*10**6 # 1.6M - 2004.8M in Yona experiment 8/(1.0/(np.array([2**i for i in range(8)])*1.6)).sum()
+GLOBAL_SEED = 123
+# K = 1.03125
+# A,B,C = 50, 300, 2200
+
+
+
+
+PRIOR_THRESHOLD = 0.01
+MAX_POPULATIONS = 50 # stop criterion: num of abc iterations
+A,B,C = 450, 1700, 2350
+REPS1 = int(sys.argv[1])
+SIM_SEED = int(sys.argv[2])
+K = float(sys.argv[3])
+PRIOR_TYPE = int(sys.argv[4])
+MIN_EPSILON = float(sys.argv[5]) #stop criterion
[…]

### 2022-09-12 results/all_trans2/pyabc-single-w1w1.py

@@ -1,0 +1,186 @@
+import sys
+
+if len(sys.argv)==1:
+    print('usage:\npyabc-single-w1w1.py num_reps seed tau prior_type stop_epsilon dir_name cpus')
+    print('num_reps - the number of simulations that are executed per each parameter vector')
+    print('seed - seed for random number generator that is used for each parameter vector simulations execution')
+    print('tau - mutation rate increase factor in aneuploid cells')
+    print('prior_type (see more description in the manuscript):')
+    print('   1 - informative prior for every fitness')
+    print('   2 - uninformative uniform prior for every fitness')
+    print('   3 - uninformative uniform prior for w(2n+1*) and w(2n*)')
+    print('   4 - extended informative prior')
+    print('   5 - informative prior; fixed mutation rate = 1e-5')
+    print('   6 - informative prior; fixed mutation rate = 1e-6')
+    print('   7 - informative prior; fixed mutation rate = 1e-7')
+    print('epsilon_stop - epsilon stop bound for pyabc run')
+    print('dir_name - the result files are persisted at ~/sim-data/single-model-abc/+dir_name')
+    print('cpus - num of cpu cores to use.')
+    exit()
+
+import pyabc
+import scipy
+import tempfile
+import os
+os.environ["OMP_NUM_THREADS"] = "1" #for numpy
+import numpy as np
+from model1.singleLocusModel import SingleLocusModel
+from datetime import timedelta, datetime
+from shutil import copyfile, copytree, rmtree
+from multiprocessing import  Value
+
+N = 6.425*10**6 # 1.6M - 2004.8M in Yona experiment 8/(1.0/(np.array([2**i for i in range(8)])*1.6)).sum()
+GLOBAL_SEED = 123
+# K = 1.03125
+# A,B,C = 50, 300, 2200
+
+
+
+
+PRIOR_THRESHOLD = 0.01
+MAX_POPULATIONS = 50 # stop criterion: num of abc iterations
+A,B,C = 450, 1700, 2350
+REPS1 = int(sys.argv[1])
+SIM_SEED = int(sys.argv[2])
+K = float(sys.argv[3])
+PRIOR_TYPE = int(sys.argv[4])
+MIN_EPSILON = float(sys.argv[5]) #stop criterion
[…]

### 2022-09-12 ms/ms.tex
fixed estimations for extended prior model, and Fig S7

@@
 We inferred similar parameters estimates for the model with $\mu=10^{-6}$ compared to the model with a free $\mu$ parameter, in which the inferred mutation rate is $\mu \approx 3\cdot10^{-6}$.
-Models with $\mu$=$10^{-5}$ and $\mu$=$10^{-7}$ inferred different parameters estimates, including higher aneuploidy rate and lower fitness values for \anwt\ and \eumt~(\Cref{fig:mu}).
+Models with $\mu$=$10^{-5}$ and $\mu$=$10^{-7}$ inferred different parameters estimates, including higher aneuploidy rate and the model with $\mu$=$10^{-5}$ has lower fitness value for \anwt\ (\Cref{fig:mu}).
 
@@
 % generated with MAP-and-hdi.ipynb
-with $\mu=2.475\cdot10^{-9}\ [2.419\cdot10^{-9}-2.609\cdot10^{-9}]$. Other parameter estimates are: $\delta=2.707\cdot10^{-3}\ [2.093\cdot10^{-3}-3.092\cdot10^{-3}]$,
+with $\mu=2.474\cdot10^{-9}\ [2.423\cdot10^{-9}-2.612\cdot10^{-9}]$. Other parameter estimates are: $$\delta=2.705\cdot10^{-3}\ [2.094\cdot10^{-3}-3.094\cdot10^{-3}]$,
 $w_{\anwt}=1.022\ [1.021-1.024]$,
-$w_{\anmt}=1.029\ [1.027-1.03]$,
-$w_{\eumt}=1.03\ [1.029-1.031]$. 
-Notably, the maximum posterior ratio $w_{\eumt}/w_{\anwt}=1.007$ is much lower than the maximum prior ratio of $1.033$ (\Cref{fig:growth-curves}H) and closer to the ratio of $1$ that we assume in our standard prior.
+$w_{2n+1^*}=1.052\ [1.05-1.054]$,
+$w_{2n^*}=1.053\ [1.051-1.055]$. 
+Notably, the maximum posterior ratio $w_{\eumt}/w_{\anwt}=1.0009$ is much lower than the maximum prior ratio of $1.033$ (\Cref{fig:growth-curves}H) and closer to the ratio of $1$ that we assume in our standard prior.
 Together with the posterior predictive results, we conclude that the main informative prior is preferable over the extended informative prior.
@@
 The inferred values are slightly different.
-% with back transitions
+% with back transitions MAP-and-hdi.ipynb
 The estimated mutation rate, $\mu=1.036\cdot10^{-7}\ [8.01\cdot10^{-8}-1.339\cdot10^{-7}]$, corresponds to a mutation target size of $\sim 300-500$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$ (ref.~\citep{Zhu2014}) or $3.3\cdot10^{-10}$ (ref.~\citep{Lynch2008}).
@@
     \textbf{Model with fixed mutation rate.}  \textbf{(A-D)} The inferred posterior distributions for models with free and fixed mutation rate, $\mu$.
-The MAP (maximum a posteriori) and 50\% HDI (highest density interval) for each model are:
+The MAP (maximum a posteriori) and 50\% HDI (highest density interval) for each model are: %MAP-and-hdi.ipynb
 \textbf{free $\boldsymbol{\mu}$, run 1:}
-$\delta=2.749\cdot10^{-3}\ [1.476\cdot10^{-3}-2.822\cdot10^{-3}]$,
+$\delta=1.720\cdot10^{-3}\ [1.470\cdot10^{-3}-2.786\cdot10^{-3}]$, 
 $w_{\anwt}=1.022\ [1.021-1.023]$,
-$w_{\anmt}=1.025\ [1.023-1.026]$,
-$w_{\eumt}=1.027\ [1.026-1.029]$;
+$w_{\anmt}=1.025\ [1.024-1.026]$,
+$w_{\eumt}=1.028\ [1.026-1.029]$;
 \textbf{free $\boldsymbol{\mu}$, run 2:}
-$\delta=1.938\cdot10^{-3}\ [1.338\cdot10^{-3}-2.748\cdot10^{-3}]$,
-$w_{\anwt}=1.022\ [1.02-1.023]$,
-$w_{\anmt}=1.025\ [1.023-1.026]$,
-$w_{\eumt}=1.027\ [1.026-1.029]$;
+$\delta=2.129\cdot10^{-3}\ [1.334\cdot10^{-3}-2.695\cdot10^{-3}]$,
+$w_{2n+1}=1.022\ [1.02-1.023]$,
+$w_{2n+1^*}=1.025\ [1.023-1.026]$,
+$w_{2n^*}=1.028\ [1.026-1.029]$;
 \textbf{$\boldsymbol{\mu=10^{-5}}$:}
-$\delta=3.089\cdot10^{-3}\ [2.412\cdot10^{-3}-3.169\cdot10^{-3}]$,
-$w_{\anwt}=1.022\ [1.021-1.023]$,
-$w_{\anmt}=1.024\ [1.023-1.026]$,
-$w_{\eumt}=1.027\ [1.026-1.028]$;
+$\delta=2.903\cdot10^{-3}\ [2.399\cdot10^{-3}-3.156\cdot10^{-3}]$,
+$w_{2n+1}=1.022\ [1.021-1.023]$,
+$w_{2n+1^*}=1.024\ [1.023-1.025]$,
+$w_{2n^*}=1.027\ [1.026-1.028]$;
 \textbf{$\boldsymbol{\mu=10^{-6}}$:}
-$\delta=1.413\cdot10^{-3}\ [1.04\cdot10^{-3}-2.529\cdot10^{-3}]$,
-$w_{\anwt}=1.021\ [1.02-1.023]$,
-$w_{\anmt}=1.024\ [1.023-1.026]$,
-$w_{\eumt}=1.028\ [1.026-1.029]$;
+$\delta=1.917\cdot10^{-3}\ [9.624\cdot10^{-4}-2.447\cdot10^{-3}]$,
+$w_{2n+1}=1.022\ [1.02-1.023]$,
+$w_{2n+1^*}=1.025\ [1.023-1.026]$,
+$w_{2n^*}=1.028\ [1.027-1.029]$;
 \textbf{$\boldsymbol{\mu=10^{-7}}$:}
-$\delta=3.4\cdot10^{-3}\ [2.043\cdot10^{-3}-3.578\cdot10^{-3}]$,
-$w_{\anwt}=1.019\ [1.017-1.02]$,
-$w_{\anmt}=1.026\ [1.024-1.027]$,
-$w_{\eumt}=1.027\ [1.026-1.029]$.
+$\delta=2.901\cdot10^{-3}\ [2.139\cdot10^{-3}-3.671\cdot10^{-3}]$,
+$w_{2n+1}=1.019\ [1.017-1.02]$,
+$w_{2n+1^*}=1.025\ [1.024-1.026]$,
+$w_{2n^*}=1.027\ [1.026-1.029]$.
 \label{fig:mu}

### 2022-09-13 ms/ms.tex
revise following IK comments:

@@
 This is because, without aneuploidy, a high mutation rate or fitness effect will lead to faster appearance and fixation of \eumt\ than in the experimental observations.
-Even with these lower estimates, the model fit is worse than that of a model with aneuploidy~(\Cref{fig:2n*-fixation}).
 
@@
 Changing a single parameter while keeping the rest fixed at the MAP estimate produces a worse fit to the data~(\Cref{fig:sensitivity}).
-Furthermore, we fitted models with a mutation rate fixed at $\mu$=$10^{-5}$, $10^{-6}$ and $10^{-7}$.
+Furthermore, we fitted models with a mutation rate fixed at $\mu=10^{-5}$, $10^{-6}$ and $10^{-7}$.
 We inferred similar parameters estimates for the model with $\mu=10^{-6}$ compared to the model with a free $\mu$ parameter, in which the inferred mutation rate is $\mu \approx 3\cdot10^{-6}$.
-Models with $\mu$=$10^{-5}$ and $\mu$=$10^{-7}$ inferred different parameters estimates, including higher aneuploidy rate and the model with $\mu$=$10^{-5}$ has lower fitness value for \anwt\ (\Cref{fig:mu}).
+Inference assuming $\mu=10^{-5}$ or $\mu=10^{-7}$ produced a higher estimate of the aneuploidy rate, $\delta$; assuming $\mu=10^{-7}$, the estimated fitness of \anwt\ was lower; otherwise, estimates were similar to our main model (\Cref{fig:mu}).
 
@@
 % generated with MAP-and-hdi.ipynb
-with $\mu=2.474\cdot10^{-9}\ [2.423\cdot10^{-9}-2.612\cdot10^{-9}]$. Other parameter estimates are: $$\delta=2.705\cdot10^{-3}\ [2.094\cdot10^{-3}-3.094\cdot10^{-3}]$,
+with $\mu=2.474\cdot10^{-9}\ [2.423\cdot10^{-9}-2.612\cdot10^{-9}]$. Other parameter estimates are: $\delta=2.705\cdot10^{-3}\ [2.094\cdot10^{-3}-3.094\cdot10^{-3}]$,
 $w_{\anwt}=1.022\ [1.021-1.024]$,
 $w_{2n+1^*}=1.052\ [1.05-1.054]$,
-$w_{2n^*}=1.053\ [1.051-1.055]$. 
-Notably, the maximum posterior ratio $w_{\eumt}/w_{\anwt}=1.0009$ is much lower than the maximum prior ratio of $1.033$ (\Cref{fig:growth-curves}H) and closer to the ratio of $1$ that we assume in our standard prior.
+$w_{2n^*}=1.053\ [1.051-1.055]$, the latter two being much higher compare to the main informative prior. 
+Notably, the mode of the posterior ratio $w_{\eumt}/w_{\anwt}=1.0009$ is much lower than the mode of the prior ratio of $1.033$ (\Cref{fig:growth-curves}H) and closer to the ratio of $1$ that we assume in the main informative prior.
 Together with the posterior predictive results, we conclude that the main informative prior is preferable over the extended informative prior.
 
-\subsection*{Supplementary Results}
+\paragraph{Model with transitions to less-fit genotypes}
 

### 2022-09-13 ms/ms.tex
small revision based on IK comment

@@
 We inferred similar parameters estimates for the model with $\mu=10^{-6}$ compared to the model with a free $\mu$ parameter, in which the inferred mutation rate is $\mu \approx 3\cdot10^{-6}$.
-Inference assuming $\mu=10^{-5}$ or $\mu=10^{-7}$ produced a higher estimate of the aneuploidy rate, $\delta$; assuming $\mu=10^{-7}$, the estimated fitness of \anwt\ was lower; otherwise, estimates were similar to our main model (\Cref{fig:mu}).
+Inference assuming $\mu=10^{-5}$ or $\mu=10^{-7}$ produced similar estimates except that the estimated aneuploidy rate, $\delta$, was higher, and assuming $\mu=10^{-7}$, the estimated fitness of \anwt\ was lower (\Cref{fig:mu}).
 

### 2022-09-16 ms/ms.tex
added first ideas about limitations

@@
-\documentclass[12pt]{extarticle}
+\documentclass[12pt, draft]{extarticle}
 \usepackage{geometry}
@@
 This surprising result reinforces the importance of models when making interpretations on evolutionary processes, and emphasizes the unintuitive outcomes of clonal interference during adaptive evolution.
+
+% Limitations
+\paragraph{Limitations and future directions}
+The number of measurement of the frequency of aneuploidy in \textit{S.cervisiae} is limited to three time points (0, \~450 and \~2000).
+
+The populations had fluctuating populations sizes due to dilution. We use the harmonic mean to determine a single number, on which all of the inference and simulations is based.
+
+Mathematical model is very simplistic. One mutation of fixed effect, at a single locus that accounts for all the benefit. Haploid model where yest is diploid and the aneuploidy is trisomie.
 

### 2022-09-18 ms/ms.tex
MP small comments

--- 2022-09-13 ms/ms.tex
@@
 
-The model includes the effects of natural selection, genetic drift, aneuploidy, and mutation, and follows a population of cells characterized by both their genotype: 
+The model includes the effects of natural selection, genetic drift, aneuploidy, and mutation, and follows a population of cells characterized by their genotype: 
 euploid wild-type, \euwt, is the ancestral diploid genotype; 
@@
 % without back transitions:
-The estimated mutation rate, $\mu=2.965\cdot10^{-6}\ [2.718\cdot10^{-7}-3.589\cdot10^{-6}]$, corresponds to a mutation target size of $\sim 10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$ (ref.~\citep{Zhu2014}) or $3.3\cdot10^{-10}$ (ref.~\citep{Lynch2008}).
+The estimated beneficial mutation rate, $\mu=2.965\cdot10^{-6}\ [2.718\cdot10^{-7}-3.589\cdot10^{-6}]$, corresponds to a mutation target size of $\sim 10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$ (ref.~\citep{Zhu2014}) or $3.3\cdot10^{-10}$ (ref.~\citep{Lynch2008}).
 The estimated aneuploidy rate, $\delta=1.72\cdot10^{-3}\ [1.47\cdot10^{-3}-2.786\cdot10^{-3}]$ is higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
@@
 However, in the experiment of \citet{Yona2012}, both the wild-type and the aneuploid were under heat stress, which may explain why we did not find evidence for an increased mutation rate.
+
+% TODO caveats - MP
 

### 2022-09-18 ms/ms.tex
revised after AY comments

@@
 Aneuploidy is common in eukaryotes, often leading to decreased growth and fitness.
-However, evidence from yeast and fungi, as well as human tumour cells, suggests that aneuploidy can be beneficial under stressful conditions and facilitate  adaptation.
-In a prominent example, an evolutionary experiment with yeast, populations evolving under heat stress had become aneuploid, only to later revert back to euploid after genetic mutations have accumulated.
+However, evidence from yeast and fungi, as well as human tumour cells, suggests that specific aneuploidies can be beneficial under stressful conditions and facilitate  adaptation.
+In a prominent example, an evolutionary experiment with yeast, populations evolving under heat stress had become aneuploid (chromosome III), only to later revert back to euploid after genetic mutations have accumulated.
 It has therefore been suggested that aneuploidy serves as a "stepping stone" on the path to adaptation.
+
 Here, we test this hypothesis. First, we apply DNA sequencing to show that mutant alleles common in aneuploid cells are uncommon in the evolved euploid population. 
@@
 We then predict the genotype frequency dynamics during the experiment, demonstrating that the majority of the evolved euploid population likely did not descend from aneuploid cells, but rather directly from the euploid wild-type population.
-Together, our results suggest that aneuploidy is an evolutionary "detour" rather than a "stepping stone": it delays rather than facilitates the adaptation of the population, and cells that become aneuploid leave less descendants compared to cells that remain diploid.
+Our model further predicts that if the experiment was repeated with smaller populations, then a larger fraction of the evolved population would descend from aneuploid cells.
+
+Together, our results suggest that aneuploidy can be an evolutionary "detour" rather than a "stepping stone": it can delay, rather than facilitate, the adaptation of the population, and cells that become aneuploid may leave less descendants compared to cells that remain diploid.
 \end{abstract}
@@
 %\paragraph{Aneuploidy is generally deleterious.}
-Experiments with human and mouse embryos found that aneuploidy is usually lethal.
+Experiments with human and mouse embryos found that most aneuploidies are lethal.
 It is also associated with developmental defects and lethality in other multicellular organisms~\citep{Sheltzer2011}.
@@
 Here, we test the hypothesis that aneuploidy is a an \emph{evolutionary stepping stone} that facilitates adaptive evolution by genetic mutations.
-First, we analyze previously unpublished sequencing data from the original experimental populations of \citet{Yona2012} to assess if the evolved euploid population is descended from the aneuploid population.
+First, we analyze previously unpublished sequencing data from the original experimental populations \citep{Yona2012} to assess if the evolved euploid population is descended from the aneuploid population.
 Second, we develop an evolutionary genetic model and fit it to the experimental results of \citet{Yona2012} in order to predict the genotype frequency dynamics in the experimental populations, thereby estimating the frequency of evolved euploid cells that descended from aneuploid cells.
@@
 Two of the repetitions, marked $H2$ and $H4$, carried no large-scale duplications other than a chromosome III trisomy.
-These two repetitions continued to evolve under the same conditions, wherein aneuploidy was eliminated by generation 1,700 and 2,350.
+These two repetitions continued to evolve under the same conditions, wherein aneuploidy was eliminated by generation 1,700 and 2,350 in $H4$ and $H2$, respectively.
 
@@
 $w_{\eumt}=1.028\ [1.026-1.029]$, all relative to the fitness of \euwt, which is set to $w_{\euwt}=1$. 
-Thus, we can infer that the cost of trisomy is $c = w_{\eumt}-w_{\anmt} = 0.003$ (or 0.3\%) and the benefit of trisomy is $w_{\anwt}-1-c=0.019$ (1.9\%), whereas the benefit of the beneficial mutation is $w_{\eumt}-1=0.028$ (2.8\%).
+Thus, we can infer that the cost of chromosome III trisomy is $c = w_{\eumt}-w_{\anmt} = 0.003$ (or 0.3\%) and the benefit of trisomy is $w_{\anwt}-1-c=0.019$ (1.9\%), whereas the benefit of the beneficial mutation is $w_{\eumt}-1=0.028$ (2.8\%).
 
@@
   \centering
-\includegraphics[width=0.9\textwidth]{../figures/posterior.pdf}
+\includegraphics[width=0.8\textwidth]{../figures/posterior.pdf}
   \caption{
@@
 WAIC (lower is better, see Methods) is lowest for $\tau=33/32$ and $\tau=1$ (\Cref{table:WAIC}).
-Therefore, evidence does not support an increase in mutation rate in aneuploid cells, and moreover, unless the increase is strong ($\tau \ge 10$), it does not seem to affect our inference.
+Therefore, our results do not support an increase in mutation rate in aneuploid cells.
+This may be because, unless the increase is strong ($\tau \ge 10$), it does not seem to affect our inference; or because chromosome III is one of the smallest chromosomes \citep{Gilchrist2019}.
 We also checked the differences in genotype frequency dynamics for different $\tau$ values.
@@
 \paragraph{DNA sequencing.}
-BLA BLA BLA
-% TODO AY
+Whole-genome sequencing of the ancestral diploid strain (\euwt) was performed on a sample generated from a single colony of the ancestor.
+Whole-genome sequencing of the four evolving populations ($H2$ after 450 and 2,350 generations, and $H4$ after 450 and 1,700 generations) was performed at the population level.
+A liquid sample from each evolving population was taken for sequencing from each culture as it grows under the same conditions as the lab evolution experiments (YPD at \SI{39}{\celsius}).
+DNA was purified from the collected cells for each population and downstream deep sequencing was performed according to the Illumina protocol for amplicon sequencing.
 
@@
 \caption*{
-	Mutant alleles identified in the ancestor (generation 0), aneuploid (generation 450), and evolved (generation 1,700) of population $H2$. See supplementary file.
+	Mutant alleles identified in the ancestor (generation 0), aneuploid (generation 450), and evolved (generation 2,350) of population $H2$. See supplementary file.
 } 
@@
 \caption*{
-	Mutant alleles identified in the ancestor (generation 0), aneuploid (generation 450), and evolved (generation 2,350) of population $H4$. See supplementary file.
+	Mutant alleles identified in the ancestor (generation 0), aneuploid (generation 450), and evolved (generation 1,700) of population $H4$. See supplementary file.
 } 

### 2022-09-19 ms/ms.tex
small revisions with AY

@@
 % Chromosomal duplication is a transient evolutionary solution to stress
-\title{Chromosomal duplication can be an evolutionary detour on the path to adaptation}
+% \title{Chromosomal duplication can be an evolutionary detour on the path to adaptation}
+\title{Aneuploidy can be an evolutionary detour \\ on the path to adaptation}
 
@@
 
-Keywords: aneuploidy, evolutionary model, adaptive evolution
+Keywords: whole-chromosome duplication, evolutionary model, adaptive evolution
 
@@
 
-Together, our results suggest that aneuploidy can be an evolutionary "detour" rather than a "stepping stone": it can delay, rather than facilitate, the adaptation of the population, and cells that become aneuploid may leave less descendants compared to cells that remain diploid.
+Together, our results suggest that aneuploidy can be an inevitable evolutionary "detour" rather than a "stepping stone": it can delay, rather than facilitate, the adaptation of the population, and cells that become aneuploid may leave less descendants compared to cells that remain diploid.
 \end{abstract}
@@
 We estimate that only about 10-15\% of the euploid cells descended from aneuploid cells, whereas the rest are direct descendants of the wild-type euploid cells.
+Thus, aneuploidy can delay, rather than accelerate, adaptation, and and cells that become aneuploid may leave less descendants than cells that remain euploid.
 This surprising result reinforces the importance of models when making interpretations on evolutionary processes, and emphasizes the unintuitive outcomes of clonal interference during adaptive evolution.
@@
 \paragraph{DNA sequencing.}
-Whole-genome sequencing of the ancestral diploid strain (\euwt) was performed on a sample generated from a single colony of the ancestor.
-Whole-genome sequencing of the four evolving populations ($H2$ after 450 and 2,350 generations, and $H4$ after 450 and 1,700 generations) was performed at the population level.
-A liquid sample from each evolving population was taken for sequencing from each culture as it grows under the same conditions as the lab evolution experiments (YPD at \SI{39}{\celsius}).
+Whole-genome sequencing of the ancestral diploid strain (\euwt) was performed on a sample from a single colony of the ancestor.
+Whole-genome sequencing of the four evolving populations ($H2$ after 450 and 2,350 generations, and $H4$ after 450 and 1,700 generations) was performed on a sample from these populations (rather than from single colonies).
+The liquid sample from each evolving population was taken from each culture as it grows under the same conditions as the lab evolution experiments (YPD at \SI{39}{\celsius}).
 DNA was purified from the collected cells for each population and downstream deep sequencing was performed according to the Illumina protocol for amplicon sequencing.

### 2022-09-20 ms/ms.tex
Update ms.tex

--- 2022-09-18 ms/ms.tex
@@
 
-The model includes the effects of natural selection, genetic drift, aneuploidy, and mutation, and follows a population of cells characterized by their genotype: 
+The model includes the effects of natural selection, genetic drift, aneuploidy, and mutation, and follows a population of cells characterized by both their genotype: 
 euploid wild-type, \euwt, is the ancestral diploid genotype; 
@@
 % without back transitions:
-The estimated beneficial mutation rate, $\mu=2.965\cdot10^{-6}\ [2.718\cdot10^{-7}-3.589\cdot10^{-6}]$, corresponds to a mutation target size of $\sim 10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$ (ref.~\citep{Zhu2014}) or $3.3\cdot10^{-10}$ (ref.~\citep{Lynch2008}).
+The estimated mutation rate, $\mu=2.965\cdot10^{-6}\ [2.718\cdot10^{-7}-3.589\cdot10^{-6}]$, corresponds to a mutation target size of $\sim 10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$ (ref.~\citep{Zhu2014}) or $3.3\cdot10^{-10}$ (ref.~\citep{Lynch2008}).
 The estimated aneuploidy rate, $\delta=1.72\cdot10^{-3}\ [1.47\cdot10^{-3}-2.786\cdot10^{-3}]$ is higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
@@
 
-% TODO caveats - MP
-
 % Conclusions
@@
 This surprising result reinforces the importance of models when making interpretations on evolutionary processes, and emphasizes the unintuitive outcomes of clonal interference during adaptive evolution.
+
+% Limitations
+\paragraph{Limitations and future directions}
+The number of measurement of the frequency of aneuploidy in \textit{S.cervisiae} is limited to three time points (0, \~450 and \~2000).
+A more thorough time series could improve inference of model parameters.
+
+The mathematical model is rather simplistic.
+It assumes one mutation of fixed effect, at a single locus that accounts for all the benefit.
+In the original experiment, the replicates had fluctuating population sizes due to dilution.
+We use the harmonic mean to determine a single number, on which all of the inference and simulations is based.
+Extending the model and the simulations to include fluctuating population sizes and adding other features (e.g.
+dominance, more than one locus) should be done in the future as it is beyond the scope of the current paper.
 

### 2022-10-02 ms/ms.tex
updated manuscript
target size discussion

--- 2022-09-19 ms/ms.tex
@@
 % without back transitions:
-The estimated beneficial mutation rate, $\mu=2.965\cdot10^{-6}\ [2.718\cdot10^{-7}-3.589\cdot10^{-6}]$, corresponds to a mutation target size of $\sim 10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$ (ref.~\citep{Zhu2014}) or $3.3\cdot10^{-10}$ (ref.~\citep{Lynch2008}).
+The estimated beneficial mutation rate is $\mu=2.965\cdot10^{-6}\ [2.718\cdot10^{-7}-3.589\cdot10^{-6}]$.
+From the literature, the mutation rate per base pair is roughly $2-3\cdot10^{-10}$ (refs.~\citep{Zhu2014,Lynch2008}), but it may be higher under heat stress, as several stresses may cause hypermutation in yeast~\citep{Heidenreich2007}.
+If we assume a 10-fold increase in mutation rate, then the estimated beneficial mutation rate can be explained by a target size of 1,000 base pairs.
+\citet{Yona2012} found at least 10 genes on chromosome III with mutations that led to increased heat tolerance.
+Assuming that other chromosomes also have a similar number of heat-tolerance genes (i.e., chromosome III is one of the smaller chromosomes~\citep{Gilchrist2019}), and that each gene has a target size of ten base pairs, we easily get to a target size above 1,600.
+
 The estimated aneuploidy rate, $\delta=1.72\cdot10^{-3}\ [1.47\cdot10^{-3}-2.786\cdot10^{-3}]$ is higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
+However, this difference may be explained by an increased aneuploidy rate during heat stress~\citep{Chen2012a}.
+
 The estimated fitness values are $w_{\anwt}=1.022\ [1.021-1.023]$,
@@
 Previous empirical studies have suggested that genetic instability (e.g., elevated mutation rates) in aneuploid cells is due to stress associated with the aneuploid state~\citep{Bouchonville2009, Chen2012b, Zhu2012, Ippolito2021b}.
-However, in the experiment of \citet{Yona2012}, both the wild-type and the aneuploid were under heat stress, which may explain why we did not find evidence for an increased mutation rate.
-
-% TODO caveats - MP
+However, in the experiment of \citet{Yona2012}, both the wild-type and the aneuploid were under heat stress, which may explain why we did not find evidence for an increased mutation rate specifically in aneuploid cells.
 
@@
 	Appearance of \anmt, and therefore \eumtA, is stochastic, however, the frequency dynamics are deterministic above a frequency of roughly 0.001. 
+	Note that the \eumtM\ and the \eumtA+\eumtM\ lines are overlapping for much of their trajectories.
 	}

### 2022-10-13 ms/ms.tex
update authors, acks, affil

@@
 
-\author[1,2,*]{Ilia Kohanovski}
-\author[1,*]{Martin Pontz}
-\author[3]{Avihu H. Yona}
-\author[1,$\dagger$]{Yoav Ram}
+\author[a,b,*]{Ilia Kohanovski}
+\author[a,*]{Martin Pontz}
+\author[c]{Orna Dahan}
+\author[c]{Yitzhak Pilpel}
+\author[d]{Avihu H. Yona}
+\author[a,1]{Yoav Ram}
 
-\affil[1]{School of Zoology, Faculty of Life Sciences, Tel Aviv University, Tel Aviv, Israel}
-\affil[2]{School of Computer Science, Reichman University, Herzliya, Israel}
-\affil[3]{Institute of Biochemistry, Food Science and Nutrition,
+\affil[a]{School of Zoology, Faculty of Life Sciences, Tel Aviv University, Tel Aviv, Israel}
+\affil[b]{School of Computer Science, Reichman University, Herzliya, Israel}
+\affil[c]{Department of Molecular Genetics, Weizmann Institute of Science, Rehovot 76100, Israel}
+\affil[d]{Institute of Biochemistry, Food Science and Nutrition,
 Robert H. Smith Faculty of Agriculture, Food and Environment,
@@
 \affil[*]{These authors contributed equally to this work}
-\affil[$\dagger$]{Corresponding author: yoav@yoavram.com}
+\affil[1]{Corresponding author: yoav@yoavram.com}
 
@@
 \section*{Acknowledgements}
-We thank Yitzhak Pilpel, Orna Dahan, Lilach Hadany, Judith Berman, David Gresham, Shay Covo, Martin Kupiec, Uri Obolski, Daniel Weissman, and Tal Simon for discussions and comments.
+We thank Lilach Hadany, Judith Berman, David Gresham, Shay Covo, Martin Kupiec, Uri Obolski, Daniel Weissman, and Tal Simon for discussions and comments.
 This work was supported in part by 

### 2022-10-13 ms/ms.tex
added FA vs mu

@@
 To test the hypothesis that aneuploidy facilitates adaptation, we estimated $F_A$, the expected frequency of \eumt\ that arose from \anwt, computed as the average frequency of such \eumtA\ cells at the end of simulations using the MAP estimate parameters.
-Surprisingly, we observe that the majority of \eumt\ cells are \eumtM, a product of a direct mutation in \euwt\ cells, rather than descending from \anwt\ cells ($F_A^{MAP} = 0.106$, \Cref{fig:FA}A). 
+Surprisingly, we observe that the majority of \eumt\ cells are \eumtM, a product of a direct mutation in \euwt\ cells, rather than descending from \anwt\ cells ($F_A^{MAP} = 0.106$, average end point of 50 purple lines in \Cref{fig:FA}A).
 This is despite the fact that the \anwt\ genotype reaches high frequencies in the population (at least 0.98, \Cref{fig:FA}A).
@@
 This result is not unique to the MAP parameter estimate. 
-We simulated genotype frequency dynamics using parameter samples from the posterior distribution (\Cref{fig:posterior}), and computed the posterior distribution of $F_A$ (\Cref{fig:FA}B). 
-The mean $F_A$ was just 0.1673 [0.0154-0.370 95\% CI] and only in 489 of 100,000 posterior samples (0.489\%) $F_A$ was  larger than 0.5 (see Supporting Material for results when transitions to less-fit genotypes are allowed, such as \eumt\ to \anmt).
+We simulated genotype frequency dynamics using parameter samples from the posterior distribution, and computed the posterior distribution of $F_A$ (\Cref{fig:FA}B). 
+The posterior mode $F_A$ was just 0.147 [0.0154-0.370 95\% CI] and only in 489 of 100,000 posterior samples (0.489\%) $F_A$ was  larger than 0.5 (see Supporting Material for results when transitions to less-fit genotypes are allowed, such as \eumt\ to \anmt).
 Thus, if we sample a random cell from the evolved \eumt\ population, it is more likely to have descended directly from an euploid cell than from an aneuploid cell.
-The probability of \eumt\ descending from \anwt\ ($F_A$) increases with the aneuploidy rate, $\delta$, and decreases with the mutation rate, $\mu$, and population size $N$ (\Cref{fig:FA}C,D).
+The probability of \eumt\ descending from \anwt\ ($F_A$) increases with the aneuploidy rate, $\delta$, and decreases with both the population size $N$ and the mutation rate, $\mu$ (\Cref{fig:FA}C,D).
 In some cases it can also be affected by the fitness parameters (\Cref{fig:posterior_FA_supp}).
@@
 % Fig dynamics
-\begin{figure}[h!]
+\begin{figure}[p]
       \centering
@@
     \textbf{(B)} Posterior distribution of $F_A$, the expected frequency of \eumt\ cells descended from \anwt\ cells, computed as the average frequency at the end of 100 simulations for 100,000 samples from the parameter posterior distribution.
-    Solid and dashed lines show the mean and 95\% CI.
+    Solid and dashed lines show the mode and 95\% CI.
 	\textbf{(C)} $F_A$ values (color coded) from panel B, with their corresponding mutation rate $\mu$ on x-axis and aneuploidy rate $\delta$ on the y-axis.
 	White star shows the MAP estimate. See also \Cref{fig:posterior_FA_supp}.
-	\textbf{(D)} $F_A$ as a function of the population size, $N$, in posterior predictions with MAP parameters.
-Markers show $F_A$ in 250 simulations per population size.
-Error bars show mean $F_A$ with 95\% CI (bootstrap, $n=10,000$).
-Vertical dashed line for population size in the experiment, $6.425\cdot10^6$.
-Horizontal line for $F_A^{MAP}=0.106$.
+	\textbf{(D)} $F_A$ as a function of the population size ($N$, bottom x-axis) and the beneficial mutation rate ($\mu$, top x-axis) in posterior predictions with MAP parameters. 
+	Markers show $F_A$ in 250 simulations per population size or mutation rate value.
+	Error bars show mean $F_A$ with 95\% CI (bootstrap, $n=10,000$). Blue and red bars for varying population size and mutation rate, respectively.
+	Vertical dashed line for population size in the experiment, $6.425\cdot10^6$, and the MAP mutation rate, $2.965\cdot10^{-6}$. Horizontal line for $F_A^{MAP}=0.106$. 
 	\textbf{(E)} Population mean fitness in a model without drift using MAP estimate parameters.
@@
 
-% population size
-\paragraph{Predictions for small populations.}
-We examined the effect of the population size, $N$, on the frequency of \anwt\ descendants in the evolved population, $F_A$.
-We found that $F_A$ is expected to decrease as the population size increases (\Cref{fig:FA}D), ranging from about 90\% when the population size is 10,000, to about 10\% when the population size is above 1,000,000 (less than the experimental population size, which was 6,425,000).
-Thus, our model provides a testable prediction: if the experiment was repeated under a lower population size (via stronger daily dilutions or in a smaller volume), than the fraction of the population descending from aneuploid cells would be much higher.
+% mutation supply
+\paragraph{Predictions for small populations and low mutation rates.}
+We examined the effect of the population size, $N$, and the beneficial mutation rate, $\mu$, on the frequency of \anwt\ descendants in the evolved population, $F_A$.
+We found that $F_A$ is expected to decrease as the population size or mutation rate increase (\Cref{fig:FA}D), ranging from >90\% when the population size is 10,000 or the mutation rate is $6\cdot10^{-9}$, to about 10\% when the population size is above 1,000,000 (less than the experimental population size, which was 6,425,000) or the mutation rate is above $2\cdot10^{-6}$ (less than the inferred mutation rate, which is $2.965\cdot10^{-6}$).
+Thus, our model provides a testable prediction: if the experiment was repeated under a lower population size (via stronger daily dilutions or in a smaller volume) or a lower mutation rate (via a non-mutagenic stress or stress with a smaller target size), then the fraction of the population descending from aneuploid cells would be much higher.
 
@@
 We inferred the rates of aneuploidy and mutation and their effects on fitness. 
-We estimate that the aneuploidy rate (i.e., number of chromosome gains per generation) is $1.7 \cdot 10^{-3}$, higher than a previous estimate of $6.7 \cdot 10^{-6}$~\citep{Zhu2016}.
+We estimate that the aneuploidy rate (i.e., number of chromosome gains per generation) is $1.7 \cdot 10^{-3}$, higher than a previous estimate of $6.7 \cdot 10^{-6}$ (ref~~\citep{Zhu2016}).
 This may be due to genetic instability caused by heat stress~\citep{Chen2012a}.

### 2022-10-30 ms/ms.tex
OD revised sequencing paragraph

@@
 \paragraph{DNA sequencing.}
-Whole-genome sequencing of the ancestral diploid strain (\euwt) was performed on a sample from a single colony of the ancestor.
-Whole-genome sequencing of the four evolving populations ($H2$ after 450 and 2,350 generations, and $H4$ after 450 and 1,700 generations) was performed on a sample from these populations (rather than from single colonies).
-The liquid sample from each evolving population was taken from each culture as it grows under the same conditions as the lab evolution experiments (YPD at \SI{39}{\celsius}).
-DNA was purified from the collected cells for each population and downstream deep sequencing was performed according to the Illumina protocol for amplicon sequencing.
+Whole-genome sequencing of the ancestral diploid strain (\euwt) was performed
+on a single colony of the ancestor.
+Whole-genome sequencing of the four evolving populations ($H2$ after 450 and 2,350 generations, and $H4$ after 450 and 1,700 generations) was performed on a sample from these populations (rather than from single colonies) in order to maintain the population diversity.
+Cells were grown in 5ml of YPD medium, either at \SI{30}{\celsius} (ancestral diploid) or \SI{39}{\celsius} (evolved populations) in shaking conditions (200rpm) until reaching stationary phase.
+Following growth, 3ml of each culture were centrifuge (14,000rpm) and cell pellets were used for DNA extraction.
+Genomic DNA was extracted using “MasterPure Yeast DNA Purification Kit” (Lucigen) according to the manufacture instructions.
+Following extraction, DNA concentrations were determined by Qubit assay (Thermo Fisher) and ~1µg DNA was used for library preparation using Illumina sample preparation kit (Illumina).
+Samples were sequenced using a 100 bp pair end read output run using Illumina HiSeq2500.
 

### 2022-10-30 ms/ms.tex
OD comments - still some left in TODOs

@@
 Evidence suggests aneuploidy is very common in eukaryotes, e.g. animals~\citep{Santaguida2015review, Naylor2016, Bakhoum2017}, and fungi~\citep{Pavelka2010, Zhu2016, Robbins2017, Todd2017}.
-Aneuploidy has been implicated in cancer formation, progression, and resistance~\citep{Boveri2008, Schvartzman2010, Santaguida2015review, Ippolito2021b}.
+Aneuploidy has been implicated in cancer formation, progression, and drug resistance~\citep{Boveri2008, Schvartzman2010, Santaguida2015review, Ippolito2021b}.
 It is also common in protozoan pathogens of the Leishmania genus, a major global health concern~\citep{Mannaert2012}, and contributes to the emergence of drug resistance~\citep{Selmecki2009} and virulence~\citep{Moller2018} in fungal pathogens, which are under-studied~\citep{Rodrigues2018}, despite infecting a billion people per year, causing significant morbidity in >150 million and death in >1.5 million people per year~\citep{Selmecki2009, Rodrigues2018}.
@@
 Experiments with human and mouse embryos found that most aneuploidies are lethal.
-It is also associated with developmental defects and lethality in other multicellular organisms~\citep{Sheltzer2011}.
+It is also associated with developmental defects and lethality in other multicellular organisms~\citep{Sheltzer2011}. % trisomy 21 and Klinefelter and Turner's syndromes
 For example, aneuploid mouse embryonic cells grow slower than euploid cells~\citep{Williams2008}.
@@
 However, aneuploidy can be beneficial under stressful conditions due to the wide range of phenotypes it can produce, some of which are advantageous~\citep{Pavelka2010, Yang2021}.
+Indeed, in a survey of 1,011 yeast strain, aneuploidy has been detected in about 19\%\citep{Peter2018}.
 Thus, aneuploidy can lead to rapid adaptation in unicellular eukaryotes~\citep{Gerstein2015,Torres2010, Hong2014, Rancati2008}, as well as to rapid growth of somatic tumour cells~\citep{Schvartzman2010, Sheltzer2017}.
@@
 Here, we test the hypothesis that aneuploidy is a an \emph{evolutionary stepping stone} that facilitates adaptive evolution by genetic mutations.
-First, we analyze previously unpublished sequencing data from the original experimental populations \citep{Yona2012} to assess if the evolved euploid population is descended from the aneuploid population.
+First, we sequenced the genomes of evolved populations reported in \citep{Yona2012} and analyzed their mutant allele frequencies to assess if the evolved euploid cells are descended from aneuploid cells.
 Second, we develop an evolutionary genetic model and fit it to the experimental results of \citet{Yona2012} in order to predict the genotype frequency dynamics in the experimental populations, thereby estimating the frequency of evolved euploid cells that descended from aneuploid cells.
-Our results show that aneuploidy reached high frequencies in the experimental populations, but nevertheless, the majority of cells in the evolved euploid populations likely did not descend from aneuploid cells, but rather directly from wild-type euploid cells.
-These results suggest that aneuploidy can be an evolutionary detour, rather than a stepping stone, on the path to adaptation.
+Our results show that aneuploidy reached high frequencies in the experimental populations, but nevertheless, the majority of cells in the evolved euploid population likely did not descend from aneuploid cells, but rather directly from wild-type euploid cells.
+These results suggest that at the lineage level, aneuploidy may be an evolutionary detour, rather than a stepping stone, on the path to adaptation.
 
@@
 \paragraph{Empirical frequencies of mutant alleles.} 
-For each of two evolved populations ($H2$ and $H4$) we sequenced the ancestral population (generation 0), the aneuploid population (generation 450), and the evolved euploid population (generation 1,700 or 2,350) to estimate the mutant allele frequencies (\Cref{table:mutations_H2,table:mutations_H4}).
+For each of two evolved populations ($H2$ and $H4$) we sequenced the ancestral diploid population (generation 0), the aneuploid population (generation 450), and the evolved euploid population (generation 1,700 or 2,350) to estimate the mutant allele frequencies (\Cref{table:mutations_H2,table:mutations_H4}).
 Overall, between 100 and 173 mutant alleles were detected with at least a single read in the six populations that were sampled. 
@@
 Surprisingly, out of all these mutant alleles, none was present at a frequency >20\% in both the aneuploid and the evolved euploid populations. 
-Furthermore, a high mutant allele frequency in the aneuploid population implies a low frequency in the evolved euploid population, and vice-versa (Spearman's correlation coefficient $\rho=-0.64$ and $-0.66$ in the two experimental repetitions; \Cref{fig:mutation_frequencies}), such that mutant alleles frequent in the aneuploid populations decreased in frequency when aneuploidy was lost.
-Moreover, for the 18 mutant alleles with high frequency in the aneuploid populations (>20\%), the highest frequencies in the euploid populations were 15.4\%, 16\%, 16.3\% and 19.6\% (the rest were below 15\%).
-Similarly, for the 48 mutant alleles with high frequency in the evolved euploid populations, the highest frequencies in the aneuploid populations were 2.7\%, 7.7\%, and 11.1\% (the rest were below 1\%).
+More importantly, a high mutant allele frequency in the aneuploid population was associated with a low frequency in the evolved euploid population, and vice-versa (Spearman's correlation coefficient $\rho=-0.64$ and $-0.66$ in the two experimental repetitions; \Cref{fig:mutation_frequencies}), suggesting that mutant alleles frequent in the aneuploid populations decreased in frequency when aneuploidy was lost.
 These results suggest evolved euploid cells are unlikely to descend from aneuploid~cells.
@@
 Solid black line is a linear orthogonal distance regression line (slope=$-0.559$, intercept=$0.164$; a regression through alleles that reach at least 20\% in one of the populations has slope=$-0.645$ and intercept=$0.297$).
-Dashed vertical and horizontal lines show allele frequencies 20\%. Error bars show standard error of the mean accounting for the number of reads.
+Dashed vertical and horizontal lines show allele frequencies 20\%.
+Error bars show standard error of the mean accounting for the number of reads, which may be large when the total number of reads is small.
+For the 18 mutant alleles with high frequency in the aneuploid populations (>20\%), the highest frequencies in the euploid populations were 15.4\%, 16\%, 16.3\% and 19.6\% (the rest were below 15\%).
+Similarly, for the 48 mutant alleles with high frequency in the evolved euploid populations, the highest frequencies in the aneuploid populations were 2.7\%, 7.7\%, and 11.1\% (the rest were below 1\%).
 }
   \label{fig:mutation_frequencies}
+  %TODO OD: suggest to put more stringent cut off here as a single read can be also a result of sequencing mistake.
+I think cut off of 10 reads is better and i think will not change the overall message
 \end{figure}
@@
 From the literature, the mutation rate per base pair is roughly $2-3\cdot10^{-10}$ (refs.~\citep{Zhu2014,Lynch2008}), but it may be higher under heat stress, as several stresses may cause hypermutation in yeast~\citep{Heidenreich2007}.
-If we assume a 10-fold increase in mutation rate, then the estimated beneficial mutation rate can be explained by a target size of 1,000 base pairs.
-\citet{Yona2012} found at least 10 genes on chromosome III with mutations that led to increased heat tolerance.
-Assuming that other chromosomes also have a similar number of heat-tolerance genes (i.e., chromosome III is one of the smaller chromosomes~\citep{Gilchrist2019}), and that each gene has a target size of ten base pairs, we easily get to a target size above 1,600.
+If we assume a 10-fold increase over the mutation rate reported in the literature, then the estimated beneficial mutation rate can be explained by a target size of 1,000 base pairs.
+\citet{Yona2012} found at least 10 genes on chromosome III that increased heat tolerance when over-expressed.
+Assuming that other chromosomes also have a similar number of heat-tolerance genes (i.e., chromosome III is one of the smaller chromosomes~\citep{Gilchrist2019}), and that each gene has a target size of ten base pairs, % TODO OD: on what this assumption based ?
+maybe look for fitness landscape analysis of other proteins and try to based estimation on that ?
+we easily get to a target size above 1,600.
 
@@
 
-Our model fits the data well: in simulations using the MAP parameter estimates, \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\Cref{fig:2n*-fixation}B).
+Our model fits the data well: in simulations using the MAP parameter estimates, \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\Cref{fig:2n*-fixation}).
 
@@
      }
+     % TODO OD: I missed it in the figure - make white star more visible 
   \label{fig:FA}
@@
 \paragraph{Genetic instability in aneuploid cells.}
+% TODO OD: I have some issues understanding this section (only got it right after we talked ....) I suggest to try explain better and also add the logic (why focusing on the effect on aneuploidy on mutation rate rather then mis-segregation rate ?).
+maybe also add sentence saying you also model the assumption that aneuplody affect mis-segregation rate (not shown ) 
 It has been suggested that aneuploidy increases genetic instability~\citep{Sheltzer2011b, Ippolito2021b}.
@@
 Additionally, such a replacement also occurred when they initiated evolutionary experiments with a population in which all cells carry a chromosome III trisomy.
-They hypothesized that the euploid mutant cells evolved by heat-resistance mutations in aneuploid cells followed by reversion of trisomy due to a chromosome loss event.
+They hypothesized % TODO OD: To be accurate, we never state implicitly that we think it occur in the same lineage.  
+maybe you can soften here a bit "One possible/ likely scenario ...." 
+ that the euploid mutant cells evolved by heat-resistance mutations in aneuploid cells followed by reversion of trisomy due to a chromosome loss event.
 
@@
 \paragraph{Conclusions.}
-Here, we tested the hypothesis that aneuploid cells are an evolutionary "stepping stone", or adaptive intermediate, between wild-type euploid cells and mutant euploid cells. 
+Here, we tested the hypothesis that aneuploidy cells are an evolutionary "stepping stone", or adaptive intermediate, between wild-type euploid cells and mutant euploid cells. 
 Our results suggest that, although it seems the population goes from euploid to aneuploid and back, this is not the case at the individual level.

### 2022-10-30 ms/ms.tex
YP abstract comments

@@
 Aneuploidy is common in eukaryotes, often leading to decreased growth and fitness.
-However, evidence from yeast and fungi, as well as human tumour cells, suggests that specific aneuploidies can be beneficial under stressful conditions and facilitate  adaptation.
+However, evidence from yeast and other fungi, as well as human tumour cells, suggests that specific aneuploidies can be beneficial under stressful conditions and facilitate  adaptation.
 In a prominent example, an evolutionary experiment with yeast, populations evolving under heat stress had become aneuploid (chromosome III), only to later revert back to euploid after genetic mutations have accumulated.
@@
 We then predict the genotype frequency dynamics during the experiment, demonstrating that the majority of the evolved euploid population likely did not descend from aneuploid cells, but rather directly from the euploid wild-type population.
-Our model further predicts that if the experiment was repeated with smaller populations, then a larger fraction of the evolved population would descend from aneuploid cells.
-
-Together, our results suggest that aneuploidy can be an inevitable evolutionary "detour" rather than a "stepping stone": it can delay, rather than facilitate, the adaptation of the population, and cells that become aneuploid may leave less descendants compared to cells that remain diploid.
+%Our model further predicts that if the experiment was repeated with smaller populations, then a larger fraction of the evolved population would descend from aneuploid cells.
+Our model shows how the beneficial mutation supply--the product of population size and beneficial mutation rate--determine the evolutionary dynamics, such that with a lower mutation supply a larger fraction of the evolved population may descend from aneuploid cells.
+However, with high mutation supply, beneficial mutations are generated before fixation of aneuploidy, and can outcompete aneuploidy due to its inherent fitness cost.
+
+Together, our results suggest that aneuploidy can be an evolutionary "detour" rather than a "stepping stone": it can delay, rather than facilitate, the adaptation of the population, and cells that become aneuploid may leave less descendants compared to cells that remain diploid.
 \end{abstract}

### 2022-11-07 ms/ms.tex
finished OD comments; added MR fig 1 draft

@@
 However, aneuploidy can be beneficial under stressful conditions due to the wide range of phenotypes it can produce, some of which are advantageous~\citep{Pavelka2010, Yang2021}.
-Indeed, in a survey of 1,011 yeast strain, aneuploidy has been detected in about 19\%\citep{Peter2018}.
+Indeed, in a survey of 1,011 yeast strains, aneuploidy has been detected in about 19\%\citep{Peter2018}.
 Thus, aneuploidy can lead to rapid adaptation in unicellular eukaryotes~\citep{Gerstein2015,Torres2010, Hong2014, Rancati2008}, as well as to rapid growth of somatic tumour cells~\citep{Schvartzman2010, Sheltzer2017}.
@@
   \centering
-    \includegraphics[width=\textwidth]{../figures/mutation_frequencies.pdf}      
+    \includegraphics[width=0.9\textwidth]{../figures/mutation_frequencies.pdf}      
   \caption{
@@
 Solid black line is a linear orthogonal distance regression line (slope=$-0.559$, intercept=$0.164$; a regression through alleles that reach at least 20\% in one of the populations has slope=$-0.645$ and intercept=$0.297$).
-Dashed vertical and horizontal lines show allele frequencies 20\%.
-Error bars show standard error of the mean accounting for the number of reads, which may be large when the total number of reads is small.
+Dashed vertical and horizontal lines show allele frequencies of 20\%.
+Error bars show SEM (standard error of the mean) assuming the number of reads in Binomially distributed; the SEM may be large when the total number of reads is small.
 For the 18 mutant alleles with high frequency in the aneuploid populations (>20\%), the highest frequencies in the euploid populations were 15.4\%, 16\%, 16.3\% and 19.6\% (the rest were below 15\%).
@@
   \label{fig:mutation_frequencies}
-  %TODO OD: suggest to put more stringent cut off here as a single read can be also a result of sequencing mistake.
-I think cut off of 10 reads is better and i think will not change the overall message
 \end{figure}
@@
   \centering
-    \includegraphics[height=1.8in]{../figures/Fig1-A.pdf}      
+    \includegraphics[width=\textwidth]{../figures/Fig1-A.png}      
   \caption{
@@
 The estimated beneficial mutation rate is $\mu=2.965\cdot10^{-6}\ [2.718\cdot10^{-7}-3.589\cdot10^{-6}]$.
-From the literature, the mutation rate per base pair is roughly $2-3\cdot10^{-10}$ (refs.~\citep{Zhu2014,Lynch2008}), but it may be higher under heat stress, as several stresses may cause hypermutation in yeast~\citep{Heidenreich2007}.
-If we assume a 10-fold increase over the mutation rate reported in the literature, then the estimated beneficial mutation rate can be explained by a target size of 1,000 base pairs.
-\citet{Yona2012} found at least 10 genes on chromosome III that increased heat tolerance when over-expressed.
-Assuming that other chromosomes also have a similar number of heat-tolerance genes (i.e., chromosome III is one of the smaller chromosomes~\citep{Gilchrist2019}), and that each gene has a target size of ten base pairs, % TODO OD: on what this assumption based ?
-maybe look for fitness landscape analysis of other proteins and try to based estimation on that ?
-we easily get to a target size above 1,600.
+From the literature, the mutation rate per base pair is roughly $2-3\cdot10^{-10}$ (refs.~\citep{Zhu2014,Lynch2008}), but it may be higher under heat stress, as several stresses~\citep{Heidenreich2007}, including heat~\citep{Huang2018}, may cause hypermutation in yeast.
+If we assume a 10-fold increase over the mutation rate reported in the literature, then the estimated beneficial mutation rate can be explained by a genomic target size of 1,000 base pairs.
+Supporting this, \citet{Flynn2020} used a deep mutational scan of a single protein, Hsp90, to find 465 amino-acid variants that increased growth rate in \SI{37}{\celsius}.
+Furthermore, \citet{Yona2012} found at least 10 genes on chromosome III that increased heat tolerance when over-expressed.
+Assuming that other chromosomes also have a similar number of heat-tolerance genes (i.e., chromosome III is one of the smaller chromosomes~\citep{Gilchrist2019}), we get a total of 160 heat-tolerance genes in the genome.
+Indeed, mutations were found in 97 genes in an evolutionary experiment with yeast under heat stress~\citep{Huang2018}.
+Thus, to get a genomic target size of 1,000, it is enough that the average gene target size is 6.25 base pairs.
+For example, \citet{Kohn2014} found a target size of 11 in a proton exporter gene (\emph{PMA1}) that contributes to high-salt adaptation.
 
@@
 	\textbf{(C)} $F_A$ values (color coded) from panel B, with their corresponding mutation rate $\mu$ on x-axis and aneuploidy rate $\delta$ on the y-axis.
-	White star shows the MAP estimate. See also \Cref{fig:posterior_FA_supp}.
+	Yellow star shows the MAP estimate. See also \Cref{fig:posterior_FA_supp}.
 	\textbf{(D)} $F_A$ as a function of the population size ($N$, bottom x-axis) and the beneficial mutation rate ($\mu$, top x-axis) in posterior predictions with MAP parameters. 
@@
      }
-     % TODO OD: I missed it in the figure - make white star more visible 
   \label{fig:FA}
@@
 \paragraph{Genetic instability in aneuploid cells.}
-% TODO OD: I have some issues understanding this section (only got it right after we talked ....) I suggest to try explain better and also add the logic (why focusing on the effect on aneuploidy on mutation rate rather then mis-segregation rate ?).
-maybe also add sentence saying you also model the assumption that aneuplody affect mis-segregation rate (not shown ) 
 It has been suggested that aneuploidy increases genetic instability~\citep{Sheltzer2011b, Ippolito2021b}.
@@
 We found that the posterior distribution was similar for $\tau=$1, 33/32, 2, and 5 (\Cref{fig:tau}).
+For each $\tau$, we computed the WAIC, a criterion for model selection (lower is better, see Methods), and found WAIC is lowest for $\tau=33/32$ and $\tau=1$ (\Cref{table:WAIC}).
+
 %% generated with MAP-and-hdi.ipynb
-With $\tau=100$, the estimated mutation rate was about 7-8-fold lower compared to $\tau=1$ ($\mu=4.094\cdot10^{-7}\ [6.252\cdot10^{-8}-6.046\cdot10^{-7}]$) and the aneuploidy rate was about 2-3-fold lower ($\delta=0.744\cdot10^{-3}\ [0.506\cdot10^{-3}-1.827\cdot10^{-3}]$).
-With $\tau=10$, the estimated mutation rate was only slightly lower compared to $\tau=1$ ($\mu=1.67\cdot10^{-6}\ [2.836\cdot10^{-8}-2.245\cdot10^{-6}]$). 
-WAIC (lower is better, see Methods) is lowest for $\tau=33/32$ and $\tau=1$ (\Cref{table:WAIC}).
-Therefore, our results do not support an increase in mutation rate in aneuploid cells.
+Assuming a strong increase of the mutation rate in aneuploid cells, i.e.
+$\tau=100$, the inferred a mutation rate was $\mu=4.094\cdot10^{-7}\ [6.252\cdot10^{-8}-6.046\cdot10^{-7}]$), and the inferred aneuploidy rate that was  $\delta=0.744\cdot10^{-3}\ [0.506\cdot10^{-3}-1.827\cdot10^{-3}]$.
+Compared to inference made assuming no effect of aneuploidy on the mutation rate, these rates were about 7-8-fold lower about 2-3-fold lower.
+Assuming $\tau=10$, the inferred a mutation rate was only slightly lower compared to $\tau=1$ ($\mu=1.67\cdot10^{-6}\ [2.836\cdot10^{-8}-2.245\cdot10^{-6}]$).
+
+Therefore, we do not find any evidence of an increase in mutation rate in aneuploid cells.
 This may be because, unless the increase is strong ($\tau \ge 10$), it does not seem to affect our inference; or because chromosome III is one of the smallest chromosomes \citep{Gilchrist2019}.
@@
 We observe $\tau=100$ could be distinguished if accurate data was available for the waiting time until the frequency of \euwt\ to decrease below 95\%~(\Cref{fig:tau-plots}A) or for waiting time for the frequency of \anwt\ to either reach or go below 95\%~(\Cref{fig:tau-plots}B).
+We also did not find evidence for an increase in the aneuploidy rate in aneuploid cells (data not shown).
 
@@
 Additionally, such a replacement also occurred when they initiated evolutionary experiments with a population in which all cells carry a chromosome III trisomy.
-They hypothesized % TODO OD: To be accurate, we never state implicitly that we think it occur in the same lineage.  
-maybe you can soften here a bit "One possible/ likely scenario ...." 
- that the euploid mutant cells evolved by heat-resistance mutations in aneuploid cells followed by reversion of trisomy due to a chromosome loss event.
+They hypothesized that aneuploidy is a ``useful yet short-lived intermediate that facilitates further adaptation'', suggesting that the euploid mutant cells evolved by heat-resistance mutations in aneuploid cells followed by reversion of trisomy due to a chromosome loss event.
 
@@
 We found that $F_A$ is expected to decrease as the population size or mutation rate increase (\Cref{fig:FA}D), ranging from >90\% when the population size is 10,000 or the mutation rate is $6\cdot10^{-9}$, to about 10\% when the population size is above 1,000,000 (less than the experimental population size, which was 6,425,000) or the mutation rate is above $2\cdot10^{-6}$ (less than the inferred mutation rate, which is $2.965\cdot10^{-6}$).
-Thus, our model provides a testable prediction: if the experiment was repeated under a lower population size (via stronger daily dilutions or in a smaller volume) or a lower mutation rate (via a non-mutagenic stress or stress with a smaller target size), then the fraction of the population descending from aneuploid cells would be much higher.
+Thus, our model provides a testable prediction: if the experiment was repeated under a lower population size (via stronger daily dilutions or in a smaller volume) or a lower mutation rate (via a non-mutagenic stress or stress with a smaller target size such as drug resistance), then the fraction of the population descending from aneuploid cells would be much higher.
 
@@
 That is, we assume chromosomes are gained and lost at the same rate, and we neglect events that form a less-fit genotype (i.e., \anwt\; to \euwt\; and \eumt\; to \anmt).
+A model that assumed increased aneuploidy rates in aneuploid cells did not perform well and was abandoned.
 

### 2022-11-08 ms/ms.tex
revise by YP comments

@@
 \usepackage[super,comma]{natbib}
+%\usepackage[numbers,super,compress]{natbib}
 %\usepackage[round,colon,authoryear]{natbib}
@@
 %Our model further predicts that if the experiment was repeated with smaller populations, then a larger fraction of the evolved population would descend from aneuploid cells.
-Our model shows how the beneficial mutation supply--the product of population size and beneficial mutation rate--determine the evolutionary dynamics, such that with a lower mutation supply a larger fraction of the evolved population may descend from aneuploid cells.
-However, with high mutation supply, beneficial mutations are generated before fixation of aneuploidy, and can outcompete aneuploidy due to its inherent fitness cost.
+Our model shows how the beneficial mutation supply--the product of population size and beneficial mutation rate--determine the evolutionary dynamics: with a low mutation supply, a large fraction of the evolved population may descend from aneuploid cells; but with a high mutation supply, beneficial mutations are generated before fixation of aneuploidy, and can outcompete aneuploidy due to the latter's inherent fitness cost.
 
@@
 %\paragraph{Aneuploidy is generally deleterious.}
-Experiments with human and mouse embryos found that most aneuploidies are lethal.
-It is also associated with developmental defects and lethality in other multicellular organisms~\citep{Sheltzer2011}. % trisomy 21 and Klinefelter and Turner's syndromes
+Experiments with human and mouse embryos found that most grem-line aneuploidies are lethal.
+Aneuploidies are also associated with developmental defects and lethality in other multicellular organisms~\citep{Sheltzer2011}. % trisomy 21 and Klinefelter and Turner's syndromes
 For example, aneuploid mouse embryonic cells grow slower than euploid cells~\citep{Williams2008}.
-Similarly, in unicellular eukaryotes growing in benign conditions, aneuploidy usually leads to slower growth and decreased overall fitness~\citep{Niwa2006, Torres2007, Pavelka2010, Sheltzer2011, Kasuga2016, Yang2021}, in part due to proteotoxic stress caused by increased expression in aneuploid cells~\citep{Pavelka2010, Santaguida2015, Zhu2018} and hypo-osmotic-like stress~\citep{Tsai2019}.
+Similarly, in unicellular eukaryotes growing in benign conditions, aneuploidy usually leads to slower growth and decreased overall fitness, in part due to proteotoxic stress due to increased expression, gene dosage imbalance, and hypo-osmotic-like stress~\citep{Niwa2006, Torres2007, Pavelka2010, Sheltzer2011, Santaguida2015, Kasuga2016, Zhu2018, Tsai2019, Yang2021}.
 
@@
 More importantly, a high mutant allele frequency in the aneuploid population was associated with a low frequency in the evolved euploid population, and vice-versa (Spearman's correlation coefficient $\rho=-0.64$ and $-0.66$ in the two experimental repetitions; \Cref{fig:mutation_frequencies}), suggesting that mutant alleles frequent in the aneuploid populations decreased in frequency when aneuploidy was lost.
-These results suggest evolved euploid cells are unlikely to descend from aneuploid~cells.
+These results suggest evolved euploid cells are unlikely to have descended from aneuploid cells.
 
@@
 aneuploid wild-type, \anwt, has an extra chromosome due to a chromosome duplication event; and
-aneuploid mutant, \anmt, has and extra chromosome and a beneficial mutation.
-Fitness values of the different genotypes are denoted by $w_{\euwt}$, $w_{\eumt}$, $w_{\anwt}$, and $w_{\anmt}$, and the rate of mutation and aneuploidy are denoted by $\mu$ and $\delta$.
+aneuploid mutant, \anmt, has and extra chromosome (like \anwt) and a beneficial mutation (like \eumt).
+Fitness values of the different genotypes are denoted by $w_{\euwt}$, $w_{\eumt}$, $w_{\anwt}$, and $w_{\anmt}$, and the rate of mutation and aneuploidy are denoted by $\mu$ and $\delta$, respectively.
 See \Cref{fig:models} for an illustration of the model.
@@
 From the literature, the mutation rate per base pair is roughly $2-3\cdot10^{-10}$ (refs.~\citep{Zhu2014,Lynch2008}), but it may be higher under heat stress, as several stresses~\citep{Heidenreich2007}, including heat~\citep{Huang2018}, may cause hypermutation in yeast.
-If we assume a 10-fold increase over the mutation rate reported in the literature, then the estimated beneficial mutation rate can be explained by a genomic target size of 1,000 base pairs.
+If we assume a 10-fold increase over the mutation rate reported in the literature, then the estimated beneficial mutation rate can be explained by a genomic target size of 1,000 base pairs (i.e., 1,000 base pairs across the genome in which a mutation would provide a fitness benefit).
 Supporting this, \citet{Flynn2020} used a deep mutational scan of a single protein, Hsp90, to find 465 amino-acid variants that increased growth rate in \SI{37}{\celsius}.
 Furthermore, \citet{Yona2012} found at least 10 genes on chromosome III that increased heat tolerance when over-expressed.
-Assuming that other chromosomes also have a similar number of heat-tolerance genes (i.e., chromosome III is one of the smaller chromosomes~\citep{Gilchrist2019}), we get a total of 160 heat-tolerance genes in the genome.
+Assuming that other chromosomes also have a similar number of heat-tolerance genes (and even more, as chromosome III is one of the smallest chromosomes~\citep{Gilchrist2019}), we get a total of 160 heat-tolerance genes in the genome.
 Indeed, mutations were found in 97 genes in an evolutionary experiment with yeast under heat stress~\citep{Huang2018}.
@@
 The estimated aneuploidy rate, $\delta=1.72\cdot10^{-3}\ [1.47\cdot10^{-3}-2.786\cdot10^{-3}]$ is higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
-However, this difference may be explained by an increased aneuploidy rate during heat stress~\citep{Chen2012a}.
+However, this difference may be partly explained by an increased aneuploidy rate during heat stress: heat shock can increase the rate of chromosome fragment loss by 2-3 orders of magnitude~\citep{Chen2012a}.
 
@@
 
-Our model fits the data well: in simulations using the MAP parameter estimates, \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\Cref{fig:2n*-fixation}).
+To assess the fit of our model to the data, we use posterior predictive checks, in which we simulate the frequency dynamics using MAP parameter estimates and compare them to the data.
+Our model fits the data well: \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\Cref{fig:2n*-fixation}).
 
@@
 %% generated with MAP-and-hdi.ipynb
-The estimated mutation rate without aneuploidy is $\mu=7.98\cdot10^{-9}\ [7.906\cdot10^{-9}-8.138\cdot10^{-9}]$, much lower compared to a model with aneuploidy and suggesting a target size of just 40.
+The estimated mutation rate without aneuploidy is $\mu=7.98\cdot10^{-9}\ [7.906\cdot10^{-9}-8.138\cdot10^{-9}]$, much lower compared to a model with aneuploidy.
 The fitness of the mutant is also much lower at $w_{\eumt}=1.013\ [1.012-1.013]$.

### 2022-11-22 ms/ms.tex
update fig 2 to pdf; more YP comments

@@
 \affil[b]{School of Computer Science, Reichman University, Herzliya, Israel}
-\affil[c]{Department of Molecular Genetics, Weizmann Institute of Science, Rehovot 76100, Israel}
+\affil[c]{Department of Molecular Genetics, Weizmann Institute of Science, Rehovot, Israel}
 \affil[d]{Institute of Biochemistry, Food Science and Nutrition,
@@
 %\paragraph{The present study.}
-Here, we test the hypothesis that aneuploidy is a an \emph{evolutionary stepping stone} that facilitates adaptive evolution by genetic mutations.
+Here, we test the hypothesis that aneuploidy is a an \emph{evolutionary stepping stone} that facilitates adaptive evolution by genetic mutations~\citet{Yona2015}.
 First, we sequenced the genomes of evolved populations reported in \citep{Yona2012} and analyzed their mutant allele frequencies to assess if the evolved euploid cells are descended from aneuploid cells.
@@
 %%%%%%%% Fig illustration
-\begin{figure}[t]
-  \centering
-    \includegraphics[width=\textwidth]{../figures/Fig1-A.png}      
+\begin{figure}[h]
+  \centering
+    \includegraphics[width=0.7\textwidth]{../figures/model_illustration.pdf}      
   \caption{
@@
     Arrows denote transitions between genotypes, with transitions rates $\mu$ for the beneficial mutation rate and $\delta$ for the aneuploidy rate.
+    Elevation differences illustrate the expected, rather than the assumed, fitness differences between the genotypes.
    }
@@
 
-In a landmark study on the role of chromosome duplication in adaptive evolution, 
+In a study on the role of chromosome duplication in adaptive evolution, 
 \citet{Yona2012} found that a chromosome III trisomy was acquired by \yeast populations evolving under heat stress, only to be later replaced by euploid mutant cells that carry "refined" solutions to the stress.
@@
 \paragraph{Conclusions.}
-Here, we tested the hypothesis that aneuploidy cells are an evolutionary "stepping stone", or adaptive intermediate, between wild-type euploid cells and mutant euploid cells. 
+Here, we tested the hypothesis that aneuploidy cells are an evolutionary "stepping stone", or adaptive intermediate, between wild-type euploid cells and mutant euploid cells~\citep{Yona2015}. 
 Our results suggest that, although it seems the population goes from euploid to aneuploid and back, this is not the case at the individual level.
 We estimate that only about 10-15\% of the euploid cells descended from aneuploid cells, whereas the rest are direct descendants of the wild-type euploid cells.
-Thus, aneuploidy can delay, rather than accelerate, adaptation, and and cells that become aneuploid may leave less descendants than cells that remain euploid.
-This surprising result reinforces the importance of models when making interpretations on evolutionary processes, and emphasizes the unintuitive outcomes of clonal interference during adaptive evolution.
+Thus, aneuploidy can delay, rather than accelerate, adaptation, and cells that become aneuploid may leave less descendants than cells that remain euploid. 
+This surprising result reinforces the importance of mathematical models when interpreting evolutionary dynamics.
+Moreover, our study emphasizes the unintuitive outcomes of clonal interference  between mechanisms for generation of variation that differ in their rate of formation and distribution of fitness effects, including mutation, copy number variation, horizontal gene transfer, and epigenetic modifications.
 

### 2022-11-22 ms/ms.tex
update ack

@@
 the Israel Science Foundation~(ISF, YR 552/19),
-the US--Israel Binational Science Foundation~(BSF, YR 2021276)
-Minerva Stiftung center for Lab Evolution~(YR),
+the US--Israel Binational Science Foundation~(BSF, YR 2021276),
+Minerva Stiftung center for Lab Evolution~(YP, YR),
 Minerva Stiftung short-term research grant~(MP).

### 2022-11-22 ms/ms.tex
update ack

--- 2022-11-22 ms/ms.tex
@@
 the Israel Science Foundation~(ISF, YR 552/19),
-the US--Israel Binational Science Foundation~(BSF, YR 2021276)
-Minerva Stiftung center for Lab Evolution~(YR),
+the US--Israel Binational Science Foundation~(BSF, YR 2021276),
+Minerva Center for Live Emulation of Evolution in the Lab~(YP, YR),
 Minerva Stiftung short-term research grant~(MP).

### 2022-11-24 ms/ms.tex
typo

@@
 
-The estimated aneuploidy rate, $\delta=1.72\cdot10^{-3}\ [1.47\cdot10^{-3}-2.786\cdot10^{-3}]$ is higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
+The estimated aneuploidy rate, $\delta=1.72\cdot10^{-3}\ [1.47\cdot10^{-3}-2.786\cdot10^{-3}]$ is higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0\cdot10^{-5}-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
 However, this difference may be partly explained by an increased aneuploidy rate during heat stress: heat shock can increase the rate of chromosome fragment loss by 2-3 orders of magnitude~\citep{Chen2012a}.
@@
 Following growth, 3ml of each culture were centrifuge (14,000rpm) and cell pellets were used for DNA extraction.
-Genomic DNA was extracted using “MasterPure Yeast DNA Purification Kit” (Lucigen) according to the manufacture instructions.
-Following extraction, DNA concentrations were determined by Qubit assay (Thermo Fisher) and ~1µg DNA was used for library preparation using Illumina sample preparation kit (Illumina).
+Genomic DNA was extracted using "MasterPure Yeast DNA Purification Kit" (Lucigen) according to the manufacture instructions.
+Following extraction, DNA concentrations were determined by Qubit assay (Thermo Fisher) and $\sim1\mu g$ DNA was used for library preparation using Illumina sample preparation kit (Illumina).
 Samples were sequenced using a 100 bp pair end read output run using Illumina HiSeq2500.

### 2022-11-30 ms/ms.tex
submit NatGen

@@
 We then predict the genotype frequency dynamics during the experiment, demonstrating that the majority of the evolved euploid population likely did not descend from aneuploid cells, but rather directly from the euploid wild-type population.
-%Our model further predicts that if the experiment was repeated with smaller populations, then a larger fraction of the evolved population would descend from aneuploid cells.
 Our model shows how the beneficial mutation supply--the product of population size and beneficial mutation rate--determine the evolutionary dynamics: with a low mutation supply, a large fraction of the evolved population may descend from aneuploid cells; but with a high mutation supply, beneficial mutations are generated before fixation of aneuploidy, and can outcompete aneuploidy due to the latter's inherent fitness cost.
 
-Together, our results suggest that aneuploidy can be an evolutionary "detour" rather than a "stepping stone": it can delay, rather than facilitate, the adaptation of the population, and cells that become aneuploid may leave less descendants compared to cells that remain diploid.
+Together, our results suggest that despite its potential fitness benefits under stress, aneuploidy can be an evolutionary "detour" rather than a "stepping stone": it can delay, rather than facilitate, the adaptation of the population, and cells that become aneuploid may leave less descendants compared to cells that remain diploid.
 \end{abstract}

### 2022-12-14 ms/ms.tex
shorter abstract

@@
 Aneuploidy is common in eukaryotes, often leading to decreased growth and fitness.
-However, evidence from yeast and other fungi, as well as human tumour cells, suggests that specific aneuploidies can be beneficial under stressful conditions and facilitate  adaptation.
-In a prominent example, an evolutionary experiment with yeast, populations evolving under heat stress had become aneuploid (chromosome III), only to later revert back to euploid after genetic mutations have accumulated.
+However, evidence from yeast and other fungi, as well as human tumour cells, suggest that specific aneuploidies can be beneficial under stressful conditions and facilitate adaptation.
+In a prominent experimental evolutionary study with yeast, populations evolving under heat stress became aneuploid, only to later revert back to euploidy after genetic mutations have accumulated.
 It has therefore been suggested that aneuploidy serves as a "stepping stone" on the path to adaptation.
-
 Here, we test this hypothesis. First, we apply DNA sequencing to show that mutant alleles common in aneuploid cells are uncommon in the evolved euploid population. 
-Second, we develop an evolutionary model with both aneuploidy and  mutation, and fit it to the results of the experiment using a Bayesian inference framework.
-We then predict the genotype frequency dynamics during the experiment, demonstrating that the majority of the evolved euploid population likely did not descend from aneuploid cells, but rather directly from the euploid wild-type population.
-Our model shows how the beneficial mutation supply--the product of population size and beneficial mutation rate--determine the evolutionary dynamics: with a low mutation supply, a large fraction of the evolved population may descend from aneuploid cells; but with a high mutation supply, beneficial mutations are generated before fixation of aneuploidy, and can outcompete aneuploidy due to the latter's inherent fitness cost.
-
+Second, we develop an evolutionary model with both aneuploidy and mutation and fit it to the experimental results using Bayesian inference.
+We then predict the genotype frequency dynamics during the experiment, demonstrating that the majority of the evolved euploid population likely did not descend from aneuploid cells, but rather from the euploid wildtype population.
 Together, our results suggest that despite its potential fitness benefits under stress, aneuploidy can be an evolutionary "detour" rather than a "stepping stone": it can delay, rather than facilitate, the adaptation of the population, and cells that become aneuploid may leave less descendants compared to cells that remain diploid.
 \end{abstract}
-
-% TODO Significance statement 120 words
 

### 2023-01-25 ms/ms.tex
added todo

@@
+% TODO
+% cite https://www.nature.com/articles/ng.3967
 \documentclass[12pt]{extarticle}
@@
 \begin{abstract} 
-Aneuploidy is common in eukaryotes, often leading to decreased growth and fitness.
-However, evidence from yeast and other fungi, as well as human tumour cells, suggest that specific aneuploidies can be beneficial under stressful conditions and facilitate adaptation.
-In a prominent experimental evolutionary study with yeast, populations evolving under heat stress became aneuploid, only to later revert back to euploidy after genetic mutations have accumulated.
-It has therefore been suggested that aneuploidy serves as a "stepping stone" on the path to adaptation.
+Aneuploidy is common in eukaryotes, often leading to decreased fitness.
+However, evidence from fungi and human tumour cells suggests that specific aneuploidies can be beneficial under stressful conditions and facilitate adaptation.
+In a previous evolutionary experiment with yeast, populations evolving under heat stress became aneuploid, only to later revert to euploidy after beneficial mutations accumulated.
+It was therefore suggested that aneuploidy is a "stepping stone" on the path to adaptation.
 Here, we test this hypothesis. First, we apply DNA sequencing to show that mutant alleles common in aneuploid cells are uncommon in the evolved euploid population. 
-Second, we develop an evolutionary model with both aneuploidy and mutation and fit it to the experimental results using Bayesian inference.
-We then predict the genotype frequency dynamics during the experiment, demonstrating that the majority of the evolved euploid population likely did not descend from aneuploid cells, but rather from the euploid wildtype population.
+Second, we use Bayesian inference to fit an evolutionary model with both aneuploidy and mutation the experimental results.
+We then predict the genotype frequency dynamics during the experiment, demonstrating that most of the evolved euploid population likely did not descend from aneuploid cells, but rather from the euploid wildtype population.
+Our model shows how the beneficial mutation supply--the product of population size and beneficial mutation rate--determines the evolutionary dynamics: with low supply, much of the evolved population descends from aneuploid cells; but with high supply, beneficial mutations are generated fast enough to outcompete aneuploidy due to its inherent fitness cost.
 Together, our results suggest that despite its potential fitness benefits under stress, aneuploidy can be an evolutionary "detour" rather than a "stepping stone": it can delay, rather than facilitate, the adaptation of the population, and cells that become aneuploid may leave less descendants compared to cells that remain diploid.

### 2023-05-30 ms/ms.tex
rev2 comments

@@
 %\paragraph{Aneuploidy is generally deleterious.}
-Experiments with human and mouse embryos found that most grem-line aneuploidies are lethal.
+Experiments with human and mouse embryos found that most germ-line aneuploidies are lethal.
 Aneuploidies are also associated with developmental defects and lethality in other multicellular organisms~\citep{Sheltzer2011}. % trisomy 21 and Klinefelter and Turner's syndromes
@@
 Dashed vertical and horizontal lines show allele frequencies of 20\%.
-Error bars show SEM (standard error of the mean) assuming the number of reads in Binomially distributed; the SEM may be large when the total number of reads is small.
+Error bars show SEM (standard error of the mean) assuming the number of reads is Binomially distributed; the SEM may be large when the total number of reads is small.
 For the 18 mutant alleles with high frequency in the aneuploid populations (>20\%), the highest frequencies in the euploid populations were 15.4\%, 16\%, 16.3\% and 19.6\% (the rest were below 15\%).
@@
 
-We fitted this model to the experimental results\citep{Yona2012} -- time for fixation (>95\%) and for loss (<5\%) of aneuploidy -- using approximate Bayesian computation with sequential Monte Carol (ABC-SMC)~\citep{Sisson2009}, thereby inferring the model parameters: rates aneuploidy and mutation and the fitness of all genotypes.
+We fitted this model to the experimental results\citep{Yona2012} -- time for fixation (>95\%) and for loss (<5\%) of aneuploidy -- using approximate Bayesian computation with sequential Monte Carlo (ABC-SMC)~\citep{Sisson2009}, thereby inferring the model parameters: rates aneuploidy and mutation and the fitness of all genotypes.
 We then sampled posterior predictions for the genotype frequency dynamics using the estimated parameter values and compared different versions of the model to test additional hypotheses about the evolutionary process.
@@
 % without back transitions:
-The estimated beneficial mutation rate is $\mu=2.965\cdot10^{-6}\ [2.718\cdot10^{-7}-3.589\cdot10^{-6}]$.
+The estimated beneficial mutation rate is $\mu=2.965\cdot10^{-6}\ [2.718\cdot10^{-7}-3.589\cdot10^{-6}]$ per genome per generation.
 From the literature, the mutation rate per base pair is roughly $2-3\cdot10^{-10}$ (refs.~\citep{Zhu2014,Lynch2008}), but it may be higher under heat stress, as several stresses~\citep{Heidenreich2007}, including heat~\citep{Huang2018}, may cause hypermutation in yeast.
-If we assume a 10-fold increase over the mutation rate reported in the literature, then the estimated beneficial mutation rate can be explained by a genomic target size of 1,000 base pairs (i.e., 1,000 base pairs across the genome in which a mutation would provide a fitness benefit).
+If we assume a 10-fold increase over the mutation rate reported in the literature, then the estimated beneficial mutation rate can be explained by a genomic target size of 1,000 base pairs (that is, 1,000 base pairs across the genome in which a mutation would provide a fitness benefit).
 Supporting this, \citet{Flynn2020} used a deep mutational scan of a single protein, Hsp90, to find 465 amino-acid variants that increased growth rate in \SI{37}{\celsius}.
@@
 Indeed, mutations were found in 97 genes in an evolutionary experiment with yeast under heat stress~\citep{Huang2018}.
-Thus, to get a genomic target size of 1,000, it is enough that the average gene target size is 6.25 base pairs.
+Thus, to get a genomic target size of 1,000, it is enough that the average gene target size (number of base pairs in a gene in which a mutation is beneficial) is 6.25 base pairs.
 For example, \citet{Kohn2014} found a target size of 11 in a proton exporter gene (\emph{PMA1}) that contributes to high-salt adaptation.
@@
 To assess the fit of our model to the data, we use posterior predictive checks, in which we simulate the frequency dynamics using MAP parameter estimates and compare them to the data.
-Our model fits the data well: \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\Cref{fig:2n*-fixation}).
+Our model fits the data well: \eumt\ fixed in 63\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\Cref{fig:2n*-fixation}).
 
@@
 \paragraph{Genetic instability in aneuploid cells.}
-It has been suggested that aneuploidy increases genetic instability~\citep{Sheltzer2011b, Ippolito2021b}.
+It has been suggested that aneuploidy increases genetic instability: \citet{Sheltzer2011b} have demonstrated a fold increase of between 2.2 and 7.1 in mutation rate.
 Therefore, we inferred model parameters under the assumption that the mutation rate increases in aneuploid cells by a factor $\tau$ = 1, 33/32 (due to an additional chromosome), 2, 5, 10, or 100 (due to genetic instability).
 We found that the posterior distribution was similar for $\tau=$1, 33/32, 2, and 5 (\Cref{fig:tau}).
-For each $\tau$, we computed the WAIC, a criterion for model selection (lower is better, see Methods), and found WAIC is lowest for $\tau=33/32$ and $\tau=1$ (\Cref{table:WAIC}).
+Furthermore, we computed the WAIC, a criterion for model selection (Methods).
+The WAIC values were similar for all $\tau$ values (\Cref{table:WAIC}).
 
@@
 $\tau=100$, the inferred a mutation rate was $\mu=4.094\cdot10^{-7}\ [6.252\cdot10^{-8}-6.046\cdot10^{-7}]$), and the inferred aneuploidy rate that was  $\delta=0.744\cdot10^{-3}\ [0.506\cdot10^{-3}-1.827\cdot10^{-3}]$.
-Compared to inference made assuming no effect of aneuploidy on the mutation rate, these rates were about 7-8-fold lower about 2-3-fold lower.
+Compared to inference made assuming no effect of aneuploidy on the mutation rate, these rates were about 7-8-fold and 2-3-fold lower for $\mu$ and $\delta$, respectively.
 Assuming $\tau=10$, the inferred a mutation rate was only slightly lower compared to $\tau=1$ ($\mu=1.67\cdot10^{-6}\ [2.836\cdot10^{-8}-2.245\cdot10^{-6}]$).
@@
 We observe $\tau=100$ could be distinguished if accurate data was available for the waiting time until the frequency of \euwt\ to decrease below 95\%~(\Cref{fig:tau-plots}A) or for waiting time for the frequency of \anwt\ to either reach or go below 95\%~(\Cref{fig:tau-plots}B).
-We also did not find evidence for an increase in the aneuploidy rate in aneuploid cells (data not shown).
+Similarly, we did not find evidence for an increase in the aneuploidy rate in aneuploid cells~\citep{Sheltzer2011b}, probably due to lack of statistical power.
 
@@
 That is, we assume chromosomes are gained and lost at the same rate, and we neglect events that form a less-fit genotype (i.e., \anwt\; to \euwt\; and \eumt\; to \anmt).
-A model that assumed increased aneuploidy rates in aneuploid cells did not perform well and was abandoned.
+A model that assumed increased aneuploidy rates in aneuploid cells (as in \citet{Sheltzer2011b}) did not perform well, probably due to lack of statistical power, and was abandoned.
 
@@
 
-Where posterior plots are very similar and the number of parameters is the same, we use WAIC, or the widely applicable information criterion ~\citep{gelman2013bayesian}, defined as
+Where posterior plots are very similar and the number of parameters is the same, we use WAIC, or the widely applicable information criterion~\citep{gelman2013bayesian}, defined as
 \begin{equation} \label{eq:WAIC}
@@
 We validated that upon resampling WAIC values do not significantly change and that differences in WAIC between models are preserved.
-WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy~\citep{Kass1995}.
+WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy.
 
@@
 \caption{
-\textbf{WAIC values for different $\tau$ values.}}
+\textbf{WAIC values for different $\tau$ values.}
+Differences of less than 6 are considered of weak significance~\citep{Kass1995}.
+}
 \pgfplotstabletypeset[

### 2023-06-05 ms/ms.tex
rev1 comments

@@
 Evidence suggests aneuploidy is very common in eukaryotes, e.g. animals~\citep{Santaguida2015review, Naylor2016, Bakhoum2017}, and fungi~\citep{Pavelka2010, Zhu2016, Robbins2017, Todd2017}.
-Aneuploidy has been implicated in cancer formation, progression, and drug resistance~\citep{Boveri2008, Schvartzman2010, Santaguida2015review, Ippolito2021b}.
+Aneuploidy has been implicated in cancer formation, progression, and drug resistance~\citep{Boveri2008, Schvartzman2010, Santaguida2015review, Rutledge2016, Ippolito2021b, Lukow2021}.
 It is also common in protozoan pathogens of the Leishmania genus, a major global health concern~\citep{Mannaert2012}, and contributes to the emergence of drug resistance~\citep{Selmecki2009} and virulence~\citep{Moller2018} in fungal pathogens, which are under-studied~\citep{Rodrigues2018}, despite infecting a billion people per year, causing significant morbidity in >150 million and death in >1.5 million people per year~\citep{Selmecki2009, Rodrigues2018}.
@@
 \citet{Yona2012} demonstrated experimentally the importance of aneuploidy in adaptive evolution. They evolved populations of \yeast under strong heat stress.
-The populations adapted to the heat stress within 450 generations, and this adaptation was determined to be due a duplication of chromosome III.
+The populations adapted to the heat stress within 450 generations, and this adaptation was determined to be due a duplication of Chromosome III.
 Later on, after more than 1,500 generations, the populations reverted back to an euploid state, while remaining adapted to the heat stress. 
@@
 Aneuploidy fixed in all four experimental repetitions in the first 450 generations.
-Two of the repetitions, marked $H2$ and $H4$, carried no large-scale duplications other than a chromosome III trisomy.
+Two of the repetitions, marked $H2$ and $H4$, carried no large-scale duplications other than a Chromosome III trisomy.
 These two repetitions continued to evolve under the same conditions, wherein aneuploidy was eliminated by generation 1,700 and 2,350 in $H4$ and $H2$, respectively.
@@
 Disregarding 45 and 40 alleles that were present in the ancestral populations at a frequency >10\%, the aneuploid and euploid populations carried a large number of mutant alleles: 82 and 95, respectively, in repetition $H2$, and 60 and 66 in repetition $H4$.
+37 mutant alleles were shared between the two evolved euploid populations, and their frequencies were correlated (Spearman's correlation: $r=0.4$, $p=0.15$).
+Of these 37, only two were identified in the Saccharomyces Genome Database~\citep{Cherry2012}: YDR034C-A on Chromosome IV and YGR265W on Chromosome VII.
+Neither has a known function, and the former was previously identified as slightly \emph{decreasing} heat tolerance~\citep{Jarolim2013}.
 
@@
 If we assume a 10-fold increase over the mutation rate reported in the literature, then the estimated beneficial mutation rate can be explained by a genomic target size of 1,000 base pairs (that is, 1,000 base pairs across the genome in which a mutation would provide a fitness benefit).
-Supporting this, \citet{Flynn2020} used a deep mutational scan of a single protein, Hsp90, to find 465 amino-acid variants that increased growth rate in \SI{37}{\celsius}.
-Furthermore, \citet{Yona2012} found at least 10 genes on chromosome III that increased heat tolerance when over-expressed.
-Assuming that other chromosomes also have a similar number of heat-tolerance genes (and even more, as chromosome III is one of the smallest chromosomes~\citep{Gilchrist2019}), we get a total of 160 heat-tolerance genes in the genome.
+Supporting this, \citet{Jarolim2013} found 279 genes that contributed to survival after a sudden shift from \SI{30}{\celsius} to \SI{50}{\celsius}, and \citet{Flynn2020} used a deep mutational scan of a single protein, Hsp90, to find 465 amino-acid variants that increased growth rate in \SI{37}{\celsius}.
+Furthermore, \citet{Yona2012} found at least 10 genes on Chromosome III that increased heat tolerance when over-expressed.
+Assuming that other chromosomes also have a similar number of heat-tolerance genes (and even more, as Chromosome III is one of the smallest chromosomes~\citep{Gilchrist2019}), we get a total of 160 heat-tolerance genes in the genome.
 Indeed, mutations were found in 97 genes in an evolutionary experiment with yeast under heat stress~\citep{Huang2018}.
@@
 
-The estimated aneuploidy rate, $\delta=1.72\cdot10^{-3}\ [1.47\cdot10^{-3}-2.786\cdot10^{-3}]$ is higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0\cdot10^{-5}-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
+The estimated aneuploidy rate, $\delta=1.72\cdot10^{-3}\ [1.47\cdot10^{-3}-2.786\cdot10^{-3}]$ is higher than in previous studies: for Chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0\cdot10^{-5}-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
 However, this difference may be partly explained by an increased aneuploidy rate during heat stress: heat shock can increase the rate of chromosome fragment loss by 2-3 orders of magnitude~\citep{Chen2012a}.
@@
 $w_{\eumt}=1.028\ [1.026-1.029]$, all relative to the fitness of \euwt, which is set to $w_{\euwt}=1$. 
-Thus, we can infer that the cost of chromosome III trisomy is $c = w_{\eumt}-w_{\anmt} = 0.003$ (or 0.3\%) and the benefit of trisomy is $w_{\anwt}-1-c=0.019$ (1.9\%), whereas the benefit of the beneficial mutation is $w_{\eumt}-1=0.028$ (2.8\%).
+Thus, we can infer that the cost of Chromosome III trisomy is $c = w_{\eumt}-w_{\anmt} = 0.003$ (or 0.3\%) and the benefit of trisomy is $w_{\anwt}-1-c=0.019$ (1.9\%), whereas the benefit of the beneficial mutation is $w_{\eumt}-1=0.028$ (2.8\%).
 
@@
 Therefore, we do not find any evidence of an increase in mutation rate in aneuploid cells.
-This may be because, unless the increase is strong ($\tau \ge 10$), it does not seem to affect our inference; or because chromosome III is one of the smallest chromosomes \citep{Gilchrist2019}.
+This may be because, unless the increase is strong ($\tau \ge 10$), it does not seem to affect our inference; or because Chromosome III is one of the smallest chromosomes \citep{Gilchrist2019}.
 We also checked the differences in genotype frequency dynamics for different $\tau$ values.
@@
 In a study on the role of chromosome duplication in adaptive evolution, 
-\citet{Yona2012} found that a chromosome III trisomy was acquired by \yeast populations evolving under heat stress, only to be later replaced by euploid mutant cells that carry "refined" solutions to the stress.
-Additionally, such a replacement also occurred when they initiated evolutionary experiments with a population in which all cells carry a chromosome III trisomy.
+\citet{Yona2012} found that a Chromosome III trisomy was acquired by \yeast populations evolving under heat stress, only to be later replaced by euploid mutant cells that carry "refined" solutions to the stress.
+Additionally, such a replacement also occurred when they initiated evolutionary experiments with a population in which all cells carry a Chromosome III trisomy.
 They hypothesized that aneuploidy is a ``useful yet short-lived intermediate that facilitates further adaptation'', suggesting that the euploid mutant cells evolved by heat-resistance mutations in aneuploid cells followed by reversion of trisomy due to a chromosome loss event.
@@
 If indeed the evolved euploid population is descended from the aneuploid population, then mutant alleles that were common in the aneuploid populations should also be common in the evolved euploid population.
-However, we found that this is not the case (\Cref{fig:mutation_frequencies}): mutant allele frequencies in the aneuploid and euploid populations are negatively correlated, such that common alleles in the former are rare in the later.
+However, this is likely not the case (\Cref{fig:mutation_frequencies}): mutant allele frequencies in the aneuploid and euploid populations are negatively correlated, such that common alleles in the former are rare in the later.
 Furthermore, we developed an evolutionary genetic model of adaptive evolution by aneuploidy and mutation (\Cref{fig:models}), fitted it to the experimental results of \citet{Yona2012}, and used it to predict the genotype frequency dynamics.
-The model predicted that only about 10-15\% of the evolved euploid population descended from aneuploid cells--that is, the majority of the euploid population are not descended from aneuploid cells, but rather are direct descendants of the ancestral wild-type population (\Cref{fig:FA}).
+The model predicted that only about 10-15\% of the evolved euploid population descended from aneuploid cells by acquiring a mutation and losing the extra chromosome--that is, the majority of the euploid population are not descended from aneuploid cells, but rather are direct descendants of the ancestral wild-type population (\Cref{fig:FA}).
 
@@
 We focus on beneficial genetic modifications, neglecting the effects of deleterious and neutral mutations or karyotypic changes.
-The model allows for a single aneuploid karyotype (e.g., chromosome III duplication) and a single mutation to accumulate in the genotype.
+The model allows for a single aneuploid karyotype (e.g., Chromosome III duplication) and a single mutation to accumulate in the genotype.
 Thus, the model follows four genotypes~(\Cref{fig:models}): euploid wild-type, \euwt, the initial genotype; 
@@
 
-For a model without aneuploidy (that is, when the aneuploidy rate is fixed at zero, $\delta=0$), we disregard the increased expression in chromosome III and the growth advantage measured in generation 450, and focus on the growth advantage measured in later generations, presumably due to a beneficial mutation.
+For a model without aneuploidy (that is, when the aneuploidy rate is fixed at zero, $\delta=0$), we disregard the increased expression in Chromosome III and the growth advantage measured in generation 450, and focus on the growth advantage measured in later generations, presumably due to a beneficial mutation.
 Therefore, the likelihood is approximated by
@@
 The estimated mutation rate, $\mu=1.036\cdot10^{-7}\ [8.01\cdot10^{-8}-1.339\cdot10^{-7}]$, corresponds to a mutation target size of $\sim 300-500$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$ (ref.~\citep{Zhu2014}) or $3.3\cdot10^{-10}$ (ref.~\citep{Lynch2008}).
-The estimated aneuploidy rate, $\delta=2.358\cdot10^{-4}\ [1.766\cdot10^{-4}-2.837\cdot10^{-4}]$ is 5-35-fold higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
+The estimated aneuploidy rate, $\delta=2.358\cdot10^{-4}\ [1.766\cdot10^{-4}-2.837\cdot10^{-4}]$ is 5-35-fold higher than in previous studies: for Chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
 The estimated fitness values are 

### 2023-07-18 ms/ms.tex
added citations and fixed bib

@@
 For example, aneuploid mouse embryonic cells grow slower than euploid cells~\citep{Williams2008}.
-Similarly, in unicellular eukaryotes growing in benign conditions, aneuploidy usually leads to slower growth and decreased overall fitness, in part due to proteotoxic stress due to increased expression, gene dosage imbalance, and hypo-osmotic-like stress~\citep{Niwa2006, Torres2007, Pavelka2010, Sheltzer2011, Santaguida2015, Kasuga2016, Zhu2018, Tsai2019, Yang2021}.
+Similarly, in unicellular eukaryotes growing in benign conditions, aneuploidy usually leads to slower growth and decreased overall fitness, in part due to proteotoxic stress due to increased expression, gene dosage imbalance, and hypo-osmotic-like stress~\citep{Niwa2006, Torres2007, Pavelka2010, Sheltzer2011, Santaguida2015, Kasuga2016, Zhu2018, Tsai2019, Yang2021, Robinson2023}.
 
@@
 Thus, aneuploidy can lead to rapid adaptation in unicellular eukaryotes~\citep{Gerstein2015,Torres2010, Hong2014, Rancati2008}, as well as to rapid growth of somatic tumour cells~\citep{Schvartzman2010, Sheltzer2017}.
-For example, aneuploidy in \yeast facilitates adaptation to a variety of stressful conditions like heat and pH~\citep{Yona2012}, copper~\citep{Covo2014, Gerstein2015}, salt~\citep{Dhar2011}, and nutrient limitation~\citep{Dunham2002, Gresham2008, Avecilla2022}, with similar results in \calbicans~\citep{Yang2021}.
-Importantly, aneuploidy can also lead to drug resistance in pathogenic fungi such as \calbicans~\citep{Selmecki2008, Selmecki2010, Gerstein2018} and \cneoformans~\citep{Sionov2010}, which cause candidiasis and meningoencephalitis, respectively.
+For example, aneuploidy in \yeast facilitates adaptation to a variety of stressful conditions like heat and pH~\citep{Yona2012}, copper~\citep{Covo2014, Gerstein2015}, salt~\citep{Dhar2011, Robinson2023}, and nutrient limitation~\citep{Dunham2002, Gresham2008, Avecilla2022}, with similar results in \calbicans~\citep{Yang2021}.
+Importantly, aneuploidy can also lead to drug resistance in pathogenic fungi such as \calbicans~\citep{Selmecki2008, Selmecki2010, Gerstein2020} and \cneoformans~\citep{Sionov2010}, which cause candidiasis and meningoencephalitis, respectively.
+Although we focus here on aneuploidy, a similar phenomena of adaptation via gene duplication or amplification has been observed in yeast \citep{Lauer2018}, bacteria \citep{Sonti1989}, and DNA viruses \citep{Elde2012}.
 

### 2023-11-23 ms/ms.tex
remove seq data

@@
 
-\author[a,b,*]{Ilia Kohanovski}
-\author[a,*]{Martin Pontz}
+\author[a,b,1]{Ilia Kohanovski}
+\author[a,1]{Martin Pontz}
 \author[c]{Orna Dahan}
@@
 \author[d]{Avihu H. Yona}
-\author[a,1]{Yoav Ram}
+\author[a,*]{Yoav Ram}
 
@@
 The Hebrew University of Jerusalem, Israel}
-\affil[*]{These authors contributed equally to this work}
-\affil[1]{Corresponding author: yoav@yoavram.com}
+\affil[1]{These authors contributed equally to this work}
+\affil[*]{Corresponding author: yoav@yoavram.com}
 
@@
 It was therefore suggested that aneuploidy is a "stepping stone" on the path to adaptation.
-Here, we test this hypothesis. First, we apply DNA sequencing to show that mutant alleles common in aneuploid cells are uncommon in the evolved euploid population. 
-Second, we use Bayesian inference to fit an evolutionary model with both aneuploidy and mutation the experimental results.
+Here, we test this hypothesis.
+We use Bayesian inference to fit an evolutionary model with both aneuploidy and mutation the experimental results.
 We then predict the genotype frequency dynamics during the experiment, demonstrating that most of the evolved euploid population likely did not descend from aneuploid cells, but rather from the euploid wildtype population.
 Our model shows how the beneficial mutation supply--the product of population size and beneficial mutation rate--determines the evolutionary dynamics: with low supply, much of the evolved population descends from aneuploid cells; but with high supply, beneficial mutations are generated fast enough to outcompete aneuploidy due to its inherent fitness cost.
-Together, our results suggest that despite its potential fitness benefits under stress, aneuploidy can be an evolutionary "detour" rather than a "stepping stone": it can delay, rather than facilitate, the adaptation of the population, and cells that become aneuploid may leave less descendants compared to cells that remain diploid.
+Our results suggest that despite its potential fitness benefits under stress, aneuploidy can be an evolutionary "detour" rather than a "stepping stone": it can delay, rather than facilitate, the adaptation of the population, and cells that become aneuploid may leave less descendants compared to cells that remain diploid.
 \end{abstract}
@@
 Here, we test the hypothesis that aneuploidy is a an \emph{evolutionary stepping stone} that facilitates adaptive evolution by genetic mutations~\citet{Yona2015}.
-First, we sequenced the genomes of evolved populations reported in \citep{Yona2012} and analyzed their mutant allele frequencies to assess if the evolved euploid cells are descended from aneuploid cells.
-Second, we develop an evolutionary genetic model and fit it to the experimental results of \citet{Yona2012} in order to predict the genotype frequency dynamics in the experimental populations, thereby estimating the frequency of evolved euploid cells that descended from aneuploid cells.
-Our results show that aneuploidy reached high frequencies in the experimental populations, but nevertheless, the majority of cells in the evolved euploid population likely did not descend from aneuploid cells, but rather directly from wild-type euploid cells.
-These results suggest that at the lineage level, aneuploidy may be an evolutionary detour, rather than a stepping stone, on the path to adaptation.
+We develop an evolutionary genetic model and fit it to the experimental results of \citet{Yona2012} to predict the genotype frequency dynamics in the experimental populations, thereby estimating the frequency of evolved euploid cells that descended from aneuploid cells.
+Our results show that although aneuploidy reached high frequencies in the experimental populations, the majority of cells in the evolved euploid population likely did not descend from aneuploid cells, but rather directly from wild-type euploid cells.
+These suggests that at the lineage level, aneuploidy may be an evolutionary detour, rather than a stepping stone, on the path to adaptation.
 
@@
 These two repetitions continued to evolve under the same conditions, wherein aneuploidy was eliminated by generation 1,700 and 2,350 in $H4$ and $H2$, respectively.
-
-\paragraph{Empirical frequencies of mutant alleles.} 
-For each of two evolved populations ($H2$ and $H4$) we sequenced the ancestral diploid population (generation 0), the aneuploid population (generation 450), and the evolved euploid population (generation 1,700 or 2,350) to estimate the mutant allele frequencies (\Cref{table:mutations_H2,table:mutations_H4}).
-Overall, between 100 and 173 mutant alleles were detected with at least a single read in the six populations that were sampled. 
-Disregarding 45 and 40 alleles that were present in the ancestral populations at a frequency >10\%, the aneuploid and euploid populations carried a large number of mutant alleles: 82 and 95, respectively, in repetition $H2$, and 60 and 66 in repetition $H4$.
-37 mutant alleles were shared between the two evolved euploid populations, and their frequencies were correlated (Spearman's correlation: $r=0.4$, $p=0.15$).
-Of these 37, only two were identified in the Saccharomyces Genome Database~\citep{Cherry2012}: YDR034C-A on Chromosome IV and YGR265W on Chromosome VII.
-Neither has a known function, and the former was previously identified as slightly \emph{decreasing} heat tolerance~\citep{Jarolim2013}.
-
-Surprisingly, out of all these mutant alleles, none was present at a frequency >20\% in both the aneuploid and the evolved euploid populations. 
-More importantly, a high mutant allele frequency in the aneuploid population was associated with a low frequency in the evolved euploid population, and vice-versa (Spearman's correlation coefficient $\rho=-0.64$ and $-0.66$ in the two experimental repetitions; \Cref{fig:mutation_frequencies}), suggesting that mutant alleles frequent in the aneuploid populations decreased in frequency when aneuploidy was lost.
-These results suggest evolved euploid cells are unlikely to have descended from aneuploid cells.
-
-%%%%%%%% Fig mutation frequencies
-\begin{figure}[h]
-  \centering
-    \includegraphics[width=0.9\textwidth]{../figures/mutation_frequencies.pdf}      
-  \caption{
-    \textbf{Frequencies of mutant alleles in the experimental populations are negatively correlated.} 
-    Frequencies of mutant alleles when trisomy was widespread in the population (y-axis) and after it was eliminated (x-axis) in two experimental repetitions (circles for $H2$ and triangles for $H4$) from \citet{Yona2012}.
-Mutant alleles with >20\% in the aneuploid population were <20\% in the euploid population, and vice versa (the upper-right quadrant is empty), suggesting that the majority of evolved euploid cells did not descend from the most common aneuploid genotypes.
-Alleles with frequency below and above 10\% in the ancestral populations are in black and gray, respectively. 
-Solid black line is a linear orthogonal distance regression line (slope=$-0.559$, intercept=$0.164$; a regression through alleles that reach at least 20\% in one of the populations has slope=$-0.645$ and intercept=$0.297$).
-Dashed vertical and horizontal lines show allele frequencies of 20\%.
-Error bars show SEM (standard error of the mean) assuming the number of reads is Binomially distributed; the SEM may be large when the total number of reads is small.
-For the 18 mutant alleles with high frequency in the aneuploid populations (>20\%), the highest frequencies in the euploid populations were 15.4\%, 16\%, 16.3\% and 19.6\% (the rest were below 15\%).
-Similarly, for the 48 mutant alleles with high frequency in the evolved euploid populations, the highest frequencies in the aneuploid populations were 2.7\%, 7.7\%, and 11.1\% (the rest were below 1\%).
-}
-  \label{fig:mutation_frequencies}
-\end{figure}
 
@@
 
-If indeed the evolved euploid population is descended from the aneuploid population, then mutant alleles that were common in the aneuploid populations should also be common in the evolved euploid population.
-However, this is likely not the case (\Cref{fig:mutation_frequencies}): mutant allele frequencies in the aneuploid and euploid populations are negatively correlated, such that common alleles in the former are rare in the later.
-Furthermore, we developed an evolutionary genetic model of adaptive evolution by aneuploidy and mutation (\Cref{fig:models}), fitted it to the experimental results of \citet{Yona2012}, and used it to predict the genotype frequency dynamics.
+We developed an evolutionary genetic model of adaptive evolution by aneuploidy and mutation (\Cref{fig:models}), fitted it to the experimental results of \citet{Yona2012}, and used it to predict the genotype frequency dynamics.
 The model predicted that only about 10-15\% of the evolved euploid population descended from aneuploid cells by acquiring a mutation and losing the extra chromosome--that is, the majority of the euploid population are not descended from aneuploid cells, but rather are direct descendants of the ancestral wild-type population (\Cref{fig:FA}).
@@
 %%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph{DNA sequencing.}
-Whole-genome sequencing of the ancestral diploid strain (\euwt) was performed
-on a single colony of the ancestor.
-Whole-genome sequencing of the four evolving populations ($H2$ after 450 and 2,350 generations, and $H4$ after 450 and 1,700 generations) was performed on a sample from these populations (rather than from single colonies) in order to maintain the population diversity.
-Cells were grown in 5ml of YPD medium, either at \SI{30}{\celsius} (ancestral diploid) or \SI{39}{\celsius} (evolved populations) in shaking conditions (200rpm) until reaching stationary phase.
-Following growth, 3ml of each culture were centrifuge (14,000rpm) and cell pellets were used for DNA extraction.
-Genomic DNA was extracted using "MasterPure Yeast DNA Purification Kit" (Lucigen) according to the manufacture instructions.
-Following extraction, DNA concentrations were determined by Qubit assay (Thermo Fisher) and $\sim1\mu g$ DNA was used for library preparation using Illumina sample preparation kit (Illumina).
-Samples were sequenced using a 100 bp pair end read output run using Illumina HiSeq2500.
-
 \paragraph{Evolutionary genetic model.}
@@
 \section*{Acknowledgements}
-We thank Lilach Hadany, Judith Berman, David Gresham, Shay Covo, Martin Kupiec, Uri Obolski, Daniel Weissman, and Tal Simon for discussions and comments.
+We thank Lilach Hadany, Judith Berman, David Gresham, Shay Covo, Martin Kupiec, Anna Selmecki, Petra Vande Zande, Uri Obolski, Daniel Weissman, and Tal Simon for discussions and comments.
 This work was supported in part by 
@@
 the US--Israel Binational Science Foundation~(BSF, YR 2021276),
-Minerva Center for Live Emulation of Evolution in the Lab~(YP, YR),
+Minerva Center for Live Emulation of Evolution in the Lab~(YR, YP),
 Minerva Stiftung short-term research grant~(MP).
@@
 
-% Table - mutations
-%% generated with mutation_analysis.ipynb
-\begin{table}[h]
-\centering
-\caption{
-\textbf{Mutant alleles in population $H2$.}}
-\caption*{
-	Mutant alleles identified in the ancestor (generation 0), aneuploid (generation 450), and evolved (generation 2,350) of population $H2$. See supplementary file.
-} 
-\label{table:mutations_H2}
-\end{table}
-
-% Table - mutations
-%% generated with mutation_analysis.ipynb
-\begin{table}[h]
-\centering
-\caption{
-\textbf{Mutant alleles in population $H4$.}}
-\caption*{
-	Mutant alleles identified in the ancestor (generation 0), aneuploid (generation 450), and evolved (generation 1,700) of population $H4$. See supplementary file.
-} 
-\label{table:mutations_H4}
-\end{table}
-
 %% generated with WAIC.ipynb 

### 2023-11-23 ms/ms.tex
submit to mbe

@@
 \section*{Acknowledgements}
-We thank Lilach Hadany, Judith Berman, David Gresham, Shay Covo, Martin Kupiec, Anna Selmecki, Petra Vande Zande, Uri Obolski, Daniel Weissman, and Tal Simon for discussions and comments.
+We thank Lilach Hadany, Judith Berman, David Gresham, Shay Covo, Martin Kupiec, Uri Obolski, Daniel Weissman, and Tal Simon for discussions and comments.
 This work was supported in part by 

### 2024-01-11 ms/ms.tex
add anna and petra; revise title

@@
 % \title{Chromosomal duplication can be an evolutionary detour on the path to adaptation}
-\title{Aneuploidy can be an evolutionary detour \\ on the path to adaptation}
+\title{Aneuploidy can be an evolutionary diversion\\ on the path to adaptation}
 
@@
 \author[a,1]{Martin Pontz}
-\author[c]{Orna Dahan}
-\author[c]{Yitzhak Pilpel}
-\author[d]{Avihu H. Yona}
+\author[c]{P\'{e}tra Vande Zande}
+\author[c]{Anna Selmecki}
+\author[d]{Orna Dahan}
+\author[d]{Yitzhak Pilpel}
+\author[e]{Avihu H. Yona}
 \author[a,*]{Yoav Ram}
@@
 \affil[b]{School of Computer Science, Reichman University, Herzliya, Israel}
-\affil[c]{Department of Molecular Genetics, Weizmann Institute of Science, Rehovot, Israel}
-\affil[d]{Institute of Biochemistry, Food Science and Nutrition,
-Robert H. Smith Faculty of Agriculture, Food and Environment,
-The Hebrew University of Jerusalem, Israel}
+\affil[c]{Department of Microbiology and Immunology, University of Minnesota Medical School, Minneapolis, MN}
+\affil[d]{Department of Molecular Genetics, Weizmann Institute of Science, Rehovot, Israel}
+\affil[e]{Institute of Biochemistry, Food Science and Nutrition, Robert H. Smith Faculty of Agriculture, Food and Environment, The Hebrew University of Jerusalem, Israel}
 \affil[1]{These authors contributed equally to this work}

### 2024-01-11 ms/ms.tex
fix figure ref

@@
 
-We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate an informative prior distribution for $w_{\anwt}$~(\Cref{fig:growth-curves}-D, assuming $w_{\euwt}=1$).
+We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate an informative prior distribution for $w_{\anwt}$~(\Cref{fig:growth-curves}D, assuming $w_{\euwt}=1$).
 In this prior distribution, we used the same prior for $w_{\anmt}$ and $w_{\eumt}$. 
@@
 \paragraph{Extended informative prior distribution.}
-In an extended informative prior distribution, we used additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{\eumt}/w_{\anwt}$~(\Cref{fig:growth-curves}L).
+In an extended informative prior distribution, we used additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{\eumt}/w_{\anwt}$~(\Cref{fig:growth-curves}H).
 The same distribution was used for $w_{\eumt}/w_{\anmt}$.

### 2024-01-21 ms/ms.tex
finish rev 1

@@
 \title{Aneuploidy can be an evolutionary diversion\\ on the path to adaptation}
+%\title{Aneuploidy can delay rather than accelerate adaptation}
+%\title{Aneuploidy can impede rather than spur adaptation}
 
@@
 Our model shows how the beneficial mutation supply--the product of population size and beneficial mutation rate--determines the evolutionary dynamics: with low supply, much of the evolved population descends from aneuploid cells; but with high supply, beneficial mutations are generated fast enough to outcompete aneuploidy due to its inherent fitness cost.
-Our results suggest that despite its potential fitness benefits under stress, aneuploidy can be an evolutionary "detour" rather than a "stepping stone": it can delay, rather than facilitate, the adaptation of the population, and cells that become aneuploid may leave less descendants compared to cells that remain diploid.
+Our results suggest that despite its potential fitness benefits under stress, aneuploidy can be an evolutionary "diversion" rather than a "stepping stone": it can delay, rather than facilitate, the adaptation of the population, and cells that become aneuploid may leave less descendants compared to cells that remain diploid.
 \end{abstract}
@@
 Our results show that although aneuploidy reached high frequencies in the experimental populations, the majority of cells in the evolved euploid population likely did not descend from aneuploid cells, but rather directly from wild-type euploid cells.
-These suggests that at the lineage level, aneuploidy may be an evolutionary detour, rather than a stepping stone, on the path to adaptation.
+These suggests that at the lineage level, aneuploidy may be an evolutionary diversion, rather than a stepping stone, on the path to adaptation.
 
@@
 % without back transitions:
-The estimated beneficial mutation rate is $\mu=2.965\cdot10^{-6}\ [2.718\cdot10^{-7}-3.589\cdot10^{-6}]$ per genome per generation.
+The estimated beneficial mutation rate is $\mu=2.965\cdot10^{-6}\ [2.718\cdot10^{-7}-3.589\cdot10^{-6}]$ per genome per generation (that is, roughly 3 out of $10^6$ cell divisions produce a mutant cell with a fitness advantage).
 From the literature, the mutation rate per base pair is roughly $2-3\cdot10^{-10}$ (refs.~\citep{Zhu2014,Lynch2008}), but it may be higher under heat stress, as several stresses~\citep{Heidenreich2007}, including heat~\citep{Huang2018}, may cause hypermutation in yeast.
-If we assume a 10-fold increase over the mutation rate reported in the literature, then the estimated beneficial mutation rate can be explained by a genomic target size of 1,000 base pairs (that is, 1,000 base pairs across the genome in which a mutation would provide a fitness benefit).
-Supporting this, \citet{Jarolim2013} found 279 genes that contributed to survival after a sudden shift from \SI{30}{\celsius} to \SI{50}{\celsius}, and \citet{Flynn2020} used a deep mutational scan of a single protein, Hsp90, to find 465 amino-acid variants that increased growth rate in \SI{37}{\celsius}.
+If we assume a 10-fold increase over the mutation rate reported in the literature, then the estimated beneficial mutation rate can be explained by a genomic target size of 1,000 base pairs (that is, 1,000 base pairs across the genome in which a mutation would provide a fitness advantage): $3\cdot10^{-10} \times 10 \times 1,000=3\cdot10^{-6}$.
+Supporting this, \citet{Jarolim2013} found 279 genes that contributed to survival after a sudden shift from \SI{30}{\celsius} to \SI{50}{\celsius}, and \citet{Flynn2020} used a deep mutational scan of a single protein, Hsp90, to find 465 amino-acid variants (out of 14,160) that significantly increased growth rate in \SI{37}{\celsius}.
 Furthermore, \citet{Yona2012} found at least 10 genes on Chromosome III that increased heat tolerance when over-expressed.
@@
 $w_{\eumt}=1.028\ [1.026-1.029]$, all relative to the fitness of \euwt, which is set to $w_{\euwt}=1$. 
-Thus, we can infer that the cost of Chromosome III trisomy is $c = w_{\eumt}-w_{\anmt} = 0.003$ (or 0.3\%) and the benefit of trisomy is $w_{\anwt}-1-c=0.019$ (1.9\%), whereas the benefit of the beneficial mutation is $w_{\eumt}-1=0.028$ (2.8\%).
-
 If we allow for transitions (mutation, chromosome loss and gain) to less-fit genotypes (e.g., \eumt\ to \anmt), then we infer similar but slightly different values, see Supplementary Material.
@@
 We observe $\tau=100$ could be distinguished if accurate data was available for the waiting time until the frequency of \euwt\ to decrease below 95\%~(\Cref{fig:tau-plots}A) or for waiting time for the frequency of \anwt\ to either reach or go below 95\%~(\Cref{fig:tau-plots}B).
-Similarly, we did not find evidence for an increase in the aneuploidy rate in aneuploid cells~\citep{Sheltzer2011b}, probably due to lack of statistical power.
+Similarly, we did not find evidence for an increase in the rate of chromosome loss in aneuploid cells~\citep{Sheltzer2011b}, probably due to lack of statistical power.
+
+Nevertheless, increasing the rate of chromosome loss (transitions from $2n+1^*$ to $2n^*$) without increasing the rate of chromosome gain (transitions from $2n$ to $2n+1$) increases $F_A$ (\Cref{fig:FA_rates}B), but not to the same extent as increasing the rate of chromosome gain.
+In contrast, increasing the mutation rate in aneuploid cells can have a marked effect on the dynamics: when using the MAP parameter estimates, $F_A$ increases to  0.52 when the mutation rate in aneuploid cells increases 10-fold (\Cref{fig:FA_rates}C).
 
@@
 That is, we assume chromosomes are gained and lost at the same rate, and we neglect events that form a less-fit genotype (i.e., \anwt\; to \euwt\; and \eumt\; to \anmt).
-A model that assumed increased aneuploidy rates in aneuploid cells (as in \citet{Sheltzer2011b}) did not perform well, probably due to lack of statistical power, and was abandoned.
+A model that assumed increased rate of chromosome loss in aneuploid cells (as in \citet{Sheltzer2011b}) did not perform well, probably due to lack of statistical power, and was abandoned.
 
@@
       &f^m_{2n+1^*} &=&\; \mu f^s_{2n+1} + (1-\delta) f^s_{2n+1^*}  \;,\\
-      &f^m_{2n^*} &=&\; \mu f^s_{2n} + \delta f^s_{2n+1} + f^s_{2n^*}  \;.
+      &f^m_{2n^*} &=&\; \mu f^s_{2n} + \delta f^s_{2n+1^*} + f^s_{2n^*}  \;.
     \end{aligned}
@@
 Hereafter, fixation or elimination of a genotype \emph{by generation $t$} means that more than 95\% or less than 5\% of the population carry the genotype at generation $t$, and possibly earlier.
-From re-analysis of data not published in the original paper, aneuploidy did not fix before at least 200 generations elapsed.
+In the original analysis of \citet{Yona2012}, samples were routinely extracted from the evolving populations and tested for heat-shock tolerance.
+The first generation in which such indication was found was generation 200.
+Therefore, we determine that aneuploidy did not fix before generation 200.
 The experiment continued with two populations, in which aneuploidy was eliminated by generation 1,700 and 2,350 while still under the same conditions of elevated heat (\SI{39}{\celsius}).
@@
 \paragraph{Prior distributions.} \label{par:prior}
-We used informative prior distributions for $w_{\anwt}=1-c+b$, $w_{\anmt}=(1+s)(1-c)+b$ and $w_{\eumt}=1+s$, which we estimated from growth curves data from mono-culture growth experiments previously reported by \citet[Figs.
+We used informative prior distributions for $w_{\anwt}$, $w_{\anmt}$ and $w_{\eumt}$ (we set $w_{2n}=1$), which we estimated from growth curves data from mono-culture growth experiments previously reported by \citet[Figs.
 3C, 4A, and S2]{Yona2012}.
@@
 $w_{\eumt}=1.032\ [1.031-1.033]$, all relative to the fitness of \euwt, which is set to $w_{\euwt}=1$.
-Thus, we can infer that the cost of trisomy is $c = w_{\eumt}-w_{\anmt} = 0.007$ (or 0.7\%) and the benefit of trisomy is $w_{\anwt}-1-c=0.017$ (1.7\%), whereas the benefit of beneficial mutation is $w_{\eumt}-1=0.032$ (3.2\%).
 
@@
 
+% FIG posterior F_A with high rates
+\begin{figure}[p]
+      \centering
+      \includegraphics[width=\textwidth]{../figures/FA_rates.pdf}      
+	\caption{\textbf{Effect of genomic instability on genotype frequencies.}
+	Genotype frequencies in the deterministic model without drift and with MAP parameter estimates \textbf{(A)}, with 100-fold increase in rate of chromosome loss (transition from $2n+1^*$ to $2n^*$) \textbf{(B)}, and with 10-fold increase in mutation rate in aneuploid cells \textbf{(C)}.
+Corresponding $F_A$ values are 0.098, 0.223, and 0.519.
+	}
+    \label{fig:FA_rates}
+\end{figure}
+
 \pagebreak

### 2024-01-21 ms/ms.tex
started with rev 2

@@
-% TODO
-% cite https://www.nature.com/articles/ng.3967
 \documentclass[12pt]{extarticle}
@@
 \citet{Yona2012} demonstrated experimentally the importance of aneuploidy in adaptive evolution. They evolved populations of \yeast under strong heat stress.
-The populations adapted to the heat stress within 450 generations, and this adaptation was determined to be due a duplication of Chromosome III.
+The populations adapted to the heat stress within 450 generations, and this adaptation was determined to be due a duplication of Chromosome~III.
 Later on, after more than 1,500 generations, the populations reverted back to an euploid state, while remaining adapted to the heat stress. 
-Aneuploidy was therefore suggested to be a \emph{transient adaptive solution}, because it can rapidly appear and fixate in the population under stressful conditions, and can then be rapidly lost when the cost of aneuploidy outweighs its benefit---after the stress is removed, or after "refined" beneficial mutations appear and fixate~\citep{Yona2012}.
-Furthermore, it has been suggested that aneuploidy is an evolutionary "stepping stone" that facilitates future adaptation by genetic mutations, which require more time to evolve~\citep{Yona2012,Yona2015}.
+Aneuploidy was therefore suggested to be a transient adaptive solution, because it can rapidly appear and take over the population under stressful conditions, and can then be rapidly lost when the cost of aneuploidy outweighs its benefit---after the stress is removed, or after refined beneficial mutations appear and fixate~\citep{Yona2012}.
+Furthermore, it has been suggested that aneuploidy is an evolutionary ``stepping stone'' that facilitates future adaptation by genetic mutations, which require more time to evolve~\citep{Yona2012,Yona2015}.
 
 %\paragraph{The present study.}
-Here, we test the hypothesis that aneuploidy is a an \emph{evolutionary stepping stone} that facilitates adaptive evolution by genetic mutations~\citet{Yona2015}.
+Here, we test the hypothesis that aneuploidy is a an evolutionary stepping stone that facilitates adaptive evolution by genetic mutations~\citet{Yona2015}.
 We develop an evolutionary genetic model and fit it to the experimental results of \citet{Yona2012} to predict the genotype frequency dynamics in the experimental populations, thereby estimating the frequency of evolved euploid cells that descended from aneuploid cells.
 Our results show that although aneuploidy reached high frequencies in the experimental populations, the majority of cells in the evolved euploid population likely did not descend from aneuploid cells, but rather directly from wild-type euploid cells.
-These suggests that at the lineage level, aneuploidy may be an evolutionary diversion, rather than a stepping stone, on the path to adaptation.
+These suggests that at the lineage level, aneuploidy may be an ``evolutionary diversion'', rather than a stepping stone, on the path to adaptation.
 
@@
 In the heat-stress experiment of \citet{Yona2012}, four populations of \yeast evolved under \SI{39}{\celsius}.
-Aneuploidy fixed in all four experimental repetitions in the first 450 generations.
+Aneuploidy reached high frequency (>95\%) in all four experimental repetitions in the first 450 generations.
 Two of the repetitions, marked $H2$ and $H4$, carried no large-scale duplications other than a Chromosome III trisomy.
@@
 
-The model includes the effects of natural selection, genetic drift, aneuploidy, and mutation, and follows a population of cells characterized by their genotype: 
+The model includes the effects of natural selection, genetic drift, aneuploidy, and mutation (i.e., other genetic variants), and follows a population of cells characterized by their genotype: 
 euploid wild-type, \euwt, is the ancestral diploid genotype; 
@@
 aneuploid mutant, \anmt, has and extra chromosome (like \anwt) and a beneficial mutation (like \eumt).
+Note that `mutation' here refers to point mutations and other genetic variants unrelated to aneuploidy.
 Fitness values of the different genotypes are denoted by $w_{\euwt}$, $w_{\eumt}$, $w_{\anwt}$, and $w_{\anmt}$, and the rate of mutation and aneuploidy are denoted by $\mu$ and $\delta$, respectively.
@@
 
-We fitted this model to the experimental results\citep{Yona2012} -- time for fixation (>95\%) and for loss (<5\%) of aneuploidy -- using approximate Bayesian computation with sequential Monte Carlo (ABC-SMC)~\citep{Sisson2009}, thereby inferring the model parameters: rates aneuploidy and mutation and the fitness of all genotypes.
+We fitted this model to the experimental results\citep{Yona2012}--time for fixation (frequency >95\%) and for loss (frequency <5\%) of aneuploidy--using approximate Bayesian computation with sequential Monte Carlo (ABC-SMC)~\citep{Sisson2009}, thereby inferring the model parameters: rates of aneuploidy (i.e., mis-segregation, non-disjunction) and mutation and the fitness of all genotypes.
 We then sampled posterior predictions for the genotype frequency dynamics using the estimated parameter values and compared different versions of the model to test additional hypotheses about the evolutionary process.
@@
 
-The estimated aneuploidy rate, $\delta=1.72\cdot10^{-3}\ [1.47\cdot10^{-3}-2.786\cdot10^{-3}]$ is higher than in previous studies: for Chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0\cdot10^{-5}-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
+The estimated rate of aneuploidy (i.e., mis-segregation, non-disjunction), $\delta=1.72\cdot10^{-3}\ [1.47\cdot10^{-3}-2.786\cdot10^{-3}]$ is higher than in previous studies: for Chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0\cdot10^{-5}-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
 However, this difference may be partly explained by an increased aneuploidy rate during heat stress: heat shock can increase the rate of chromosome fragment loss by 2-3 orders of magnitude~\citep{Chen2012a}.
@@
 We estimate that the aneuploidy rate (i.e., number of chromosome gains per generation) is $1.7 \cdot 10^{-3}$, higher than a previous estimate of $6.7 \cdot 10^{-6}$ (ref~~\citep{Zhu2016}).
-This may be due to genetic instability caused by heat stress~\citep{Chen2012a}.
+This may be due to genetic instability caused by heat stress~\citep{Chen2012a}, but we note that there is a general scarcity of empirical data on aneuploidy rates.
 In addition, we find no evidence for increased mutation rates in aneuploid cells. 
@@
 In their heat-stress experiment, four populations of \yeast evolved under \SI{39}{\celsius}. 
-Aneuploidy fixed in all four population in the first 450 generations.
-Hereafter, fixation or elimination of a genotype \emph{by generation $t$} means that more than 95\% or less than 5\% of the population carry the genotype at generation $t$, and possibly earlier.
+Aneuploidy fixed (frequency >95\%) in all four population in the first 450 generations. 
+Hereafter, fixation or elimination of a genotype by generation $t$ means that more than 95\% or less than 5\% of the population carry the genotype at generation $t$, and possibly earlier. 
 In the original analysis of \citet{Yona2012}, samples were routinely extracted from the evolving populations and tested for heat-shock tolerance.
 The first generation in which such indication was found was generation 200.
-Therefore, we determine that aneuploidy did not fix before generation 200.
+Therefore, we determine that aneuploidy did not reach high frequency before generation 200.
 The experiment continued with two populations, in which aneuploidy was eliminated by generation 1,700 and 2,350 while still under the same conditions of elevated heat (\SI{39}{\celsius}).
@@
 where $!\{\ldots\}$ is the "logical not" operator, $P^4(\ldots)$ is the 4th power of $P(\ldots)$, and all probabilities $P_{\tilde{\vec X}}(\ldots)$ are approximated from the results of the simulations $\tilde{\vec X}$.
-For example, $P_{\tilde{\vec X}}\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)$ is approximated by taking simulations in which \anwt\; fixed before generation 450 but not before generation 200, and computing the fraction of such simulations in which \eumt\; did not fix by generation 1,700, and hence aneuploidy did not extinct before generation 1,700.
+For example, $P_{\tilde{\vec X}}\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)$ is approximated by taking simulations in which \anwt\; fixed (reached >95\%) before generation 450 but not before generation 200, and computing the fraction of such simulations in which \eumt\; did not fix by generation 1,700, and hence aneuploidy did not extinct (reach <5\%) before generation 1,700.
 \Cref{fig:seeds} compares results with less and more simulated experiments, demonstrating that 1,000 simulations are likely sufficient.

### 2024-01-22 ms/ms.tex
finish rev 2 and added citations

@@
 \paragraph{Genetic instability in aneuploid cells.}
-It has been suggested that aneuploidy increases genetic instability: \citet{Sheltzer2011b} have demonstrated a fold increase of between 2.2 and 7.1 in mutation rate.
+It has been suggested that aneuploidy increases genetic instability: \citet{Sheltzer2011b} have demonstrated a fold increase of between 2.2 and 7.1 in the mutation rate of disomic yeast (rather than trisomic yeast, the focus of our analysis).
 Therefore, we inferred model parameters under the assumption that the mutation rate increases in aneuploid cells by a factor $\tau$ = 1, 33/32 (due to an additional chromosome), 2, 5, 10, or 100 (due to genetic instability).
@@
 We observe $\tau=100$ could be distinguished if accurate data was available for the waiting time until the frequency of \euwt\ to decrease below 95\%~(\Cref{fig:tau-plots}A) or for waiting time for the frequency of \anwt\ to either reach or go below 95\%~(\Cref{fig:tau-plots}B).
+
 Similarly, we did not find evidence for an increase in the rate of chromosome loss in aneuploid cells~\citep{Sheltzer2011b}, probably due to lack of statistical power.
-
 Nevertheless, increasing the rate of chromosome loss (transitions from $2n+1^*$ to $2n^*$) without increasing the rate of chromosome gain (transitions from $2n$ to $2n+1$) increases $F_A$ (\Cref{fig:FA_rates}B), but not to the same extent as increasing the rate of chromosome gain.
@@
 This may be due to genetic instability caused by heat stress~\citep{Chen2012a}, but we note that there is a general scarcity of empirical data on aneuploidy rates.
-In addition, we find no evidence for increased mutation rates in aneuploid cells. 
+In addition, we did not find evidence for increased mutation rates in aneuploid cells. 
 Previous empirical studies have suggested that genetic instability (e.g., elevated mutation rates) in aneuploid cells is due to stress associated with the aneuploid state~\citep{Bouchonville2009, Chen2012b, Zhu2012, Ippolito2021b}.
-However, in the experiment of \citet{Yona2012}, both the wild-type and the aneuploid were under heat stress, which may explain why we did not find evidence for an increased mutation rate specifically in aneuploid cells.
+However, in the experiment of \citet{Yona2012}, both the wild-type and the aneuploid were under heat stress, which may explain why we didn't find evidence for an increased mutation rate specifically in aneuploid cells.
+
+% haploids
+\paragraph{Effect of ploidy.}
+The evolutionary dynamics may change in haploid yeast, in which aneuploidy results in a second, rather than third, chromosome copy. 
+For example, it has been demonstrated that in haploid yeast drug resistance mainly evolves via recessive mutations and aneuploidy~\citep{Soncini2020}, whereas in diploid yeast it evolves via dominant mutations, aneuploidy, and gene/segmental duplications~\citep{Barney2021}.
+Thus, the second chromosome of disomic yeast may facilitate further adaptation via duplications, rearrangements, and increased mutational tolerance~\citep{Avecilla2023}, while decreasing the chance for adaptation via recessive mutations.
+Future models and experiments can consider this more complex genomic contexts.
 

### 2024-01-22 ms/ms.tex
revise effect of ploidy paragraph

@@
 The evolutionary dynamics may change in haploid yeast, in which aneuploidy results in a second, rather than third, chromosome copy. 
-For example, it has been demonstrated that in haploid yeast drug resistance mainly evolves via recessive mutations and aneuploidy~\citep{Soncini2020}, whereas in diploid yeast it evolves via dominant mutations, aneuploidy, and gene/segmental duplications~\citep{Barney2021}.
-Thus, the second chromosome of disomic yeast may facilitate further adaptation via duplications, rearrangements, and increased mutational tolerance~\citep{Avecilla2023}, while decreasing the chance for adaptation via recessive mutations.
-Future models and experiments can consider this more complex genomic contexts.
+For example, it has been demonstrated that drug resistance mainly evolves via recessive mutations and aneuploidy in haploid yeast~\citep{Soncini2020}, whereas in diploid yeast it evolves via dominant mutations, aneuploidy, and gene/segmental duplications~\citep{Barney2021}.
+Thus, the second chromosome copy of disomic yeast may facilitate further adaptation via duplications, rearrangements, and increased mutational tolerance~\citep{Avecilla2023}, while decreasing the chance for adaptation via recessive mutations.
+Future models and experiments can consider how ploidy and other genomic contexts affect the role of aneuploidy in adaptive evolution.
 

### 2024-01-22 ms/ms.tex
update line numbers; edited changes

@@
 In a previous evolutionary experiment with yeast, populations evolving under heat stress became aneuploid, only to later revert to euploidy after beneficial mutations accumulated.
-It was therefore suggested that aneuploidy is a "stepping stone" on the path to adaptation.
+It was therefore suggested that aneuploidy is a ``stepping stone'' on the path to adaptation.
 Here, we test this hypothesis.
 We use Bayesian inference to fit an evolutionary model with both aneuploidy and mutation the experimental results.
-We then predict the genotype frequency dynamics during the experiment, demonstrating that most of the evolved euploid population likely did not descend from aneuploid cells, but rather from the euploid wildtype population.
+We then predict the genotype frequency dynamics during the experiment, demonstrating that most of the evolved euploid population likely did not descend from aneuploid cells, but rather from the euploid wild-type population.
 Our model shows how the beneficial mutation supply--the product of population size and beneficial mutation rate--determines the evolutionary dynamics: with low supply, much of the evolved population descends from aneuploid cells; but with high supply, beneficial mutations are generated fast enough to outcompete aneuploidy due to its inherent fitness cost.
-Our results suggest that despite its potential fitness benefits under stress, aneuploidy can be an evolutionary "diversion" rather than a "stepping stone": it can delay, rather than facilitate, the adaptation of the population, and cells that become aneuploid may leave less descendants compared to cells that remain diploid.
+Our results suggest that despite its potential fitness benefits under stress, aneuploidy can be an evolutionary ``diversion'' rather than a ``stepping stone'': it can delay, rather than facilitate, the adaptation of the population, and cells that become aneuploid may leave less descendants compared to cells that remain diploid.
 \end{abstract}
@@
 %\paragraph{The present study.}
-Here, we test the hypothesis that aneuploidy is a an evolutionary stepping stone that facilitates adaptive evolution by genetic mutations~\citet{Yona2015}.
+Here, we test the hypothesis that aneuploidy is a an evolutionary ``stepping stone'' that facilitates adaptive evolution by genetic mutations~\citet{Yona2015}.
 We develop an evolutionary genetic model and fit it to the experimental results of \citet{Yona2012} to predict the genotype frequency dynamics in the experimental populations, thereby estimating the frequency of evolved euploid cells that descended from aneuploid cells.
 Our results show that although aneuploidy reached high frequencies in the experimental populations, the majority of cells in the evolved euploid population likely did not descend from aneuploid cells, but rather directly from wild-type euploid cells.
-These suggests that at the lineage level, aneuploidy may be an ``evolutionary diversion'', rather than a stepping stone, on the path to adaptation.
+These suggests that at the lineage level, aneuploidy may be an ``evolutionary diversion'', rather than a ``stepping stone'', on the path to adaptation.
 
@@
 
-Therefore, we do not find any evidence of an increase in mutation rate in aneuploid cells.
+Therefore, we do not find evidence of an increase in mutation rate in aneuploid cells.
 This may be because, unless the increase is strong ($\tau \ge 10$), it does not seem to affect our inference; or because Chromosome III is one of the smallest chromosomes \citep{Gilchrist2019}.
@@
 We inferred the rates of aneuploidy and mutation and their effects on fitness. 
-We estimate that the aneuploidy rate (i.e., number of chromosome gains per generation) is $1.7 \cdot 10^{-3}$, higher than a previous estimate of $6.7 \cdot 10^{-6}$ (ref~~\citep{Zhu2016}).
+We estimate that the aneuploidy rate (i.e., number of chromosome gains per generation) is $1.7 \cdot 10^{-3}$, higher than a previous estimate of $6.7 \cdot 10^{-6}$ (ref.~~\citep{Zhu2016}).
 This may be due to genetic instability caused by heat stress~\citep{Chen2012a}, but we note that there is a general scarcity of empirical data on aneuploidy rates.
@@
 \paragraph{Conclusions.}
-Here, we tested the hypothesis that aneuploidy cells are an evolutionary "stepping stone", or adaptive intermediate, between wild-type euploid cells and mutant euploid cells~\citep{Yona2015}. 
+Here, we tested the hypothesis that aneuploidy cells are an evolutionary ``stepping stone'', or adaptive intermediate, between wild-type euploid cells and mutant euploid cells~\citep{Yona2015}. 
 Our results suggest that, although it seems the population goes from euploid to aneuploid and back, this is not the case at the individual level.
@@
 That is, we assume chromosomes are gained and lost at the same rate, and we neglect events that form a less-fit genotype (i.e., \anwt\; to \euwt\; and \eumt\; to \anmt).
-A model that assumed increased rate of chromosome loss in aneuploid cells (as in \citet{Sheltzer2011b}) did not perform well, probably due to lack of statistical power, and was abandoned.
+A model that assumes an increased rate of chromosome loss in aneuploid cells (as in \citet{Sheltzer2011b}) did not perform well, probably due to lack of statistical power, and was abandoned.
 
@@
 Hereafter, fixation or elimination of a genotype by generation $t$ means that more than 95\% or less than 5\% of the population carry the genotype at generation $t$, and possibly earlier. 
-In the original analysis of \citet{Yona2012}, samples were routinely extracted from the evolving populations and tested for heat-shock tolerance.
+In the original analysis of \citet{Yona2012}, samples were routinely extracted from the evolving populations and tested for indication of heat-shock tolerance.
 The first generation in which such indication was found was generation 200.
@@
 	Genotype frequencies in the deterministic model without drift and with MAP parameter estimates \textbf{(A)}, with 100-fold increase in rate of chromosome loss (transition from $2n+1^*$ to $2n^*$) \textbf{(B)}, and with 10-fold increase in mutation rate in aneuploid cells \textbf{(C)}.
-Corresponding $F_A$ values are 0.098, 0.223, and 0.519.
+Corresponding $F_A$ values (purple line at generation 2,500) are 0.098, 0.223, and 0.519.
 	}

### 2024-01-28 ms/ms.tex
fixed genotypes labels

@@
 % genotype commands
-\newcommand{\euwt}{\emph{2n}}
-\newcommand{\anwt}{\emph{2n+1}}
-\newcommand{\eumt}{\emph{2n*}}
+\newcommand{\euwt}{\emph{$2n$}}
+\newcommand{\anwt}{\emph{$2n+1$}}
+\newcommand{\eumt}{\emph{$2n^*$}}
 \newcommand{\eumtM}{\emph{$2n^*_M$}}
 \newcommand{\eumtA}{\emph{$2n^*_A$}}
-\newcommand{\anmt}{\emph{2n+1*}}
+\newcommand{\anmt}{\emph{$2n+1^*$}}
 
@@
 We simulated 50 replicate genotype frequency dynamics using the MAP estimate parameters. 
-\Cref{fig:FA}A shows the simulated frequencies of the four genotypes (\euwt, \anwt, \anmt\ and \eumt), as well as the frequencies of \eumt\ cells that arose from either \anwt\ cells via a sequences of mutation and chromosome loss events ($2n^*_A$), or directly from \euwt\ cells via a mutation event (\eumtM).
+\Cref{fig:FA}A shows the simulated frequencies of the four genotypes (\euwt, \anwt, \anmt\ and \eumt), as well as the frequencies of \eumt\ cells that arose from either \anwt\ cells via a sequences of mutation and chromosome loss events (\eumtA), or directly from \euwt\ cells via a mutation event (\eumtM).
 We find that \anmt\ never reaches substantial frequency as it is quickly replaced by \eumt\ in a process similar to \emph{stochastic tunneling}~\citep{Iwasa2004,Komarova2003}.
@@
     \textbf{Predicted frequency of aneuploid-descended cells.}     
-    \textbf{(A)} Posterior predicted genotype frequencies over time, including the source of \eumt: $2n^*_A$ arose from $2n+1$, whereas $2n^*_M$ arose directly from $2n$.
+    \textbf{(A)} Posterior predicted genotype frequencies over time, including the source of \eumt: \eumtA\ arose from \anwt, whereas \eumtM\ arose directly from \euwt.
 Colored curves are 50 simulations using the MAP estimate parameters.
@@
 Similarly, we did not find evidence for an increase in the rate of chromosome loss in aneuploid cells~\citep{Sheltzer2011b}, probably due to lack of statistical power.
-Nevertheless, increasing the rate of chromosome loss (transitions from $2n+1^*$ to $2n^*$) without increasing the rate of chromosome gain (transitions from $2n$ to $2n+1$) increases $F_A$ (\Cref{fig:FA_rates}B), but not to the same extent as increasing the rate of chromosome gain.
+Nevertheless, increasing the rate of chromosome loss (transitions from \anmt\ to \eumt) without increasing the rate of chromosome gain (transitions from \euwt\ to \anwt) increases $F_A$ (\Cref{fig:FA_rates}B), but not to the same extent as increasing the rate of chromosome gain.
 In contrast, increasing the mutation rate in aneuploid cells can have a marked effect on the dynamics: when using the MAP parameter estimates, $F_A$ increases to  0.52 when the mutation rate in aneuploid cells increases 10-fold (\Cref{fig:FA_rates}C).
@@
 The mean $F_A$ in this case is just 0.0189 [0.0004 - 0.1214 95\% CI], lower than without the transitions to less-fit genotypes.
-Here, $F_A$ is the sum of frequencies of both \eumtA\ and $2n+1^*_A$, which reaches a frequency of 0.0007.
+Here, $F_A$ is the sum of frequencies of both \eumtA\ and \anmt, which reaches a frequency of 0.0007.
 Out of 100,000 posterior samples, none had $F_A$ above 0.05 (i.e., 5\% of the population).
@@
 	\caption{\textbf{Effect of genomic instability on genotype frequencies.}
-	Genotype frequencies in the deterministic model without drift and with MAP parameter estimates \textbf{(A)}, with 100-fold increase in rate of chromosome loss (transition from $2n+1^*$ to $2n^*$) \textbf{(B)}, and with 10-fold increase in mutation rate in aneuploid cells \textbf{(C)}.
+	Genotype frequencies in the deterministic model without drift and with MAP parameter estimates \textbf{(A)}, with 100-fold increase in rate of chromosome loss (transition from \anmt\ to \eumt) \textbf{(B)}, and with 10-fold increase in mutation rate in aneuploid cells \textbf{(C)}.
 Corresponding $F_A$ values (purple line at generation 2,500) are 0.098, 0.223, and 0.519.

### 2024-01-28 ms/ms.tex
finish IK comments

@@
 Aneuploidy is common in eukaryotes, often leading to decreased fitness.
-However, evidence from fungi and human tumour cells suggests that specific aneuploidies can be beneficial under stressful conditions and facilitate adaptation.
+However, evidence from fungi and human tumur cells suggests that specific aneuploidies can be beneficial under stressful conditions and facilitate adaptation.
 In a previous evolutionary experiment with yeast, populations evolving under heat stress became aneuploid, only to later revert to euploidy after beneficial mutations accumulated.
@@
 Here, we test this hypothesis.
-We use Bayesian inference to fit an evolutionary model with both aneuploidy and mutation the experimental results.
+We use Bayesian inference to fit an evolutionary model with both aneuploidy and mutation to the experimental results.
 We then predict the genotype frequency dynamics during the experiment, demonstrating that most of the evolved euploid population likely did not descend from aneuploid cells, but rather from the euploid wild-type population.
@@
 Similarly, we did not find evidence for an increase in the rate of chromosome loss in aneuploid cells~\citep{Sheltzer2011b}, probably due to lack of statistical power.
-Nevertheless, increasing the rate of chromosome loss (transitions from \anmt\ to \eumt) without increasing the rate of chromosome gain (transitions from \euwt\ to \anwt) increases $F_A$ (\Cref{fig:FA_rates}B), but not to the same extent as increasing the rate of chromosome gain.
-In contrast, increasing the mutation rate in aneuploid cells can have a marked effect on the dynamics: when using the MAP parameter estimates, $F_A$ increases to  0.52 when the mutation rate in aneuploid cells increases 10-fold (\Cref{fig:FA_rates}C).
+Nevertheless, increasing the rate of chromosome loss (transitions from \anmt\ to \eumt) without increasing the rate of chromosome gain (transitions from \euwt\ to \anwt) increases $F_A$ (\Cref{fig:FA_rates}B), but not to the same extent as increasing the rate of chromosome gain (\Cref{fig:posterior_FA_supp}).
+In contrast, increasing the mutation rate in aneuploid cells can have a marked effect on the dynamics: when using the MAP parameter estimates, $F_A$ increases from  0.1 to 0.52 when the mutation rate in aneuploid cells increases 10-fold (\Cref{fig:FA_rates}C).
 
@@
 	\caption{\textbf{Effect of genomic instability on genotype frequencies.}
-	Genotype frequencies in the deterministic model without drift and with MAP parameter estimates \textbf{(A)}, with 100-fold increase in rate of chromosome loss (transition from \anmt\ to \eumt) \textbf{(B)}, and with 10-fold increase in mutation rate in aneuploid cells \textbf{(C)}.
-Corresponding $F_A$ values (purple line at generation 2,500) are 0.098, 0.223, and 0.519.
+	Genotype frequencies in the deterministic model without drift and \textbf{(A)} with MAP parameter estimates; \textbf{(B)} with 100-fold increase in rate of chromosome loss (transition from \anmt\ to \eumt); or \textbf{(C)} with 10-fold increase in mutation rate in aneuploid cells (transition from \anwt\ to \anmt).
+Corresponding $F_A$ values (purple line at generation 2,500) are 0.098, 0.223, and 0.519, respectively.
 	}

### 2024-01-28 ms/ms.tex
AS

@@
 Later on, after more than 1,500 generations, the populations reverted back to an euploid state, while remaining adapted to the heat stress. 
-Aneuploidy was therefore suggested to be a transient adaptive solution, because it can rapidly appear and take over the population under stressful conditions, and can then be rapidly lost when the cost of aneuploidy outweighs its benefit---after the stress is removed, or after refined beneficial mutations appear and fixate~\citep{Yona2012}.
+Aneuploidy was therefore suggested to be a transient adaptive solution, because it can rapidly appear and take over the population under stressful conditions, and can then be rapidly lost when the cost of aneuploidy outweighs its benefit---after the stress is removed, or after refined beneficial mutations appear and fix~\citep{Yona2012}.
 Furthermore, it has been suggested that aneuploidy is an evolutionary ``stepping stone'' that facilitates future adaptation by genetic mutations, which require more time to evolve~\citep{Yona2012,Yona2015}.
@@
 %\paragraph{The present study.}
-Here, we test the hypothesis that aneuploidy is a an evolutionary ``stepping stone'' that facilitates adaptive evolution by genetic mutations~\citet{Yona2015}.
+Here, we test the hypothesis that aneuploidy is an evolutionary ``stepping stone'' that facilitates adaptive evolution by genetic mutations~\citep{Yona2015}.
 We develop an evolutionary genetic model and fit it to the experimental results of \citet{Yona2012} to predict the genotype frequency dynamics in the experimental populations, thereby estimating the frequency of evolved euploid cells that descended from aneuploid cells.
@@
 Furthermore, \citet{Yona2012} found at least 10 genes on Chromosome III that increased heat tolerance when over-expressed.
-Assuming that other chromosomes also have a similar number of heat-tolerance genes (and even more, as Chromosome III is one of the smallest chromosomes~\citep{Gilchrist2019}), we get a total of 160 heat-tolerance genes in the genome.
+Assuming that other chromosomes also have a similar number of heat-tolerance genes (and even more, as Chromosome III is one of the smallest chromosomes~\citep{Gilchrist2019}), we estimate a total of 160 heat-tolerance genes in the genome.
 Indeed, mutations were found in 97 genes in an evolutionary experiment with yeast under heat stress~\citep{Huang2018}.
@@
 The estimated rate of aneuploidy (i.e., mis-segregation, non-disjunction), $\delta=1.72\cdot10^{-3}\ [1.47\cdot10^{-3}-2.786\cdot10^{-3}]$ is higher than in previous studies: for Chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0\cdot10^{-5}-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
-However, this difference may be partly explained by an increased aneuploidy rate during heat stress: heat shock can increase the rate of chromosome fragment loss by 2-3 orders of magnitude~\citep{Chen2012a}.
+However, this difference may be partly explained by an increased rate of aneuploidy during heat stress: heat shock can increase the rate of chromosome fragment loss by 2-3 orders of magnitude~\citep{Chen2012a}.
 
@@
 \paragraph{Genetic instability in aneuploid cells.}
-It has been suggested that aneuploidy increases genetic instability: \citet{Sheltzer2011b} have demonstrated a fold increase of between 2.2 and 7.1 in the mutation rate of disomic yeast (rather than trisomic yeast, the focus of our analysis).
+It has been suggested that aneuploidy increases genomic instability: \citet{Sheltzer2011b} have demonstrated a fold increase of between 2.2 and 7.1 in the mutation rate of disomic yeast (rather than trisomic yeast, the focus of our analysis).
 Therefore, we inferred model parameters under the assumption that the mutation rate increases in aneuploid cells by a factor $\tau$ = 1, 33/32 (due to an additional chromosome), 2, 5, 10, or 100 (due to genetic instability).
@@
 This happens despite the fact that the mean fitness initially increases faster in the presence of aneuploidy (\Cref{fig:FA}E). 
-This is because once \anwt\ is common, selection for the mutant strain (\anmt\ or \eumt) is weaker compared to when \eumt\ competes directly with \euwt.
+Aneuploidy increases adaptation time because once \anwt\ is common, selection for the mutant strain (\anmt\ or \eumt) is weaker compared to when \eumt\ competes directly with \euwt.
+This is an interesting example of clonal interference~\citep{Good2012} but between fast and slow mutational processes~\citep{Kronholm2016}.
 
@@
 \paragraph{Conclusions.}
-Here, we tested the hypothesis that aneuploidy cells are an evolutionary ``stepping stone'', or adaptive intermediate, between wild-type euploid cells and mutant euploid cells~\citep{Yona2015}. 
+Here, we tested the hypothesis that aneuploid cells are an evolutionary ``stepping stone'', or adaptive intermediate, between wild-type euploid cells and mutant euploid cells in the evolutionary experiment of~\citet{Yona2015}.
 Our results suggest that, although it seems the population goes from euploid to aneuploid and back, this is not the case at the individual level.
@@
 Minerva Center for Live Emulation of Evolution in the Lab~(YR, YP),
-Minerva Stiftung short-term research grant~(MP).
+Minerva Stiftung short-term research grant~(MP),
+the Burroughs Wellcome Fund (AS),
+and the Jane Coffin Childs Memorial Fund (PVZ).
 }

### 2024-01-29 ms/ms.tex
format for MBE

@@
-\documentclass[12pt]{extarticle}
+\documentclass[12pt]{article}
 \usepackage{geometry}
@@
 \usepackage{longtable}
-\usepackage{gensymb,siunitx}
+\usepackage{gensymb}
 \usepackage{cleveref}
@@
 % NatBib
-\usepackage[super,comma]{natbib}
-%\usepackage[numbers,super,compress]{natbib}
-%\usepackage[round,colon,authoryear]{natbib}
+\usepackage[round,colon,authoryear]{natbib}
 
 % Title page
-% Chromosomal duplication is a transient evolutionary solution to stress
-% \title{Chromosomal duplication can be an evolutionary detour on the path to adaptation}
 \title{Aneuploidy can be an evolutionary diversion\\ on the path to adaptation}
-%\title{Aneuploidy can delay rather than accelerate adaptation}
-%\title{Aneuploidy can impede rather than spur adaptation}
 
@@
 However, aneuploidy can be beneficial under stressful conditions due to the wide range of phenotypes it can produce, some of which are advantageous~\citep{Pavelka2010, Yang2021}.
-Indeed, in a survey of 1,011 yeast strains, aneuploidy has been detected in about 19\%\citep{Peter2018}.
+Indeed, in a survey of 1,011 yeast strains, aneuploidy has been detected in about 19\% \citep{Peter2018}.
 Thus, aneuploidy can lead to rapid adaptation in unicellular eukaryotes~\citep{Gerstein2015,Torres2010, Hong2014, Rancati2008}, as well as to rapid growth of somatic tumour cells~\citep{Schvartzman2010, Sheltzer2017}.
@@
 %\paragraph{The present study.}
-Here, we test the hypothesis that aneuploidy is an evolutionary ``stepping stone'' that facilitates adaptive evolution by genetic mutations~\citep{Yona2015}.
+Here, we test the hypothesis that aneuploidy is an evolutionary ``stepping stone'' that facilitates adaptive evolution by genetic mutations~\citep{Yona2012}.
 We develop an evolutionary genetic model and fit it to the experimental results of \citet{Yona2012} to predict the genotype frequency dynamics in the experimental populations, thereby estimating the frequency of evolved euploid cells that descended from aneuploid cells.
@@
 \section*{Results}
-In the heat-stress experiment of \citet{Yona2012}, four populations of \yeast evolved under \SI{39}{\celsius}.
+In the heat-stress experiment of \citet{Yona2012}, four populations of \yeast evolved under 39\degree C.
 Aneuploidy reached high frequency (>95\%) in all four experimental repetitions in the first 450 generations.
@@
   \centering
-    \includegraphics[width=0.7\textwidth]{../figures/model_illustration.pdf}      
+    \includegraphics[width=0.7\textwidth]{model_illustration.pdf}      
   \caption{
@@
 The estimated beneficial mutation rate is $\mu=2.965\cdot10^{-6}\ [2.718\cdot10^{-7}-3.589\cdot10^{-6}]$ per genome per generation (that is, roughly 3 out of $10^6$ cell divisions produce a mutant cell with a fitness advantage).
-From the literature, the mutation rate per base pair is roughly $2-3\cdot10^{-10}$ (refs.~\citep{Zhu2014,Lynch2008}), but it may be higher under heat stress, as several stresses~\citep{Heidenreich2007}, including heat~\citep{Huang2018}, may cause hypermutation in yeast.
+From the literature, the mutation rate per base pair is roughly $2-3\cdot10^{-10}$~\citep{Zhu2014,Lynch2008}, but it may be higher under heat stress, as several stresses~\citep{Heidenreich2007}, including heat \citep{Huang2018}, may cause hypermutation in yeast.
 If we assume a 10-fold increase over the mutation rate reported in the literature, then the estimated beneficial mutation rate can be explained by a genomic target size of 1,000 base pairs (that is, 1,000 base pairs across the genome in which a mutation would provide a fitness advantage): $3\cdot10^{-10} \times 10 \times 1,000=3\cdot10^{-6}$.
-Supporting this, \citet{Jarolim2013} found 279 genes that contributed to survival after a sudden shift from \SI{30}{\celsius} to \SI{50}{\celsius}, and \citet{Flynn2020} used a deep mutational scan of a single protein, Hsp90, to find 465 amino-acid variants (out of 14,160) that significantly increased growth rate in \SI{37}{\celsius}.
+Supporting this, \citet{Jarolim2013} found 279 genes that contributed to survival after a sudden shift from 30\degree C to 50\degree C, and \citet{Flynn2020} used a deep mutational scan of a single protein, Hsp90, to find 465 amino-acid variants (out of 14,160) that significantly increased growth rate in 37\degree C.
 Furthermore, \citet{Yona2012} found at least 10 genes on Chromosome III that increased heat tolerance when over-expressed.
@@
   \centering
-\includegraphics[width=0.8\textwidth]{../figures/posterior.pdf}
+\includegraphics[width=0.8\textwidth]{posterior.pdf}
   \caption{
@@
       \centering
-      \includegraphics[width=\textwidth]{../figures/fixation-plot-a.pdf}      
+      \includegraphics[width=\textwidth]{fixation-plot-a.pdf}      
       \label{fig:fit}
@@
       \centering
-      \includegraphics[width=\textwidth]{../figures/fixation-plot-b.pdf}      
+      \includegraphics[width=\textwidth]{fixation-plot-b.pdf}      
       \label{fig:fit-cumulative}
@@
 The probability of \eumt\ descending from \anwt\ ($F_A$) increases with the aneuploidy rate, $\delta$, and decreases with both the population size $N$ and the mutation rate, $\mu$ (\Cref{fig:FA}C,D).
-In some cases it can also be affected by the fitness parameters (\Cref{fig:posterior_FA_supp}).
+In some cases it can also be affected by the fitness parameters (Figure S10).
 
@@
       \centering
-      \includegraphics[width=0.8\textwidth]{../figures/FA.pdf}      
+      \includegraphics[width=0.8\textwidth]{FA.pdf}      
   \caption{
@@
 Black dashed curves are the expected genotype frequencies without genetic drift (from a deterministic model).
-     See \Cref{fig:MAP_freq_dynamics_log} for log-log scale, in which the sequence of events is easier to observe.
+     See Figure S9 for log-log scale, in which the sequence of events is easier to observe.
     \textbf{(B)} Posterior distribution of $F_A$, the expected frequency of \eumt\ cells descended from \anwt\ cells, computed as the average frequency at the end of 100 simulations for 100,000 samples from the parameter posterior distribution.
@@
 	\textbf{(C)} $F_A$ values (color coded) from panel B, with their corresponding mutation rate $\mu$ on x-axis and aneuploidy rate $\delta$ on the y-axis.
-	Yellow star shows the MAP estimate. See also \Cref{fig:posterior_FA_supp}.
+	Yellow star shows the MAP estimate. See also Figure S10.
 	\textbf{(D)} $F_A$ as a function of the population size ($N$, bottom x-axis) and the beneficial mutation rate ($\mu$, top x-axis) in posterior predictions with MAP parameters. 
@@
 Therefore, we inferred model parameters under the assumption that the mutation rate increases in aneuploid cells by a factor $\tau$ = 1, 33/32 (due to an additional chromosome), 2, 5, 10, or 100 (due to genetic instability).
-We found that the posterior distribution was similar for $\tau=$1, 33/32, 2, and 5 (\Cref{fig:tau}).
+We found that the posterior distribution was similar for $\tau=$1, 33/32, 2, and 5 (Figure S4).
 Furthermore, we computed the WAIC, a criterion for model selection (Methods).
-The WAIC values were similar for all $\tau$ values (\Cref{table:WAIC}).
+The WAIC values were similar for all $\tau$ values (Table S1).
 
@@
 We also checked the differences in genotype frequency dynamics for different $\tau$ values.
-We observe $\tau=100$ could be distinguished if accurate data was available for the waiting time until the frequency of \euwt\ to decrease below 95\%~(\Cref{fig:tau-plots}A) or for waiting time for the frequency of \anwt\ to either reach or go below 95\%~(\Cref{fig:tau-plots}B).
+We observe $\tau=100$ could be distinguished if accurate data was available for the waiting time until the frequency of \euwt\ to decrease below 95\%~(Figure S5A) or for waiting time for the frequency of \anwt\ to either reach or go below 95\%~(Figure S5B).
 
 Similarly, we did not find evidence for an increase in the rate of chromosome loss in aneuploid cells~\citep{Sheltzer2011b}, probably due to lack of statistical power.
-Nevertheless, increasing the rate of chromosome loss (transitions from \anmt\ to \eumt) without increasing the rate of chromosome gain (transitions from \euwt\ to \anwt) increases $F_A$ (\Cref{fig:FA_rates}B), but not to the same extent as increasing the rate of chromosome gain (\Cref{fig:posterior_FA_supp}).
-In contrast, increasing the mutation rate in aneuploid cells can have a marked effect on the dynamics: when using the MAP parameter estimates, $F_A$ increases from  0.1 to 0.52 when the mutation rate in aneuploid cells increases 10-fold (\Cref{fig:FA_rates}C).
+Nevertheless, increasing the rate of chromosome loss (transitions from \anmt\ to \eumt) without increasing the rate of chromosome gain (transitions from \euwt\ to \anwt) increases $F_A$ (Figure S11B), but not to the same extent as increasing the rate of chromosome gain (Figure S10).
+In contrast, increasing the mutation rate in aneuploid cells can have a marked effect on the dynamics: when using the MAP parameter estimates, $F_A$ increases from  0.1 to 0.52 when the mutation rate in aneuploid cells increases 10-fold (Figure S11C).
 
@@
 Initially, the supply rates of \anwt\ and \eumtM\ are $N \delta \approx 11,000$ and $N \mu \approx 19$, respectively (assuming MAP parameter estimates). 
-Therefore, both genotypes are expected to appear immediately at the beginning of the experiment (\Cref{fig:MAP_freq_dynamics_log}).
+Therefore, both genotypes are expected to appear immediately at the beginning of the experiment (Figure S9).
 However, \anwt\ appears at a much higher frequency as $\delta \gg \mu$ by 2-3 orders of magnitude.
@@
 For the aneuploid lineage to compete with the mutant lineage, it must produce \eumtA\ via a mutation followed by chromosome loss.
-Although this is a stochastic process (due to drift), our results show that the time until \eumtA\ reaches a frequency of ~0.1\% is roughly 450 generations, without much variation (intersection of purple lines and vertical dashed line in \Cref{fig:MAP_freq_dynamics_log}).
+Although this is a stochastic process (due to drift), our results show that the time until \eumtA\ reaches a frequency of ~0.1\% is roughly 450 generations, without much variation (intersection of purple lines and vertical dashed line in Figure S9.
 However, by that time \eumtM\ is already at a roughly 10-fold higher frequency (1.86\%), and since both mutants have the same fitness, their relative frequency remains roughly the same until the end of the experiment.
@@
 We inferred the rates of aneuploidy and mutation and their effects on fitness. 
-We estimate that the aneuploidy rate (i.e., number of chromosome gains per generation) is $1.7 \cdot 10^{-3}$, higher than a previous estimate of $6.7 \cdot 10^{-6}$ (ref.~~\citep{Zhu2016}).
+We estimate that the aneuploidy rate (i.e., number of chromosome gains per generation) is $1.7 \cdot 10^{-3}$, higher than a previous estimate of $6.7 \cdot 10^{-6}$ \citep{Zhu2016}.
 This may be due to genetic instability caused by heat stress~\citep{Chen2012a}, but we note that there is a general scarcity of empirical data on aneuploidy rates.
@@
 \paragraph{Conclusions.}
-Here, we tested the hypothesis that aneuploid cells are an evolutionary ``stepping stone'', or adaptive intermediate, between wild-type euploid cells and mutant euploid cells in the evolutionary experiment of~\citet{Yona2015}.
+Here, we tested the hypothesis that aneuploid cells are an evolutionary ``stepping stone'', or adaptive intermediate, between wild-type euploid cells and mutant euploid cells in the evolutionary experiment of~\citet{Yona2012}.
 Our results suggest that, although it seems the population goes from euploid to aneuploid and back, this is not the case at the individual level.
@@
 We use the results of evolutionary experiments reported by \citet{Yona2012}.
-In their heat-stress experiment, four populations of \yeast evolved under \SI{39}{\celsius}. 
+In their heat-stress experiment, four populations of \yeast evolved under 39\degree C. 
 Aneuploidy fixed (frequency >95\%) in all four population in the first 450 generations. 
@@
 Therefore, we determine that aneuploidy did not reach high frequency before generation 200.
-The experiment continued with two populations, in which aneuploidy was eliminated by generation 1,700 and 2,350 while still under the same conditions of elevated heat (\SI{39}{\celsius}).
+The experiment continued with two populations, in which aneuploidy was eliminated by generation 1,700 and 2,350 while still under the same conditions of elevated heat (39\degree C).
 
@@
 For example, $P_{\tilde{\vec X}}\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)$ is approximated by taking simulations in which \anwt\; fixed (reached >95\%) before generation 450 but not before generation 200, and computing the fraction of such simulations in which \eumt\; did not fix by generation 1,700, and hence aneuploidy did not extinct (reach <5\%) before generation 1,700.
-\Cref{fig:seeds} compares results with less and more simulated experiments, demonstrating that 1,000 simulations are likely sufficient.
+Figure S1 compares results with less and more simulated experiments, demonstrating that 1,000 simulations are likely sufficient.
 
@@
 We terminated the inference iterations when the change in $\epsilon$ value from one iteration to the next was small.
-With our standard prior and model, we reached $\epsilon=0.13$ (or $\likelihood=0.87$) after six iterations, with $n_6=982$ accepted parameter vectors and effective sample size ESS=651 (\Cref{fig:convergence}).
-Running the inference algorithm with different initialization seeds and less or more simulations for approximating the likelihood produced similar posterior distributions~(\Cref{fig:seeds}).
+With our standard prior and model, we reached $\epsilon=0.13$ (or $\likelihood=0.87$) after six iterations, with $n_6=982$ accepted parameter vectors and effective sample size ESS=651 (Figure S2).
+Running the inference algorithm with different initialization seeds and less or more simulations for approximating the likelihood produced similar posterior distributions~(Figure S1).
 
@@
 From these predictions, relative fitness values can be computed.
-Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values to be used as a prior distribution by sampling 10,000 samples from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix~(\Cref{fig:growth-curves}).
+Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values to be used as a prior distribution by sampling 10,000 samples from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix~(Figure S3).
 
-We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate an informative prior distribution for $w_{\anwt}$~(\Cref{fig:growth-curves}D, assuming $w_{\euwt}=1$).
+We used growth curves of \euwt\ and \anwt\ in 39\degree C to estimate an informative prior distribution for $w_{\anwt}$~(Figure S3D, assuming $w_{\euwt}=1$).
 In this prior distribution, we used the same prior for $w_{\anmt}$ and $w_{\eumt}$. 
 To increase computational efficiency, we also assumed $w_\eumt>w_\anmt>w_\anwt>w_\euwt$; running the inference without this assumption produced similar results.
-See \emph{supporting material} for an extended informative prior distribution that uses growth curves of \eumt\ and \anwt\ growing in \SI{39}{\celsius}; this prior distribution proved to be less useful.
+See \emph{supporting material} for an extended informative prior distribution that uses growth curves of \eumt\ and \anwt\ growing in 39\degree C; this prior distribution proved to be less useful.
 
@@
 
-\newpage
-
-\section*{Supplementary Material}
-\beginsupplement % https://support.authorea.com/en-us/article/how-to-create-an-appendix-section-or-supplementary-information-1g25i5a/
-
-\subsection*{Supplementary Analysis}
-\label{sec:supp_analysis}
-
-\paragraph{Sensitivity analysis.} 
-Changing a single parameter while keeping the rest fixed at the MAP estimate produces a worse fit to the data~(\Cref{fig:sensitivity}).
-Furthermore, we fitted models with a mutation rate fixed at $\mu=10^{-5}$, $10^{-6}$ and $10^{-7}$.
-We inferred similar parameters estimates for the model with $\mu=10^{-6}$ compared to the model with a free $\mu$ parameter, in which the inferred mutation rate is $\mu \approx 3\cdot10^{-6}$.
-Inference assuming $\mu=10^{-5}$ or $\mu=10^{-7}$ produced similar estimates except that the estimated aneuploidy rate, $\delta$, was higher, and assuming $\mu=10^{-7}$, the estimated fitness of \anwt\ was lower (\Cref{fig:mu}).
-
-\paragraph{Extended informative prior distribution.}
-In an extended informative prior distribution, we used additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{\eumt}/w_{\anwt}$~(\Cref{fig:growth-curves}H).
-The same distribution was used for $w_{\eumt}/w_{\anmt}$.
-Thus, our main informative prior uses a single prior distribution for fitness values of \anwt, \anmt, and \eumt, whereas the extended informative prior uses one distribution for \anwt, and another distribution for both \anmt\ and \eumt.
-
-% generated with alt-prior.ipynb
-We estimated the parameters under this extended informative prior.
-Inference took much longer to run but the posterior distribution seemed to converge, as it did not change much in the final iterations. 
-The posterior predictive plot shows that inference with this extended prior produces a posterior distribution that fails to explain the empirical observations~(pink in Figure~\ref{fig:2n*-fixation}).
-However, the inferred posterior distribution is considerably narrower (compare \Cref{fig:posterior,fig:posterior-alt}) and therefore parameter estimates are less variable.
-The estimated mutation rate was much lower compared to the main informative prior, 
-% generated with MAP-and-hdi.ipynb
-with $\mu=2.474\cdot10^{-9}\ [2.423\cdot10^{-9}-2.612\cdot10^{-9}]$. Other parameter estimates are: $\delta=2.705\cdot10^{-3}\ [2.094\cdot10^{-3}-3.094\cdot10^{-3}]$,
-$w_{\anwt}=1.022\ [1.021-1.024]$,
-$w_{2n+1^*}=1.052\ [1.05-1.054]$,
-$w_{2n^*}=1.053\ [1.051-1.055]$, the latter two being much higher compare to the main informative prior. 
-Notably, the mode of the posterior ratio $w_{\eumt}/w_{\anwt}=1.0009$ is much lower than the mode of the prior ratio of $1.033$ (\Cref{fig:growth-curves}H) and closer to the ratio of $1$ that we assume in the main informative prior.
-Together with the posterior predictive results, we conclude that the main informative prior is preferable over the extended informative prior.
-
-\paragraph{Model with transitions to less-fit genotypes}
-
-We also estimated the parameters of a version of the model that includes transitions (mutation, chromosome loss and gain) to less-fit genotypes (e.g., \eumt\ to \anmt),
-\begin{equation} \label{eq:mutation-aneuploidy-single}
-    \begin{aligned}
-      &f^m_{2n} &=&\; (1 - \delta - \mu) f^s_{2n} + \delta f^s_{2n+1} + \mu f^s_{2n^*} \;,\\
-      &f^m_{2n+1} &=&\; \delta f^s_{2n} + (1 - \delta - \mu) f^s_{2n+1} + \mu f^s_{2n+1^*}  \;,\\
-      &f^m_{2n+1^*} &=&\; \mu f^s_{2n+1} + (1 - \delta - \mu) f^s_{2n+1^*} + \delta f^s_{2n^*}  \;,\\
-      &f^m_{2n^*} &=&\; \mu f^s_{2n} + \delta f^s_{2n+1^*} + (1 - \delta - \mu) f^s_{2n^*}  \;.
-    \end{aligned}
-    \end{equation}
-
-The inferred values are slightly different.
-% with back transitions MAP-and-hdi.ipynb
-The estimated mutation rate, $\mu=1.036\cdot10^{-7}\ [8.01\cdot10^{-8}-1.339\cdot10^{-7}]$, corresponds to a mutation target size of $\sim 300-500$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$ (ref.~\citep{Zhu2014}) or $3.3\cdot10^{-10}$ (ref.~\citep{Lynch2008}).
-The estimated aneuploidy rate, $\delta=2.358\cdot10^{-4}\ [1.766\cdot10^{-4}-2.837\cdot10^{-4}]$ is 5-35-fold higher than in previous studies: for Chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
-The estimated fitness values are 
-$w_{\anwt}=1.024\ [1.023-1.025]$,
-$w_{\anmt}=1.025\ [1.024-1.026]$,
-$w_{\eumt}=1.032\ [1.031-1.033]$, all relative to the fitness of \euwt, which is set to $w_{\euwt}=1$.
-
-We simulated genotype frequency dynamics using parameter samples from the posterior distribution, and computed the posterior distribution of $F_A$. 
-The mean $F_A$ in this case is just 0.0189 [0.0004 - 0.1214 95\% CI], lower than without the transitions to less-fit genotypes.
-Here, $F_A$ is the sum of frequencies of both \eumtA\ and \anmt, which reaches a frequency of 0.0007.
-Out of 100,000 posterior samples, none had $F_A$ above 0.05 (i.e., 5\% of the population).
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage
-\subsection*{Supplementary Figures \& Tables}
-
-% Fig Validation of posteriors
-%% generated with diff-runs.ipynb
-\begin{figure}[h]
-    \centering
-      \includegraphics[width=0.45\textwidth]{../figures/runs-A.pdf}      
-      \includegraphics[width=0.45\textwidth]{../figures/runs-B.pdf}    
-      \includegraphics[width=0.325\textwidth]{../figures/runs-C.pdf}      
-      \includegraphics[width=0.325\textwidth]{../figures/runs-D.pdf}      
-      \includegraphics[width=0.325\textwidth]{../figures/runs-E.pdf} 
-       \caption{
-    \textbf{Posterior distribution validation.}
-    The posterior distribution of model parameters is roughly the same regardless of the number of simulations (4-10,000 replicates) used to approximate the likelihood (\cref{eq:heatstress-likelihood}).
-    }
-
-     \label{fig:seeds}
- \end{figure}
-
-% Fig ABC iterations, convergence
-%% generated with convergence.ipynb
-\begin{figure}[p]
-
-  \begin{subfigure}{1\textwidth}
-    \centering
-      \includegraphics[width=0.45\textwidth]{../figures/convergence-p1_mr.pdf}      
-      \includegraphics[width=0.45\textwidth]{../figures/convergence-p2_tr.pdf} \\     
-      \includegraphics[width=0.325\textwidth]{../figures/convergence-p3_w1.pdf}      
-      \includegraphics[width=0.325\textwidth]{../figures/convergence-p4_w2.pdf}      
-      \includegraphics[width=0.325\textwidth]{../figures/convergence-p5_w3.pdf}      
-     \label{fig:convergence-A}
-    \end{subfigure}
-    \\    
-  \begin{subfigure}{1\textwidth}
-  	\includegraphics[width=\textwidth]{../figures/ess.pdf}
-     \label{fig:convergence-B}
-  \end{subfigure}
-    \caption{
-    \textbf{Inference convergence.} 
-    The ABC-SMC algorithm was used to infer the model parameters.
-\textbf{(A-E)} The approximate posterior distributions of model parameters at each iteration of the ABC-SMC algorithm demonstrates convergence, as the posterior did not significantly change after the first iteration, $t=1$.
-    \textbf{(F-I)} ABC-SMC measures of convergence.
-After iteration number 6, the acceptance threshold was $\epsilon=0.13$ (i.e., $\likelihood=0.87$, \cref{eq:heatstress-likelihood}), the acceptance rate was $0.018$, the number of particles was 982, and the effective sample size ESS=651.
-}
-    \label{fig:convergence}
-\end{figure}
-
-% Fig Curveball 
-\begin{figure}[p]
-	\centering
-	% generated with growth_curves.ipynb and deg = '39' 
-	\includegraphics[width=.95\textwidth]{../figures/evo39_fitness_39deg.pdf} 
-	% generated with growth_curves_refined.ipynb
-	\includegraphics[width=0.95\textwidth]{../figures/refined_vs_evo39_fitness_39deg.pdf}
-\caption{
-    \textbf{Fitness estimation from growth curves.}
-    \textbf{(A-D)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{39}{\celsius}. $ w_{\anwt}/w_{\euwt}$=1.024 (95\% CI: 0.959 - 1.115). \texttt{Curveball} 
-    \textbf{(E-H)} Fitness estimation from growth curves of \anwt\ and \eumt\ at \SI{39}{\celsius}. $ w_{\eumt}/w_{\anwt}$=1.033 (95\% CI: 1.027 - 1.041).
-    Growth curves previously described in \citet[Figs. 3C, 4A, and S2]{Yona2012}.
-	Fitness estimated from growth curves using \texttt{Curveball}, a method for predicting results of competition experiments from growth curve data~\citep[\href{https://curveball.yoavram.com}{curveball.yoavram.com}]{Ram2019}.
-See \emph{Models and Methods, Prior distributions} for more details.
-\textbf{(A,B;E,F)} Mono-culture growth curve data (markers) and best-fit growth models (lines).
-\textbf{(C,G)} The mixed-culture prediction for the strains from A,B and E,F respectively, 6,375 generated curves.
-\textbf{(D,H)} The relative fitness distribution for \anwt\ relative to \euwt\ (panel D) and \eumt\ relative to \anwt\ (panel H).
-Figures generated by \texttt{Curveball}.
-} 
-\label{fig:growth-curves}
-\end{figure}
-
-% Fig tau comparisons
-%% generated with diff-tau.ipynb
-\begin{figure}[p]
-  \centering
-  \begin{subfigure}{0.45\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-A.pdf}      
-  \end{subfigure}
-  \begin{subfigure}{0.45\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-B.pdf}      
-  \end{subfigure}
-  \\
-   \begin{subfigure}{0.325\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-C.pdf}      
-  \end{subfigure}
-    \begin{subfigure}{0.325\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-D.pdf}      
-  \end{subfigure}
-    \begin{subfigure}{0.325\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-E.pdf}      
-  \end{subfigure}
-  \\
-  \begin{subfigure}{0.325\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-joint-F.pdf}      
-  \end{subfigure}
-\begin{subfigure}{0.325\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-joint-G.pdf}      
-  \end{subfigure}
-\begin{subfigure}{0.325\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-joint-H.pdf}      
-  \end{subfigure}
-\begin{subfigure}{0.325\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-joint-I.pdf}      
-  \end{subfigure}
-\begin{subfigure}{0.325\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-joint-J.pdf}      
-  \end{subfigure}
-\begin{subfigure}{0.325\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-joint-K.pdf}      
-  \end{subfigure}
-  \caption{
-    \textbf{Model with elevated mutation rate in aneuploid cells.}  \textbf{(A-E)} The inferred posterior distributions for models with different values of $\tau$, the fold-increase in mutation rate in aneuploid cells (\anwt\ and \anmt).
-Vertical dashed lines represent the MAP (maximum a posteriori) of each distribution.
-When the increase in mutation rate is high, $\tau=10$ and $\tau=100$, the inferred mutation (A) and aneuploidy (B) rates tend to be lower.
-    \textbf{(F-K)} The inferred joint posterior distribution of mutation rate ($\mu$) and aneuploidy rate ($\delta$) with different $\tau$ values (dark purple and bright yellow for low and high density, respectively).
-  \label{fig:tau}
-  }
-  \end{figure}
-
-% Fig tau plots
-%% generated by diff-tau-more-plots.ipynb
-\begin{figure}[p]
-  \centering
-  \begin{subfigure}{0.75\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-diff-a.pdf}      
-  \end{subfigure}
-  \begin{subfigure}{0.75\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-diff-b.pdf}      
-  \end{subfigure}
-   \begin{subfigure}{0.75\textwidth}
-      \includegraphics[width=\textwidth]{../figures/tau-diff-c.pdf}      
-  \end{subfigure}
-   \caption{
-    \textbf{Genotype fixations for models with increased genetic instability.} We estimated the parameters for different models, each assuming a different value of $\tau$, the fold-increase in mutation rate in aneuploid cells.
-We then generated 10,000 simulations using the MAP estimate of each model and evaluated the fraction of simulations in which the frequency of genotype \euwt\ (\textbf{A}), \anwt\ (\textbf{B}), and \eumt\ (\textbf{C}) is above 95\% (y-axis) at each generation (x-axis).
-Note that \anmt\ did not fix.
-We can see that $\tau=100$ can be distinguished if the waiting time for $f_{\euwt}<95\%$ is known (panel~A) or if the waiting time for $f_{\anwt}>95\%$ or $f_{\anwt}<95\%$ is known (panel~B).
-It is harder to distinguish between $1\leq \tau \leq10$.
-  }
-  \label{fig:tau-plots}
-\end{figure}
-
-% Fig likelihood profile
-%% generated by sensitivity-analysis.ipynb
-\begin{figure}[p]
-  \centering
-  \begin{subfigure}{0.3\textwidth}
-      \includegraphics[width=\textwidth]{../figures/sensitivity-A.pdf}      
-      \label{fig:sensitivity-mutation}
-  \end{subfigure}
-  \begin{subfigure}{0.3\textwidth}
-      \includegraphics[width=\textwidth]{../figures/sensitivity-B.pdf}      
-      \label{fig:sensitivity-aneuploidy}
-  \end{subfigure}
-  \\
-   \begin{subfigure}{0.3\textwidth}
-      \includegraphics[width=\textwidth]{../figures/sensitivity-C.pdf}      
-      \label{fig:sensitivity-anwt}
-  \end{subfigure}
-    \begin{subfigure}{0.3\textwidth}
-      \includegraphics[width=\textwidth]{../figures/sensitivity-D.pdf}      
-      \label{fig:sensitivity-anmt}
-  \end{subfigure}
-    \begin{subfigure}{0.3\textwidth}
-      \includegraphics[width=\textwidth]{../figures/sensitivity-E.pdf}      
-      \label{fig:sensitivity-eumt}
-  \end{subfigure}
-  \caption{
-    \textbf{Likelihood profiles.} Sensitivity of the model approximate likelihood, $\likelihood(\theta)$, to changing a single parameter while the other parameters remain fixed at their MAP estimates.
-Dashed vertical line represents the MAP value.
-The prior distributions for the mutation rate and aneuploidy rate are $\mu \sim U(10^{-9}, 10^{-5})$ and $\delta \sim U(10^{-6}, 10^{-2})$, respectively.
-  }
-
-  \label{fig:sensitivity}
-\end{figure}
-
-% Fig mu comparisons
-%% generated with diff-mutation-rate.ipynb
-\begin{figure}[p]
-  \centering
-  \begin{subfigure}{0.45\textwidth}
-      \includegraphics[width=\textwidth]{../figures/mu-A.pdf}      
-  \end{subfigure}
-  \begin{subfigure}{0.45\textwidth}
-      \includegraphics[width=\textwidth]{../figures/mu-B.pdf}      
-  \end{subfigure}
-  \\
-   \begin{subfigure}{0.45\textwidth}
-      \includegraphics[width=\textwidth]{../figures/mu-C.pdf}      
-  \end{subfigure}
-    \begin{subfigure}{0.45\textwidth}
-      \includegraphics[width=\textwidth]{../figures/mu-D.pdf}      
-    \end{subfigure}
-  \caption{
-    \textbf{Model with fixed mutation rate.}  \textbf{(A-D)} The inferred posterior distributions for models with free and fixed mutation rate, $\mu$.
-The MAP (maximum a posteriori) and 50\% HDI (highest density interval) for each model are: %MAP-and-hdi.ipynb
-\textbf{free $\boldsymbol{\mu}$, run 1:}
-$\delta=1.720\cdot10^{-3}\ [1.470\cdot10^{-3}-2.786\cdot10^{-3}]$, 
-$w_{\anwt}=1.022\ [1.021-1.023]$,
-$w_{\anmt}=1.025\ [1.024-1.026]$,
-$w_{\eumt}=1.028\ [1.026-1.029]$;
-\textbf{free $\boldsymbol{\mu}$, run 2:}
-$\delta=2.129\cdot10^{-3}\ [1.334\cdot10^{-3}-2.695\cdot10^{-3}]$,
-$w_{2n+1}=1.022\ [1.02-1.023]$,
-$w_{2n+1^*}=1.025\ [1.023-1.026]$,
-$w_{2n^*}=1.028\ [1.026-1.029]$;
-\textbf{$\boldsymbol{\mu=10^{-5}}$:}
-$\delta=2.903\cdot10^{-3}\ [2.399\cdot10^{-3}-3.156\cdot10^{-3}]$,
-$w_{2n+1}=1.022\ [1.021-1.023]$,
-$w_{2n+1^*}=1.024\ [1.023-1.025]$,
-$w_{2n^*}=1.027\ [1.026-1.028]$;
-\textbf{$\boldsymbol{\mu=10^{-6}}$:}
-$\delta=1.917\cdot10^{-3}\ [9.624\cdot10^{-4}-2.447\cdot10^{-3}]$,
-$w_{2n+1}=1.022\ [1.02-1.023]$,
-$w_{2n+1^*}=1.025\ [1.023-1.026]$,
-$w_{2n^*}=1.028\ [1.027-1.029]$;
-\textbf{$\boldsymbol{\mu=10^{-7}}$:}
-$\delta=2.901\cdot10^{-3}\ [2.139\cdot10^{-3}-3.671\cdot10^{-3}]$,
-$w_{2n+1}=1.019\ [1.017-1.02]$,
-$w_{2n+1^*}=1.025\ [1.024-1.026]$,
-$w_{2n^*}=1.027\ [1.026-1.029]$.
-\label{fig:mu}
-  }
-\end{figure}
-
-% FIG posterior alt-prior
-%% generated with alt-prior.ipynb
-\begin{figure}[p]
-  \centering
-  \includegraphics[width=0.9\textwidth]{../figures/posterior-alt.pdf}
-  \caption{
-  \textbf{Posterior distribution of parameters inferred with the extended prior distribution.}
-On the diagonal, the inferred posterior distribution of each model parameter. 
-Below the diagonal, the inferred joint posterior distribution of pairs of model parameters (dark purple and bright yellow for low and high density, respectively).
-Red markers and orange lines for the joint MAP estimate (which may differ from the marginal MAP, as the marginal distribution integrates over all other parameters).
-} 
-  \label{fig:posterior-alt}
-\end{figure}
-
-% FIG MAP frequency dynamics in log scale
-\begin{figure}[p]
-    \centering
-	\includegraphics[width=\textwidth]{../figures/MAP_freq_dynamics.pdf}
-	\caption{\textbf{Posterior predicted genotype frequencies in log-log scale.}
-	Frequency dynamics of the different genotypes with MAP parameter estimates, same as \Cref{fig:FA}A, but in log-log scale. Black dashed curves for a deterministic model without genetic drift. 
-	Clearly, appearance of \anwt\ and \eumtM\ is deterministic.
-	Appearance of \anmt, and therefore \eumtA, is stochastic, however, the frequency dynamics are deterministic above a frequency of roughly 0.001. 
-	Note that the \eumtM\ and the \eumtA+\eumtM\ lines are overlapping for much of their trajectories.
-	}
-    \label{fig:MAP_freq_dynamics_log}    
-\end{figure}
-
-% FIG posterior F_A supplementary
-\begin{figure}[p]
-	\begin{subfigure}{0.5\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../figures/F_A_delta-mu-ratio_1.pdf}      
-	\end{subfigure}
-	  \begin{subfigure}{0.5\textwidth}
-      \centering
-      \includegraphics[width=\textwidth]{../figures/F_A_delta-mu-ratio_2.pdf}
-  \end{subfigure}
-	\caption{\textbf{Posterior distribution of $F_A$.}
-	\textbf{(A,B)} $F_A$ values (color coded) as in \Cref{fig:FA} for different parameter choices on the x- and y-axes. White star denotes the MAP estimate.
-	}
-    \label{fig:posterior_FA_supp}
-\end{figure}
-
-% FIG posterior F_A with high rates
-\begin{figure}[p]
-      \centering
-      \includegraphics[width=\textwidth]{../figures/FA_rates.pdf}      
-	\caption{\textbf{Effect of genomic instability on genotype frequencies.}
-	Genotype frequencies in the deterministic model without drift and \textbf{(A)} with MAP parameter estimates; \textbf{(B)} with 100-fold increase in rate of chromosome loss (transition from \anmt\ to \eumt); or \textbf{(C)} with 10-fold increase in mutation rate in aneuploid cells (transition from \anwt\ to \anmt).
-Corresponding $F_A$ values (purple line at generation 2,500) are 0.098, 0.223, and 0.519, respectively.
-	}
-    \label{fig:FA_rates}
-\end{figure}
-
-\pagebreak
-
-%% generated with WAIC.ipynb 
-\begin{table}[h]
-\centering
-\caption{
-\textbf{WAIC values for different $\tau$ values.}
-Differences of less than 6 are considered of weak significance~\citep{Kass1995}.
-}
-\pgfplotstabletypeset[
-    col sep=semicolon,
-    string type,
-    every head row/.style={before row=\hline,after row=\hline},
-    every last row/.style={after row=\hline},
-    ]{../figures/Table_WAIC.csv} 
-\caption*{
-	WAIC defined in \cref{eq:WAIC}.
-} 
-\label{table:WAIC}
-\end{table}
-
 \end{document}

### 2024-01-29 ms/si.tex
format for MBE

\documentclass[12pt]{extarticle}
\usepackage{geometry}
\geometry{
a4paper,
total={170mm,257mm},
left=20mm,
top=20mm,
headheight=12pt
}
\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx} % Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
% TeX will automatically convert eps --> pdf in pdflatex

\usepackage{amssymb,amsmath,amsthm}
\usepackage{commath}
\usepackage{longtable}
\usepackage[hyphens]{url}
\usepackage[dvipsnames]{xcolor}
\usepackage[unicode=true,colorlinks=true,urlcolor=CadetBlue,citecolor=black,linkcolor=black]{hyperref}
\usepackage{booktabs}
\usepackage{pgfplotstable} % http://mirrors.ctan.org/graphics/pgf/contrib/pgfplots/doc/pgfplotstable.pdf
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage{authblk}
\usepackage{subcaption}
\captionsetup[subfigure]{labelformat=empty}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{longtable}
\usepackage{gensymb,siunitx}
\usepackage{cleveref}

%SetFonts
% newtxtext+newtxmath
\usepackage{newtxtext} %loads helv for ss, txtt for tt
\usepackage{amsmath}
\usepackage[bigdelims]{newtxmath}
\usepackage[T1]{fontenc}
\usepackage{textcomp}
%SetFonts

\renewcommand{\baselinestretch}{1.5}

% less space before sections 
% \@startsection {NAME}{LEVEL}{INDENT}{BEFORESKIP}{AFTERSKIP}{STYLE} 
%            optional * [ALTHEADING]{HEADING} 
\makeatletter
 \renewcommand\section{\@startsection {section}{1}{\z@}%
     {-2.5ex \@plus -1ex \@minus -.2ex}%
     {1.3ex \@plus.2ex}%
    {\Large\bfseries}}

% Species names
%% Meta-Command for defining new species macros
\usepackage{xspace}

\newcommand{\species}[3]{%
  \newcommand{#1}{\gdef#1{\textit{#3}\xspace}\textit{#2}\xspace}}

\species{\yeast}{Saccharomyces cerevisiae}{S.~cerevisiae}
\species{\calbicans}{Candida albicans}{C.~albicans}
\species{\cneoformans}{Cryptococcus neoformans}{C.~neoformans}

% Yoav & Lee commands
\newcommand*{\tr}{^\intercal}
\let\vec\mathbf
\newcommand{\matrx}[1]{{\left[ \stackrel{}{#1}\right]}}
\newcommand{\diag}[1]{\mbox{diag}\matrx{#1}}
\newcommand{\goesto}{\rightarrow}
\newcommand{\dspfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2} }
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{remark}{Remark}
\newtheorem{result}{Result}
\renewcommand\qedsymbol{} % no square at end of proof
\newcommand{\cl}{\mathbf{L}}
\newcommand{\cj}{\mathbf{J}}
\newcommand{\ci}{I}
\newcommand{\likelihood}{\mathcal{L}}

% line numbers
 \usepackage[displaymath, mathlines]{lineno}
 \renewcommand\linenumberfont{\normalfont\small\sffamily}
\linenumbers
\modulolinenumbers[2]

% genotype commands
\newcommand{\euwt}{\emph{$2n$}}
\newcommand{\anwt}{\emph{$2n+1$}}
\newcommand{\eumt}{\emph{$2n^*$}}
\newcommand{\eumtM}{\emph{$2n^*_M$}}
\newcommand{\eumtA}{\emph{$2n^*_A$}}
\newcommand{\anmt}{\emph{$2n+1^*$}}

% Supplementary
\newcommand{\beginsupplement}{%
      	\setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
}

% NatBib
\usepackage[round,colon,authoryear]{natbib}

% Title page
\title{Supplementary Material}

% Authors
\renewcommand\Affilfont{\small}

\author{Ilia Kohanovski, Martin Pontz, P\'{e}tra Vande Zande, Anna Selmecki, Orna Dahan, Yitzhak Pilpel, Avihu H. Yona, Yoav Ram}

% Document
\begin{document}
\maketitle

\beginsupplement % https://support.authorea.com/en-us/article/how-to-create-an-appendix-section-or-supplementary-information-1g25i5a/

\subsection*{Supplementary Analysis}
\label{sec:supp_analysis}

\paragraph{Sensitivity analysis.} 
Changing a single parameter while keeping the rest fixed at the MAP estimate produces a worse fit to the data~(\Cref{fig:sensitivity}).
Furthermore, we fitted models with a mutation rate fixed at $\mu=10^{-5}$, $10^{-6}$ and $10^{-7}$.
We inferred similar parameters estimates for the model with $\mu=10^{-6}$ compared to the model with a free $\mu$ parameter, in which the inferred mutation rate is $\mu \approx 3\cdot10^{-6}$.
Inference assuming $\mu=10^{-5}$ or $\mu=10^{-7}$ produced similar estimates except that the estimated aneuploidy rate, $\delta$, was higher, and assuming $\mu=10^{-7}$, the estimated fitness of \anwt\ was lower (\Cref{fig:mu}).

\paragraph{Extended informative prior distribution.}
In an extended informative prior distribution, we used additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{\eumt}/w_{\anwt}$~(\Cref{fig:growth-curves}H). The same distribution was used for $w_{\eumt}/w_{\anmt}$. 
Thus, our main informative prior uses a single prior distribution for fitness values of \anwt, \anmt, and \eumt, whereas the extended informative prior uses one distribution for \anwt, and another distribution for both \anmt\ and \eumt.

% generated with alt-prior.ipynb
We estimated the parameters under this extended informative prior.
Inference took much longer to run but the posterior distribution seemed to converge, as it did not change much in the final iterations. 
The posterior predictive plot shows that inference with this extended prior produces a posterior distribution that fails to explain the empirical observations~(pink in Figure 3).
However, the inferred posterior distribution is considerably narrower (compare Figure 2 and \Cref{fig:posterior-alt}) and therefore parameter estimates are less variable.
The estimated mutation rate was much lower compared to the main informative prior, 
% generated with MAP-and-hdi.ipynb
with $\mu=2.474\cdot10^{-9}\ [2.423\cdot10^{-9}-2.612\cdot10^{-9}]$. Other parameter estimates are: $\delta=2.705\cdot10^{-3}\ [2.094\cdot10^{-3}-3.094\cdot10^{-3}]$,
$w_{\anwt}=1.022\ [1.021-1.024]$,
$w_{2n+1^*}=1.052\ [1.05-1.054]$,
$w_{2n^*}=1.053\ [1.051-1.055]$, the latter two being much higher compare to the main informative prior. 
Notably, the mode of the posterior ratio $w_{\eumt}/w_{\anwt}=1.0009$ is much lower than the mode of the prior ratio of $1.033$ (\Cref{fig:growth-curves}H) and closer to the ratio of $1$ that we assume in the main informative prior.
Together with the posterior predictive results, we conclude that the main informative prior is preferable over the extended informative prior.

\paragraph{Model with transitions to less-fit genotypes}

We also estimated the parameters of a version of the model that includes transitions (mutation, chromosome loss and gain) to less-fit genotypes (e.g., \eumt\ to \anmt),
\begin{equation} \label{eq:mutation-aneuploidy-single}
    \begin{aligned}
      &f^m_{2n} &=&\; (1 - \delta - \mu) f^s_{2n} + \delta f^s_{2n+1} + \mu f^s_{2n^*} \;,\\
      &f^m_{2n+1} &=&\; \delta f^s_{2n} + (1 - \delta - \mu) f^s_{2n+1} + \mu f^s_{2n+1^*}  \;,\\
      &f^m_{2n+1^*} &=&\; \mu f^s_{2n+1} + (1 - \delta - \mu) f^s_{2n+1^*} + \delta f^s_{2n^*}  \;,\\
      &f^m_{2n^*} &=&\; \mu f^s_{2n} + \delta f^s_{2n+1^*} + (1 - \delta - \mu) f^s_{2n^*}  \;.
    \end{aligned}
    \end{equation}

The inferred values are slightly different.
% with back transitions MAP-and-hdi.ipynb
The estimated mutation rate, $\mu=1.036\cdot10^{-7}\ [8.01\cdot10^{-8}-1.339\cdot10^{-7}]$, corresponds to a mutation target size of $\sim 300-500$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$ \citep{Zhu2014} or $3.3\cdot10^{-10}$ \citep{Lynch2008}.
The estimated aneuploidy rate, $\delta=2.358\cdot10^{-4}\ [1.766\cdot10^{-4}-2.837\cdot10^{-4}]$ is 5-35-fold higher than in previous studies: for Chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
The estimated fitness values are 
$w_{\anwt}=1.024\ [1.023-1.025]$,
$w_{\anmt}=1.025\ [1.024-1.026]$,
$w_{\eumt}=1.032\ [1.031-1.033]$, all relative to the fitness of \euwt, which is set to $w_{\euwt}=1$.

We simulated genotype frequency dynamics using parameter samples from the posterior distribution, and computed the posterior distribution of $F_A$. 
The mean $F_A$ in this case is just 0.0189 [0.0004 - 0.1214 95\% CI], lower than without the transitions to less-fit genotypes. Here, $F_A$ is the sum of frequencies of both \eumtA\ and \anmt, which reaches a frequency of 0.0007. Out of 100,000 posterior samples, none had $F_A$ above 0.05 (i.e., 5\% of the population).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
%\subsection*{Supplementary Figures \& Tables}

% Fig Validation of posteriors
%% generated with diff-runs.ipynb
\begin{figure}[p]
    \centering
      \includegraphics[width=0.45\textwidth]{runs-A.pdf}      
      \includegraphics[width=0.45\textwidth]{runs-B.pdf}    
      \includegraphics[width=0.325\textwidth]{runs-C.pdf}      
      \includegraphics[width=0.325\textwidth]{runs-D.pdf}      
      \includegraphics[width=0.325\textwidth]{runs-E.pdf} 
       \caption{
    \textbf{Posterior distribution validation.}
    The posterior distribution of model parameters is roughly the same regardless of the number of simulations (4-10,000 replicates) used to approximate the likelihood (eq. 4).
    }

     \label{fig:seeds}
 \end{figure}

% Fig ABC iterations, convergence
%% generated with convergence.ipynb
\begin{figure}[p]
  \begin{subfigure}{1\textwidth}
    \centering
      \includegraphics[width=0.45\textwidth]{convergence-p1_mr.pdf}      
      \includegraphics[width=0.45\textwidth]{convergence-p2_tr.pdf} \\     
      \includegraphics[width=0.325\textwidth]{convergence-p3_w1.pdf}      
      \includegraphics[width=0.325\textwidth]{convergence-p4_w2.pdf}      
      \includegraphics[width=0.325\textwidth]{convergence-p5_w3.pdf}      
     \label{fig:convergence-A}
    \end{subfigure}
    \\    
  \begin{subfigure}{1\textwidth}
  	\includegraphics[width=\textwidth]{ess.pdf}
     \label{fig:convergence-B}
  \end{subfigure}
    \caption{
    \textbf{Inference convergence.} 
    The ABC-SMC algorithm was used to infer the model parameters. \textbf{(A-E)} The approximate posterior distributions of model parameters at each iteration of the ABC-SMC algorithm demonstrates convergence, as the posterior did not significantly change after the first iteration, $t=1$.
    \textbf{(F-I)} ABC-SMC measures of convergence. After iteration number 6, the acceptance threshold was $\epsilon=0.13$ (i.e., $\likelihood=0.87$, eq.~4), the acceptance rate was $0.018$, the number of particles was 982, and the effective sample size ESS=651.
}
    \label{fig:convergence}
\end{figure}

% Fig Curveball 
\begin{figure}[p]
	\centering
	% generated with growth_curves.ipynb and deg = '39' 
	\includegraphics[width=.95\textwidth]{evo39_fitness_39deg.pdf} 
	% generated with growth_curves_refined.ipynb
	\includegraphics[width=0.95\textwidth]{refined_vs_evo39_fitness_39deg.pdf}
\caption{
    \textbf{Fitness estimation from growth curves.}
    \textbf{(A-D)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{39}{\celsius}. $ w_{\anwt}/w_{\euwt}$=1.024 (95\% CI: 0.959 - 1.115). \texttt{Curveball} 
    \textbf{(E-H)} Fitness estimation from growth curves of \anwt\ and \eumt\ at \SI{39}{\celsius}. $ w_{\eumt}/w_{\anwt}$=1.033 (95\% CI: 1.027 - 1.041).
    Growth curves previously described in \citet[Figs. 3C, 4A, and S2]{Yona2012}.
	Fitness estimated from growth curves using \texttt{Curveball}, a method for predicting results of competition experiments from growth curve data~\citep[\href{https://curveball.yoavram.com}{curveball.yoavram.com}]{Ram2019}. See \emph{Models and Methods, Prior distributions} for more details.  \textbf{(A,B;E,F)} Mono-culture growth curve data (markers) and best-fit growth models (lines).
\textbf{(C,G)} The mixed-culture prediction for the strains from A,B and E,F respectively, 6,375 generated curves. \textbf{(D,H)} The relative fitness distribution for \anwt\ relative to \euwt\ (panel D) and \eumt\ relative to \anwt\ (panel H). Figures generated by \texttt{Curveball}.
} 
\label{fig:growth-curves}
\end{figure}

% Fig tau comparisons
%% generated with diff-tau.ipynb
\begin{figure}[p]
  \centering
  \begin{subfigure}{0.45\textwidth}
      \includegraphics[width=\textwidth]{tau-A.pdf}      
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
      \includegraphics[width=\textwidth]{tau-B.pdf}      
  \end{subfigure}
  \\
   \begin{subfigure}{0.325\textwidth}
      \includegraphics[width=\textwidth]{tau-C.pdf}      
  \end{subfigure}
    \begin{subfigure}{0.325\textwidth}
      \includegraphics[width=\textwidth]{tau-D.pdf}      
  \end{subfigure}
    \begin{subfigure}{0.325\textwidth}
      \includegraphics[width=\textwidth]{tau-E.pdf}      
  \end{subfigure}
  \\
  \begin{subfigure}{0.325\textwidth}
      \includegraphics[width=\textwidth]{tau-joint-F.pdf}      
  \end{subfigure}
\begin{subfigure}{0.325\textwidth}
      \includegraphics[width=\textwidth]{tau-joint-G.pdf}      
  \end{subfigure}
\begin{subfigure}{0.325\textwidth}
      \includegraphics[width=\textwidth]{tau-joint-H.pdf}      
  \end{subfigure}
\begin{subfigure}{0.325\textwidth}
      \includegraphics[width=\textwidth]{tau-joint-I.pdf}      
  \end{subfigure}
\begin{subfigure}{0.325\textwidth}
      \includegraphics[width=\textwidth]{tau-joint-J.pdf}      
  \end{subfigure}
\begin{subfigure}{0.325\textwidth}
      \includegraphics[width=\textwidth]{tau-joint-K.pdf}      
  \end{subfigure}
  \caption{
    \textbf{Model with elevated mutation rate in aneuploid cells.}  \textbf{(A-E)} The inferred posterior distributions for models with different values of $\tau$, the fold-increase in mutation rate in aneuploid cells (\anwt\ and \anmt). Vertical dashed lines represent the MAP (maximum a posteriori) of each distribution. When the increase in mutation rate is high, $\tau=10$ and $\tau=100$, the inferred mutation (A) and aneuploidy (B) rates tend to be lower. 
    \textbf{(F-K)} The inferred joint posterior distribution of mutation rate ($\mu$) and aneuploidy rate ($\delta$) with different $\tau$ values (dark purple and bright yellow for low and high density, respectively).
  \label{fig:tau}
  }
  \end{figure}

% Fig tau plots
%% generated by diff-tau-more-plots.ipynb
\begin{figure}[p]
  \centering
  \begin{subfigure}{0.75\textwidth}
      \includegraphics[width=\textwidth]{tau-diff-a.pdf}      
  \end{subfigure}
  \begin{subfigure}{0.75\textwidth}
      \includegraphics[width=\textwidth]{tau-diff-b.pdf}      
  \end{subfigure}
   \begin{subfigure}{0.75\textwidth}
      \includegraphics[width=\textwidth]{tau-diff-c.pdf}      
  \end{subfigure}
   \caption{
    \textbf{Genotype fixations for models with increased genetic instability.} We estimated the parameters for different models, each assuming a different value of $\tau$, the fold-increase in mutation rate in aneuploid cells. We then generated 10,000 simulations using the MAP estimate of each model and evaluated the fraction of simulations in which the frequency of genotype \euwt\ (\textbf{A}), \anwt\ (\textbf{B}), and \eumt\ (\textbf{C}) is above 95\% (y-axis) at each generation (x-axis). Note that \anmt\ did not fix. We can see that $\tau=100$ can be distinguished if the waiting time for $f_{\euwt}<95\%$ is known (panel~A) or if the waiting time for $f_{\anwt}>95\%$ or $f_{\anwt}<95\%$ is known (panel~B). It is harder to distinguish between $1\leq \tau \leq10$.
  }
  \label{fig:tau-plots}
\end{figure}

% Fig likelihood profile
%% generated by sensitivity-analysis.ipynb
\begin{figure}[p]
  \centering
  \begin{subfigure}{0.3\textwidth}
      \includegraphics[width=\textwidth]{sensitivity-A.pdf}      
      \label{fig:sensitivity-mutation}
  \end{subfigure}
  \begin{subfigure}{0.3\textwidth}
      \includegraphics[width=\textwidth]{sensitivity-B.pdf}      
      \label{fig:sensitivity-aneuploidy}
  \end{subfigure}
  \\
   \begin{subfigure}{0.3\textwidth}
      \includegraphics[width=\textwidth]{sensitivity-C.pdf}      
      \label{fig:sensitivity-anwt}
  \end{subfigure}
    \begin{subfigure}{0.3\textwidth}
      \includegraphics[width=\textwidth]{sensitivity-D.pdf}      
      \label{fig:sensitivity-anmt}
  \end{subfigure}
    \begin{subfigure}{0.3\textwidth}
      \includegraphics[width=\textwidth]{sensitivity-E.pdf}      
      \label{fig:sensitivity-eumt}
  \end{subfigure}
  \caption{
    \textbf{Likelihood profiles.} Sensitivity of the model approximate likelihood, $\likelihood(\theta)$, to changing a single parameter while the other parameters remain fixed at their MAP estimates. Dashed vertical line represents the MAP value. The prior distributions for the mutation rate and aneuploidy rate are $\mu \sim U(10^{-9}, 10^{-5})$ and $\delta \sim U(10^{-6}, 10^{-2})$, respectively. 
  }

  \label{fig:sensitivity}
\end{figure}

% Fig mu comparisons
%% generated with diff-mutation-rate.ipynb
\begin{figure}[p]
  \centering
  \begin{subfigure}{0.45\textwidth}
      \includegraphics[width=\textwidth]{mu-A.pdf}      
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
      \includegraphics[width=\textwidth]{mu-B.pdf}      
  \end{subfigure}
  \\
   \begin{subfigure}{0.45\textwidth}
      \includegraphics[width=\textwidth]{mu-C.pdf}      
  \end{subfigure}
    \begin{subfigure}{0.45\textwidth}
      \includegraphics[width=\textwidth]{mu-D.pdf}      
    \end{subfigure}
  \caption{
    \textbf{Model with fixed mutation rate.}  \textbf{(A-D)} The inferred posterior distributions for models with free and fixed mutation rate, $\mu$. The MAP (maximum a posteriori) and 50\% HDI (highest density interval) for each model are: %MAP-and-hdi.ipynb
\textbf{free $\boldsymbol{\mu}$, run 1:}
$\delta=1.720\cdot10^{-3}\ [1.470\cdot10^{-3}-2.786\cdot10^{-3}]$, 
$w_{\anwt}=1.022\ [1.021-1.023]$,
$w_{\anmt}=1.025\ [1.024-1.026]$,
$w_{\eumt}=1.028\ [1.026-1.029]$;
\textbf{free $\boldsymbol{\mu}$, run 2:}
$\delta=2.129\cdot10^{-3}\ [1.334\cdot10^{-3}-2.695\cdot10^{-3}]$,
$w_{2n+1}=1.022\ [1.02-1.023]$,
$w_{2n+1^*}=1.025\ [1.023-1.026]$,
$w_{2n^*}=1.028\ [1.026-1.029]$;
\textbf{$\boldsymbol{\mu=10^{-5}}$:}
$\delta=2.903\cdot10^{-3}\ [2.399\cdot10^{-3}-3.156\cdot10^{-3}]$,
$w_{2n+1}=1.022\ [1.021-1.023]$,
$w_{2n+1^*}=1.024\ [1.023-1.025]$,
$w_{2n^*}=1.027\ [1.026-1.028]$;
\textbf{$\boldsymbol{\mu=10^{-6}}$:}
$\delta=1.917\cdot10^{-3}\ [9.624\cdot10^{-4}-2.447\cdot10^{-3}]$,
$w_{2n+1}=1.022\ [1.02-1.023]$,
$w_{2n+1^*}=1.025\ [1.023-1.026]$,
$w_{2n^*}=1.028\ [1.027-1.029]$;
\textbf{$\boldsymbol{\mu=10^{-7}}$:}
$\delta=2.901\cdot10^{-3}\ [2.139\cdot10^{-3}-3.671\cdot10^{-3}]$,
$w_{2n+1}=1.019\ [1.017-1.02]$,
$w_{2n+1^*}=1.025\ [1.024-1.026]$,
$w_{2n^*}=1.027\ [1.026-1.029]$.
\label{fig:mu}
  }
\end{figure}

% FIG posterior alt-prior
%% generated with alt-prior.ipynb
\begin{figure}[p]
  \centering
  \includegraphics[width=0.9\textwidth]{posterior-alt.pdf}
  \caption{
  \textbf{Posterior distribution of parameters inferred with the extended prior distribution.}
On the diagonal, the inferred posterior distribution of each model parameter. 
Below the diagonal, the inferred joint posterior distribution of pairs of model parameters (dark purple and bright yellow for low and high density, respectively). Red markers and orange lines for the joint MAP estimate (which may differ from the marginal MAP, as the marginal distribution integrates over all other parameters).
} 
  \label{fig:posterior-alt}
\end{figure}

% FIG MAP frequency dynamics in log scale
\begin{figure}[p]
    \centering
	\includegraphics[width=\textwidth]{MAP_freq_dynamics.pdf}
	\caption{\textbf{Posterior predicted genotype frequencies in log-log scale.}
	Frequency dynamics of the different genotypes with MAP parameter estimates, same as Figure 4A, but in log-log scale. Black dashed curves for a deterministic model without genetic drift. 
	Clearly, appearance of \anwt\ and \eumtM\ is deterministic.
	Appearance of \anmt, and therefore \eumtA, is stochastic, however, the frequency dynamics are deterministic above a frequency of roughly 0.001. 
	Note that the \eumtM\ and the \eumtA+\eumtM\ lines are overlapping for much of their trajectories.
	}
    \label{fig:MAP_freq_dynamics_log}    
\end{figure}

% FIG posterior F_A supplementary
\begin{figure}[p]
	\begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{F_A_delta-mu-ratio_1.pdf}      
	\end{subfigure}
	  \begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{F_A_delta-mu-ratio_2.pdf}
  \end{subfigure}
	\caption{\textbf{Posterior distribution of $F_A$.}
	\textbf{(A,B)} $F_A$ values (color coded) as in Figure 4C for different parameter choices on the x- and y-axes. Yellow star shows the MAP estimate.
	}
    \label{fig:posterior_FA_supp}
\end{figure}

% FIG posterior F_A with high rates
\begin{figure}[p]
      \centering
      \includegraphics[width=\textwidth]{FA_rates.pdf}      
	\caption{\textbf{Effect of genomic instability on genotype frequencies.}
	Genotype frequencies in the deterministic model without drift and \textbf{(A)} with MAP parameter estimates; \textbf{(B)} with 100-fold increase in rate of chromosome loss (transition from \anmt\ to \eumt); or \textbf{(C)} with 10-fold increase in mutation rate in aneuploid cells (transition from \anwt\ to \anmt). Corresponding $F_A$ values (purple line at generation 2,500) are 0.098, 0.223, and 0.519, respectively.
	}
    \label{fig:FA_rates}
\end{figure}

\pagebreak

%% generated with WAIC.ipynb 
\begin{table}[p]
\centering
\caption{
\textbf{WAIC values for different $\tau$ values.}
Differences of less than 6 are considered of weak significance~\citep{Kass1995}.
}
\pgfplotstabletypeset[
    col sep=semicolon,
    string type,
    every head row/.style={before row=\hline,after row=\hline},
    every last row/.style={after row=\hline},
    ]{Table_WAIC.csv} 
\caption*{
	WAIC defined in eq.~6.
} 
\label{table:WAIC}
\end{table}

\newpage
\bibliographystyle{agsm}
\bibliography{ms.bib}

\end{document}
